---
title: "BCC125_sea_lamprey_scRNA"
author: "Stephanie Hickey"
date: "`r Sys.Date()`"
output: html_document
params:
    project_dir: /mnt/ufs18/rs-013/bioinformaticsCore/projects/chong_davidson/BCC125_sea_lamprey_scRNA
    project_name: BCC125_sea_lamprey_scRNA
    lab: Weiming Li
    sampleID: TF_C00004337
    expected_cells: 20000
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

```{r, echo=FALSE, results="asis"}
## parameterized text
cat("\n")
cat("#### Lab:", params$lab, " \n")
cat("#### Project:", params$project_name, " \n")
cat("#### Sample:", params$sampleID, " \n")
```

# Set up

Load libraries
```{r}
library(tidyverse)
library(Matrix)
library(SingleCellExperiment)
library(monocle3)
library(DropletUtils)
library(scater)
library(ape)
source(paste0(params$project_dir, "/src/scRNA_seq_functions.R"))
```

Set Paths
```{r}
results_dir = paste0(params$project_dir, "/results")
# results_dir = "/mnt/ufs18/rs-013/bioinformaticsCore/projects/chong_davidson/BCC125_sea_lamprey_scRNA/results"
data_dir = paste0(params$project_dir, "/data")
# data_dir = "/mnt/ufs18/rs-013/bioinformaticsCore/projects/chong_davidson/BCC125_sea_lamprey_scRNA/data"
```

# Sample quality control

Load data
```{r}
sampleID = params$sampleID
mat = readMM(paste0(data_dir, "/", sampleID, "/raw_matrix/matrix.mtx.gz"))
barcodes = read.delim(paste0(data_dir, "/", sampleID, "/raw_matrix/barcodes.tsv.gz"), header=F)
barcodes$V1 = paste0(sampleID, "_", barcodes$V1)

genes = read.delim(paste0(data_dir,"/", sampleID, "/raw_matrix/features.tsv.gz"), header=F)  

rownames(mat) = genes$V1
colnames(mat) = barcodes$V1
```

Create `SingleCellExperiment` object
```{r}
sce <- SingleCellExperiment(assays = list(counts = mat))
```

## Detect empty droplets with `DropletUtils`

Empty droplets often contain RNA from the ambient solution, resulting in non-zero counts after debarcoding. The `emptyDrops` function is designed to distinguish between empty droplets and cells. It does so by testing each barcode’s expression profile for significant deviation from the ambient profile.

```{r}
# Set a seed. This can be any number. Using the same number every time ensures that a sequence of random numbers generated is the same every time you run your code.
set.seed(100) 

# set `by.rank` to the max number of cells you expect
e.out <- emptyDrops(counts(sce), 
                    by.rank = as.numeric(params$expected_cells)) 
e.out

# add barcodes to a column
e.out$barcode = rownames(e.out)

# make a vector of significant barcodes
# use a false discovery rate (FDR) of 0.1%,
# meaning that no more than 0.1% of our called 
# barcodes should be empty droplets on average.
sig_barcodes <- 
  as_tibble(e.out) %>%
  filter(FDR < .001) %>% 
  pull(barcode)

# how many cells?
length(sig_barcodes)

sce <- sce[,sig_barcodes]
```

A useful diagnostic for droplet-based data is the barcode rank plot, which shows the (log-)total UMI count for each barcode on the y-axis and the (log-)rank on the x-axis. This is effectively a transposed empirical cumulative density plot with log-transformed axes. It is useful as it allows users to examine the distribution of total counts across barcodes, focusing on those with the largest counts. 

```{r}
br.out <- barcodeRanks(counts(sce))
head(br.out)

# make the rownames a column
br.out$barcodes <- rownames(br.out)
head(br.out)

br.df =
  as.data.frame(br.out) %>%
  mutate(keep_cells = case_when(
   barcodes %in% sig_barcodes ~ "TRUE",
   !barcodes %in% sig_barcodes ~ "FALSE"
  ))

# Making the knee plot.
ggplot(br.df, aes(x=rank, y=total)) +
  geom_point(aes(color = keep_cells)) +
  scale_colour_manual(values = c("black","red")) +
  geom_hline(aes(yintercept=metadata(br.out)$knee,
             linetype="knee"),
             color = "dodgerblue") +
  geom_hline(aes(yintercept=metadata(br.out)$inflection,
             linetype="inflection"),
             color = "forestgreen") +
  scale_linetype_manual(name = "legend", 
                        values = c(2, 2), 
                        guide = guide_legend(override.aes = list(color = c("forestgreen", "dodgerblue")))) +
  scale_x_continuous(trans='log10') +
  scale_y_continuous(trans='log10') +
  xlab("Rank") +
  ylab("Total Counts")

ggsave(file = paste0(results_dir,"/", samp, "/kneeplot.png"))
```

## Identifying low quality cells

Common metrics include for identifying low quality cells include:

  * Total counts    
  * Number of expressed features (genes)   
  * Proportion of reads mapped to genes in the mitochondrial genome    
  
For each cell, we calculate these QC metrics using the `perCellQCMetrics()` function from the `scater` package.
```{r}
# Identifying the mitochondrial transcripts in our SingleCellExperiment.

# mouse mito genes start with "mt"
# human is "MT"
# change for different species
# this is a numeric vector of row numbers
# corresponding to mito genes
# NC_001626.1
all_genes = read.delim("/mnt/research/bioinformaticsCore/shared/Genomes/Petromyzon_marinus/GCF_010993605.1_kPetMar1.pri_genes.txt", header=F)

mito_genes = 
  all_genes %>% 
  filter(V1 == "NC_001626.1") %>%
  pull(V6)

mito_genes = gsub("gene ", "", mito_genes)

is.mito = which(rownames(sce) %in% mito_genes)

# show the gene names
rownames(sce)[is.mito]

# get the per cell stats
stats <- perCellQCMetrics(sce,
                          subsets=list(Mito=is.mito))

head(stats)
```

'sum' is the the sum of counts for each cell. 'detected' is the number of genes with at least one count in each cell. 'subsets_Mito_percent' is the percent of the total counts ('sum') that come from mitochrondria genes: 'subsets_Mito_sum'.

### Summarize the stats

```{r}
summary(stats$sum)

summary(stats$detected)

summary(stats$subsets_Mito_percent)
```

Normally, I would filter out cells with high percent mitochondria, but female germ cells likely have biologically high percent mitochondria. So I will cluster first and attempt to identify female germ cells before performing any QC filtering.

# Normalization and transformation

```{r}
counts = counts(sce)
cds <- new_cell_data_set(counts,
                         cell_metadata = stats)

cds <- preprocess_cds(cds, num_dim = 100)
plot_pc_variance_explained(cds)
```
# Clustering with UMAP
```{r}
cds <- reduce_dimension(cds)
```

## Plot percent mitochodria reads
```{r}
plot_cells(cds, color_cells_by="subsets_Mito_percent")
```
```{r}
cds <- cluster_cells(cds, resolution=1e-5)
plot_cells(cds)
ggsave(file = paste0(results_dir, "/", sampleID, "_cluster_UMAP.png"))
```

## Check for enrichment of genes of interest in each cluster

```{r}
marker_test_res <- top_markers(cds, 
                               genes_to_test_per_group = 100,
                               group_cells_by="cluster", cores=8)

write.csv(marker_test_res, file = paste0(results_dir, "/", sampleID, "_cluster_marker_genes.csv"), row.names = FALSE) 

# load known marker genes
library(readr)
chung_davidson_cell_type_markers <- read_csv(paste0(data_dir, "/chung_davidson_cell_type_markers.csv"))
chung_davidson_cell_type_markers = 
  chung_davidson_cell_type_markers %>%
  select(-GeneName)

marker_test_genes = 
  marker_test_res %>%
  select(gene_id, cell_group)

cluster_ol =  phyperOverlapSets(chung_davidson_cell_type_markers, 
                                marker_test_genes, nrow(genes))

write.csv(cluster_ol$intersecting.genes, file = paste0(results_dir, "/", sampleID, "_marker_gene_overlaps.csv"), row.names = FALSE) 

scores = cluster_ol$pval.scores

scores = 
  scores %>%
  rename(Group1 = "CellTypes",
         Group2 = "Clusters") %>%
  group_by(CellTypes) %>%
    mutate(FDR = p.adjust(Pval, "BH"))

scores$FDR[scores$FDR > .01] = 1

plot1 =
    ggplot(scores,
           aes(Clusters, CellTypes)) + 
           geom_tile(aes(fill = -log10(FDR)), 
                     color = "light gray") +
        #Creating legend
        guides(fill=guide_colorbar(title = "-log10(FDR)",
                                   barheight = 10,
                                   barwidth = 2,
                                   default.unit = "mm")) +
        #Creating color range
        scale_fill_gradientn(colors=c("white",
                                      "#fd9668",
                                      "#f1605d",
                                      "#de4968",
                                      "#9e2f7f"),
                             guide="colorbar",
                             limits = c(0, 23)) +
        #Rotating labels
        theme(axis.text.x = element_text(angle = 30, 
                                         hjust = 0,vjust=-0.05),
              ) +
        theme(text = element_text(size = 7),
              axis.text = element_text(size = 7)) +
        geom_text(aes(label = nSharedGenes), size = 1) +
        coord_fixed()

plot1

ggsave(plot1, file = paste0(results_dir, "/", sampleID, "_overlap_heatmap.png"))

```

## Plot Mitochondira gene percent per cluster
```{r}

colData(cds)$cluster = clusters(cds)

mito_vln = 
  ggplot(colData(cds), aes(y = subsets_Mito_percent, x = cluster)) +
  geom_violin()

mito_vln
ggsave(mito_vln, file = paste0(results_dir, "/", sampleID, "_percent_mito_vln.png"))
```
# get rid of high mito cells that are not of interest