---
title: "01_all_cells_clustering_annotation_no_doublets"
author: "Stephanie Hickey"
date: "`r Sys.Date()`"
output:
  html_document:
    code_folding: hide
    toc: true
    toc_float:
      toc_collapsed: true
params:
    project_dir: /mnt/ufs18/rs-013/bioinformaticsCore/projects/chong_davidson/BCC125_sea_lamprey_scRNA
    project_name: BCC125_sea_lamprey_scRNA
    lab: Weiming Li
    dbr: 0.08
    dbr_sd: 0.02
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

```{r, echo=FALSE, results="asis"}
cat("\n")
cat("#### Lab:", params$lab, " \n")
cat("#### Project:", params$project_name, " \n")
```

```{r}
library(tidyverse)
library(SingleCellExperiment)
library(scater)
library(scran)
library(patchwork)
library(RColorBrewer)
library(readxl)
library(monocle3)
library(batchelor)
library(DT)
library(rtracklayer)
library(BiocSingular)
library(ggnewscale)
library(scDblFinder)
source(paste0(params$project_dir, "/src/scRNA_seq_functions.R"))
```

This is a copy of `01_all_cells_clustering_annotation.Rmd` that removes cells
called doublets by `scDblFinder` before clustering, integration or trajectory
analysis, so the effect of removing them on clustering and annotation can be
assessed directly. `scDblFinder` is rerun here, per sample, rather than
reusing scores from `04_doublet_qc.Rmd`, so this document is standalone and
does not depend on that one having been run first. Because `scDblFinder`'s
doublet simulation is stochastic, the doublet calls made here are not
guaranteed to be identical to the ones reported in `04_doublet_qc.Rmd`, even
with the same seed, if package versions differ between runs.

```{r}
results_dir = paste0(params$project_dir, "/results/20260813_no_doublets")
if(!dir.exists(results_dir)){dir.create(results_dir)}
data_dir    = paste0(params$project_dir, "/data")

# Each expensive object is computed when it is absent and read back when it is
# present, so the first knit against a new results_dir builds everything and a
# reknit is cheap. The previous arrangement, eval=FALSE compute chunks feeding
# an unguarded load chunk, only worked while results_dir pointed at a directory
# some earlier interactive session had already filled.
all_sce_file   = paste0(results_dir, "/all.sce.Rds")
merged_file    = paste0(results_dir, "/merged.sce.Rds")
combined_file  = paste0(results_dir, "/combined.sce.Rds")
cds_file       = paste0(results_dir, "/merged.cds.Rds")
dbl_stats_file = paste0(results_dir, "/doublet_removal_stats.csv")

build_all_sce  = !file.exists(all_sce_file)
build_merged   = !file.exists(merged_file)
build_combined = !(file.exists(combined_file) && file.exists(cds_file))

# the doublet rate is a parameter, so a results directory built at one rate must
# not be silently reused at another
rate_stamp = paste0(results_dir, "/doublet_rate_used.txt")
if (build_all_sce) {
  writeLines(paste0("dbr=", params$dbr, " dbr_sd=", params$dbr_sd), rate_stamp)
} else if (file.exists(rate_stamp)) {
  prev = readLines(rate_stamp)[1]
  now  = paste0("dbr=", params$dbr, " dbr_sd=", params$dbr_sd)
  if (!identical(prev, now))
    stop("all.sce.Rds in ", results_dir, " was built at ", prev,
         " but this knit specifies ", now,
         ". Point results_dir at a new directory or delete the cached objects.")
}
```

# Collect individual samples into one object

```{r, eval=build_all_sce}
sample_dirs  = list.dirs(data_dir, full.names = TRUE, recursive = FALSE)
sample_dirs = sample_dirs[grep("smith_annotation", sample_dirs, invert = TRUE)]
sample_names = basename(sample_dirs)

cds_list = list()
for (i in 1:length(sample_dirs)) {
  cds_list[[i]] = readRDS(paste0(sample_dirs[i], "/", sample_names[i],
                                  "_cds_mito_filter.Rds"))
}
names(cds_list) = sample_names

all.sce = list()
for (cds in names(cds_list)) {
  raw_counts          = counts(cds_list[[cds]])
  all.sce[[cds]]      = SingleCellExperiment(list(counts = raw_counts),
                                              colData = cds_list[[cds]]@colData)
  all.sce[[cds]]@colData$sample = cds
}
```

## Remove cells called doublets

Doublets are removed here, per sample, before any normalization, HVG selection or PCA, and before the samples are merged.

## Expected doublet rate

`scDblFinder` documents `dbr` as *"The expected doublet rate. By default this is
assumed to be 1% per thousand cells captured (so 4% among 4000 thousand cells),
which is appropriate for 10x datasets."* At roughly 16,700 cells per library
that default expects about 16.7% doublets.

These libraries are PIPseq T20 3' v4.0. Cell concentration was adjusted to
4 x 10^6 cells/ml and 10 ul loaded, which is 40,000 cells, the standard
protocol. The kit user guide states that this gives recovery above 20,000 cells
and a multiplet rate below 8%.

`dbr` places the cutoff on the doublet score, so expecting roughly twice the
rate the chemistry produces pushes the cutoff down and flags more cells than
warranted. The rate is therefore set from the platform specification. Under 8%
is an upper bound rather than a point estimate, so `r params$dbr` is
conservative and `dbr.sd` sets the tolerance; both are document parameters.


```{r, eval=build_all_sce}
set.seed(100)
all.sce.dbl = lapply(all.sce, function(x)
  scDblFinder(x, dbr = params$dbr, dbr.sd = params$dbr_sd))

n_before = sapply(all.sce, ncol)
n_dbl    = sapply(all.sce.dbl, function(x) sum(x$scDblFinder.class == "doublet"))

for (cds in names(all.sce)) {
  keep = colData(all.sce.dbl[[cds]])$scDblFinder.class == "singlet"
  all.sce[[cds]] = all.sce[[cds]][, keep]
}

n_after = sapply(all.sce, ncol)

# written out rather than only printed, so the rate actually achieved is still
# reportable on a reknit that loads the cached objects
write.csv(data.frame(sample = names(n_before), n_before = n_before,
                     n_doublet = n_dbl, n_after = n_after,
                     pct_doublet = round(100 * n_dbl / n_before, 2),
                     row.names = NULL),
          file = dbl_stats_file, row.names = FALSE)
```

```{r, eval=build_all_sce}
all.sce  <- lapply(all.sce, logNormCounts)
all.dec  <- lapply(all.sce, modelGeneVar)
all.hvgs <- lapply(all.dec, getTopHVGs, prop = 0.1)
all.sce  <- mapply(FUN = runPCA, x = all.sce, subset_row = all.hvgs)
saveRDS(all.sce, file = all_sce_file)
```

```{r}
if (!build_all_sce) all.sce = readRDS(all_sce_file)
```

Cells removed as doublets, and the rate achieved at `dbr` = `r params$dbr`.

```{r}
datatable(read.csv(dbl_stats_file), rownames = FALSE,
          options = list(dom = "t", ordering = FALSE))
```

# Sample summary table

```{r}
df_list = list()
for (samp in names(all.sce)) {
  ncells     = ncol(all.sce[[samp]])
  med_counts = round(median(colData(all.sce[[samp]])$sum))
  med_genes  = round(median(colData(all.sce[[samp]])$detected))
  med_mito   = round(median(colData(all.sce[[samp]])$subsets_Mito_percent), digits = 2)

  df_list[[samp]] = data.frame(Sample                 = samp,
                                nCells                 = ncells,
                                medianCounts           = med_counts,
                                medianGenesDetected    = med_genes,
                                medianPercentMitoReads = med_mito)
}

df_all = do.call(rbind, df_list)
write.csv(df_all, file = paste0(results_dir, "/median_stats.csv"))
datatable(df_all, rownames = FALSE)
```

# Merge samples, removing batch effect

```{r}
if (build_merged) {
  # recomputed here rather than carried from the chunk above, which does not run
  # when all.sce is loaded from cache
  all.dec      <- lapply(all.sce, modelGeneVar)
  universe     <- Reduce(intersect, lapply(all.sce, rownames))
  all.sce2     <- lapply(all.sce, "[", i = universe,)
  all.dec2     <- lapply(all.dec, "[", i = universe,)
  normed.sce   <- do.call(multiBatchNorm, all.sce2)
  combined.dec <- do.call(combineVar, all.dec2)
  combined.hvg <- getTopHVGs(combined.dec, n = 5000)

  set.seed(1)
  merged.sce <- do.call(fastMNN, c(normed.sce,
      list(subset.row = combined.hvg, BSPARAM = RandomParam())))

  set.seed(1)
  g <- buildSNNGraph(merged.sce, use.dimred = "corrected", k = 30)
  set.seed(1)
  colLabels(merged.sce) <- factor(igraph::cluster_louvain(g)$membership)

  set.seed(1)
  merged.sce <- runUMAP(merged.sce, dimred = "corrected")
  set.seed(1)
  merged.sce <- runTSNE(merged.sce, dimred = "corrected")

  saveRDS(merged.sce, file = merged_file)
} else {
  merged.sce <- readRDS(merged_file)
}
```

The plot below shows the merged data in TSNE space separated by sample.

```{r, fig.width=8, fig.height=8}
p <- lapply(unique(merged.sce$batch), function(f) {
  sce.f = merged.sce[, which(merged.sce$batch == f)]
  plotTSNE(sce.f, point_size = .05) + ggtitle(f)
})
wrap_plots(p, ncol = 2)
```

Clusters identified in batch-corrected expression space.

```{r, fig.width=6, fig.height=6}
plotTSNE(merged.sce, colour_by = "label", text_by = "label",
         text_colour = "red", point_size = .05) +
  guides(color = guide_legend(override.aes = list(size = 5)))

ggsave(file = paste0(results_dir, "/all_cells_tsne.png"))
```

Proportion of cells from each sample within each cluster.

```{r}
meta_df = as.data.frame(colData(merged.sce))

sample_tally = meta_df %>%
  group_by(label, batch) %>%
  tally()

ggplot(sample_tally, aes(fill = batch, y = label, x = n)) +
  geom_bar(position = "fill", stat = "identity") +
  xlab("Proportion of cells")

ggsave(file = paste0(results_dir, "/cells_per_cluster_per_sample.png"))
```

The same data as cell counts rather than proportions, saved in
`cells_per_cluster_per_sample.csv`.

```{r}
sample_tally_wide =
  sample_tally %>%
  tidyr::pivot_wider(names_from = batch, values_from = n, values_fill = 0) %>%
  ungroup() %>%
  mutate(label = as.character(label))

sample_tally_wide$Total = rowSums(sample_tally_wide[, -1])

sample_tally_wide =
  bind_rows(sample_tally_wide,
            sample_tally_wide %>%
              summarize(across(where(is.numeric), sum)) %>%
              mutate(label = "Total"))

write.csv(sample_tally_wide,
          file = paste0(results_dir, "/cells_per_cluster_per_sample.csv"),
          row.names = FALSE)

datatable(sample_tally_wide, rownames = FALSE)
```

```{r}
# build uncorrected combined SCE and monocle CDS for downstream plotting
if (build_combined) {
  all.dec  <- lapply(all.sce, modelGeneVar)
  universe <- Reduce(intersect, lapply(all.sce, rownames))
  all.sce2 <- lapply(all.sce, "[", i = universe,)
  all.dec2 <- lapply(all.dec, "[", i = universe,)

  all.sce2 <- lapply(all.sce2, function(x) {
    colData(x)    <- colData(x)[, "sample", drop = FALSE]
    reducedDims(x) <- list()
    rowData(x)    <- rowData(x)[, character(0)]
    x
  })

  combined.sce <- do.call(cbind, all.sce2)
  saveRDS(combined.sce, file = combined_file)

  merged.cds <- new_cell_data_set(counts(combined.sce),
                                  cell_metadata = merged.sce@colData)
  reducedDims(merged.cds)$UMAP <- reducedDims(merged.sce)$UMAP
  reducedDims(merged.cds)$tSNE <- reducedDims(merged.sce)$TSNE
  rowData(merged.cds)$gene_short_name <- row.names(merged.cds)

  saveRDS(merged.cds, file = cds_file)
} else {
  combined.sce <- readRDS(combined_file)
  merged.cds   <- readRDS(cds_file)
}

# the clustering and the counts must describe the same cells in the same order
stopifnot(identical(colnames(combined.sce), colnames(merged.sce)))
```

## Gene descriptions from Wang et al. and the kPetMar1.pri gtf

Summarized from [Wang et al, 2024](https://www.nature.com/articles/s41467-024-55019-x#Sec16):
LOC IDs are re-annotated to human gene symbols via BLAST; genes with human matches
are UPPERCASE, others lowercase.

```{r}
gene_description =
  read_excel(paste0(data_dir, "/41467_2024_55019_MOESM3_ESM.xlsx"), skip = 1) %>%
  dplyr::rename(WangGeneID = "CountMatrix_GeneID",
                Gene        = "NCBI_GTF_GENEID",
                description = "NCBI_GTF_ProteinInfo") %>%
  dplyr::select(Gene, WangGeneID, description) %>%
  filter(!is.na(Gene)) %>%
  filter(!WangGeneID %in% c("LOC116945425.1", "S100A1.1"))

gtf = as.data.frame(readGFF("/mnt/research/bioinformaticsCore/shared/Genomes/Petromyzon_marinus/GCF_010993605.1_kPetMar1.pri_genomic.gtf"))

gtf_filt =
  gtf %>%
  filter(gbkey %in% c("mRNA", "misc_RNA", "ncRNA", "tRNA", "rRNA")) %>%
  mutate(product = gsub(", transcript variant.*", "", product)) %>%
  dplyr::select(gene_id, product) %>%
  distinct() %>%
  dplyr::rename(Gene = "gene_id", description = "product") %>%
  mutate(WangGeneID = Gene)

notWang      = gtf_filt$Gene[!gtf_filt$Gene %in% gene_description$Gene]
gtf_add      = gtf_filt %>% filter(Gene %in% notWang)
gene_description = rbind(gene_description, gtf_add)

write.csv(gene_description,
          file = paste0(results_dir, "/gene_descritption.csv"), row.names = FALSE)
datatable(gene_description)
```

## Known cell-type markers

The curated list is `canonical_markers_long_2026-08-14.csv`, in long format: one
row per gene per cell type, with the NCBI identifier already resolved. It is the
collaborators' S1 sheet with the identifiers repaired, plus additions only where
a source was retrieved for the set.

Every marker is one of three things, recorded in `provenance`:

* **S1**, taken unchanged from `S1(Cell_Type_Marker_Genes_12-12-25).csv`.
* **S1, symbol corrected**, where S1's spelling resolves against nothing in the
  gene table and a corrected spelling resolves. Three cases, each named in
  `repair_note`.
* **added, set cited**, an addition to a set for which a citation was retrieved
  and checked. Sixteen genes across three sets.

Nine genes present in the 2026-08-12 draft were added without a retrieved source
and have been removed. A confidence tier column was also dropped: it graded 40
genes as having direct support in a non-mammalian vertebrate while the
supporting text named one for six of them, so it asserted evidence that had not
been gathered. The per-gene descriptive notes went with it for the same reason.
What remains is either from the collaborators or checked against a source.

```{r}
canonical = read.csv(paste0(data_dir, "/canonical_markers_long_2026-08-14.csv"),
                     stringsAsFactors = FALSE) %>%
  mutate(symbol = trimws(symbol), loc_id = trimws(loc_id)) %>%
  filter(symbol != "")

# loc_id is the NCBI identifier and is populated for every row, so it is used
# directly. The symbol is kept for labelling.
chung_davidson_markers = canonical %>%
  dplyr::select(CellType = CellTypeMarkerSet, Gene = loc_id,
                WangGeneID = symbol) %>%
  distinct()

n_missing = sum(!chung_davidson_markers$Gene %in% rownames(combined.sce))
```

### Sources

Citations attach to the marker **set**, which is the level at which they were
retrieved, and cover three of the fourteen sets. There is no per-gene citation
and none is invented here. Each reference was checked against PubMed by author,
year, journal and volume; the two without a PMID could not be confirmed that way
and say so.

```{r}
datatable(canonical %>%
            filter(sources != "") %>%
            dplyr::count(CellTypeMarkerSet, sources, name = "n_markers") %>%
            dplyr::select(CellTypeMarkerSet, n_markers, sources),
          rownames = FALSE, options = list(dom = "t", ordering = FALSE))
```

### The marker list

```{r}
datatable(canonical %>%
            dplyr::select(CellTypeMarkerSet, symbol, loc_id, provenance,
                          repair_note) %>%
            arrange(CellTypeMarkerSet, symbol),
          filter = "top", rownames = FALSE)
```

```{r}
set_sizes = canonical %>%
  dplyr::count(CellTypeMarkerSet, name = "n_markers") %>%
  mutate(below_coherence_floor = n_markers < 3)

datatable(set_sizes, rownames = FALSE, options = list(dom = "t"))

thin_sets = set_sizes %>% filter(below_coherence_floor)
```

`r if (nrow(thin_sets) == 0) "Every set has at least three markers." else paste0("**", nrow(thin_sets), " set(s) hold fewer than three markers: ", paste(thin_sets$CellTypeMarkerSet, collapse = ", "), ".** A set that small cannot pass a coherence threshold, so these report NA rather than a verdict wherever coherence is scored. That is what S1 provides for those cell types once unsourced additions are excluded, not a loss created by excluding them.")`

`r paste0(nrow(chung_davidson_markers), " marker genes across ", dplyr::n_distinct(chung_davidson_markers$CellType), " cell types. ", if (n_missing == 0) "All resolve against the count matrix." else paste0(n_missing, " do not appear in the count matrix and are dropped from the plots below."))`

## Marker expression across all clusters

Every curated marker against every cluster, on uncorrected depth-normalised
values. Dot size is the percentage of cells in the cluster expressing the gene,
colour is the mean log expression.

This is the plot to annotate from. The overlap heatmap further down measures
whether a cell type's markers are *enriched* among a cluster's strongest
data-derived markers, which is a different question: a gene can be enriched in a
cluster relative to the others while barely being transcribed there.

```{r}
marker_ids = intersect(unique(chung_davidson_markers$Gene),
                       rownames(combined.sce))

bubble_mat = as.matrix(logcounts(combined.sce)[marker_ids, , drop = FALSE])
bubble_lab = as.character(colData(merged.sce)$label)

marker_bubble = purrr::map_dfr(sort(unique(as.integer(bubble_lab))), function(k) {
  sub = bubble_mat[, bubble_lab == as.character(k), drop = FALSE]
  data.frame(cluster        = k,
             Gene           = marker_ids,
             n_cells        = ncol(sub),
             pct_expressing = 100 * rowMeans(sub > 0),
             mean_logexpr   = rowMeans(sub),
             stringsAsFactors = FALSE)
}) %>%
  left_join(chung_davidson_markers, by = "Gene",
            relationship = "many-to-many") %>%
  mutate(CellType = factor(CellType),
         cluster  = factor(cluster, levels = sort(unique(cluster))))

write.csv(marker_bubble,
          file = paste0(results_dir, "/canonical_marker_expression_per_cluster.csv"),
          row.names = FALSE)
```

```{r, fig.width=11, fig.height=13}
p_bubble = ggplot(marker_bubble,
                  aes(x = cluster, y = WangGeneID,
                      size = pct_expressing, colour = mean_logexpr)) +
  geom_point() +
  scale_colour_viridis_c() +
  scale_size_continuous(range = c(0.4, 5)) +
  facet_grid(CellType ~ ., scales = "free_y", space = "free_y") +
  theme(strip.text.y = element_text(angle = 0, size = 6),
        axis.text.y  = element_text(size = 6),
        axis.text.x  = element_text(size = 8),
        panel.grid.major = element_line(linewidth = .2)) +
  xlab("cluster") + ylab(NULL) +
  labs(size = "% expressing", colour = "mean logexpr")

p_bubble
ggsave(p_bubble, file = paste0(results_dir, "/canonical_marker_bubble_all_clusters.png"),
       width = 11, height = 13)
```

## Cluster enriched genes

Top ~100 cluster enriched genes per cluster. All significant genes (FDR < 0.01) are
saved in `scatter_cluster_marker_genes.csv`.

**Columns:** `cluster`, `Gene`, `p.value`, `FDR`, `summary.logFC`

```{r}
scatter_markers = findMarkers(combined.sce,
                               merged.sce$label,
                               test.type = "t",
                               block     = combined.sce$sample,
                               direction = "up",
                               pval.type = "some")

saveRDS(scatter_markers, file = paste0(results_dir, "/scatter_markers.Rds"))

scatter_summary = list()
for (i in 1:length(scatter_markers)) {
  scatter_markers[[i]]$cluster = i
  scatter_markers[[i]]$Gene    = rownames(scatter_markers[[i]])
  scatter_summary[[i]] = as.data.frame(scatter_markers[[i]]) %>%
    select(cluster, Gene, p.value, FDR, summary.logFC)
}

scatter_markers_df =
  do.call(rbind, scatter_summary) %>%
  left_join(gene_description, by = "Gene") %>%
  filter(FDR < .01)

write.csv(scatter_markers_df,
          file = paste0(results_dir, "/scatter_cluster_marker_genes.csv"))

scatter_markers_top100 = scatter_markers_df %>%
  group_by(cluster) %>%
  slice_min(FDR, n = 100, with_ties = TRUE)

datatable(scatter_markers_top100, filter = "top", rownames = FALSE)
```

## Comparison with known markers

Rows = cell types with known markers; columns = clusters. Color = $-log10(FDR)$ of
overlap significance. Number in each box = shared genes.

- Overlap p-values: `scatter_cluster_marker_overlap_pvalues.csv`
- Overlapping genes: `scatter_cluster_marker_gene_overlaps.csv`

```{r, fig.width=8}
marker_test_genes =
  scatter_markers_top100 %>%
  select(Gene, cluster) %>%
  mutate(cluster = paste0("cluster", cluster))

chung_davidson_cell_type_markers = chung_davidson_markers %>%
  select(Gene, CellType)

plot_marker_overlap_heatmap(
  known_markers  = chung_davidson_cell_type_markers,
  test_genes     = marker_test_genes,
  background_n   = length(merged.sce),
  gene_description = gene_description,
  overlaps_file  = paste0(results_dir, "/scatter_cluster_marker_gene_overlaps.csv"),
  pvals_file     = paste0(results_dir, "/scatter_cluster_marker_overlap_pvalues.csv"),
  plot_file      = paste0(results_dir, "/scatter_cluster_overlap_heatmap.png")
)
```

## Gene ontology of top 100 cluster enriched genes

**Columns:** `GOID`, `GOName`, `nGOTermGenes`, `Clusters`, `nClusterEnrichedGenes`,
`nSharedGenes`, `Pval`, `FDR`

- Overlap p-values: `scatter_cluster_marker_gene_GO_overlap_pvalues.csv`
- Overlapping genes: `scatter_cluster_marker_gene_GO_overlaps.csv`

```{r}
gaf = read.delim("/mnt/research/bioinformaticsCore/shared/Genomes/Petromyzon_marinus/GCF_010993605.1_kPetMar1.pri_gene_ontology.gaf",
                 header = FALSE, comment = "!")

gaf <- gaf %>%
  filter(!grepl("NOT", V4)) %>%
  dplyr::select(V3, V5) %>%
  dplyr::rename(GOID = "V5", Gene = "V3") %>%
  distinct()

keep_go = gaf %>%
  group_by(GOID) %>% tally() %>%
  filter(n > 10, n < 200) %>%
  pull(GOID)

gaf = gaf %>% filter(GOID %in% keep_go)

go_names = read.delim("/mnt/research/bioinformaticsCore/shared/Genomes/Petromyzon_marinus/GOs_names.txt",
                      header = FALSE)
colnames(go_names) = c("GOID", "GOName")

scores = run_go_enrichment(
  gaf             = gaf,
  gene_description = gene_description,
  go_names        = go_names,
  test_genes      = marker_test_genes,
  background_n    = length(merged.cds),
  overlaps_file   = paste0(results_dir, "/scatter_cluster_marker_gene_GO_overlaps.csv"),
  pvals_file      = paste0(results_dir, "/scatter_cluster_marker_gene_GO_overlap_pvalues.csv"),
  pval_filter     = 0.01
)

datatable(scores, filter = "top", rownames = FALSE)
```

## Plot known cell-type markers

Early germ cell markers.

```{r}
rowData(merged.cds)$gene_short_name = row.names(merged.cds)

# a gene assigned to two sets would be faceted ambiguously, so only genes
# belonging to exactly one set are plotted
unique_markers = chung_davidson_markers %>%
  group_by(Gene) %>% tally() %>%
  filter(n == 1) %>%
  pull(Gene)

genes_for_plot = chung_davidson_markers %>%
  select(Gene, CellType) %>%
  filter(Gene %in% unique_markers, Gene %in% rownames(merged.cds))

# Set names are spelled out here rather than inline in each plot_cells call, so
# they can be checked against the curated list before anything is drawn. The
# previous version carried the wide sheet's syntactic names (Pre.granulosa.cells
# and so on); one panel matched nothing and errored, and two matched a single
# set each and would have rendered a partial figure under a heading promising
# all of them.
marker_panels = list(
  early_germ   = "Primordial germ cells",
  pan_germline = "Germ cells (pan-germline)",
  female       = c("Oocyte", "Granulosa cells", "Pre-granulosa cells"),
  male         = c("Spermatocytes", "Spermatogonial stem cells"))

known_sets = unique(chung_davidson_markers$CellType)
unknown    = setdiff(unlist(marker_panels, use.names = FALSE), known_sets)
if (length(unknown) > 0)
  stop("marker set(s) requested for plotting but absent from the curated list: ",
       paste(unknown, collapse = ", "),
       ". Sets available: ", paste(sort(known_sets), collapse = ", "))

# plot_cells errors with "None of the provided genes were found in the cds" when
# handed an empty set, which does not say which panel or why
plot_marker_panel = function(sets, ...) {
  g = genes_for_plot %>% filter(CellType %in% sets)
  if (nrow(g) == 0)
    stop("no plottable genes for ", paste(sets, collapse = " / "),
         ". Every marker in these sets is either shared with another set or ",
         "absent from the count matrix.")
  plot_cells(merged.cds, reduction_method = "tSNE", genes = g, ...)
}

panel_counts = purrr::map_dfr(names(marker_panels), function(p)
  data.frame(panel     = p,
             sets      = paste(marker_panels[[p]], collapse = "; "),
             n_genes   = sum(genes_for_plot$CellType %in% marker_panels[[p]])))

datatable(panel_counts, rownames = FALSE,
          options = list(dom = "t", ordering = FALSE))
```

```{r, fig.width=8, fig.height=4}
plot_marker_panel(marker_panels$early_germ)
```

Pan-germline markers.

```{r, fig.width=12, fig.height=6}
plot_marker_panel(marker_panels$pan_germline,
                  show_trajectory_graph = FALSE,
                  label_cell_groups     = FALSE,
                  label_leaves          = FALSE,
                  label_branch_points   = FALSE)
```

Female cell markers.

```{r, fig.height=4, fig.width=12}
plot_marker_panel(marker_panels$female,
                  show_trajectory_graph = FALSE,
                  label_cell_groups     = FALSE,
                  label_leaves          = FALSE,
                  label_branch_points   = FALSE)
```

Male cell markers.

```{r, fig.height=4, fig.width=12}
plot_marker_panel(marker_panels$male,
                  show_trajectory_graph = FALSE,
                  label_cell_groups     = FALSE,
                  label_leaves          = FALSE,
                  label_branch_points   = FALSE)
```

# Annotations, parked

**Nothing in this section runs.** Every chunk below is `eval=FALSE` and the
document produces no cell type labels, no annotated TSNE and no bubble plots
until the annotation is settled with the collaborator.

**Everything below describes the `20260803` clustering, not this run.** The
labels, the cluster numbers, the correspondence table and the cell counts were
all derived from the earlier run at `scDblFinder`'s 10x-calibrated default rate.
Louvain renumbers on every clustering, so none of it transfers on its own. It is
kept here as the record of how the previous annotation was reached, and as the
starting point for redoing it, not as a description of the clusters this
document just produced.

To revive it: rerun `05_compare_doublet_removal.Rmd` against this run, rebuild
the correspondence from `cluster_contingency_table.csv` regrouped on `label_B`,
settle the labels with the collaborator, then set these chunks back to `eval=TRUE`.

## What the previous annotation was, and how it was reached

Cluster numbering did not match `01_all_cells_clustering_annotation.Rmd`.
Doublet removal yielded 20 clusters rather than 19, and the numbering was
reassigned by Louvain, so the labels below were re-derived rather than copied.

Each was assigned from two independent criteria computed in
`05_compare_doublet_removal.Rmd`: which with-doublets cluster contributed the
most cells, and which with-doublets cluster shares the most top-100 marker
genes. The two criteria agreed for 19 of the 20 clusters. Cluster 13 was the
exception and is discussed below.

Three labels differed from the with-doublets analysis in substance, not just in
number:

  * **Cluster 11 and cluster 12** were both hemopoietic stem cells. The single
    HSC cluster in the with-doublets analysis divided roughly evenly (1,302 and
    1,224 cells), each essentially pure. What distinguishes them was not
    established.
  * **Cluster 7** was podocyte-dominated by marker overlap (49 shared genes
    against 37 for the next-best match), but it also absorbed 321 cells from
    the cluster previously annotated "Primordial germ cells (migrating)". It is
    a genuine mixture and was labeled as such.
  * **Cluster 13** held 72% of the cells from that same previously-annotated
    PGC cluster, but shared almost no markers with it (9 genes, against 40 with
    the erythrocyte cluster). Its own top markers were globins, ferritin and
    PCLAF, it drew cells from 12 different with-doublets clusters, and it had
    the lowest purity of any cluster. It carried no germline markers at all. It
    was labeled unresolved rather than given a cell type.

No cluster in that run carried an early germline program. Canonical PGC markers
(PRDM14, DND1, TFAP2C, SOX17, NANOG) were absent or negligible throughout.
Germline identity was confined to cluster 8 (ZP3 and ZP4 family, logFC up to
3.9) and clusters 17 to 20 (PIWIL1, SYCP3, SYCP2, BOLL, TDRD7, TDRD9, MAEL,
TEX15). Whether that still holds is a question for the rerun, not something
this section establishes.

```{r, eval=FALSE}
anno = c(
  "1. Fibroblasts (migrating)",
  "2. Mesonephric cells",
  "3. Granulosa cells (migrating)",
  "4. Granulosa cells",
  "5. Pre-granulosa cells (migrating)",
  "6. Endothelial cells",
  "7. Podocytes (mixed)",
  "8. Oogonia/Oocytes",
  "9. Erythrocytes",
  "10. Leukocytes",
  "11. Hemopoietic stem cells (1)",
  "12. Hemopoietic stem cells (2)",
  "13. Unresolved (globin-high, mixed origin)",
  "14. Macrophages",
  "15. Pre-granulosa cells (mitotic)",
  "16. Immune cells (migrating, mitotic)",
  "17. Spermatocytes (cytokinetic)",
  "18. Male germ cells (migrating)",
  "19. Spermatocytes (apoptotic)",
  "20. Spermatocytes (GPCR-activated)"
)

# These names were assigned to the 20260803 clustering, using the two criteria
# described above as computed in 05_compare_doublet_removal.Rmd. Louvain
# renumbers on every clustering, so they do not transfer to a rerun on their
# own. A count mismatch is caught here; a renumbering that still yields 20
# clusters is not, and the correspondence table below would then be asserting
# matches that were never computed for this run. Rerun 05 and re-derive both
# before trusting the labels.
n_clusters = nlevels(colData(merged.sce)$label)
if (length(anno) != n_clusters)
  stop("anno has ", length(anno), " labels but this clustering has ",
       n_clusters, " clusters. Re-derive the annotations from ",
       "05_compare_doublet_removal.Rmd before relabeling.")
```

| no-doublets cluster | best match by cells | purity | annotation |
| --- | --- | --- | --- |
| 1 | 1 | 99.7% | Fibroblasts (migrating) |
| 2 | 2 | 100.0% | Mesonephric cells |
| 3 | 3 | 94.9% | Granulosa cells (migrating) |
| 4 | 4 | 98.5% | Granulosa cells |
| 5 | 9 | 95.5% | Pre-granulosa cells (migrating) |
| 6 | 6 | 99.8% | Endothelial cells |
| 7 | 12 + 7 | mixed | Podocytes (mixed) |
| 8 | 5 | 99.7% | Oogonia/Oocytes |
| 9 | 8 | 99.5% | Erythrocytes |
| 10 | 10 | 99.8% | Leukocytes |
| 11 | 11 | 100.0% | Hemopoietic stem cells (1) |
| 12 | 11 | 99.9% | Hemopoietic stem cells (2) |
| 13 | 7 (markers disagree) | 72.4% | Unresolved (globin-high, mixed origin) |
| 14 | 13 | 99.8% | Macrophages |
| 15 | 14 | 92.7% | Pre-granulosa cells (mitotic) |
| 16 | 15 | 97.1% | Immune cells (migrating, mitotic) |
| 17 | 16 | 98.9% | Spermatocytes (cytokinetic) |
| 18 | 17 | 94.3% | Male germ cells (migrating) |
| 19 | 18 | 96.9% | Spermatocytes (apoptotic) |
| 20 | 19 | 93.8% | Spermatocytes (GPCR-activated) |

## Labeled TSNE, pending

```{r, eval=FALSE, fig.width=8, fig.height=6}
ct = anno[colData(merged.sce)$label]
merged.sce$CellType = ct
merged.sce$CellType = factor(merged.sce$CellType, levels = anno)

plotTSNE(merged.sce, colour_by = "CellType", text_by = "label",
         text_colour = "red", point_size = .05) +
  guides(color = guide_legend(override.aes = list(size = 5))) +
  theme(panel.background = element_rect(fill = "black"),
        axis.title = element_blank(),
        axis.text  = element_blank(),
        axis.ticks = element_blank())

ggsave(file = paste0(results_dir, "/all_cells_tsne_black.png"), height = 6, width = 8)
```

## Bubble plots per cell type, pending

```{r, eval=FALSE, fig.show="hide"}
nested_df   <- chung_davidson_markers %>% group_by(CellType) %>% tidyr::nest()
list_of_dfs <- nested_df$data
names(list_of_dfs) <- nested_df$CellType
names(list_of_dfs)  = gsub("\\(", "", names(list_of_dfs))

file_names = gsub(" ", "_", names(list_of_dfs))
file_names = paste0(file_names, "_all_cell_clusters")

plot.sce = combined.sce
colData(plot.sce) <- colData(merged.sce)

ct = anno[colData(plot.sce)$label]
plot.sce$CellType = ct
plot.sce$CellType = factor(plot.sce$CellType, levels = anno)
plot.sce$cluster  = plot.sce$label

plot_dir = paste0(results_dir, "/all_cell_bubble")
if (!dir.exists(plot_dir)) dir.create(plot_dir)

scatter_markers_df = read.csv(
  file    = paste0(results_dir, "/scatter_cluster_marker_genes.csv"),
  row.names = 1)

cluster_index              = as.numeric(gsub("cluster_", "", scatter_markers_df$cluster))
scatter_markers_df$CellType = anno[cluster_index]

markers = scatter_markers_df %>%
  select(WangGeneID, CellType, FDR, summary.logFC) %>%
  dplyr::rename(Gene = "WangGeneID")

for (i in 1:length(list_of_dfs)) {
  plot_marker_bubbles(
    sce              = plot.sce,
    markers          = markers,
    gene_description = gene_description,
    cell_type_name   = names(list_of_dfs)[i],
    cell_type_genes  = list_of_dfs[[i]],
    anno             = anno,
    results_dir      = plot_dir,
    file_name        = file_names[i]
  )
}
```

# Session info

```{r}
sessionInfo()
```
