---
title: "01_all_cells_clustering_annotation_no_doublets"
author: "Stephanie Hickey"
date: "`r Sys.Date()`"
output:
  html_document:
    code_folding: hide
    toc: true
    toc_float:
      toc_collapsed: true
params:
    project_dir: /mnt/ufs18/rs-013/bioinformaticsCore/projects/chong_davidson/BCC125_sea_lamprey_scRNA
    project_name: BCC125_sea_lamprey_scRNA
    lab: Weiming Li
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

```{r, echo=FALSE, results="asis"}
cat("\n")
cat("#### Lab:", params$lab, " \n")
cat("#### Project:", params$project_name, " \n")
```

```{r}
library(tidyverse)
library(SingleCellExperiment)
library(scater)
library(scran)
library(patchwork)
library(RColorBrewer)
library(readxl)
library(monocle3)
library(batchelor)
library(DT)
library(rtracklayer)
library(BiocSingular)
library(ggnewscale)
library(scDblFinder)
source(paste0(params$project_dir, "/src/scRNA_seq_functions.R"))
```

This is a copy of `01_all_cells_clustering_annotation.Rmd` that removes cells
called doublets by `scDblFinder` before clustering, integration or trajectory
analysis, so the effect of removing them on clustering and annotation can be
assessed directly. `scDblFinder` is rerun here, per sample, rather than
reusing scores from `04_doublet_qc.Rmd`, so this document is standalone and
does not depend on that one having been run first. Because `scDblFinder`'s
doublet simulation is stochastic, the doublet calls made here are not
guaranteed to be identical to the ones reported in `04_doublet_qc.Rmd`, even
with the same seed, if package versions differ between runs.

```{r}
results_dir = paste0(params$project_dir, "/results/20260803_no_doublets")
data_dir    = paste0(params$project_dir, "/data")
```

```{r, echo=FALSE}
# load pre-computed objects for rendering; produced by the eval=FALSE chunks
# below, once they have been run at least once against the doublet-filtered
# cells
all.sce      = readRDS(paste0(results_dir, "/all.sce.Rds"))
merged.sce   = readRDS(paste0(results_dir, "/merged.sce.Rds"))
merged.cds   = readRDS(paste0(results_dir, "/merged.cds.Rds"))
combined.sce = readRDS(paste0(results_dir, "/combined.sce.Rds"))
```

# Collect individual samples into one object

```{r, eval=FALSE}
sample_dirs  = list.dirs(data_dir, full.names = TRUE, recursive = FALSE)
sample_names = basename(sample_dirs)

cds_list = list()
for (i in 1:length(sample_dirs)) {
  cds_list[[i]] = readRDS(paste0(sample_dirs[i], "/", sample_names[i],
                                  "_cds_mito_filter.Rds"))
}
names(cds_list) = sample_names

all.sce = list()
for (cds in names(cds_list)) {
  raw_counts          = counts(cds_list[[cds]])
  all.sce[[cds]]      = SingleCellExperiment(list(counts = raw_counts),
                                              colData = cds_list[[cds]]@colData)
  all.sce[[cds]]@colData$sample = cds
}
```

## Remove cells called doublets

`scDblFinder` is run separately on each sample, i.e. per library, without a
`samples =` argument, since each list element already is one library. Doublets
are removed here, per sample, before any normalization, HVG selection or PCA,
and before the samples are merged.

```{r, eval=FALSE}
set.seed(100)
all.sce.dbl = lapply(all.sce, scDblFinder)

n_before = sapply(all.sce, ncol)
n_dbl    = sapply(all.sce.dbl, function(x) sum(x$scDblFinder.class == "doublet"))

for (cds in names(all.sce)) {
  keep = colData(all.sce.dbl[[cds]])$scDblFinder.class == "singlet"
  all.sce[[cds]] = all.sce[[cds]][, keep]
}

n_after = sapply(all.sce, ncol)
data.frame(sample = names(n_before), n_before = n_before,
           n_doublet = n_dbl, n_after = n_after, row.names = NULL)
```

```{r, eval=FALSE}
all.sce  <- lapply(all.sce, logNormCounts)
all.dec  <- lapply(all.sce, modelGeneVar)
all.hvgs <- lapply(all.dec, getTopHVGs, prop = 0.1)
all.sce  <- mapply(FUN = runPCA, x = all.sce, subset_row = all.hvgs)
saveRDS(all.sce, file = paste0(results_dir, "/all.sce.Rds"))
```

# Sample summary table

```{r}
df_list = list()
for (samp in names(all.sce)) {
  ncells     = ncol(all.sce[[samp]])
  med_counts = round(median(colData(all.sce[[samp]])$sum))
  med_genes  = round(median(colData(all.sce[[samp]])$detected))
  med_mito   = round(median(colData(all.sce[[samp]])$subsets_Mito_percent), digits = 2)

  df_list[[samp]] = data.frame(Sample                 = samp,
                                nCells                 = ncells,
                                medianCounts           = med_counts,
                                medianGenesDetected    = med_genes,
                                medianPercentMitoReads = med_mito)
}

df_all = do.call(rbind, df_list)
write.csv(df_all, file = paste0(results_dir, "/median_stats.csv"))
datatable(df_all, rownames = FALSE)
```

# Merge samples, removing batch effect

```{r, eval=FALSE}
universe     <- Reduce(intersect, lapply(all.sce, rownames))
all.sce2     <- lapply(all.sce, "[", i = universe,)
all.dec2     <- lapply(all.dec, "[", i = universe,)
normed.sce   <- do.call(multiBatchNorm, all.sce2)
combined.dec <- do.call(combineVar, all.dec2)
combined.hvg <- getTopHVGs(combined.dec, n = 5000)

set.seed(1)
merged.sce <- do.call(fastMNN, c(normed.sce,
    list(subset.row = combined.hvg, BSPARAM = RandomParam())))

set.seed(1)
g <- buildSNNGraph(merged.sce, use.dimred = "corrected", k = 30)
set.seed(1)
colLabels(merged.sce) <- factor(igraph::cluster_louvain(g)$membership)

set.seed(1)
merged.sce <- runUMAP(merged.sce, dimred = "corrected")
set.seed(1)
merged.sce <- runTSNE(merged.sce, dimred = "corrected")

saveRDS(merged.sce, file = paste0(results_dir, "/merged.sce.Rds"))
```

The plot below shows the merged data in TSNE space separated by sample.

```{r, fig.width=8, fig.height=8}
p <- lapply(unique(merged.sce$batch), function(f) {
  sce.f = merged.sce[, which(merged.sce$batch == f)]
  plotTSNE(sce.f, point_size = .05) + ggtitle(f)
})
wrap_plots(p, ncol = 2)
```

Clusters identified in batch-corrected expression space.

```{r, fig.width=6, fig.height=6}
plotTSNE(merged.sce, colour_by = "label", text_by = "label",
         text_colour = "red", point_size = .05) +
  guides(color = guide_legend(override.aes = list(size = 5)))

ggsave(file = paste0(results_dir, "/all_cells_tsne.png"))
```

Proportion of cells from each sample within each cluster.

```{r}
meta_df = as.data.frame(colData(merged.sce))

sample_tally = meta_df %>%
  group_by(label, batch) %>%
  tally()

ggplot(sample_tally, aes(fill = batch, y = label, x = n)) +
  geom_bar(position = "fill", stat = "identity") +
  xlab("Proportion of cells")
```

The same data as cell counts rather than proportions, saved in
`cells_per_cluster_per_sample.csv`.

```{r}
sample_tally_wide =
  sample_tally %>%
  tidyr::pivot_wider(names_from = batch, values_from = n, values_fill = 0) %>%
  ungroup() %>%
  mutate(label = as.character(label))

sample_tally_wide$Total = rowSums(sample_tally_wide[, -1])

sample_tally_wide =
  bind_rows(sample_tally_wide,
            sample_tally_wide %>%
              summarize(across(where(is.numeric), sum)) %>%
              mutate(label = "Total"))

write.csv(sample_tally_wide,
          file = paste0(results_dir, "/cells_per_cluster_per_sample.csv"),
          row.names = FALSE)

datatable(sample_tally_wide, rownames = FALSE)
```

```{r, eval=FALSE}
# build uncorrected combined SCE and monocle CDS for downstream plotting
all.dec  <- lapply(all.sce, modelGeneVar)
universe <- Reduce(intersect, lapply(all.sce, rownames))
all.sce2 <- lapply(all.sce, "[", i = universe,)
all.dec2 <- lapply(all.dec, "[", i = universe,)

all.sce2 <- lapply(all.sce2, function(x) {
  colData(x)    <- colData(x)[, "sample", drop = FALSE]
  reducedDims(x) <- list()
  rowData(x)    <- rowData(x)[, character(0)]
  x
})

combined.sce <- do.call(cbind, all.sce2)
saveRDS(combined.sce, file = paste0(results_dir, "/combined.sce.Rds"))

merged.cds <- new_cell_data_set(counts(combined.sce),
                                cell_metadata = merged.sce@colData)
reducedDims(merged.cds)$UMAP <- reducedDims(merged.sce)$UMAP
reducedDims(merged.cds)$tSNE <- reducedDims(merged.sce)$TSNE
rowData(merged.cds)$gene_short_name <- row.names(merged.cds)

saveRDS(merged.cds, file = paste0(results_dir, "/merged.cds.Rds"))
```

## Gene descriptions from Wang et al. and the kPetMar1.pri gtf

Summarized from [Wang et al, 2024](https://www.nature.com/articles/s41467-024-55019-x#Sec16):
LOC IDs are re-annotated to human gene symbols via BLAST; genes with human matches
are UPPERCASE, others lowercase.

```{r}
gene_description =
  read_excel(paste0(data_dir, "/41467_2024_55019_MOESM3_ESM.xlsx"), skip = 1) %>%
  dplyr::rename(WangGeneID = "CountMatrix_GeneID",
                Gene        = "NCBI_GTF_GENEID",
                description = "NCBI_GTF_ProteinInfo") %>%
  dplyr::select(Gene, WangGeneID, description) %>%
  filter(!is.na(Gene)) %>%
  filter(!WangGeneID %in% c("LOC116945425.1", "S100A1.1"))

gtf = as.data.frame(readGFF("/mnt/research/bioinformaticsCore/shared/Genomes/Petromyzon_marinus/GCF_010993605.1_kPetMar1.pri_genomic.gtf"))

gtf_filt =
  gtf %>%
  filter(gbkey %in% c("mRNA", "misc_RNA", "ncRNA", "tRNA", "rRNA")) %>%
  mutate(product = gsub(", transcript variant.*", "", product)) %>%
  dplyr::select(gene_id, product) %>%
  distinct() %>%
  dplyr::rename(Gene = "gene_id", description = "product") %>%
  mutate(WangGeneID = Gene)

notWang      = gtf_filt$Gene[!gtf_filt$Gene %in% gene_description$Gene]
gtf_add      = gtf_filt %>% filter(Gene %in% notWang)
gene_description = rbind(gene_description, gtf_add)

write.csv(gene_description,
          file = paste0(results_dir, "/gene_descritption.csv"), row.names = FALSE)
datatable(gene_description)
```

## Known cell-type markers

```{r}
markers_wide = read.csv(paste0(data_dir, "/Cell_Type_Marker_Genes_12-12-25.csv"))
markers_wide = markers_wide[, 2:ncol(markers_wide)]

markers_long <- markers_wide %>%
  pivot_longer(cols = everything(), names_to = "CellType", values_to = "Gene",
               values_drop_na = TRUE) %>%
  mutate(main_part  = str_remove(Gene, "\\s*\\(.*\\)"),
         paren_part = str_extract(Gene, "(?<=\\().*(?=\\))")) %>%
  tidyr::separate_rows(main_part, sep = "/") %>%
  mutate(main_part  = trimws(main_part),
         paren_part = trimws(paren_part)) %>%
  filter(Gene != "")

chung_davidson_markers = markers_long %>%
  dplyr::select(CellType, main_part) %>%
  dplyr::rename(WangGeneID = "main_part") %>%
  filter(!WangGeneID == "") %>%
  left_join(gene_description) %>%
  filter(!is.na(Gene))

datatable(chung_davidson_markers %>% arrange(CellType), rownames = FALSE)
```

## Cluster enriched genes

Top ~100 cluster enriched genes per cluster. All significant genes (FDR < 0.01) are
saved in `scatter_cluster_marker_genes.csv`.

**Columns:** `cluster`, `Gene`, `p.value`, `FDR`, `summary.logFC`

```{r}
scatter_markers = findMarkers(combined.sce,
                               merged.sce$label,
                               test.type = "t",
                               block     = combined.sce$sample,
                               direction = "up",
                               pval.type = "some")

saveRDS(scatter_markers, file = paste0(results_dir, "/scatter_markers.Rds"))

scatter_summary = list()
for (i in 1:length(scatter_markers)) {
  scatter_markers[[i]]$cluster = i
  scatter_markers[[i]]$Gene    = rownames(scatter_markers[[i]])
  scatter_summary[[i]] = as.data.frame(scatter_markers[[i]]) %>%
    select(cluster, Gene, p.value, FDR, summary.logFC)
}

scatter_markers_df =
  do.call(rbind, scatter_summary) %>%
  left_join(gene_description, by = "Gene") %>%
  filter(FDR < .01)

write.csv(scatter_markers_df,
          file = paste0(results_dir, "/scatter_cluster_marker_genes.csv"))

scatter_markers_top100 = scatter_markers_df %>%
  group_by(cluster) %>%
  slice_min(FDR, n = 100, with_ties = TRUE)

datatable(scatter_markers_top100, filter = "top", rownames = FALSE)
```

## Comparison with known markers

Rows = cell types with known markers; columns = clusters. Color = $-log10(FDR)$ of
overlap significance. Number in each box = shared genes.

- Overlap p-values: `scatter_cluster_marker_overlap_pvalues.csv`
- Overlapping genes: `scatter_cluster_marker_gene_overlaps.csv`

```{r, fig.width=8}
marker_test_genes =
  scatter_markers_top100 %>%
  select(Gene, cluster) %>%
  mutate(cluster = paste0("cluster", cluster))

chung_davidson_cell_type_markers = chung_davidson_markers %>%
  select(Gene, CellType)

plot_marker_overlap_heatmap(
  known_markers  = chung_davidson_cell_type_markers,
  test_genes     = marker_test_genes,
  background_n   = length(merged.sce),
  gene_description = gene_description,
  overlaps_file  = paste0(results_dir, "/scatter_cluster_marker_gene_overlaps.csv"),
  pvals_file     = paste0(results_dir, "/scatter_cluster_marker_overlap_pvalues.csv"),
  plot_file      = paste0(results_dir, "/scatter_cluster_overlap_heatmap.png")
)
```

## Gene ontology of top 100 cluster enriched genes

**Columns:** `GOID`, `GOName`, `nGOTermGenes`, `Clusters`, `nClusterEnrichedGenes`,
`nSharedGenes`, `Pval`, `FDR`

- Overlap p-values: `scatter_cluster_marker_gene_GO_overlap_pvalues.csv`
- Overlapping genes: `scatter_cluster_marker_gene_GO_overlaps.csv`

```{r}
gaf = read.delim("/mnt/research/bioinformaticsCore/shared/Genomes/Petromyzon_marinus/GCF_010993605.1_kPetMar1.pri_gene_ontology.gaf",
                 header = FALSE, comment = "!")

gaf <- gaf %>%
  filter(!grepl("NOT", V4)) %>%
  dplyr::select(V3, V5) %>%
  dplyr::rename(GOID = "V5", Gene = "V3") %>%
  distinct()

keep_go = gaf %>%
  group_by(GOID) %>% tally() %>%
  filter(n > 10, n < 200) %>%
  pull(GOID)

gaf = gaf %>% filter(GOID %in% keep_go)

go_names = read.delim("/mnt/research/bioinformaticsCore/shared/Genomes/Petromyzon_marinus/GOs_names.txt",
                      header = FALSE)
colnames(go_names) = c("GOID", "GOName")

scores = run_go_enrichment(
  gaf             = gaf,
  gene_description = gene_description,
  go_names        = go_names,
  test_genes      = marker_test_genes,
  background_n    = length(merged.cds),
  overlaps_file   = paste0(results_dir, "/scatter_cluster_marker_gene_GO_overlaps.csv"),
  pvals_file      = paste0(results_dir, "/scatter_cluster_marker_gene_GO_overlap_pvalues.csv"),
  pval_filter     = 0.01
)

datatable(scores, filter = "top", rownames = FALSE)
```

## Plot known cell-type markers

Early germ cell markers.

```{r, fig.width=8, fig.height=4}
rowData(merged.cds)$gene_short_name = row.names(merged.cds)

unique_markers = chung_davidson_markers %>%
  group_by(Gene) %>% tally() %>%
  filter(n == 1) %>%
  pull(Gene)

genes_for_plot = chung_davidson_markers %>%
  select(Gene, CellType) %>%
  filter(Gene %in% unique_markers)

plot_cells(merged.cds, reduction_method = "tSNE",
           genes = genes_for_plot %>%
             filter(CellType %in% c("Migrating.germ.cells", "Primordial.germ.cells")))
```

Female cell markers.

```{r, fig.height=4, fig.width=12}
plot_cells(merged.cds, reduction_method = "tSNE",
           genes = genes_for_plot %>%
             filter(CellType %in% c("Oogonia", "Oocyte",
                                    "Granulosa.cells", "Pre.granulosa.cells")),
           show_trajectory_graph = FALSE,
           label_cell_groups     = FALSE,
           label_leaves          = FALSE,
           label_branch_points   = FALSE)
```

Male cell markers.

```{r, fig.height=4, fig.width=12}
plot_cells(merged.cds, reduction_method = "tSNE",
           genes = genes_for_plot %>%
             filter(CellType %in% c("Spermatocytes", "Spermatogonia",
                                    "Spermatogonial.stem.cells")),
           show_trajectory_graph = FALSE,
           label_cell_groups     = FALSE,
           label_leaves          = FALSE,
           label_branch_points   = FALSE)
```

## Annotations

Cluster numbering here does not match `01_all_cells_clustering_annotation.Rmd`.
Doublet removal yields 20 clusters rather than 19, and the numbering is
reassigned by Louvain, so the labels below were re-derived rather than copied.

Each was assigned from two independent criteria computed in
`05_compare_doublet_removal.Rmd`: which with-doublets cluster contributed the
most cells, and which with-doublets cluster shares the most top-100 marker
genes. The two criteria agree for 19 of the 20 clusters. Cluster 13 is the
exception and is discussed below.

Three labels differ from the with-doublets analysis in substance, not just in
number:

  * **Cluster 11 and cluster 12** are both hemopoietic stem cells. The single
    HSC cluster in the with-doublets analysis divides roughly evenly here
    (1,302 and 1,224 cells), each essentially pure. What distinguishes them has
    not been established.
  * **Cluster 7** is podocyte-dominated by marker overlap (49 shared genes
    against 37 for the next-best match), but it also absorbed 321 cells from
    the cluster previously annotated "Primordial germ cells (migrating)". It is
    a genuine mixture and is labeled as such.
  * **Cluster 13** holds 72% of the cells from that same previously-annotated
    PGC cluster, but shares almost no markers with it (9 genes, against 40 with
    the erythrocyte cluster). Its own top markers are globins, ferritin and
    PCLAF, it draws cells from 12 different with-doublets clusters, and it has
    the lowest purity of any cluster. It carries no germline markers at all. It
    is labeled unresolved rather than given a cell type.

No cluster in this analysis carries an early germline program. Canonical PGC
markers (PRDM14, DND1, TFAP2C, SOX17, NANOG) are absent or negligible
throughout. Germline identity is confined to cluster 8 (ZP3 and ZP4 family,
logFC up to 3.9) and clusters 17 to 20 (PIWIL1, SYCP3, SYCP2, BOLL, TDRD7,
TDRD9, MAEL, TEX15).

```{r}
anno = c(
  "1. Fibroblasts (migrating)",
  "2. Mesonephric cells",
  "3. Granulosa cells (migrating)",
  "4. Granulosa cells",
  "5. Pre-granulosa cells (migrating)",
  "6. Endothelial cells",
  "7. Podocytes (mixed)",
  "8. Oogonia/Oocytes",
  "9. Erythrocytes",
  "10. Leukocytes",
  "11. Hemopoietic stem cells (1)",
  "12. Hemopoietic stem cells (2)",
  "13. Unresolved (globin-high, mixed origin)",
  "14. Macrophages",
  "15. Pre-granulosa cells (mitotic)",
  "16. Immune cells (migrating, mitotic)",
  "17. Spermatocytes (cytokinetic)",
  "18. Male germ cells (migrating)",
  "19. Spermatocytes (apoptotic)",
  "20. Spermatocytes (GPCR-activated)"
)
```

| no-doublets cluster | best match by cells | purity | annotation |
| --- | --- | --- | --- |
| 1 | 1 | 99.7% | Fibroblasts (migrating) |
| 2 | 2 | 100.0% | Mesonephric cells |
| 3 | 3 | 94.9% | Granulosa cells (migrating) |
| 4 | 4 | 98.5% | Granulosa cells |
| 5 | 9 | 95.5% | Pre-granulosa cells (migrating) |
| 6 | 6 | 99.8% | Endothelial cells |
| 7 | 12 + 7 | mixed | Podocytes (mixed) |
| 8 | 5 | 99.7% | Oogonia/Oocytes |
| 9 | 8 | 99.5% | Erythrocytes |
| 10 | 10 | 99.8% | Leukocytes |
| 11 | 11 | 100.0% | Hemopoietic stem cells (1) |
| 12 | 11 | 99.9% | Hemopoietic stem cells (2) |
| 13 | 7 (markers disagree) | 72.4% | Unresolved (globin-high, mixed origin) |
| 14 | 13 | 99.8% | Macrophages |
| 15 | 14 | 92.7% | Pre-granulosa cells (mitotic) |
| 16 | 15 | 97.1% | Immune cells (migrating, mitotic) |
| 17 | 16 | 98.9% | Spermatocytes (cytokinetic) |
| 18 | 17 | 94.3% | Male germ cells (migrating) |
| 19 | 18 | 96.9% | Spermatocytes (apoptotic) |
| 20 | 19 | 93.8% | Spermatocytes (GPCR-activated) |

```{r, fig.width=8, fig.height=6}
ct = anno[colData(merged.sce)$label]
merged.sce$CellType = ct
merged.sce$CellType = factor(merged.sce$CellType, levels = anno)

plotTSNE(merged.sce, colour_by = "CellType", text_by = "label",
         text_colour = "red", point_size = .05) +
  guides(color = guide_legend(override.aes = list(size = 5))) +
  theme(panel.background = element_rect(fill = "black"),
        axis.title = element_blank(),
        axis.text  = element_blank(),
        axis.ticks = element_blank())

ggsave(file = paste0(results_dir, "/all_cells_tsne_black.png"), height = 6, width = 8)
```

## Bubble plots per cell type

```{r, eval=TRUE, fig.show="hide"}
nested_df   <- chung_davidson_markers %>% group_by(CellType) %>% tidyr::nest()
list_of_dfs <- nested_df$data
names(list_of_dfs) <- nested_df$CellType
names(list_of_dfs)  = gsub("\\(", "", names(list_of_dfs))

file_names = gsub(" ", "_", names(list_of_dfs))
file_names = paste0(file_names, "_all_cell_clusters")

plot.sce = combined.sce
colData(plot.sce) <- colData(merged.sce)

ct = anno[colData(plot.sce)$label]
plot.sce$CellType = ct
plot.sce$CellType = factor(plot.sce$CellType, levels = anno)
plot.sce$cluster  = plot.sce$label

plot_dir = paste0(results_dir, "/all_cell_bubble")
if (!dir.exists(plot_dir)) dir.create(plot_dir)

scatter_markers_df = read.csv(
  file    = paste0(results_dir, "/scatter_cluster_marker_genes.csv"),
  row.names = 1)

cluster_index              = as.numeric(gsub("cluster_", "", scatter_markers_df$cluster))
scatter_markers_df$CellType = anno[cluster_index]

markers = scatter_markers_df %>%
  select(WangGeneID, CellType, FDR, summary.logFC) %>%
  dplyr::rename(Gene = "WangGeneID")

for (i in 1:length(list_of_dfs)) {
  plot_marker_bubbles(
    sce              = plot.sce,
    markers          = markers,
    gene_description = gene_description,
    cell_type_name   = names(list_of_dfs)[i],
    cell_type_genes  = list_of_dfs[[i]],
    anno             = anno,
    results_dir      = plot_dir,
    file_name        = file_names[i]
  )
}
```

# Session info

```{r}
sessionInfo()
```
