---
title: "02_sex_cell_subanalysis_no_doublets"
author: "Stephanie Hickey"
date: "`r Sys.Date()`"
output:
  html_document:
    code_folding: hide
    toc: true
    toc_float:
      toc_collapsed: true
params:
    project_dir: /mnt/ufs18/rs-013/bioinformaticsCore/projects/chong_davidson/BCC125_sea_lamprey_scRNA
    project_name: BCC125_sea_lamprey_scRNA
    lab: Weiming Li
    no_dbl_date: "20260813"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

```{r, echo=FALSE, results="asis"}
cat("\n")
cat("#### Lab:", params$lab, " \n")
cat("#### Project:", params$project_name, " \n")
```

```{r}
library(tidyverse)
library(SingleCellExperiment)
library(scater)
library(scran)
library(patchwork)
library(pheatmap)
library(RColorBrewer)
library(readxl)
library(monocle3)
library(batchelor)
library(DT)
library(rtracklayer)
library(BiocSingular)
library(ggnewscale)
library(ggforce)
library(ggraph)
library(tidygraph)
library(scatterpie)
library(ggrepel)
source(paste0(params$project_dir, "/src/scRNA_seq_functions.R"))
```

This document inputs the doublet-filtered, merged all-cells object produced by
`13_atlas_and_stage_correspondance.Rmd`. 

```{r}
# Which doublet-filtered run this reads and writes. 20260813 uses the PIPseq
# specified doublet rate; 20260803 used scDblFinder's 10x-calibrated default.
# The two produce different cells, so a mismatch between this and the run that
# produced merged.sce would compare objects built from different filters.
results_dir = paste0(params$project_dir, "/results/",
                     params$no_dbl_date, "_no_doublets")
data_dir    = paste0(params$project_dir, "/data")
```

`r paste0("Reading and writing ", basename(results_dir), ".")`

```{r, echo=FALSE}
# load pre-computed objects for rendering; produced by the eval=FALSE chunks
# below, once they have been run at least once
Robject_dir    = results_dir
all.sex.sce    = readRDS(paste0(Robject_dir, "/no.support.sex.sce.Rds"))
merged.sex.sce = readRDS(paste0(Robject_dir, "/no.support.merged.sex.sce.Rds"))
scatter_markers = readRDS(paste0(Robject_dir, "/no.support_sex_scatter_markers.Rds"))
merged.sce   = readRDS(paste0(results_dir, "/merged.sce.Rds"))
```

```{r}
# gene descriptions — rebuilt from source files 
gene_description =
  read_excel(paste0(data_dir, "/41467_2024_55019_MOESM3_ESM.xlsx"), skip = 1) %>%
  dplyr::rename(WangGeneID = "CountMatrix_GeneID",
                Gene        = "NCBI_GTF_GENEID",
                description = "NCBI_GTF_ProteinInfo") %>%
  dplyr::select(Gene, WangGeneID, description) %>%
  filter(!is.na(Gene)) %>%
  filter(!WangGeneID %in% c("LOC116945425.1", "S100A1.1"))

gtf = as.data.frame(readGFF("/mnt/research/bioinformaticsCore/shared/Genomes/Petromyzon_marinus/GCF_010993605.1_kPetMar1.pri_genomic.gtf"))

gtf_filt =
  gtf %>%
  filter(gbkey %in% c("mRNA", "misc_RNA", "ncRNA", "tRNA", "rRNA")) %>%
  mutate(product = gsub(", transcript variant.*", "", product)) %>%
  dplyr::select(gene_id, product) %>%
  distinct() %>%
  dplyr::rename(Gene = "gene_id", description = "product") %>%
  mutate(WangGeneID = Gene)

notWang      = gtf_filt$Gene[!gtf_filt$Gene %in% gene_description$Gene]
gtf_add      = gtf_filt %>% filter(Gene %in% notWang)
gene_description = rbind(gene_description, gtf_add)
```

```{r}
# known cell-type markers
markers_wide = read.csv(paste0(data_dir, "/Cell_Type_Marker_Genes_12-12-25.csv"))
markers_wide = markers_wide[, 2:ncol(markers_wide)]

markers_long <- markers_wide %>%
  pivot_longer(cols = everything(), names_to = "CellType", values_to = "Gene",
               values_drop_na = TRUE) %>%
  mutate(main_part  = str_remove(Gene, "\\s*\\(.*\\)"),
         paren_part = str_extract(Gene, "(?<=\\().*(?=\\))")) %>%
  tidyr::separate_rows(main_part, sep = "/") %>%
  mutate(main_part  = trimws(main_part),
         paren_part = trimws(paren_part)) %>%
  filter(Gene != "")

chung_davidson_markers = markers_long %>%
  dplyr::select(CellType, main_part) %>%
  dplyr::rename(WangGeneID = "main_part") %>%
  filter(!WangGeneID == "") %>%
  left_join(gene_description) %>%
  filter(!is.na(Gene))

chung_davidson_cell_type_markers = chung_davidson_markers %>%
  select(Gene, CellType)
```

```{r}
# GO annotation
gaf = read.delim("/mnt/research/bioinformaticsCore/shared/Genomes/Petromyzon_marinus/GCF_010993605.1_kPetMar1.pri_gene_ontology.gaf",
                 header = FALSE, comment = "!")

gaf <- gaf %>%
  filter(!grepl("NOT", V4)) %>%
  dplyr::select(V3, V5) %>%
  dplyr::rename(GOID = "V5", Gene = "V3") %>%
  distinct()

keep_go = gaf %>%
  group_by(GOID) %>% tally() %>%
  filter(n > 10, n < 200) %>%
  pull(GOID)

gaf = gaf %>% filter(GOID %in% keep_go)

go_names = read.delim("/mnt/research/bioinformaticsCore/shared/Genomes/Petromyzon_marinus/GOs_names.txt",
                      header = FALSE)
colnames(go_names) = c("GOID", "GOName")
```

```{r}
# broad annotations (15 sex cell clusters)
# TODO: copied from 02_sex_cell_subanalysis.Rmd; likely wrong post-doublet-removal.
# Re-derive from this document's own marker/GO results once the correct
# germ-cell cluster numbers below have been identified.
anno = c(
  "FGC.migrating",
  "PGC",
  "Oogonia.stem.cells",
  "Oocytes",
  "Cyst.epithelial.cells",
  "Oogonia",
  "SPC.cytokinetic",
  "MGC.SPC.group2",
  "MGC.SPC.group1",
  "MGC.SPC.group3",
  "MGC.SPC.group2",
  "MGC.SPC.group1",
  "MGC.SPC.group3",
  "MGC.SPC.group2",
  "MGC.SPC.group3"
)

colData(merged.sex.sce)$broad = anno[colData(merged.sex.sce)$label]

narrow = c(
  "Female germ cells (migrating)",
  "Primordial germ cells (migrating, differentiated, cyst)",
  "Oogonia (mitotic, cyst, stem cells)",
  "Oocytes",
  "Cyst epithelial cells",
  "Oogonia (mitotic, cyst, cell fate committed)",
  "Spermatocytes (cytokinetic)",
  "Male germ cells (migrating, cell fate committed, differentiated)",
  "Spermatocytes (meiosis arrested)",
  "Spermatocytes (migrating, S-phase, cyst connected with cell bridges, cell fate committed, GPCR-activated)",
  "Spermatocytes (apoptotic, meiosis arrested)",
  "Male germ cells (migrating, differentiated)",
  "Spermatocytes (angiogenic, cell fate committed, GPCR-activated)",
  "Spermatocytes (cell fate committed)",
  "Male germ cells (migrating, cell fate committed, differentiated, wounded)"
)

colData(merged.sex.sce)$narrow = narrow[colData(merged.sex.sce)$label]
```

# Sex cell cluster analysis

The with-doublets analysis extracted all-cell clusters 5, 7, 16, 17, 18, 19.
The equivalent germ cell clusters in the doublet-filtered clustering are
**8, 17, 18, 19, 20**: the oocyte cluster and the four meiotic male clusters,
all confirmed by canonical germline markers (ZP3/ZP4 for cluster 8; PIWIL1,
SYCP3, SYCP2, BOLL, TDRD7/9, MAEL, TEX15 for 17 to 20).

The set is one cluster smaller than before, because former cluster 7
("Primordial germ cells (migrating)") has no germline counterpart after doublet
removal. Its cells now sit in filtered clusters 13 (globin-high, no germline
markers) and 7 (podocyte-dominated), neither of which belongs in a germ cell
subanalysis. Filtered cluster 7 does contain 321 cells from that population,
so including it would import podocytes to recover cells with no demonstrated
germline identity.

**Consequence for the trajectory:** the with-doublets germ cell trajectory was
rooted at a PGC subcluster derived from all-cells cluster 7, and subclusters 2,
3 and 5 all came from it. Those subclusters were also the only material
connecting the female and male branches. With that population excluded, this
subanalysis may yield two disconnected trajectories rather than one bifurcation.
Resolve that before relying on `03_trajectory_analysis_no_doublets.Rmd`.

### All-cells cluster numbering, with and without doublets

Reproduced from `01_all_cells_clustering_annotation_no_doublets.Rmd`, derived in
`05_compare_doublet_removal.Rmd`. This is the source of the cluster numbers used
in the filter below. Germ cell clusters selected for this subanalysis are in
bold.

| no-doublets cluster | best match by cells | purity | annotation |
| --- | --- | --- | --- |
| 1 | 1 | 99.7% | Fibroblasts (migrating) |
| 2 | 2 | 100.0% | Mesonephric cells |
| 3 | 3 | 94.9% | Granulosa cells (migrating) |
| 4 | 4 | 98.5% | Granulosa cells |
| 5 | 9 | 95.5% | Pre-granulosa cells (migrating) |
| 6 | 6 | 99.8% | Endothelial cells |
| 7 | 12 + 7 | mixed | Podocytes (mixed) |
| **8** | 5 | 99.7% | **Oogonia/Oocytes** |
| 9 | 8 | 99.5% | Erythrocytes |
| 10 | 10 | 99.8% | Leukocytes |
| 11 | 11 | 100.0% | Hemopoietic stem cells (1) |
| 12 | 11 | 99.9% | Hemopoietic stem cells (2) |
| 13 | 7 (markers disagree) | 72.4% | Unresolved (globin-high, mixed origin) |
| 14 | 13 | 99.8% | Macrophages |
| 15 | 14 | 92.7% | Pre-granulosa cells (mitotic) |
| 16 | 15 | 97.1% | Immune cells (migrating, mitotic) |
| **17** | 16 | 98.9% | **Spermatocytes (cytokinetic)** |
| **18** | 17 | 94.3% | **Male germ cells (migrating)** |
| **19** | 18 | 96.9% | **Spermatocytes (apoptotic)** |
| **20** | 19 | 93.8% | **Spermatocytes (GPCR-activated)** |

**Note that this table describes the all-cells clustering only.** The sex
subcluster mapping and annotation tables later in this document are still
carried over from `02_sex_cell_subanalysis.Rmd` and will change once this
document is rerun on the doublet-filtered data, since the subclustering has not
yet been recomputed.

## Filter and re-cluster

```{r, eval=FALSE}
sample_dirs  = list.dirs(data_dir, full.names = TRUE, recursive = FALSE)
sample_dirs = sample_dirs[grep("smith_annotation", sample_dirs, invert=TRUE)]
sample_names = basename(sample_dirs)

cds_list = list()
for (i in 1:length(sample_dirs)) {
  cds_list[[i]] = readRDS(paste0(sample_dirs[i], "/", sample_names[i],
                                  "_cds_mito_filter.Rds"))
}
names(cds_list) = sample_names

merged.sce    = readRDS(paste0(results_dir, "/merged.sce.Rds"))
meta_df       = as.data.frame(colData(merged.sce))
meta_df$cells = rownames(meta_df)

all.sex.sce = list()
for (cds in names(cds_list)) {
  raw_counts = counts(cds_list[[cds]])
  all.sce    = SingleCellExperiment(list(counts = raw_counts),
                                    colData = cds_list[[cds]]@colData)
  all.sce@colData$sample = cds

  # TODO: cluster numbers below are from the with-doublets analysis; confirm
  # against the doublet-filtered annotation before relying on this filter.
  sex.cells = meta_df %>%
    filter(label %in% c(17, 18, 19, 20, 8), batch == cds) %>%
    pull(cells)

  all.sex.sce[[cds]] = all.sce[, sex.cells]
}

set.seed(1)
all.sex.sce  <- lapply(all.sex.sce, logNormCounts)
all.sex.dec  <- lapply(all.sex.sce, modelGeneVar)
all.sex.hvgs <- lapply(all.sex.dec, getTopHVGs, prop = 0.1)
all.sex.sce  <- mapply(FUN = runPCA, x = all.sex.sce, subset_row = all.sex.hvgs)

saveRDS(all.sex.sce, file = paste0(results_dir, "/no.support.sex.sce.Rds"))
```

```{r, eval=FALSE}
all.sex.dec <- lapply(all.sex.sce, modelGeneVar)

universe    <- Reduce(intersect, lapply(all.sex.sce, rownames))
all.sex.sce2 <- lapply(all.sex.sce, "[", i = universe,)
all.sex.dec2 <- lapply(all.sex.dec, "[", i = universe,)

normed.sce   <- do.call(multiBatchNorm, all.sex.sce2)
combined.dec <- do.call(combineVar, all.sex.dec2)
combined.hvg <- getTopHVGs(combined.dec, n = 5000)

set.seed(1)
merged.sex.sce <- do.call(fastMNN, c(normed.sce,
    list(subset.row = combined.hvg, BSPARAM = RandomParam())))

set.seed(1)
g <- buildSNNGraph(merged.sex.sce, use.dimred = "corrected", k = 8)
set.seed(1)
colLabels(merged.sex.sce) <- factor(igraph::cluster_louvain(g)$membership)

all.scatter.clust =
  meta_df %>%
  filter(label %in% c(17, 18, 19, 20, 8)) %>%
  select(cells, label) %>%
  tibble::deframe()

all.scatter.clust              <- all.scatter.clust[colnames(merged.sex.sce)]
merged.sex.sce$all.scatter.clust = all.scatter.clust

merged.sex.sce <- runUMAP(merged.sex.sce, dimred = "corrected")
merged.sex.sce <- runTSNE(merged.sex.sce, dimred = "corrected")

saveRDS(merged.sex.sce, file = paste0(results_dir, "/no.support.merged.sex.sce.Rds"))
```

## Plot new clusters in TSNE space

```{r, fig.width=10, fig.height=4}
# Both colour variables are factors inherited from the parent object, so they
# carry levels that are not present here (all.scatter.clust keeps all 19
# all-cells levels though only 6 appear) and their levels sort as text, giving
# a legend ordered 1, 10, 11, ... 2, 3. Rebuild both with only the levels
# present, ordered numerically.
numeric_factor = function(x) {
  x = as.character(x)
  factor(x, levels = as.character(sort(unique(as.numeric(x)))))
}

colData(merged.sex.sce)$sex_cluster = numeric_factor(colData(merged.sex.sce)$label)
colData(merged.sex.sce)$all_cluster = numeric_factor(colData(merged.sex.sce)$all.scatter.clust)

# point_size .05 leaves the legend keys too small to match against the plot
key_size = guides(colour = guide_legend(override.aes = list(size = 4)))

tsne_new = plotTSNE(merged.sex.sce, colour_by = "sex_cluster",
                    text_by = "sex_cluster", text_colour = "black",
                    point_size = .05) +
  key_size + ggtitle("Sex cell subclusters")

tsne_old = plotTSNE(merged.sex.sce, colour_by = "all_cluster",
                    text_by = "all_cluster", text_colour = "black",
                    point_size = .05) +
  key_size + ggtitle("Originating all-cells clusters")

tsne_new + tsne_old
```

### Original to new cluster assignment

```{r}
df = as.data.frame(merged.sex.sce@colData)

df_count =
  df %>%
  select(label, all.scatter.clust) %>%
  dplyr::rename(sex.merged.clusters = "label") %>%
  dplyr::count(sex.merged.clusters, all.scatter.clust) %>%
  group_by(all.scatter.clust) %>%
  mutate(total_m1 = sum(n)) %>%
  ungroup() %>%
  group_by(sex.merged.clusters) %>%
  mutate(total_m2 = sum(n)) %>%
  ungroup() %>%
  mutate(
    pct_of_all.clusters = round((n / total_m1) * 100, 2),
    pct_of_sex.clusters = round((n / total_m2) * 100, 2)) %>%
  select(-total_m1, -total_m2)

datatable(df_count, filter = "top", rownames = FALSE)
write.csv(df_count, paste0(results_dir, "/cluster_to_subcluster_map.csv"))
```

## Cluster distance heatmap (batch corrected space)

```{r}
embeddings <- reducedDim(merged.sex.sce, "corrected")[, 1:30]
clusters   <- colData(merged.sex.sce)$label

centroids  <- aggregate(embeddings, by = list(cluster = clusters), FUN = mean)
rownames(centroids) <- centroids$cluster
centroids  <- centroids[, -1]

dist_matrix <- as.matrix(dist(centroids, method = "euclidean"))

pheatmap(dist_matrix,
         color          = colorRampPalette(c("lightyellow", "orange", "darkred"))(100),
         main           = "Pairwise distances between cluster centroids (batch corrected space)",
         display_numbers = TRUE,
         number_format  = "%.1f")
```

### Sex subcluster to all-cell cluster mapping

| sex subclusters | all cell clusters | Broad Annotation |
| --- | --- | --- |
| 1 | Oogonia/Oocytes | ? |
| 2 | Male germ cells (migrating) / Spermatocytes (GPCR-activated) | noise |
| 3 | Spermatocytes (cytokinetic) | ? |
| 4 | Male germ cells (migrating) | MGC |
| 5 | Male germ cells (migrating) | MGC |
| 6 | Spermatocytes (apoptotic) | MGC-SPC |
| 7 | Spermatocytes (GPCR-activated) | SPC-GPCR |
| 8 | Male germ cells (migrating) | MGC |
| 9 | Male germ cells (migrating) | MGC-SPC |
| 10 | Spermatocytes (GPCR-activated) | SPC-GPCR |
| 11 | Male germ cells (migrating) / Spermatocytes (apoptotic) | MGC-SPC |
| 12 | Spermatocytes (GPCR-activated) | SPC-GPCR |
| 13 | Male germ cells (migrating) / Spermatocytes (GPCR-activated) | ? |
| 14 | Oogonia/Oocytes | noise |

## Cells per broad annotation by sample

```{r, fig.height=8, fig.width=12}
p <- lapply(unique(merged.sex.sce$batch), function(f) {
  sce.f = merged.sex.sce[, which(merged.sex.sce$batch == f)]
  plotTSNE(sce.f, colour_by = "broad", text_colour = "black", point_size = .05) +
    ggtitle(f)
})
wrap_plots(p, ncol = 2)
```

```{r}
meta_df = as.data.frame(colData(merged.sex.sce))

sample_tally =
  meta_df %>%
  mutate(label = as.factor(label), batch = as.factor(batch)) %>%
  group_by(label, batch, .drop = FALSE) %>%
  tally()

ggplot(sample_tally, aes(fill = batch, y = label, x = n)) +
  geom_bar(position = "fill", stat = "identity") +
  xlab("Proportion of cells")
```

## Build combined (uncorrected) sex cell SCE

```{r, eval=TRUE}
all.sex.dec  <- lapply(all.sex.sce, modelGeneVar)

universe     <- Reduce(intersect, lapply(all.sex.sce, rownames))
all.sex.sce2 <- lapply(all.sex.sce, "[", i = universe,)
all.sex.dec2 <- lapply(all.sex.dec, "[", i = universe,)

all.sex.sce2[["TM_C00004374"]]@colData$cluster = "noclust"

combined.sex.sce <- do.call(cbind, all.sex.sce2)
```

# Cluster enriched genes

Top ~100 cluster enriched genes per broad annotation (all in `no.support_sex_scatter_cluster_marker_genes.csv`).

**Columns:** `cluster`, `Gene`, `p.value`, `FDR`, `summary.logFC`, `WangGeneID`, `description`

```{r, eval=FALSE}
scatter_markers = findMarkers(combined.sex.sce,
                               merged.sex.sce$label,
                               test.type = "t",
                               block     = combined.sex.sce$sample,
                               direction = "up",
                               pval.type = "some")

saveRDS(scatter_markers,
        file = paste0(results_dir, "/no.support_sex_scatter_markers.Rds"))
```

```{r}
scatter_summary = list()
for (i in 1:length(scatter_markers)) {
  scatter_markers[[i]]$cluster = names(scatter_markers)[i]
  scatter_markers[[i]]$Gene    = rownames(scatter_markers[[i]])
  scatter_summary[[i]] = as.data.frame(scatter_markers[[i]]) %>%
    select(cluster, Gene, p.value, FDR, summary.logFC)
}

scatter_markers_df =
  do.call(rbind, scatter_summary) %>%
  left_join(gene_description, by = "Gene") %>%
  filter(FDR < .01)

write.csv(scatter_markers_df,
          file = paste0(results_dir, "/no.support_sex_scatter_cluster_marker_genes.csv"))

scatter_markers_top100 = scatter_markers_df %>%
  group_by(cluster) %>%
  slice_min(FDR, n = 100, with_ties = TRUE) %>%
  mutate(p.value       = formatC(signif(p.value, digits = 3)),
         FDR           = formatC(signif(FDR, digits = 3)),
         summary.logFC = round(summary.logFC, digits = 3))

datatable(scatter_markers_top100, filter = "top", rownames = FALSE)
```

## Comparison with known markers

```{r, fig.width=8}
marker_test_genes =
  scatter_markers_top100 %>%
  select(Gene, cluster)

plot_marker_overlap_heatmap(
  known_markers    = chung_davidson_cell_type_markers,
  test_genes       = marker_test_genes,
  background_n     = length(merged.sex.sce),
  gene_description = gene_description,
  overlaps_file    = paste0(results_dir, "/no.support_sex_scatter_cluster_marker_gene_overlaps.csv"),
  pvals_file       = paste0(results_dir, "/no.support_sex_scatter_cluster_marker_overlap_pvalues.csv"),
  plot_file        = paste0(results_dir, "/no.support_sex_scatter_cluster_overlap_heatmap.png")
)
```

## Gene ontology of top 100 cluster enriched genes

**Columns:** `GOID`, `GOName`, `nGOTermGenes`, `Clusters`, `nClusterEnrichedGenes`,
`nSharedGenes`, `Pval`, `FDR`

```{r}
scores = run_go_enrichment(
  gaf             = gaf,
  gene_description = gene_description,
  go_names        = go_names,
  test_genes      = marker_test_genes,
  background_n    = length(merged.sex.sce),
  overlaps_file   = paste0(results_dir, "/no.support_sex_scatter_cluster_top_100_marker_gene_GO_overlaps.csv"),
  pvals_file      = paste0(results_dir, "/no.support_sex_scatter_cluster_top_100_marker_gene_GO_pvals.csv"),
  require_shared  = TRUE
)

datatable(scores, filter = "top", rownames = FALSE)
```

## Gene ontology of all significant cluster enriched genes (FDR < 0.01)

```{r}
marker_test_genes_all =
  scatter_markers_df %>%
  select(Gene, cluster)

scores = run_go_enrichment(
  gaf             = gaf,
  gene_description = gene_description,
  go_names        = go_names,
  test_genes      = marker_test_genes_all,
  background_n    = length(merged.sex.sce),
  overlaps_file   = paste0(results_dir, "/no.support_sex_scatter_cluster_all_sig_marker_gene_GO_overlaps_all.csv"),
  pvals_file      = paste0(results_dir, "/no.support_sex_scatter_all_sig_cluster_marker_gene_GO_pvals.csv"),
  require_shared  = TRUE
)

datatable(scores, filter = "top", rownames = FALSE)
```

## Known marker bubble plots

```{r, eval=TRUE, fig.show="hide"}
nested_df   <- chung_davidson_markers %>% group_by(CellType) %>% tidyr::nest()
list_of_dfs <- nested_df$data
names(list_of_dfs) <- nested_df$CellType
names(list_of_dfs)  = gsub("\\(", "", names(list_of_dfs))

file_names = gsub(" ", "_", names(list_of_dfs))
file_names = paste0(file_names, "_germ_cell_clusters")

ct = colData(merged.sex.sce)$label
combined.sex.sce$CellType = ct
combined.sex.sce$CellType = factor(combined.sex.sce$CellType, levels = unique(ct))
combined.sex.sce$cluster  = ct

plot_dir = paste0(results_dir, "/germ_cell_bubble")
if (!dir.exists(plot_dir)) dir.create(plot_dir)

scatter_markers_df_read = read.csv(
  file    = paste0(results_dir, "/no.support_sex_scatter_cluster_marker_genes.csv"),
  row.names = 1)

markers = scatter_markers_df_read %>%
  dplyr::rename(CellType = "cluster") %>%
  select(WangGeneID, CellType, FDR, summary.logFC) %>%
  dplyr::rename(Gene = "WangGeneID")

for (i in 1:length(list_of_dfs)) {
  plot_marker_bubbles(
    sce              = combined.sex.sce,
    markers          = markers,
    gene_description = gene_description,
    cell_type_name   = names(list_of_dfs)[i],
    cell_type_genes  = list_of_dfs[[i]],
    anno             = unique(ct),
    results_dir      = plot_dir,
    file_name        = file_names[i]
  )
}
```

# Sex cluster annotations

| sex subcluster | Narrow annotation |
| --- | --- |
| 1 | Female germ cells (migrating) |
| 2 | Primordial germ cells (migrating, differentiated, cyst) |
| 3 | Oogonia (mitotic, cyst, stem cells) |
| 4 | Oocytes |
| 5 | Cyst epithelial cells |
| 6 | Oogonia (mitotic, cyst, cell fate committed) |
| 7 | Spermatocytes (cytokinetic) |
| 8 | Male germ cells (migrating, cell fate committed, differentiated) |
| 9 | Spermatocytes (meiosis arrested) |
| 10 | Spermatocytes (migrating, S-phase, cyst connected with cell bridges, cell fate committed, GPCR-activated) |
| 11 | Spermatocytes (apoptotic, meiosis arrested) |
| 12 | Male germ cells (migrating, differentiated) |
| 13 | Spermatocytes (angiogenic, cell fate committed, GPCR-activated) |
| 14 | Spermatocytes (cell fate committed) |
| 15 | Male germ cells (migrating, cell fate committed, differentiated, wounded) |

# Pairwise differential expression

Each section subsets two cell types, runs `findMarkers`, and reports GO enrichment
of the top 100 and all significant (FDR < 0.01) DE genes.

**TODO: the `cluster_labels` numbers and section titles below (2 vs 3, 2 vs
1, 2 vs 12, read as sex-cell subcluster numbers from `merged.sex.sce$label`,
not all-cell clusters) are copied unchanged from `02_sex_cell_subanalysis.Rmd`
and are almost certainly wrong here.** Doublet removal changes which
subcluster number corresponds to which cell type in this document's own
re-clustering above; re-identify the correct pairs (e.g. PGC vs. Oogonia stem
cells) from this document's cluster-to-subcluster mapping and marker tables
before trusting any of the comparisons below.

## PGC vs Oogonia stem cells (clusters 2 vs 3)

```{r}
de_23 = run_pairwise_de(
  combined_sce   = combined.sex.sce,
  merged_sce     = merged.sex.sce,
  cluster_labels = c(2, 3),
  gene_description = gene_description,
  results_dir    = results_dir,
  file_prefix    = "no.support_sex_scatter_PCGvsOSC"
)

datatable(de_23$top100, filter = "top", rownames = FALSE)
```

### GO enrichment — top 100 DE genes

```{r}
scores = run_go_enrichment(
  gaf             = gaf,
  gene_description = gene_description,
  go_names        = go_names,
  test_genes      = de_23$top100 %>% select(Gene, cluster),
  background_n    = length(merged.sex.sce),
  overlaps_file   = paste0(results_dir, "/no.support_sex_scatter_cluster_2vs3_top_100_DE_gene_GO_overlaps.csv"),
  pvals_file      = paste0(results_dir, "/no.support_sex_scatter_cluster_2vs3_top_100_DE_gene_GO_pvals.csv"),
  require_shared  = TRUE
)

datatable(scores, filter = "top", rownames = FALSE)
```

### GO enrichment — all significant DE genes (FDR < 0.01)

```{r}
scores = run_go_enrichment(
  gaf             = gaf,
  gene_description = gene_description,
  go_names        = go_names,
  test_genes      = de_23$all %>% select(Gene, cluster),
  background_n    = length(merged.sex.sce),
  overlaps_file   = paste0(results_dir, "/no.support_sex_scatter_cluster_2vs3_sig_DE_gene_GO_overlaps.csv"),
  pvals_file      = paste0(results_dir, "/no.support_sex_scatter_cluster_2vs3_sig_DE_gene_GO_pvals.csv"),
  require_shared  = TRUE
)

datatable(scores, filter = "top", rownames = FALSE)
```

## PGC vs Female germ cells (clusters 2 vs 1)

```{r}
de_21 = run_pairwise_de(
  combined_sce   = combined.sex.sce,
  merged_sce     = merged.sex.sce,
  cluster_labels = c(2, 1),
  gene_description = gene_description,
  results_dir    = results_dir,
  file_prefix    = "no.support_sex_scatter_PCGvsFGC"
)

datatable(de_21$top100, filter = "top", rownames = FALSE)
```

### GO enrichment — top 100 DE genes

```{r}
scores = run_go_enrichment(
  gaf             = gaf,
  gene_description = gene_description,
  go_names        = go_names,
  test_genes      = de_21$top100 %>% select(Gene, cluster),
  background_n    = length(merged.sex.sce),
  overlaps_file   = paste0(results_dir, "/no.support_sex_scatter_cluster_2v1_top_100_DE_gene_GO_overlaps.csv"),
  pvals_file      = paste0(results_dir, "/no.support_sex_scatter_cluster_2vs1_top_100_DE_gene_GO_pvals.csv"),
  require_shared  = TRUE
)

datatable(scores, filter = "top", rownames = FALSE)
```

### GO enrichment — all significant DE genes (FDR < 0.01)

```{r}
scores = run_go_enrichment(
  gaf             = gaf,
  gene_description = gene_description,
  go_names        = go_names,
  test_genes      = de_21$all %>% select(Gene, cluster),
  background_n    = length(merged.sex.sce),
  overlaps_file   = paste0(results_dir, "/no.support_sex_scatter_cluster_2vs1_sig_DE_gene_GO_overlaps.csv"),
  pvals_file      = paste0(results_dir, "/no.support_sex_scatter_cluster_2vs1_sig_DE_gene_GO_pvals.csv"),
  require_shared  = TRUE
)

datatable(scores, filter = "top", rownames = FALSE)
```

## PGC vs Male germ cells (clusters 2 vs 12)

```{r}
de_212 = run_pairwise_de(
  combined_sce   = combined.sex.sce,
  merged_sce     = merged.sex.sce,
  cluster_labels = c(2, 12),
  gene_description = gene_description,
  results_dir    = results_dir,
  file_prefix    = "no.support_sex_scatter_PCGvsMGC"
)

datatable(de_212$top100, filter = "top", rownames = FALSE)
```

### GO enrichment — top 100 DE genes

```{r}
scores = run_go_enrichment(
  gaf             = gaf,
  gene_description = gene_description,
  go_names        = go_names,
  test_genes      = de_212$top100 %>% select(Gene, cluster),
  background_n    = length(merged.sex.sce),
  overlaps_file   = paste0(results_dir, "/no.support_sex_scatter_cluster_2v12_top_100_DE_gene_GO_overlaps.csv"),
  pvals_file      = paste0(results_dir, "/no.support_sex_scatter_cluster_2vs12_top_100_DE_gene_GO_pvals.csv"),
  require_shared  = TRUE
)

datatable(scores, filter = "top", rownames = FALSE)
```

### GO enrichment — all significant DE genes (FDR < 0.01)

```{r}
scores = run_go_enrichment(
  gaf             = gaf,
  gene_description = gene_description,
  go_names        = go_names,
  test_genes      = de_212$all %>% select(Gene, cluster),
  background_n    = length(merged.sex.sce),
  overlaps_file   = paste0(results_dir, "/no.support_sex_scatter_cluster_2vs12_sig_DE_gene_GO_overlaps.csv"),
  pvals_file      = paste0(results_dir, "/no.support_sex_scatter_cluster_2vs12_sig_DE_gene_GO_pvals.csv"),
  require_shared  = TRUE
)

datatable(scores, filter = "top", rownames = FALSE)
```

# Session info

```{r}
sessionInfo()
```
