---
title: "02_sex_cell_subanalysis"
author: "Stephanie Hickey"
date: "`r Sys.Date()`"
output:
  html_document:
    code_folding: hide
    toc: true
    toc_float:
      toc_collapsed: true
params:
    project_dir: /mnt/ufs18/rs-013/bioinformaticsCore/projects/chong_davidson/BCC125_sea_lamprey_scRNA
    project_name: BCC125_sea_lamprey_scRNA
    lab: Weiming Li
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

```{r, echo=FALSE, results="asis"}
cat("\n")
cat("#### Lab:", params$lab, " \n")
cat("#### Project:", params$project_name, " \n")
```

```{r}
library(tidyverse)
library(SingleCellExperiment)
library(scater)
library(scran)
library(patchwork)
library(pheatmap)
library(RColorBrewer)
library(readxl)
library(monocle3)
library(batchelor)
library(DT)
library(rtracklayer)
library(BiocSingular)
library(ggnewscale)
library(ggforce)
library(ggraph)
library(tidygraph)
library(scatterpie)
library(ggrepel)
source(paste0(params$project_dir, "/src/scRNA_seq_functions.R"))
```

```{r}
results_dir = paste0(params$project_dir, "/results/20260615")
data_dir    = paste0(params$project_dir, "/data")
```

```{r, echo=FALSE}
# load pre-computed objects for rendering
Robject_dir    = paste0(params$project_dir, "/results/20260312/Robjects")
all.sex.sce    = readRDS(paste0(Robject_dir, "/no.support.sex.sce.Rds"))
merged.sex.sce = readRDS(paste0(Robject_dir, "/no.support.merged.sex.sce.Rds"))
scatter_markers = readRDS(paste0(Robject_dir, "/no.support_sex_scatter_markers.Rds"))
```

```{r}
# gene descriptions — rebuilt from source files 
gene_description =
  read_excel(paste0(data_dir, "/41467_2024_55019_MOESM3_ESM.xlsx"), skip = 1) %>%
  dplyr::rename(WangGeneID = "CountMatrix_GeneID",
                Gene        = "NCBI_GTF_GENEID",
                description = "NCBI_GTF_ProteinInfo") %>%
  dplyr::select(Gene, WangGeneID, description) %>%
  filter(!is.na(Gene)) %>%
  filter(!WangGeneID %in% c("LOC116945425.1", "S100A1.1"))

gtf = as.data.frame(readGFF("/mnt/research/bioinformaticsCore/shared/Genomes/Petromyzon_marinus/GCF_010993605.1_kPetMar1.pri_genomic.gtf"))

gtf_filt =
  gtf %>%
  filter(gbkey %in% c("mRNA", "misc_RNA", "ncRNA", "tRNA", "rRNA")) %>%
  mutate(product = gsub(", transcript variant.*", "", product)) %>%
  dplyr::select(gene_id, product) %>%
  distinct() %>%
  dplyr::rename(Gene = "gene_id", description = "product") %>%
  mutate(WangGeneID = Gene)

notWang      = gtf_filt$Gene[!gtf_filt$Gene %in% gene_description$Gene]
gtf_add      = gtf_filt %>% filter(Gene %in% notWang)
gene_description = rbind(gene_description, gtf_add)
```

```{r}
# known cell-type markers
markers_wide = read.csv(paste0(data_dir, "/Cell_Type_Marker_Genes_12-12-25.csv"))
markers_wide = markers_wide[, 2:ncol(markers_wide)]

markers_long <- markers_wide %>%
  pivot_longer(cols = everything(), names_to = "CellType", values_to = "Gene",
               values_drop_na = TRUE) %>%
  mutate(main_part  = str_remove(Gene, "\\s*\\(.*\\)"),
         paren_part = str_extract(Gene, "(?<=\\().*(?=\\))")) %>%
  tidyr::separate_rows(main_part, sep = "/") %>%
  mutate(main_part  = trimws(main_part),
         paren_part = trimws(paren_part)) %>%
  filter(Gene != "")

chung_davidson_markers = markers_long %>%
  dplyr::select(CellType, main_part) %>%
  dplyr::rename(WangGeneID = "main_part") %>%
  filter(!WangGeneID == "") %>%
  left_join(gene_description) %>%
  filter(!is.na(Gene))

chung_davidson_cell_type_markers = chung_davidson_markers %>%
  select(Gene, CellType)
```

```{r}
# GO annotation
gaf = read.delim("/mnt/research/bioinformaticsCore/shared/Genomes/Petromyzon_marinus/GCF_010993605.1_kPetMar1.pri_gene_ontology.gaf",
                 header = FALSE, comment = "!")

gaf <- gaf %>%
  filter(!grepl("NOT", V4)) %>%
  dplyr::select(V3, V5) %>%
  dplyr::rename(GOID = "V5", Gene = "V3") %>%
  distinct()

keep_go = gaf %>%
  group_by(GOID) %>% tally() %>%
  filter(n > 10, n < 200) %>%
  pull(GOID)

gaf = gaf %>% filter(GOID %in% keep_go)

go_names = read.delim("/mnt/research/bioinformaticsCore/shared/Genomes/Petromyzon_marinus/GOs_names.txt",
                      header = FALSE)
colnames(go_names) = c("GOID", "GOName")
```

```{r}
# broad annotations (15 sex cell clusters)
anno = c(
  "FGC.migrating",
  "PGC",
  "Oogonia.stem.cells",
  "Oocytes",
  "Cyst.epithelial.cells",
  "Oogonia",
  "SPC.cytokinetic",
  "MGC.SPC.group2",
  "MGC.SPC.group1",
  "MGC.SPC.group3",
  "MGC.SPC.group2",
  "MGC.SPC.group1",
  "MGC.SPC.group3",
  "MGC.SPC.group2",
  "MGC.SPC.group3"
)

colData(merged.sex.sce)$broad = anno[colData(merged.sex.sce)$label]

narrow = c(
  "Female germ cells (migrating)",
  "Primordial germ cells (migrating, differentiated, cyst)",
  "Oogonia (mitotic, cyst, stem cells)",
  "Oocytes",
  "Cyst epithelial cells",
  "Oogonia (mitotic, cyst, cell fate committed)",
  "Spermatocytes (cytokinetic)",
  "Male germ cells (migrating, cell fate committed, differentiated)",
  "Spermatocytes (meiosis arrested)",
  "Spermatocytes (migrating, S-phase, cyst connected with cell bridges, cell fate committed, GPCR-activated)",
  "Spermatocytes (apoptotic, meiosis arrested)",
  "Male germ cells (migrating, differentiated)",
  "Spermatocytes (angiogenic, cell fate committed, GPCR-activated)",
  "Spermatocytes (cell fate committed)",
  "Male germ cells (migrating, cell fate committed, differentiated, wounded)"
)

colData(merged.sex.sce)$narrow = narrow[colData(merged.sex.sce)$label]
```

# Sex cell cluster analysis

Cells from all-cell clusters 5, 7, 16, 17, 18, 19 were re-extracted and re-clustered.

## Filter and re-cluster 

```{r, eval=FALSE}
sample_dirs  = list.dirs(data_dir, full.names = TRUE, recursive = FALSE)
sample_names = basename(sample_dirs)

cds_list = list()
for (i in 1:length(sample_dirs)) {
  cds_list[[i]] = readRDS(paste0(sample_dirs[i], "/", sample_names[i],
                                  "_cds_mito_filter.Rds"))
}
names(cds_list) = sample_names

meta_df       = as.data.frame(colData(merged.sce))
meta_df$cells = rownames(meta_df)

all.sex.sce = list()
for (cds in names(cds_list)) {
  raw_counts = counts(cds_list[[cds]])
  all.sce    = SingleCellExperiment(list(counts = raw_counts),
                                    colData = cds_list[[cds]]@colData)
  all.sce@colData$sample = cds

  sex.cells = meta_df %>%
    filter(label %in% c(16, 17, 18, 19, 5, 7), batch == cds) %>%
    pull(cells)

  all.sex.sce[[cds]] = all.sce[, sex.cells]
}

set.seed(1)
all.sex.sce  <- lapply(all.sex.sce, logNormCounts)
all.sex.dec  <- lapply(all.sex.sce, modelGeneVar)
all.sex.hvgs <- lapply(all.sex.dec, getTopHVGs, prop = 0.1)
all.sex.sce  <- mapply(FUN = runPCA, x = all.sex.sce, subset_row = all.sex.hvgs)

saveRDS(all.sex.sce, file = paste0(results_dir, "/no.support.sex.sce.Rds"))
```

```{r, eval=FALSE}
all.sex.dec <- lapply(all.sex.sce, modelGeneVar)

universe    <- Reduce(intersect, lapply(all.sex.sce, rownames))
all.sex.sce2 <- lapply(all.sex.sce, "[", i = universe,)
all.sex.dec2 <- lapply(all.sex.dec, "[", i = universe,)

normed.sce   <- do.call(multiBatchNorm, all.sex.sce2)
combined.dec <- do.call(combineVar, all.sex.dec2)
combined.hvg <- getTopHVGs(combined.dec, n = 5000)

set.seed(1)
merged.sex.sce <- do.call(fastMNN, c(normed.sce,
    list(subset.row = combined.hvg, BSPARAM = RandomParam())))

set.seed(1)
g <- buildSNNGraph(merged.sex.sce, use.dimred = "corrected", k = 8)
set.seed(1)
colLabels(merged.sex.sce) <- factor(igraph::cluster_louvain(g)$membership)

all.scatter.clust =
  meta_df %>%
  filter(label %in% c(16, 17, 18, 19, 5, 7)) %>%
  select(cells, label) %>%
  tibble::deframe()

all.scatter.clust              <- all.scatter.clust[colnames(merged.sex.sce)]
merged.sex.sce$all.scatter.clust = all.scatter.clust

merged.sex.sce <- runUMAP(merged.sex.sce, dimred = "corrected")
merged.sex.sce <- runTSNE(merged.sex.sce, dimred = "corrected")

saveRDS(merged.sex.sce, file = paste0(results_dir, "/no.support.merged.sex.sce.Rds"))
```

## Plot new clusters in TSNE space

```{r, fig.width=10, fig.height=4}
tsne_new = plotTSNE(merged.sex.sce, colour_by = "label",
                    text_by = "label", text_colour = "black", point_size = .05)
tsne_old = plotTSNE(merged.sex.sce, colour_by = "all.scatter.clust",
                    text_by = "all.scatter.clust", text_colour = "black", point_size = .05)
tsne_new + tsne_old
```

### Original to new cluster assignment

```{r}
df = as.data.frame(merged.sex.sce@colData)

df_count =
  df %>%
  select(label, all.scatter.clust) %>%
  dplyr::rename(sex.merged.clusters = "label") %>%
  dplyr::count(sex.merged.clusters, all.scatter.clust) %>%
  group_by(all.scatter.clust) %>%
  mutate(total_m1 = sum(n)) %>%
  ungroup() %>%
  group_by(sex.merged.clusters) %>%
  mutate(total_m2 = sum(n)) %>%
  ungroup() %>%
  mutate(
    pct_of_all.clusters = round((n / total_m1) * 100, 2),
    pct_of_sex.clusters = round((n / total_m2) * 100, 2)) %>%
  select(-total_m1, -total_m2)

datatable(df_count, filter = "top", rownames = FALSE)
write.csv(df_count, paste0(results_dir, "/cluster_to_subcluster_map.csv"))
```

## Cluster distance heatmap (batch corrected space)

```{r}
embeddings <- reducedDim(merged.sex.sce, "corrected")[, 1:30]
clusters   <- colData(merged.sex.sce)$label

centroids  <- aggregate(embeddings, by = list(cluster = clusters), FUN = mean)
rownames(centroids) <- centroids$cluster
centroids  <- centroids[, -1]

dist_matrix <- as.matrix(dist(centroids, method = "euclidean"))

pheatmap(dist_matrix,
         color          = colorRampPalette(c("lightyellow", "orange", "darkred"))(100),
         main           = "Pairwise distances between cluster centroids (batch corrected space)",
         display_numbers = TRUE,
         number_format  = "%.1f")
```

### Sex subcluster to all-cell cluster mapping

| sex subclusters | all cell clusters | Broad Annotation |
| --- | --- | --- |
| 1 | Oogonia/Oocytes | FGC (migrating) |
| 2 | Migrating PGCs | PGC |
| 3 | Migrating PGCs | Oogonia (stem cells) |
| 4 | Oogonia/Oocytes | Oocytes |
| 5 | Migrating PGCs | Cyst epithelial cell |
| 6 | Oogonia/Oocytes | Oogonia |
| 7 | Spermatocytes (cytokinetic) | SPC (cytokinetic) |
| 8 | Male germ cells (migrating) | MGC/SPC group2 |
| 9 | Spermatocytes (apoptotic) | MGC/SPC group1 |
| 10 | Male germ cells (migrating) / Spermatocytes (GPCR-activated) | MGC/SPC group3 |
| 11 | Male germ cells (migrating) / Spermatocytes (apoptotic) | MGC/SPC group2 |
| 12 | Male germ cells (migrating) / Spermatocytes (apoptotic) | MGC/SPC group1 |
| 13 | Spermatocytes (GPCR-activated) | MGC/SPC group3 |
| 14 | Spermatocytes (cytokinetic) / Male germ cells (migrating) / Spermatocytes (apoptotic) | MGC/SPC group2 |
| 15 | Male germ cells (migrating) / Spermatocytes (apoptotic) / Spermatocytes (GPCR-activated) | MGC/SPC group3 |

## Cells per broad annotation by sample

```{r, fig.height=8, fig.width=12}
p <- lapply(unique(merged.sex.sce$batch), function(f) {
  sce.f = merged.sex.sce[, which(merged.sex.sce$batch == f)]
  plotTSNE(sce.f, colour_by = "broad", text_colour = "black", point_size = .05) +
    ggtitle(f)
})
wrap_plots(p, ncol = 2)
```

```{r}
meta_df = as.data.frame(colData(merged.sex.sce))

sample_tally =
  meta_df %>%
  mutate(broad = as.factor(broad), batch = as.factor(batch)) %>%
  group_by(broad, batch, .drop = FALSE) %>%
  tally()

ggplot(sample_tally, aes(fill = batch, y = broad, x = n)) +
  geom_bar(position = "fill", stat = "identity") +
  xlab("Proportion of cells")
```

## Build combined (uncorrected) sex cell SCE

```{r, eval=TRUE}
all.sex.dec  <- lapply(all.sex.sce, modelGeneVar)

universe     <- Reduce(intersect, lapply(all.sex.sce, rownames))
all.sex.sce2 <- lapply(all.sex.sce, "[", i = universe,)
all.sex.dec2 <- lapply(all.sex.dec, "[", i = universe,)

all.sex.sce2[["TM_C00004374"]]@colData$cluster = "noclust"

combined.sex.sce <- do.call(cbind, all.sex.sce2)
```

# Cluster enriched genes

Top ~100 cluster enriched genes per broad annotation (all in `no.support_sex_scatter_cluster_marker_genes.csv`).

**Columns:** `cluster`, `Gene`, `p.value`, `FDR`, `summary.logFC`, `WangGeneID`, `description`

```{r, eval=FALSE}
scatter_markers = findMarkers(combined.sex.sce,
                               merged.sex.sce$broad,
                               test.type = "t",
                               block     = combined.sex.sce$sample,
                               direction = "up",
                               pval.type = "some")

saveRDS(scatter_markers,
        file = paste0(results_dir, "/no.support_sex_scatter_markers.Rds"))
```

```{r}
scatter_summary = list()
for (i in 1:length(scatter_markers)) {
  scatter_markers[[i]]$cluster = names(scatter_markers)[i]
  scatter_markers[[i]]$Gene    = rownames(scatter_markers[[i]])
  scatter_summary[[i]] = as.data.frame(scatter_markers[[i]]) %>%
    select(cluster, Gene, p.value, FDR, summary.logFC)
}

scatter_markers_df =
  do.call(rbind, scatter_summary) %>%
  left_join(gene_description, by = "Gene") %>%
  filter(FDR < .01)

write.csv(scatter_markers_df,
          file = paste0(results_dir, "/no.support_sex_scatter_cluster_marker_genes.csv"))

scatter_markers_top100 = scatter_markers_df %>%
  group_by(cluster) %>%
  slice_min(FDR, n = 100, with_ties = TRUE) %>%
  mutate(p.value       = formatC(signif(p.value, digits = 3)),
         FDR           = formatC(signif(FDR, digits = 3)),
         summary.logFC = round(summary.logFC, digits = 3))

datatable(scatter_markers_top100, filter = "top", rownames = FALSE)
```

## Comparison with known markers

```{r, fig.width=8}
marker_test_genes =
  scatter_markers_top100 %>%
  select(Gene, cluster)

plot_marker_overlap_heatmap(
  known_markers    = chung_davidson_cell_type_markers,
  test_genes       = marker_test_genes,
  background_n     = length(merged.sex.sce),
  gene_description = gene_description,
  overlaps_file    = paste0(results_dir, "/no.support_sex_scatter_cluster_marker_gene_overlaps.csv"),
  pvals_file       = paste0(results_dir, "/no.support_sex_scatter_cluster_marker_overlap_pvalues.csv"),
  plot_file        = paste0(results_dir, "/no.support_sex_scatter_cluster_overlap_heatmap.png")
)
```

## Gene ontology of top 100 cluster enriched genes

**Columns:** `GOID`, `GOName`, `nGOTermGenes`, `Clusters`, `nClusterEnrichedGenes`,
`nSharedGenes`, `Pval`, `FDR`

```{r}
scores = run_go_enrichment(
  gaf             = gaf,
  gene_description = gene_description,
  go_names        = go_names,
  test_genes      = marker_test_genes,
  background_n    = length(merged.sex.sce),
  overlaps_file   = paste0(results_dir, "/no.support_sex_scatter_cluster_top_100_marker_gene_GO_overlaps.csv"),
  pvals_file      = paste0(results_dir, "/no.support_sex_scatter_cluster_top_100_marker_gene_GO_pvals.csv"),
  require_shared  = TRUE
)

datatable(scores, filter = "top", rownames = FALSE)
```

## Gene ontology of all significant cluster enriched genes (FDR < 0.01)

```{r}
marker_test_genes_all =
  scatter_markers_df %>%
  select(Gene, cluster)

scores = run_go_enrichment(
  gaf             = gaf,
  gene_description = gene_description,
  go_names        = go_names,
  test_genes      = marker_test_genes_all,
  background_n    = length(merged.sex.sce),
  overlaps_file   = paste0(results_dir, "/no.support_sex_scatter_cluster_all_sig_marker_gene_GO_overlaps_all.csv"),
  pvals_file      = paste0(results_dir, "/no.support_sex_scatter_all_sig_cluster_marker_gene_GO_pvals.csv"),
  require_shared  = TRUE
)

datatable(scores, filter = "top", rownames = FALSE)
```

## Known marker bubble plots

```{r, eval=TRUE, fig.show="hide"}
nested_df   <- chung_davidson_markers %>% group_by(CellType) %>% tidyr::nest()
list_of_dfs <- nested_df$data
names(list_of_dfs) <- nested_df$CellType
names(list_of_dfs)  = gsub("\\(", "", names(list_of_dfs))

file_names = gsub(" ", "_", names(list_of_dfs))
file_names = paste0(file_names, "_germ_cell_clusters")

ct = colData(merged.sex.sce)$broad
combined.sex.sce$CellType = ct
combined.sex.sce$CellType = factor(combined.sex.sce$CellType, levels = unique(ct))
combined.sex.sce$cluster  = ct

plot_dir = paste0(results_dir, "/germ_cell_bubble")
if (!dir.exists(plot_dir)) dir.create(plot_dir)

scatter_markers_df_read = read.csv(
  file    = paste0(results_dir, "/no.support_sex_scatter_cluster_marker_genes.csv"),
  row.names = 1)

markers = scatter_markers_df_read %>%
  dplyr::rename(CellType = "cluster") %>%
  select(WangGeneID, CellType, FDR, summary.logFC) %>%
  dplyr::rename(Gene = "WangGeneID")

for (i in 1:length(list_of_dfs)) {
  plot_marker_bubbles(
    sce              = combined.sex.sce,
    markers          = markers,
    gene_description = gene_description,
    cell_type_name   = names(list_of_dfs)[i],
    cell_type_genes  = list_of_dfs[[i]],
    anno             = unique(ct),
    results_dir      = plot_dir,
    file_name        = file_names[i]
  )
}
```

# Sex cluster annotations

| sex subcluster | Narrow annotation |
| --- | --- |
| 1 | Female germ cells (migrating) |
| 2 | Primordial germ cells (migrating, differentiated, cyst) |
| 3 | Oogonia (mitotic, cyst, stem cells) |
| 4 | Oocytes |
| 5 | Cyst epithelial cells |
| 6 | Oogonia (mitotic, cyst, cell fate committed) |
| 7 | Spermatocytes (cytokinetic) |
| 8 | Male germ cells (migrating, cell fate committed, differentiated) |
| 9 | Spermatocytes (meiosis arrested) |
| 10 | Spermatocytes (migrating, S-phase, cyst connected with cell bridges, cell fate committed, GPCR-activated) |
| 11 | Spermatocytes (apoptotic, meiosis arrested) |
| 12 | Male germ cells (migrating, differentiated) |
| 13 | Spermatocytes (angiogenic, cell fate committed, GPCR-activated) |
| 14 | Spermatocytes (cell fate committed) |
| 15 | Male germ cells (migrating, cell fate committed, differentiated, wounded) |

# Pairwise differential expression

Each section subsets two cell types, runs `findMarkers`, and reports GO enrichment
of the top 100 and all significant (FDR < 0.01) DE genes.

## PGC vs Oogonia stem cells (clusters 2 vs 3)

```{r}
de_23 = run_pairwise_de(
  combined_sce   = combined.sex.sce,
  merged_sce     = merged.sex.sce,
  cluster_labels = c(2, 3),
  gene_description = gene_description,
  results_dir    = results_dir,
  file_prefix    = "no.support_sex_scatter_PCGvsOSC"
)

datatable(de_23$top100, filter = "top", rownames = FALSE)
```

### GO enrichment — top 100 DE genes

```{r}
scores = run_go_enrichment(
  gaf             = gaf,
  gene_description = gene_description,
  go_names        = go_names,
  test_genes      = de_23$top100 %>% select(Gene, cluster),
  background_n    = length(merged.sex.sce),
  overlaps_file   = paste0(results_dir, "/no.support_sex_scatter_cluster_2vs3_top_100_DE_gene_GO_overlaps.csv"),
  pvals_file      = paste0(results_dir, "/no.support_sex_scatter_cluster_2vs3_top_100_DE_gene_GO_pvals.csv"),
  require_shared  = TRUE
)

datatable(scores, filter = "top", rownames = FALSE)
```

### GO enrichment — all significant DE genes (FDR < 0.01)

```{r}
scores = run_go_enrichment(
  gaf             = gaf,
  gene_description = gene_description,
  go_names        = go_names,
  test_genes      = de_23$all %>% select(Gene, cluster),
  background_n    = length(merged.sex.sce),
  overlaps_file   = paste0(results_dir, "/no.support_sex_scatter_cluster_2vs3_sig_DE_gene_GO_overlaps.csv"),
  pvals_file      = paste0(results_dir, "/no.support_sex_scatter_cluster_2vs3_sig_DE_gene_GO_pvals.csv"),
  require_shared  = TRUE
)

datatable(scores, filter = "top", rownames = FALSE)
```

## PGC vs Female germ cells (clusters 2 vs 1)

```{r}
de_21 = run_pairwise_de(
  combined_sce   = combined.sex.sce,
  merged_sce     = merged.sex.sce,
  cluster_labels = c(2, 1),
  gene_description = gene_description,
  results_dir    = results_dir,
  file_prefix    = "no.support_sex_scatter_PCGvsFGC"
)

datatable(de_21$top100, filter = "top", rownames = FALSE)
```

### GO enrichment — top 100 DE genes

```{r}
scores = run_go_enrichment(
  gaf             = gaf,
  gene_description = gene_description,
  go_names        = go_names,
  test_genes      = de_21$top100 %>% select(Gene, cluster),
  background_n    = length(merged.sex.sce),
  overlaps_file   = paste0(results_dir, "/no.support_sex_scatter_cluster_2v1_top_100_DE_gene_GO_overlaps.csv"),
  pvals_file      = paste0(results_dir, "/no.support_sex_scatter_cluster_2vs1_top_100_DE_gene_GO_pvals.csv"),
  require_shared  = TRUE
)

datatable(scores, filter = "top", rownames = FALSE)
```

### GO enrichment — all significant DE genes (FDR < 0.01)

```{r}
scores = run_go_enrichment(
  gaf             = gaf,
  gene_description = gene_description,
  go_names        = go_names,
  test_genes      = de_21$all %>% select(Gene, cluster),
  background_n    = length(merged.sex.sce),
  overlaps_file   = paste0(results_dir, "/no.support_sex_scatter_cluster_2vs1_sig_DE_gene_GO_overlaps.csv"),
  pvals_file      = paste0(results_dir, "/no.support_sex_scatter_cluster_2vs1_sig_DE_gene_GO_pvals.csv"),
  require_shared  = TRUE
)

datatable(scores, filter = "top", rownames = FALSE)
```

## PGC vs Male germ cells (clusters 2 vs 12)

```{r}
de_212 = run_pairwise_de(
  combined_sce   = combined.sex.sce,
  merged_sce     = merged.sex.sce,
  cluster_labels = c(2, 12),
  gene_description = gene_description,
  results_dir    = results_dir,
  file_prefix    = "no.support_sex_scatter_PCGvsMGC"
)

datatable(de_212$top100, filter = "top", rownames = FALSE)
```

### GO enrichment — top 100 DE genes

```{r}
scores = run_go_enrichment(
  gaf             = gaf,
  gene_description = gene_description,
  go_names        = go_names,
  test_genes      = de_212$top100 %>% select(Gene, cluster),
  background_n    = length(merged.sex.sce),
  overlaps_file   = paste0(results_dir, "/no.support_sex_scatter_cluster_2v12_top_100_DE_gene_GO_overlaps.csv"),
  pvals_file      = paste0(results_dir, "/no.support_sex_scatter_cluster_2vs12_top_100_DE_gene_GO_pvals.csv"),
  require_shared  = TRUE
)

datatable(scores, filter = "top", rownames = FALSE)
```

### GO enrichment — all significant DE genes (FDR < 0.01)

```{r}
scores = run_go_enrichment(
  gaf             = gaf,
  gene_description = gene_description,
  go_names        = go_names,
  test_genes      = de_212$all %>% select(Gene, cluster),
  background_n    = length(merged.sex.sce),
  overlaps_file   = paste0(results_dir, "/no.support_sex_scatter_cluster_2vs12_sig_DE_gene_GO_overlaps.csv"),
  pvals_file      = paste0(results_dir, "/no.support_sex_scatter_cluster_2vs12_sig_DE_gene_GO_pvals.csv"),
  require_shared  = TRUE
)

datatable(scores, filter = "top", rownames = FALSE)
```

# Session info

```{r}
sessionInfo()
```
