---
title: "03_trajectory_analysis"
author: "Stephanie Hickey"
date: "`r Sys.Date()`"
output:
  html_document:
    code_folding: hide
    toc: true
    toc_float:
      toc_collapsed: true
params:
    project_dir: /mnt/ufs18/rs-013/bioinformaticsCore/projects/chong_davidson/BCC125_sea_lamprey_scRNA
    project_name: BCC125_sea_lamprey_scRNA
    lab: Weiming Li
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

```{r, echo=FALSE, results="asis"}
cat("\n")
cat("#### Lab:", params$lab, " \n")
cat("#### Project:", params$project_name, " \n")
```

```{r}
library(tidyverse)
library(SingleCellExperiment)
library(scater)
library(scran)
library(patchwork)
library(RColorBrewer)
library(readxl)
library(monocle3)
library(batchelor)
library(DT)
library(TSCAN)
library(rtracklayer)
library(BiocSingular)
library(ggforce)
library(ggraph)
library(tidygraph)
library(scatterpie)
library(ggrepel)
source(paste0(params$project_dir, "/src/scRNA_seq_functions.R"))
```

```{r}
results_dir = paste0(params$project_dir, "/results/20260615")
data_dir    = paste0(params$project_dir, "/data")
```

```{r, echo=FALSE}
# load pre-computed objects for rendering
Robject_dir    = paste0(params$project_dir, "/results/20260312/Robjects")
all.sex.sce    = readRDS(paste0(Robject_dir, "/no.support.sex.sce.Rds"))
merged.sex.sce = readRDS(paste0(Robject_dir, "/no.support.merged.sex.sce.Rds"))
pseudo         = readRDS(paste0(Robject_dir, "/no.support.merged.sex.pseudo.Rds"))
```

```{r}
# broad + narrow annotations for sex clusters
anno = c(
  "FGC.migrating",
  "PGC",
  "Oogonia.stem.cells",
  "Oocytes",
  "Cyst.epithelial.cells",
  "Oogonia",
  "SPC.cytokinetic",
  "MGC.SPC.group2",
  "MGC.SPC.group1",
  "MGC.SPC.group3",
  "MGC.SPC.group2",
  "MGC.SPC.group1",
  "MGC.SPC.group3",
  "MGC.SPC.group2",
  "MGC.SPC.group3"
)

colData(merged.sex.sce)$broad = anno[colData(merged.sex.sce)$label]

narrow = c(
  "Female germ cells (migrating)",
  "Primordial germ cells (migrating, differentiated, cyst)",
  "Oogonia (mitotic, cyst, stem cells)",
  "Oocytes",
  "Cyst epithelial cells",
  "Oogonia (mitotic, cyst, cell fate committed)",
  "Spermatocytes (cytokinetic)",
  "Male germ cells (migrating, cell fate committed, differentiated)",
  "Spermatocytes (meiosis arrested)",
  "Spermatocytes (migrating, S-phase, cyst connected with cell bridges, cell fate committed, GPCR-activated)",
  "Spermatocytes (apoptotic, meiosis arrested)",
  "Male germ cells (migrating, differentiated)",
  "Spermatocytes (angiogenic, cell fate committed, GPCR-activated)",
  "Spermatocytes (cell fate committed)",
  "Male germ cells (migrating, cell fate committed, differentiated, wounded)"
)

colData(merged.sex.sce)$narrow = narrow[colData(merged.sex.sce)$label]
```

```{r}
# reconstruct uncorrected combined sex SCE from all.sex.sce
all.sex.dec  <- lapply(all.sex.sce, modelGeneVar)

universe     <- Reduce(intersect, lapply(all.sex.sce, rownames))
all.sex.sce2 <- lapply(all.sex.sce, "[", i = universe,)

all.sex.sce2[["TM_C00004374"]]@colData$cluster = "noclust"

combined.sex.sce <- do.call(cbind, all.sex.sce2)
colData(combined.sex.sce)$broad = colData(merged.sex.sce)$broad
colData(combined.sex.sce)$label = colData(merged.sex.sce)$label
```

```{r}
# gene descriptions
gene_description =
  read_excel(paste0(data_dir, "/41467_2024_55019_MOESM3_ESM.xlsx"), skip = 1) %>%
  dplyr::rename(WangGeneID = "CountMatrix_GeneID",
                Gene        = "NCBI_GTF_GENEID",
                description = "NCBI_GTF_ProteinInfo") %>%
  dplyr::select(Gene, WangGeneID, description) %>%
  filter(!is.na(Gene)) %>%
  filter(!WangGeneID %in% c("LOC116945425.1", "S100A1.1"))

gtf = as.data.frame(readGFF("/mnt/research/bioinformaticsCore/shared/Genomes/Petromyzon_marinus/GCF_010993605.1_kPetMar1.pri_genomic.gtf"))

gtf_filt =
  gtf %>%
  filter(gbkey %in% c("mRNA", "misc_RNA", "ncRNA", "tRNA", "rRNA")) %>%
  mutate(product = gsub(", transcript variant.*", "", product)) %>%
  dplyr::select(gene_id, product) %>%
  distinct() %>%
  dplyr::rename(Gene = "gene_id", description = "product") %>%
  mutate(WangGeneID = Gene)

notWang      = gtf_filt$Gene[!gtf_filt$Gene %in% gene_description$Gene]
gtf_add      = gtf_filt %>% filter(Gene %in% notWang)
gene_description = rbind(gene_description, gtf_add)
```

```{r}
# GO annotation
gaf = read.delim("/mnt/research/bioinformaticsCore/shared/Genomes/Petromyzon_marinus/GCF_010993605.1_kPetMar1.pri_gene_ontology.gaf",
                 header = FALSE, comment = "!")

gaf <- gaf %>%
  filter(!grepl("NOT", V4)) %>%
  dplyr::select(V3, V5) %>%
  dplyr::rename(GOID = "V5", Gene = "V3") %>%
  distinct()

keep_go = gaf %>%
  group_by(GOID) %>% tally() %>%
  filter(n > 10, n < 200) %>%
  pull(GOID)

gaf = gaf %>% filter(GOID %in% keep_go)

go_names = read.delim("/mnt/research/bioinformaticsCore/shared/Genomes/Petromyzon_marinus/GOs_names.txt",
                      header = FALSE)
colnames(go_names) = c("GOID", "GOName")
```

# Sex cell developmental trajectory analysis

## Trajectory in TSNE space

```{r}
set.seed(1)
by.cluster <- aggregateAcrossCells(merged.sex.sce,
                                    ids = merged.sex.sce$broad,
                                    use.assay.type = "reconstructed")

centroids <- reducedDim(by.cluster, "corrected")
mst       <- createClusterMST(centroids, clusters = NULL)
line.data <- reportEdges(by.cluster, mst = mst, clusters = NULL, use.dimred = "TSNE")

plotTSNE(merged.sex.sce, colour_by = "broad", point_size = .5) +
  geom_line(data = line.data, mapping = aes(x = TSNE1, y = TSNE2, group = edge)) +
  guides(color = guide_legend(override.aes = list(size = 5)))
```

## Trajectory in batch-corrected gene expression space

The trajectory and clustering were calculated in batch-corrected expression space,
so this view is more interpretable than TSNE.

```{r}
line.data_corrected <- reportEdges(by.cluster, mst = mst, clusters = NULL,
                                    use.dimred = "corrected")

plotReducedDim(merged.sex.sce, colour_by = "broad", dimred = "corrected",
               point_size = .4) +
  geom_line(data = line.data_corrected,
            mapping = aes(x = dim1, y = dim2, group = edge)) +
  guides(color = guide_legend(override.aes = list(size = 5)))
```

## Pseudotime calculation

```{r, eval=FALSE}
set.seed(1)
pseudo <- quickPseudotime(merged.sex.sce,
                           clusters   = merged.sex.sce$broad,
                           use.dimred = "corrected",
                           outgroup   = FALSE,
                           start      = "PGC")

saveRDS(pseudo, file = paste0(results_dir, "/no.support.merged.sex.pseudo.Rds"))
```

## Trajectory node graph

Top: Pie charts show the proportion of cells from each sample per node. Bottom: Node color corresponds to pseudotim.e

```{r, fig.width=8, fig.height=7}
meta_df = as.data.frame(colData(merged.sex.sce))

sample_tally =
  meta_df %>%
  mutate(broad = as.factor(broad), batch = as.factor(batch)) %>%
  group_by(broad, batch, .drop = FALSE) %>%
  tally()

gr <- as_tbl_graph(pseudo$mst)

node_meta =
  sample_tally %>%
  mutate(batch = case_when(
    batch == "TF_C00004337"   ~ "Female",
    batch == "TM_C00004374"   ~ "Male",
    batch == "UND1_C00004336" ~ "Undifferentiated.1",
    batch == "UND2_C00004362" ~ "Undifferentiated.2"
  )) %>%
  tidyr::pivot_wider(names_from = batch, values_from = n) %>%
  mutate(totalCells = sum(Female, Male, Undifferentiated.1, Undifferentiated.2,
                          na.rm = TRUE)) %>%
  dplyr::rename(name = "broad")

common.pseudo <- averagePseudotime(pseudo$ordering)
merged.sex.sce$Pseudotime <- common.pseudo

mean_pseudo =
  as.data.frame(colData(merged.sex.sce)) %>%
  group_by(broad) %>%
  summarise(mean_pseudo = mean(Pseudotime)) %>%
  dplyr::rename(name = "broad")

gr = gr %>%
  activate(nodes) %>%
  left_join(node_meta) %>%
  arrange(desc(totalCells)) %>%
  left_join(mean_pseudo)

layout_data <- create_layout(gr, layout = "stress")

gr_plot =
  ggraph(layout_data) +
  geom_edge_link() +
  geom_scatterpie(
    aes(x, y),
    cols     = c("Female", "Male", "Undifferentiated.1", "Undifferentiated.2"),
    data     = layout_data,
    colour   = NA,
    pie_scale = 1.5
  ) +
  scale_fill_manual(
    name   = "Sample",
    values = c(
      "Female"           = "#FF6DB6",
      "Male"             = "#006ddb",
      "Undifferentiated.1" = "#009292",
      "Undifferentiated.2" = "#b66ddf"
    )
  ) +
  guides(fill = guide_legend(keywidth = .8, keyheight = .8)) +
  geom_node_text(aes(label = name), color = "black") +
  theme_void(base_size = 8) +
  theme(
    legend.position    = "left",
    legend.justification = "center",
    legend.box.just    = "center",
    legend.title       = element_text(size = 8),
    legend.text        = element_text(size = 8)
  )

gr_pseudo =
  ggraph(layout_data) +
  geom_edge_link() +
  geom_node_point(aes(color = mean_pseudo), size = 13) +
  scale_color_viridis_c(option = "viridis") +
  geom_node_text(aes(label = name), vjust = .75, size = 2.8, color = "red") +
  scale_x_continuous(expand = expansion(mult = 0.15)) +
  scale_y_continuous(expand = expansion(mult = 0.15)) +
  theme_void(base_size = 8) +
  theme(
    legend.position    = "left",
    legend.justification = "center",
    legend.box.just    = "center",
    legend.title       = element_text(size = 8),
    legend.text        = element_text(size = 8)
  )

(gr_plot + coord_fixed()) / (gr_pseudo + coord_fixed())

ggsave(file = paste0(results_dir, "/germ.only_pseudo_mst.pdf"), width = 8, height = 7)
```

# Pseudotime-associated genes

Top ~100 genes by FDR changing along each path. All significant genes are saved in
`no.support_sex_pseudotime_de_genes.csv`.

**Columns:** `Path`, `SYMBOL`, `WangGeneID`, `description`, `logFC`, `p.value`, `FDR`

```{r}
pseudo_mat  <- pathStat(pseudo$ordering)
test_pseudo <- TSCAN::testPseudotime(combined.sex.sce, pseudotime = pseudo_mat)

test_pseudo_filt = list()
for (i in 1:length(test_pseudo)) {
  test_pseudo[[i]]$SYMBOL <- rownames(combined.sex.sce)
  test_pseudo[[i]]$Path   <- names(test_pseudo)[i]

  test_pseudo_filt[[i]] =
    as.data.frame(test_pseudo[[i]]) %>%
    filter(FDR < .01) %>%
    arrange(desc(abs(logFC)))
}

symbol_description = gene_description %>% dplyr::rename(SYMBOL = "Gene")

test_pseudo_filt_df =
  do.call(rbind, test_pseudo_filt) %>%
  left_join(symbol_description, by = "SYMBOL")

test_pseudo_filt_df = test_pseudo_filt_df[,
  c("Path", "SYMBOL", "WangGeneID", "description", "logFC", "p.value", "FDR")]

write.csv(test_pseudo_filt_df,
          file = paste0(results_dir, "/no.support_sex_pseudotime_de_genes.csv"))

test_pseudo100 =
  test_pseudo_filt_df %>%
  mutate(Path = case_when(
    logFC > 0 ~ paste0(Path, ".Upregulated"),
    logFC < 0 ~ paste0(Path, ".Downregulated"))) %>%
  group_by(Path) %>%
  slice_min(order_by = FDR, n = 100) %>%
  mutate(logFC   = round(logFC, digits = 2),
         p.value = formatC(signif(p.value, digits = 2)),
         FDR     = formatC(signif(FDR, digits = 2)))

test_pseudo =
  test_pseudo_filt_df %>%
  mutate(Path = case_when(
    logFC > 0 ~ paste0(Path, ".Upregulated"),
    logFC < 0 ~ paste0(Path, ".Downregulated"))) %>%
  group_by(Path) %>%
  mutate(logFC   = round(logFC, digits = 2),
         p.value = formatC(signif(p.value, digits = 2)),
         FDR     = formatC(signif(FDR, digits = 2)))

datatable(test_pseudo100, filter = "top", rownames = FALSE)
```

## Trajectory heatmaps (top 25 genes per path and direction)

```{r}
colData(combined.sex.sce)$pseudotime_TSCAN <- averagePseudotime(pseudo$ordering)

test_pseudo_SYMBOL_list <-
  test_pseudo_filt_df %>%
  dplyr::mutate(Path = dplyr::case_when(
    logFC > 0 ~ paste0(Path, ".Upregulated"),
    logFC < 0 ~ paste0(Path, ".Downregulated"))) %>%
  dplyr::group_by(Path) %>%
  arrange(FDR, desc(abs(logFC))) %>%
  dplyr::slice_head(n = 25) %>%
  dplyr::summarise(SYMBOL = list(SYMBOL), .groups = "drop") %>%
  tibble::deframe()

p_list      <- list()
n_cells_list <- list()
for (p in names(test_pseudo_SYMBOL_list)) {
  gene_ids = test_pseudo_SYMBOL_list[[p]]

  gene_labels = gene_description %>%
    filter(Gene %in% gene_ids) %>%
    pull(WangGeneID)

  path_name = sub("\\.(Upregulated|Downregulated)$", "", p)
  cell_subset = as.data.frame(pseudo_mat) %>%
    rownames_to_column(var = "Barcode") %>%
    filter(!is.na(.data[[path_name]])) %>%
    pull(Barcode)

  n_cells_list[[p]] <- length(cell_subset)

  p_list[[p]] =
    plot_pseudotime_heatmap(combined.sex.sce,
                             gene_ids,
                             gene_labels,
                             cell_subset,
                             pseudotime_column = "pseudotime_TSCAN")
}


```

### PGC → SPC.cytokinetic path

```{r, fig.width=7, fig.height=4}
make_heatmap_pair <- function(p_list, path_name, n_cells_list = NULL) {
  up_name   <- paste0(path_name, ".Upregulated")
  down_name <- paste0(path_name, ".Downregulated")

  up_label   <- if (!is.null(n_cells_list))
    paste0(up_name,   " (n=", n_cells_list[[up_name]],   ")") else up_name
  down_label <- if (!is.null(n_cells_list))
    paste0(down_name, " (n=", n_cells_list[[down_name]], ")") else down_name

  title_up   <- ggplot2::ggplot() +
    ggplot2::labs(title = up_label) + ggplot2::theme_void() +
    ggplot2::theme(plot.title = ggplot2::element_text(size = 8, face = "bold"),
                   plot.margin = ggplot2::margin(t = 10))

  title_down <- ggplot2::ggplot() +
    ggplot2::labs(title = down_label) + ggplot2::theme_void() +
    ggplot2::theme(plot.title = ggplot2::element_text(size = 8, face = "bold"),
                   plot.margin = ggplot2::margin(t = 10))

  col1 <- (title_up   / p_list[[up_name]][[1]]   / p_list[[up_name]][[2]]) +
    patchwork::plot_layout(heights = c(0.02, 0.025, 1))
  col2 <- (title_down / p_list[[down_name]][[1]] / p_list[[down_name]][[2]]) +
    patchwork::plot_layout(heights = c(0.02, 0.025, 1))

  (col1 | col2) +
    patchwork::plot_layout(guides = "collect") &
    ggplot2::theme(legend.position = "right")
}

make_heatmap_pair(p_list, "SPC.cytokinetic", n_cells_list)
ggsave(file = paste0(results_dir, "/SPC.cytokinetic_heatmaps.png"), width = 8, height = 5)
```

### PGC → Oocytes path

```{r, fig.width=7, fig.height=4}
make_heatmap_pair(p_list, "Oocytes", n_cells_list)
ggsave(file = paste0(results_dir, "/Oocytes_heatmaps.png"), width = 8, height = 5)
```

# Gene ontology of pseudotime-associated genes

**Columns:** `GOID`, `GOName`, `nGOTermGenes`, `Path`, `nPathDEGenes`,
`nSharedGenes`, `Pval`, `FDR`

## All significant genes (FDR < 0.01)

- Overlap p-values: `no.support_sex_pseudotime_de_GO_pvals.csv`
- Overlapping genes: `no.support_sex_pseudotime_de_GO_overlaps.csv`

```{r}
test_pseudo_genes = test_pseudo %>% select(SYMBOL, Path)

go_ol <- phyperOverlapSets(gaf, test_pseudo_genes, length(merged.sex.sce))

shared_desc <- gene_description %>% dplyr::rename(SharedGene = "Gene")

go_int =
  go_ol$intersecting.genes %>%
  left_join(shared_desc, by = "SharedGene") %>%
  filter(!SharedGene == "none") %>%
  dplyr::rename(GOID = "Group1", Path = "Group2") %>%
  left_join(go_names, by = "GOID") %>%
  relocate(GOName, .after = GOID)

write.csv(go_int,
          file = paste0(results_dir, "/no.support_sex_pseudotime_de_GO_overlaps.csv"),
          row.names = FALSE)

scores =
  go_ol$pval.scores %>%
  dplyr::rename(GOID           = "Group1",
                Path           = "Group2",
                nGOTermGenes   = "nGroup1",
                nPathDEGenes   = "nGroup2") %>%
  mutate(FDR = p.adjust(Pval, "BH")) %>%
  select(-Enrichment)

scores =
  left_join(go_names, scores, by = "GOID") %>%
  filter(!is.na(nGOTermGenes)) %>%
  filter(Pval < .05)

write.csv(scores,
          file = paste0(results_dir, "/no.support_sex_pseudotime_de_GO_pvals.csv"),
          row.names = FALSE)

datatable(scores, filter = "top", rownames = FALSE)
```

## Top 100 genes per path and direction

- Overlap p-values: `no.support_sex_pseudotime_top100_de_GO_pvals.csv`
- Overlapping genes: `no.support_sex_pseudotime_top100_de_GO_overlaps.csv`

```{r}
test_pseudo_genes100 = test_pseudo100 %>% select(SYMBOL, Path)

go_ol <- phyperOverlapSets(gaf, test_pseudo_genes100, length(merged.sex.sce))

go_int =
  go_ol$intersecting.genes %>%
  left_join(shared_desc, by = "SharedGene") %>%
  filter(!SharedGene == "none") %>%
  dplyr::rename(GOID = "Group1", Path = "Group2") %>%
  left_join(go_names, by = "GOID") %>%
  relocate(GOName, .after = GOID)

write.csv(go_int,
          file = paste0(results_dir, "/no.support_sex_pseudotime_top100_de_GO_overlaps.csv"),
          row.names = FALSE)

scores =
  go_ol$pval.scores %>%
  dplyr::rename(GOID           = "Group1",
                Path           = "Group2",
                nGOTermGenes   = "nGroup1",
                nPathDEGenes   = "nGroup2") %>%
  mutate(FDR = p.adjust(Pval, "BH")) %>%
  select(-Enrichment)

scores =
  left_join(go_names, scores, by = "GOID") %>%
  filter(!is.na(nGOTermGenes)) %>%
  filter(Pval < .05)

write.csv(scores,
          file = paste0(results_dir, "/no.support_sex_pseudotime_top100_de_GO_pvals.csv"),
          row.names = FALSE)

datatable(scores, filter = "top", rownames = FALSE)
```

# Cell metadata export

```{r}
germ_meta = colData(merged.sex.sce)
germ_meta$batch           = NULL
germ_meta$all.scatter.clust = NULL

colnames(germ_meta) = c("germ.cluster", "germ.broad.CellType",
                         "germ.narrow.CellType", "germ.Pseudotime")

germ_tsne      = reducedDim(merged.sex.sce, "TSNE")
germ_corrected = reducedDim(merged.sex.sce, "corrected")
colnames(germ_corrected) = paste0("germ.fastMNN.corrected.", 1:50)

germ_meta$germ.TSNE1 = germ_tsne[, "TSNE1"]
germ_meta$germ.TSNE2 = germ_tsne[, "TSNE2"]
germ_meta = cbind(germ_meta, germ_corrected)
germ_meta$barcode = rownames(germ_meta)

# load the all-cells merged object for the full metadata join
merged.sce    = readRDS(paste0(Robject_dir, "/merged.sce.Rds"))

all_meta = colData(merged.sce)
all_meta$monocole_cluster = NULL
all_meta$barcode = rownames(all_meta)

all_tsne      = reducedDim(merged.sce, "TSNE")
all_corrected = reducedDim(merged.sce, "corrected")
colnames(all_corrected) = paste0("fastMNN.corrected.", 1:50)

all_meta$TSNE1 = all_tsne[, "TSNE1"]
all_meta$TSNE2 = all_tsne[, "TSNE2"]
all_meta = cbind(all_meta, all_corrected)

meta_final =
  left_join(as.data.frame(all_meta), as.data.frame(germ_meta), by = "barcode") %>%
  relocate(barcode, .before = "batch")

write.table(meta_final,
            file      = paste0(results_dir, "/cell_metadata.tsv"),
            row.names = FALSE,
            sep       = "\t",
            quote     = FALSE)

zip(zipfile = paste0(results_dir, "/cell_metadata.tsv.gz"),
    files   = paste0(results_dir, "/cell_metadata.tsv"))
```

# Session info

```{r}
sessionInfo()
```
