---
title: "05_compare_doublet_removal"
author: "Stephanie Hickey"
date: "`r Sys.Date()`"
output:
  html_document:
    code_folding: hide
    toc: true
    toc_float:
      toc_collapsed: true
params:
    project_dir: /mnt/ufs18/rs-013/bioinformaticsCore/projects/chong_davidson/BCC125_sea_lamprey_scRNA
    project_name: BCC125_sea_lamprey_scRNA
    lab: Weiming Li
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

```{r, echo=FALSE, results="asis"}
cat("\n")
cat("#### Lab:", params$lab, " \n")
cat("#### Project:", params$project_name, " \n")
```

This document compares the with-doublets analysis
(`01_all_cells_clustering_annotation.Rmd`) to the doublet-filtered reanalysis
(`01_all_cells_clustering_annotation_no_doublets.Rmd`), using two independent
lines of evidence: whether individual cells common to both analyses land in
the same or different clusters, and whether each analysis's cluster marker
genes overlap. These two lines of evidence are meant to resolve the `TODO`
notes left in `02_sex_cell_subanalysis_no_doublets.Rmd` and
`03_trajectory_analysis_no_doublets.Rmd` about which no-doublets cluster
numbers correspond to which with-doublets cluster numbers.

```{r}
library(tidyverse)
library(SingleCellExperiment)
library(DT)
source(paste0(params$project_dir, "/src/scRNA_seq_functions.R"))
```

```{r}
results_dir       = paste0(params$project_dir, "/results/20260803_comparison")
with_dbl_robjects = paste0(params$project_dir, "/results/20260312/Robjects")
with_dbl_results  = paste0(params$project_dir, "/results/20260615")
no_dbl_dir        = paste0(params$project_dir, "/results/20260803_no_doublets")
if (!dir.exists(results_dir)) dir.create(results_dir, recursive = TRUE)
```

With-doublets `merged.sce`/`combined.sce` come from `results/20260312/Robjects`,
the same source `04_doublet_qc.Rmd` treats as canonical, and with-doublets
marker gene tables come from `results/20260615`, where
`01_all_cells_clustering_annotation.Rmd` currently writes them. No-doublets
objects and tables all come from the single directory
`results/20260803_no_doublets`.

```{r}
merged.sce.A   = readRDS(paste0(with_dbl_robjects, "/merged.sce.Rds"))
combined.sce.A = readRDS(paste0(with_dbl_robjects, "/combined.sce.Rds"))

merged.sce.B   = readRDS(paste0(no_dbl_dir, "/merged.sce.Rds"))
combined.sce.B = readRDS(paste0(no_dbl_dir, "/combined.sce.Rds"))

gene_description = read.csv(paste0(with_dbl_results, "/gene_descritption.csv"))
```

# Cell-level cluster correspondence

For cells present in both analyses, this cross-tabulates the with-doublets
cluster label against the no-doublets cluster label. Cells only in the
with-doublets object are candidates for having been removed as doublets (or
otherwise dropped); cells only in the no-doublets object would indicate a
barcode mismatch and are investigated rather than assumed away.

```{r}
common_cells    = intersect(colnames(merged.sce.A), colnames(merged.sce.B))
only_A          = setdiff(colnames(merged.sce.A), colnames(merged.sce.B))
only_B          = setdiff(colnames(merged.sce.B), colnames(merged.sce.A))
```

* `r length(common_cells)` cells are common to both analyses.
* `r length(only_A)` cells are in the with-doublets analysis only (candidates for having been removed as doublets).
* `r length(only_B)` cells are in the no-doublets analysis only.

```{r}
if (length(only_B) > 0) {
  warning(paste0(length(only_B), " cell(s) appear in the no-doublets analysis ",
                 "but not the with-doublets analysis. This should not happen if ",
                 "the no-doublets pipeline only removes cells from the same ",
                 "starting set; investigate before trusting the comparison below."))
}
```

```{r}
label_df = data.frame(
  barcode = common_cells,
  label_A = as.integer(as.character(
    colData(merged.sce.A)$label[match(common_cells, colnames(merged.sce.A))])),
  label_B = as.integer(as.character(
    colData(merged.sce.B)$label[match(common_cells, colnames(merged.sce.B))]))
)

contingency = label_df %>%
  dplyr::count(label_A, label_B, name = "n_cells") %>%
  group_by(label_A) %>%
  mutate(pct_of_A_cluster = 100 * n_cells / sum(n_cells)) %>%
  ungroup()

write.csv(contingency, file = paste0(results_dir, "/cluster_contingency_table.csv"),
          row.names = FALSE)

datatable(contingency %>% mutate(pct_of_A_cluster = round(pct_of_A_cluster, 1)),
          filter = "top", rownames = FALSE)
```

## Best-matching no-doublets cluster per with-doublets cluster

For each with-doublets cluster, the no-doublets cluster that captures the
largest share of its cells. A low `pct_of_A_cluster` means that
with-doublets cluster's cells are split across several no-doublets clusters,
rather than mapping cleanly to one.

```{r}
best_match = contingency %>%
  group_by(label_A) %>%
  slice_max(order_by = n_cells, n = 1, with_ties = FALSE) %>%
  ungroup() %>%
  arrange(label_A) %>%
  dplyr::rename(best_match_label_B = label_B)

write.csv(best_match, file = paste0(results_dir, "/cluster_best_match_by_cells.csv"),
          row.names = FALSE)

datatable(best_match %>% mutate(pct_of_A_cluster = round(pct_of_A_cluster, 1)),
          rownames = FALSE)
```

## Heatmap

Each tile is the percent of a with-doublets cluster's cells (rows) that fall
into a given no-doublets cluster (columns). Rows sum to 100%.

Every cluster pair gets a tile, including pairs sharing no cells, so that an
empty tile is unambiguous. A blank tile means no cells are shared. A tile
labeled `<1` means cells are shared but they are under 0.5% of that
with-doublets cluster, which would otherwise round to a misleading `0`.

```{r, fig.width=8, fig.height=6}
# dplyr::count only returns observed combinations, so fill in the cluster
# pairs with no shared cells before plotting
contingency_plot = contingency %>%
  tidyr::complete(label_A, label_B, fill = list(n_cells = 0, pct_of_A_cluster = 0)) %>%
  mutate(tile_label = case_when(
    n_cells == 0          ~ "",
    pct_of_A_cluster < .5 ~ "<1",
    TRUE                  ~ as.character(round(pct_of_A_cluster))
  ))

ggplot(contingency_plot, aes(x = factor(label_B), y = factor(label_A))) +
  geom_tile(aes(fill = pct_of_A_cluster), color = "light gray") +
  scale_fill_gradientn(colors = c("white", "#fd9668", "#f1605d", "#de4968", "#9e2f7f"),
                       limits = c(0, 100),
                       name = "% of with-doublets\ncluster's cells") +
  geom_text(aes(label = tile_label), size = 2) +
  xlab("No-doublets cluster") +
  ylab("With-doublets cluster") +
  coord_fixed()

ggsave(file = paste0(results_dir, "/cluster_contingency_heatmap.png"), width = 8, height = 6)
```

# Marker gene overlap between cluster marker sets

Independent of cell-level correspondence: does each with-doublets cluster's
top marker genes overlap significantly with a no-doublets cluster's top
marker genes. This still works if the two analyses have different numbers of
clusters, and it is a check on the cell-level correspondence above: the same
cluster pairing should generally show up both by shared cells and by shared
marker genes. Where the two disagree, that pairing is worth a closer look
rather than trusting either line of evidence alone.

```{r}
markers_A = read.csv(paste0(with_dbl_results, "/scatter_cluster_marker_genes.csv"))
markers_B = read.csv(paste0(no_dbl_dir,       "/scatter_cluster_marker_genes.csv"))

top100_A = markers_A %>% group_by(cluster) %>% slice_min(FDR, n = 100, with_ties = TRUE) %>% ungroup()
top100_B = markers_B %>% group_by(cluster) %>% slice_min(FDR, n = 100, with_ties = TRUE) %>% ungroup()

genes_A = top100_A %>% transmute(Gene, Group = paste0("wDbl_", cluster))
genes_B = top100_B %>% transmute(Gene, Group = paste0("noDbl_", cluster))
```

```{r}
shared_genes = intersect(rownames(combined.sce.A), rownames(combined.sce.B))
background_n = length(shared_genes)
```

Background gene universe for the overlap test is the `r background_n` genes
shared by both analyses' combined, uncorrected count matrices (the same
universe `findMarkers` was run over in each analysis).

`phyperOverlapSets` is called directly here rather than through
`plot_marker_overlap_heatmap`. Comparing an analysis to a near-identical
reanalysis produces overlaps so significant that the hypergeometric p-values
underflow to exactly 0, making `-log10(FDR)` infinite and breaking the
`hclust` row and column ordering inside that helper. The output files are the
same ones the helper would have written, so the sections below are unaffected.

```{r}
cluster_ol = phyperOverlapSets(genes_A, genes_B, background_n)

shared_desc = gene_description %>% dplyr::rename(SharedGene = "Gene")

write.csv(cluster_ol$intersecting.genes %>% left_join(shared_desc, by = "SharedGene"),
          file = paste0(results_dir, "/cluster_marker_gene_overlaps.csv"),
          row.names = FALSE)

overlap_scores_all = cluster_ol$pval.scores %>%
  dplyr::rename(CellTypes = "Group1", Clusters = "Group2") %>%
  group_by(CellTypes) %>%
  mutate(FDR = p.adjust(Pval, "BH")) %>%
  ungroup()

write.csv(overlap_scores_all,
          file = paste0(results_dir, "/cluster_marker_overlap_pvalues.csv"),
          row.names = FALSE)

n_underflow = sum(overlap_scores_all$FDR == 0)
```

`r n_underflow` of `r nrow(overlap_scores_all)` cluster pairs have an FDR that
underflowed to 0. For plotting only, these are capped at the largest finite
`-log10(FDR)` observed, so they are shown as the most significant color rather
than dropped. The exported p-value table keeps the raw values.

```{r, fig.width=10, fig.height=8}
plot_scores = overlap_scores_all %>% mutate(logFDR = -log10(FDR))

finite_max = suppressWarnings(max(plot_scores$logFDR[is.finite(plot_scores$logFDR)]))
if (!is.finite(finite_max)) finite_max = 300

plot_scores = plot_scores %>%
  mutate(logFDR = ifelse(is.finite(logFDR), logFDR, finite_max))

FDR_wide = plot_scores %>%
  tidyr::pivot_wider(id_cols = CellTypes, names_from = Clusters, values_from = logFDR)

CellTypes          = FDR_wide$CellTypes
FDR_wide$CellTypes = NULL
FDR_mat            = as.matrix(FDR_wide)
rownames(FDR_mat)  = CellTypes
FDR_mat[!is.finite(FDR_mat)] = 0

# hclust needs at least 3 items and cannot take NA/Inf; fall back to the
# existing order when there are too few clusters to cluster
ord_row = if (nrow(FDR_mat) > 2) hclust(dist(FDR_mat),    method = "ward.D")$order else seq_len(nrow(FDR_mat))
ord_col = if (ncol(FDR_mat) > 2) hclust(dist(t(FDR_mat)), method = "ward.D")$order else seq_len(ncol(FDR_mat))

plot_scores$CellTypes = factor(plot_scores$CellTypes, levels = rownames(FDR_mat)[ord_row])
plot_scores$Clusters  = factor(plot_scores$Clusters,  levels = colnames(FDR_mat)[ord_col])

ggplot(plot_scores, aes(Clusters, CellTypes)) +
  geom_tile(aes(fill = logFDR), color = "light gray") +
  guides(fill = guide_colorbar(title = "-log10(FDR)",
                               barheight = 10, barwidth = 2,
                               default.unit = "mm")) +
  scale_fill_gradientn(colors = c("white", "#fd9668", "#f1605d", "#de4968", "#9e2f7f"),
                       guide = "colorbar") +
  theme(axis.text.x = element_text(angle = 30, hjust = 1, vjust = 1)) +
  theme(text = element_text(size = 7), axis.text = element_text(size = 7)) +
  geom_text(aes(label = nSharedGenes), size = 1) +
  xlab("No-doublets cluster") +
  ylab("With-doublets cluster") +
  coord_fixed()

ggsave(file = paste0(results_dir, "/cluster_marker_overlap_heatmap.png"),
       width = 10, height = 8)
```

## Significant marker gene overlaps (FDR < 0.01)

```{r}
overlap_scores = read.csv(paste0(results_dir, "/cluster_marker_overlap_pvalues.csv")) %>%
  filter(FDR < .01) %>%
  arrange(CellTypes, FDR)

datatable(overlap_scores %>% mutate(FDR = signif(FDR, 3), Pval = signif(Pval, 3)),
          filter = "top", rownames = FALSE)
```

# Combined correspondence table

Brings the two lines of evidence together per with-doublets cluster: the
no-doublets cluster it shares the most cells with, and the no-doublets
cluster its marker genes overlap most significantly with. Agreement between
the two is reassurance that the pairing is real; disagreement means the
cluster likely changed substantially (e.g. split, merged with a different
cell type, or lost its defining markers after doublets were removed) and
should not be assumed to correspond to anything one-to-one.

```{r}
# many FDRs tie at 0 after underflow, so break ties on the number of shared
# genes rather than letting slice_min pick arbitrarily
best_by_markers = overlap_scores %>%
  group_by(CellTypes) %>%
  arrange(FDR, desc(nSharedGenes), .by_group = TRUE) %>%
  slice_head(n = 1) %>%
  ungroup() %>%
  transmute(label_A = as.integer(gsub("wDbl_", "", CellTypes)),
            best_match_by_markers = gsub("noDbl_", "", Clusters),
            marker_overlap_FDR = FDR,
            marker_overlap_nSharedGenes = nSharedGenes)

combined_correspondence = best_match %>%
  dplyr::select(label_A, best_match_by_cells = best_match_label_B, pct_of_A_cluster) %>%
  full_join(best_by_markers, by = "label_A") %>%
  mutate(agree = as.character(best_match_by_cells) == as.character(best_match_by_markers)) %>%
  arrange(label_A)

write.csv(combined_correspondence,
          file = paste0(results_dir, "/cluster_combined_correspondence.csv"),
          row.names = FALSE)

datatable(combined_correspondence %>% mutate(pct_of_A_cluster = round(pct_of_A_cluster, 1),
                                              marker_overlap_FDR = signif(marker_overlap_FDR, 3)),
          rownames = FALSE)
```

`r sum(!combined_correspondence$agree, na.rm = TRUE)` of `r nrow(combined_correspondence)`
with-doublets clusters have a disagreement between the cell-overlap best
match and the marker-overlap best match; those rows are the ones to inspect
manually before filling in the `TODO`s in
`02_sex_cell_subanalysis_no_doublets.Rmd` and
`03_trajectory_analysis_no_doublets.Rmd`. A blank
`best_match_by_markers`/`marker_overlap_FDR` means that with-doublets
cluster had no no-doublets cluster with a significant (FDR < 0.01) marker
overlap at all, which is itself worth noting rather than a disagreement to
resolve.

# Session info

```{r}
sessionInfo()
```
