---
title: "05_compare_doublet_removal"
author: "Stephanie Hickey"
date: "`r Sys.Date()`"
output:
  html_document:
    code_folding: hide
    toc: true
    toc_float:
      toc_collapsed: true
params:
    project_dir: /mnt/ufs18/rs-013/bioinformaticsCore/projects/chong_davidson/BCC125_sea_lamprey_scRNA
    project_name: BCC125_sea_lamprey_scRNA
    lab: Weiming Li
    no_dbl_date: "20260813"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

```{r, echo=FALSE, results="asis"}
cat("\n")
cat("#### Lab:", params$lab, " \n")
cat("#### Project:", params$project_name, " \n")
```

This document compares the with-doublets analysis
(`01_all_cells_clustering_annotation.Rmd`) to the doublet-filtered reanalysis
(`01_all_cells_clustering_annotation_no_doublets.Rmd`), using two independent
lines of evidence: whether individual cells common to both analyses land in
the same or different clusters, and whether each analysis's cluster marker
genes overlap. 

```{r}
library(tidyverse)
library(SingleCellExperiment)
library(DT)
source(paste0(params$project_dir, "/src/scRNA_seq_functions.R"))
```

```{r}
results_dir       = paste0(params$project_dir, "/results/", params$no_dbl_date, "_comparison")
with_dbl_robjects = paste0(params$project_dir, "/results/20260312/Robjects")
with_dbl_results  = paste0(params$project_dir, "/results/20260615")
# which no-doublets run to compare against. Set no_dbl_date to 20260803 to
# reproduce the comparison against the earlier run, which used scDblFinder's
# 10x-calibrated default rate rather than the PIPseq specification.
no_dbl_dir        = paste0(params$project_dir, "/results/", params$no_dbl_date, "_no_doublets")
if (!dir.exists(results_dir)) dir.create(results_dir, recursive = TRUE)
```

```{r}
merged.sce.A   = readRDS(paste0(with_dbl_robjects, "/merged.sce.Rds"))
combined.sce.A = readRDS(paste0(with_dbl_robjects, "/combined.sce.Rds"))

merged.sce.B   = readRDS(paste0(no_dbl_dir, "/merged.sce.Rds"))
combined.sce.B = readRDS(paste0(no_dbl_dir, "/combined.sce.Rds"))

gene_description = read.csv(paste0(with_dbl_results, "/gene_descritption.csv"))
```

# Cell-level cluster correspondence

For cells present in both analyses, this cross-tabulates the with-doublets
cluster label against the no-doublets cluster label. Cells only in the
with-doublets object are candidates for having been removed as doublets (or
otherwise dropped); cells only in the no-doublets object would indicate a
barcode mismatch and are investigated rather than assumed away.

```{r}
common_cells    = intersect(colnames(merged.sce.A), colnames(merged.sce.B))
only_A          = setdiff(colnames(merged.sce.A), colnames(merged.sce.B))
only_B          = setdiff(colnames(merged.sce.B), colnames(merged.sce.A))
```

* `r length(common_cells)` cells are common to both analyses.
* `r length(only_A)` cells are in the with-doublets analysis only (candidates for having been removed as doublets).
* `r length(only_B)` cells are in the no-doublets analysis only.

```{r}
if (length(only_B) > 0) {
  warning(paste0(length(only_B), " cell(s) appear in the no-doublets analysis ",
                 "but not the with-doublets analysis. This should not happen if ",
                 "the no-doublets pipeline only removes cells from the same ",
                 "starting set; investigate before trusting the comparison below."))
}
```

```{r}
label_df = data.frame(
  barcode = common_cells,
  label_A = as.integer(as.character(
    colData(merged.sce.A)$label[match(common_cells, colnames(merged.sce.A))])),
  label_B = as.integer(as.character(
    colData(merged.sce.B)$label[match(common_cells, colnames(merged.sce.B))]))
)

contingency = label_df %>%
  dplyr::count(label_A, label_B, name = "n_cells") %>%
  group_by(label_A) %>%
  mutate(pct_of_A_cluster = 100 * n_cells / sum(n_cells)) %>%
  ungroup()

write.csv(contingency, file = paste0(results_dir, "/cluster_contingency_table.csv"),
          row.names = FALSE)

datatable(contingency %>% mutate(pct_of_A_cluster = round(pct_of_A_cluster, 1)),
          filter = "top", rownames = FALSE)
```

## Best-matching no-doublets cluster per with-doublets cluster

For each with-doublets cluster, the no-doublets cluster that captures the
largest share of its cells. A low `pct_of_A_cluster` means that
with-doublets cluster's cells are split across several no-doublets clusters,
rather than mapping cleanly to one.

```{r}
best_match = contingency %>%
  group_by(label_A) %>%
  slice_max(order_by = n_cells, n = 1, with_ties = FALSE) %>%
  ungroup() %>%
  arrange(label_A) %>%
  dplyr::rename(best_match_label_B = label_B)

write.csv(best_match, file = paste0(results_dir, "/cluster_best_match_by_cells.csv"),
          row.names = FALSE)

datatable(best_match %>% mutate(pct_of_A_cluster = round(pct_of_A_cluster, 1)),
          rownames = FALSE)
```

## Heatmap

Each tile is the percent of a with-doublets cluster's cells (rows) that fall
into a given no-doublets cluster (columns). Rows sum to 100%.

Every cluster pair gets a tile, including pairs sharing no cells, so that an
empty tile is unambiguous. A blank tile means no cells are shared. A tile
labeled `<1` means cells are shared but they are under 0.5% of that
with-doublets cluster, which would otherwise round to a misleading `0`.

```{r, fig.width=8, fig.height=6}
# dplyr::count only returns observed combinations, so fill in the cluster
# pairs with no shared cells before plotting
contingency_plot = contingency %>%
  tidyr::complete(label_A, label_B, fill = list(n_cells = 0, pct_of_A_cluster = 0)) %>%
  mutate(tile_label = case_when(
    n_cells == 0          ~ "",
    pct_of_A_cluster < .5 ~ "<1",
    TRUE                  ~ as.character(round(pct_of_A_cluster))
  ))

ggplot(contingency_plot, aes(x = factor(label_B), y = factor(label_A))) +
  geom_tile(aes(fill = pct_of_A_cluster), color = "light gray") +
  scale_fill_gradientn(colors = c("white", "#fd9668", "#f1605d", "#de4968", "#9e2f7f"),
                       limits = c(0, 100),
                       name = "% of with-doublets\ncluster's cells") +
  geom_text(aes(label = tile_label), size = 2) +
  xlab("No-doublets cluster") +
  ylab("With-doublets cluster") +
  coord_fixed()

ggsave(file = paste0(results_dir, "/cluster_contingency_heatmap.png"), width = 8, height = 6)
```

# Marker gene overlap between cluster marker sets

Independent of cell-level correspondence: does each with-doublets cluster's
top marker genes overlap significantly with a no-doublets cluster's top
marker genes. This still works if the two analyses have different numbers of
clusters, and it is a check on the cell-level correspondence above: the same
cluster pairing should generally show up both by shared cells and by shared
marker genes. Where the two disagree, that pairing is worth a closer look
rather than trusting either line of evidence alone.

```{r}
markers_A = read.csv(paste0(with_dbl_results, "/scatter_cluster_marker_genes.csv"))
markers_B = read.csv(paste0(no_dbl_dir,       "/scatter_cluster_marker_genes.csv"))

top100_A = markers_A %>% group_by(cluster) %>% slice_min(FDR, n = 100, with_ties = TRUE) %>% ungroup()
top100_B = markers_B %>% group_by(cluster) %>% slice_min(FDR, n = 100, with_ties = TRUE) %>% ungroup()

genes_A = top100_A %>% transmute(Gene, Group = paste0("wDbl_", cluster))
genes_B = top100_B %>% transmute(Gene, Group = paste0("noDbl_", cluster))
```

```{r}
shared_genes = intersect(rownames(combined.sce.A), rownames(combined.sce.B))
background_n = length(shared_genes)
```

Background gene universe for the overlap test is the `r background_n` genes
shared by both analyses' combined, uncorrected count matrices (the same
universe `findMarkers` was run over in each analysis).

```{r}
cluster_ol = phyperOverlapSets(genes_A, genes_B, background_n)

shared_desc = gene_description %>% dplyr::rename(SharedGene = "Gene")

write.csv(cluster_ol$intersecting.genes %>% left_join(shared_desc, by = "SharedGene"),
          file = paste0(results_dir, "/cluster_marker_gene_overlaps.csv"),
          row.names = FALSE)

overlap_scores_all = cluster_ol$pval.scores %>%
  dplyr::rename(CellTypes = "Group1", Clusters = "Group2") %>%
  group_by(CellTypes) %>%
  mutate(FDR = p.adjust(Pval, "BH")) %>%
  ungroup()

write.csv(overlap_scores_all,
          file = paste0(results_dir, "/cluster_marker_overlap_pvalues.csv"),
          row.names = FALSE)

n_underflow = sum(overlap_scores_all$FDR == 0)
```

```{r, fig.width=10, fig.height=8}
plot_scores = overlap_scores_all %>% mutate(logFDR = -log10(FDR))

finite_max = suppressWarnings(max(plot_scores$logFDR[is.finite(plot_scores$logFDR)]))
if (!is.finite(finite_max)) finite_max = 300

plot_scores = plot_scores %>%
  mutate(logFDR = ifelse(is.finite(logFDR), logFDR, finite_max))

FDR_wide = plot_scores %>%
  tidyr::pivot_wider(id_cols = CellTypes, names_from = Clusters, values_from = logFDR)

CellTypes          = FDR_wide$CellTypes
FDR_wide$CellTypes = NULL
FDR_mat            = as.matrix(FDR_wide)
rownames(FDR_mat)  = CellTypes
FDR_mat[!is.finite(FDR_mat)] = 0

# hclust needs at least 3 items and cannot take NA/Inf; fall back to the
# existing order when there are too few clusters to cluster
ord_row = if (nrow(FDR_mat) > 2) hclust(dist(FDR_mat),    method = "ward.D")$order else seq_len(nrow(FDR_mat))
ord_col = if (ncol(FDR_mat) > 2) hclust(dist(t(FDR_mat)), method = "ward.D")$order else seq_len(ncol(FDR_mat))

plot_scores$CellTypes = factor(plot_scores$CellTypes, levels = rownames(FDR_mat)[ord_row])
plot_scores$Clusters  = factor(plot_scores$Clusters,  levels = colnames(FDR_mat)[ord_col])

ggplot(plot_scores, aes(Clusters, CellTypes)) +
  geom_tile(aes(fill = logFDR), color = "light gray") +
  guides(fill = guide_colorbar(title = "-log10(FDR)",
                               barheight = 10, barwidth = 2,
                               default.unit = "mm")) +
  scale_fill_gradientn(colors = c("white", "#fd9668", "#f1605d", "#de4968", "#9e2f7f"),
                       guide = "colorbar") +
  theme(axis.text.x = element_text(angle = 30, hjust = 1, vjust = 1)) +
  theme(text = element_text(size = 7), axis.text = element_text(size = 7)) +
  geom_text(aes(label = nSharedGenes), size = 1) +
  xlab("No-doublets cluster") +
  ylab("With-doublets cluster") +
  coord_fixed()

ggsave(file = paste0(results_dir, "/cluster_marker_overlap_heatmap.png"),
       width = 10, height = 8)
```

## Significant marker gene overlaps (FDR < 0.01)

```{r}
overlap_scores = read.csv(paste0(results_dir, "/cluster_marker_overlap_pvalues.csv")) %>%
  filter(FDR < .01) %>%
  arrange(CellTypes, FDR)

datatable(overlap_scores %>% mutate(FDR = signif(FDR, 3), Pval = signif(Pval, 3)),
          filter = "top", rownames = FALSE)
```

# Combined correspondence table

Brings the two lines of evidence together per with-doublets cluster: the
no-doublets cluster it shares the most cells with, and the no-doublets
cluster its marker genes overlap most significantly with. Agreement between
the two is reassurance that the pairing is real; disagreement means the
cluster likely changed substantially (e.g. split, merged with a different
cell type, or lost its defining markers after doublets were removed) and
should not be assumed to correspond to anything one-to-one.

```{r}
# many FDRs tie at 0 after underflow, so break ties on the number of shared
# genes rather than letting slice_min pick arbitrarily
best_by_markers = overlap_scores %>%
  group_by(CellTypes) %>%
  arrange(FDR, desc(nSharedGenes), .by_group = TRUE) %>%
  slice_head(n = 1) %>%
  ungroup() %>%
  transmute(label_A = as.integer(gsub("wDbl_", "", CellTypes)),
            best_match_by_markers = gsub("noDbl_", "", Clusters),
            marker_overlap_FDR = FDR,
            marker_overlap_nSharedGenes = nSharedGenes)

combined_correspondence = best_match %>%
  dplyr::select(label_A, best_match_by_cells = best_match_label_B, pct_of_A_cluster) %>%
  full_join(best_by_markers, by = "label_A") %>%
  mutate(agree = as.character(best_match_by_cells) == as.character(best_match_by_markers)) %>%
  arrange(label_A)

write.csv(combined_correspondence,
          file = paste0(results_dir, "/cluster_combined_correspondence.csv"),
          row.names = FALSE)

datatable(combined_correspondence %>% mutate(pct_of_A_cluster = round(pct_of_A_cluster, 1),
                                              marker_overlap_FDR = signif(marker_overlap_FDR, 3)),
          rownames = FALSE)
```

`r sum(!combined_correspondence$agree, na.rm = TRUE)` of `r nrow(combined_correspondence)`
with-doublets clusters have a disagreement between the cell-overlap best
match and the marker-overlap best match.

# Session info

```{r}
sessionInfo()
```
