---
title: "06_stage_composition"
author: "Stephanie Hickey"
date: "`r Sys.Date()`"
output:
  html_document:
    code_folding: hide
    toc: true
    toc_float:
      toc_collapsed: true
params:
    project_dir: /mnt/ufs18/rs-013/bioinformaticsCore/projects/chong_davidson/BCC125_sea_lamprey_scRNA
    project_name: BCC125_sea_lamprey_scRNA
    lab: Weiming Li
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

```{r, echo=FALSE, results="asis"}
cat("\n")
cat("#### Lab:", params$lab, " \n")
cat("#### Project:", params$project_name, " \n")
```

A reviewer noted that developmental stage is confounded with collection date,
season and technical batch: the two larval libraries were collected in July 2023
and the two transformer libraries in November 2023.

**What does not discriminate.** Finding that early trajectory positions are
larval and late positions transformer does not settle the question, because two
different explanations predict it equally well. One is a residual batch effect.
The other is that the underlying populations genuinely do not co-occur, if
primordial germ cells are largely absent from transformers and differentiated
germ cells largely absent from larvae. Whether that is true of sea lamprey is a
biological question this dataset was not designed to answer, and it is treated
here as an open possibility rather than an established fact. Either way, the
consequence for the analysis is the same: observed segregation along pseudotime
is uninformative, because it is what both explanations predict.

The reverse is not symmetric. Cells of both stages occupying the *same* state is
not predicted by a batch effect, so overlap, where it occurs, does carry
information. Section 4 is that comparison, and it also bears on the biological
question above: if larval and transformer cells are found in the same germ cell
populations, then those states are not stage-exclusive.

Sections 1 to 3 report the pseudotime distribution descriptively, for
transparency about how the libraries sit in the data. No argument is drawn from
segregation in either direction.

**What does discriminate.** Two things, in sections 4 and 5:

  * Which populations are shared between stages. A cell state occupied by both
    larval and transformer cells is not predicted by a batch effect. If the
    shared population is a germ cell state, that is considerably stronger than
    a shared somatic one, and stronger still if the same genes mark that
    population when larval and transformer cells are examined separately.
  * How annotation markers behave along pseudotime, in uncorrected expression.
    If a marker rises or falls monotonically along the inferred ordering, and
    does so within individual libraries rather than only across them, that
    ordering carries biological signal that does not come from the library of
    origin.

Nothing is recomputed here. The clustering, integration and trajectory objects
are loaded as produced by `01_all_cells_clustering_annotation.Rmd`,
`02_sex_cell_subanalysis.Rmd` and `03_trajectory_analysis.Rmd`.

```{r}
library(tidyverse)
library(SingleCellExperiment)
library(scater)
library(scran)
library(patchwork)
library(monocle3)
library(TSCAN)
library(readxl)
library(rtracklayer)
library(DT)
source(paste0(params$project_dir, "/src/scRNA_seq_functions.R"))
```

```{r}
results_dir     = paste0(params$project_dir, "/results/20260804_stage_composition")
Robject_out_dir = paste0(results_dir, "/Robjects")
if (!dir.exists(Robject_out_dir)) dir.create(Robject_out_dir, recursive = TRUE)
data_dir        = paste0(params$project_dir, "/data")
```

```{r, echo=FALSE}
# load pre-computed objects for rendering
Robject_dir    = paste0(params$project_dir, "/results/20260312/Robjects")
all.sex.sce    = readRDS(paste0(Robject_dir, "/no.support.sex.sce.Rds"))
merged.sex.sce = readRDS(paste0(Robject_dir, "/no.support.merged.sex.sce.Rds"))
pseudo         = readRDS(paste0(Robject_dir, "/no.support.merged.sex.pseudo.Rds"))
merged.sce     = readRDS(paste0(Robject_dir, "/merged.sce.Rds"))
combined.sce   = readRDS(paste0(Robject_dir, "/combined.sce.Rds"))
```

```{r}
# broad annotations, as assigned in 02 and 03. These are the MST node names.
anno_germ = c(
  "FGC.migrating",
  "PGC",
  "Oogonia.stem.cells",
  "Oocytes",
  "Cyst.epithelial.cells",
  "Oogonia",
  "SPC.cytokinetic",
  "MGC.SPC.group2",
  "MGC.SPC.group1",
  "MGC.SPC.group3",
  "MGC.SPC.group2",
  "MGC.SPC.group1",
  "MGC.SPC.group3",
  "MGC.SPC.group2",
  "MGC.SPC.group3"
)

narrow_germ = c(
  "Female germ cells (migrating)",
  "Primordial germ cells (migrating, differentiated, cyst)",
  "Oogonia (mitotic, cyst, stem cells)",
  "Oocytes",
  "Cyst epithelial cells",
  "Oogonia (mitotic, cyst, cell fate committed)",
  "Spermatocytes (cytokinetic)",
  "Male germ cells (migrating, cell fate committed, differentiated)",
  "Spermatocytes (meiosis arrested)",
  "Spermatocytes (migrating, S-phase, cyst connected with cell bridges, cell fate committed, GPCR-activated)",
  "Spermatocytes (apoptotic, meiosis arrested)",
  "Male germ cells (migrating, differentiated)",
  "Spermatocytes (angiogenic, cell fate committed, GPCR-activated)",
  "Spermatocytes (cell fate committed)",
  "Male germ cells (migrating, cell fate committed, differentiated, wounded)"
)

colData(merged.sex.sce)$broad  = anno_germ[colData(merged.sex.sce)$label]
colData(merged.sex.sce)$narrow = narrow_germ[colData(merged.sex.sce)$label]

# all-cells annotations, as assigned in 01
anno_all = c(
  "1. Fibroblasts (migrating)",
  "2. Mesonephric cells",
  "3. Granulosa cells (migrating)",
  "4. Granulosa cells",
  "5. Oogonia/Oocytes",
  "6. Endothelial cells",
  "7. Primordial germ cells (migrating)",
  "8. Erythrocytes",
  "9. Pre-granulosa cells (migrating)",
  "10. Leukocytes",
  "11. Hemopoietic stem cells",
  "12. Podocytes",
  "13. Macrophages",
  "14. Pre-granulosa cells (mitotic)",
  "15. Immune cells (migrating, mitotic)",
  "16. Spermatocytes (cytokinetic)",
  "17. Male germ cells (migrating)",
  "18. Spermatocytes (apoptotic)",
  "19. Spermatocytes (GPCR-activated)"
)
```

# Library and stage assignment

Library identifiers are read from the objects rather than assumed. The mapping
below is checked against what is actually present, and the document stops if they
disagree.

```{r}
library_meta = data.frame(
  batch      = c("UND1_C00004336", "UND2_C00004362", "TF_C00004337", "TM_C00004374"),
  library    = c("Undifferentiated.1", "Undifferentiated.2", "Female", "Male"),
  stage      = c("larva", "larva", "transformer", "transformer"),
  collection = c("July 2023", "July 2023", "November 2023", "November 2023"),
  stringsAsFactors = FALSE
)

observed_germ = sort(unique(as.character(colData(merged.sex.sce)$batch)))
observed_all  = sort(unique(as.character(colData(merged.sce)$batch)))

unexpected = setdiff(union(observed_germ, observed_all), library_meta$batch)
missing_germ = setdiff(library_meta$batch, observed_germ)
```

* Libraries in `merged.sex.sce`: `r paste(observed_germ, collapse = ", ")`
* Libraries in `merged.sce`: `r paste(observed_all, collapse = ", ")`
* Identifiers present in the objects but not in the mapping: `r if (length(unexpected) > 0) paste(unexpected, collapse = ", ") else "none"`
* Libraries in the mapping with no cells in the germ cell object: `r if (length(missing_germ) > 0) paste(missing_germ, collapse = ", ") else "none"`

```{r}
if (length(unexpected) > 0) {
  stop(paste0("Library identifier(s) found in the objects that are not in ",
              "library_meta: ", paste(unexpected, collapse = ", "),
              ". Add them before proceeding, rather than silently dropping cells."))
}

datatable(library_meta, rownames = FALSE)
```

Stage is derived from the library, and the library identifier is kept as a
separate variable throughout. This matters because the male and female branches
each come from a single library, so a population that is stage-pure is also
library-pure by construction, and the two facts are not independent evidence.

```{r}
germ_meta = as.data.frame(colData(merged.sex.sce))
germ_meta$barcode = rownames(germ_meta)
germ_meta = germ_meta %>% left_join(library_meta, by = "batch")

common.pseudo = averagePseudotime(pseudo$ordering)
stopifnot(length(common.pseudo) == nrow(germ_meta))
germ_meta$Pseudotime = common.pseudo

pseudo_mat = pathStat(pseudo$ordering)
barcodes_match = identical(rownames(pseudo_mat), colnames(merged.sex.sce))
```

Pseudotime matrix row names `r if (barcodes_match) "match" else "do NOT match"` the
column names of `merged.sex.sce`, in the same order.

```{r}
stopifnot(barcodes_match)
branch_names = colnames(pseudo_mat)
```

Branches returned by `quickPseudotime`: `r paste(branch_names, collapse = ", ")`.

```{r}
all_meta = as.data.frame(colData(merged.sce))
all_meta$barcode = rownames(all_meta)
all_meta = all_meta %>% left_join(library_meta, by = "batch")
all_meta$CellType = anno_all[all_meta$label]
```

# Descriptive: how the libraries are distributed

These three sections report where each library's cells fall along the
trajectory. They are included for transparency: Figure 7A shows this as pie
charts, but the underlying numbers are not written down anywhere.

They are not evidence for or against the reviewer's concern, in either
direction. Suppose the cells do sort by stage, larval early and transformer
late. That is what a real developmental ordering would produce. It is also what
would happen if `fastMNN` did not fully remove the difference between the July
and the November libraries, leaving enough residual structure for the July cells
to sit at one end. Seeing the pattern therefore does not tell you which of the
two you are looking at, so no conclusion is drawn from it here. The comparisons
that can tell them apart are in sections 4 and 5.

## 1. Composition of each MST node

The MST nodes are the broad annotation groups the trajectory was built over.
Figure 7A shows these as pie charts; the underlying numbers are tabulated here.
Every node is reported, sorted by percentage larval, descending.

```{r}
node_lib = germ_meta %>%
  dplyr::count(broad, library, name = "n") %>%
  group_by(broad) %>%
  mutate(pct = 100 * n / sum(n)) %>%
  ungroup()

node_stage = germ_meta %>%
  dplyr::count(broad, stage, name = "n") %>%
  group_by(broad) %>%
  mutate(pct = 100 * n / sum(n)) %>%
  ungroup()

node_lib_wide = node_lib %>%
  tidyr::pivot_wider(id_cols = broad, names_from = library,
                     values_from = c(n, pct), values_fill = 0)

node_stage_wide = node_stage %>%
  tidyr::pivot_wider(id_cols = broad, names_from = stage,
                     values_from = c(n, pct), values_fill = 0)

node_pseudo = germ_meta %>%
  group_by(broad) %>%
  summarize(n_cells = n(),
            mean_pseudotime   = mean(Pseudotime, na.rm = TRUE),
            median_pseudotime = median(Pseudotime, na.rm = TRUE),
            .groups = "drop")

node_dominant = node_lib %>%
  group_by(broad) %>%
  slice_max(pct, n = 1, with_ties = FALSE) %>%
  ungroup() %>%
  transmute(broad,
            dominant_library = library,
            dominant_library_pct = pct,
            single_library_dominated = pct >= 90)

node_comp = node_pseudo %>%
  left_join(node_stage_wide, by = "broad") %>%
  left_join(node_lib_wide,   by = "broad") %>%
  left_join(node_dominant,   by = "broad") %>%
  arrange(desc(pct_larva))

write.csv(node_comp, file = paste0(results_dir, "/SuppTable_node_composition.csv"),
          row.names = FALSE)

datatable(node_comp %>% mutate(across(where(is.numeric), ~round(., 2))),
          filter = "top", rownames = FALSE)
```

```{r}
dominated = node_comp %>% filter(single_library_dominated)
```

`r if (nrow(dominated) == 0) "No node draws 90% or more of its cells from a single library." else paste0("Nodes drawing 90% or more of their cells from a single library: ", paste0(dominated$broad, " (", round(dominated$dominant_library_pct, 1), "% ", dominated$dominant_library, ")", collapse = "; "), ".")`

```{r, fig.width=9, fig.height=5}
node_order = node_comp %>% arrange(mean_pseudotime) %>% pull(broad)

p_node = ggplot(node_stage %>% mutate(broad = factor(broad, levels = node_order)),
                aes(x = broad, y = pct, fill = stage)) +
  geom_bar(stat = "identity") +
  scale_fill_manual(values = c(larva = "#009292", transformer = "#FF6DB6")) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  xlab("MST node, ordered by mean pseudotime") +
  ylab("Percent of cells")

p_node
ggsave(p_node, file = paste0(results_dir, "/node_stage_composition.png"),
       width = 9, height = 5)
```

## 2. Stage composition along pseudotime, per branch

```{r}
branch_long =
  as.data.frame(pseudo_mat) %>%
  tibble::rownames_to_column("barcode") %>%
  tidyr::pivot_longer(-barcode, names_to = "branch", values_to = "pseudotime") %>%
  filter(!is.na(pseudotime)) %>%
  left_join(germ_meta %>% dplyr::select(barcode, library, stage, broad, label),
            by = "barcode")

branch_summary = branch_long %>%
  group_by(branch, stage) %>%
  summarize(n_cells = n(),
            min     = min(pseudotime),
            p10     = quantile(pseudotime, 0.10),
            median  = median(pseudotime),
            p90     = quantile(pseudotime, 0.90),
            max     = max(pseudotime),
            .groups = "drop")

overlap_tab = branch_summary %>%
  dplyr::select(branch, stage, p10, p90) %>%
  tidyr::pivot_wider(id_cols = branch, names_from = stage,
                     values_from = c(p10, p90)) %>%
  mutate(overlap_lo       = pmax(p10_larva, p10_transformer),
         overlap_hi       = pmin(p90_larva, p90_transformer),
         overlap_width    = pmax(0, overlap_hi - overlap_lo),
         union_width      = pmax(p90_larva, p90_transformer) -
                            pmin(p10_larva, p10_transformer),
         overlap_fraction = overlap_width / union_width)

pseudotime_table = branch_summary %>%
  left_join(overlap_tab %>%
              dplyr::select(branch, overlap_width, union_width, overlap_fraction),
            by = "branch")

write.csv(pseudotime_table,
          file = paste0(results_dir, "/SuppTable_pseudotime_by_stage.csv"),
          row.names = FALSE)

datatable(pseudotime_table %>% mutate(across(where(is.numeric), ~round(., 3))),
          rownames = FALSE)
```

```{r, fig.width=10, fig.height=4}
p_density = ggplot(branch_long, aes(x = pseudotime, fill = stage, color = stage)) +
  geom_density(alpha = 0.4) +
  scale_fill_manual(values  = c(larva = "#009292", transformer = "#FF6DB6")) +
  scale_color_manual(values = c(larva = "#009292", transformer = "#FF6DB6")) +
  facet_wrap(~ branch, scales = "free") +
  xlab("Pseudotime") + ylab("Density")

p_density
ggsave(p_density, file = paste0(results_dir, "/pseudotime_density_by_stage.png"),
       width = 10, height = 4)
```

Pseudotime binned into ten equal-count bins per branch.

```{r}
decile_long = branch_long %>%
  group_by(branch) %>%
  mutate(decile = ntile(pseudotime, 10)) %>%
  ungroup() %>%
  dplyr::count(branch, decile, stage, name = "n") %>%
  group_by(branch, decile) %>%
  mutate(pct = 100 * n / sum(n)) %>%
  ungroup()

decile_wide = decile_long %>%
  tidyr::pivot_wider(id_cols = c(branch, decile), names_from = stage,
                     values_from = c(n, pct), values_fill = 0) %>%
  arrange(branch, decile)

write.csv(decile_wide,
          file = paste0(results_dir, "/SuppTable_pseudotime_deciles.csv"),
          row.names = FALSE)

datatable(decile_wide %>% mutate(across(where(is.numeric), ~round(., 2))),
          rownames = FALSE)
```

```{r, fig.width=10, fig.height=4}
p_decile = ggplot(decile_long, aes(x = factor(decile), y = pct, fill = stage)) +
  geom_bar(stat = "identity") +
  scale_fill_manual(values = c(larva = "#009292", transformer = "#FF6DB6")) +
  facet_wrap(~ branch) +
  xlab("Pseudotime decile") + ylab("Percent of cells")

p_decile
ggsave(p_decile, file = paste0(results_dir, "/pseudotime_decile_by_stage.png"),
       width = 10, height = 4)
```

## 3. Subcluster composition

```{r}
germ_sub = germ_meta %>%
  dplyr::count(label, library, stage, name = "n") %>%
  group_by(label) %>%
  mutate(pct = 100 * n / sum(n)) %>%
  ungroup() %>%
  mutate(clustering = "germ_subcluster", label = as.character(label))

all_sub = all_meta %>%
  dplyr::count(label, library, stage, name = "n") %>%
  group_by(label) %>%
  mutate(pct = 100 * n / sum(n)) %>%
  ungroup() %>%
  mutate(clustering = "all_cells_cluster", label = as.character(label))

subcluster_comp = bind_rows(germ_sub, all_sub) %>%
  dplyr::select(clustering, label, library, stage, n, pct)

write.csv(subcluster_comp,
          file = paste0(results_dir, "/SuppTable_subcluster_composition.csv"),
          row.names = FALSE)

datatable(subcluster_comp %>% mutate(across(where(is.numeric), ~round(., 2))),
          filter = "top", rownames = FALSE)
```

```{r, fig.width=9, fig.height=5}
lib_cols = c("Female" = "#FF6DB6", "Male" = "#006ddb",
             "Undifferentiated.1" = "#009292", "Undifferentiated.2" = "#b66ddf")

p_sub = ggplot(germ_sub, aes(x = factor(label, levels = sort(unique(as.numeric(label)))),
                             y = pct, fill = library)) +
  geom_bar(stat = "identity") +
  scale_fill_manual(values = lib_cols) +
  xlab("Germ cell subcluster") + ylab("Percent of cells")

p_sub
ggsave(p_sub, file = paste0(results_dir, "/subcluster_composition_by_library.png"),
       width = 9, height = 5)
```

```{r, fig.width=9, fig.height=5}
p_sub_all = ggplot(all_sub, aes(x = factor(label, levels = sort(unique(as.numeric(label)))),
                                y = pct, fill = library)) +
  geom_bar(stat = "identity") +
  scale_fill_manual(values = lib_cols) +
  xlab("All-cells cluster") + ylab("Percent of cells")

p_sub_all
ggsave(p_sub_all, file = paste0(results_dir, "/all_cluster_composition_by_library.png"),
       width = 9, height = 5)
```

# 4. Which populations are shared between stages

A batch effect moves cells apart; it does not create a population occupied by
both batches. Populations holding a substantial share of cells from both stages
are therefore evidence that the integration is aligning real common biology.

This also gives an empirical read on the assumption discussed at the top of the
document. If germ cell populations turn out to be shared, then those states are
not stage-exclusive, and the segregation seen along pseudotime is not simply a
consequence of the populations existing at only one stage. If instead every germ
cell population is stage-dominated, that is consistent both with stage-exclusive
biology and with a batch effect, and the two remain unseparated by this evidence.

Each cluster is classified as shared (at least 10% of its cells from each stage),
larva-dominated (over 90% larval), transformer-dominated (over 90% transformer),
or intermediate where it meets none of those thresholds.

```{r}
# germ or somatic, from the annotation text
classify_population = function(x) {
  x_l = tolower(x)
  ifelse(grepl("germ|oogonia|oocyte|spermatocyt|spermatogon|pgc|fgc|mgc|spc", x_l),
         "germ", "somatic")
}

stage_share = function(meta, label_col, name_col, clustering) {
  meta %>%
    dplyr::count(.data[[label_col]], .data[[name_col]], stage, name = "n") %>%
    dplyr::rename(label = 1, population = 2) %>%
    group_by(label, population) %>%
    mutate(pct = 100 * n / sum(n)) %>%
    ungroup() %>%
    tidyr::pivot_wider(id_cols = c(label, population), names_from = stage,
                       values_from = c(n, pct), values_fill = 0) %>%
    mutate(clustering      = clustering,
           n_cells         = n_larva + n_transformer,
           population_type = classify_population(population),
           sharing = case_when(
             pct_larva >= 10 & pct_transformer >= 10 ~ "shared",
             pct_larva > 90                          ~ "larva-dominated",
             pct_transformer > 90                    ~ "transformer-dominated",
             TRUE                                    ~ "intermediate"))
}

share_germ = stage_share(germ_meta %>% mutate(label_chr = as.character(label)),
                         "label_chr", "narrow", "germ_subcluster")
share_all  = stage_share(all_meta %>% mutate(label_chr = as.character(label)),
                         "label_chr", "CellType", "all_cells_cluster")

population_sharing = bind_rows(share_germ, share_all) %>%
  dplyr::select(clustering, label, population, population_type, n_cells,
                n_larva, n_transformer, pct_larva, pct_transformer, sharing)
```

```{r}
sharing_counts = population_sharing %>%
  dplyr::count(clustering, population_type, sharing, name = "n_populations")

datatable(sharing_counts, rownames = FALSE)
```

```{r}
shared_pops      = population_sharing %>% filter(sharing == "shared")
shared_germ_pops = shared_pops %>% filter(population_type == "germ")
shared_som_pops  = shared_pops %>% filter(population_type == "somatic")
```

`r nrow(shared_pops)` populations are shared, of which `r nrow(shared_germ_pops)`
are germ cell populations and `r nrow(shared_som_pops)` are somatic.

**Shared germ cell populations:**
`r if (nrow(shared_germ_pops) == 0) "none." else paste0(paste0(shared_germ_pops$population, " (", shared_germ_pops$clustering, " ", shared_germ_pops$label, ", ", round(shared_germ_pops$pct_larva, 1), "% larval, ", shared_germ_pops$n_cells, " cells)"), collapse = "; ")`

**Shared somatic populations:**
`r if (nrow(shared_som_pops) == 0) "none." else paste0(paste0(shared_som_pops$population, " (", round(shared_som_pops$pct_larva, 1), "% larval)"), collapse = "; ")`

## Do shared populations have the same markers in each stage?

Marker genes are computed separately in different groups of cells, and the
agreement between those gene lists is measured. Two metrics are used throughout:

  * **Top 50 overlap**, the percentage of each cluster's fifty strongest markers
    shared between two groups. Easy to read, but it depends on an arbitrary
    cutoff and marker rankings shuffle even between random halves of identical
    data, so a raw percentage means nothing without a reference.
  * **Spearman rank correlation across all tested genes.** It uses all the
    information rather than a cutoff, and is far less sensitive to differences
    in cell number.

Neither is interpretable in isolation, so every comparison is reported against a
reference computed the same way. Two comparisons follow, at different coverage:

  * **Pairwise library discordance**, which requires cells in all four
    libraries. Its reference is the two larval replicate libraries, which differ
    only by animal and library preparation, so it is a true replicate floor.
    This is the stronger design, but sex-restricted populations cannot enter it.
  * **The two-stage comparison**, which requires cells in both stages only, and
    so reaches the germ cell populations excluded above. Its reference is a
    random split of each stage into halves, which measures sampling noise rather
    than replicate variation and therefore sits higher than a true floor.

Cell numbers are matched and every comparison is repeated over independent
draws, in both.

Two implementation choices matter for how these numbers come out.

**The top 50 is sorted explicitly**, by FDR then p-value, rather than relying on
the order `findMarkers` happens to return. Nothing would error if that order
changed; the top 50 would silently become an arbitrary 50 and every overlap
figure in this section would be wrong.

**The rank correlation uses the mean of the pairwise log fold changes, not
`summary.logFC`.** Under `pval.type = "some"`, `summary.logFC` is not a fixed
quantity. `findMarkers` computes a log fold change for every pairwise comparison
of a cluster against each other cluster, then reports the one belonging to the
comparison that supplied the summary p-value, which under `"some"` is drawn from
the middle of the ranked p-values. Which comparison that is varies from gene to
gene, and varies between the two runs being compared: the same gene in the same
cluster may report its log fold change against one cluster in the larval run and
against a different cluster in the transformer run. Correlating those vectors
compares different contrasts. Because both runs are now restricted to the same
`common` cluster set, they carry identical `logFC.*` columns, so the mean across
those columns is an average effect size against the same clusters on both sides
and means the same thing in each.

The noise `summary.logFC` introduced was unstructured, so it attenuated the
correlation rather than inflating it. Any value computed under the previous
version was a lower bound on agreement.

```{r}
top_markers_for = function(marker_list, lab, n = 50) {
  if (is.null(marker_list)) return(character(0))
  if (!lab %in% names(marker_list)) return(character(0))
  df = marker_list[[lab]]
  if (nrow(df) == 0) return(character(0))

  # sort explicitly rather than trusting the order findMarkers returns
  ord = if (all(c("FDR", "p.value") %in% colnames(df))) {
    order(df$FDR, df$p.value)
  } else if ("FDR" %in% colnames(df)) {
    order(df$FDR)
  } else if ("p.value" %in% colnames(df)) {
    order(df$p.value)
  } else {
    seq_len(nrow(df))
  }
  rownames(df)[ord][seq_len(min(n, nrow(df)))]
}

# per-gene effect size: the mean log fold change across every pairwise
# comparison, which is comparable between runs because both use the same
# cluster set. summary.logFC is only a fallback; see the note above.
marker_effect = function(df) {
  lf = grep("^logFC\\.", colnames(df), value = TRUE)
  if (length(lf) > 0) {
    v = rowMeans(as.matrix(as.data.frame(df[, lf, drop = FALSE])), na.rm = TRUE)
    names(v) = rownames(df)
    return(v)
  }
  if ("summary.logFC" %in% colnames(df)) {
    v = as.numeric(df$summary.logFC)
    names(v) = rownames(df)
    return(v)
  }
  NULL
}

top_overlap_pct = function(m1, m2, lab, n = 50) {
  t1 = top_markers_for(m1, lab, n)
  t2 = top_markers_for(m2, lab, n)
  denom = min(length(t1), length(t2))
  if (denom == 0) return(NA_real_)
  100 * length(intersect(t1, t2)) / denom
}

stat_rank_cor = function(m1, m2, lab) {
  if (is.null(m1) || is.null(m2))                 return(NA_real_)
  if (!lab %in% names(m1) || !lab %in% names(m2)) return(NA_real_)
  e1 = marker_effect(m1[[lab]])
  e2 = marker_effect(m2[[lab]])
  if (is.null(e1) || is.null(e2)) return(NA_real_)
  g = intersect(names(e1), names(e2))
  if (length(g) < 10) return(NA_real_)
  suppressWarnings(cor(e1[g], e2[g], method = "spearman"))
}
```

Both comparisons below use uncorrected expression. The germ cell object is
rebuilt here from the per-sample objects, as in `03`. It is only a subset and a
`cbind`, so it is not cached. Rather than patching one sample's `colData` by
name, the columns common to all samples are kept, which is all `cbind` requires
and does not depend on which sample happens to be missing a field.

```{r}
universe     = Reduce(intersect, lapply(all.sex.sce, rownames))
all.sex.sce2 = lapply(all.sex.sce, "[", i = universe, )

common_cols  = Reduce(intersect, lapply(all.sex.sce2,
                                        function(x) colnames(colData(x))))
all.sex.sce2 = lapply(all.sex.sce2, function(x) {
  colData(x) = colData(x)[, common_cols, drop = FALSE]
  x
})

combined.sex.sce = do.call(cbind, all.sex.sce2)
stopifnot(identical(colnames(combined.sex.sce), colnames(merged.sex.sce)))
```

### Pairwise library discordance

This is the primary comparison. Marker genes are computed separately within each
single library, and the agreement between pairs of libraries is measured with the
same top 50 overlap and Spearman rank correlation used elsewhere in this section.
Three pairs are compared:

  * **Undifferentiated.1 versus Undifferentiated.2.** Same stage, same
    collection date, no sex difference. Whatever disagreement appears here is
    produced by replicate animals and library preparation alone, so it is the
    reference floor. Everything else is reported relative to it.
  * **Female versus Male transformer.** Same stage and collection date,
    different sex.
  * **Larvae pooled versus transformers pooled.** The stage contrast, which is
    also the collection date contrast.

Cell numbers are matched across every group and every comparison, since
disagreement between marker lists otherwise tracks statistical power rather than
biology. Each cluster is subsampled to a common `n` in every group, and the whole
procedure is repeated over independent draws. `findMarkers` arguments are
identical to those used above, and library is deliberately not blocked, since
library is the variable being measured.

```{r}
library_discordance = function(sce, labels, library_vec, min_cells = 20,
                               n_draws = 10, n_top = 50, seed = 1) {
  labels      = as.character(labels)
  library_vec = as.character(library_vec)

  singles = c("Undifferentiated.1", "Undifferentiated.2", "Female", "Male")
  group_defs = list(
    "Undifferentiated.1"  = "Undifferentiated.1",
    "Undifferentiated.2"  = "Undifferentiated.2",
    "Female"              = "Female",
    "Male"                = "Male",
    "larvae_pooled"       = c("Undifferentiated.1", "Undifferentiated.2"),
    "transformers_pooled" = c("Female", "Male"))

  comparisons = list(
    UND1_vs_UND2           = c("Undifferentiated.1", "Undifferentiated.2"),
    Female_vs_Male         = c("Female", "Male"),
    larvae_vs_transformers = c("larvae_pooled", "transformers_pooled"))

  clusters_all = sort(unique(labels))
  cnt = sapply(singles, function(lb)
    sapply(clusters_all, function(cl) sum(labels == cl & library_vec == lb)))
  if (is.null(dim(cnt))) cnt = matrix(cnt, nrow = length(clusters_all),
                                      dimnames = list(clusters_all, singles))
  rownames(cnt) = clusters_all

  eligibility = as.data.frame(cnt)
  eligibility$label = clusters_all
  eligibility$min_across_libraries = apply(cnt, 1, min)
  eligibility$eligible = eligibility$min_across_libraries >= min_cells
  rownames(eligibility) = NULL

  eligible = clusters_all[eligibility$eligible]
  if (length(eligible) < 2) {
    warning("Fewer than two clusters have enough cells in every library; ",
            "pairwise library discordance cannot be computed.")
    return(list(results = NULL, eligibility = eligibility,
                n_by_cluster = integer(0), eligible = eligible))
  }

  # Each cluster is matched at its own depth: the smallest count available for
  # it in any single library. What the comparison requires is equal power
  # between the groups being compared, and that holds cluster by cluster. A
  # single global n would instead force every cluster down to the sparsest one,
  # discarding most of the data for no gain, since clusters are never compared
  # against each other here.
  n_by_cluster = apply(cnt[eligible, , drop = FALSE], 1, min)
  names(n_by_cluster) = eligible

  set.seed(seed)
  res = list()

  for (d in seq_len(n_draws)) {
    markers_by_group = list()

    for (gname in names(group_defs)) {
      libs = group_defs[[gname]]
      idx  = integer(0)
      ok   = TRUE
      for (cl in eligible) {
        pool = which(labels == cl & library_vec %in% libs)
        if (length(pool) < n_by_cluster[[cl]]) { ok = FALSE; break }
        idx = c(idx, pool[sample(length(pool), n_by_cluster[[cl]])])
      }
      if (!ok) next
      markers_by_group[[gname]] = findMarkers(
        sce[, idx], droplevels(factor(labels[idx])),
        test.type = "t", direction = "up", pval.type = "some")
    }

    for (cmp in names(comparisons)) {
      g1 = comparisons[[cmp]][1]
      g2 = comparisons[[cmp]][2]
      if (is.null(markers_by_group[[g1]]) || is.null(markers_by_group[[g2]])) next
      res[[length(res) + 1]] = data.frame(
        comparison  = cmp,
        draw        = d,
        label       = eligible,
        n_per_group = as.integer(n_by_cluster[eligible]),
        overlap_pct = vapply(eligible, function(cl)
          top_overlap_pct(markers_by_group[[g1]], markers_by_group[[g2]], cl, n_top),
          numeric(1)),
        rank_cor    = vapply(eligible, function(cl)
          stat_rank_cor(markers_by_group[[g1]], markers_by_group[[g2]], cl),
          numeric(1)),
        stringsAsFactors = FALSE)
    }
  }

  out = do.call(rbind, res)
  if (!is.null(out)) rownames(out) = NULL
  list(results = out, eligibility = eligibility, n_by_cluster = n_by_cluster,
       eligible = eligible)
}
```

```{r}
discordance_file = paste0(Robject_out_dir, "/library_discordance.Rds")

if (file.exists(discordance_file)) {
  discordance = readRDS(discordance_file)
} else {
  message("Computing pairwise library discordance. Cached in ",
          discordance_file, ".")
  discordance = list(
    germ = library_discordance(combined.sex.sce, colData(merged.sex.sce)$label,
                               germ_meta$library),
    all  = library_discordance(combined.sce, colData(merged.sce)$label,
                               all_meta$library))
  saveRDS(discordance, file = discordance_file)
}
```

#### Cells used, and clusters excluded

```{r}
disc_eligibility = bind_rows(
  discordance$germ$eligibility %>% mutate(clustering = "germ_subcluster"),
  discordance$all$eligibility  %>% mutate(clustering = "all_cells_cluster")) %>%
  dplyr::select(clustering, label, everything())

write.csv(disc_eligibility,
          file = paste0(results_dir, "/SuppTable_discordance_eligibility.csv"),
          row.names = FALSE)

datatable(disc_eligibility, filter = "top", rownames = FALSE)
```

Each cluster is subsampled to its own depth, the smallest count available for it
in any single library, and that depth is used identically in every group and
every comparison for that cluster.

```{r}
# a clustering with no cluster passing the four-library threshold yields an
# empty n_by_cluster, which is a result rather than an error
n_used_df = function(d, clustering) {
  if (length(d$n_by_cluster) == 0) {
    return(data.frame(clustering = character(0), label = character(0),
                      n_per_group = integer(0), stringsAsFactors = FALSE))
  }
  data.frame(clustering  = clustering,
             label       = names(d$n_by_cluster),
             n_per_group = as.integer(d$n_by_cluster),
             stringsAsFactors = FALSE)
}

n_used_tab = bind_rows(n_used_df(discordance$germ, "germ_subcluster"),
                       n_used_df(discordance$all,  "all_cells_cluster"))

if (nrow(n_used_tab) > 0) {
  n_used_tab = n_used_tab %>%
    mutate(low_power_flag = n_per_group < 50) %>%
    arrange(n_per_group)
} else {
  n_used_tab$low_power_flag = logical(0)
}

datatable(n_used_tab, rownames = FALSE)
```

```{r}
depth_line = function(d, what) {
  if (length(d$n_by_cluster) == 0) {
    return(paste0(what, ": ", length(d$eligible),
                  " eligible, so no depth could be set and this clustering ",
                  "does not enter the analysis."))
  }
  paste0(what, ": ", length(d$eligible), " eligible, ",
         min(d$n_by_cluster), " to ", max(d$n_by_cluster),
         " cells per cluster per group (median ",
         round(median(d$n_by_cluster)), ").")
}
```

* `r depth_line(discordance$germ, "Germ cell subclusters")`
* `r depth_line(discordance$all, "All-cells clusters")`

`r if (length(discordance$germ$n_by_cluster) == 0) "**The germ cell subclusters cannot support this analysis.** Fewer than two of them have enough cells in all four libraries at once. That is a structural consequence of the biology rather than a shortage of data: the female germ cell subclusters have almost no cells in the male library and the male subclusters almost none in the female library, so requiring presence in all four libraries excludes nearly all of them. The pairwise library comparison therefore runs only on the all-cells clustering, where somatic populations are present in every library, and it speaks to the batch effect in those populations rather than in the germ cell states along the trajectory." else ""`

`r if (nrow(n_used_tab) == 0) "" else if (sum(n_used_tab$low_power_flag) == 0) "Every cluster that does enter the analysis is matched at 50 or more cells per group." else paste0(sum(n_used_tab$low_power_flag), " cluster(s) are matched at fewer than 50 cells per group and are flagged above. Marker rankings are noisy at that depth, so their discordance values carry little information in either direction and should not be read as evidence of agreement or disagreement. Raising min_cells excludes them and does not affect the depth used for any other cluster.")`

```{r}
disc_excluded = disc_eligibility %>% filter(!eligible)
```

`r if (nrow(disc_excluded) == 0) "No cluster was excluded; every cluster has enough cells in all four libraries." else paste0(nrow(disc_excluded), " cluster(s) were excluded for having fewer than the required cells in at least one library. A cluster confined to one sex, such as the oocyte and spermatocyte clusters, cannot appear in this analysis at all, since the Female versus Male comparison requires cells of that cluster in both transformer libraries.")`

#### Discordance per cluster

Both metrics are measures of agreement, so lower values mean greater
discordance. The ratio columns express each comparison against the
Undifferentiated.1 versus Undifferentiated.2 floor for the same cluster: a ratio
below 1 means that comparison agrees less well than replicate libraries of the
same stage do.

```{r}
disc_results = bind_rows(
  if (!is.null(discordance$germ$results))
    discordance$germ$results %>% mutate(clustering = "germ_subcluster"),
  if (!is.null(discordance$all$results))
    discordance$all$results  %>% mutate(clustering = "all_cells_cluster"))

# if neither clustering produced results, give the downstream code a table with
# the right columns and no rows rather than a zero-column object
if (nrow(disc_results) == 0) {
  disc_results = data.frame(
    comparison = character(0), draw = integer(0), label = character(0),
    n_per_group = integer(0), overlap_pct = numeric(0), rank_cor = numeric(0),
    clustering = character(0), stringsAsFactors = FALSE)
}

disc_per_cluster = disc_results %>%
  group_by(clustering, label, comparison) %>%
  summarize(n_per_group    = dplyr::first(n_per_group),
            n_draws        = dplyr::n(),
            overlap_median = median(overlap_pct, na.rm = TRUE),
            overlap_min    = min(overlap_pct, na.rm = TRUE),
            overlap_max    = max(overlap_pct, na.rm = TRUE),
            rankcor_median = median(rank_cor, na.rm = TRUE),
            rankcor_min    = min(rank_cor, na.rm = TRUE),
            rankcor_max    = max(rank_cor, na.rm = TRUE),
            .groups = "drop")

floor_per_cluster = disc_per_cluster %>%
  filter(comparison == "UND1_vs_UND2") %>%
  dplyr::select(clustering, label,
                floor_overlap = overlap_median,
                floor_rankcor = rankcor_median)

disc_per_cluster = disc_per_cluster %>%
  left_join(floor_per_cluster, by = c("clustering", "label")) %>%
  mutate(overlap_ratio_to_floor = ifelse(!is.na(floor_overlap) & floor_overlap > 0,
                                         overlap_median / floor_overlap, NA_real_),
         rankcor_ratio_to_floor = ifelse(!is.na(floor_rankcor) & floor_rankcor > 0,
                                         rankcor_median / floor_rankcor, NA_real_))

write.csv(disc_per_cluster,
          file = paste0(results_dir, "/SuppTable_library_discordance.csv"),
          row.names = FALSE)

datatable(disc_per_cluster %>% mutate(across(where(is.numeric), ~round(., 3))),
          filter = "top", rownames = FALSE)
```

#### Pooled levels per comparison

```{r}
disc_overall = disc_results %>%
  group_by(clustering, comparison) %>%
  summarize(n_obs          = dplyr::n(),
            overlap_median = median(overlap_pct, na.rm = TRUE),
            overlap_min    = min(overlap_pct, na.rm = TRUE),
            overlap_max    = max(overlap_pct, na.rm = TRUE),
            rankcor_median = median(rank_cor, na.rm = TRUE),
            rankcor_min    = min(rank_cor, na.rm = TRUE),
            rankcor_max    = max(rank_cor, na.rm = TRUE),
            .groups = "drop")

floor_overall = disc_overall %>%
  filter(comparison == "UND1_vs_UND2") %>%
  dplyr::select(clustering,
                floor_overlap = overlap_median, floor_rankcor = rankcor_median)

disc_overall = disc_overall %>%
  left_join(floor_overall, by = "clustering") %>%
  mutate(overlap_ratio_to_floor = overlap_median / floor_overlap,
         rankcor_ratio_to_floor = rankcor_median / floor_rankcor)

datatable(disc_overall %>% mutate(across(where(is.numeric), ~round(., 3))),
          rownames = FALSE)
```

#### Paired comparison against the floor

Every cluster has two numbers: how well the two larval replicate libraries agree
for that cluster, which is its floor, and how well larvae and transformers agree
for that same cluster. This test compares those two numbers **within each
cluster**, then asks how the differences come out across clusters.

Pairing matters because clusters are not alike. A cluster with sharp,
distinctive markers will show high agreement in every comparison, and a diffuse
one will show low agreement in every comparison. Collapsing all clusters into a
single number for each comparison lets that between-cluster spread swamp the
difference being measured. Comparing each cluster against itself removes it, in
the same way a paired test removes subject-to-subject variation.

Three things are reported per metric: the median of the per-cluster ratios, how
many clusters fall below their own floor, and a Wilcoxon signed rank test on the
pairs.

```{r}
paired = disc_per_cluster %>%
  dplyr::select(clustering, label, comparison, n_per_group,
                overlap_median, rankcor_median) %>%
  tidyr::pivot_longer(c(overlap_median, rankcor_median),
                      names_to = "metric", values_to = "value") %>%
  tidyr::pivot_wider(id_cols = c(clustering, label, n_per_group, metric),
                     names_from = comparison, values_from = value) %>%
  filter(!is.na(UND1_vs_UND2))

paired_test = paired %>%
  tidyr::pivot_longer(c(Female_vs_Male, larvae_vs_transformers),
                      names_to = "comparison", values_to = "value") %>%
  filter(!is.na(value)) %>%
  group_by(clustering, metric, comparison) %>%
  summarize(
    n_clusters        = dplyr::n(),
    median_floor      = median(UND1_vs_UND2),
    median_comparison = median(value),
    median_ratio      = median(value / UND1_vs_UND2, na.rm = TRUE),
    n_below_floor     = sum(value < UND1_vs_UND2),
    wilcox_p          = if (dplyr::n() >= 5)
      suppressWarnings(wilcox.test(value, UND1_vs_UND2, paired = TRUE)$p.value)
      else NA_real_,
    .groups = "drop") %>%
  mutate(exceeds_floor = median_ratio < 1 &
                         n_below_floor > n_clusters / 2)

write.csv(paired_test,
          file = paste0(results_dir, "/SuppTable_discordance_paired_test.csv"),
          row.names = FALSE)

datatable(paired_test %>% mutate(across(where(is.numeric), ~signif(., 3))),
          rownames = FALSE)
```

`exceeds_floor` is TRUE when a comparison agrees less well than the replicate
floor: the median per-cluster ratio is below 1 and more than half the clusters
individually fall below their own floor. Lower agreement means greater
discordance.

```{r, fig.width=10, fig.height=6}
cmp_cols = c(UND1_vs_UND2 = "#009292", Female_vs_Male = "#006ddb",
             larvae_vs_transformers = "#9e2f7f")

p_disc = ggplot(disc_per_cluster,
                aes(x = factor(label), y = overlap_median, color = comparison)) +
  geom_linerange(aes(ymin = overlap_min, ymax = overlap_max),
                 position = position_dodge(width = 0.6), linewidth = 0.6) +
  geom_point(position = position_dodge(width = 0.6), size = 2) +
  scale_color_manual(values = cmp_cols) +
  facet_wrap(~ clustering, scales = "free_x", ncol = 1) +
  xlab("Cluster") +
  ylab("Top 50 marker overlap (%)") +
  labs(caption = paste0("Points are medians across draws, bars the range. ",
                        "UND1 vs UND2 is the replicate floor."))

p_disc
ggsave(p_disc, file = paste0(results_dir, "/library_discordance_by_cluster.png"),
       width = 10, height = 6)
```

```{r}
disc_read = function(clust) {
  d = disc_overall %>% filter(clustering == clust)
  if (nrow(d) == 0) return(NULL)
  list(
    floor = d %>% filter(comparison == "UND1_vs_UND2"),
    fm    = d %>% filter(comparison == "Female_vs_Male"),
    lt    = d %>% filter(comparison == "larvae_vs_transformers"))
}

dg = disc_read("germ_subcluster")
da = disc_read("all_cells_cluster")

disc_sentence = function(x, what) {
  if (is.null(x) || nrow(x$floor) == 0) return(paste0("No ", what, " result available."))
  parts = c()
  fl = x$floor
  parts = c(parts, paste0(
    "Replicate floor (Undifferentiated.1 versus Undifferentiated.2): median top 50 overlap ",
    round(fl$overlap_median, 1), "% (range ", round(fl$overlap_min, 1), " to ",
    round(fl$overlap_max, 1), "%), median rank correlation ",
    round(fl$rankcor_median, 2), "."))
  if (nrow(x$fm) > 0) {
    parts = c(parts, paste0(
      "Female versus Male, reported for completeness only: median overlap ",
      round(x$fm$overlap_median, 1), "%, rank correlation ",
      round(x$fm$rankcor_median, 2), "."))
  }
  if (nrow(x$lt) > 0) {
    parts = c(parts, paste0(
      "Larvae versus transformers: median overlap ", round(x$lt$overlap_median, 1),
      "%, which is ", round(x$lt$overlap_ratio_to_floor, 2),
      " times the floor, and rank correlation ", round(x$lt$rankcor_median, 2),
      "."))
  }
  paste(parts, collapse = " ")
}

# verdict from the paired per-cluster test, not from a pooled spread
lt_paired = function(clust, met = "overlap_median") {
  paired_test %>%
    filter(clustering == clust, metric == met,
           comparison == "larvae_vs_transformers")
}

lt_g = lt_paired("germ_subcluster")
lt_a = lt_paired("all_cells_cluster")

# the germ subclusters are usually unavailable here, since sex-restricted
# populations cannot be present in all four libraries. Fall back to the
# all-cells clustering, and record which one the verdict came from.
lt_primary    = if (nrow(lt_g) > 0) lt_g else lt_a
lt_source     = if (nrow(lt_g) > 0) "germ cell subclusters" else "all-cells clusters"
lt_available  = nrow(lt_primary) > 0
lt_exceeds    = lt_available && isTRUE(lt_primary$exceeds_floor)
lt_within     = lt_available && !isTRUE(lt_primary$exceeds_floor)

paired_sentence = function(p) {
  if (nrow(p) == 0) return("No paired comparison available.")
  paste0("Paired across ", p$n_clusters, " cluster(s): median floor ",
         round(p$median_floor, 1), "%, median stage contrast ",
         round(p$median_comparison, 1), "%, median per-cluster ratio ",
         round(p$median_ratio, 2), ". ", p$n_below_floor, " of ", p$n_clusters,
         " clusters agree less well than their own floor",
         if (!is.na(p$wilcox_p)) paste0(" (Wilcoxon signed rank p = ",
                                        signif(p$wilcox_p, 2), ")") else "",
         ".")
}
```

#### Interpretation

The two readouts below answer different questions and are computed differently.

The **levels** pool every cluster and every draw into one distribution per
comparison and report its median. They tell you the absolute amount of agreement
in play, which is what decides whether the measure is informative at all. If the
replicate floor itself is low, marker agreement is too noisy here to support any
conclusion, whatever the comparisons do relative to it.

The **paired test** reduces each cluster to its own median first, forms the
ratio against that same cluster's floor, and then summarizes across clusters. It
is the actual comparison against the floor. A ratio of medians and a median of
ratios are different quantities, so these two blocks can disagree: a few
high-agreement clusters can hold up the pooled median while most clusters
individually sit below their own floor.

**Pooled levels, germ cell subclusters.** `r disc_sentence(dg, "germ subcluster")`

**Pooled levels, all-cells clusters.** `r disc_sentence(da, "all-cells cluster")`

**Paired test, germ cell subclusters.** `r paired_sentence(lt_g)`

**Paired test, all-cells clusters.** `r paired_sentence(lt_a)`

#### How to read these numbers

The larvae versus transformers comparison is the one that bears on the
reviewer's concern. Markers here are computed on uncorrected expression, so this
comparison contains the July to November difference in full. If it does not
exceed the replicate floor, then in these clusters the batch difference adds no
more systematic variation than two animals of the same stage do, which is
evidence that the batch effect is modest where it can be measured.

The two outcomes are not symmetric. Exceeding the floor is ambiguous, since real
stage biology and batch effect push in the same direction and this comparison
cannot separate them. Not exceeding it is the informative outcome.

Two limits on how far that generalizes:

  * The testable clusters are those with enough cells in all four libraries,
    which excludes any population confined to one stage or one sex. Most of the
    germ cell states along the trajectory are of that kind. The batch effect is
    therefore being measured in the populations where it can be measured, not in
    the ones the trajectory claim depends on.
  * A cluster containing cells from all four libraries is one that `fastMNN`
    succeeded in merging. Clusters were defined on the corrected values, so
    co-membership by itself is not evidence of a small batch effect: the
    correction exists to produce it. What is evidence is that the *uncorrected*
    marker profiles within those clusters agree across libraries, since the
    correction never touched those values. The residual gap is populations where
    the batch effect was large enough that correction failed and the libraries
    formed separate clusters, which would be excluded here. That gap is
    checkable, and is addressed immediately below.

#### Are there clusters the correction failed to merge?

If a batch effect were large, cells of the same type from different libraries
would fail to merge and would appear as separate, library-dominated clusters.
Those clusters would be excluded from the discordance analysis, hiding the worst
cases. The check is whether any stage-dominated cluster lacks a biological
explanation.

```{r}
stage_dominated = population_sharing %>%
  filter(sharing %in% c("larva-dominated", "transformer-dominated")) %>%
  dplyr::select(clustering, label, population, population_type,
                n_larva, n_transformer, pct_larva, sharing)

datatable(stage_dominated %>% mutate(across(where(is.numeric), ~round(., 1))),
          filter = "top", rownames = FALSE)
```

`r if (nrow(stage_dominated) == 0) "No cluster is dominated by a single stage, so there is no population that the correction failed to merge, and the discordance analysis is not missing a set of high batch effect clusters." else paste0(nrow(stage_dominated), " cluster(s) are dominated by a single stage, of which ", sum(stage_dominated$population_type == "germ"), " are germ cell populations and ", sum(stage_dominated$population_type == "somatic"), " are somatic. Stage-restricted germ cell populations are expected on biological grounds, since oocytes and spermatocytes have no larval counterpart. A stage-dominated **somatic** cluster is the informative case: somatic cell types should be present at both stages, so if one is split by stage that is a candidate for a merge the correction did not achieve. Inspect the somatic rows above; if there are none, the discordance estimate is not missing a set of high batch effect clusters.")`

The Female versus Male comparison is computed and included in the tables and
figure for completeness, but nothing is concluded from it. The clusters testable
here are largely somatic and need not differ between sexes, so it serves neither
as a positive control nor as a result.

`r if (!lt_available) "**The larvae versus transformers comparison could not be computed.** No clustering had two or more clusters with enough cells in all four libraries, so there is no paired test to report and nothing follows in either direction." else if (lt_within) paste0("**The larvae versus transformers comparison does not exceed the replicate floor**, in the ", lt_source, ". ", paired_sentence(lt_primary), " In the clusters that can be tested, the stage and collection date contrast produces no more marker disagreement than two larval replicate libraries do. Since these markers are computed on uncorrected expression, that is direct evidence the batch effect is small in these populations, subject to the two limits above.") else paste0("**The larvae versus transformers comparison exceeds the replicate floor**, in the ", lt_source, ". ", paired_sentence(lt_primary), " The stage and collection date contrast agrees less well than two larval replicate libraries do. This comparison cannot say how much of that shortfall is real stage biology and how much is batch, since both push in the same direction.")`

### Two-stage comparison, covering populations the four-library filter excludes

The comparison above requires enough cells in all four libraries, which excludes
any population confined to one sex. Most germ cell subclusters are of that kind,
so they cannot enter it. This section runs the same comparison at a looser
requirement, cells in both stages rather than in all four libraries, which
reaches those populations.

The reference cannot be the larval replicate floor here, because a cluster with
few larval cells cannot be divided between the two larval libraries. Instead each
stage is split at random into two halves the same size as the comparison groups,
and agreement between those halves is the reference. That measures sampling noise
at matched depth. It is a weaker reference than a replicate floor, since it holds
animal and library constant rather than varying them, so it will sit higher than
the true floor and this comparison is correspondingly conservative.

Every group holds the same number of cells per cluster, `n`, drawn so that each
stage supplies two disjoint sets of `n`. The procedure is repeated over
independent draws, and `findMarkers` arguments are unchanged.

```{r}
stage_discordance = function(sce, labels, stage_vec, min_cells = 20,
                             n_draws = 10, n_top = 50, seed = 1) {
  labels    = as.character(labels)
  stage_vec = as.character(stage_vec)
  stages    = c("larva", "transformer")

  clusters_all = sort(unique(labels))
  cnt = sapply(stages, function(st)
    sapply(clusters_all, function(cl) sum(labels == cl & stage_vec == st)))
  if (is.null(dim(cnt)))
    cnt = matrix(cnt, nrow = length(clusters_all),
                 dimnames = list(clusters_all, stages))
  rownames(cnt) = clusters_all

  # each stage must supply two disjoint draws of n cells per cluster, so n is
  # half the smaller stage count. Every group then holds exactly n.
  n_by_cluster = floor(apply(cnt, 1, min) / 2)

  eligibility = as.data.frame(cnt)
  colnames(eligibility)   = paste0("n_", stages)
  eligibility$label       = clusters_all
  eligibility$n_per_group = as.integer(n_by_cluster)
  eligibility$eligible    = eligibility$n_per_group >= min_cells
  rownames(eligibility)   = NULL

  eligible = clusters_all[eligibility$eligible]
  if (length(eligible) < 2) {
    warning("Fewer than two clusters support the two-stage comparison.")
    return(list(results = NULL, eligibility = eligibility,
                n_by_cluster = integer(0), eligible = eligible))
  }
  n_by_cluster = n_by_cluster[eligible]

  cmps = list(
    larvae_vs_transformers = c("larva_h1", "transformer_h1"),
    larva_splithalf        = c("larva_h1", "larva_h2"),
    transformer_splithalf  = c("transformer_h1", "transformer_h2"))

  set.seed(seed)
  res = list()

  for (d in seq_len(n_draws)) {
    idx = list()
    ok  = TRUE
    for (st in stages) {
      a = integer(0); b = integer(0)
      for (cl in eligible) {
        pool = which(labels == cl & stage_vec == st)
        need = 2 * n_by_cluster[[cl]]
        if (length(pool) < need) { ok = FALSE; break }
        pick = pool[sample(length(pool), need)]
        a = c(a, pick[seq_len(n_by_cluster[[cl]])])
        b = c(b, pick[n_by_cluster[[cl]] + seq_len(n_by_cluster[[cl]])])
      }
      if (!ok) break
      idx[[paste0(st, "_h1")]] = a
      idx[[paste0(st, "_h2")]] = b
    }
    if (!ok) next

    markers = lapply(idx, function(ii)
      findMarkers(sce[, ii], droplevels(factor(labels[ii])),
                  test.type = "t", direction = "up", pval.type = "some"))

    for (cmp in names(cmps)) {
      g1 = cmps[[cmp]][1]
      g2 = cmps[[cmp]][2]
      res[[length(res) + 1]] = data.frame(
        comparison  = cmp,
        draw        = d,
        label       = eligible,
        n_per_group = as.integer(n_by_cluster[eligible]),
        overlap_pct = vapply(eligible, function(cl)
          top_overlap_pct(markers[[g1]], markers[[g2]], cl, n_top), numeric(1)),
        rank_cor    = vapply(eligible, function(cl)
          stat_rank_cor(markers[[g1]], markers[[g2]], cl), numeric(1)),
        stringsAsFactors = FALSE)
    }
  }

  out = do.call(rbind, res)
  if (!is.null(out)) rownames(out) = NULL
  list(results = out, eligibility = eligibility,
       n_by_cluster = n_by_cluster, eligible = eligible)
}
```

```{r}
sd_file = paste0(Robject_out_dir, "/stage_discordance.Rds")

if (file.exists(sd_file)) {
  stage_disc = readRDS(sd_file)
} else {
  message("Computing the two-stage comparison. Cached in ", sd_file, ".")
  stage_disc = list(
    germ = stage_discordance(combined.sex.sce, colData(merged.sex.sce)$label,
                             germ_meta$stage),
    all  = stage_discordance(combined.sce, colData(merged.sce)$label,
                             all_meta$stage))
  saveRDS(stage_disc, file = sd_file)
}
```

#### Cells used, and clusters excluded

```{r}
sd_elig = bind_rows(
  stage_disc$germ$eligibility %>% mutate(clustering = "germ_subcluster"),
  stage_disc$all$eligibility  %>% mutate(clustering = "all_cells_cluster")) %>%
  dplyr::select(clustering, label, everything())

write.csv(sd_elig,
          file = paste0(results_dir, "/SuppTable_stage_discordance_eligibility.csv"),
          row.names = FALSE)

datatable(sd_elig, filter = "top", rownames = FALSE)
```

* `r depth_line(stage_disc$germ, "Germ cell subclusters")`
* `r depth_line(stage_disc$all, "All-cells clusters")`

```{r}
sd_results = bind_rows(
  if (!is.null(stage_disc$germ$results))
    stage_disc$germ$results %>% mutate(clustering = "germ_subcluster"),
  if (!is.null(stage_disc$all$results))
    stage_disc$all$results  %>% mutate(clustering = "all_cells_cluster"))

if (nrow(sd_results) == 0) {
  sd_results = data.frame(
    comparison = character(0), draw = integer(0), label = character(0),
    n_per_group = integer(0), overlap_pct = numeric(0), rank_cor = numeric(0),
    clustering = character(0), stringsAsFactors = FALSE)
}

sd_per_cluster = sd_results %>%
  group_by(clustering, label, comparison) %>%
  summarize(n_per_group    = dplyr::first(n_per_group),
            n_draws        = dplyr::n(),
            overlap_median = median(overlap_pct, na.rm = TRUE),
            overlap_min    = min(overlap_pct, na.rm = TRUE),
            overlap_max    = max(overlap_pct, na.rm = TRUE),
            rankcor_median = median(rank_cor, na.rm = TRUE),
            .groups = "drop")

# the reference is the noisier of the two within-stage split halves, since a
# cluster cannot agree across stages better than it agrees with itself
sd_ref = sd_per_cluster %>%
  filter(comparison %in% c("larva_splithalf", "transformer_splithalf")) %>%
  group_by(clustering, label) %>%
  summarize(ref_overlap = min(overlap_median, na.rm = TRUE),
            ref_rankcor = min(rankcor_median, na.rm = TRUE),
            .groups = "drop")

sd_stage = sd_per_cluster %>%
  filter(comparison == "larvae_vs_transformers") %>%
  left_join(sd_ref, by = c("clustering", "label")) %>%
  mutate(overlap_ratio_to_ref = ifelse(!is.na(ref_overlap) & ref_overlap > 0,
                                       overlap_median / ref_overlap, NA_real_),
         rankcor_ratio_to_ref = ifelse(!is.na(ref_rankcor) & ref_rankcor > 0,
                                       rankcor_median / ref_rankcor, NA_real_),
         below_ref = overlap_median < ref_overlap)

write.csv(sd_stage,
          file = paste0(results_dir, "/SuppTable_stage_discordance.csv"),
          row.names = FALSE)

datatable(sd_stage %>% mutate(across(where(is.numeric), ~round(., 3))),
          filter = "top", rownames = FALSE)
```

#### Paired test against the within-stage reference

```{r}
sd_paired = sd_stage %>%
  filter(!is.na(ref_overlap)) %>%
  group_by(clustering) %>%
  summarize(n_clusters   = dplyr::n(),
            median_ref   = median(ref_overlap),
            median_stage = median(overlap_median),
            median_ratio = median(overlap_ratio_to_ref, na.rm = TRUE),
            n_below_ref  = sum(below_ref, na.rm = TRUE),
            wilcox_p     = if (dplyr::n() >= 5)
              suppressWarnings(wilcox.test(overlap_median, ref_overlap,
                                           paired = TRUE)$p.value) else NA_real_,
            .groups = "drop") %>%
  mutate(exceeds_ref = median_ratio < 1 & n_below_ref > n_clusters / 2)

datatable(sd_paired %>% mutate(across(where(is.numeric), ~signif(., 3))),
          rownames = FALSE)
```

```{r}
sd_sentence = function(clust) {
  p = sd_paired %>% filter(clustering == clust)
  if (nrow(p) == 0) return("No two-stage comparison available.")
  paste0("Paired across ", p$n_clusters, " cluster(s): median within-stage ",
         "reference ", round(p$median_ref, 1), "%, median stage contrast ",
         round(p$median_stage, 1), "%, median per-cluster ratio ",
         round(p$median_ratio, 2), ". ", p$n_below_ref, " of ", p$n_clusters,
         " clusters agree across stages less well than they agree with ",
         "themselves",
         if (!is.na(p$wilcox_p)) paste0(" (Wilcoxon signed rank p = ",
                                        signif(p$wilcox_p, 2), ")") else "",
         ".")
}

sd_germ = sd_paired %>% filter(clustering == "germ_subcluster")
sd_all  = sd_paired %>% filter(clustering == "all_cells_cluster")
sd_germ_available = nrow(sd_germ) > 0
sd_germ_exceeds   = sd_germ_available && isTRUE(sd_germ$exceeds_ref)
```

**Germ cell subclusters.** `r sd_sentence("germ_subcluster")`

**All-cells clusters.** `r sd_sentence("all_cells_cluster")`

`r if (!sd_germ_available) "**No germ cell subcluster supports this comparison.** Fewer than two have enough cells in both stages once the depth requirement is applied, so marker agreement cannot be assessed for germ cell populations at all, by this or any other comparison in this document. That is a limitation to state in the response rather than a result." else if (sd_germ_exceeds) "**Germ cell populations agree across stages less well than they agree with themselves.** More than half of them fall below their own within-stage reference. Since the reference holds animal, library and collection date constant and the stage contrast does not, the shortfall is consistent with either a stage difference or a batch difference, and this comparison cannot separate them." else "**Germ cell populations agree across stages about as well as they agree with themselves.** Their cross-stage marker agreement is not detectably worse than agreement between random halves of the same stage, at matched depth. Since these markers are computed on uncorrected expression, that is evidence the same genes define these populations at both stages."`

```{r, fig.width=10, fig.height=6}
sd_plot = sd_per_cluster %>%
  mutate(comparison = factor(comparison,
                             levels = c("larva_splithalf", "transformer_splithalf",
                                        "larvae_vs_transformers")))

sd_cols = c(larva_splithalf = "#009292", transformer_splithalf = "#FF6DB6",
            larvae_vs_transformers = "#9e2f7f")

p_sd = ggplot(sd_plot, aes(x = factor(label), y = overlap_median,
                           color = comparison)) +
  geom_linerange(aes(ymin = overlap_min, ymax = overlap_max),
                 position = position_dodge(width = 0.6), linewidth = 0.6) +
  geom_point(position = position_dodge(width = 0.6), size = 2) +
  scale_color_manual(values = sd_cols) +
  facet_wrap(~ clustering, scales = "free_x", ncol = 1) +
  xlab("Cluster") + ylab("Top 50 marker overlap (%)") +
  labs(caption = paste0("Points are medians across draws, bars the range. ",
                        "The two split-half comparisons are the within-stage reference."))

p_sd
ggsave(p_sd, file = paste0(results_dir, "/stage_discordance_by_cluster.png"),
       width = 10, height = 6)
```

#### The lowest-agreement population

```{r}
sd_germ_clusters = sd_stage %>%
  filter(clustering == "germ_subcluster", !is.na(overlap_median))

worst = if (nrow(sd_germ_clusters) > 0)
  sd_germ_clusters %>% slice_min(overlap_median, n = 1, with_ties = FALSE) else NULL
```

`r if (is.null(worst)) "No germ cell subcluster entered the two-stage comparison, so there is no lowest-agreement population to report." else paste0("The germ cell subcluster with the lowest cross-stage agreement is **", worst$label, "**, at ", round(worst$overlap_median, 1), "% top 50 overlap with ", worst$n_per_group, " cells per group. Its own within-stage reference is ", round(worst$ref_overlap, 1), "%, a ratio of ", round(worst$overlap_ratio_to_ref, 2), ". ", if (!is.na(worst$overlap_ratio_to_ref) && worst$overlap_ratio_to_ref > 0.8) "Its markers reproduce across stages about as well as they reproduce within one stage, so the low absolute figure reflects the instability of marker lists at this depth rather than a difference between stages." else "Its markers reproduce within a stage but not across stages, which is not explained by depth or instability. The genes defining this population differ between larvae and transformers, and its identity is worth re-examining.")`

#### Adding these results to the population table

```{r}
population_sharing = population_sharing %>%
  left_join(sd_stage %>%
              dplyr::select(clustering, label, n_per_group,
                            stage_overlap_pct = overlap_median,
                            stage_rank_cor    = rankcor_median,
                            ref_overlap_pct   = ref_overlap,
                            ref_rank_cor      = ref_rankcor,
                            overlap_ratio_to_ref),
            by = c("clustering", "label")) %>%
  mutate(min_stage_n   = pmin(n_larva, n_transformer),
         low_cell_flag = min_stage_n < 50)

write.csv(population_sharing,
          file = paste0(results_dir, "/SuppTable_population_sharing.csv"),
          row.names = FALSE)

datatable(population_sharing %>%
            mutate(across(where(is.numeric), ~round(., 2))),
          filter = "top", rownames = FALSE)
```

#### Cell counts for every shared population

Overlap percentages are not interpretable without them. Populations where one
stage contributes fewer than 50 cells are flagged.

```{r}
datatable(population_sharing %>%
            filter(sharing == "shared") %>%
            dplyr::select(clustering, label, population, population_type,
                          n_larva, n_transformer, min_stage_n, low_cell_flag,
                          n_per_group, stage_overlap_pct, ref_overlap_pct,
                          overlap_ratio_to_ref, stage_rank_cor) %>%
            mutate(across(where(is.numeric), ~round(., 2))),
          filter = "top", rownames = FALSE)
```


# 5. Marker behavior along pseudotime

Uncorrected, per-library normalized log-expression is used here, not the
`fastMNN` output, so any trend cannot be an artifact of the correction. Each
marker is plotted against pseudotime per branch and colored by library, so it is
visible whether a trend is carried by one sample or holds within several.

Note that normalization is per library (`logNormCounts` applied to each sample in
`02`), so absolute levels are not strictly comparable across libraries. The
within-library trend, reported separately below, does not depend on that
comparability; the pooled trend does.

```{r}
gene_description =
  read_excel(paste0(data_dir, "/41467_2024_55019_MOESM3_ESM.xlsx"), skip = 1) %>%
  dplyr::rename(WangGeneID = "CountMatrix_GeneID",
                Gene        = "NCBI_GTF_GENEID",
                description = "NCBI_GTF_ProteinInfo") %>%
  dplyr::select(Gene, WangGeneID, description) %>%
  filter(!is.na(Gene))

markers_wide = read.csv(paste0(data_dir, "/Cell_Type_Marker_Genes_12-12-25.csv"))
markers_wide = markers_wide[, 2:ncol(markers_wide)]

markers_long = markers_wide %>%
  pivot_longer(cols = everything(), names_to = "CellType", values_to = "Gene",
               values_drop_na = TRUE) %>%
  mutate(main_part = str_remove(Gene, "\\s*\\(.*\\)")) %>%
  tidyr::separate_rows(main_part, sep = "/") %>%
  mutate(main_part = trimws(main_part)) %>%
  filter(main_part != "")

germ_cell_types = c("Migrating.germ.cells", "Primordial.germ.cells", "Oocyte",
                    "Oogonia", "Spermatogonia", "Spermatocytes",
                    "Spermatogonial.stem.cells")

requested = unique(c("DDX4", "PIWIL1", "SYCP3", "BOLL",
                     markers_long %>%
                       filter(CellType %in% germ_cell_types) %>%
                       pull(main_part)))

resolve_marker = function(symbol, sce_rownames, gd) {
  if (symbol %in% sce_rownames) return(symbol)
  g = gd$Gene[gd$WangGeneID == symbol]
  g = g[g %in% sce_rownames]
  if (length(g) > 0) return(g[1])
  NA_character_
}

marker_map = data.frame(symbol = requested, stringsAsFactors = FALSE) %>%
  rowwise() %>%
  mutate(gene_id = resolve_marker(symbol, rownames(combined.sex.sce),
                                  gene_description)) %>%
  ungroup()

resolved   = marker_map %>% filter(!is.na(gene_id))
unresolved = marker_map %>% filter(is.na(gene_id))
```

* Annotation markers requested: `r nrow(marker_map)`
* Resolved in `rownames(combined.sex.sce)`: `r nrow(resolved)`
* Not found, and therefore excluded: `r if (nrow(unresolved) == 0) "none" else paste(unresolved$symbol, collapse = ", ")`

```{r}
stopifnot(identical(colnames(combined.sex.sce), colnames(merged.sex.sce)))

expr_long =
  as.matrix(logcounts(combined.sex.sce)[resolved$gene_id, , drop = FALSE]) %>%
  t() %>%
  as.data.frame() %>%
  tibble::rownames_to_column("barcode") %>%
  tidyr::pivot_longer(-barcode, names_to = "gene_id", values_to = "logexpr") %>%
  left_join(resolved, by = "gene_id")

expr_branch = branch_long %>%
  dplyr::select(barcode, branch, pseudotime, library, stage) %>%
  inner_join(expr_long, by = "barcode")
```

```{r}
safe_rho = function(x, y) {
  if (length(x) < 10) return(NA_real_)
  if (length(unique(x)) < 2 || length(unique(y)) < 2) return(NA_real_)
  suppressWarnings(cor(x, y, method = "spearman"))
}

safe_slope = function(x, y) {
  if (length(x) < 10 || length(unique(x)) < 2) return(NA_real_)
  unname(coef(lm(y ~ x))[2])
}

trend_overall = expr_branch %>%
  group_by(symbol, gene_id, branch) %>%
  summarize(n_cells        = n(),
            pct_expressing = 100 * mean(logexpr > 0),
            mean_logexpr   = mean(logexpr),
            spearman_rho   = safe_rho(pseudotime, logexpr),
            slope_per_unit_pseudotime = safe_slope(pseudotime, logexpr),
            .groups = "drop") %>%
  mutate(direction = case_when(
    is.na(spearman_rho)     ~ "undetermined",
    spearman_rho >=  0.10   ~ "increases",
    spearman_rho <= -0.10   ~ "decreases",
    TRUE                    ~ "flat"))

trend_by_library = expr_branch %>%
  group_by(symbol, branch, library) %>%
  summarize(n_cells = n(),
            rho     = safe_rho(pseudotime, logexpr),
            .groups = "drop") %>%
  filter(!is.na(rho), n_cells >= 20)

trend_consistency = trend_by_library %>%
  left_join(trend_overall %>% dplyr::select(symbol, branch, spearman_rho),
            by = c("symbol", "branch")) %>%
  filter(!is.na(spearman_rho)) %>%
  group_by(symbol, branch) %>%
  summarize(n_libraries_tested = n(),
            n_libraries_same_sign = sum(sign(rho) == sign(spearman_rho)),
            min_library_rho = min(rho),
            max_library_rho = max(rho),
            .groups = "drop") %>%
  mutate(consistent_within_libraries =
           n_libraries_tested >= 2 & n_libraries_same_sign == n_libraries_tested)

marker_trends = trend_overall %>%
  left_join(trend_consistency, by = c("symbol", "branch")) %>%
  arrange(branch, desc(abs(spearman_rho)))

write.csv(marker_trends,
          file = paste0(results_dir, "/SuppTable_marker_trends.csv"),
          row.names = FALSE)

datatable(marker_trends %>% mutate(across(where(is.numeric), ~round(., 3))),
          filter = "top", rownames = FALSE)
```

## Core markers

Points are individual cells, lines are loess fits per library. One row per
marker, one column per branch.

```{r, fig.width=10, fig.height=8}
core_markers = intersect(c("DDX4", "PIWIL1", "SYCP3", "BOLL"), resolved$symbol)

p_marker = ggplot(expr_branch %>% filter(symbol %in% core_markers),
                  aes(x = pseudotime, y = logexpr, color = library)) +
  geom_point(size = 0.15, alpha = 0.12) +
  geom_smooth(method = "loess", se = FALSE, linewidth = 1) +
  scale_color_manual(values = lib_cols) +
  facet_grid(symbol ~ branch, scales = "free") +
  theme(legend.position = "top") +
  guides(color = guide_legend(override.aes = list(alpha = 1, linewidth = 2))) +
  xlab("Pseudotime") + ylab("Uncorrected log-expression")

p_marker
ggsave(p_marker, file = paste0(results_dir, "/marker_expression_vs_pseudotime.png"),
       width = 10, height = 8)
```

## All annotation markers, as a trend heatmap

Plotting every marker as its own panel is unreadable once there are more than a
handful, so the full set is shown as the statistic itself. Each tile is the
Spearman correlation between expression and pseudotime, computed within a single
library. Red is increasing, blue is decreasing, white is flat, and a blank tile
means that library had fewer than 20 cells on that branch.

A marker whose row is the same color across libraries has a trend that holds
within each sample separately. A marker whose row changes color between
libraries does not, and its pooled trend is likely a difference between samples.

```{r, fig.width=9, fig.height=10}
marker_order = trend_by_library %>%
  group_by(symbol) %>%
  summarize(m = mean(rho, na.rm = TRUE), .groups = "drop") %>%
  arrange(m) %>%
  pull(symbol)

p_rho = ggplot(trend_by_library %>%
                 mutate(symbol = factor(symbol, levels = marker_order)),
               aes(x = library, y = symbol, fill = rho)) +
  geom_tile(color = "grey90") +
  geom_text(aes(label = sprintf("%.2f", rho)), size = 1.9) +
  scale_fill_gradient2(low = "#2166ac", mid = "white", high = "#b2182b",
                       midpoint = 0, limits = c(-1, 1)) +
  facet_wrap(~ branch) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  xlab("Library") + ylab("Marker") +
  labs(fill = "Spearman rho")

p_rho
ggsave(p_rho, file = paste0(results_dir, "/marker_trend_heatmap.png"),
       width = 9, height = 10)
```

## All annotation markers, expression against pseudotime

The same markers as curves, split across pages of six so each panel stays large
enough to read. Points are omitted here; only the per-library loess fits are
drawn.

```{r, fig.width=10, fig.height=7}
# markers with no library passing the cell threshold are absent from
# marker_order, so append them rather than dropping them from the figures
ordered_symbols = c(marker_order, setdiff(resolved$symbol, marker_order))
marker_pages    = split(ordered_symbols,
                        ceiling(seq_along(ordered_symbols) / 6))

for (i in seq_along(marker_pages)) {
  p_page = ggplot(expr_branch %>% filter(symbol %in% marker_pages[[i]]),
                  aes(x = pseudotime, y = logexpr, color = library)) +
    geom_smooth(method = "loess", se = FALSE, linewidth = 0.8) +
    scale_color_manual(values = lib_cols) +
    facet_grid(symbol ~ branch, scales = "free") +
    theme(legend.position = "top") +
    xlab("Pseudotime") + ylab("Uncorrected log-expression")

  print(p_page)
  ggsave(p_page,
         file = paste0(results_dir, "/marker_trends_page", i, ".png"),
         width = 10, height = 7)
}
```

# 6. Summary

```{r}
n_shared       = nrow(shared_pops)
n_shared_germ  = nrow(shared_germ_pops)
n_shared_som   = nrow(shared_som_pops)

n_directional  = sum(marker_trends$direction %in% c("increases", "decreases"))
n_flat         = sum(marker_trends$direction == "flat")
n_tested       = nrow(marker_trends)
n_consistent   = sum(marker_trends$consistent_within_libraries, na.rm = TRUE)

germ_markers_ok = n_shared_germ > 0
markers_ok      = n_directional > 0 && n_consistent > 0
```

**Pairwise library discordance, the primary result.**
`r disc_sentence(dg, "germ subcluster")`

`r if (!lt_available) "The pairwise library comparison could not be computed: no clustering had two or more clusters with enough cells in all four libraries. Nothing follows from it in either direction, and the population sharing and marker trend results below carry the argument on their own." else if (lt_within) paste0("The stage and collection date contrast does not agree less well than two larval replicate libraries do, compared cluster by cluster against each cluster's own floor, in the ", lt_source, ". ", paired_sentence(lt_primary), " Because these markers are computed on uncorrected expression, and the correction never touched those values, that is evidence the batch effect is small in the clusters where it can be measured. ", if (sum(stage_dominated$population_type == "somatic") == 0) "No somatic cluster is stage-dominated, so there is no population that the correction failed to merge and no set of high batch effect clusters excluded from this estimate." else paste0(sum(stage_dominated$population_type == "somatic"), " somatic cluster(s) are stage-dominated and should be inspected, since a somatic type split by stage is a candidate for a merge the correction did not achieve."), " The result does not extend automatically to the germ cell states along the trajectory, most of which are confined to one stage and so cannot enter this comparison.") else paste0("The stage and collection date contrast agrees less well than two larval replicate libraries do, compared cluster by cluster against each cluster's own floor, in the ", lt_source, ". ", paired_sentence(lt_primary), " This comparison cannot apportion that shortfall between real stage biology and batch, since both act in the same direction.")`

**Population sharing.** `r n_shared` of `r nrow(population_sharing)` populations
hold at least 10% of their cells from each stage: `r n_shared_germ` germ cell
populations and `r n_shared_som` somatic populations.

`r if (n_shared_germ == 0) "**No germ cell population is shared between stages.** Every germ cell cluster and subcluster is dominated by one stage. Sharing among somatic populations shows the integration can align cells collected on different dates, but it says nothing about whether the germ cell ordering is developmental, because no germ cell state is jointly occupied. On this evidence the trajectory claim is not supported." else paste0("Germ cell populations occupied by both stages are the strongest evidence available here, because a batch effect does not create a jointly occupied state. ", sd_sentence("germ_subcluster"))`

**Marker behavior.** Of `r n_tested` marker and branch combinations tested,
`r n_directional` show a directional trend along pseudotime (Spearman |rho| >=
0.10), `r n_flat` are flat, and `r n_consistent` have a trend of consistent sign
within every library tested separately.

A trend computed inside a single library holds the confound constant. Every cell
in that calculation came from the same animal, the same collection date and the
same capture, so none of those can be what produces the trend. If a marker rises
with pseudotime among the cells of one library alone, that rise is not the July
to November difference, because there is no July to November difference inside
one library. Trends of that kind bear on the ordering independently of the
confound; a trend seen only when libraries are pooled does not, since it may
just be a difference between libraries.

Note the asymmetry when reading the table. A library may occupy only part of the
pseudotime range, in which case its within-library correlation is computed over a
short stretch and will be attenuated. A strong within-library trend is therefore
good evidence for the ordering, but a weak one is not strong evidence against it.

`r if (!markers_ok) "**On these numbers, marker expression does not support the inferred ordering.** Either no marker changes monotonically along pseudotime, or no trend survives within individual libraries, meaning any apparent trend is carried by differences between samples rather than by position along the trajectory." else "**On these numbers, marker behavior supports the inferred ordering**, at least for the markers and branches where a within-library trend was detected."`

`r if (germ_markers_ok && markers_ok) "Taken together, shared germ cell populations and within-library marker trends provide evidence for the ordering that does not reduce to collection batch." else "Taken together, the evidence needed to separate development from collection batch is incomplete. State this limitation explicitly rather than resting the trajectory claim on the pseudotime composition, which cannot distinguish the two."`

## For the reviewer response

```{r, results="asis"}
cat(paste0(
  "We agree that developmental stage is confounded with collection date and ",
  "batch, and that the distribution of libraries along the trajectory cannot ",
  "by itself address this. Segregation of larval and transformer cells along ",
  "pseudotime is predicted both by a residual batch effect and by the ",
  "possibility that the underlying populations do not co-occur across stages, ",
  "so it does not distinguish them. We therefore report that distribution ",
  "descriptively ",
  "(Supplementary Tables 1 to 3) and base our argument on two analyses that ",
  "do distinguish the possibilities. First, ", n_shared,
  " populations contain at least 10% of their cells from each stage, of which ",
  n_shared_germ, " are germ cell populations",
  if (sd_germ_available) paste0(
    "; for these populations, markers computed separately in larvae and in ",
    "transformers at matched cell numbers overlap by a median of ",
    round(sd_germ$median_stage, 1),
    "% in the top 50, against a within-stage reference of ",
    round(sd_germ$median_ref, 1),
    "% obtained by splitting each stage into random halves of the same size, ",
    "a ratio of ", round(sd_germ$median_ratio, 2), ", with ",
    sd_germ$n_below_ref, " of ", sd_germ$n_clusters,
    " clusters falling below their own reference")
  else "",
  ". Second, of ", n_tested, " marker and branch combinations, ", n_directional,
  " show a directional trend along pseudotime and ", n_consistent,
  " retain a consistent trend within every individual library, computed on ",
  "uncorrected expression. ",
  if (germ_markers_ok && markers_ok) {
    "Because these trends are computed within a single library, collection date and batch are held constant by construction, so the trends cannot be produced by the difference between the July and the November samples."
  } else {
    "We note that this evidence is weaker than we would like, and we have revised the strength of our trajectory claim accordingly."
  }))
```

# Session info

```{r}
sessionInfo()
```
