---
title: "06_stage_composition"
author: "Stephanie Hickey"
date: "`r Sys.Date()`"
output:
  html_document:
    code_folding: hide
    toc: true
    toc_float:
      toc_collapsed: true
params:
    project_dir: /mnt/ufs18/rs-013/bioinformaticsCore/projects/chong_davidson/BCC125_sea_lamprey_scRNA
    project_name: BCC125_sea_lamprey_scRNA
    lab: Weiming Li
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

```{r, echo=FALSE, results="asis"}
cat("\n")
cat("#### Lab:", params$lab, " \n")
cat("#### Project:", params$project_name, " \n")
```

A reviewer noted that developmental stage is confounded with collection date,
season and technical batch: the two larval libraries were collected in July 2023
and the two transformer libraries in November 2023.

**What does not discriminate.** Finding that early trajectory positions are
larval and late positions transformer does not settle the question, because two
different explanations predict it equally well. One is a residual batch effect.
The other is that the underlying populations genuinely do not co-occur, if
primordial germ cells are largely absent from transformers and differentiated
germ cells largely absent from larvae. Whether that is true of sea lamprey is a
biological question this dataset was not designed to answer, and it is treated
here as an open possibility rather than an established fact. Either way, the
consequence for the analysis is the same: observed segregation along pseudotime
is uninformative, because it is what both explanations predict.

The reverse is not symmetric. Cells of both stages occupying the *same* state is
not predicted by a batch effect, so overlap, where it occurs, does carry
information. Section 4 is that comparison, and it also bears on the biological
question above: if larval and transformer cells are found in the same germ cell
populations, then those states are not stage-exclusive.

Sections 1 to 3 report the pseudotime distribution descriptively, for
transparency about how the libraries sit in the data. No argument is drawn from
segregation in either direction.

**What does discriminate.** Two things, in sections 4 and 5:

  * Which populations are shared between stages. A cell state occupied by both
    larval and transformer cells is not predicted by a batch effect. If the
    shared population is a germ cell state, that is considerably stronger than
    a shared somatic one, and stronger still if the same genes mark that
    population when larval and transformer cells are examined separately.
  * How annotation markers behave along pseudotime, in uncorrected expression.
    If a marker rises or falls monotonically along the inferred ordering, and
    does so within individual libraries rather than only across them, that
    ordering carries biological signal that does not come from the library of
    origin.

Nothing is recomputed here. The clustering, integration and trajectory objects
are loaded as produced by `01_all_cells_clustering_annotation.Rmd`,
`02_sex_cell_subanalysis.Rmd` and `03_trajectory_analysis.Rmd`.

```{r}
library(tidyverse)
library(SingleCellExperiment)
library(scater)
library(scran)
library(patchwork)
library(monocle3)
library(TSCAN)
library(readxl)
library(rtracklayer)
library(DT)
source(paste0(params$project_dir, "/src/scRNA_seq_functions.R"))
```

```{r}
results_dir     = paste0(params$project_dir, "/results/20260804_stage_composition")
Robject_out_dir = paste0(results_dir, "/Robjects")
if (!dir.exists(Robject_out_dir)) dir.create(Robject_out_dir, recursive = TRUE)
data_dir        = paste0(params$project_dir, "/data")
```

```{r, echo=FALSE}
# load pre-computed objects for rendering
Robject_dir    = paste0(params$project_dir, "/results/20260312/Robjects")
all.sex.sce    = readRDS(paste0(Robject_dir, "/no.support.sex.sce.Rds"))
merged.sex.sce = readRDS(paste0(Robject_dir, "/no.support.merged.sex.sce.Rds"))
pseudo         = readRDS(paste0(Robject_dir, "/no.support.merged.sex.pseudo.Rds"))
merged.sce     = readRDS(paste0(Robject_dir, "/merged.sce.Rds"))
combined.sce   = readRDS(paste0(Robject_dir, "/combined.sce.Rds"))
```

```{r}
# broad annotations, as assigned in 02 and 03. These are the MST node names.
anno_germ = c(
  "FGC.migrating",
  "PGC",
  "Oogonia.stem.cells",
  "Oocytes",
  "Cyst.epithelial.cells",
  "Oogonia",
  "SPC.cytokinetic",
  "MGC.SPC.group2",
  "MGC.SPC.group1",
  "MGC.SPC.group3",
  "MGC.SPC.group2",
  "MGC.SPC.group1",
  "MGC.SPC.group3",
  "MGC.SPC.group2",
  "MGC.SPC.group3"
)

narrow_germ = c(
  "Female germ cells (migrating)",
  "Primordial germ cells (migrating, differentiated, cyst)",
  "Oogonia (mitotic, cyst, stem cells)",
  "Oocytes",
  "Cyst epithelial cells",
  "Oogonia (mitotic, cyst, cell fate committed)",
  "Spermatocytes (cytokinetic)",
  "Male germ cells (migrating, cell fate committed, differentiated)",
  "Spermatocytes (meiosis arrested)",
  "Spermatocytes (migrating, S-phase, cyst connected with cell bridges, cell fate committed, GPCR-activated)",
  "Spermatocytes (apoptotic, meiosis arrested)",
  "Male germ cells (migrating, differentiated)",
  "Spermatocytes (angiogenic, cell fate committed, GPCR-activated)",
  "Spermatocytes (cell fate committed)",
  "Male germ cells (migrating, cell fate committed, differentiated, wounded)"
)

colData(merged.sex.sce)$broad  = anno_germ[colData(merged.sex.sce)$label]
colData(merged.sex.sce)$narrow = narrow_germ[colData(merged.sex.sce)$label]

# all-cells annotations, as assigned in 01
anno_all = c(
  "1. Fibroblasts (migrating)",
  "2. Mesonephric cells",
  "3. Granulosa cells (migrating)",
  "4. Granulosa cells",
  "5. Oogonia/Oocytes",
  "6. Endothelial cells",
  "7. Primordial germ cells (migrating)",
  "8. Erythrocytes",
  "9. Pre-granulosa cells (migrating)",
  "10. Leukocytes",
  "11. Hemopoietic stem cells",
  "12. Podocytes",
  "13. Macrophages",
  "14. Pre-granulosa cells (mitotic)",
  "15. Immune cells (migrating, mitotic)",
  "16. Spermatocytes (cytokinetic)",
  "17. Male germ cells (migrating)",
  "18. Spermatocytes (apoptotic)",
  "19. Spermatocytes (GPCR-activated)"
)
```

# Library and stage assignment

Library identifiers are read from the objects rather than assumed. The mapping
below is checked against what is actually present, and the document stops if they
disagree.

```{r}
library_meta = data.frame(
  batch      = c("UND1_C00004336", "UND2_C00004362", "TF_C00004337", "TM_C00004374"),
  library    = c("Undifferentiated.1", "Undifferentiated.2", "Female", "Male"),
  stage      = c("larva", "larva", "transformer", "transformer"),
  collection = c("July 2023", "July 2023", "November 2023", "November 2023"),
  stringsAsFactors = FALSE
)

observed_germ = sort(unique(as.character(colData(merged.sex.sce)$batch)))
observed_all  = sort(unique(as.character(colData(merged.sce)$batch)))

unexpected = setdiff(union(observed_germ, observed_all), library_meta$batch)
missing_germ = setdiff(library_meta$batch, observed_germ)
```

* Libraries in `merged.sex.sce`: `r paste(observed_germ, collapse = ", ")`
* Libraries in `merged.sce`: `r paste(observed_all, collapse = ", ")`
* Identifiers present in the objects but not in the mapping: `r if (length(unexpected) > 0) paste(unexpected, collapse = ", ") else "none"`
* Libraries in the mapping with no cells in the germ cell object: `r if (length(missing_germ) > 0) paste(missing_germ, collapse = ", ") else "none"`

```{r}
if (length(unexpected) > 0) {
  stop(paste0("Library identifier(s) found in the objects that are not in ",
              "library_meta: ", paste(unexpected, collapse = ", "),
              ". Add them before proceeding, rather than silently dropping cells."))
}

datatable(library_meta, rownames = FALSE)
```

Stage is derived from the library, and the library identifier is kept as a
separate variable throughout. This matters because the male and female branches
each come from a single library, so a population that is stage-pure is also
library-pure by construction, and the two facts are not independent evidence.

```{r}
germ_meta = as.data.frame(colData(merged.sex.sce))
germ_meta$barcode = rownames(germ_meta)
germ_meta = germ_meta %>% left_join(library_meta, by = "batch")

common.pseudo = averagePseudotime(pseudo$ordering)
stopifnot(length(common.pseudo) == nrow(germ_meta))
germ_meta$Pseudotime = common.pseudo

pseudo_mat = pathStat(pseudo$ordering)
barcodes_match = identical(rownames(pseudo_mat), colnames(merged.sex.sce))
```

Pseudotime matrix row names `r if (barcodes_match) "match" else "do NOT match"` the
column names of `merged.sex.sce`, in the same order.

```{r}
stopifnot(barcodes_match)
branch_names = colnames(pseudo_mat)
```

Branches returned by `quickPseudotime`: `r paste(branch_names, collapse = ", ")`.

```{r}
all_meta = as.data.frame(colData(merged.sce))
all_meta$barcode = rownames(all_meta)
all_meta = all_meta %>% left_join(library_meta, by = "batch")
all_meta$CellType = anno_all[all_meta$label]
```

# Descriptive: how the libraries are distributed

These three sections report where each library's cells fall along the
trajectory. They are included for transparency: Figure 7A shows this as pie
charts, but the underlying numbers are not written down anywhere.

They are not evidence for or against the reviewer's concern, in either
direction. Suppose the cells do sort by stage, larval early and transformer
late. That is what a real developmental ordering would produce. It is also what
would happen if `fastMNN` did not fully remove the difference between the July
and the November libraries, leaving enough residual structure for the July cells
to sit at one end. Seeing the pattern therefore does not tell you which of the
two you are looking at, so no conclusion is drawn from it here. The comparisons
that can tell them apart are in sections 4 and 5.

## 1. Composition of each MST node

The MST nodes are the broad annotation groups the trajectory was built over.
Figure 7A shows these as pie charts; the underlying numbers are tabulated here.
Every node is reported, sorted by percentage larval, descending.

```{r}
node_lib = germ_meta %>%
  dplyr::count(broad, library, name = "n") %>%
  group_by(broad) %>%
  mutate(pct = 100 * n / sum(n)) %>%
  ungroup()

node_stage = germ_meta %>%
  dplyr::count(broad, stage, name = "n") %>%
  group_by(broad) %>%
  mutate(pct = 100 * n / sum(n)) %>%
  ungroup()

node_lib_wide = node_lib %>%
  tidyr::pivot_wider(id_cols = broad, names_from = library,
                     values_from = c(n, pct), values_fill = 0)

node_stage_wide = node_stage %>%
  tidyr::pivot_wider(id_cols = broad, names_from = stage,
                     values_from = c(n, pct), values_fill = 0)

node_pseudo = germ_meta %>%
  group_by(broad) %>%
  summarize(n_cells = n(),
            mean_pseudotime   = mean(Pseudotime, na.rm = TRUE),
            median_pseudotime = median(Pseudotime, na.rm = TRUE),
            .groups = "drop")

node_dominant = node_lib %>%
  group_by(broad) %>%
  slice_max(pct, n = 1, with_ties = FALSE) %>%
  ungroup() %>%
  transmute(broad,
            dominant_library = library,
            dominant_library_pct = pct,
            single_library_dominated = pct >= 90)

node_comp = node_pseudo %>%
  left_join(node_stage_wide, by = "broad") %>%
  left_join(node_lib_wide,   by = "broad") %>%
  left_join(node_dominant,   by = "broad") %>%
  arrange(desc(pct_larva))

write.csv(node_comp, file = paste0(results_dir, "/SuppTable_node_composition.csv"),
          row.names = FALSE)

datatable(node_comp %>% mutate(across(where(is.numeric), ~round(., 2))),
          filter = "top", rownames = FALSE)
```

```{r}
dominated = node_comp %>% filter(single_library_dominated)
```

`r if (nrow(dominated) == 0) "No node draws 90% or more of its cells from a single library." else paste0("Nodes drawing 90% or more of their cells from a single library: ", paste0(dominated$broad, " (", round(dominated$dominant_library_pct, 1), "% ", dominated$dominant_library, ")", collapse = "; "), ".")`

```{r, fig.width=9, fig.height=5}
node_order = node_comp %>% arrange(mean_pseudotime) %>% pull(broad)

p_node = ggplot(node_stage %>% mutate(broad = factor(broad, levels = node_order)),
                aes(x = broad, y = pct, fill = stage)) +
  geom_bar(stat = "identity") +
  scale_fill_manual(values = c(larva = "#009292", transformer = "#FF6DB6")) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  xlab("MST node, ordered by mean pseudotime") +
  ylab("Percent of cells")

p_node
ggsave(p_node, file = paste0(results_dir, "/node_stage_composition.png"),
       width = 9, height = 5)
```

## 2. Stage composition along pseudotime, per branch

```{r}
branch_long =
  as.data.frame(pseudo_mat) %>%
  tibble::rownames_to_column("barcode") %>%
  tidyr::pivot_longer(-barcode, names_to = "branch", values_to = "pseudotime") %>%
  filter(!is.na(pseudotime)) %>%
  left_join(germ_meta %>% dplyr::select(barcode, library, stage, broad, label),
            by = "barcode")

branch_summary = branch_long %>%
  group_by(branch, stage) %>%
  summarize(n_cells = n(),
            min     = min(pseudotime),
            p10     = quantile(pseudotime, 0.10),
            median  = median(pseudotime),
            p90     = quantile(pseudotime, 0.90),
            max     = max(pseudotime),
            .groups = "drop")

overlap_tab = branch_summary %>%
  dplyr::select(branch, stage, p10, p90) %>%
  tidyr::pivot_wider(id_cols = branch, names_from = stage,
                     values_from = c(p10, p90)) %>%
  mutate(overlap_lo       = pmax(p10_larva, p10_transformer),
         overlap_hi       = pmin(p90_larva, p90_transformer),
         overlap_width    = pmax(0, overlap_hi - overlap_lo),
         union_width      = pmax(p90_larva, p90_transformer) -
                            pmin(p10_larva, p10_transformer),
         overlap_fraction = overlap_width / union_width)

pseudotime_table = branch_summary %>%
  left_join(overlap_tab %>%
              dplyr::select(branch, overlap_width, union_width, overlap_fraction),
            by = "branch")

write.csv(pseudotime_table,
          file = paste0(results_dir, "/SuppTable_pseudotime_by_stage.csv"),
          row.names = FALSE)

datatable(pseudotime_table %>% mutate(across(where(is.numeric), ~round(., 3))),
          rownames = FALSE)
```

```{r, fig.width=10, fig.height=4}
p_density = ggplot(branch_long, aes(x = pseudotime, fill = stage, color = stage)) +
  geom_density(alpha = 0.4) +
  scale_fill_manual(values  = c(larva = "#009292", transformer = "#FF6DB6")) +
  scale_color_manual(values = c(larva = "#009292", transformer = "#FF6DB6")) +
  facet_wrap(~ branch, scales = "free") +
  xlab("Pseudotime") + ylab("Density")

p_density
ggsave(p_density, file = paste0(results_dir, "/pseudotime_density_by_stage.png"),
       width = 10, height = 4)
```

Pseudotime binned into ten equal-count bins per branch.

```{r}
decile_long = branch_long %>%
  group_by(branch) %>%
  mutate(decile = ntile(pseudotime, 10)) %>%
  ungroup() %>%
  dplyr::count(branch, decile, stage, name = "n") %>%
  group_by(branch, decile) %>%
  mutate(pct = 100 * n / sum(n)) %>%
  ungroup()

decile_wide = decile_long %>%
  tidyr::pivot_wider(id_cols = c(branch, decile), names_from = stage,
                     values_from = c(n, pct), values_fill = 0) %>%
  arrange(branch, decile)

write.csv(decile_wide,
          file = paste0(results_dir, "/SuppTable_pseudotime_deciles.csv"),
          row.names = FALSE)

datatable(decile_wide %>% mutate(across(where(is.numeric), ~round(., 2))),
          rownames = FALSE)
```

```{r, fig.width=10, fig.height=4}
p_decile = ggplot(decile_long, aes(x = factor(decile), y = pct, fill = stage)) +
  geom_bar(stat = "identity") +
  scale_fill_manual(values = c(larva = "#009292", transformer = "#FF6DB6")) +
  facet_wrap(~ branch) +
  xlab("Pseudotime decile") + ylab("Percent of cells")

p_decile
ggsave(p_decile, file = paste0(results_dir, "/pseudotime_decile_by_stage.png"),
       width = 10, height = 4)
```

## 3. Subcluster composition

```{r}
germ_sub = germ_meta %>%
  dplyr::count(label, library, stage, name = "n") %>%
  group_by(label) %>%
  mutate(pct = 100 * n / sum(n)) %>%
  ungroup() %>%
  mutate(clustering = "germ_subcluster", label = as.character(label))

all_sub = all_meta %>%
  dplyr::count(label, library, stage, name = "n") %>%
  group_by(label) %>%
  mutate(pct = 100 * n / sum(n)) %>%
  ungroup() %>%
  mutate(clustering = "all_cells_cluster", label = as.character(label))

subcluster_comp = bind_rows(germ_sub, all_sub) %>%
  dplyr::select(clustering, label, library, stage, n, pct)

write.csv(subcluster_comp,
          file = paste0(results_dir, "/SuppTable_subcluster_composition.csv"),
          row.names = FALSE)

datatable(subcluster_comp %>% mutate(across(where(is.numeric), ~round(., 2))),
          filter = "top", rownames = FALSE)
```

```{r, fig.width=9, fig.height=5}
lib_cols = c("Female" = "#FF6DB6", "Male" = "#006ddb",
             "Undifferentiated.1" = "#009292", "Undifferentiated.2" = "#b66ddf")

p_sub = ggplot(germ_sub, aes(x = factor(label, levels = sort(unique(as.numeric(label)))),
                             y = pct, fill = library)) +
  geom_bar(stat = "identity") +
  scale_fill_manual(values = lib_cols) +
  xlab("Germ cell subcluster") + ylab("Percent of cells")

p_sub
ggsave(p_sub, file = paste0(results_dir, "/subcluster_composition_by_library.png"),
       width = 9, height = 5)
```

```{r, fig.width=9, fig.height=5}
p_sub_all = ggplot(all_sub, aes(x = factor(label, levels = sort(unique(as.numeric(label)))),
                                y = pct, fill = library)) +
  geom_bar(stat = "identity") +
  scale_fill_manual(values = lib_cols) +
  xlab("All-cells cluster") + ylab("Percent of cells")

p_sub_all
ggsave(p_sub_all, file = paste0(results_dir, "/all_cluster_composition_by_library.png"),
       width = 9, height = 5)
```

# 4. Which populations are shared between stages

A batch effect moves cells apart; it does not create a population occupied by
both batches. Populations holding a substantial share of cells from both stages
are therefore evidence that the integration is aligning real common biology.

This also gives an empirical read on the assumption discussed at the top of the
document. If germ cell populations turn out to be shared, then those states are
not stage-exclusive, and the segregation seen along pseudotime is not simply a
consequence of the populations existing at only one stage. If instead every germ
cell population is stage-dominated, that is consistent both with stage-exclusive
biology and with a batch effect, and the two remain unseparated by this evidence.

Each cluster is classified as shared (at least 10% of its cells from each stage),
larva-dominated (over 90% larval), transformer-dominated (over 90% transformer),
or intermediate where it meets none of those thresholds.

```{r}
# germ or somatic, from the annotation text
classify_population = function(x) {
  x_l = tolower(x)
  ifelse(grepl("germ|oogonia|oocyte|spermatocyt|spermatogon|pgc|fgc|mgc|spc", x_l),
         "germ", "somatic")
}

stage_share = function(meta, label_col, name_col, clustering) {
  meta %>%
    dplyr::count(.data[[label_col]], .data[[name_col]], stage, name = "n") %>%
    dplyr::rename(label = 1, population = 2) %>%
    group_by(label, population) %>%
    mutate(pct = 100 * n / sum(n)) %>%
    ungroup() %>%
    tidyr::pivot_wider(id_cols = c(label, population), names_from = stage,
                       values_from = c(n, pct), values_fill = 0) %>%
    mutate(clustering      = clustering,
           n_cells         = n_larva + n_transformer,
           population_type = classify_population(population),
           sharing = case_when(
             pct_larva >= 10 & pct_transformer >= 10 ~ "shared",
             pct_larva > 90                          ~ "larva-dominated",
             pct_transformer > 90                    ~ "transformer-dominated",
             TRUE                                    ~ "intermediate"))
}

share_germ = stage_share(germ_meta %>% mutate(label_chr = as.character(label)),
                         "label_chr", "narrow", "germ_subcluster")
share_all  = stage_share(all_meta %>% mutate(label_chr = as.character(label)),
                         "label_chr", "CellType", "all_cells_cluster")

population_sharing = bind_rows(share_germ, share_all) %>%
  dplyr::select(clustering, label, population, population_type, n_cells,
                n_larva, n_transformer, pct_larva, pct_transformer, sharing)
```

```{r}
sharing_counts = population_sharing %>%
  dplyr::count(clustering, population_type, sharing, name = "n_populations")

datatable(sharing_counts, rownames = FALSE)
```

```{r}
shared_pops      = population_sharing %>% filter(sharing == "shared")
shared_germ_pops = shared_pops %>% filter(population_type == "germ")
shared_som_pops  = shared_pops %>% filter(population_type == "somatic")
```

`r nrow(shared_pops)` populations are shared, of which `r nrow(shared_germ_pops)`
are germ cell populations and `r nrow(shared_som_pops)` are somatic.

**Shared germ cell populations:**
`r if (nrow(shared_germ_pops) == 0) "none." else paste0(paste0(shared_germ_pops$population, " (", shared_germ_pops$clustering, " ", shared_germ_pops$label, ", ", round(shared_germ_pops$pct_larva, 1), "% larval, ", shared_germ_pops$n_cells, " cells)"), collapse = "; ")`

**Shared somatic populations:**
`r if (nrow(shared_som_pops) == 0) "none." else paste0(paste0(shared_som_pops$population, " (", round(shared_som_pops$pct_larva, 1), "% larval)"), collapse = "; ")`

## Do shared populations have the same markers in each stage?

For each shared population, marker genes are computed twice: once using only
larval cells and once using only transformer cells, in each case comparing that
population against all other cells of the same stage in the same clustering.
Uncorrected expression is used. If the two gene lists agree, the population is
being aligned on common biology rather than merged by the correction.

<<<<<<< HEAD
**The two calls must use the same set of competing clusters.** `findMarkers`
scores a cluster against the other clusters present in the object it is given.
If a cluster is dropped from one stage for having too few cells but retained in
the other, the two stages are no longer being compared against the same
background, and a gene can appear as a marker in one stage only because a
competing cluster was missing there. That would inflate apparent disagreement
between stages, which is precisely the quantity this section reports, so the
eligible cluster set is computed as the intersection across stages and both
calls are restricted to it.

The asymmetry is expected to be real here: the meiotic male clusters and the
differentiated oocyte cluster are likely sparse in larvae, so without this the
larval background would be systematically smaller.

Clusters excluded by the intersection are not discarded from the report. A
cluster with cells in one stage and almost none in the other is a stage
restricted population, which is a finding, so the exclusions are tabulated
below.

Markers are not blocked on library within stage. Each stage contains two
libraries, and in the transformer stage cluster identity is heavily confounded
with library, since the male and female clusters come from different animals.
Blocking would therefore drop the comparisons of interest. Both stages are
computed the same way, so the overlap comparison remains internally consistent.

The uncorrected germ cell object is rebuilt here rather than cached, as in `03`,
since it is only a subset and a `cbind`. Rather than patching one sample's
`colData` by name, the columns common to all samples are kept, which is all
`cbind` requires and does not depend on which sample happens to be missing a
field.

```{r}
universe     = Reduce(intersect, lapply(all.sex.sce, rownames))
all.sex.sce2 = lapply(all.sex.sce, "[", i = universe, )

common_cols  = Reduce(intersect, lapply(all.sex.sce2,
                                        function(x) colnames(colData(x))))
all.sex.sce2 = lapply(all.sex.sce2, function(x) {
  colData(x) = colData(x)[, common_cols, drop = FALSE]
  x
})

combined.sex.sce = do.call(cbind, all.sex.sce2)
stopifnot(identical(colnames(combined.sex.sce), colnames(merged.sex.sce)))
```

```{r}
stage_split_markers = function(sce, labels, stage_vec, min_cells = 20) {
  labels    = as.character(labels)
  stage_vec = as.character(stage_vec)
  stages    = sort(unique(stage_vec))

  # cells per cluster per stage
  counts  = table(labels, stage_vec)
  elig_df = as.data.frame.matrix(counts)
  colnames(elig_df) = paste0("n_", colnames(elig_df))
  elig_df$label = rownames(elig_df)

  # eligible clusters within each stage, then the intersection actually used
  eligible = lapply(stages, function(st) rownames(counts)[counts[, st] >= min_cells])
  names(eligible) = stages
  common = Reduce(intersect, eligible)

  for (st in stages) {
    elig_df[[paste0("eligible_", st)]] = counts[elig_df$label, st] >= min_cells
  }
  elig_mat = as.matrix(elig_df[, paste0("eligible_", stages), drop = FALSE])
  elig_df$failed_in = apply(elig_mat, 1, function(r) paste(stages[!r], collapse = "; "))
  elig_df$used_in_comparison = elig_df$label %in% common
  rownames(elig_df) = NULL

  markers = list()
  if (length(common) < 2) {
    warning(paste0("Fewer than two clusters are eligible in every stage, so ",
                   "stage-split markers cannot be computed."))
  } else {
    for (st in stages) {
      keep = stage_vec == st & labels %in% common
      markers[[st]] = findMarkers(sce[, keep],
                                  droplevels(factor(labels[keep])),
                                  test.type = "t",
                                  direction = "up",
                                  pval.type = "some")
    }
  }

  list(markers           = markers,
       eligibility       = elig_df,
       eligible_by_stage = eligible,
       common            = common,
       min_cells         = min_cells)
}
```

The `findMarkers` calls are the expensive part, so they are computed once and
cached. On the first run the chunk below computes and saves them; on later runs
it loads the saved result. Delete `stage_split_markers.Rds` to force a
recomputation.

```{r}
markers_file = paste0(Robject_out_dir, "/stage_split_markers.Rds")

if (file.exists(markers_file)) {
  stage_split = readRDS(markers_file)
} else {
  message("Computing stage-split markers. This is slow, and the result is ",
          "cached in ", markers_file, " for later runs.")
  stage_split = list(
    germ = stage_split_markers(combined.sex.sce,
                               colData(merged.sex.sce)$label,
                               germ_meta$stage),
    all  = stage_split_markers(combined.sce,
                               colData(merged.sce)$label,
                               all_meta$stage))
  saveRDS(stage_split, file = markers_file)
}
```

### Which clusters were eligible in each stage

```{r}
eligibility = bind_rows(
  stage_split$germ$eligibility %>% mutate(clustering = "germ_subcluster"),
  stage_split$all$eligibility  %>% mutate(clustering = "all_cells_cluster")) %>%
  dplyr::select(clustering, label, everything())

write.csv(eligibility,
          file = paste0(results_dir, "/SuppTable_marker_background_eligibility.csv"),
          row.names = FALSE)

datatable(eligibility, filter = "top", rownames = FALSE)
```

```{r, results="asis"}
for (cl in c("germ", "all")) {
  s   = stage_split[[cl]]
  nm  = if (cl == "germ") "Germ cell subclusters" else "All-cells clusters"
  cat("**", nm, "** (minimum ", s$min_cells, " cells per stage)\n\n", sep = "")
  for (st in names(s$eligible_by_stage)) {
    cat("* Eligible in ", st, ": ",
        paste(sort(as.numeric(s$eligible_by_stage[[st]])), collapse = ", "),
        "\n", sep = "")
  }
  cat("* Intersection used for both calls: ",
      paste(sort(as.numeric(s$common)), collapse = ", "), "\n\n", sep = "")
}
```

### Clusters excluded from the marker comparison

A cluster excluded here has too few cells in at least one stage. That is a
stage restricted population, and is reported rather than dropped silently.

```{r}
excluded = eligibility %>%
  filter(!used_in_comparison) %>%
  dplyr::select(clustering, label, starts_with("n_"), failed_in)

datatable(excluded, rownames = FALSE)
```

`r if (nrow(excluded) == 0) "No cluster was excluded; every cluster met the cell threshold in both stages." else paste0(nrow(excluded), " cluster(s) were excluded from the marker comparison because they fall below the cell threshold in at least one stage.")`

### Two metrics, and why a control is needed

Top 50 overlap is easy to read but unstable: marker rankings shuffle between
random halves of identical data, so a raw overlap percentage cannot be judged
without knowing how much overlap is achievable at all. Two things are added
here.

  * **A split-half control.** Within a single stage, the cells of the shared
    clusters are split at random into two halves, markers are computed on each
    half, and the top 50 overlap is calculated by the identical function. This
    establishes the ceiling. Cross-stage overlap should be read against it, not
    on its own.
  * **A rank correlation across all genes.** Spearman correlation between the
    larval and transformer test statistics over every tested gene. It uses all
    the information rather than an arbitrary cutoff, and is far less sensitive
    to differences in cell number.

Both metrics are computed identically for the cross-stage comparison and for
the split-half control.

```{r}
top_markers_for = function(marker_list, lab, n = 50) {
  if (is.null(marker_list)) return(character(0))
  if (!lab %in% names(marker_list)) return(character(0))
  rownames(marker_list[[lab]])[seq_len(min(n, nrow(marker_list[[lab]])))]
}

# the statistic findMarkers returns for the relevant comparison
pick_stat = function(df) {
  if ("summary.logFC" %in% colnames(df)) return("summary.logFC")
  lf = grep("^logFC", colnames(df), value = TRUE)
  if (length(lf) > 0) return(lf[1])
  NA_character_
}

top_overlap_pct = function(m1, m2, lab, n = 50) {
  t1 = top_markers_for(m1, lab, n)
  t2 = top_markers_for(m2, lab, n)
  denom = min(length(t1), length(t2))
  if (denom == 0) return(NA_real_)
  100 * length(intersect(t1, t2)) / denom
}

stat_rank_cor = function(m1, m2, lab) {
  if (is.null(m1) || is.null(m2))            return(NA_real_)
  if (!lab %in% names(m1) || !lab %in% names(m2)) return(NA_real_)
  d1 = m1[[lab]]; d2 = m2[[lab]]
  s1 = pick_stat(d1); s2 = pick_stat(d2)
  if (is.na(s1) || is.na(s2)) return(NA_real_)
  g = intersect(rownames(d1), rownames(d2))
  if (length(g) < 10) return(NA_real_)
  suppressWarnings(cor(as.numeric(d1[g, s1]), as.numeric(d2[g, s2]),
                       method = "spearman"))
}
```

### Pairwise library discordance

This is the primary comparison. Marker genes are computed separately within each
single library, and the agreement between pairs of libraries is measured with the
same top 50 overlap and Spearman rank correlation used elsewhere in this section.
Three pairs are compared:

  * **Undifferentiated.1 versus Undifferentiated.2.** Same stage, same
    collection date, no sex difference. Whatever disagreement appears here is
    produced by replicate animals and library preparation alone, so it is the
    reference floor. Everything else is reported relative to it.
  * **Female versus Male transformer.** Same stage and collection date,
    different sex.
  * **Larvae pooled versus transformers pooled.** The stage contrast, which is
    also the collection date contrast.

Cell numbers are matched across every group and every comparison, since
disagreement between marker lists otherwise tracks statistical power rather than
biology. Each cluster is subsampled to a common `n` in every group, and the whole
procedure is repeated over independent draws. `findMarkers` arguments are
identical to those used above, and library is deliberately not blocked, since
library is the variable being measured.

```{r}
library_discordance = function(sce, labels, library_vec, min_cells = 20,
                               n_draws = 10, n_top = 50, seed = 1) {
  labels      = as.character(labels)
  library_vec = as.character(library_vec)

  singles = c("Undifferentiated.1", "Undifferentiated.2", "Female", "Male")
  group_defs = list(
    "Undifferentiated.1"  = "Undifferentiated.1",
    "Undifferentiated.2"  = "Undifferentiated.2",
    "Female"              = "Female",
    "Male"                = "Male",
    "larvae_pooled"       = c("Undifferentiated.1", "Undifferentiated.2"),
    "transformers_pooled" = c("Female", "Male"))

  comparisons = list(
    UND1_vs_UND2           = c("Undifferentiated.1", "Undifferentiated.2"),
    Female_vs_Male         = c("Female", "Male"),
    larvae_vs_transformers = c("larvae_pooled", "transformers_pooled"))

  clusters_all = sort(unique(labels))
  cnt = sapply(singles, function(lb)
    sapply(clusters_all, function(cl) sum(labels == cl & library_vec == lb)))
  if (is.null(dim(cnt))) cnt = matrix(cnt, nrow = length(clusters_all),
                                      dimnames = list(clusters_all, singles))
  rownames(cnt) = clusters_all

  eligibility = as.data.frame(cnt)
  eligibility$label = clusters_all
  eligibility$min_across_libraries = apply(cnt, 1, min)
  eligibility$eligible = eligibility$min_across_libraries >= min_cells
  rownames(eligibility) = NULL

  eligible = clusters_all[eligibility$eligible]
  if (length(eligible) < 2) {
    warning("Fewer than two clusters have enough cells in every library; ",
            "pairwise library discordance cannot be computed.")
    return(list(results = NULL, eligibility = eligibility, n_used = NA_integer_,
                eligible = eligible))
  }

  # every group, in every comparison, gets exactly this many cells per cluster
  n_used = min(cnt[eligible, , drop = FALSE])

  set.seed(seed)
  res = list()

  for (d in seq_len(n_draws)) {
    markers_by_group = list()

    for (gname in names(group_defs)) {
      libs = group_defs[[gname]]
      idx  = integer(0)
      ok   = TRUE
      for (cl in eligible) {
        pool = which(labels == cl & library_vec %in% libs)
        if (length(pool) < n_used) { ok = FALSE; break }
        idx = c(idx, pool[sample(length(pool), n_used)])
      }
      if (!ok) next
      markers_by_group[[gname]] = findMarkers(
        sce[, idx], droplevels(factor(labels[idx])),
        test.type = "t", direction = "up", pval.type = "some")
    }

    for (cmp in names(comparisons)) {
      g1 = comparisons[[cmp]][1]
      g2 = comparisons[[cmp]][2]
      if (is.null(markers_by_group[[g1]]) || is.null(markers_by_group[[g2]])) next
      res[[length(res) + 1]] = data.frame(
        comparison  = cmp,
        draw        = d,
        label       = eligible,
        overlap_pct = vapply(eligible, function(cl)
          top_overlap_pct(markers_by_group[[g1]], markers_by_group[[g2]], cl, n_top),
          numeric(1)),
        rank_cor    = vapply(eligible, function(cl)
          stat_rank_cor(markers_by_group[[g1]], markers_by_group[[g2]], cl),
          numeric(1)),
        stringsAsFactors = FALSE)
    }
  }

  out = do.call(rbind, res)
  if (!is.null(out)) rownames(out) = NULL
  list(results = out, eligibility = eligibility, n_used = n_used,
       eligible = eligible)
}
```

```{r}
discordance_file = paste0(Robject_out_dir, "/library_discordance.Rds")

if (file.exists(discordance_file)) {
  discordance = readRDS(discordance_file)
} else {
  message("Computing pairwise library discordance. Cached in ",
          discordance_file, ".")
  discordance = list(
    germ = library_discordance(combined.sex.sce, colData(merged.sex.sce)$label,
                               germ_meta$library),
    all  = library_discordance(combined.sce, colData(merged.sce)$label,
                               all_meta$library))
  saveRDS(discordance, file = discordance_file)
}
```

#### Cells used, and clusters excluded

```{r}
disc_eligibility = bind_rows(
  discordance$germ$eligibility %>% mutate(clustering = "germ_subcluster"),
  discordance$all$eligibility  %>% mutate(clustering = "all_cells_cluster")) %>%
  dplyr::select(clustering, label, everything())

write.csv(disc_eligibility,
          file = paste0(results_dir, "/SuppTable_discordance_eligibility.csv"),
          row.names = FALSE)

datatable(disc_eligibility, filter = "top", rownames = FALSE)
```

* Germ cell subclusters: `r length(discordance$germ$eligible)` eligible, subsampled to **`r discordance$germ$n_used` cells per cluster per group**.
* All-cells clusters: `r length(discordance$all$eligible)` eligible, subsampled to **`r discordance$all$n_used` cells per cluster per group**.

```{r}
disc_excluded = disc_eligibility %>% filter(!eligible)
```

`r if (nrow(disc_excluded) == 0) "No cluster was excluded; every cluster has enough cells in all four libraries." else paste0(nrow(disc_excluded), " cluster(s) were excluded for having fewer than the required cells in at least one library. A cluster confined to one sex, such as the oocyte and spermatocyte clusters, cannot appear in this analysis at all, since the Female versus Male comparison requires cells of that cluster in both transformer libraries.")`

#### Discordance per cluster

Both metrics are measures of agreement, so lower values mean greater
discordance. The ratio columns express each comparison against the
Undifferentiated.1 versus Undifferentiated.2 floor for the same cluster: a ratio
below 1 means that comparison agrees less well than replicate libraries of the
same stage do.

```{r}
disc_results = bind_rows(
  if (!is.null(discordance$germ$results))
    discordance$germ$results %>% mutate(clustering = "germ_subcluster"),
  if (!is.null(discordance$all$results))
    discordance$all$results  %>% mutate(clustering = "all_cells_cluster"))

disc_per_cluster = disc_results %>%
  group_by(clustering, label, comparison) %>%
  summarize(n_draws        = dplyr::n(),
            overlap_median = median(overlap_pct, na.rm = TRUE),
            overlap_min    = min(overlap_pct, na.rm = TRUE),
            overlap_max    = max(overlap_pct, na.rm = TRUE),
            rankcor_median = median(rank_cor, na.rm = TRUE),
            rankcor_min    = min(rank_cor, na.rm = TRUE),
            rankcor_max    = max(rank_cor, na.rm = TRUE),
            .groups = "drop")

floor_per_cluster = disc_per_cluster %>%
  filter(comparison == "UND1_vs_UND2") %>%
  dplyr::select(clustering, label,
                floor_overlap = overlap_median,
                floor_rankcor = rankcor_median)

disc_per_cluster = disc_per_cluster %>%
  left_join(floor_per_cluster, by = c("clustering", "label")) %>%
  mutate(overlap_ratio_to_floor = ifelse(!is.na(floor_overlap) & floor_overlap > 0,
                                         overlap_median / floor_overlap, NA_real_),
         rankcor_ratio_to_floor = ifelse(!is.na(floor_rankcor) & floor_rankcor > 0,
                                         rankcor_median / floor_rankcor, NA_real_))

write.csv(disc_per_cluster,
          file = paste0(results_dir, "/SuppTable_library_discordance.csv"),
          row.names = FALSE)

datatable(disc_per_cluster %>% mutate(across(where(is.numeric), ~round(., 3))),
          filter = "top", rownames = FALSE)
```

#### Overall

```{r}
disc_overall = disc_results %>%
  group_by(clustering, comparison) %>%
  summarize(n_obs          = dplyr::n(),
            overlap_median = median(overlap_pct, na.rm = TRUE),
            overlap_min    = min(overlap_pct, na.rm = TRUE),
            overlap_max    = max(overlap_pct, na.rm = TRUE),
            rankcor_median = median(rank_cor, na.rm = TRUE),
            rankcor_min    = min(rank_cor, na.rm = TRUE),
            rankcor_max    = max(rank_cor, na.rm = TRUE),
            .groups = "drop")

floor_overall = disc_overall %>%
  filter(comparison == "UND1_vs_UND2") %>%
  dplyr::select(clustering,
                floor_overlap = overlap_median, floor_rankcor = rankcor_median,
                floor_overlap_min = overlap_min, floor_overlap_max = overlap_max)

disc_overall = disc_overall %>%
  left_join(floor_overall, by = "clustering") %>%
  mutate(overlap_ratio_to_floor = overlap_median / floor_overlap,
         rankcor_ratio_to_floor = rankcor_median / floor_rankcor,
         within_floor_range = overlap_median >= floor_overlap_min &
                              overlap_median <= floor_overlap_max)

datatable(disc_overall %>% mutate(across(where(is.numeric), ~round(., 3))),
          rownames = FALSE)
```

```{r, fig.width=10, fig.height=6}
cmp_cols = c(UND1_vs_UND2 = "#009292", Female_vs_Male = "#006ddb",
             larvae_vs_transformers = "#9e2f7f")

p_disc = ggplot(disc_per_cluster,
                aes(x = factor(label), y = overlap_median, color = comparison)) +
  geom_linerange(aes(ymin = overlap_min, ymax = overlap_max),
                 position = position_dodge(width = 0.6), linewidth = 0.6) +
  geom_point(position = position_dodge(width = 0.6), size = 2) +
  scale_color_manual(values = cmp_cols) +
  facet_wrap(~ clustering, scales = "free_x", ncol = 1) +
  xlab("Cluster") +
  ylab("Top 50 marker overlap (%)") +
  labs(caption = paste0("Points are medians across draws, bars the range. ",
                        "UND1 vs UND2 is the replicate floor."))

p_disc
ggsave(p_disc, file = paste0(results_dir, "/library_discordance_by_cluster.png"),
       width = 10, height = 6)
```

```{r}
disc_read = function(clust) {
  d = disc_overall %>% filter(clustering == clust)
  if (nrow(d) == 0) return(NULL)
  list(
    floor = d %>% filter(comparison == "UND1_vs_UND2"),
    fm    = d %>% filter(comparison == "Female_vs_Male"),
    lt    = d %>% filter(comparison == "larvae_vs_transformers"))
}

dg = disc_read("germ_subcluster")
da = disc_read("all_cells_cluster")

disc_sentence = function(x, what) {
  if (is.null(x) || nrow(x$floor) == 0) return(paste0("No ", what, " result available."))
  parts = c()
  fl = x$floor
  parts = c(parts, paste0(
    "Replicate floor (Undifferentiated.1 versus Undifferentiated.2): median top 50 overlap ",
    round(fl$overlap_median, 1), "% (range ", round(fl$overlap_min, 1), " to ",
    round(fl$overlap_max, 1), "%), median rank correlation ",
    round(fl$rankcor_median, 2), "."))
  if (nrow(x$fm) > 0) {
    parts = c(parts, paste0(
      "Female versus Male, reported for completeness only: median overlap ",
      round(x$fm$overlap_median, 1), "%, rank correlation ",
      round(x$fm$rankcor_median, 2), "."))
  }
  if (nrow(x$lt) > 0) {
    parts = c(parts, paste0(
      "Larvae versus transformers: median overlap ", round(x$lt$overlap_median, 1),
      "%, which is ", round(x$lt$overlap_ratio_to_floor, 2),
      " times the floor, and rank correlation ", round(x$lt$rankcor_median, 2),
      ". This ", if (isTRUE(x$lt$within_floor_range))
        "falls inside the range of the larval replicate floor, so the stage contrast produces no more disagreement than replicate variation alone."
      else if (x$lt$overlap_median < x$lt$floor_overlap_min)
        "falls below the floor range, so the stage contrast produces more disagreement than replicate variation alone."
      else "falls above the floor range."))
  }
  paste(parts, collapse = " ")
}
```

**Germ cell subclusters.** `r disc_sentence(dg, "germ subcluster")`

**All-cells clusters.** `r disc_sentence(da, "all-cells cluster")`

```{r}
lt_within = !is.null(dg) && nrow(dg$lt) > 0 && isTRUE(dg$lt$within_floor_range)
```

#### How to read these numbers

The larvae versus transformers comparison is the one that bears on the
reviewer's concern. Markers here are computed on uncorrected expression, so this
comparison contains the July to November difference in full. If it does not
exceed the replicate floor, then in these clusters the batch difference adds no
more systematic variation than two animals of the same stage do, which is
evidence that the batch effect is modest where it can be measured.

The two outcomes are not symmetric. Exceeding the floor is ambiguous, since real
stage biology and batch effect push in the same direction and this comparison
cannot separate them. Not exceeding it is the informative outcome.

Two limits on how far that generalizes:

  * The testable clusters are those with enough cells in all four libraries,
    which excludes any population confined to one stage or one sex. Most of the
    germ cell states along the trajectory are of that kind. The batch effect is
    therefore being measured in the populations where it can be measured, not in
    the ones the trajectory claim depends on.
  * A cluster containing cells from all four libraries is one that `fastMNN`
    succeeded in merging. Clusters were defined on the corrected values, so
    co-membership by itself is not evidence of a small batch effect: the
    correction exists to produce it. What is evidence is that the *uncorrected*
    marker profiles within those clusters agree across libraries, since the
    correction never touched those values. The residual gap is populations where
    the batch effect was large enough that correction failed and the libraries
    formed separate clusters, which would be excluded here. That gap is
    checkable, and is addressed immediately below.

#### Are there clusters the correction failed to merge?

If a batch effect were large, cells of the same type from different libraries
would fail to merge and would appear as separate, library-dominated clusters.
Those clusters would be excluded from the discordance analysis, hiding the worst
cases. The check is whether any stage-dominated cluster lacks a biological
explanation.

```{r}
stage_dominated = population_sharing %>%
  filter(sharing %in% c("larva-dominated", "transformer-dominated")) %>%
  dplyr::select(clustering, label, population, population_type,
                n_larva, n_transformer, pct_larva, sharing)

datatable(stage_dominated %>% mutate(across(where(is.numeric), ~round(., 1))),
          filter = "top", rownames = FALSE)
```

`r if (nrow(stage_dominated) == 0) "No cluster is dominated by a single stage, so there is no population that the correction failed to merge, and the discordance analysis is not missing a set of high batch effect clusters." else paste0(nrow(stage_dominated), " cluster(s) are dominated by a single stage, of which ", sum(stage_dominated$population_type == "germ"), " are germ cell populations and ", sum(stage_dominated$population_type == "somatic"), " are somatic. Stage-restricted germ cell populations are expected on biological grounds, since oocytes and spermatocytes have no larval counterpart. A stage-dominated **somatic** cluster is the informative case: somatic cell types should be present at both stages, so if one is split by stage that is a candidate for a merge the correction did not achieve. Inspect the somatic rows above; if there are none, the discordance estimate is not missing a set of high batch effect clusters.")`

The Female versus Male comparison is computed and included in the tables and
figure for completeness, but nothing is concluded from it. The clusters testable
here are largely somatic and need not differ between sexes, so it serves neither
as a positive control nor as a result.

`r if (lt_within) "**The larvae versus transformers comparison falls within the replicate floor.** In the clusters that can be tested, the stage and collection date contrast produces no more marker disagreement than two larval replicate libraries do. Since these markers are computed on uncorrected expression, that is direct evidence the batch effect is small in these populations, subject to the two limits above." else "**The larvae versus transformers comparison exceeds the replicate floor.** The stage and collection date contrast produces more marker disagreement than two larval replicate libraries do. This comparison cannot say how much of that excess is real stage biology and how much is batch, since both push in the same direction."`

### Cross-stage agreement, as a supporting comparison

The comparison below pools libraries by stage and asks whether the same genes
mark each shared population in larvae and in transformers. It is reported after
the pairwise analysis above because it cannot separate stage from library on its
own: it has no reference floor built into it. Read it against the replicate floor
established above.

```{r}
marker_agreement = shared_pops %>%
  rowwise() %>%
  mutate(
    grp = list(if (clustering == "germ_subcluster") stage_split$germ
               else stage_split$all),
    in_common       = as.character(label) %in% grp$common,
    top_larva       = list(top_markers_for(grp$markers[["larva"]],
                                           as.character(label))),
    top_transformer = list(top_markers_for(grp$markers[["transformer"]],
                                           as.character(label))),
    n_top_larva       = length(top_larva),
    n_top_transformer = length(top_transformer),
    n_shared_markers  = length(intersect(top_larva, top_transformer)),
    cross_stage_overlap_pct = ifelse(in_common,
                                     top_overlap_pct(grp$markers[["larva"]],
                                                     grp$markers[["transformer"]],
                                                     as.character(label)),
                                     NA_real_),
    cross_stage_rank_cor = ifelse(in_common,
                                  stat_rank_cor(grp$markers[["larva"]],
                                                grp$markers[["transformer"]],
                                                as.character(label)),
                                  NA_real_),
    shared_marker_genes = paste(intersect(top_larva, top_transformer),
                                collapse = ";")) %>%
  ungroup() %>%
  dplyr::select(-grp, -top_larva, -top_transformer)
```

### Split-half control

Cells are split within each cluster, so both halves contain every cluster at
roughly half size and the comparison background is unchanged. The same
`findMarkers` arguments and the same overlap and correlation functions are used.
This is the expensive step in the document and is cached separately; delete
`split_half_control.Rds` to recompute.

```{r}
split_half_control = function(sce, labels, stage_vec, common, stage,
                              n_splits = 10, n_top = 50, seed = 1) {
  labels    = as.character(labels)
  stage_vec = as.character(stage_vec)

  in_stage = which(stage_vec == stage & labels %in% common)
  lab_s    = labels[in_stage]
  if (length(unique(lab_s)) < 2) return(NULL)

  # seeded once, outside the loop, so each of the n_splits iterations draws a
  # different partition while the set as a whole stays reproducible
  set.seed(seed)
  res = list()

  for (r in seq_len(n_splits)) {
    # a fresh stratified 50/50 split within each cluster on every replicate
    by_clust = split(seq_along(lab_s), lab_s)
    pick     = unlist(lapply(by_clust, function(ii) {
      ii[sample(length(ii), floor(length(ii) / 2))]
    }), use.names = FALSE)

    is_half1 = rep(FALSE, length(lab_s))
    is_half1[pick] = TRUE

    lab1 = droplevels(factor(lab_s[is_half1]))
    lab2 = droplevels(factor(lab_s[!is_half1]))
    if (nlevels(lab1) < 2 || nlevels(lab2) < 2) next
    if (min(table(lab1)) < 2 || min(table(lab2)) < 2) next

    m1 = findMarkers(sce[, in_stage[is_half1]],  lab1,
                     test.type = "t", direction = "up", pval.type = "some")
    m2 = findMarkers(sce[, in_stage[!is_half1]], lab2,
                     test.type = "t", direction = "up", pval.type = "some")

    res[[r]] = data.frame(
      stage       = stage,
      replicate   = r,
      label       = common,
      overlap_pct = vapply(common, function(lb) top_overlap_pct(m1, m2, lb, n_top),
                           numeric(1)),
      rank_cor    = vapply(common, function(lb) stat_rank_cor(m1, m2, lb),
                           numeric(1)),
      stringsAsFactors = FALSE)
  }

  out = do.call(rbind, res)
  if (!is.null(out)) rownames(out) = NULL
  out
}
```

```{r}
split_half_file = paste0(Robject_out_dir, "/split_half_control.Rds")

if (file.exists(split_half_file)) {
  split_half = readRDS(split_half_file)
} else {
  message("Running the split-half control. This is the slowest step in the ",
          "document; the result is cached in ", split_half_file, ".")
  split_half = bind_rows(
    bind_rows(lapply(c("larva", "transformer"), function(st)
      split_half_control(combined.sex.sce, colData(merged.sex.sce)$label,
                         germ_meta$stage, stage_split$germ$common, st))) %>%
      mutate(clustering = "germ_subcluster"),
    bind_rows(lapply(c("larva", "transformer"), function(st)
      split_half_control(combined.sce, colData(merged.sce)$label,
                         all_meta$stage, stage_split$all$common, st))) %>%
      mutate(clustering = "all_cells_cluster"))
  saveRDS(split_half, file = split_half_file)
}
```

```{r}
split_half_summary = split_half %>%
  group_by(clustering, label, stage) %>%
  summarize(n_replicates      = dplyr::n(),
            splithalf_overlap_median = median(overlap_pct, na.rm = TRUE),
            splithalf_overlap_min    = min(overlap_pct, na.rm = TRUE),
            splithalf_overlap_max    = max(overlap_pct, na.rm = TRUE),
            splithalf_rankcor_median = median(rank_cor, na.rm = TRUE),
            .groups = "drop")

datatable(split_half_summary %>% mutate(across(where(is.numeric), ~round(., 2))),
          filter = "top", rownames = FALSE)
```

Overall, across all replicates and clusters:

```{r}
split_half_overall = split_half %>%
  left_join(population_sharing %>%
              dplyr::select(clustering, label, population_type),
            by = c("clustering", "label")) %>%
  group_by(clustering, population_type, stage) %>%
  summarize(n_obs = dplyr::n(),
            overlap_median = median(overlap_pct, na.rm = TRUE),
            overlap_min    = min(overlap_pct, na.rm = TRUE),
            overlap_max    = max(overlap_pct, na.rm = TRUE),
            rankcor_median = median(rank_cor, na.rm = TRUE),
            .groups = "drop")

datatable(split_half_overall %>% mutate(across(where(is.numeric), ~round(., 2))),
          rownames = FALSE)
```

### Cross-stage agreement against the ceiling

```{r}
sh_wide = split_half_summary %>%
  tidyr::pivot_wider(id_cols = c(clustering, label),
                     names_from = stage,
                     values_from = c(splithalf_overlap_median,
                                     splithalf_overlap_min,
                                     splithalf_overlap_max,
                                     splithalf_rankcor_median))

population_sharing = population_sharing %>%
  left_join(marker_agreement %>%
              dplyr::select(clustering, label, in_common, n_top_larva,
                            n_top_transformer, n_shared_markers,
                            cross_stage_overlap_pct, cross_stage_rank_cor,
                            shared_marker_genes),
            by = c("clustering", "label")) %>%
  left_join(sh_wide, by = c("clustering", "label")) %>%
  mutate(ceiling_overlap = pmin(splithalf_overlap_median_larva,
                                splithalf_overlap_median_transformer,
                                na.rm = TRUE),
         overlap_vs_ceiling = ifelse(!is.na(ceiling_overlap) & ceiling_overlap > 0,
                                     cross_stage_overlap_pct / ceiling_overlap,
                                     NA_real_),
         min_stage_n = pmin(n_larva, n_transformer),
         low_cell_flag = min_stage_n < 50)

write.csv(population_sharing,
          file = paste0(results_dir, "/SuppTable_population_sharing.csv"),
          row.names = FALSE)

datatable(population_sharing %>%
            dplyr::select(-shared_marker_genes) %>%
            mutate(across(where(is.numeric), ~round(., 2))),
          filter = "top", rownames = FALSE)
```

```{r, fig.width=10, fig.height=6}
plot_df = population_sharing %>%
  filter(in_common, !is.na(cross_stage_overlap_pct)) %>%
  mutate(pop_label = paste0(clustering, " ", label, ": ", population))

sh_points = split_half_summary %>%
  inner_join(plot_df %>% dplyr::select(clustering, label, pop_label,
                                       population_type),
             by = c("clustering", "label"))

p_ceiling = ggplot() +
  geom_linerange(data = sh_points,
                 aes(x = pop_label, ymin = splithalf_overlap_min,
                     ymax = splithalf_overlap_max, color = stage),
                 position = position_dodge(width = 0.5), linewidth = 0.6) +
  geom_point(data = sh_points,
             aes(x = pop_label, y = splithalf_overlap_median, color = stage),
             position = position_dodge(width = 0.5), size = 2) +
  geom_point(data = plot_df,
             aes(x = pop_label, y = cross_stage_overlap_pct),
             shape = 4, size = 3, stroke = 1.2, color = "black") +
  scale_color_manual(values = c(larva = "#009292", transformer = "#FF6DB6")) +
  coord_flip() +
  facet_wrap(~ population_type, scales = "free_y", ncol = 1) +
  ylab("Top 50 marker overlap (%)") +
  xlab(NULL) +
  labs(color = "Split-half within stage",
       caption = "Cross (x) is the cross-stage overlap. Dots and bars are the within-stage split-half median and range.")

p_ceiling
ggsave(p_ceiling, file = paste0(results_dir, "/overlap_vs_splithalf_ceiling.png"),
       width = 10, height = 6)
```

```{r}
germ_agree = population_sharing %>%
  filter(population_type == "germ", !is.na(cross_stage_overlap_pct))

germ_ceiling = germ_agree %>% filter(!is.na(ceiling_overlap))
```

`r if (nrow(germ_agree) == 0) "No shared germ cell population had enough cells of both stages to compute stage-split markers, so marker agreement could not be assessed." else paste0("Across ", nrow(germ_agree), " shared germ cell population(s), cross-stage top 50 overlap ranges from ", round(min(germ_agree$cross_stage_overlap_pct), 1), "% to ", round(max(germ_agree$cross_stage_overlap_pct), 1), "% (median ", round(median(germ_agree$cross_stage_overlap_pct), 1), "%). The within-stage split-half ceiling for the same clusters has a median of ", if (nrow(germ_ceiling) > 0) round(median(germ_ceiling$ceiling_overlap), 1) else NA, "%, so cross-stage overlap reaches a median of ", if (nrow(germ_ceiling) > 0) round(100 * median(germ_ceiling$overlap_vs_ceiling, na.rm = TRUE), 0) else NA, "% of what is achievable within a single stage. Cross-stage rank correlation across all tested genes has a median of ", round(median(germ_agree$cross_stage_rank_cor, na.rm = TRUE), 2), ".")`

```{r}
ceiling_low = nrow(germ_ceiling) > 0 && median(germ_ceiling$ceiling_overlap) < 50
```

`r if (nrow(germ_ceiling) == 0) "" else if (ceiling_low) "**The ceiling is low.** Marker lists do not reproduce well even between random halves of the same cells, so top 50 overlap is not a usable measure of agreement in this dataset. Read the rank correlation across all genes instead, which does not depend on an arbitrary cutoff." else "The ceiling is high enough that top 50 overlap is interpretable here, and the cross-stage figure should be read as a fraction of it rather than in absolute terms."`

### The lowest-agreement population

```{r}
worst = germ_agree %>% slice_min(cross_stage_overlap_pct, n = 1, with_ties = FALSE)

worst_sh = split_half_summary %>%
  filter(clustering == worst$clustering, label == worst$label)

worst_sh_min = if (nrow(worst_sh) > 0) min(worst_sh$splithalf_overlap_median, na.rm = TRUE) else NA_real_

# low cross-stage overlap is a power or instability artifact if the same cluster
# also fails to reproduce within a single stage, or if one stage is thin
worst_is_artifact = (!is.na(worst_sh_min) && worst_sh_min < 2 * worst$cross_stage_overlap_pct) ||
                     worst$min_stage_n < 50
```

The shared germ cell population with the lowest cross-stage agreement is
**`r worst$population`** (`r worst$clustering` `r worst$label`), with
`r worst$n_larva` larval and `r worst$n_transformer` transformer cells, and a
cross-stage top 50 overlap of `r round(worst$cross_stage_overlap_pct, 1)`%.

```{r}
datatable(worst_sh %>% mutate(across(where(is.numeric), ~round(., 2))),
          rownames = FALSE)
```

`r if (nrow(worst_sh) == 0) "No split-half control is available for this cluster, so the low overlap cannot be attributed." else if (worst_is_artifact) paste0("Its own split-half overlap is only ", round(worst_sh_min, 1), "% within a single stage, and the sparser stage contributes ", worst$min_stage_n, " cells. Its markers therefore do not reproduce even between random halves of the same cells, so the low cross-stage figure reflects instability and limited power rather than a biological difference between stages. It should be described that way and not used as evidence that the population differs between stages.") else paste0("Its split-half overlap is ", round(worst_sh_min, 1), "% within a single stage, and both stages contribute ample cells (minimum ", worst$min_stage_n, "). Markers reproduce within a stage but not across stages, which is a substantive finding: the cells assigned to this cluster are not defined by the same genes in larvae and in transformers, and the identity of this cluster is questionable.")`

### Cell counts for every shared population

Overlap percentages are not interpretable without them. Populations where one
stage contributes fewer than 50 cells are flagged.

```{r}
datatable(population_sharing %>%
            filter(sharing == "shared") %>%
            dplyr::select(clustering, label, population, population_type,
                          n_larva, n_transformer, min_stage_n, low_cell_flag,
                          cross_stage_overlap_pct, ceiling_overlap,
                          overlap_vs_ceiling, cross_stage_rank_cor) %>%
            mutate(across(where(is.numeric), ~round(., 2))),
          filter = "top", rownames = FALSE)
```

# 5. Marker behavior along pseudotime

Uncorrected, per-library normalized log-expression is used here, not the
`fastMNN` output, so any trend cannot be an artifact of the correction. Each
marker is plotted against pseudotime per branch and colored by library, so it is
visible whether a trend is carried by one sample or holds within several.

Note that normalization is per library (`logNormCounts` applied to each sample in
`02`), so absolute levels are not strictly comparable across libraries. The
within-library trend, reported separately below, does not depend on that
comparability; the pooled trend does.

```{r}
gene_description =
  read_excel(paste0(data_dir, "/41467_2024_55019_MOESM3_ESM.xlsx"), skip = 1) %>%
  dplyr::rename(WangGeneID = "CountMatrix_GeneID",
                Gene        = "NCBI_GTF_GENEID",
                description = "NCBI_GTF_ProteinInfo") %>%
  dplyr::select(Gene, WangGeneID, description) %>%
  filter(!is.na(Gene))

markers_wide = read.csv(paste0(data_dir, "/Cell_Type_Marker_Genes_12-12-25.csv"))
markers_wide = markers_wide[, 2:ncol(markers_wide)]

markers_long = markers_wide %>%
  pivot_longer(cols = everything(), names_to = "CellType", values_to = "Gene",
               values_drop_na = TRUE) %>%
  mutate(main_part = str_remove(Gene, "\\s*\\(.*\\)")) %>%
  tidyr::separate_rows(main_part, sep = "/") %>%
  mutate(main_part = trimws(main_part)) %>%
  filter(main_part != "")

germ_cell_types = c("Migrating.germ.cells", "Primordial.germ.cells", "Oocyte",
                    "Oogonia", "Spermatogonia", "Spermatocytes",
                    "Spermatogonial.stem.cells")

requested = unique(c("DDX4", "PIWIL1", "SYCP3", "BOLL",
                     markers_long %>%
                       filter(CellType %in% germ_cell_types) %>%
                       pull(main_part)))

resolve_marker = function(symbol, sce_rownames, gd) {
  if (symbol %in% sce_rownames) return(symbol)
  g = gd$Gene[gd$WangGeneID == symbol]
  g = g[g %in% sce_rownames]
  if (length(g) > 0) return(g[1])
  NA_character_
}

marker_map = data.frame(symbol = requested, stringsAsFactors = FALSE) %>%
  rowwise() %>%
  mutate(gene_id = resolve_marker(symbol, rownames(combined.sex.sce),
                                  gene_description)) %>%
  ungroup()

resolved   = marker_map %>% filter(!is.na(gene_id))
unresolved = marker_map %>% filter(is.na(gene_id))
```

* Annotation markers requested: `r nrow(marker_map)`
* Resolved in `rownames(combined.sex.sce)`: `r nrow(resolved)`
* Not found, and therefore excluded: `r if (nrow(unresolved) == 0) "none" else paste(unresolved$symbol, collapse = ", ")`

```{r}
stopifnot(identical(colnames(combined.sex.sce), colnames(merged.sex.sce)))

expr_long =
  as.matrix(logcounts(combined.sex.sce)[resolved$gene_id, , drop = FALSE]) %>%
  t() %>%
  as.data.frame() %>%
  tibble::rownames_to_column("barcode") %>%
  tidyr::pivot_longer(-barcode, names_to = "gene_id", values_to = "logexpr") %>%
  left_join(resolved, by = "gene_id")

expr_branch = branch_long %>%
  dplyr::select(barcode, branch, pseudotime, library, stage) %>%
  inner_join(expr_long, by = "barcode")
```

```{r}
safe_rho = function(x, y) {
  if (length(x) < 10) return(NA_real_)
  if (length(unique(x)) < 2 || length(unique(y)) < 2) return(NA_real_)
  suppressWarnings(cor(x, y, method = "spearman"))
}

safe_slope = function(x, y) {
  if (length(x) < 10 || length(unique(x)) < 2) return(NA_real_)
  unname(coef(lm(y ~ x))[2])
}

trend_overall = expr_branch %>%
  group_by(symbol, gene_id, branch) %>%
  summarize(n_cells        = n(),
            pct_expressing = 100 * mean(logexpr > 0),
            mean_logexpr   = mean(logexpr),
            spearman_rho   = safe_rho(pseudotime, logexpr),
            slope_per_unit_pseudotime = safe_slope(pseudotime, logexpr),
            .groups = "drop") %>%
  mutate(direction = case_when(
    is.na(spearman_rho)     ~ "undetermined",
    spearman_rho >=  0.10   ~ "increases",
    spearman_rho <= -0.10   ~ "decreases",
    TRUE                    ~ "flat"))

trend_by_library = expr_branch %>%
  group_by(symbol, branch, library) %>%
  summarize(n_cells = n(),
            rho     = safe_rho(pseudotime, logexpr),
            .groups = "drop") %>%
  filter(!is.na(rho), n_cells >= 20)

trend_consistency = trend_by_library %>%
  left_join(trend_overall %>% dplyr::select(symbol, branch, spearman_rho),
            by = c("symbol", "branch")) %>%
  filter(!is.na(spearman_rho)) %>%
  group_by(symbol, branch) %>%
  summarize(n_libraries_tested = n(),
            n_libraries_same_sign = sum(sign(rho) == sign(spearman_rho)),
            min_library_rho = min(rho),
            max_library_rho = max(rho),
            .groups = "drop") %>%
  mutate(consistent_within_libraries =
           n_libraries_tested >= 2 & n_libraries_same_sign == n_libraries_tested)

marker_trends = trend_overall %>%
  left_join(trend_consistency, by = c("symbol", "branch")) %>%
  arrange(branch, desc(abs(spearman_rho)))

write.csv(marker_trends,
          file = paste0(results_dir, "/SuppTable_marker_trends.csv"),
          row.names = FALSE)

datatable(marker_trends %>% mutate(across(where(is.numeric), ~round(., 3))),
          filter = "top", rownames = FALSE)
```

## Core markers

Points are individual cells, lines are loess fits per library. One row per
marker, one column per branch.

```{r, fig.width=10, fig.height=8}
core_markers = intersect(c("DDX4", "PIWIL1", "SYCP3", "BOLL"), resolved$symbol)

p_marker = ggplot(expr_branch %>% filter(symbol %in% core_markers),
                  aes(x = pseudotime, y = logexpr, color = library)) +
  geom_point(size = 0.15, alpha = 0.12) +
  geom_smooth(method = "loess", se = FALSE, linewidth = 1) +
  scale_color_manual(values = lib_cols) +
  facet_grid(symbol ~ branch, scales = "free") +
  theme(legend.position = "top") +
  guides(color = guide_legend(override.aes = list(alpha = 1, linewidth = 2))) +
  xlab("Pseudotime") + ylab("Uncorrected log-expression")

p_marker
ggsave(p_marker, file = paste0(results_dir, "/marker_expression_vs_pseudotime.png"),
       width = 10, height = 8)
```

## All annotation markers, as a trend heatmap

Plotting every marker as its own panel is unreadable once there are more than a
handful, so the full set is shown as the statistic itself. Each tile is the
Spearman correlation between expression and pseudotime, computed within a single
library. Red is increasing, blue is decreasing, white is flat, and a blank tile
means that library had fewer than 20 cells on that branch.

A marker whose row is the same color across libraries has a trend that holds
within each sample separately. A marker whose row changes color between
libraries does not, and its pooled trend is likely a difference between samples.

```{r, fig.width=9, fig.height=10}
marker_order = trend_by_library %>%
  group_by(symbol) %>%
  summarize(m = mean(rho, na.rm = TRUE), .groups = "drop") %>%
  arrange(m) %>%
  pull(symbol)

p_rho = ggplot(trend_by_library %>%
                 mutate(symbol = factor(symbol, levels = marker_order)),
               aes(x = library, y = symbol, fill = rho)) +
  geom_tile(color = "grey90") +
  geom_text(aes(label = sprintf("%.2f", rho)), size = 1.9) +
  scale_fill_gradient2(low = "#2166ac", mid = "white", high = "#b2182b",
                       midpoint = 0, limits = c(-1, 1)) +
  facet_wrap(~ branch) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  xlab("Library") + ylab("Marker") +
  labs(fill = "Spearman rho")

p_rho
ggsave(p_rho, file = paste0(results_dir, "/marker_trend_heatmap.png"),
       width = 9, height = 10)
```

## All annotation markers, expression against pseudotime

The same markers as curves, split across pages of six so each panel stays large
enough to read. Points are omitted here; only the per-library loess fits are
drawn.

```{r, fig.width=10, fig.height=7}
# markers with no library passing the cell threshold are absent from
# marker_order, so append them rather than dropping them from the figures
ordered_symbols = c(marker_order, setdiff(resolved$symbol, marker_order))
marker_pages    = split(ordered_symbols,
                        ceiling(seq_along(ordered_symbols) / 6))

for (i in seq_along(marker_pages)) {
  p_page = ggplot(expr_branch %>% filter(symbol %in% marker_pages[[i]]),
                  aes(x = pseudotime, y = logexpr, color = library)) +
    geom_smooth(method = "loess", se = FALSE, linewidth = 0.8) +
    scale_color_manual(values = lib_cols) +
    facet_grid(symbol ~ branch, scales = "free") +
    theme(legend.position = "top") +
    xlab("Pseudotime") + ylab("Uncorrected log-expression")

  print(p_page)
  ggsave(p_page,
         file = paste0(results_dir, "/marker_trends_page", i, ".png"),
         width = 10, height = 7)
}
```

# 6. Summary

```{r}
n_shared       = nrow(shared_pops)
n_shared_germ  = nrow(shared_germ_pops)
n_shared_som   = nrow(shared_som_pops)

n_directional  = sum(marker_trends$direction %in% c("increases", "decreases"))
n_flat         = sum(marker_trends$direction == "flat")
n_tested       = nrow(marker_trends)
n_consistent   = sum(marker_trends$consistent_within_libraries, na.rm = TRUE)

germ_markers_ok = n_shared_germ > 0
markers_ok      = n_directional > 0 && n_consistent > 0
```

**Pairwise library discordance, the primary result.**
`r disc_sentence(dg, "germ subcluster")`

`r if (lt_within) paste0("The stage and collection date contrast falls within the range produced by two larval replicate libraries. Because these markers are computed on uncorrected expression, and the correction never touched those values, that is evidence the batch effect is small in the clusters where it can be measured. ", if (sum(stage_dominated$population_type == "somatic") == 0) "No somatic cluster is stage-dominated, so there is no population that the correction failed to merge and no set of high batch effect clusters excluded from this estimate." else paste0(sum(stage_dominated$population_type == "somatic"), " somatic cluster(s) are stage-dominated and should be inspected, since a somatic type split by stage is a candidate for a merge the correction did not achieve."), " The result does not extend automatically to the germ cell states along the trajectory, most of which are confined to one stage and so cannot enter this comparison.") else "The stage and collection date contrast exceeds the replicate floor. This comparison cannot apportion that excess between real stage biology and batch, since both act in the same direction."`

**Population sharing.** `r n_shared` of `r nrow(population_sharing)` populations
hold at least 10% of their cells from each stage: `r n_shared_germ` germ cell
populations and `r n_shared_som` somatic populations.

`r if (n_shared_germ == 0) "**No germ cell population is shared between stages.** Every germ cell cluster and subcluster is dominated by one stage. Sharing among somatic populations shows the integration can align cells collected on different dates, but it says nothing about whether the germ cell ordering is developmental, because no germ cell state is jointly occupied. On this evidence the trajectory claim is not supported." else paste0("Germ cell populations occupied by both stages are the strongest evidence available here, because a batch effect does not create a jointly occupied state. ", if (nrow(germ_ceiling) > 0) paste0("Markers computed separately within each stage overlap across stages by a median of ", round(median(germ_agree$cross_stage_overlap_pct), 1), "% in the top 50. That figure is only interpretable against the within-stage split-half ceiling, which is ", round(median(germ_ceiling$ceiling_overlap), 1), "%: cross-stage agreement therefore reaches ", round(100 * median(germ_ceiling$overlap_vs_ceiling, na.rm = TRUE), 0), "% of what is achievable between random halves of the same cells. Rank correlation of the test statistic across all tested genes has a median of ", round(median(germ_agree$cross_stage_rank_cor, na.rm = TRUE), 2), ".") else if (nrow(germ_agree) > 0) "Marker agreement was computed but no split-half control was available, so the overlap figures cannot be placed against a ceiling." else "Marker agreement could not be assessed for these populations because at least one stage had too few cells.")`

**Marker behavior.** Of `r n_tested` marker and branch combinations tested,
`r n_directional` show a directional trend along pseudotime (Spearman |rho| >=
0.10), `r n_flat` are flat, and `r n_consistent` have a trend of consistent sign
within every library tested separately.

A trend computed inside a single library holds the confound constant. Every cell
in that calculation came from the same animal, the same collection date and the
same capture, so none of those can be what produces the trend. If a marker rises
with pseudotime among the cells of one library alone, that rise is not the July
to November difference, because there is no July to November difference inside
one library. Trends of that kind bear on the ordering independently of the
confound; a trend seen only when libraries are pooled does not, since it may
just be a difference between libraries.

Note the asymmetry when reading the table. A library may occupy only part of the
pseudotime range, in which case its within-library correlation is computed over a
short stretch and will be attenuated. A strong within-library trend is therefore
good evidence for the ordering, but a weak one is not strong evidence against it.

`r if (!markers_ok) "**On these numbers, marker expression does not support the inferred ordering.** Either no marker changes monotonically along pseudotime, or no trend survives within individual libraries, meaning any apparent trend is carried by differences between samples rather than by position along the trajectory." else "**On these numbers, marker behavior supports the inferred ordering**, at least for the markers and branches where a within-library trend was detected."`

`r if (germ_markers_ok && markers_ok) "Taken together, shared germ cell populations and within-library marker trends provide evidence for the ordering that does not reduce to collection batch." else "Taken together, the evidence needed to separate development from collection batch is incomplete. State this limitation explicitly rather than resting the trajectory claim on the pseudotime composition, which cannot distinguish the two."`

## For the reviewer response

```{r, results="asis"}
cat(paste0(
  "We agree that developmental stage is confounded with collection date and ",
  "batch, and that the distribution of libraries along the trajectory cannot ",
  "by itself address this. Segregation of larval and transformer cells along ",
  "pseudotime is predicted both by a residual batch effect and by the ",
  "possibility that the underlying populations do not co-occur across stages, ",
  "so it does not distinguish them. We therefore report that distribution ",
  "descriptively ",
  "(Supplementary Tables 1 to 3) and base our argument on two analyses that ",
  "do distinguish the possibilities. First, ", n_shared,
  " populations contain at least 10% of their cells from each stage, of which ",
  n_shared_germ, " are germ cell populations",
  if (nrow(germ_ceiling) > 0) paste0(
    "; markers computed separately within larval and within transformer cells ",
    "for these populations overlap by a median of ",
    round(median(germ_agree$cross_stage_overlap_pct), 1),
    "% in the top 50, against a within-stage split-half ceiling of ",
    round(median(germ_ceiling$ceiling_overlap), 1),
    "%, that is ",
    round(100 * median(germ_ceiling$overlap_vs_ceiling, na.rm = TRUE), 0),
    "% of the agreement achievable between random halves of the same cells, ",
    "with a median rank correlation of the test statistic across all tested ",
    "genes of ", round(median(germ_agree$cross_stage_rank_cor, na.rm = TRUE), 2))
  else "",
  ". Second, of ", n_tested, " marker and branch combinations, ", n_directional,
  " show a directional trend along pseudotime and ", n_consistent,
  " retain a consistent trend within every individual library, computed on ",
  "uncorrected expression. ",
  if (germ_markers_ok && markers_ok) {
    "Because these trends are computed within a single library, collection date and batch are held constant by construction, so the trends cannot be produced by the difference between the July and the November samples."
  } else {
    "We note that this evidence is weaker than we would like, and we have revised the strength of our trajectory claim accordingly."
  }))
```

# Session info

```{r}
sessionInfo()
```
