---
title: "08_female_germ_cell_continuum"
author: "Stephanie Hickey"
date: "`r Sys.Date()`"
output:
  html_document:
    code_folding: hide
    toc: true
    toc_float:
      toc_collapsed: true
params:
    project_dir: /mnt/ufs18/rs-013/bioinformaticsCore/projects/chong_davidson/BCC125_sea_lamprey_scRNA
    project_name: BCC125_sea_lamprey_scRNA
    lab: Weiming Li
    female_all_cluster: 8
    min_library_cells: 50
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

```{r, echo=FALSE, results="asis"}
cat("\n")
cat("#### Lab:", params$lab, " \n")
cat("#### Project:", params$project_name, " \n")
```

Roughly 1,700 female germ cells survive doublet filtering, and
`buildSNNGraph(k = 8)` followed by Louvain does not split them. No attempt is
made here to force clusters by lowering `k` or raising resolution, since tuning
until clusters appear would be selecting a parameter to obtain a desired result.

The document runs two analyses.

**Combined.** All female germ cells together. The leading diffusion component
turns out to separate larval from transformer cells rather than describing a
gradient. Stage and collection date vary together, so that separation is
confounded, but a batch effect and a differentiation program are distinguishable
by whether the marker changes are **coherent**: a batch effect does not
coordinately move an entire known gene family in the biologically expected
direction. That coherence test decides how the separation is reported.

**Per library.** The same continuum test run inside each library separately.
There, animal, collection date and batch are constant by construction, so a
maturation gradient has nothing to be confounded with, and a correlation between
the leading component and detection rate is unambiguously technical.

**A pseudotime can be computed from any blob of cells**, so an ordering proves
nothing on its own. The technical and marker checks decide, not the existence of
an ordering.

Everything uses uncorrected, normalized log-expression. Running this on
`fastMNN` output would remove the library signal by construction and make the
central checks meaningless.

```{r}
library(tidyverse)
library(SingleCellExperiment)
library(scater)
library(scran)
library(batchelor)
library(patchwork)
library(readxl)
library(destiny)
library(DT)
source(paste0(params$project_dir, "/src/scRNA_seq_functions.R"))
```

```{r}
results_dir     = paste0(params$project_dir, "/results/20260804_female_continuum")
Robject_out_dir = paste0(results_dir, "/Robjects")
if (!dir.exists(Robject_out_dir)) dir.create(Robject_out_dir, recursive = TRUE)
data_dir        = paste0(params$project_dir, "/data")
```

```{r, echo=FALSE}
Robject_dir    = paste0(params$project_dir, "/results/20260803_no_doublets")
all.sex.sce    = readRDS(paste0(Robject_dir, "/no.support.sex.sce.Rds"))
merged.sex.sce = readRDS(paste0(Robject_dir, "/no.support.merged.sex.sce.Rds"))

# all cells, uncorrected, for the ambient RNA control in section 1
merged.all.sce   = readRDS(paste0(Robject_dir, "/merged.sce.Rds"))
combined.all.sce = readRDS(paste0(Robject_dir, "/combined.sce.Rds"))
```

```{r}
library_meta = data.frame(
  batch   = c("UND1_C00004336", "UND2_C00004362", "TF_C00004337", "TM_C00004374"),
  library = c("Undifferentiated.1", "Undifferentiated.2", "Female", "Male"),
  stage   = c("larva", "larva", "transformer", "transformer"),
  stringsAsFactors = FALSE)

lib_cols = c("Female" = "#FF6DB6", "Male" = "#006ddb",
             "Undifferentiated.1" = "#009292", "Undifferentiated.2" = "#b66ddf")
```

# Identify the female germ cells

```{r}
germ_meta = as.data.frame(colData(merged.sex.sce))
germ_meta$barcode = rownames(germ_meta)

if (!"all.scatter.clust" %in% colnames(germ_meta)) {
  stop(paste0("all.scatter.clust is not present, so female germ cells cannot ",
              "be identified by originating cluster. Check that ",
              "02_sex_cell_subanalysis_no_doublets.Rmd was run to completion."))
}

germ_meta$origin_cluster = as.character(germ_meta$all.scatter.clust)
female_cells = germ_meta$barcode[germ_meta$origin_cluster ==
                                   as.character(params$female_all_cluster)]
```

`r length(female_cells)` cells come from all-cells cluster
`r params$female_all_cluster`, the doublet-filtered oogonia and oocyte cluster.

## Libraries contributing, and which are dropped

```{r}
fem_list_all = lapply(all.sex.sce, function(x) {
  keep = intersect(colnames(x), female_cells)
  if (length(keep) == 0) return(NULL)
  x[, keep]
})
fem_list_all = fem_list_all[!vapply(fem_list_all, is.null, logical(1))]

lib_counts = data.frame(batch   = names(fem_list_all),
                        n_cells = vapply(fem_list_all, ncol, integer(1)),
                        stringsAsFactors = FALSE) %>%
  left_join(library_meta, by = "batch") %>%
  mutate(kept = n_cells >= params$min_library_cells) %>%
  dplyr::select(batch, library, stage, n_cells, kept) %>%
  arrange(desc(n_cells))

datatable(lib_counts, rownames = FALSE)

dropped = lib_counts %>% filter(!kept)
fem_list = fem_list_all[lib_counts$batch[lib_counts$kept]]
```

`r if (nrow(dropped) == 0) "Every library contributes enough cells to be analyzed." else paste0("Dropped for contributing fewer than ", params$min_library_cells, " cells: ", paste0(dropped$library, " (", dropped$n_cells, ")", collapse = "; "), ". At that size they cannot support a per-library analysis and would only add a colour to the plots.")`

```{r}
universe = Reduce(intersect, lapply(fem_list, rownames))
fem_list = lapply(fem_list, "[", i = universe, )

common_cols = Reduce(intersect, lapply(fem_list, function(x) colnames(colData(x))))
fem_list    = lapply(fem_list, function(x) {
  colData(x)     = colData(x)[, common_cols, drop = FALSE]
  reducedDims(x) = list()
  x
})

fem_list_norm = if (length(fem_list) > 1) do.call(multiBatchNorm, fem_list) else fem_list

fem.sce = do.call(cbind, fem_list_norm)
fem.sce$batch   = rep(names(fem_list), times = vapply(fem_list, ncol, integer(1)))
fem.sce$library = library_meta$library[match(fem.sce$batch, library_meta$batch)]
fem.sce$stage   = library_meta$stage[match(fem.sce$batch, library_meta$batch)]
```

`r ncol(fem.sce)` cells retained across `r length(fem_list)` libraries.
`multiBatchNorm` rescales for sequencing depth only and does not adjust
expression structure, so library differences remain visible and testable.

# Markers

```{r}
gene_description =
  read_excel(paste0(data_dir, "/41467_2024_55019_MOESM3_ESM.xlsx"), skip = 1) %>%
  dplyr::rename(WangGeneID = "CountMatrix_GeneID",
                Gene        = "NCBI_GTF_GENEID",
                description = "NCBI_GTF_ProteinInfo") %>%
  dplyr::select(Gene, WangGeneID, description) %>%
  filter(!is.na(Gene))

resolve_marker = function(symbol, rn, gd) {
  if (symbol %in% rn) return(symbol)
  g = gd$Gene[gd$WangGeneID == symbol]
  g = g[g %in% rn]
  if (length(g) > 0) return(g[1])
  NA_character_
}

# families are declared, so coherence can be tested within a known programme
zp_wang  = unique(gene_description$WangGeneID[grepl("^ZP[0-9]", gene_description$WangGeneID)])
aep_wang = unique(gene_description$WangGeneID[grepl("^aep1", gene_description$WangGeneID,
                                                    ignore.case = TRUE)])

marker_families = bind_rows(
  data.frame(symbol = zp_wang,  family = "ZP (oocyte)",        stringsAsFactors = FALSE),
  data.frame(symbol = aep_wang, family = "AEP1",               stringsAsFactors = FALSE),
  data.frame(symbol = c("DDX4", "PIWIL1", "SYCP3"),
             family = "germline, general",                     stringsAsFactors = FALSE),
  data.frame(symbol = c("ALPL", "PRDM14"),
             family = "early germline",                        stringsAsFactors = FALSE),
  data.frame(symbol = c("CUZD1", "S100A1"),
             family = "female-associated",                     stringsAsFactors = FALSE)) %>%
  distinct(symbol, .keep_all = TRUE)

rn = rownames(fem.sce)
marker_families = marker_families %>%
  rowwise() %>%
  mutate(gene_id = resolve_marker(symbol, rn, gene_description)) %>%
  ungroup()

resolved   = marker_families %>% filter(!is.na(gene_id)) %>%
  distinct(gene_id, .keep_all = TRUE)
unresolved = marker_families %>% filter(is.na(gene_id))
```

* Requested, after expanding the ZP and AEP families: `r nrow(marker_families)`
* Resolved in `rownames`: `r nrow(resolved)`
* Not found, and therefore excluded: `r if (nrow(unresolved) == 0) "none" else paste(unresolved$symbol, collapse = ", ")`

```{r}
datatable(resolved %>% dplyr::count(family, name = "n_resolved"), rownames = FALSE)
```

```{r}
# shared helpers, used by both the combined and per-library analyses
safe_rho = function(x, y) {
  if (length(x) < 10) return(NA_real_)
  if (length(unique(x)) < 2 || length(unique(y)) < 2) return(NA_real_)
  suppressWarnings(cor(x, y, method = "spearman"))
}
safe_p = function(x, y) {
  if (length(x) < 10) return(NA_real_)
  if (length(unique(x)) < 2 || length(unique(y)) < 2) return(NA_real_)
  suppressWarnings(cor.test(x, y, method = "spearman")$p.value)
}

embed_cells = function(sce, n_hvg = 2000, n_pc = 30, n_dc_from = 15, block = NULL) {
  dec  = modelGeneVar(sce, block = block)
  hvgs = getTopHVGs(dec, n = min(n_hvg, nrow(sce) - 1))
  set.seed(1)
  npc = min(n_pc, ncol(sce) - 1)
  tmp = runPCA(sce, subset_row = hvgs, ncomponents = npc)
  set.seed(1)
  dm  = destiny::DiffusionMap(reducedDim(tmp, "PCA")[, seq_len(min(n_dc_from, npc))])
  list(pca        = reducedDim(tmp, "PCA"),
       percentVar = attr(reducedDim(tmp, "PCA"), "percentVar"),
       dc         = destiny::eigenvectors(dm)[, 1:min(5, ncol(destiny::eigenvectors(dm)))],
       dm         = dm)
}

marker_expression = function(sce, resolved) {
  m = as.matrix(logcounts(sce)[resolved$gene_id, , drop = FALSE])
  rownames(m) = resolved$symbol
  t(m) %>% as.data.frame() %>%
    tibble::rownames_to_column("barcode") %>%
    tidyr::pivot_longer(-barcode, names_to = "symbol", values_to = "logexpr")
}

trend_table = function(expr_long, axis_df, axis_col) {
  expr_long %>%
    left_join(axis_df, by = "barcode") %>%
    group_by(symbol) %>%
    summarize(n_cells        = dplyr::n(),
              pct_expressing = 100 * mean(logexpr > 0),
              rho            = safe_rho(.data[[axis_col]], logexpr),
              p_value        = safe_p(.data[[axis_col]], logexpr),
              .groups = "drop") %>%
    mutate(FDR = p.adjust(p_value, "BH"),
           direction = case_when(
             is.na(rho)      ~ "undetermined",
             FDR >= 0.05     ~ "flat",
             rho >=  0.3     ~ "increases",
             rho <= -0.3     ~ "decreases",
             TRUE            ~ "weak")) %>%
    left_join(resolved %>% dplyr::select(symbol, family), by = "symbol") %>%
    arrange(desc(abs(rho)))
}
```

# 1. Combined analysis

```{r}
combined_file = paste0(Robject_out_dir, "/female_combined_embedding.Rds")

if (file.exists(combined_file)) {
  emb = readRDS(combined_file)
} else {
  message("Computing the combined embedding. Cached in ", combined_file, ".")
  emb = embed_cells(fem.sce,
                    block = if (length(fem_list) > 1) fem.sce$batch else NULL)
  saveRDS(emb, file = combined_file)
}

colnames(emb$dc) = paste0("DC", seq_len(ncol(emb$dc)))
reducedDim(fem.sce, "PCA")          = emb$pca
reducedDim(fem.sce, "DiffusionMap") = emb$dc

cd = as.data.frame(colData(fem.sce))
cd$barcode = colnames(fem.sce)
cd$DC1 = emb$dc[, "DC1"]; cd$DC2 = emb$dc[, "DC2"]
cd$PC1 = emb$pca[, 1];    cd$PC2 = emb$pca[, 2]
```

PC1 explains `r round(emb$percentVar[1], 1)`% of variance, PC2
`r round(emb$percentVar[2], 1)`%.

## Is DC1 bimodal?

A correlation computed across a bimodal distribution measures the gap between
the two modes, not a gradient within them. Bimodality is therefore tested first,
because it changes what every subsequent statistic means.

```{r}
set.seed(1)
km = kmeans(matrix(cd$DC1, ncol = 1), centers = 2, nstart = 25)
cd$mode = paste0("mode", km$cluster)

# relabel so mode1 is always the lower end of DC1
if (mean(cd$DC1[cd$mode == "mode1"]) > mean(cd$DC1[cd$mode == "mode2"])) {
  cd$mode = ifelse(cd$mode == "mode1", "mode2", "mode1")
}

between_ratio = km$betweenss / km$totss
mode_gap = min(cd$DC1[cd$mode == "mode2"]) - max(cd$DC1[cd$mode == "mode1"])
dc1_range = diff(range(cd$DC1))
gap_frac = mode_gap / dc1_range

bimodal = between_ratio >= 0.75
```

A two-means split of DC1 accounts for `r round(100 * between_ratio, 1)`% of its
total variance, and the gap between the modes spans
`r round(100 * gap_frac, 1)`% of the DC1 range.

`r if (bimodal) "**DC1 is bimodal.** The cells fall into two separated groups rather than lying along a gradient, so DC1 is read below as a group separation and not as a pseudotime." else "DC1 is not clearly bimodal, so it is read below as a continuous axis."`

## What separates the two modes

```{r}
tech_vars = intersect(c("sum", "detected", "subsets_Mito_percent"), colnames(cd))

mode_tab = cd %>%
  group_by(mode) %>%
  summarize(n_cells = dplyr::n(),
            across(all_of(tech_vars), median, .names = "median_{.col}"),
            .groups = "drop")

mode_lib = cd %>%
  dplyr::count(mode, library, name = "n") %>%
  group_by(mode) %>% mutate(pct = 100 * n / sum(n)) %>% ungroup()

mode_stage = cd %>%
  dplyr::count(mode, stage, name = "n") %>%
  group_by(mode) %>% mutate(pct = 100 * n / sum(n)) %>% ungroup()

datatable(mode_tab %>% mutate(across(where(is.numeric), ~round(., 1))),
          rownames = FALSE)
datatable(mode_lib %>% mutate(pct = round(pct, 1)), rownames = FALSE)
```

```{r}
stage_r2 = if (length(unique(cd$stage)) > 1)
  summary(lm(cd$DC1 ~ cd$stage))$r.squared else NA_real_
lib_r2 = if (length(unique(cd$library)) > 1)
  summary(lm(cd$DC1 ~ cd$library))$r.squared else NA_real_

mode_is_stage = !is.na(stage_r2) && stage_r2 >= 0.5
```

Stage explains `r round(100 * stage_r2, 1)`% of the variance in DC1, and library
of origin `r round(100 * lib_r2, 1)`%.

`r if (mode_is_stage) "**The two modes are the larval and transformer cells.** Because collection date varies with stage, this separation is confounded: it is equally consistent with a genuine difference between larval oogonia and transformer oocytes and with a July to November batch effect. The marker coherence test below is what distinguishes them." else "The modes do not correspond to stage, so whatever separates them is not the stage and batch contrast."`

`r if (!is.na(lib_r2) && !is.na(stage_r2) && lib_r2 - stage_r2 < 0.05) "Note that library adds essentially nothing beyond stage, meaning the two libraries within each stage sit together. Were this per-library technical noise rather than a stage-level difference, all four libraries would have separated." else ""`

## Technical covariates, within each mode

This is the diagnostic that the pooled correlation cannot give. If DC1 tracked
detection rate as a gradient, the association would persist inside each mode. If
it is only the two groups differing, the within-mode correlations collapse.

```{r}
within_tech = bind_rows(lapply(unique(cd$mode), function(m) {
  sub = cd[cd$mode == m, ]
  bind_rows(lapply(tech_vars, function(v) {
    data.frame(mode = m, covariate = v,
               rho_within = safe_rho(sub$DC1, sub[[v]]),
               stringsAsFactors = FALSE)
  }))
}))

pooled_tech = bind_rows(lapply(tech_vars, function(v) {
  data.frame(covariate = v, rho_pooled = safe_rho(cd$DC1, cd[[v]]),
             stringsAsFactors = FALSE)
}))

tech_compare = within_tech %>%
  tidyr::pivot_wider(id_cols = covariate, names_from = mode,
                     values_from = rho_within, names_prefix = "rho_") %>%
  left_join(pooled_tech, by = "covariate")

write.csv(tech_compare,
          file = paste0(results_dir, "/SuppTable_technical_within_mode.csv"),
          row.names = FALSE)

datatable(tech_compare %>% mutate(across(where(is.numeric), ~round(., 3))),
          rownames = FALSE)
```

```{r}
max_within = max(abs(within_tech$rho_within), na.rm = TRUE)
max_pooled = max(abs(pooled_tech$rho_pooled), na.rm = TRUE)
technical_gradient = isTRUE(max_within >= 0.4)
```

The strongest pooled correlation is `r round(max_pooled, 2)`; the strongest
within either mode is `r round(max_within, 2)`.

`r if (technical_gradient) "**A technical gradient is present within a mode**, so detection rate structures the data even after the group separation is accounted for. Treat any ordering with suspicion." else "**The pooled association is entirely the group difference.** Within each mode, DC1 has essentially no relationship to detection rate or composition. The two groups differ in RNA content, which is not the same thing as DC1 being a technical artifact."`

## Do any later components avoid the stage split?

DC1 is the stage separation, so the natural question is whether DC2 or a later
component carries within-stage structure instead. Every component is scanned
against the same tests.

Two caveats on reading this table. Later components are constrained to be
orthogonal to DC1, which is a mathematical requirement rather than a biological
one: a maturation gradient would partly align with the stage axis, since larval
oogonia are less mature than transformer oocytes, so forcing it perpendicular to
DC1 distorts the signal being looked for. And a diffusion component can be
produced entirely by a few weakly connected cells, which is why outlier
dominance is reported alongside. The per-library analysis in section 2 is the
cleaner form of this question, because within a library there is no stage axis
to be orthogonal to.

```{r}
expr_long_all = marker_expression(fem.sce, resolved)

outlier_share = function(v) {
  d = abs(v - median(v))
  k = max(1, ceiling(0.01 * length(v)))
  sum(sort(d, decreasing = TRUE)[seq_len(k)]^2) / sum(d^2)
}

dc_scan = bind_rows(lapply(colnames(emb$dc), function(dcn) {
  v  = emb$dc[, dcn]
  ax = data.frame(barcode = cd$barcode, axis = v, stringsAsFactors = FALSE)

  tt = expr_long_all %>%
    left_join(ax, by = "barcode") %>%
    group_by(symbol) %>%
    summarize(rho = safe_rho(axis, logexpr),
              p   = safe_p(axis, logexpr), .groups = "drop") %>%
    mutate(FDR = p.adjust(p, "BH"))

  data.frame(
    component     = dcn,
    eigenvalue    = as.numeric(destiny::eigenvalues(emb$dm)[match(dcn, colnames(emb$dc))]),
    stage_r2      = if (length(unique(cd$stage)) > 1)
                      summary(lm(v ~ cd$stage))$r.squared else NA_real_,
    library_r2    = if (length(unique(cd$library)) > 1)
                      summary(lm(v ~ cd$library))$r.squared else NA_real_,
    max_tech_rho  = suppressWarnings(max(abs(vapply(tech_vars,
                      function(t) safe_rho(v, cd[[t]]), numeric(1))), na.rm = TRUE)),
    n_directional = sum(tt$FDR < 0.05 & abs(tt$rho) >= 0.3, na.rm = TRUE),
    outlier_share = outlier_share(v),
    stringsAsFactors = FALSE)
})) %>%
  mutate(batch_free    = !is.na(stage_r2) & stage_r2 < 0.25,
         outlier_driven = outlier_share >= 0.5,
         candidate     = batch_free & !outlier_driven &
                         max_tech_rho < 0.4 & n_directional >= 3)

write.csv(dc_scan, file = paste0(results_dir, "/SuppTable_dc_scan.csv"),
          row.names = FALSE)

datatable(dc_scan %>% mutate(across(where(is.numeric), ~signif(., 3))),
          rownames = FALSE)
```

`outlier_share` is the fraction of a component's total squared deviation
contributed by its most extreme 1% of cells. A value near 1 means the component
describes a few cells rather than the population.

```{r}
candidates = dc_scan %>% filter(candidate)
later_batch_free = dc_scan %>% filter(component != "DC1", batch_free)
```

`r if (nrow(candidates) > 0) paste0("**", paste(candidates$component, collapse = ", "), "** avoid the stage split, are not outlier driven, are not associated with detection rate, and carry monotonic marker changes. Worth examining as a candidate within-stage axis, subject to the orthogonality caveat above.") else if (nrow(later_batch_free) > 0) paste0(paste(later_batch_free$component, collapse = ", "), " avoid the stage split, but none of them pass the remaining checks: they are either driven by a small number of outlier cells, associated with detection rate, or carry no monotonic marker changes. No later component provides a usable within-stage axis.") else "No component other than DC1 avoids the stage split, so the combined embedding offers no batch-free axis at all. Any within-stage structure has to be looked for inside a single library, which is section 2."`

```{r, fig.width=11, fig.height=3.5}
dc_long = as.data.frame(emb$dc) %>%
  mutate(barcode = cd$barcode, library = cd$library) %>%
  tidyr::pivot_longer(starts_with("DC"), names_to = "component", values_to = "value")

p_dcs = ggplot(dc_long, aes(x = value, fill = library, color = library)) +
  geom_density(alpha = 0.3) +
  scale_fill_manual(values = lib_cols) +
  scale_color_manual(values = lib_cols) +
  facet_wrap(~ component, scales = "free", nrow = 1) +
  xlab(NULL) + ylab("density")

p_dcs
ggsave(p_dcs, file = paste0(results_dir, "/all_DCs_by_library.png"),
       width = 11, height = 3.5)
```

## Marker expression across the modes

Coherence within a declared gene family narrows the possibilities but does not
by itself separate a differentiation programme from a technical difference. A
batch effect can move a family as a set, by at least three routes: ambient RNA
from an abundant cell type contaminating every droplet of one library,
paralogous genes sharing length and GC so that a capture bias hits them
together, and dissociation stress inducing the immediate-early and heat shock
families coordinately.

Ambient RNA is the serious candidate here. Oocytes are large, fragile and
RNA-rich, zona pellucida transcripts are among the most abundant things in them,
and the library showing the signal is the female transformer. The control for it
is in the next section, and it is what the interpretation actually rests on.

```{r}
expr_long = marker_expression(fem.sce, resolved)

axis_df = cd %>% dplyr::select(barcode, DC1, mode, library, stage)

mode_markers = expr_long %>%
  left_join(axis_df, by = "barcode") %>%
  group_by(symbol, mode) %>%
  summarize(mean_logexpr = mean(logexpr),
            pct_expressing = 100 * mean(logexpr > 0), .groups = "drop") %>%
  tidyr::pivot_wider(id_cols = symbol,
                     names_from = mode,
                     values_from = c(mean_logexpr, pct_expressing)) %>%
  left_join(resolved %>% dplyr::select(symbol, family), by = "symbol") %>%
  mutate(log2fc_mode2_over_mode1 = mean_logexpr_mode2 - mean_logexpr_mode1)

combined_trends = trend_table(expr_long, axis_df, "DC1")

write.csv(combined_trends,
          file = paste0(results_dir, "/SuppTable_marker_trends_combined.csv"),
          row.names = FALSE)

datatable(combined_trends %>% mutate(across(where(is.numeric), ~signif(., 3))),
          filter = "top", rownames = FALSE)
```

A family with fewer than three resolved genes cannot meet the criteria no
matter what the data show, so those are reported as `NA`, meaning not testable,
rather than `FALSE`, which would read as a negative result.

```{r}
min_family_genes = 3

family_coherence = combined_trends %>%
  filter(!is.na(rho)) %>%
  group_by(family) %>%
  summarize(n_genes        = dplyr::n(),
            median_rho     = median(rho),
            n_same_sign    = max(sum(rho > 0), sum(rho < 0)),
            pct_same_sign  = 100 * max(sum(rho > 0), sum(rho < 0)) / dplyr::n(),
            n_significant  = sum(FDR < 0.05 & abs(rho) >= 0.3),
            .groups = "drop") %>%
  mutate(testable = n_genes >= min_family_genes,
         coherent = ifelse(testable,
                           pct_same_sign >= 80 & n_significant >= 3,
                           NA)) %>%
  arrange(desc(n_genes))

write.csv(family_coherence,
          file = paste0(results_dir, "/SuppTable_family_coherence.csv"),
          row.names = FALSE)

datatable(family_coherence %>% mutate(across(where(is.numeric), ~round(., 2))),
          rownames = FALSE)
```

```{r}
zp_row = family_coherence %>% filter(family == "ZP (oocyte)")
zp_coherent = nrow(zp_row) > 0 && isTRUE(zp_row$coherent)
n_testable  = sum(family_coherence$testable)
n_coherent_families = sum(family_coherence$coherent, na.rm = TRUE)
n_not_testable = sum(!family_coherence$testable)
programme_coherent = n_coherent_families >= 1

# families that miss only on the sign threshold, reported rather than lost
near_miss = family_coherence %>%
  filter(testable, !coherent, n_significant >= 3, pct_same_sign >= 65)
```

`r n_coherent_families` of `r n_testable` testable families move coherently
along DC1, defined as at least 80% of genes agreeing in sign and at least three
reaching |rho| >= 0.3 with FDR < 0.05.
`r if (n_not_testable > 0) paste0(n_not_testable, " famil", if (n_not_testable == 1) "y has" else "ies have", " fewer than ", min_family_genes, " resolved genes and could not be tested either way.") else ""`

`r if (zp_coherent) paste0("The ZP family is coherent: ", zp_row$n_genes, " genes, median rho ", round(zp_row$median_rho, 2), ", ", round(zp_row$pct_same_sign), "% agreeing in sign, ", zp_row$n_significant, " significant.") else "The ZP family does not move coherently along DC1."`

`r if (nrow(near_miss) > 0) paste0("Reported rather than discarded: ", paste0(near_miss$family, " (", near_miss$n_genes, " genes, median rho ", round(near_miss$median_rho, 2), ", ", round(near_miss$pct_same_sign), "% same sign, ", near_miss$n_significant, " significant)", collapse = "; "), ". ", if (nrow(near_miss) == 1) "This family falls" else "These families fall", " below the 80% sign threshold while still showing a strong majority direction and many significant genes. For a large paralogous family there is no reason to expect every member to share the same regulation, so a minority moving the other way is plausibly biological rather than noise. The threshold is a convention, not a finding.") else ""`

## Direction test: do early and late programmes move oppositely?

Coherence within one family is weaker evidence than two families moving in
**opposite** directions. Ambient RNA, capture bias and sequencing depth all push
a set of genes the same way. None of them raise an oocyte programme while
lowering a general germline programme, in the order that matches known
development.

```{r}
late_families  = c("ZP (oocyte)", "AEP1")
early_families = c("germline, general", "early germline")

dir_summary = combined_trends %>%
  filter(!is.na(rho)) %>%
  mutate(programme = case_when(family %in% late_families  ~ "late (oocyte)",
                               family %in% early_families ~ "early (germline)",
                               TRUE                       ~ "other")) %>%
  filter(programme != "other") %>%
  group_by(programme) %>%
  summarize(n_genes       = dplyr::n(),
            median_rho    = median(rho),
            n_significant = sum(FDR < 0.05 & abs(rho) >= 0.3),
            .groups = "drop")

write.csv(dir_summary,
          file = paste0(results_dir, "/SuppTable_direction_test.csv"),
          row.names = FALSE)

datatable(dir_summary %>% mutate(across(where(is.numeric), ~round(., 3))),
          rownames = FALSE)
```

```{r}
late_row  = dir_summary %>% filter(programme == "late (oocyte)")
early_row = dir_summary %>% filter(programme == "early (germline)")

both_present = nrow(late_row) == 1 && nrow(early_row) == 1

opposite_directions = both_present &&
  sign(late_row$median_rho) != sign(early_row$median_rho) &&
  late_row$n_significant  >= 3 &&
  early_row$n_significant >= 1

# a Wilcoxon on the two sets of per-gene correlations, to test the split
# rather than eyeballing two medians
dir_p = if (both_present) {
  a = combined_trends %>% filter(family %in% late_families,  !is.na(rho)) %>% pull(rho)
  b = combined_trends %>% filter(family %in% early_families, !is.na(rho)) %>% pull(rho)
  suppressWarnings(wilcox.test(a, b)$p.value)
} else NA_real_
```

`r if (!both_present) "One of the two programmes has no resolved genes, so the direction test cannot be run." else paste0("The late oocyte programme has a median rho of ", round(late_row$median_rho, 2), " across ", late_row$n_genes, " genes (", late_row$n_significant, " significant), and the early germline programme a median rho of ", round(early_row$median_rho, 2), " across ", early_row$n_genes, " genes (", early_row$n_significant, " significant). Wilcoxon on the two sets of per-gene correlations: p = ", signif(dir_p, 2), ".")`

`r if (opposite_directions) "**The two programmes move in opposite directions along DC1.** Oocyte genes are high at one end and general germline identity genes at the other. A technical difference between libraries acts on a set of genes in one direction; it does not raise one programme while lowering another in the sequence that matches oogonia giving way to oocytes. This is stronger evidence for a differentiation transition than coherence within either family alone." else if (both_present) "The two programmes do not move in opposite directions along DC1, so the separation does not show the early to late transition expected of a differentiation transition." else ""`

```{r, fig.width=11, fig.height=8}
top_markers = combined_trends %>% filter(!is.na(rho)) %>%
  slice_max(abs(rho), n = 12) %>% pull(symbol)

p_mark = ggplot(expr_long %>% filter(symbol %in% top_markers) %>%
                  left_join(axis_df, by = "barcode"),
                aes(x = DC1, y = logexpr)) +
  geom_point(aes(color = library), size = 0.3, alpha = 0.35) +
  geom_smooth(method = "loess", se = FALSE, color = "black", linewidth = 0.7) +
  scale_color_manual(values = lib_cols) +
  guides(color = guide_legend(override.aes = list(size = 3, alpha = 1))) +
  facet_wrap(~ symbol, scales = "free_y") +
  ylab("Uncorrected log-expression")

p_mark
ggsave(p_mark, file = paste0(results_dir, "/markers_vs_DC1_combined.png"),
       width = 11, height = 8)
```

```{r, fig.width=10, fig.height=4}
p_dc = ggplot(cd, aes(x = DC1, y = DC2, color = library)) +
  geom_point(size = 0.6, alpha = 0.8) +
  scale_color_manual(values = lib_cols) +
  guides(color = guide_legend(override.aes = list(size = 3))) +
  ggtitle("Diffusion map, by library")

p_dens = ggplot(cd, aes(x = DC1, fill = library, color = library)) +
  geom_density(alpha = 0.35) +
  scale_fill_manual(values = lib_cols) +
  scale_color_manual(values = lib_cols) +
  ggtitle("DC1 by library")

p_dc + p_dens
ggsave(p_dc + p_dens, file = paste0(results_dir, "/DC1_by_library.png"),
       width = 10, height = 4)
```

## Ambient RNA control

Ambient contamination is uniform across all droplets from a library, so if the
zona pellucida signal were free-floating oocyte RNA in the female transformer
library it would appear in every cell captured from that library, including the
fibroblasts, immune cells and endothelium that cannot transcribe it. Real
transcription is confined to the cells that make it.

The test therefore compares ZP expression in **non-germ** cells of the female
library against non-germ cells of the other libraries. Elevated there means
ambient. Flat there, with a high signal confined to the germ cluster, means
transcription.

```{r}
stopifnot(identical(colnames(combined.all.sce), colnames(merged.all.sce)))

zp_ids = resolved$gene_id[resolved$family == "ZP (oocyte)"]

all_meta = data.frame(
  barcode = colnames(combined.all.sce),
  cluster = as.character(colData(merged.all.sce)$label),
  batch   = as.character(colData(merged.all.sce)$batch),
  stringsAsFactors = FALSE) %>%
  left_join(library_meta, by = "batch") %>%
  mutate(compartment = ifelse(cluster == as.character(params$female_all_cluster),
                              "germ (cluster of interest)", "other clusters"))

all_meta$zp_mean = colMeans(as.matrix(
  logcounts(combined.all.sce)[zp_ids, , drop = FALSE]))

ambient_tab = all_meta %>%
  group_by(library, compartment) %>%
  summarize(n_cells        = dplyr::n(),
            mean_ZP        = mean(zp_mean),
            pct_expressing = 100 * mean(zp_mean > 0),
            .groups = "drop") %>%
  arrange(compartment, desc(mean_ZP))

write.csv(ambient_tab,
          file = paste0(results_dir, "/SuppTable_ambient_control.csv"),
          row.names = FALSE)

datatable(ambient_tab %>% mutate(across(where(is.numeric), ~round(., 3))),
          rownames = FALSE)
```

```{r}
other_zp = ambient_tab %>% filter(compartment == "other clusters")
fem_other = other_zp$mean_ZP[other_zp$library == "Female"]
rest_other = other_zp$mean_ZP[other_zp$library != "Female"]

germ_zp = ambient_tab %>% filter(compartment == "germ (cluster of interest)")
fem_germ = germ_zp$mean_ZP[germ_zp$library == "Female"]

ambient_ratio = if (length(fem_other) == 1 && length(rest_other) > 0 &&
                    max(rest_other) > 0) fem_other / max(rest_other) else NA_real_
germ_vs_other = if (length(fem_germ) == 1 && length(fem_other) == 1 && fem_other > 0)
  fem_germ / fem_other else if (length(fem_germ) == 1) Inf else NA_real_

ambient_suspected = isTRUE(ambient_ratio >= 3)
```

Mean ZP expression in non-germ cells of the female library is
`r round(fem_other, 3)`, against a maximum of `r round(max(rest_other), 3)` in
the non-germ cells of any other library. In the germ cluster of the female
library it is `r round(fem_germ, 3)`.

`r if (ambient_suspected) "**Ambient contamination is indicated.** Cells in the female library that cannot transcribe zona pellucida nevertheless carry it, so the germ cell signal is inflated by free-floating oocyte RNA and the size of the difference between the modes cannot be taken at face value." else "**Ambient contamination is ruled out.** Cells in the female library that cannot transcribe zona pellucida do not carry it, at levels comparable to the same compartment in every other library, while the germ cluster of that same library carries it strongly. Free-floating oocyte RNA would have appeared in every droplet from the library, so the signal is transcription confined to the germ cells."`

```{r, fig.width=8, fig.height=4}
p_amb = ggplot(all_meta, aes(x = library, y = zp_mean, fill = compartment)) +
  geom_boxplot(outlier.size = 0.3) +
  scale_fill_manual(values = c("germ (cluster of interest)" = "#FF6DB6",
                               "other clusters" = "grey70")) +
  theme(axis.text.x = element_text(angle = 30, hjust = 1)) +
  ylab("Mean ZP family log-expression") + xlab(NULL)

p_amb
ggsave(p_amb, file = paste0(results_dir, "/ambient_control_ZP.png"),
       width = 8, height = 4)
```

## Verdict for the combined analysis

```{r}
combined_verdict = if (technical_gradient) {
  "technical"
} else if (bimodal && ambient_suspected) {
  "two groups, ambient confounded"
} else if (bimodal && programme_coherent) {
  "two states"
} else if (bimodal) {
  "two groups, unexplained"
} else if (!bimodal && sum(combined_trends$direction %in% c("increases", "decreases")) >= 3) {
  "continuum"
} else {
  "no structure"
}
```

**`r combined_verdict`.**

`r switch(combined_verdict,
  "technical" = "DC1 retains a strong association with detection rate inside each mode, so it is structured by technical variation and nothing should be built on it.",
  "two groups, unexplained" = "The cells form two separated groups, but no declared marker family moves coherently across them, so the separation cannot be attributed to a differentiation programme and may be technical or batch driven.",
  "two groups, ambient confounded" = "The cells form two separated groups and marker families move coherently, but non-germ cells of the female library also carry zona pellucida transcripts, which they cannot transcribe. The signal is therefore inflated by ambient oocyte RNA and the separation cannot be read as a differentiation difference on this evidence.",
  "two states" = paste0("The female germ cells form two separated groups, not a gradient, and the separation carries a coherent differentiation programme that is not explained by ambient RNA: zona pellucida expression is confined to the germ cluster and is absent from the other cells of the same library, so it is transcribed rather than free-floating. ", if (opposite_directions) "The oocyte and general germline programmes also move in opposite directions along the axis, which no technical difference between libraries would produce. " else "", "Louvain at k = 8 not splitting these cells is a finding about the clustering rather than evidence of a continuum, and is expected given the subclustering in 02 runs on fastMNN corrected values where this between-library difference is removed by design. Stage and collection date remain confounded in general, but the specific alternative that the marker signal is a library artifact has been tested and rejected."),
  "continuum" = "DC1 is unimodal, is not explained by detection rate, and maturation markers vary monotonically along it, which supports a continuous gradient.",
  "no structure" = "DC1 is neither bimodal nor accompanied by monotonic marker changes, so there is no evidence of either discrete states or a gradient.")`

# 2. Per-library analysis

Within a single library, animal, collection date and batch are constant, so a
gradient there cannot be a batch effect and a correlation with detection rate is
unambiguously technical. This is the clean test for a maturation continuum.

```{r}
perlib_file = paste0(Robject_out_dir, "/female_perlibrary.Rds")

if (file.exists(perlib_file)) {
  perlib = readRDS(perlib_file)
} else {
  message("Computing per-library embeddings. Cached in ", perlib_file, ".")
  perlib = lapply(names(fem_list), function(b) {
    sce_b = fem_list[[b]]
    if (ncol(sce_b) < 100) return(NULL)
    e = embed_cells(sce_b)
    list(batch = b, emb = e, barcodes = colnames(sce_b))
  })
  names(perlib) = names(fem_list)
  perlib = perlib[!vapply(perlib, is.null, logical(1))]
  saveRDS(perlib, file = perlib_file)
}
```

```{r}
perlib_summary = bind_rows(lapply(names(perlib), function(b) {
  e     = perlib[[b]]$emb
  bc    = perlib[[b]]$barcodes
  sce_b = fem_list[[b]]

  cdb = as.data.frame(colData(sce_b))
  cdb$barcode = bc
  cdb$DC1 = e$dc[, 1]

  tv = intersect(c("sum", "detected", "subsets_Mito_percent"), colnames(cdb))
  tech_max = max(abs(vapply(tv, function(v) safe_rho(cdb$DC1, cdb[[v]]),
                            numeric(1))), na.rm = TRUE)

  el = marker_expression(sce_b, resolved)
  tt = trend_table(el, cdb %>% dplyr::select(barcode, DC1), "DC1")

  fam = tt %>% filter(!is.na(rho)) %>% group_by(family) %>%
    summarize(n_genes = dplyr::n(),
              pct_same_sign = 100 * max(sum(rho > 0), sum(rho < 0)) / dplyr::n(),
              n_significant = sum(FDR < 0.05 & abs(rho) >= 0.3), .groups = "drop") %>%
    mutate(coherent = n_genes >= 3 & pct_same_sign >= 80 & n_significant >= 3)

  data.frame(
    batch            = b,
    library          = library_meta$library[match(b, library_meta$batch)],
    n_cells          = ncol(sce_b),
    percentVar_PC1   = e$percentVar[1],
    max_tech_rho     = tech_max,
    n_directional    = sum(tt$direction %in% c("increases", "decreases")),
    n_markers_tested = nrow(tt),
    n_coherent_fam   = sum(fam$coherent, na.rm = TRUE),
    stringsAsFactors = FALSE)
})) %>%
  mutate(technical_confound = max_tech_rho >= 0.4,
         continuum_supported = !technical_confound & n_directional >= 3)

write.csv(perlib_summary,
          file = paste0(results_dir, "/SuppTable_perlibrary_summary.csv"),
          row.names = FALSE)

datatable(perlib_summary %>% mutate(across(where(is.numeric), ~round(., 3))),
          rownames = FALSE)
```

```{r, results="asis"}
for (i in seq_len(nrow(perlib_summary))) {
  r = perlib_summary[i, ]
  cat("**", r$library, "** (", r$n_cells, " cells): PC1 explains ",
      round(r$percentVar_PC1, 1), "% of variance; strongest DC1 to technical ",
      "covariate correlation ", round(r$max_tech_rho, 2), "; ", r$n_directional,
      " of ", r$n_markers_tested, " markers monotonic; ", r$n_coherent_fam,
      " coherent marker famil", if (r$n_coherent_fam == 1) "y" else "ies", ". ",
      if (r$technical_confound) {
        "Within this single library, where batch cannot contribute, the leading component still tracks detection rate, so it is technical and no ordering should be built on it."
      } else if (r$continuum_supported) {
        "Within this single library the leading component is not technical and markers vary monotonically along it, which supports a maturation gradient inside this stage."
      } else {
        "The leading component is not technical, but markers do not vary monotonically along it, so there is no evidence of a gradient within this library."
      },
      "\n\n", sep = "")
}
```

```{r, fig.width=11, fig.height=4}
perlib_dc = bind_rows(lapply(names(perlib), function(b) {
  data.frame(barcode = perlib[[b]]$barcodes,
             DC1     = perlib[[b]]$emb$dc[, 1],
             DC2     = perlib[[b]]$emb$dc[, 2],
             library = library_meta$library[match(b, library_meta$batch)],
             stringsAsFactors = FALSE)
}))

p_perlib = ggplot(perlib_dc, aes(x = DC1, y = DC2, color = library)) +
  geom_point(size = 0.5, alpha = 0.8) +
  scale_color_manual(values = lib_cols) +
  facet_wrap(~ library, scales = "free", nrow = 1) +
  guides(color = "none") +
  ggtitle("Diffusion map computed within each library separately")

p_perlib
ggsave(p_perlib, file = paste0(results_dir, "/perlibrary_diffusion.png"),
       width = 11, height = 4)
```

## Marker expression in the Female library alone

This is the cleanest test in the document. Inside the female transformer
library, animal, collection date, batch and sex are all constant, so nothing
here can be a batch effect and a correlation with detection rate is
unambiguously technical. If oogonia and oocytes coexist in this sample, the
oocyte and general germline programmes should move in opposite directions along
its internal DC1, exactly as they do across the stage split, but with the
confound removed.

```{r}
female_batch     = library_meta$batch[library_meta$library == "Female"]
female_available = length(female_batch) == 1 && female_batch %in% names(perlib)
```

`r if (!female_available) "The Female library is not available for this analysis, either because it contributes too few cells or because it was dropped above. The section below is skipped." else ""`

```{r}
if (female_available) {
  sce_f = fem_list[[female_batch]]
  e_f   = perlib[[female_batch]]$emb

  cdf = as.data.frame(colData(sce_f))
  cdf$barcode = perlib[[female_batch]]$barcodes
  cdf$DC1 = e_f$dc[, 1]
  cdf$DC2 = e_f$dc[, 2]

  tv_f = intersect(c("sum", "detected", "subsets_Mito_percent"), colnames(cdf))
  tech_f = data.frame(
    covariate = tv_f,
    rho_DC1   = vapply(tv_f, function(v) safe_rho(cdf$DC1, cdf[[v]]), numeric(1)),
    stringsAsFactors = FALSE)

  datatable(tech_f %>% mutate(across(where(is.numeric), ~round(., 3))),
            rownames = FALSE)
}
```

```{r}
if (female_available) {
  fem_tech_max = max(abs(tech_f$rho_DC1), na.rm = TRUE)
  fem_technical = isTRUE(fem_tech_max >= 0.4)
} else {
  fem_tech_max = NA_real_; fem_technical = FALSE
}
```

`r if (female_available) paste0("Strongest correlation between the internal DC1 and a technical covariate: ", round(fem_tech_max, 2), ". ", if (fem_technical) "**That is a technical gradient.** With batch held constant there is nothing else it could be, so the internal axis is structured by detection rate and no biological reading is available." else "No technical gradient, so the internal axis is not detection rate.") else ""`

### Marker trends within the Female library

```{r}
if (female_available) {
  expr_f   = marker_expression(sce_f, resolved)
  trends_f = trend_table(expr_f, cdf %>% dplyr::select(barcode, DC1), "DC1")

  write.csv(trends_f,
            file = paste0(results_dir, "/SuppTable_marker_trends_female.csv"),
            row.names = FALSE)

  datatable(trends_f %>% mutate(across(where(is.numeric), ~signif(., 3))),
            filter = "top", rownames = FALSE)
}
```

```{r}
if (female_available) {
  fam_f = trends_f %>%
    filter(!is.na(rho)) %>%
    group_by(family) %>%
    summarize(n_genes       = dplyr::n(),
              median_rho    = median(rho),
              pct_same_sign = 100 * max(sum(rho > 0), sum(rho < 0)) / dplyr::n(),
              n_significant = sum(FDR < 0.05 & abs(rho) >= 0.3),
              .groups = "drop") %>%
    mutate(testable = n_genes >= min_family_genes,
           coherent = ifelse(testable,
                             pct_same_sign >= 80 & n_significant >= 3, NA)) %>%
    arrange(desc(n_genes))

  write.csv(fam_f,
            file = paste0(results_dir, "/SuppTable_family_coherence_female.csv"),
            row.names = FALSE)

  datatable(fam_f %>% mutate(across(where(is.numeric), ~round(., 2))),
            rownames = FALSE)
}
```

### Direction test within the Female library

```{r}
if (female_available) {
  dir_f = trends_f %>%
    filter(!is.na(rho)) %>%
    mutate(programme = case_when(family %in% late_families  ~ "late (oocyte)",
                                 family %in% early_families ~ "early (germline)",
                                 TRUE                       ~ "other")) %>%
    filter(programme != "other") %>%
    group_by(programme) %>%
    summarize(n_genes       = dplyr::n(),
              median_rho    = median(rho),
              n_significant = sum(FDR < 0.05 & abs(rho) >= 0.3),
              .groups = "drop")

  datatable(dir_f %>% mutate(across(where(is.numeric), ~round(., 3))),
            rownames = FALSE)

  late_f  = dir_f %>% filter(programme == "late (oocyte)")
  early_f = dir_f %>% filter(programme == "early (germline)")

  both_f = nrow(late_f) == 1 && nrow(early_f) == 1
  fem_opposite = both_f &&
    sign(late_f$median_rho) != sign(early_f$median_rho) &&
    late_f$n_significant >= 3 && early_f$n_significant >= 1

  dir_p_f = if (both_f) {
    a = trends_f %>% filter(family %in% late_families,  !is.na(rho)) %>% pull(rho)
    b = trends_f %>% filter(family %in% early_families, !is.na(rho)) %>% pull(rho)
    suppressWarnings(wilcox.test(a, b)$p.value)
  } else NA_real_
} else {
  both_f = FALSE; fem_opposite = FALSE; dir_p_f = NA_real_
}
```

`r if (!female_available) "" else if (!both_f) "One programme has no resolved genes within this library, so the direction test cannot be run here." else paste0("Within the Female library alone, the late oocyte programme has a median rho of ", round(late_f$median_rho, 2), " across ", late_f$n_genes, " genes (", late_f$n_significant, " significant) and the early germline programme ", round(early_f$median_rho, 2), " across ", early_f$n_genes, " genes (", early_f$n_significant, " significant). Wilcoxon p = ", signif(dir_p_f, 2), ".")`

`r if (female_available && fem_opposite) "**The two programmes move in opposite directions within a single library.** Batch, animal, collection date and sex are all constant here, so this cannot be a library artifact. Oogonia and oocytes coexist in this sample and are ordered along the internal axis, which is direct evidence of a maturation gradient rather than only a difference between stages." else if (female_available && both_f) "**The two programmes do not move in opposite directions within this library.** Whatever the internal axis describes, it is not the early to late germ cell transition. The opposite-direction result in the combined analysis therefore reflects the difference between the stages rather than a gradient inside the transformer sample." else ""`

```{r, fig.width=11, fig.height=8}
if (female_available) {
  top_f = trends_f %>% filter(!is.na(rho)) %>%
    slice_max(abs(rho), n = 12) %>% pull(symbol)

  p_fem = ggplot(expr_f %>% filter(symbol %in% top_f) %>%
                   left_join(cdf %>% dplyr::select(barcode, DC1), by = "barcode"),
                 aes(x = DC1, y = logexpr)) +
    geom_point(size = 0.4, alpha = 0.3, color = "#FF6DB6") +
    geom_smooth(method = "loess", se = FALSE, color = "black", linewidth = 0.7) +
    facet_wrap(~ symbol, scales = "free_y") +
    xlab("DC1, computed within the Female library only") +
    ylab("Uncorrected log-expression")

  print(p_fem)
  ggsave(p_fem, file = paste0(results_dir, "/markers_vs_DC1_female.png"),
         width = 11, height = 8)
}
```

```{r, fig.width=10, fig.height=4}
if (female_available) {
  zp_f  = resolved$gene_id[resolved$family == "ZP (oocyte)"]
  erly  = resolved$gene_id[resolved$family == "germline, general"]

  cdf$ZP_mean       = colMeans(as.matrix(logcounts(sce_f)[zp_f, , drop = FALSE]))
  cdf$germline_mean = colMeans(as.matrix(logcounts(sce_f)[erly, , drop = FALSE]))

  p_zp = ggplot(cdf, aes(x = DC1, y = DC2, color = ZP_mean)) +
    geom_point(size = 0.8) +
    scale_color_viridis_c() +
    ggtitle("ZP family mean")

  p_gl = ggplot(cdf, aes(x = DC1, y = DC2, color = germline_mean)) +
    geom_point(size = 0.8) +
    scale_color_viridis_c(option = "magma") +
    ggtitle("General germline mean")

  print(p_zp + p_gl)
  ggsave(p_zp + p_gl,
         file = paste0(results_dir, "/female_programmes_on_DC.png"),
         width = 10, height = 4)
}
```

# 3. Summary

```{r, results="asis"}
any_within_continuum = any(perlib_summary$continuum_supported)

cat(paste0(
  ncol(fem.sce), " female germ cells across ", length(fem_list), " libraries. ",
  "**Combined:** DC1 is ", if (bimodal) "bimodal" else "unimodal",
  ", with stage explaining ", round(100 * stage_r2, 1),
  "% of its variance. The strongest correlation with a technical covariate is ",
  round(max_pooled, 2), " pooled but only ", round(max_within, 2),
  " within a mode, so the pooled association is the group difference rather ",
  "than a gradient. ", n_coherent_families, " of ", n_testable,
  " testable marker famil",
  if (n_coherent_families == 1) "ies moves" else "ies move",
  " coherently across the separation",
  if (opposite_directions)
    ", and the oocyte and general germline programmes move in opposite directions along it, which a technical difference between libraries would not produce"
  else "",
  ". Verdict: ", combined_verdict, ". ",
  "**Per library:** ",
  if (any_within_continuum) {
    paste0(sum(perlib_summary$continuum_supported), " of ", nrow(perlib_summary),
           " libraries support a maturation gradient internally, which cannot ",
           "be a batch effect because batch is constant within a library.")
  } else {
    paste0("no library supports a maturation gradient internally. Within a ",
           "single collection, where batch cannot contribute, these cells do ",
           "not resolve into an ordered progression, so the structure in the ",
           "combined analysis is between the stages rather than within them.")
  },
  if (!female_available) "" else if (fem_opposite) {
    paste0(" In the Female library alone, the oocyte and general germline ",
           "programmes move in opposite directions along the internal axis, ",
           "so oogonia and oocytes coexist in that sample and are ordered ",
           "within it. That result carries no confound at all, since batch, ",
           "animal, collection date and sex are constant.")
  } else {
    paste0(" In the Female library alone the two programmes do not separate, ",
           "so the early to late transition seen in the combined analysis is ",
           "between the stages rather than within the transformer sample.")
  }))
```

# Session info

```{r}
sessionInfo()
```
