---
title: "08_female_germ_cell_continuum"
author: "Stephanie Hickey"
date: "`r Sys.Date()`"
output:
  html_document:
    code_folding: hide
    toc: true
    toc_float:
      toc_collapsed: true
params:
    project_dir: /mnt/ufs18/rs-013/bioinformaticsCore/projects/chong_davidson/BCC125_sea_lamprey_scRNA
    project_name: BCC125_sea_lamprey_scRNA
    lab: Weiming Li
    results_date: "20260813"
    female_cluster: "5"
    min_library_cells: 50
    sub_k_range: "5,10,15,20,25,30,40,50"
    sub_k: 0
    sub_n_iterations: 10
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

```{r, echo=FALSE, results="asis"}
cat("\n")
cat("#### Lab:", params$lab, " \n")
cat("#### Project:", params$project_name, " \n")
```

This document analyses one cluster from the all-cells labelling, set by the
`female_cluster` parameter. The cluster contains cells from both stages, and its
larval cells come from the two libraries this project calls Undifferentiated.
**The annotation describes the cluster, not each cell in it.** Nothing below
establishes that the larval cells are female, and a trajectory through the
cluster cannot assume it.

The document asks two questions with different requirements.

**Is there a continuum?** This needs no clusters. The diffusion component
analyses treat the cells as one set and do not subcluster or tune a resolution.

**Can a trajectory be fitted?** A minimum spanning tree is a graph over nodes,
and a single cluster is a single node, so this needs subclusters. The
subclusters supply nodes and are not evidence of discrete subpopulations. `k` is
chosen by the subsampling stability sweep used in
`13_atlas_and_stage_correspondence.Rmd`, and what the subclusters correspond to
is reported before anything is built on them.

That correspondence matters because in earlier runs the leading diffusion
component in these cells separated larvae from transformers. If the subclusters
recover that split, a trajectory over them orders cells along the stage and
collection boundary.

The continuum question is asked twice.

**Combined.** All cells in the cluster together. The leading diffusion component
is tested for bimodality, technical association and stage. Marker coherence, an
ambient RNA control and a direction test decide how any separation is reported.

**Per library.** The same tests inside each library separately. Animal,
collection date and batch are constant within a library, so a gradient there
cannot be a batch effect, and a correlation with detection rate is technical.

A pseudotime can be computed from any set of cells, so an ordering alone shows
nothing. The technical and marker checks decide.

Every test uses uncorrected, normalized log-expression, because `fastMNN`
correction would remove the library signal the tests measure. The subclustering
section is the only exception. It uses corrected space to define nodes and feeds
no result back into the tests.

```{r}
library(tidyverse)
library(SingleCellExperiment)
library(scater)
library(scran)
library(batchelor)
library(patchwork)
library(readxl)
library(destiny)
library(BiocSingular)
library(DT)
source(paste0(params$project_dir, "/src/scRNA_seq_functions.R"))
```

```{r}
object_dir      = paste0(params$project_dir, "/results/", params$results_date,
                        "_no_doublets")
atlas_dir       = paste0(params$project_dir, "/results/", params$results_date,
                        "_atlas")
results_dir     = paste0(params$project_dir, "/results/", params$results_date,
                        "_female_continuum")
Robject_out_dir = paste0(results_dir, "/Robjects")
if (!dir.exists(Robject_out_dir)) dir.create(Robject_out_dir, recursive = TRUE)
data_dir        = paste0(params$project_dir, "/data")
```

```{r, echo=FALSE}
# Cells come from the labelling written by 13_atlas_and_stage_correspondence.Rmd
# rather than from 02's sex subanalysis. 02 selects its cells by cluster numbers
# taken from the with-doublets annotation, which do not apply to this
# clustering, and nothing here needs its subclustering: this document only ever
# used it to look up which all-cells cluster a cell came from, which 13 now
# states directly.
label_file = paste0(atlas_dir, "/cell_cluster_labels.csv")
if (!file.exists(label_file))
  stop("missing ", label_file,
       ". Run 13_atlas_and_stage_correspondence.Rmd first.")

cell_labels = read.csv(label_file, stringsAsFactors = FALSE)

all.sce          = readRDS(paste0(object_dir, "/all.sce.Rds"))
merged.all.sce   = readRDS(paste0(object_dir, "/merged.sce.Rds"))
combined.all.sce = readRDS(paste0(object_dir, "/combined.sce.Rds"))
```

```{r}
library_meta = data.frame(
  batch   = c("UND1_C00004336", "UND2_C00004362", "TF_C00004337", "TM_C00004374"),
  library = c("Undifferentiated.1", "Undifferentiated.2", "Female", "Male"),
  stage   = c("larva", "larva", "transformer", "transformer"),
  stringsAsFactors = FALSE)

lib_cols = c("Female" = "#FF6DB6", "Male" = "#006ddb",
             "Undifferentiated.1" = "#009292", "Undifferentiated.2" = "#b66ddf")
```

# Cells in the cluster

```{r}
if (!as.character(params$female_cluster) %in% cell_labels$final_label)
  stop("cluster ", params$female_cluster, " is not in ", label_file,
       ". Labels present: ",
       paste(sort(unique(cell_labels$final_label)), collapse = ", "),
       ". Cluster numbering changes on every rerun of the clustering, so set ",
       "the female_cluster parameter from the current labelling.")

female_cells = cell_labels$barcode[cell_labels$final_label ==
                                     as.character(params$female_cluster)]

female_type = unique(cell_labels$cell_type[cell_labels$final_label ==
                                             as.character(params$female_cluster)])
```

`r paste0(length(female_cells), " cells come from cluster ", params$female_cluster, ", annotated ", female_type, ".")`

**The stage boundary cannot be resolved inside this dataset.** Larvae were
collected in July and transformers in November, so any ordering recovered here
is equally consistent with maturation and with a collection batch effect. No
alternative trajectory algorithm changes that. Genes that change along the
ordering are to be checked against independently collected bulk data. Nothing
below should be described as developmental time on the strength of this
document alone.

## Libraries contributing

```{r}
fem_list_all = lapply(all.sce, function(x) {
  keep = intersect(colnames(x), female_cells)
  if (length(keep) == 0) return(NULL)
  x[, keep]
})
fem_list_all = fem_list_all[!vapply(fem_list_all, is.null, logical(1))]

lib_counts = data.frame(batch   = names(fem_list_all),
                        n_cells = vapply(fem_list_all, ncol, integer(1)),
                        stringsAsFactors = FALSE) %>%
  left_join(library_meta, by = "batch") %>%
  mutate(kept = n_cells >= params$min_library_cells) %>%
  dplyr::select(batch, library, stage, n_cells, kept) %>%
  arrange(desc(n_cells))

datatable(lib_counts, rownames = FALSE)

dropped = lib_counts %>% filter(!kept)
fem_list = fem_list_all[lib_counts$batch[lib_counts$kept]]
```

`r if (nrow(dropped) == 0) "Every library contributes enough cells to be analyzed." else paste0("Dropped for contributing fewer than ", params$min_library_cells, " cells: ", paste0(dropped$library, " (", dropped$n_cells, ")", collapse = "; "), ". They are too small for a per-library analysis.")`

```{r}
universe = Reduce(intersect, lapply(fem_list, rownames))
fem_list = lapply(fem_list, "[", i = universe, )

common_cols = Reduce(intersect, lapply(fem_list, function(x) colnames(colData(x))))
fem_list    = lapply(fem_list, function(x) {
  colData(x)     = colData(x)[, common_cols, drop = FALSE]
  reducedDims(x) = list()
  x
})

fem_list_norm = if (length(fem_list) > 1) do.call(multiBatchNorm, fem_list) else fem_list

fem.sce = do.call(cbind, fem_list_norm)
fem.sce$batch   = rep(names(fem_list), times = vapply(fem_list, ncol, integer(1)))
fem.sce$library = library_meta$library[match(fem.sce$batch, library_meta$batch)]
fem.sce$stage   = library_meta$stage[match(fem.sce$batch, library_meta$batch)]
```

`r ncol(fem.sce)` cells retained across `r length(fem_list)` libraries.
`multiBatchNorm` rescales for sequencing depth only and does not adjust
expression structure, so library differences remain visible and testable.

# Subclustering

Every other section runs on `fem.sce`, the uncorrected object built above. This
section is the exception.

Subclustering uncorrected cells from several libraries recovers the libraries,
because they are the largest source of variation. A trajectory over nodes
defined that way would be a graph of libraries. The subclusters are therefore
built in `fastMNN` corrected space, computed on this cluster's cells alone
rather than inherited from the all-cells correction. The correction runs across
every library that contributes cells to this cluster, larval libraries included.

The corrected object defines subcluster membership and nothing else. The
continuum tests stay on uncorrected expression.

```{r}
sub_file = paste0(Robject_out_dir, "/fem.corrected.sce.Rds")

if (file.exists(sub_file)) {
  fem.corrected.sce = readRDS(sub_file)
} else {
  fem_dec  = lapply(fem_list, modelGeneVar)
  comb_dec = if (length(fem_dec) > 1) do.call(combineVar, fem_dec) else fem_dec[[1]]
  comb_hvg = getTopHVGs(comb_dec, n = 5000)

  set.seed(1)
  fem.corrected.sce =
    if (length(fem_list_norm) > 1) {
      do.call(fastMNN, c(fem_list_norm,
                         list(subset.row = comb_hvg, BSPARAM = RandomParam())))
    } else {
      # A single surviving library has nothing to correct between. PCA stands in
      # for the corrected space and is given the same reducedDim name so that
      # everything downstream is indifferent to which branch ran.
      x = fem_list_norm[[1]]
      if (!"logcounts" %in% assayNames(x)) x = logNormCounts(x)
      set.seed(1)
      x = runPCA(x, subset_row = comb_hvg, ncomponents = 50, BSPARAM = RandomParam())
      reducedDim(x, "corrected") = reducedDim(x, "PCA")
      # fastMNN would have added this; the metadata lookup below needs it either way
      x$batch = names(fem_list_norm)[1]
      x
    }

  set.seed(1)
  fem.corrected.sce = runUMAP(fem.corrected.sce, dimred = "corrected")

  saveRDS(fem.corrected.sce, file = sub_file)
}

# metadata attached by barcode, never by position
fem.corrected.sce$library = library_meta$library[
  match(as.character(fem.corrected.sce$batch), library_meta$batch)]
fem.corrected.sce$stage = library_meta$stage[
  match(as.character(fem.corrected.sce$batch), library_meta$batch)]
stopifnot(!any(is.na(fem.corrected.sce$library)),
          !any(is.na(fem.corrected.sce$stage)))
```

## Choosing k by stability

`k` is chosen by a stability sweep. Each value is scored by resampling 80% of
the cells ten times, clustering each subsample, and taking the mean adjusted
Rand index between every pair of runs on the cells they share. A `k` that gives
the same partition from different subsamples reflects structure in the data.

Only values that yield at least two clusters are eligible, since a single node
cannot carry a trajectory. The most reproducible of those is chosen. A non-zero
`sub_k` parameter overrides the sweep.

```{r}
k_range = as.integer(trimws(strsplit(as.character(params$sub_k_range), ",")[[1]]))
stopifnot(length(k_range) > 0, !any(is.na(k_range)))

stab_file = paste0(Robject_out_dir, "/fem_stable_k.Rds")
if (file.exists(stab_file)) {
  sub_stab = readRDS(stab_file)
} else {
  sub_stab = find_stable_k(fem.corrected.sce, k_range = k_range,
                           n_iterations = params$sub_n_iterations,
                           use.dimred = "corrected")
  saveRDS(sub_stab, file = stab_file)
}

stab_tab = dplyr::bind_rows(lapply(sub_stab, function(z)
  data.frame(k               = z$k,
             mean_n_clusters = round(z$mean_n_clusters, 2),
             sd_n_clusters   = round(z$sd_n_clusters, 2),
             mean_ari        = round(z$mean_ari, 3),
             min_ari         = round(z$min_ari, 3))))

eligible = stab_tab %>% dplyr::filter(mean_n_clusters >= 2, !is.na(mean_ari))

chosen_k =
  if (as.integer(params$sub_k) > 0) {
    as.integer(params$sub_k)
  } else if (nrow(eligible) > 0) {
    eligible$k[which.max(eligible$mean_ari)]
  } else {
    NA_integer_
  }

if (is.na(chosen_k))
  stop("no value of k in the sweep produced two or more clusters. The cells of ",
       "this cluster do not subdivide reproducibly at any tested resolution, so ",
       "there are no nodes to build a trajectory over. This is a result, not a ",
       "failure: report it rather than lowering k until clusters appear.")

write.csv(stab_tab, file = paste0(results_dir, "/female_subcluster_stability.csv"),
          row.names = FALSE)

datatable(stab_tab, rownames = FALSE, options = list(dom = "t", paging = FALSE))
```

```{r, fig.width=7, fig.height=3.2}
ggplot(stab_tab, aes(x = k)) +
  geom_line(aes(y = mean_ari), linewidth = 1, colour = "#004488") +
  geom_point(aes(y = mean_ari), size = 2, colour = "#004488") +
  geom_vline(xintercept = chosen_k, linetype = "dashed", colour = "grey40") +
  scale_x_continuous(breaks = stab_tab$k) +
  labs(x = "k", y = "mean pairwise ARI",
       title = paste0("Subclustering stability, chosen k = ", chosen_k)) +
  theme_bw()

ggsave(file = paste0(results_dir, "/female_subcluster_stability.png"),
       width = 7, height = 3.2, dpi = 300)
```

`r paste0("**k = ", chosen_k, "**", if (as.integer(params$sub_k) > 0) " was set by the sub_k parameter, overriding the sweep." else paste0(" was chosen as the most reproducible value giving at least two clusters, with a mean pairwise ARI of ",stab_tab$mean_ari[stab_tab$k == chosen_k], "."))`

## The subclusters

```{r}
set.seed(1)
sub_snn = buildSNNGraph(fem.corrected.sce, k = chosen_k, use.dimred = "corrected")
fem.corrected.sce$subcluster =
  factor(igraph::cluster_louvain(sub_snn)$membership)

n_sub = nlevels(fem.corrected.sce$subcluster)

sub_cd = as.data.frame(colData(fem.corrected.sce)) %>%
  tibble::rownames_to_column("barcode")
```

```{r}
sub_by_lib = sub_cd %>%
  dplyr::count(subcluster, library, name = "n") %>%
  tidyr::pivot_wider(names_from = library, values_from = n, values_fill = 0) %>%
  dplyr::mutate(Total = rowSums(dplyr::across(-subcluster)))

sub_by_stage = sub_cd %>%
  dplyr::count(subcluster, stage, name = "n") %>%
  tidyr::pivot_wider(names_from = stage, values_from = n, values_fill = 0)

sub_tab = dplyr::left_join(sub_by_lib, sub_by_stage, by = "subcluster")

# the largest share any single library takes of each subcluster
lib_names = intersect(library_meta$library, colnames(sub_tab))
sub_tab$max_one_library_pct =
  round(100 * apply(sub_tab[, lib_names, drop = FALSE], 1, max) / sub_tab$Total, 1)

write.csv(sub_tab, file = paste0(results_dir, "/female_subcluster_composition.csv"),
          row.names = FALSE)

datatable(sub_tab, rownames = FALSE, options = list(dom = "t", paging = FALSE))
```

```{r}
# label-invariant agreement between the subclustering and each covariate
ari_stage = if (length(unique(sub_cd$stage)) > 1)
  mclust::adjustedRandIndex(sub_cd$subcluster, sub_cd$stage) else NA_real_
ari_library = if (length(unique(sub_cd$library)) > 1)
  mclust::adjustedRandIndex(sub_cd$subcluster, sub_cd$library) else NA_real_

sub_is_stage   = !is.na(ari_stage)   && ari_stage   >= 0.5
sub_is_library = !is.na(ari_library) && ari_library >= 0.5

agree_tab = data.frame(
  compared_with = c("stage", "library"),
  ARI           = round(c(ari_stage, ari_library), 3))

datatable(agree_tab, rownames = FALSE, options = list(dom = "t"))
```

`r paste0("**", n_sub, " subclusters at k = ", chosen_k, ".** The adjusted Rand index compares two partitions of the same cells and ignores how either is numbered. A value of 1 means the partitions are identical, and 0 means they agree no better than chance.")`

`r if (sub_is_stage) paste0("**The subclusters are the stage split** (ARI ", round(ari_stage, 3), " against stage). They separate larval from transformer cells, and in this dataset that also separates the July and November collections. A trajectory over these nodes will order cells along that boundary. The ordering could reflect maturation within a committed female lineage, the transition out of the undifferentiated state, or a four-month collection batch effect, and this document cannot tell them apart. Report any trajectory over these nodes as the stage axis, and do not describe the larval nodes as female on its strength.") else if (sub_is_library) paste0("**The subclusters track library rather than stage** (ARI ", round(ari_library, 3), " against library, ", round(ari_stage, 3), " against stage). The correction was computed across these libraries, so subclusters that still separate by library reflect what it failed to remove. These nodes are a poor basis for a trajectory.") else paste0("**The subclusters do not correspond to stage or library** (ARI ", round(ari_stage, 3), " and ", round(ari_library, 3), " respectively). A trajectory over these nodes needs this, but it is not enough: what the subclusters are still has to be established from their markers.")`

```{r, fig.width=13, fig.height=4}
p_sub = plotReducedDim(fem.corrected.sce, dimred = "UMAP",
                       colour_by = "subcluster", text_by = "subcluster") +
  ggtitle("Subcluster")
p_stg = plotReducedDim(fem.corrected.sce, dimred = "UMAP",
                       colour_by = "stage") + ggtitle("Stage")
p_lib = plotReducedDim(fem.corrected.sce, dimred = "UMAP",
                       colour_by = "library") +
  # scater maps colour_by to fill for shape-21 points in some versions and to
  # colour in others; the unused scale is ignored
  scale_colour_manual(values = lib_cols) +
  scale_fill_manual(values = lib_cols) +
  ggtitle("Library")

p_sub + p_stg + p_lib

ggsave(file = paste0(results_dir, "/female_subcluster_umap.png"),
       width = 13, height = 4, dpi = 300)
```

```{r}
# written out so a female-only trajectory document can consume these nodes
# without repeating the clustering, and so the k that produced them is recorded
sub_labels = sub_cd %>%
  dplyr::select(barcode, batch, library, stage, subcluster)

write.csv(sub_labels,
          file = paste0(results_dir, "/female_subcluster_labels.csv"),
          row.names = FALSE)
saveRDS(setNames(as.character(sub_labels$subcluster), sub_labels$barcode),
        file = paste0(Robject_out_dir, "/female_subcluster.Rds"))
writeLines(as.character(chosen_k),
           paste0(Robject_out_dir, "/female_subcluster_k_used.txt"))
```

`r paste0("Subcluster membership for ", nrow(sub_labels), " cells written to female_subcluster_labels.csv, and the k that produced it to female_subcluster_k_used.txt. Delete Robjects/fem.corrected.sce.Rds and Robjects/fem_stable_k.Rds to force both to be recomputed.")`

# Markers

```{r}
gene_description =
  read_excel(paste0(data_dir, "/41467_2024_55019_MOESM3_ESM.xlsx"), skip = 1) %>%
  dplyr::rename(WangGeneID = "CountMatrix_GeneID",
                Gene        = "NCBI_GTF_GENEID",
                description = "NCBI_GTF_ProteinInfo") %>%
  dplyr::select(Gene, WangGeneID, description) %>%
  filter(!is.na(Gene))

resolve_marker = function(symbol, rn, gd) {
  if (symbol %in% rn) return(symbol)
  g = gd$Gene[gd$WangGeneID == symbol]
  g = g[g %in% rn]
  if (length(g) > 0) return(g[1])
  NA_character_
}

# families are declared, so coherence can be tested within a known programme
zp_wang  = unique(gene_description$WangGeneID[grepl("^ZP[0-9]", gene_description$WangGeneID)])

marker_families = bind_rows(
  data.frame(symbol = zp_wang,  family = "ZP (oocyte)",        stringsAsFactors = FALSE),
  data.frame(symbol = c("DDX4", "PIWIL1", "SYCP3"),
             family = "germline, general",                     stringsAsFactors = FALSE),
  data.frame(symbol = c("ALPL", "PRDM14"),
             family = "early germline",                        stringsAsFactors = FALSE)) %>%
  distinct(symbol, .keep_all = TRUE)

rn = rownames(fem.sce)
marker_families = marker_families %>%
  rowwise() %>%
  mutate(gene_id = resolve_marker(symbol, rn, gene_description)) %>%
  ungroup()

resolved   = marker_families %>% filter(!is.na(gene_id)) %>%
  distinct(gene_id, .keep_all = TRUE)
unresolved = marker_families %>% filter(is.na(gene_id))
```

* Requested, after expanding the ZP family: `r nrow(marker_families)`
* Resolved in `rownames`: `r nrow(resolved)`
* Not found, and therefore excluded: `r if (nrow(unresolved) == 0) "none" else paste(unresolved$symbol, collapse = ", ")`

```{r}
datatable(resolved %>% dplyr::count(family, name = "n_resolved"), rownames = FALSE)
```

```{r}
# shared helpers, used by both the combined and per-library analyses
safe_rho = function(x, y) {
  if (length(x) < 10) return(NA_real_)
  if (length(unique(x)) < 2 || length(unique(y)) < 2) return(NA_real_)
  suppressWarnings(cor(x, y, method = "spearman"))
}
safe_p = function(x, y) {
  if (length(x) < 10) return(NA_real_)
  if (length(unique(x)) < 2 || length(unique(y)) < 2) return(NA_real_)
  suppressWarnings(cor.test(x, y, method = "spearman")$p.value)
}

# partial Spearman: rank everything, regress the controls out of both sides,
# then correlate the residuals
partial_rho = function(x, y, Z) {
  Z  = as.data.frame(Z)
  ok = stats::complete.cases(x, y, Z)
  if (sum(ok) < 10) return(NA_real_)
  xr = rank(x[ok]); yr = rank(y[ok])
  if (length(unique(xr)) < 2 || length(unique(yr)) < 2) return(NA_real_)
  Zr = as.data.frame(lapply(Z[ok, , drop = FALSE], rank))
  if (any(vapply(Zr, function(z) length(unique(z)) < 2, logical(1)))) return(NA_real_)
  rx = stats::residuals(stats::lm(xr ~ ., data = Zr))
  ry = stats::residuals(stats::lm(yr ~ ., data = Zr))
  suppressWarnings(stats::cor(rx, ry))
}

# per-marker partial correlations against an axis, controlling for depth
partial_trends = function(expr_long, cdb, axis_col = "DC1",
                          controls = c("sum", "detected")) {
  ctrl = intersect(controls, colnames(cdb))
  if (length(ctrl) == 0) return(NULL)
  j = expr_long %>%
    left_join(cdb %>% dplyr::select(barcode, all_of(axis_col), all_of(ctrl)),
              by = "barcode")
  do.call(rbind, lapply(split(j, j$symbol), function(d) {
    data.frame(symbol = d$symbol[1],
               rho_partial = partial_rho(d[[axis_col]], d$logexpr,
                                         d[, ctrl, drop = FALSE]),
               stringsAsFactors = FALSE)
  }))
}

embed_cells = function(sce, n_hvg = 2000, n_pc = 30, n_dc_from = 15, block = NULL) {
  dec  = modelGeneVar(sce, block = block)
  hvgs = getTopHVGs(dec, n = min(n_hvg, nrow(sce) - 1))
  set.seed(1)
  npc = min(n_pc, ncol(sce) - 1)
  tmp = runPCA(sce, subset_row = hvgs, ncomponents = npc)
  set.seed(1)
  dm  = destiny::DiffusionMap(reducedDim(tmp, "PCA")[, seq_len(min(n_dc_from, npc))])
  list(pca        = reducedDim(tmp, "PCA"),
       percentVar = attr(reducedDim(tmp, "PCA"), "percentVar"),
       dc         = destiny::eigenvectors(dm)[, 1:min(5, ncol(destiny::eigenvectors(dm)))],
       dm         = dm)
}

marker_expression = function(sce, resolved) {
  m = as.matrix(logcounts(sce)[resolved$gene_id, , drop = FALSE])
  rownames(m) = resolved$symbol
  t(m) %>% as.data.frame() %>%
    tibble::rownames_to_column("barcode") %>%
    tidyr::pivot_longer(-barcode, names_to = "symbol", values_to = "logexpr")
}

trend_table = function(expr_long, axis_df, axis_col) {
  expr_long %>%
    left_join(axis_df, by = "barcode") %>%
    group_by(symbol) %>%
    summarize(n_cells        = dplyr::n(),
              pct_expressing = 100 * mean(logexpr > 0),
              rho            = safe_rho(.data[[axis_col]], logexpr),
              p_value        = safe_p(.data[[axis_col]], logexpr),
              .groups = "drop") %>%
    mutate(FDR = p.adjust(p_value, "BH"),
           direction = case_when(
             is.na(rho)      ~ "undetermined",
             FDR >= 0.05     ~ "flat",
             rho >=  0.3     ~ "increases",
             rho <= -0.3     ~ "decreases",
             TRUE            ~ "weak")) %>%
    left_join(resolved %>% dplyr::select(symbol, family), by = "symbol") %>%
    arrange(desc(abs(rho)))
}
```

# 1. Combined analysis

```{r}
combined_file = paste0(Robject_out_dir, "/female_combined_embedding.Rds")

if (file.exists(combined_file)) {
  emb = readRDS(combined_file)
} else {
  message("Computing the combined embedding. Cached in ", combined_file, ".")
  emb = embed_cells(fem.sce,
                    block = if (length(fem_list) > 1) fem.sce$batch else NULL)
  saveRDS(emb, file = combined_file)
}

colnames(emb$dc) = paste0("DC", seq_len(ncol(emb$dc)))
reducedDim(fem.sce, "PCA")          = emb$pca
reducedDim(fem.sce, "DiffusionMap") = emb$dc

cd = as.data.frame(colData(fem.sce))
cd$barcode = colnames(fem.sce)
cd$DC1 = emb$dc[, "DC1"]; cd$DC2 = emb$dc[, "DC2"]
cd$PC1 = emb$pca[, 1];    cd$PC2 = emb$pca[, 2]
```

PC1 explains `r round(emb$percentVar[1], 1)`% of variance, PC2
`r round(emb$percentVar[2], 1)`%.

## Bimodality of DC1

Across a bimodal distribution, a correlation measures the gap between the two
modes rather than a gradient within them. Bimodality is tested first for that
reason.

```{r}
set.seed(1)
km = kmeans(matrix(cd$DC1, ncol = 1), centers = 2, nstart = 25)
cd$mode = paste0("mode", km$cluster)

# relabel so mode1 is always the lower end of DC1
if (mean(cd$DC1[cd$mode == "mode1"]) > mean(cd$DC1[cd$mode == "mode2"])) {
  cd$mode = ifelse(cd$mode == "mode1", "mode2", "mode1")
}

between_ratio = km$betweenss / km$totss
mode_gap = min(cd$DC1[cd$mode == "mode2"]) - max(cd$DC1[cd$mode == "mode1"])
dc1_range = diff(range(cd$DC1))
gap_frac = mode_gap / dc1_range

bimodal = between_ratio >= 0.75
```

A two-means split of DC1 accounts for `r round(100 * between_ratio, 1)`% of its
total variance, and the gap between the modes spans
`r round(100 * gap_frac, 1)`% of the DC1 range.

`r if (bimodal) "**DC1 is bimodal.** The cells fall into two separated groups rather than lying along a gradient, so DC1 is read below as a separation between groups." else "DC1 is not clearly bimodal, so it is read below as a continuous axis."`

## Separation between the modes

```{r}
tech_vars = intersect(c("sum", "detected", "subsets_Mito_percent"), colnames(cd))

mode_tab = cd %>%
  group_by(mode) %>%
  summarize(n_cells = dplyr::n(),
            across(all_of(tech_vars), median, .names = "median_{.col}"),
            .groups = "drop")

mode_lib = cd %>%
  dplyr::count(mode, library, name = "n") %>%
  group_by(mode) %>% mutate(pct = 100 * n / sum(n)) %>% ungroup()

mode_stage = cd %>%
  dplyr::count(mode, stage, name = "n") %>%
  group_by(mode) %>% mutate(pct = 100 * n / sum(n)) %>% ungroup()

datatable(mode_tab %>% mutate(across(where(is.numeric), ~round(., 1))),
          rownames = FALSE)
datatable(mode_lib %>% mutate(pct = round(pct, 1)), rownames = FALSE)
```

```{r}
stage_r2 = if (length(unique(cd$stage)) > 1)
  summary(lm(cd$DC1 ~ cd$stage))$r.squared else NA_real_
lib_r2 = if (length(unique(cd$library)) > 1)
  summary(lm(cd$DC1 ~ cd$library))$r.squared else NA_real_

mode_is_stage = !is.na(stage_r2) && stage_r2 >= 0.5
```

Stage explains `r round(100 * stage_r2, 1)`% of the variance in DC1, and library
of origin `r round(100 * lib_r2, 1)`%.

`r if (mode_is_stage) "**The two modes are the larval and transformer cells.** Collection date varies with stage, so the separation could reflect a difference between larval and transformer germ cells or a July to November batch effect. The marker tests below address which." else "The modes do not correspond to stage, so the stage and batch contrast does not separate them."`

`r if (!is.na(lib_r2) && !is.na(stage_r2) && lib_r2 - stage_r2 < 0.05) "Library explains little beyond stage: the two libraries within each stage sit together. Per-library technical noise would have separated all four libraries." else ""`

## Technical covariates within each mode

If DC1 tracked detection rate as a gradient, the association would persist
inside each mode. If only the two groups differ, the within-mode correlations
collapse.

```{r}
within_tech = bind_rows(lapply(unique(cd$mode), function(m) {
  sub = cd[cd$mode == m, ]
  bind_rows(lapply(tech_vars, function(v) {
    data.frame(mode = m, covariate = v,
               rho_within = safe_rho(sub$DC1, sub[[v]]),
               stringsAsFactors = FALSE)
  }))
}))

pooled_tech = bind_rows(lapply(tech_vars, function(v) {
  data.frame(covariate = v, rho_pooled = safe_rho(cd$DC1, cd[[v]]),
             stringsAsFactors = FALSE)
}))

tech_compare = within_tech %>%
  tidyr::pivot_wider(id_cols = covariate, names_from = mode,
                     values_from = rho_within, names_prefix = "rho_") %>%
  left_join(pooled_tech, by = "covariate")

write.csv(tech_compare,
          file = paste0(results_dir, "/SuppTable_technical_within_mode.csv"),
          row.names = FALSE)

datatable(tech_compare %>% mutate(across(where(is.numeric), ~round(., 3))),
          rownames = FALSE)
```

```{r}
max_within = max(abs(within_tech$rho_within), na.rm = TRUE)
max_pooled = max(abs(pooled_tech$rho_pooled), na.rm = TRUE)
technical_gradient = isTRUE(max_within >= 0.4)
```

The strongest pooled correlation is `r round(max_pooled, 2)`; the strongest
within either mode is `r round(max_within, 2)`.

`r if (technical_gradient) "**A technical gradient is present within a mode.** Detection rate structures the data even after the group separation is accounted for, so treat any ordering with suspicion." else "**The pooled association comes from the group difference.** Within each mode, DC1 is not strongly related to detection rate. The two groups differ in RNA content, but that does not make DC1 a technical artifact."`

## Later diffusion components

If DC1 is the stage separation, DC2 or a later component might carry structure
within a stage. Every component is scanned with the same tests. Read the table
with these limits in mind:

* **Later components are orthogonal to DC1.** A maturation gradient would partly
  align with the stage axis, so forcing it perpendicular to DC1 distorts the
  signal being looked for.
* **A few cells can produce a component.** Weakly connected cells can dominate a
  diffusion component, so outlier dominance is reported alongside.

The per-library analysis in section 2 asks the same question without a stage
axis to be orthogonal to.

```{r}
expr_long_all = marker_expression(fem.sce, resolved)

outlier_share = function(v) {
  d = abs(v - median(v))
  k = max(1, ceiling(0.01 * length(v)))
  sum(sort(d, decreasing = TRUE)[seq_len(k)]^2) / sum(d^2)
}

dc_scan = bind_rows(lapply(colnames(emb$dc), function(dcn) {
  v  = emb$dc[, dcn]
  ax = data.frame(barcode = cd$barcode, axis = v, stringsAsFactors = FALSE)

  tt = expr_long_all %>%
    left_join(ax, by = "barcode") %>%
    group_by(symbol) %>%
    summarize(rho = safe_rho(axis, logexpr),
              p   = safe_p(axis, logexpr), .groups = "drop") %>%
    mutate(FDR = p.adjust(p, "BH"))

  data.frame(
    component     = dcn,
    eigenvalue    = as.numeric(destiny::eigenvalues(emb$dm)[match(dcn, colnames(emb$dc))]),
    stage_r2      = if (length(unique(cd$stage)) > 1)
                      summary(lm(v ~ cd$stage))$r.squared else NA_real_,
    library_r2    = if (length(unique(cd$library)) > 1)
                      summary(lm(v ~ cd$library))$r.squared else NA_real_,
    max_tech_rho  = suppressWarnings(max(abs(vapply(tech_vars,
                      function(t) safe_rho(v, cd[[t]]), numeric(1))), na.rm = TRUE)),
    n_directional = sum(tt$FDR < 0.05 & abs(tt$rho) >= 0.3, na.rm = TRUE),
    outlier_share = outlier_share(v),
    stringsAsFactors = FALSE)
})) %>%
  mutate(batch_free    = !is.na(stage_r2) & stage_r2 < 0.25,
         outlier_driven = outlier_share >= 0.5,
         candidate     = batch_free & !outlier_driven &
                         max_tech_rho < 0.4 & n_directional >= 3)

write.csv(dc_scan, file = paste0(results_dir, "/SuppTable_dc_scan.csv"),
          row.names = FALSE)

datatable(dc_scan %>% mutate(across(where(is.numeric), ~signif(., 3))),
          rownames = FALSE)
```

`outlier_share` is the fraction of a component's total squared deviation
contributed by its most extreme 1% of cells. A value near 1 means the component
describes a few cells rather than the population.

```{r}
candidates = dc_scan %>% filter(candidate)
later_batch_free = dc_scan %>% filter(component != "DC1", batch_free)
```

`r if (nrow(candidates) > 0) paste0("**", paste(candidates$component, collapse = ", "), "** avoid the stage split, are not driven by outliers, are not associated with detection rate, and carry monotonic marker changes. Each is a candidate within-stage axis, subject to the orthogonality limit above.") else if (nrow(later_batch_free) > 0) paste0(paste(later_batch_free$component, collapse = ", "), " avoid the stage split but fail the remaining checks: each is driven by outlier cells, associated with detection rate, or carries no monotonic marker changes. No later component gives a usable within-stage axis.") else "No component other than DC1 avoids the stage split, so the combined embedding has no batch-free axis. Within-stage structure has to be looked for inside a single library, in section 2."`

```{r, fig.width=11, fig.height=3.5}
dc_long = as.data.frame(emb$dc) %>%
  mutate(barcode = cd$barcode, library = cd$library) %>%
  tidyr::pivot_longer(starts_with("DC"), names_to = "component", values_to = "value")

p_dcs = ggplot(dc_long, aes(x = value, fill = library, color = library)) +
  geom_density(alpha = 0.3) +
  scale_fill_manual(values = lib_cols) +
  scale_color_manual(values = lib_cols) +
  facet_wrap(~ component, scales = "free", nrow = 1) +
  xlab(NULL) + ylab("density")

p_dcs
ggsave(p_dcs, file = paste0(results_dir, "/all_DCs_by_library.png"),
       width = 11, height = 3.5)
```

## Marker expression across the modes

Coherence within a declared gene family narrows the possibilities but does not
separate a differentiation programme from a technical difference. A batch effect
can move a gene family as a set through:

* ambient RNA from an abundant cell type contaminating every droplet of one
  library
* paralogous genes sharing length and GC content, so a capture bias hits them
  together
* dissociation stress inducing the immediate-early and heat shock families

Ambient RNA is the main concern here. Oocytes are large, fragile and RNA-rich,
zona pellucida transcripts are among their most abundant, and the signal appears
in the female transformer library. The ambient RNA control below tests for it.

```{r}
expr_long = marker_expression(fem.sce, resolved)

axis_df = cd %>% dplyr::select(barcode, DC1, mode, library, stage)

mode_markers = expr_long %>%
  left_join(axis_df, by = "barcode") %>%
  group_by(symbol, mode) %>%
  summarize(mean_logexpr = mean(logexpr),
            pct_expressing = 100 * mean(logexpr > 0), .groups = "drop") %>%
  tidyr::pivot_wider(id_cols = symbol,
                     names_from = mode,
                     values_from = c(mean_logexpr, pct_expressing)) %>%
  left_join(resolved %>% dplyr::select(symbol, family), by = "symbol") %>%
  mutate(log2fc_mode2_over_mode1 = mean_logexpr_mode2 - mean_logexpr_mode1)

combined_trends = trend_table(expr_long, axis_df, "DC1")

write.csv(combined_trends,
          file = paste0(results_dir, "/SuppTable_marker_trends_combined.csv"),
          row.names = FALSE)

datatable(combined_trends %>% mutate(across(where(is.numeric), ~signif(., 3))),
          filter = "top", rownames = FALSE)
```

A family with fewer than three resolved genes cannot meet the criteria, so it is
reported as `NA` (not testable) rather than `FALSE`.

```{r}
min_family_genes = 3

family_coherence = combined_trends %>%
  filter(!is.na(rho)) %>%
  group_by(family) %>%
  summarize(n_genes        = dplyr::n(),
            median_rho     = median(rho),
            n_same_sign    = max(sum(rho > 0), sum(rho < 0)),
            pct_same_sign  = 100 * max(sum(rho > 0), sum(rho < 0)) / dplyr::n(),
            n_significant  = sum(FDR < 0.05 & abs(rho) >= 0.3),
            .groups = "drop") %>%
  mutate(testable = n_genes >= min_family_genes,
         coherent = ifelse(testable,
                           pct_same_sign >= 80 & n_significant >= 3,
                           NA)) %>%
  arrange(desc(n_genes))

write.csv(family_coherence,
          file = paste0(results_dir, "/SuppTable_family_coherence.csv"),
          row.names = FALSE)

datatable(family_coherence %>% mutate(across(where(is.numeric), ~round(., 2))),
          rownames = FALSE)
```

```{r}
zp_row = family_coherence %>% filter(family == "ZP (oocyte)")
zp_coherent = nrow(zp_row) > 0 && isTRUE(zp_row$coherent)
n_testable  = sum(family_coherence$testable)
n_coherent_families = sum(family_coherence$coherent, na.rm = TRUE)
n_not_testable = sum(!family_coherence$testable)
programme_coherent = n_coherent_families >= 1

# families that miss only on the sign threshold, reported rather than lost
near_miss = family_coherence %>%
  filter(testable, !coherent, n_significant >= 3, pct_same_sign >= 65)
```

`r n_coherent_families` of `r n_testable` testable families move coherently
along DC1, defined as at least 80% of genes agreeing in sign and at least three
reaching |rho| >= 0.3 with FDR < 0.05.
`r if (n_not_testable > 0) paste0(n_not_testable, " famil", if (n_not_testable == 1) "y has" else "ies have", " fewer than ", min_family_genes, " resolved genes and could not be tested either way.") else ""`

`r if (zp_coherent) paste0("The ZP family is coherent: ", zp_row$n_genes, " genes, median rho ", round(zp_row$median_rho, 2), ", ", round(zp_row$pct_same_sign), "% agreeing in sign, ", zp_row$n_significant, " significant.") else "The ZP family does not move coherently along DC1."`

`r if (nrow(near_miss) > 0) paste0("Near misses: ", paste0(near_miss$family, " (", near_miss$n_genes, " genes, median rho ", round(near_miss$median_rho, 2), ", ", round(near_miss$pct_same_sign), "% same sign, ", near_miss$n_significant, " significant)", collapse = "; "), ". ", if (nrow(near_miss) == 1) "This family falls" else "These families fall", " below the 80% sign threshold but show a strong majority direction and many significant genes. Members of a large paralogous family need not share regulation, so a minority moving the other way may be biological.") else ""`

## Direction test

Two programmes moving in **opposite** directions is stronger evidence than
coherence within one family. Ambient RNA, capture bias and sequencing depth push
a set of genes the same way. None of them raises an oocyte programme while
lowering the general germline programme.

```{r}
late_families  = c("ZP (oocyte)")
early_families = c("germline, general", "early germline")

dir_summary = combined_trends %>%
  filter(!is.na(rho)) %>%
  mutate(programme = case_when(family %in% late_families  ~ "late (oocyte)",
                               family %in% early_families ~ "early (germline)",
                               TRUE                       ~ "other")) %>%
  filter(programme != "other") %>%
  group_by(programme) %>%
  summarize(n_genes       = dplyr::n(),
            median_rho    = median(rho),
            n_significant = sum(FDR < 0.05 & abs(rho) >= 0.3),
            .groups = "drop")

write.csv(dir_summary,
          file = paste0(results_dir, "/SuppTable_direction_test.csv"),
          row.names = FALSE)

datatable(dir_summary %>% mutate(across(where(is.numeric), ~round(., 3))),
          rownames = FALSE)
```

```{r}
late_row  = dir_summary %>% filter(programme == "late (oocyte)")
early_row = dir_summary %>% filter(programme == "early (germline)")

both_present = nrow(late_row) == 1 && nrow(early_row) == 1

opposite_directions = both_present &&
  sign(late_row$median_rho) != sign(early_row$median_rho) &&
  late_row$n_significant  >= 3 &&
  early_row$n_significant >= 1

# a Wilcoxon on the two sets of per-gene correlations, to test the split
# rather than eyeballing two medians
dir_p = if (both_present) {
  a = combined_trends %>% filter(family %in% late_families,  !is.na(rho)) %>% pull(rho)
  b = combined_trends %>% filter(family %in% early_families, !is.na(rho)) %>% pull(rho)
  suppressWarnings(wilcox.test(a, b)$p.value)
} else NA_real_
```

`r if (!both_present) "One of the two programmes has no resolved genes, so the direction test cannot be run." else paste0("The late oocyte programme has a median rho of ", round(late_row$median_rho, 2), " across ", late_row$n_genes, " genes (", late_row$n_significant, " significant), and the early germline programme a median rho of ", round(early_row$median_rho, 2), " across ", early_row$n_genes, " genes (", early_row$n_significant, " significant). Wilcoxon on the two sets of per-gene correlations: p = ", signif(dir_p, 2), ".")`

`r if (opposite_directions) "**The two programmes move in opposite directions along DC1.** Oocyte genes are high at one end and general germline genes at the other. A technical difference between libraries would move both programmes the same way." else if (both_present) "The two programmes do not move in opposite directions along DC1, so the separation does not show an early to late transition." else ""`

```{r, fig.width=11, fig.height=8}
top_markers = combined_trends %>% filter(!is.na(rho)) %>%
  slice_max(abs(rho), n = 12) %>% pull(symbol)

p_mark = ggplot(expr_long %>% filter(symbol %in% top_markers) %>%
                  left_join(axis_df, by = "barcode"),
                aes(x = DC1, y = logexpr)) +
  geom_point(aes(color = library), size = 0.3, alpha = 0.35) +
  geom_smooth(method = "loess", se = FALSE, color = "black", linewidth = 0.7) +
  scale_color_manual(values = lib_cols) +
  guides(color = guide_legend(override.aes = list(size = 3, alpha = 1))) +
  facet_wrap(~ symbol, scales = "free_y") +
  ylab("Uncorrected log-expression")

p_mark
ggsave(p_mark, file = paste0(results_dir, "/markers_vs_DC1_combined.png"),
       width = 11, height = 8)
```

```{r, fig.width=10, fig.height=4}
p_dc = ggplot(cd, aes(x = DC1, y = DC2, color = library)) +
  geom_point(size = 0.6, alpha = 0.8) +
  scale_color_manual(values = lib_cols) +
  guides(color = guide_legend(override.aes = list(size = 3))) +
  ggtitle("Diffusion map, by library")

p_dens = ggplot(cd, aes(x = DC1, fill = library, color = library)) +
  geom_density(alpha = 0.35) +
  scale_fill_manual(values = lib_cols) +
  scale_color_manual(values = lib_cols) +
  ggtitle("DC1 by library")

p_dc + p_dens
ggsave(p_dc + p_dens, file = paste0(results_dir, "/DC1_by_library.png"),
       width = 10, height = 4)
```

## Ambient RNA control

Ambient contamination is uniform across all droplets from a library, so if the
zona pellucida signal were free-floating oocyte RNA in the female transformer
library it would appear in every cell captured from that library, including the
fibroblasts, immune cells and endothelium that cannot transcribe it. Real
transcription is confined to the cells that make it.

The test therefore compares ZP expression in **non-germ** cells of the female
library against non-germ cells of the other libraries. Elevated there means
ambient. Flat there, with a high signal confined to the germ cluster, means
transcription.

```{r}
stopifnot(identical(colnames(combined.all.sce), colnames(merged.all.sce)))

zp_ids = resolved$gene_id[resolved$family == "ZP (oocyte)"]

# cluster comes from the saved labelling, not from merged.all.sce$label, which
# is the pre-split clustering and would disagree for any split cluster
all_meta = data.frame(
  barcode = colnames(combined.all.sce),
  batch   = as.character(colData(merged.all.sce)$batch),
  stringsAsFactors = FALSE) %>%
  left_join(cell_labels %>% dplyr::select(barcode, cluster = final_label),
            by = "barcode") %>%
  left_join(library_meta, by = "batch") %>%
  mutate(compartment = ifelse(cluster == as.character(params$female_cluster),
                              "germ (cluster of interest)", "other clusters"))

all_meta$zp_mean = colMeans(as.matrix(
  logcounts(combined.all.sce)[zp_ids, , drop = FALSE]))

ambient_tab = all_meta %>%
  group_by(library, compartment) %>%
  summarize(n_cells        = dplyr::n(),
            mean_ZP        = mean(zp_mean),
            pct_expressing = 100 * mean(zp_mean > 0),
            .groups = "drop") %>%
  arrange(compartment, desc(mean_ZP))

write.csv(ambient_tab,
          file = paste0(results_dir, "/SuppTable_ambient_control.csv"),
          row.names = FALSE)

datatable(ambient_tab %>% mutate(across(where(is.numeric), ~round(., 3))),
          rownames = FALSE)
```

```{r}
other_zp = ambient_tab %>% filter(compartment == "other clusters")
fem_other = other_zp$mean_ZP[other_zp$library == "Female"]
rest_other = other_zp$mean_ZP[other_zp$library != "Female"]

germ_zp = ambient_tab %>% filter(compartment == "germ (cluster of interest)")
fem_germ = germ_zp$mean_ZP[germ_zp$library == "Female"]

ambient_ratio = if (length(fem_other) == 1 && length(rest_other) > 0 &&
                    max(rest_other) > 0) fem_other / max(rest_other) else NA_real_
germ_vs_other = if (length(fem_germ) == 1 && length(fem_other) == 1 && fem_other > 0)
  fem_germ / fem_other else if (length(fem_germ) == 1) Inf else NA_real_

ambient_suspected = isTRUE(ambient_ratio >= 3)
```

Mean ZP expression in non-germ cells of the female library is
`r round(fem_other, 3)`, against a maximum of `r round(max(rest_other), 3)` in
the non-germ cells of any other library. In the germ cluster of the female
library it is `r round(fem_germ, 3)`.

`r if (ambient_suspected) "**Ambient contamination is indicated.** Non-germ cells in the female library carry ZP at three or more times the highest level in any other library. Free-floating oocyte RNA inflates the germ cell signal, so the difference between the modes cannot be taken at face value." else if (is.na(ambient_ratio)) "**The ambient control could not be computed.** Either the female library has no non-germ cells in this table, or no other library has ZP detected in its non-germ cells." else "**No sign of ambient contamination.** Non-germ cells in the female library carry ZP at less than three times the highest level in any other library. Free-floating oocyte RNA would appear in every droplet from that library, so the ZP signal reflects transcription in the germ cells."`

```{r, fig.width=8, fig.height=4}
p_amb = ggplot(all_meta, aes(x = library, y = zp_mean, fill = compartment)) +
  geom_boxplot(outlier.size = 0.3) +
  scale_fill_manual(values = c("germ (cluster of interest)" = "#FF6DB6",
                               "other clusters" = "grey70")) +
  theme(axis.text.x = element_text(angle = 30, hjust = 1)) +
  ylab("Mean ZP family log-expression") + xlab(NULL)

p_amb
ggsave(p_amb, file = paste0(results_dir, "/ambient_control_ZP.png"),
       width = 8, height = 4)
```

## Verdict for the combined analysis

```{r}
combined_verdict = if (technical_gradient) {
  "technical"
} else if (bimodal && ambient_suspected) {
  "two groups, ambient confounded"
} else if (bimodal && programme_coherent) {
  "two states"
} else if (bimodal) {
  "two groups, unexplained"
} else if (!bimodal && sum(combined_trends$direction %in% c("increases", "decreases")) >= 3) {
  "continuum"
} else {
  "no structure"
}
```

**`r combined_verdict`.**

`r switch(combined_verdict,
  "technical" = "DC1 keeps a strong association with detection rate inside each mode, so it is structured by technical variation. Nothing should be built on it.",
  "two groups, unexplained" = "The cells form two separated groups, but no declared marker family moves coherently across them. The separation cannot be attributed to a differentiation programme and may be technical or batch driven.",
  "two groups, ambient confounded" = "The cells form two separated groups and marker families move coherently, but non-germ cells of the female library also carry zona pellucida transcripts. Ambient oocyte RNA inflates the signal, so the separation cannot be read as a differentiation difference.",
  "two states" = paste0("The cells form two separated groups rather than a gradient. The separation carries a coherent marker programme, and zona pellucida expression is confined to the germ cluster, so ambient RNA does not explain it. ", if (opposite_directions) "The oocyte and general germline programmes also move in opposite directions. A uniform difference between libraries would move them the same way. " else "", "These tests rule out ambient RNA and a uniform library shift. They do not separate stage from collection date."),
  "continuum" = "DC1 is unimodal and not explained by detection rate, and markers vary monotonically along it. This supports a continuous gradient.",
  "no structure" = "DC1 is neither bimodal nor accompanied by monotonic marker changes, so there is no evidence of discrete states or a gradient.")`

# 2. Per-library analysis

Within a single library, animal, collection date and batch are constant, so a
gradient there cannot be a batch effect and a correlation with detection rate is
technical.

```{r}
perlib_file = paste0(Robject_out_dir, "/female_perlibrary.Rds")

if (file.exists(perlib_file)) {
  perlib = readRDS(perlib_file)
} else {
  message("Computing per-library embeddings. Cached in ", perlib_file, ".")
  perlib = lapply(names(fem_list), function(b) {
    sce_b = fem_list[[b]]
    if (ncol(sce_b) < 100) return(NULL)
    e = embed_cells(sce_b)
    list(batch = b, emb = e, barcodes = colnames(sce_b))
  })
  names(perlib) = names(fem_list)
  perlib = perlib[!vapply(perlib, is.null, logical(1))]
  saveRDS(perlib, file = perlib_file)
}
```

```{r}
perlib_summary = bind_rows(lapply(names(perlib), function(b) {
  e     = perlib[[b]]$emb
  bc    = perlib[[b]]$barcodes
  sce_b = fem_list[[b]]

  cdb = as.data.frame(colData(sce_b))
  cdb$barcode = bc
  cdb$DC1 = e$dc[, 1]

  tv = intersect(c("sum", "detected", "subsets_Mito_percent"), colnames(cdb))
  tech_max = max(abs(vapply(tv, function(v) safe_rho(cdb$DC1, cdb[[v]]),
                            numeric(1))), na.rm = TRUE)

  el = marker_expression(sce_b, resolved)
  tt = trend_table(el, cdb %>% dplyr::select(barcode, DC1), "DC1")

  pt = partial_trends(el, cdb)
  if (!is.null(pt)) tt = tt %>% left_join(pt, by = "symbol")

  fam = tt %>% filter(!is.na(rho)) %>% group_by(family) %>%
    summarize(n_genes = dplyr::n(),
              pct_same_sign = 100 * max(sum(rho > 0), sum(rho < 0)) / dplyr::n(),
              n_significant = sum(FDR < 0.05 & abs(rho) >= 0.3), .groups = "drop") %>%
    mutate(coherent = n_genes >= 3 & pct_same_sign >= 80 & n_significant >= 3)

  data.frame(
    batch            = b,
    library          = library_meta$library[match(b, library_meta$batch)],
    n_cells          = ncol(sce_b),
    percentVar_PC1   = e$percentVar[1],
    max_tech_rho     = tech_max,
    n_directional    = sum(tt$direction %in% c("increases", "decreases")),
    n_directional_partial = if ("rho_partial" %in% colnames(tt))
      sum(abs(tt$rho_partial) >= 0.3, na.rm = TRUE) else NA_integer_,
    n_markers_tested = nrow(tt),
    n_coherent_fam   = sum(fam$coherent, na.rm = TRUE),
    stringsAsFactors = FALSE)
})) %>%
  mutate(technical_confound  = max_tech_rho >= 0.4,
         # partialling out a covariate that already explains most of the axis
         # leaves little to correlate with, so the partial figure is only
         # informative when the axis is not near-collinear with depth
         partial_reliable    = max_tech_rho < 0.6,
         continuum_supported = !technical_confound & n_directional_partial >= 3)

write.csv(perlib_summary,
          file = paste0(results_dir, "/SuppTable_perlibrary_summary.csv"),
          row.names = FALSE)

datatable(perlib_summary %>% mutate(across(where(is.numeric), ~round(., 3))),
          rownames = FALSE)
```

Marker counts are given twice. The raw count is the number of markers correlating
with DC1. The partial count is the number still correlating after `sum` and
`detected` are regressed out of both sides. Expression correlates with depth for
almost every gene, so where DC1 tracks detection rate nearly every marker passes
the raw test and only the partial count carries information.

The partial count has its own limit. When DC1 is already highly correlated with
depth, regressing depth out leaves little variance, so a partial count near zero
is expected whether or not biology is present. `partial_reliable` marks where
the partial count can be trusted.

```{r, results="asis"}
for (i in seq_len(nrow(perlib_summary))) {
  r = perlib_summary[i, ]
  cat("**", r$library, "** (", r$n_cells, " cells): PC1 explains ",
      round(r$percentVar_PC1, 1), "% of variance; strongest DC1 to technical ",
      "covariate correlation ", round(r$max_tech_rho, 2), "; ", r$n_directional,
      " of ", r$n_markers_tested, " markers monotonic, ",
      r$n_directional_partial, " surviving control for depth; ",
      r$n_coherent_fam, " coherent marker famil",
      if (r$n_coherent_fam == 1) "y" else "ies", ". ",
      if (r$technical_confound && !r$partial_reliable) {
        paste0("Within this library the leading component tracks detection rate. Expression correlates with depth for almost every gene, so the raw marker count is not evidence of biology. DC1 and depth are nearly collinear, so the partial count cannot settle it either. This library offers no usable axis. A homogeneous population with no biological variance would give the same result.")
      } else if (r$technical_confound) {
        paste0("The leading component tracks detection rate, so the raw marker count reflects depth rather than biology. ",
               if (r$n_directional_partial >= 3)
                 "Some markers survive control for depth, so biological structure may remain beneath the technical gradient."
               else
                 "No markers survive control for depth, so no biological structure is detected beneath the technical gradient.")
      } else if (r$continuum_supported) {
        "The leading component is not technical, and markers vary monotonically along it after controlling for depth. This supports a gradient within this library."
      } else {
        "The leading component is not technical, but markers do not vary monotonically along it once depth is controlled, so there is no evidence of a gradient within this library."
      },
      "\n\n", sep = "")
}
```

```{r, fig.width=11, fig.height=4}
perlib_dc = bind_rows(lapply(names(perlib), function(b) {
  data.frame(barcode = perlib[[b]]$barcodes,
             DC1     = perlib[[b]]$emb$dc[, 1],
             DC2     = perlib[[b]]$emb$dc[, 2],
             library = library_meta$library[match(b, library_meta$batch)],
             stringsAsFactors = FALSE)
}))

p_perlib = ggplot(perlib_dc, aes(x = DC1, y = DC2, color = library)) +
  geom_point(size = 0.5, alpha = 0.8) +
  scale_color_manual(values = lib_cols) +
  facet_wrap(~ library, scales = "free", nrow = 1) +
  guides(color = "none") +
  ggtitle("Diffusion map computed within each library separately")

p_perlib
ggsave(p_perlib, file = paste0(results_dir, "/perlibrary_diffusion.png"),
       width = 11, height = 4)
```

## Marker expression in the Female library alone

Inside the female transformer library, animal, collection date, batch and sex are
constant, so structure there cannot be a batch effect. If oogonia and oocytes
coexist in this sample, the oocyte and general germline programmes should move
in opposite directions along its internal DC1.

```{r}
female_batch     = library_meta$batch[library_meta$library == "Female"]
female_available = length(female_batch) == 1 && female_batch %in% names(perlib)
```

`r if (!female_available) "The Female library is not available for this analysis, either because it contributes too few cells or because it was dropped above. The section below is skipped." else ""`

```{r}
if (female_available) {
  sce_f = fem_list[[female_batch]]
  e_f   = perlib[[female_batch]]$emb

  cdf = as.data.frame(colData(sce_f))
  cdf$barcode = perlib[[female_batch]]$barcodes
  cdf$DC1 = e_f$dc[, 1]
  cdf$DC2 = e_f$dc[, 2]

  tv_f = intersect(c("sum", "detected", "subsets_Mito_percent"), colnames(cdf))
  tech_f = data.frame(
    covariate = tv_f,
    rho_DC1   = vapply(tv_f, function(v) safe_rho(cdf$DC1, cdf[[v]]), numeric(1)),
    stringsAsFactors = FALSE)

  datatable(tech_f %>% mutate(across(where(is.numeric), ~round(., 3))),
            rownames = FALSE)
}
```

```{r}
if (female_available) {
  fem_tech_max = max(abs(tech_f$rho_DC1), na.rm = TRUE)
  fem_technical = isTRUE(fem_tech_max >= 0.4)
} else {
  fem_tech_max = NA_real_; fem_technical = FALSE
}
```

`r if (female_available) paste0("Strongest correlation between the internal DC1 and a technical covariate: ", round(fem_tech_max, 2), ". ", if (fem_technical) "**That is a technical gradient.** The internal axis tracks detection rate, so no biological reading is available." else "The internal axis is not strongly associated with detection rate.") else ""`

### Marker trends within the Female library

```{r}
if (female_available) {
  expr_f   = marker_expression(sce_f, resolved)
  trends_f = trend_table(expr_f, cdf %>% dplyr::select(barcode, DC1), "DC1")

  pt_f = partial_trends(expr_f, cdf)
  if (!is.null(pt_f)) trends_f = trends_f %>% left_join(pt_f, by = "symbol")

  write.csv(trends_f,
            file = paste0(results_dir, "/SuppTable_marker_trends_female.csv"),
            row.names = FALSE)

  datatable(trends_f %>% mutate(across(where(is.numeric), ~signif(., 3))),
            filter = "top", rownames = FALSE)

  n_dir_f_raw     = sum(trends_f$direction %in% c("increases", "decreases"))
  n_dir_f_partial = if ("rho_partial" %in% colnames(trends_f))
    sum(abs(trends_f$rho_partial) >= 0.3, na.rm = TRUE) else NA_integer_
}
```

`r if (female_available) paste0("In this library, ", n_dir_f_raw, " markers correlate with DC1 and ", n_dir_f_partial, " still do after regressing out sum and detected. ", if (fem_tech_max < 0.6) "DC1 is not strongly correlated with depth here, so the partial correlations can be read." else "DC1 is strongly correlated with depth here, so the partial correlations are underpowered.") else ""`

```{r}
if (female_available) {
  fam_f = trends_f %>%
    filter(!is.na(rho)) %>%
    group_by(family) %>%
    summarize(n_genes       = dplyr::n(),
              median_rho    = median(rho),
              pct_same_sign = 100 * max(sum(rho > 0), sum(rho < 0)) / dplyr::n(),
              n_significant = sum(FDR < 0.05 & abs(rho) >= 0.3),
              .groups = "drop") %>%
    mutate(testable = n_genes >= min_family_genes,
           coherent = ifelse(testable,
                             pct_same_sign >= 80 & n_significant >= 3, NA)) %>%
    arrange(desc(n_genes))

  write.csv(fam_f,
            file = paste0(results_dir, "/SuppTable_family_coherence_female.csv"),
            row.names = FALSE)

  datatable(fam_f %>% mutate(across(where(is.numeric), ~round(., 2))),
            rownames = FALSE)
}
```

### Direction test within the Female library

```{r}
if (female_available) {
  dir_f = trends_f %>%
    filter(!is.na(rho)) %>%
    mutate(programme = case_when(family %in% late_families  ~ "late (oocyte)",
                                 family %in% early_families ~ "early (germline)",
                                 TRUE                       ~ "other")) %>%
    filter(programme != "other") %>%
    group_by(programme) %>%
    summarize(n_genes       = dplyr::n(),
              median_rho    = median(rho),
              n_significant = sum(FDR < 0.05 & abs(rho) >= 0.3),
              .groups = "drop")

  datatable(dir_f %>% mutate(across(where(is.numeric), ~round(., 3))),
            rownames = FALSE)

  late_f  = dir_f %>% filter(programme == "late (oocyte)")
  early_f = dir_f %>% filter(programme == "early (germline)")

  both_f = nrow(late_f) == 1 && nrow(early_f) == 1
  fem_opposite = both_f &&
    sign(late_f$median_rho) != sign(early_f$median_rho) &&
    late_f$n_significant >= 3 && early_f$n_significant >= 1

  dir_p_f = if (both_f) {
    a = trends_f %>% filter(family %in% late_families,  !is.na(rho)) %>% pull(rho)
    b = trends_f %>% filter(family %in% early_families, !is.na(rho)) %>% pull(rho)
    suppressWarnings(wilcox.test(a, b)$p.value)
  } else NA_real_
} else {
  both_f = FALSE; fem_opposite = FALSE; dir_p_f = NA_real_
}
```

`r if (!female_available) "" else if (!both_f) "One programme has no resolved genes within this library, so the direction test cannot be run here." else paste0("Within the Female library alone, the late oocyte programme has a median rho of ", round(late_f$median_rho, 2), " across ", late_f$n_genes, " genes (", late_f$n_significant, " significant) and the early germline programme ", round(early_f$median_rho, 2), " across ", early_f$n_genes, " genes (", early_f$n_significant, " significant). Wilcoxon p = ", signif(dir_p_f, 2), ".")`

`r if (female_available && fem_opposite) "**The two programmes move in opposite directions within a single library.** Batch, animal, collection date and sex are constant here, so this is not a library artifact. Cells with oocyte-like and germline-like expression coexist in this sample and are ordered along the internal axis. This supports a maturation gradient within the transformer sample in addition to the difference between stages." else if (female_available && both_f) paste0("**The two programmes do not move in opposite directions within this library.** The internal axis does not show an early to late transition.", if (opposite_directions) " The opposite-direction result in the combined analysis therefore reflects the difference between stages rather than a gradient within the transformer sample." else "") else ""`

```{r, fig.width=11, fig.height=8}
if (female_available) {
  top_f = trends_f %>% filter(!is.na(rho)) %>%
    slice_max(abs(rho), n = 12) %>% pull(symbol)

  p_fem = ggplot(expr_f %>% filter(symbol %in% top_f) %>%
                   left_join(cdf %>% dplyr::select(barcode, DC1), by = "barcode"),
                 aes(x = DC1, y = logexpr)) +
    geom_point(size = 0.4, alpha = 0.3, color = "#FF6DB6") +
    geom_smooth(method = "loess", se = FALSE, color = "black", linewidth = 0.7) +
    facet_wrap(~ symbol, scales = "free_y") +
    xlab("DC1, computed within the Female library only") +
    ylab("Uncorrected log-expression")

  print(p_fem)
  ggsave(p_fem, file = paste0(results_dir, "/markers_vs_DC1_female.png"),
         width = 11, height = 8)
}
```

```{r, fig.width=10, fig.height=4}
if (female_available) {
  zp_f  = resolved$gene_id[resolved$family == "ZP (oocyte)"]
  erly  = resolved$gene_id[resolved$family == "germline, general"]

  cdf$ZP_mean       = colMeans(as.matrix(logcounts(sce_f)[zp_f, , drop = FALSE]))
  cdf$germline_mean = colMeans(as.matrix(logcounts(sce_f)[erly, , drop = FALSE]))

  p_zp = ggplot(cdf, aes(x = DC1, y = DC2, color = ZP_mean)) +
    geom_point(size = 0.8) +
    scale_color_viridis_c() +
    ggtitle("ZP family mean")

  p_gl = ggplot(cdf, aes(x = DC1, y = DC2, color = germline_mean)) +
    geom_point(size = 0.8) +
    scale_color_viridis_c(option = "magma") +
    ggtitle("General germline mean")

  print(p_zp + p_gl)
  ggsave(p_zp + p_gl,
         file = paste0(results_dir, "/female_programmes_on_DC.png"),
         width = 10, height = 4)
}
```

# 3. Summary

```{r, results="asis"}
any_within_continuum = any(perlib_summary$continuum_supported)

cat(paste0(
  ncol(fem.sce), " cells across ", length(fem_list), " libraries. ",
  "**Combined:** DC1 is ", if (bimodal) "bimodal" else "unimodal",
  ", with stage explaining ", round(100 * stage_r2, 1),
  "% of its variance. The strongest correlation with a technical covariate is ",
  round(max_pooled, 2), " pooled and ", round(max_within, 2),
  " within a mode. ", n_coherent_families, " of ", n_testable,
  " testable marker famil",
  if (n_coherent_families == 1) "ies moves" else "ies move",
  " coherently along DC1",
  if (opposite_directions)
    ", and the oocyte and general germline programmes move in opposite directions along it"
  else "",
  ". Verdict: ", combined_verdict, ". ",
  "**Per library:** ",
  if (any_within_continuum) {
    paste0(sum(perlib_summary$continuum_supported), " of ", nrow(perlib_summary),
           " libraries show a gradient internally. Batch is constant within a ",
           "library, so these gradients are not batch effects.")
  } else {
    paste0("no library shows a gradient internally, so any structure in the ",
           "combined analysis lies between the stages rather than within them.")
  },
  if (!female_available) "" else if (fem_opposite) {
    paste0(" In the Female library alone, the oocyte and general germline ",
           "programmes move in opposite directions along the internal axis. ",
           "Batch, animal, collection date and sex are constant in that library.")
  } else {
    paste0(" In the Female library alone, the two programmes do not move in ",
           "opposite directions.")
  }))
```

# Session info

```{r}
sessionInfo()
```
