---
title: "09_male_germ_cell_continuum"
author: "Stephanie Hickey"
date: "`r Sys.Date()`"
output:
  html_document:
    code_folding: hide
    toc: true
    toc_float:
      toc_collapsed: true
params:
    project_dir: /mnt/ufs18/rs-013/bioinformaticsCore/projects/chong_davidson/BCC125_sea_lamprey_scRNA
    project_name: BCC125_sea_lamprey_scRNA
    lab: Weiming Li
    results_date: "20260813"
    male_clusters: "12_transformer,15,16,17,18"
    min_library_cells: 50
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

```{r, echo=FALSE, results="asis"}
cat("\n")
cat("#### Lab:", params$lab, " \n")
cat("#### Project:", params$project_name, " \n")
```

This is the male counterpart to `08_female_germ_cell_continuum.Rmd`. It uses the
same methods to ask whether the male germ cells form an ordered progression.

**Library composition.** The male germ clusters are expected to come mostly from
the male transformer library. Within one library, animal, collection date and
batch are constant, so an ordering among those cells cannot be a batch effect.
It still describes a single animal, and comparing it with larval germ cells
brings the stage confound back. The composition check below therefore comes
first.

**A second grouping.** The male germ cells fall into several clusters defined in
batch-corrected space. The diffusion axis below is computed on uncorrected
values. If the clusters sit in sequence along that axis, two representations of
the same cells agree on the order.

Every test uses uncorrected, normalized log-expression. A pseudotime can be
computed from any set of cells, so the technical and marker checks decide
whether an ordering is real.

```{r}
library(tidyverse)
library(SingleCellExperiment)
library(scater)
library(scran)
library(batchelor)
library(patchwork)
library(readxl)
library(destiny)
library(DT)
source(paste0(params$project_dir, "/src/scRNA_seq_functions.R"))
```

```{r}
object_dir      = paste0(params$project_dir, "/results/", params$results_date,
                        "_no_doublets")
atlas_dir       = paste0(params$project_dir, "/results/", params$results_date,
                        "_atlas")
results_dir     = paste0(params$project_dir, "/results/", params$results_date,
                        "_male_continuum")
Robject_out_dir = paste0(results_dir, "/Robjects")
if (!dir.exists(Robject_out_dir)) dir.create(Robject_out_dir, recursive = TRUE)
data_dir        = paste0(params$project_dir, "/data")
```

```{r, echo=FALSE}
# Cells come from the labelling written by 13_atlas_and_stage_correspondence.Rmd
# rather than from 02's sex subanalysis, whose cluster numbers are taken from the
# with-doublets annotation and do not apply to this clustering. Nothing here
# needs 02's subclustering; it was only ever used to look up which all-cells
# cluster a cell came from.
label_file = paste0(atlas_dir, "/cell_cluster_labels.csv")
if (!file.exists(label_file))
  stop("missing ", label_file,
       ". Run 13_atlas_and_stage_correspondence.Rmd first.")

cell_labels = read.csv(label_file, stringsAsFactors = FALSE)
all.sce     = readRDS(paste0(object_dir, "/all.sce.Rds"))
```

```{r}
library_meta = data.frame(
  batch   = c("UND1_C00004336", "UND2_C00004362", "TF_C00004337", "TM_C00004374"),
  library = c("Undifferentiated.1", "Undifferentiated.2", "Female", "Male"),
  stage   = c("larva", "larva", "transformer", "transformer"),
  stringsAsFactors = FALSE)

lib_cols = c("Female" = "#FF6DB6", "Male" = "#006ddb",
             "Undifferentiated.1" = "#009292", "Undifferentiated.2" = "#b66ddf")

male_clusters = trimws(strsplit(params$male_clusters, ",")[[1]])
```

# Identify the male germ cells

Male germ cells are the cells whose cluster in the final labelling is one of
`r paste(male_clusters, collapse = ", ")`, read from `cell_cluster_labels.csv`.
Under the current annotation these are the migrating and mitotic primordial germ
cells, the two spermatocyte clusters and the migrating male germ cells.

**The stage boundary cannot be resolved inside this dataset.** Larvae were
collected in July and transformers in November. An ordering confined to the male
transformer library avoids that confound, but any comparison with larval cells
brings it back. Genes that change along the ordering are to be checked against
independently collected bulk data.

```{r}
absent = setdiff(male_clusters, cell_labels$final_label)
if (length(absent) > 0)
  stop("cluster(s) ", paste(absent, collapse = ", "), " are not in ", label_file,
       ". Labels present: ",
       paste(sort(unique(cell_labels$final_label)), collapse = ", "),
       ". Cluster numbering changes on every rerun of the clustering, so set ",
       "the male_clusters parameter from the current labelling.")

germ_meta = cell_labels %>%
  dplyr::rename(origin_cluster = final_label)
male_cells = germ_meta$barcode[germ_meta$origin_cluster %in% male_clusters]

origin_tab = germ_meta %>%
  filter(origin_cluster %in% male_clusters) %>%
  dplyr::count(origin_cluster, cell_type, name = "n_cells") %>%
  arrange(desc(n_cells))

datatable(origin_tab, rownames = FALSE)
```

`r length(male_cells)` cells across `r nrow(origin_tab)` originating clusters.

```{r}
if (length(male_cells) < 100) {
  stop(paste0("Only ", length(male_cells), " male germ cells were found. Check ",
              "params$male_clusters against the current annotation."))
}
```

## Library composition

```{r}
# library and stage already come from cell_cluster_labels.csv; joining
# library_meta again would split them into library.x and library.y
male_meta = germ_meta %>% filter(barcode %in% male_cells)
stopifnot(all(c("library", "stage") %in% colnames(male_meta)))

comp_tab = male_meta %>%
  dplyr::count(origin_cluster, library, name = "n") %>%
  group_by(origin_cluster) %>%
  mutate(pct = 100 * n / sum(n)) %>%
  ungroup()

comp_wide = comp_tab %>%
  tidyr::pivot_wider(id_cols = origin_cluster, names_from = library,
                     values_from = c(n, pct), values_fill = 0)

write.csv(comp_wide,
          file = paste0(results_dir, "/SuppTable_male_cluster_composition.csv"),
          row.names = FALSE)

datatable(comp_wide %>% mutate(across(where(is.numeric), ~round(., 1))),
          rownames = FALSE)
```

```{r}
lib_totals = male_meta %>% dplyr::count(library, name = "n_cells") %>%
  mutate(pct = 100 * n_cells / sum(n_cells)) %>%
  arrange(desc(n_cells))

datatable(lib_totals %>% mutate(pct = round(pct, 1)), rownames = FALSE)

dominant_library = lib_totals$library[1]
dominant_pct     = lib_totals$pct[1]
single_library   = dominant_pct >= 90
```

`r round(dominant_pct, 1)`% of male germ cells come from **`r dominant_library`**.

`r if (single_library) "**These cells are effectively one library.** Animal, collection date, batch and sex are constant, so structure among them cannot be a batch effect, and a correlation with detection rate is technical." else "**These cells span more than one library.** An axis that separates them may reflect stage, batch or biology, and this analysis cannot tell those apart."`

```{r, fig.width=9, fig.height=4}
p_comp = ggplot(comp_tab, aes(x = origin_cluster, y = pct, fill = library)) +
  geom_bar(stat = "identity") +
  scale_fill_manual(values = lib_cols) +
  xlab("Originating all-cells cluster") + ylab("Percent of cells")

p_comp
ggsave(p_comp, file = paste0(results_dir, "/male_cluster_composition.png"),
       width = 9, height = 4)
```

## Build an uncorrected object

```{r}
male_list_all = lapply(all.sce, function(x) {
  keep = intersect(colnames(x), male_cells)
  if (length(keep) == 0) return(NULL)
  x[, keep]
})
male_list_all = male_list_all[!vapply(male_list_all, is.null, logical(1))]

keep_batches = names(male_list_all)[
  vapply(male_list_all, ncol, integer(1)) >= params$min_library_cells]
dropped_batches = setdiff(names(male_list_all), keep_batches)
male_list = male_list_all[keep_batches]
```

`r if (length(dropped_batches) == 0) "Every library contributing male germ cells has enough cells to keep." else paste0("Dropped for contributing fewer than ", params$min_library_cells, " cells: ", paste(library_meta$library[match(dropped_batches, library_meta$batch)], collapse = ", "), ".")`

```{r}
universe  = Reduce(intersect, lapply(male_list, rownames))
male_list = lapply(male_list, "[", i = universe, )

common_cols = Reduce(intersect, lapply(male_list, function(x) colnames(colData(x))))
male_list   = lapply(male_list, function(x) {
  colData(x)     = colData(x)[, common_cols, drop = FALSE]
  reducedDims(x) = list()
  x
})

male_list_norm = if (length(male_list) > 1) do.call(multiBatchNorm, male_list) else male_list

male.sce = do.call(cbind, male_list_norm)
male.sce$batch   = rep(names(male_list), times = vapply(male_list, ncol, integer(1)))
male.sce$library = library_meta$library[match(male.sce$batch, library_meta$batch)]
male.sce$stage   = library_meta$stage[match(male.sce$batch, library_meta$batch)]
male.sce$origin_cluster = germ_meta$origin_cluster[
  match(colnames(male.sce), germ_meta$barcode)]
```

`r ncol(male.sce)` cells retained across `r length(male_list)` librar`r if (length(male_list) == 1) "y" else "ies"`.

# Markers

Marker families are declared in advance so the early and late programmes can be
tested against each other.

```{r}
gene_description =
  read_excel(paste0(data_dir, "/41467_2024_55019_MOESM3_ESM.xlsx"), skip = 1) %>%
  dplyr::rename(WangGeneID = "CountMatrix_GeneID",
                Gene        = "NCBI_GTF_GENEID",
                description = "NCBI_GTF_ProteinInfo") %>%
  dplyr::select(Gene, WangGeneID, description) %>%
  filter(!is.na(Gene))

resolve_marker = function(symbol, rn, gd) {
  if (symbol %in% rn) return(symbol)
  g = gd$Gene[gd$WangGeneID == symbol]
  g = g[g %in% rn]
  if (length(g) > 0) return(g[1])
  NA_character_
}

marker_families = bind_rows(
  data.frame(symbol = c("PAX7", "EZH2", "EPCAM", "EXOSC10", "NANOS1"),
             family = "spermatogonia / SSC (early)", stringsAsFactors = FALSE),
  data.frame(symbol = c("SYCP3", "SYCP2", "DMC1", "SPO11", "REC8",
                        "TEX11", "TEX15", "MAEL", "ASZ1", "TDRD7", "TDRD9"),
             family = "meiotic (late)",              stringsAsFactors = FALSE),
  data.frame(symbol = c("DDX4", "PIWIL1", "BOLL"),
             family = "germline, general",           stringsAsFactors = FALSE),
  data.frame(symbol = c("ALPL", "PRDM14", "PRDM1"),
             family = "early germline",              stringsAsFactors = FALSE)) %>%
  distinct(symbol, .keep_all = TRUE)

rn = rownames(male.sce)
marker_families = marker_families %>%
  rowwise() %>%
  mutate(gene_id = resolve_marker(symbol, rn, gene_description)) %>%
  ungroup()

resolved   = marker_families %>% filter(!is.na(gene_id)) %>%
  distinct(gene_id, .keep_all = TRUE)
unresolved = marker_families %>% filter(is.na(gene_id))

late_families  = "meiotic (late)"
early_families = "spermatogonia / SSC (early)"
min_family_genes = 3
```

* Requested: `r nrow(marker_families)`
* Resolved in `rownames`: `r nrow(resolved)`
* Not found, and therefore excluded: `r if (nrow(unresolved) == 0) "none" else paste(unresolved$symbol, collapse = ", ")`

```{r}
datatable(resolved %>% dplyr::count(family, name = "n_resolved"), rownames = FALSE)
```

```{r}
safe_rho = function(x, y) {
  if (length(x) < 10) return(NA_real_)
  if (length(unique(x)) < 2 || length(unique(y)) < 2) return(NA_real_)
  suppressWarnings(cor(x, y, method = "spearman"))
}
safe_p = function(x, y) {
  if (length(x) < 10) return(NA_real_)
  if (length(unique(x)) < 2 || length(unique(y)) < 2) return(NA_real_)
  suppressWarnings(cor.test(x, y, method = "spearman")$p.value)
}

partial_rho = function(x, y, Z) {
  Z  = as.data.frame(Z)
  ok = stats::complete.cases(x, y, Z)
  if (sum(ok) < 10) return(NA_real_)
  xr = rank(x[ok]); yr = rank(y[ok])
  if (length(unique(xr)) < 2 || length(unique(yr)) < 2) return(NA_real_)
  Zr = as.data.frame(lapply(Z[ok, , drop = FALSE], rank))
  if (any(vapply(Zr, function(z) length(unique(z)) < 2, logical(1)))) return(NA_real_)
  rx = stats::residuals(stats::lm(xr ~ ., data = Zr))
  ry = stats::residuals(stats::lm(yr ~ ., data = Zr))
  suppressWarnings(stats::cor(rx, ry))
}

embed_cells = function(sce, n_hvg = 2000, n_pc = 30, n_dc_from = 15, block = NULL) {
  dec  = modelGeneVar(sce, block = block)
  hvgs = getTopHVGs(dec, n = min(n_hvg, nrow(sce) - 1))
  set.seed(1)
  npc = min(n_pc, ncol(sce) - 1)
  tmp = runPCA(sce, subset_row = hvgs, ncomponents = npc)
  set.seed(1)
  dm  = destiny::DiffusionMap(reducedDim(tmp, "PCA")[, seq_len(min(n_dc_from, npc))])
  list(pca        = reducedDim(tmp, "PCA"),
       percentVar = attr(reducedDim(tmp, "PCA"), "percentVar"),
       dc         = destiny::eigenvectors(dm)[, 1:min(5, ncol(destiny::eigenvectors(dm)))],
       dm         = dm)
}

marker_expression = function(sce, resolved) {
  m = as.matrix(logcounts(sce)[resolved$gene_id, , drop = FALSE])
  rownames(m) = resolved$symbol
  t(m) %>% as.data.frame() %>%
    tibble::rownames_to_column("barcode") %>%
    tidyr::pivot_longer(-barcode, names_to = "symbol", values_to = "logexpr")
}

trend_table = function(expr_long, axis_df, axis_col) {
  expr_long %>%
    left_join(axis_df, by = "barcode") %>%
    group_by(symbol) %>%
    summarize(n_cells        = dplyr::n(),
              pct_expressing = 100 * mean(logexpr > 0),
              rho            = safe_rho(.data[[axis_col]], logexpr),
              p_value        = safe_p(.data[[axis_col]], logexpr),
              .groups = "drop") %>%
    mutate(FDR = p.adjust(p_value, "BH"),
           direction = case_when(
             is.na(rho)  ~ "undetermined",
             FDR >= 0.05 ~ "flat",
             rho >=  0.3 ~ "increases",
             rho <= -0.3 ~ "decreases",
             TRUE        ~ "weak")) %>%
    left_join(resolved %>% dplyr::select(symbol, family), by = "symbol") %>%
    arrange(desc(abs(rho)))
}

partial_trends = function(expr_long, cdb, axis_col = "DC1",
                          controls = c("sum", "detected")) {
  ctrl = intersect(controls, colnames(cdb))
  if (length(ctrl) == 0) return(NULL)
  j = expr_long %>%
    left_join(cdb %>% dplyr::select(barcode, all_of(axis_col), all_of(ctrl)),
              by = "barcode")
  do.call(rbind, lapply(split(j, j$symbol), function(d) {
    data.frame(symbol = d$symbol[1],
               rho_partial = partial_rho(d[[axis_col]], d$logexpr,
                                         d[, ctrl, drop = FALSE]),
               stringsAsFactors = FALSE)
  }))
}
```

# 1. Embedding

```{r}
embed_file = paste0(Robject_out_dir, "/male_embedding.Rds")

if (file.exists(embed_file)) {
  emb = readRDS(embed_file)
} else {
  message("Computing PCA and diffusion map. Cached in ", embed_file, ".")
  emb = embed_cells(male.sce,
                    block = if (length(male_list) > 1) male.sce$batch else NULL)
  saveRDS(emb, file = embed_file)
}

colnames(emb$dc) = paste0("DC", seq_len(ncol(emb$dc)))
reducedDim(male.sce, "PCA")          = emb$pca
reducedDim(male.sce, "DiffusionMap") = emb$dc

cd = as.data.frame(colData(male.sce))
cd$barcode = colnames(male.sce)
cd$DC1 = emb$dc[, "DC1"]; cd$DC2 = emb$dc[, "DC2"]
cd$PC1 = emb$pca[, 1];    cd$PC2 = emb$pca[, 2]
```

PC1 explains `r round(emb$percentVar[1], 1)`% of variance, PC2
`r round(emb$percentVar[2], 1)`%.

## Technical covariates

```{r}
tech_vars = intersect(c("sum", "detected", "subsets_Mito_percent"), colnames(cd))

tech_tab = data.frame(
  covariate = tech_vars,
  rho_DC1   = vapply(tech_vars, function(v) safe_rho(cd$DC1, cd[[v]]), numeric(1)),
  rho_PC1   = vapply(tech_vars, function(v) safe_rho(cd$PC1, cd[[v]]), numeric(1)),
  stringsAsFactors = FALSE)

write.csv(tech_tab,
          file = paste0(results_dir, "/SuppTable_male_technical.csv"),
          row.names = FALSE)

datatable(tech_tab %>% mutate(across(where(is.numeric), ~round(., 3))),
          rownames = FALSE)

tech_max = max(abs(tech_tab$rho_DC1), na.rm = TRUE)
technical_confound = isTRUE(tech_max >= 0.4)
partial_reliable   = tech_max < 0.6
```

Strongest absolute correlation between DC1 and a technical covariate:
`r round(tech_max, 2)`.

`r if (technical_confound) "**DC1 tracks detection rate.** Expression correlates with depth for nearly every gene, so the raw marker correlations below are inflated. Read the partial correlations instead." else "DC1 is not strongly associated with detection rate."`

## Library effect

```{r}
lib_r2 = if (length(unique(cd$library)) > 1)
  summary(lm(cd$DC1 ~ cd$library))$r.squared else NA_real_
library_confound = isTRUE(lib_r2 >= 0.25)
```

`r if (length(unique(cd$library)) < 2) "All retained male germ cells come from one library, so there is no library effect to test." else paste0("Library of origin explains ", round(100 * lib_r2, 1), "% of the variance in DC1. ", if (library_confound) "The axis is at least partly a library effect." else "The axis is not mainly a library effect.")`

# 2. Cluster order along DC1

The clusters were defined in batch-corrected space, and DC1 is computed on
uncorrected values. If the clusters sit in sequence along DC1, both
representations agree on the order.

```{r}
cluster_pos = cd %>%
  group_by(origin_cluster) %>%
  summarize(n_cells    = dplyr::n(),
            median_DC1 = median(DC1),
            q25        = quantile(DC1, 0.25),
            q75        = quantile(DC1, 0.75),
            .groups    = "drop") %>%
  arrange(median_DC1)

kw = kruskal.test(DC1 ~ factor(origin_cluster), data = cd)

write.csv(cluster_pos,
          file = paste0(results_dir, "/SuppTable_male_cluster_positions.csv"),
          row.names = FALSE)

datatable(cluster_pos %>% mutate(across(where(is.numeric), ~signif(., 3))),
          rownames = FALSE)
```

Kruskal-Wallis across clusters on DC1: p = `r signif(kw$p.value, 3)`.

```{r}
# adjacent-pair overlap: high overlap means the clusters grade into one another,
# low overlap means they are discrete states that happen to be ordered
ord = cluster_pos$origin_cluster
adjacent_overlap = if (length(ord) >= 2) {
  bind_rows(lapply(seq_len(length(ord) - 1), function(i) {
    a = cd$DC1[cd$origin_cluster == ord[i]]
    b = cd$DC1[cd$origin_cluster == ord[i + 1]]
    lo = max(min(a), min(b)); hi = min(max(a), max(b))
    inter = max(0, hi - lo)
    union_w = max(max(a), max(b)) - min(min(a), min(b))
    data.frame(pair = paste(ord[i], "to", ord[i + 1]),
               overlap_fraction = inter / union_w,
               stringsAsFactors = FALSE)
  }))
} else NULL

if (!is.null(adjacent_overlap)) {
  datatable(adjacent_overlap %>% mutate(across(where(is.numeric), ~round(., 3))),
            rownames = FALSE)
}

mean_overlap = if (!is.null(adjacent_overlap))
  mean(adjacent_overlap$overlap_fraction, na.rm = TRUE) else NA_real_
clusters_graded = isTRUE(mean_overlap >= 0.4)
clusters_ordered = isTRUE(kw$p.value < 0.05)
```

`r if (!clusters_ordered) "The clusters do not occupy distinct positions along DC1, so the axis does not match the existing grouping." else if (clusters_graded) paste0("The clusters occupy distinct positions along DC1, and adjacent clusters overlap substantially (mean adjacent overlap ", round(mean_overlap, 2), "). They are ordered and grade into one another. Clustering a continuum into groups produces this pattern.") else paste0("The clusters occupy distinct, well separated positions along DC1 (mean adjacent overlap ", round(mean_overlap, 2), "). They are ordered but do not grade into one another. This pattern fits discrete states better than a continuum.")`

```{r, fig.width=10, fig.height=4}
p_pos = ggplot(cd, aes(x = DC1, fill = origin_cluster, color = origin_cluster)) +
  geom_density(alpha = 0.3) +
  ggtitle("DC1 by originating cluster")

p_scat = ggplot(cd, aes(x = DC1, y = DC2, color = origin_cluster)) +
  geom_point(size = 0.6, alpha = 0.8) +
  guides(color = guide_legend(override.aes = list(size = 3))) +
  ggtitle("Diffusion map by originating cluster")

p_scat + p_pos
ggsave(p_scat + p_pos,
       file = paste0(results_dir, "/male_clusters_on_DC.png"),
       width = 10, height = 4)
```

# 3. Marker behavior along the axis

```{r}
expr_long = marker_expression(male.sce, resolved)
axis_df   = cd %>% dplyr::select(barcode, DC1)

trends = trend_table(expr_long, axis_df, "DC1")
pt     = partial_trends(expr_long, cd)
if (!is.null(pt)) trends = trends %>% left_join(pt, by = "symbol")

write.csv(trends, file = paste0(results_dir, "/SuppTable_male_marker_trends.csv"),
          row.names = FALSE)

datatable(trends %>% mutate(across(where(is.numeric), ~signif(., 3))),
          filter = "top", rownames = FALSE)

n_directional = sum(trends$direction %in% c("increases", "decreases"))
n_directional_partial = if ("rho_partial" %in% colnames(trends))
  sum(abs(trends$rho_partial) >= 0.3, na.rm = TRUE) else NA_integer_
```

`r n_directional` of `r nrow(trends)` markers correlate with DC1, and
`r n_directional_partial` still do after regressing out `sum` and `detected`.
`r if (!partial_reliable) "DC1 is strongly correlated with detection rate, so the partial correlations are underpowered. A low count is expected whether or not biology is present." else "DC1 is not strongly correlated with detection rate, so the partial correlations can be read."`

## Family coherence

```{r}
family_coherence = trends %>%
  filter(!is.na(rho)) %>%
  group_by(family) %>%
  summarize(n_genes       = dplyr::n(),
            median_rho    = median(rho),
            pct_same_sign = 100 * max(sum(rho > 0), sum(rho < 0)) / dplyr::n(),
            n_significant = sum(FDR < 0.05 & abs(rho) >= 0.3),
            .groups = "drop") %>%
  mutate(testable = n_genes >= min_family_genes,
         coherent = ifelse(testable, pct_same_sign >= 80 & n_significant >= 3, NA)) %>%
  arrange(desc(n_genes))

write.csv(family_coherence,
          file = paste0(results_dir, "/SuppTable_male_family_coherence.csv"),
          row.names = FALSE)

datatable(family_coherence %>% mutate(across(where(is.numeric), ~round(., 2))),
          rownames = FALSE)

n_testable = sum(family_coherence$testable)
n_coherent_families = sum(family_coherence$coherent, na.rm = TRUE)
```

## Direction test: early against late

```{r}
dir_summary = trends %>%
  filter(!is.na(rho)) %>%
  mutate(programme = case_when(family %in% late_families  ~ "late (meiotic)",
                               family %in% early_families ~ "early (spermatogonia)",
                               TRUE                       ~ "other")) %>%
  filter(programme != "other") %>%
  group_by(programme) %>%
  summarize(n_genes       = dplyr::n(),
            median_rho    = median(rho),
            n_significant = sum(FDR < 0.05 & abs(rho) >= 0.3),
            .groups = "drop")

datatable(dir_summary %>% mutate(across(where(is.numeric), ~round(., 3))),
          rownames = FALSE)

late_row  = dir_summary %>% filter(programme == "late (meiotic)")
early_row = dir_summary %>% filter(programme == "early (spermatogonia)")
both_present = nrow(late_row) == 1 && nrow(early_row) == 1

opposite_directions = both_present &&
  sign(late_row$median_rho) != sign(early_row$median_rho) &&
  late_row$n_significant >= 3 && early_row$n_significant >= 1

dir_p = if (both_present) {
  a = trends %>% filter(family %in% late_families,  !is.na(rho)) %>% pull(rho)
  b = trends %>% filter(family %in% early_families, !is.na(rho)) %>% pull(rho)
  suppressWarnings(wilcox.test(a, b)$p.value)
} else NA_real_
```

`r if (!both_present) "One programme has no resolved genes, so the direction test cannot be run." else paste0("The late meiotic programme has a median rho of ", round(late_row$median_rho, 2), " across ", late_row$n_genes, " genes (", late_row$n_significant, " significant), and the early spermatogonial programme ", round(early_row$median_rho, 2), " across ", early_row$n_genes, " genes (", early_row$n_significant, " significant). Wilcoxon p = ", signif(dir_p, 2), ".")`

`r if (opposite_directions) "**The two programmes move in opposite directions along DC1.** Spermatogonial markers are high at one end and meiotic markers at the other. That order matches spermatogenesis." else if (both_present) "The two programmes do not move in opposite directions along DC1, so the axis does not show an early to late transition." else ""`

```{r, fig.width=11, fig.height=8}
top_markers = trends %>% filter(!is.na(rho)) %>%
  slice_max(abs(rho), n = 12) %>% pull(symbol)

p_mark = ggplot(expr_long %>% filter(symbol %in% top_markers) %>%
                  left_join(cd %>% dplyr::select(barcode, DC1, origin_cluster),
                            by = "barcode"),
                aes(x = DC1, y = logexpr)) +
  geom_point(aes(color = origin_cluster), size = 0.3, alpha = 0.35) +
  geom_smooth(method = "loess", se = FALSE, color = "black", linewidth = 0.7) +
  guides(color = guide_legend(override.aes = list(size = 3, alpha = 1))) +
  facet_wrap(~ symbol, scales = "free_y") +
  ylab("Uncorrected log-expression")

p_mark
ggsave(p_mark, file = paste0(results_dir, "/male_markers_vs_DC1.png"),
       width = 11, height = 8)
```

# 4. Verdict

```{r}
markers_informative = isTRUE(n_directional_partial >= 3)

male_verdict = if (technical_confound && !partial_reliable) {
  "technical"
} else if (!markers_informative) {
  "no marker signal"
} else if (clusters_ordered && clusters_graded && opposite_directions) {
  "continuum"
} else if (clusters_ordered && opposite_directions) {
  "ordered discrete states"
} else {
  "structure without a programme"
}
```

**`r male_verdict`.**

`r switch(male_verdict,
  "technical" = "DC1 tracks detection rate and the partial correlations cannot separate biology from depth, so no ordering should be built on this axis.",
  "no marker signal" = "No markers vary monotonically along DC1 once depth is controlled, so there is no evidence that the axis corresponds to spermatogenic progression.",
  "continuum" = "The clusters occupy ordered, overlapping positions along an axis computed on uncorrected values, and the spermatogonial and meiotic programmes move in opposite directions along it. These cells form a continuous progression that the clustering has cut into groups.",
  "ordered discrete states" = "The clusters are ordered along the axis and the early and late programmes separate, but the clusters do not grade into one another. They are discrete states in a developmental sequence.",
  "structure without a programme" = "There is structure along DC1, but it does not show an ordered early to late transition. Do not describe it as spermatogenic progression.")`

`r if (single_library) "These cells come from a single library, so the ordering cannot be a batch effect." else "These cells span more than one library, so the axis may include a library effect."`

# 5. Pseudotime

```{r}
order_supported = male_verdict %in% c("continuum", "ordered discrete states")

dpt_file = paste0(Robject_out_dir, "/male_dpt.Rds")

if (order_supported) {
  if (file.exists(dpt_file)) {
    dpt = readRDS(dpt_file)
  } else {
    message("Computing diffusion pseudotime. Cached in ", dpt_file, ".")
    dpt = destiny::DPT(emb$dm)
    saveRDS(dpt, file = dpt_file)
  }
  root_cell = which.min(cd$DC1)
  cd$pseudotime = dpt[[paste0("DPT", root_cell)]]
} else {
  cd$pseudotime = NA_real_
}
```

`r if (!order_supported) "The verdict does not support an ordering, so no pseudotime is computed." else "Diffusion pseudotime is rooted at the cell with the lowest DC1. Which end of the axis is earliest is an assumption."`

```{r}
if (order_supported) {
  ord_tab = cd %>% dplyr::select(barcode, library, origin_cluster, DC1, pseudotime)
  write.csv(ord_tab, file = paste0(results_dir, "/SuppTable_male_ordering.csv"),
            row.names = FALSE)

  pt_trends = trend_table(expr_long,
                          cd %>% dplyr::select(barcode, pseudotime), "pseudotime")
  write.csv(pt_trends,
            file = paste0(results_dir, "/SuppTable_male_trends_pseudotime.csv"),
            row.names = FALSE)

  datatable(pt_trends %>% mutate(across(where(is.numeric), ~signif(., 3))),
            filter = "top", rownames = FALSE)
}
```

```{r, fig.width=10, fig.height=4}
if (order_supported) {
  p_pt = ggplot(cd, aes(x = DC1, y = DC2, color = pseudotime)) +
    geom_point(size = 0.7) + scale_color_viridis_c() +
    ggtitle("Diffusion pseudotime")

  p_pt_cl = ggplot(cd, aes(x = pseudotime, fill = origin_cluster)) +
    geom_density(alpha = 0.35) +
    ggtitle("Pseudotime by originating cluster")

  print(p_pt + p_pt_cl)
  ggsave(p_pt + p_pt_cl, file = paste0(results_dir, "/male_pseudotime.png"),
         width = 10, height = 4)
}
```

# 6. Summary

```{r, results="asis"}
cat(paste0(
  ncol(male.sce), " male germ cells from ", nrow(origin_tab),
  " originating clusters, ", round(dominant_pct, 1), "% from ",
  dominant_library, ". ",
  if (single_library)
    "They are effectively a single library, so the ordering cannot be a batch effect. "
  else
    "They span more than one library, so a library effect cannot be excluded. ",
  "PC1 explains ", round(emb$percentVar[1], 1),
  "% of variance and the strongest correlation between DC1 and a technical ",
  "covariate is ", round(tech_max, 2), ". ",
  "Kruskal-Wallis across the originating clusters on DC1 gives p = ",
  signif(kw$p.value, 3), ", with a mean adjacent overlap of ",
  round(mean_overlap, 2), ". ",
  n_directional, " markers correlate with DC1 and ", n_directional_partial,
  " survive control for depth. ",
  if (both_present) paste0(
    "The meiotic programme has a median rho of ", round(late_row$median_rho, 2),
    " and the spermatogonial programme ", round(early_row$median_rho, 2), ". ")
  else "",
  "Verdict: ", male_verdict, ". ",
  if (male_verdict == "continuum")
    "The cells can be reported as a continuous progression."
  else if (male_verdict == "ordered discrete states")
    "Report them as a sequence of discrete states."
  else
    "Do not report an ordering for these cells."))
```

# Session info

```{r}
sessionInfo()
```
