---
title: "10_four_atlases"
author: "Stephanie Hickey"
date: "`r Sys.Date()`"
output:
  html_document:
    code_folding: hide
    toc: true
    toc_float:
      toc_collapsed: true
params:
    project_dir: /mnt/ufs18/rs-013/bioinformaticsCore/projects/chong_davidson/BCC125_sea_lamprey_scRNA
    project_name: BCC125_sea_lamprey_scRNA
    lab: Weiming Li
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

```{r, echo=FALSE, results="asis"}
cat("\n")
cat("#### Lab:", params$lab, " \n")
cat("#### Project:", params$project_name, " \n")
```

```{r}
library(tidyverse)
library(SingleCellExperiment)
library(scater)
library(scran)
library(patchwork)
library(RColorBrewer)
library(readxl)
library(batchelor)
library(DT)
library(rtracklayer)
library(BiocSingular)
library(pheatmap)
source(paste0(params$project_dir, "/src/scRNA_seq_functions.R"))
```

```{r}
results_dir     = paste0(params$project_dir, "/results/20260812_four_atlases")
Robject_out_dir = paste0(results_dir, "/Robjects")
if (!dir.exists(Robject_out_dir)) dir.create(Robject_out_dir, recursive = TRUE)
data_dir        = paste0(params$project_dir, "/data")

# doublet-filtered per-sample objects, produced by
# 01_all_cells_clustering_annotation_no_doublets.Rmd
no_dbl_dir = paste0(params$project_dir, "/results/20260803_no_doublets")
```

# Purpose

Four atlases are built from the doublet-filtered cells, clustered independently,
and then compared to establish which populations are shared between them.

| atlas | libraries | stage | correction |
| --- | --- | --- | --- |
| Larva | UND1, UND2 | larva | none |
| Transformer | TF, TM | transformer | `fastMNN` |
| Female | TF | transformer | none, single library |
| Male | TM | transformer | none, single library |

Building the stages separately removes the reason the integrated atlas was
contested. Nothing here requires a correction to span the stage boundary,
because no atlas spans it.

## Design decisions, and what they cost

**The larval atlas is not corrected.** UND1 and UND2 interleave at 0.9 to 1.0
uncorrected in every cell type, so there is no batch effect to remove and
applying `fastMNN` would only risk removing real structure. `use_mnn` is a
parameter of the atlas builder, so this can be flipped and rerun if the mixing
diagnostic below disagrees.

**The transformer atlas is corrected, and this is the one place a correction
carries risk.** `fastMNN` assumes the libraries being merged share cell
populations. TF and TM do not fully: oocytes exist only in TF and spermatocytes
only in TM. Mutual nearest neighbor pairs cannot exist for a population with no
counterpart, and forcing an alignment between unmatched populations is exactly
how a correction manufactures a cluster that is not real.

The male and female atlases are the control for this. They contain the same
cells, uncorrected and unmerged, so a transformer cluster that fuses a
male-only and a female-only population will have no reciprocal match in either.
Section 5 reports that comparison directly, and it is the check to read first.

**Male and female cells are subsets of the transformer atlas.** Those three
comparisons therefore share cells and are not independent evidence. They
measure whether the integration preserved the within-library structure. Only
comparisons involving the larval atlas are between disjoint sets of cells, and
only those speak to whether a population recurs across animals and stages.

**Cross-atlas comparison is rank based throughout.** Normalization is computed
within each atlas, so absolute expression is not comparable between atlases.
Spearman correlation of cluster centroids and `SingleR`, which also ranks
within cell, are both invariant to that. No cross-atlas claim below rests on a
difference in absolute level.

# Load the doublet-filtered samples

```{r}
all.sce = readRDS(paste0(no_dbl_dir, "/all.sce.Rds"))

library_meta = data.frame(
  batch   = c("UND1_C00004336", "UND2_C00004362", "TF_C00004337", "TM_C00004374"),
  library = c("Undifferentiated.1", "Undifferentiated.2", "Female", "Male"),
  stage   = c("larva", "larva", "transformer", "transformer"),
  stringsAsFactors = FALSE)

observed   = sort(names(all.sce))
unexpected = setdiff(observed, library_meta$batch)
missing    = setdiff(library_meta$batch, observed)
```

* Samples present: `r paste(observed, collapse = ", ")`
* In the objects but not in the mapping: `r if (length(unexpected) > 0) paste(unexpected, collapse = ", ") else "none"`
* In the mapping but not in the objects: `r if (length(missing) > 0) paste(missing, collapse = ", ") else "none"`

```{r}
if (length(unexpected) > 0 || length(missing) > 0) {
  stop("Sample identifiers do not match library_meta. Resolve before proceeding ",
       "rather than silently dropping cells.")
}

atlas_def = list(
  Larva       = list(batches = c("UND1_C00004336", "UND2_C00004362"), use_mnn = FALSE),
  Transformer = list(batches = c("TF_C00004337", "TM_C00004374"),     use_mnn = TRUE),
  Female      = list(batches = "TF_C00004337",                        use_mnn = FALSE),
  Male        = list(batches = "TM_C00004374",                        use_mnn = FALSE)
)

datatable(library_meta, rownames = FALSE)
```

# 1. Build the four atlases

Each atlas is normalized across its own libraries with `multiBatchNorm`, so
expression is comparable within an atlas. Highly variable genes are selected on
the combined variance model. Clustering is Louvain on a shared nearest neighbor
graph with `k = 30`, matching `01`, so cluster granularity is comparable to the
existing analysis.

Two objects are kept per atlas: the clustering object, which for a corrected
atlas holds the `fastMNN` output, and an uncorrected expression object holding
`logcounts` for every gene. All quantification below uses the second.

```{r}
build_atlas = function(sce_list, batches, use_mnn, seed = 1, k = 30,
                       n_hvg = 5000) {

  sub      = sce_list[batches]
  universe = Reduce(intersect, lapply(sub, rownames))
  sub      = lapply(sub, "[", i = universe, )

  # strip per-sample extras so cbind does not depend on which sample carries
  # which optional colData field
  sub = lapply(sub, function(x) {
    colData(x)     = colData(x)[, "sample", drop = FALSE]
    reducedDims(x) = list()
    rowData(x)     = rowData(x)[, character(0)]
    x
  })

  dec = lapply(sub, modelGeneVar)

  if (length(sub) > 1) {
    normed = do.call(multiBatchNorm, sub)
    hvg    = getTopHVGs(do.call(combineVar, dec), n = n_hvg)
  } else {
    normed = sub
    hvg    = getTopHVGs(dec[[1]], n = n_hvg)
  }

  expr        = do.call(cbind, normed)
  expr$batch  = expr$sample

  if (use_mnn && length(normed) > 1) {
    set.seed(seed)
    clust = do.call(fastMNN, c(normed,
              list(subset.row = hvg, BSPARAM = RandomParam())))
    dimred = "corrected"
  } else {
    set.seed(seed)
    clust  = runPCA(expr, subset_row = hvg, ncomponents = 50,
                    BSPARAM = RandomParam())
    dimred = "PCA"
  }

  # cbind preserves order, so the two objects stay aligned. Assert it anyway.
  stopifnot(identical(colnames(clust), colnames(expr)))

  set.seed(seed)
  g = buildSNNGraph(clust, use.dimred = dimred, k = k)
  set.seed(seed)
  colLabels(clust) = factor(igraph::cluster_louvain(g)$membership)

  set.seed(seed)
  clust = runTSNE(clust, dimred = dimred)
  set.seed(seed)
  clust = runUMAP(clust, dimred = dimred)

  clust$batch  = expr$batch
  expr$label   = clust$label

  list(clust = clust, expr = expr, dimred = dimred, hvg = hvg,
       corrected = use_mnn && length(normed) > 1)
}

# cache; the chunk that computes also writes, so a rerun never deadlocks
atlas_file = paste0(Robject_out_dir, "/atlases.Rds")

if (file.exists(atlas_file)) {
  atlases = readRDS(atlas_file)
} else {
  atlases = lapply(names(atlas_def), function(nm) {
    build_atlas(all.sce,
                batches = atlas_def[[nm]]$batches,
                use_mnn = atlas_def[[nm]]$use_mnn)
  })
  names(atlases) = names(atlas_def)
  saveRDS(atlases, file = atlas_file)
}

atlas_names = names(atlases)
```

```{r}
atlas_summary = purrr::map_dfr(atlas_names, function(nm) {
  a = atlases[[nm]]
  data.frame(atlas      = nm,
             libraries  = paste(atlas_def[[nm]]$batches, collapse = ", "),
             corrected  = a$corrected,
             nCells     = ncol(a$clust),
             nClusters  = nlevels(a$clust$label),
             nHVG       = length(a$hvg))
})

write.csv(atlas_summary, file = paste0(results_dir, "/atlas_summary.csv"),
          row.names = FALSE)
datatable(atlas_summary, rownames = FALSE)
```

## Cluster composition by library

For the two multi-library atlases, a cluster drawn overwhelmingly from one
library is either a genuinely library-restricted population or a failure of the
merge. For the transformer atlas the distinction is the whole question, and it
is resolved in section 5, not here.

```{r, fig.width=9, fig.height=4}
comp_df = purrr::map_dfr(atlas_names, function(nm) {
  as.data.frame(colData(atlases[[nm]]$clust)) %>%
    dplyr::count(label, batch, name = "n") %>%
    mutate(atlas = nm)
}) %>%
  left_join(library_meta %>% dplyr::select(batch, library), by = "batch")

write.csv(comp_df, file = paste0(results_dir, "/cells_per_cluster_per_library.csv"),
          row.names = FALSE)

ggplot(comp_df %>% filter(atlas %in% c("Larva", "Transformer")),
       aes(fill = library, y = label, x = n)) +
  geom_bar(position = "fill", stat = "identity") +
  facet_wrap(~ atlas, scales = "free_y") +
  xlab("Proportion of cells") + ylab("Cluster")

ggsave(file = paste0(results_dir, "/cluster_composition_by_library.png"),
       width = 9, height = 4)
```

```{r}
datatable(comp_df %>%
            tidyr::pivot_wider(id_cols = c(atlas, label),
                               names_from = library, values_from = n,
                               values_fill = 0),
          filter = "top", rownames = FALSE)
```

## Embeddings

```{r, fig.width=11, fig.height=10}
p = lapply(atlas_names, function(nm) {
  plotTSNE(atlases[[nm]]$clust, colour_by = "label", text_by = "label",
           text_colour = "red", point_size = .05) +
    ggtitle(paste0(nm, if (atlases[[nm]]$corrected) " (fastMNN)" else " (uncorrected)")) +
    guides(color = guide_legend(override.aes = list(size = 4)))
})
wrap_plots(p, ncol = 2)

ggsave(file = paste0(results_dir, "/four_atlas_tsne.png"), width = 11, height = 10)
```

# 2. Marker genes per atlas

Markers are computed within each atlas, blocked on library where an atlas has
more than one, so no marker result depends on the correction. Ordering is set
explicitly by FDR then p-value rather than relying on the order `findMarkers`
returns, and the per-gene effect size is the mean across the pairwise
`logFC.*` columns. `summary.logFC` is not used: under `pval.type = "some"` it
comes from whichever pairwise comparison supplied the summary p-value, which
varies per gene and per run.

```{r}
marker_effect = function(df) {
  lf = grep("^logFC\\.", colnames(df), value = TRUE)
  if (length(lf) == 0) return(rep(NA_real_, nrow(df)))
  rowMeans(as.matrix(as.data.frame(df[, lf, drop = FALSE])), na.rm = TRUE)
}

marker_file = paste0(Robject_out_dir, "/atlas_markers.Rds")

if (file.exists(marker_file)) {
  atlas_markers = readRDS(marker_file)
} else {
  atlas_markers = lapply(atlas_names, function(nm) {
    e     = atlases[[nm]]$expr
    blk   = if (length(unique(e$batch)) > 1) e$batch else NULL
    fm    = findMarkers(e, e$label, test.type = "t", block = blk,
                        direction = "up", pval.type = "some")
    lapply(fm, function(df) {
      df$meanLogFC = marker_effect(df)
      df[order(df$FDR, df$p.value), ]
    })
  })
  names(atlas_markers) = atlas_names
  saveRDS(atlas_markers, file = marker_file)
}

top_markers = function(nm, lab, n = 100) {
  df = atlas_markers[[nm]][[as.character(lab)]]
  if (is.null(df) || nrow(df) == 0) return(character(0))
  rownames(df)[seq_len(min(n, nrow(df)))]
}
```

```{r}
marker_tbl = purrr::map_dfr(atlas_names, function(nm) {
  purrr::map_dfr(names(atlas_markers[[nm]]), function(lab) {
    df = as.data.frame(atlas_markers[[nm]][[lab]])
    df$Gene = rownames(df)
    df %>%
      filter(FDR < .01) %>%
      dplyr::select(Gene, p.value, FDR, meanLogFC) %>%
      mutate(atlas = nm, cluster = lab)
  })
})

write.csv(marker_tbl, file = paste0(results_dir, "/atlas_cluster_markers.csv"),
          row.names = FALSE)

datatable(marker_tbl %>% group_by(atlas, cluster) %>% slice_head(n = 20),
          filter = "top", rownames = FALSE)
```

## Overlap with curated cell type markers

The hypergeometric overlap from `01`, run once per atlas: rows are curated cell
types, columns are that atlas's clusters, color is the overlap significance and
the number in each tile is the count of shared genes.

The curated list is `canonical_markers_long_2026-08-12.csv`, the repaired long
form, not the S1 sheet. Three of the four erythrocyte entries and the oocyte
`ZAR1` entry in S1 resolve against nothing, so a heatmap built on S1 silently
tests fewer genes than it reports.

Two cautions on reading these. The overlap is against each cluster's top 100
enriched genes, so it measures relative enrichment, not expression; the bubble
plots below measure expression and the two can disagree. And **the four
heatmaps are not comparable to each other**: a cluster competing against 11
others yields a different marker list than one competing against 15, so overlap
counts are biased by cluster number. Compare rows within a heatmap, not tiles
between heatmaps.

```{r}
gene_description = read.csv(paste0(no_dbl_dir, "/gene_descritption.csv"),
                            stringsAsFactors = FALSE)

canonical = read.csv(paste0(data_dir, "/canonical_markers_long_2026-08-12.csv"),
                     stringsAsFactors = FALSE) %>%
  mutate(symbol = trimws(symbol), loc_id = trimws(loc_id)) %>%
  filter(symbol != "")

known_markers = canonical %>%
  dplyr::select(Gene = loc_id, CellType = CellTypeMarkerSet) %>%
  distinct()

datatable(canonical %>% dplyr::count(CellTypeMarkerSet, name = "n_markers"),
          rownames = FALSE)
```

```{r, fig.width=8, fig.height=6, results="asis"}
overlap_plots = list()

for (nm in atlas_names) {
  test_genes = marker_tbl %>%
    filter(atlas == nm) %>%
    group_by(cluster) %>%
    slice_head(n = 100) %>%
    ungroup() %>%
    dplyr::select(Gene, cluster) %>%
    mutate(cluster = paste0("cluster", cluster))

  # nrow is the gene universe the hypergeometric needs, stated explicitly
  overlap_plots[[nm]] = plot_marker_overlap_heatmap(
    known_markers    = known_markers,
    test_genes       = as.data.frame(test_genes),
    background_n     = nrow(atlases[[nm]]$expr),
    gene_description = gene_description,
    overlaps_file    = paste0(results_dir, "/", nm, "_marker_gene_overlaps.csv"),
    pvals_file       = paste0(results_dir, "/", nm, "_marker_overlap_pvalues.csv"),
    plot_file        = paste0(results_dir, "/", nm, "_marker_overlap_heatmap.png"))
}
```

```{r, fig.width=8, fig.height=5}
overlap_plots[["Larva"]] + ggtitle("Larva")
```

```{r, fig.width=8, fig.height=5}
overlap_plots[["Transformer"]] + ggtitle("Transformer")
```

```{r, fig.width=8, fig.height=5}
overlap_plots[["Female"]] + ggtitle("Female")
```

```{r, fig.width=8, fig.height=5}
overlap_plots[["Male"]] + ggtitle("Male")
```

## Canonical marker expression per cluster

The same curated genes measured as expression rather than enrichment, in the
style of the dot plot in `06`. Dot size is the percentage of cells in the
cluster expressing the gene and color is the mean log expression, computed on
uncorrected values within each atlas.

This is the plot to annotate from. A cluster that scores a significant overlap
above but shows no expression here is enriched for a marker it barely
transcribes, which is how cluster 7 in the original analysis came to be called
germline.

```{r}
canon_bubble = function(nm) {
  e   = atlases[[nm]]$expr
  ids = intersect(unique(canonical$loc_id), rownames(e))
  if (length(ids) == 0) return(NULL)

  m   = as.matrix(logcounts(e)[ids, , drop = FALSE])
  lab = as.character(e$label)

  purrr::map_dfr(sort(unique(lab)), function(k) {
    sub = m[, lab == k, drop = FALSE]
    data.frame(atlas          = nm,
               cluster        = k,
               loc_id         = ids,
               n_cells        = ncol(sub),
               pct_expressing = 100 * rowMeans(sub > 0),
               mean_logexpr   = rowMeans(sub),
               stringsAsFactors = FALSE)
  }) %>%
    left_join(canonical, by = "loc_id", relationship = "many-to-many")
}

bubble_df = purrr::map_dfr(atlas_names, canon_bubble) %>%
  mutate(cluster = factor(cluster, levels = sort(unique(as.numeric(cluster)))))

write.csv(bubble_df,
          file = paste0(results_dir, "/canonical_marker_expression_per_cluster.csv"),
          row.names = FALSE)
```

```{r}
plot_bubble = function(nm) {
  ggplot(bubble_df %>% filter(atlas == nm),
         aes(x = cluster, y = symbol,
             size = pct_expressing, color = mean_logexpr)) +
    geom_point() +
    scale_color_viridis_c() +
    scale_size_continuous(range = c(0.5, 5)) +
    facet_grid(CellTypeMarkerSet ~ ., scales = "free_y", space = "free_y") +
    theme(strip.text.y = element_text(angle = 0, size = 6),
          axis.text.y  = element_text(size = 6),
          axis.text.x  = element_text(size = 7)) +
    xlab("Cluster") + ylab(NULL) + ggtitle(nm) +
    labs(size = "% expressing", color = "mean logexpr")
}

for (nm in atlas_names) {
  ggsave(plot_bubble(nm),
         file   = paste0(results_dir, "/", nm, "_canonical_marker_bubble.png"),
         width  = 9, height = 11)
}
```

```{r, fig.width=9, fig.height=11}
plot_bubble("Larva")
```

```{r, fig.width=9, fig.height=11}
plot_bubble("Transformer")
```

```{r, fig.width=9, fig.height=11}
plot_bubble("Female")
```

```{r, fig.width=9, fig.height=11}
plot_bubble("Male")
```

# 3. Cluster centroid correlation

The first of two comparison methods. Cluster centroids are averaged
`logcounts` over genes shared by both atlases and variable in either, and
compared by Spearman correlation, which is invariant to the per-atlas
normalization.

A correlation is meaningless without a reference, so two are computed the same
way. The **ceiling** is each cluster against itself across a random split of
its own atlas into halves, which is the best agreement achievable given
sampling noise. The **floor** is the distribution of correlations between
non-matching cluster pairs.

```{r}
centroids_for = function(nm, genes, cells = NULL) {
  e = atlases[[nm]]$expr
  if (!is.null(cells)) e = e[, cells]
  # summarizeAssayByGroup keeps the matrix sparse, which matters at 5000 genes
  # by tens of thousands of cells. It returns genes by groups.
  agg = scuttle::summarizeAssayByGroup(logcounts(e)[genes, , drop = FALSE],
                                       ids = droplevels(factor(e$label)),
                                       statistics = "mean")
  as.matrix(SummarizedExperiment::assay(agg, "mean"))
}

shared_genes = function(a, b) {
  g = intersect(rownames(atlases[[a]]$expr), rownames(atlases[[b]]$expr))
  v = union(atlases[[a]]$hvg, atlases[[b]]$hvg)
  intersect(g, v)
}

cor_matrix = function(a, b) {
  g  = shared_genes(a, b)
  ca = centroids_for(a, g)
  cb = centroids_for(b, g)
  m  = cor(ca, cb, method = "spearman")
  dimnames(m) = list(colnames(ca), colnames(cb))
  m
}

atlas_pairs = t(combn(atlas_names, 2))
cor_list = setNames(
  lapply(seq_len(nrow(atlas_pairs)), function(i)
    cor_matrix(atlas_pairs[i, 1], atlas_pairs[i, 2])),
  paste(atlas_pairs[, 1], atlas_pairs[, 2], sep = " vs "))
```

```{r}
# ceiling: each cluster against itself across a random half split
set.seed(1)
self_ceiling = purrr::map_dfr(atlas_names, function(nm) {
  e     = atlases[[nm]]$expr
  g     = intersect(rownames(e), atlases[[nm]]$hvg)
  half  = sample(c(TRUE, FALSE), ncol(e), replace = TRUE)

  keep = names(which(table(e$label[half]) >= 10 & table(e$label[!half]) >= 10))
  if (length(keep) == 0)
    return(data.frame(atlas = nm, cluster = NA_character_, rho = NA_real_))

  c1 = centroids_for(nm, g, cells = which(half))
  c2 = centroids_for(nm, g, cells = which(!half))
  keep = intersect(keep, intersect(colnames(c1), colnames(c2)))

  data.frame(atlas = nm, cluster = keep,
             rho = vapply(keep, function(k)
               cor(c1[, k], c2[, k], method = "spearman"), numeric(1)))
})

ceiling_med = median(self_ceiling$rho, na.rm = TRUE)
```

Median within-atlas split-half correlation, the ceiling: **`r round(ceiling_med, 3)`**.

```{r, fig.width=10, fig.height=8}
for (nm in names(cor_list)) {
  pheatmap(cor_list[[nm]], main = nm, fontsize = 8,
           color = colorRampPalette(rev(brewer.pal(9, "RdYlBu")))(50),
           filename = paste0(results_dir, "/centroid_cor_",
                             gsub(" ", "_", nm), ".png"),
           width = 7, height = 6)
}

pheatmap(cor_list[["Larva vs Transformer"]],
         main = "Larva vs Transformer, centroid Spearman", fontsize = 8,
         color = colorRampPalette(rev(brewer.pal(9, "RdYlBu")))(50))
```

# 4. Label transfer

The second method, and independent of the first. `SingleR` is trained on one
atlas's clusters and used to classify the other atlas's cells, in both
directions. The reference is downsampled to at most 500 cells per cluster so
that large clusters do not dominate.

A cluster pair is called a **reciprocal best match** when each is the other's
most frequently assigned label. A one-directional match is reported but is
weaker: a small cluster can be the best match for a large one without the
reverse holding.

```{r}
have_singler = requireNamespace("SingleR", quietly = TRUE)
```

`r if (have_singler) "SingleR is available; label transfer runs below." else "**SingleR is not installed, so this section is skipped.** Install it with BiocManager::install('SingleR') and rerun. The centroid correlation in section 3 and the marker overlap in section 6 do not depend on it, so the document still produces a comparison without it, on two methods rather than three."`

```{r, eval=have_singler}
downsample_ref = function(nm, max_per = 500) {
  e = atlases[[nm]]$expr
  set.seed(1)
  idx = unlist(lapply(split(seq_len(ncol(e)), e$label), function(i)
    if (length(i) > max_per) sample(i, max_per) else i))
  e[, sort(idx)]
}

transfer_file = paste0(Robject_out_dir, "/label_transfer.Rds")

if (file.exists(transfer_file)) {
  transfer = readRDS(transfer_file)
} else {
  transfer = list()
  for (i in seq_len(nrow(atlas_pairs))) {
    a = atlas_pairs[i, 1]; b = atlas_pairs[i, 2]
    g = intersect(rownames(atlases[[a]]$expr), rownames(atlases[[b]]$expr))

    ref_a = downsample_ref(a); ref_b = downsample_ref(b)

    # the assay types are named explicitly: these objects carry counts as the
    # first assay, so relying on positional defaults would classify on raw
    # counts without erroring
    pred_ab = SingleR::SingleR(test = atlases[[b]]$expr[g, ],
                               ref  = ref_a[g, ],
                               labels = ref_a$label, de.method = "wilcox",
                               assay.type.test = "logcounts",
                               assay.type.ref  = "logcounts")
    pred_ba = SingleR::SingleR(test = atlases[[a]]$expr[g, ],
                               ref  = ref_b[g, ],
                               labels = ref_b$label, de.method = "wilcox",
                               assay.type.test = "logcounts",
                               assay.type.ref  = "logcounts")

    transfer[[paste(a, b, sep = " vs ")]] =
      list(a = a, b = b,
           ab = data.frame(target = as.character(atlases[[b]]$expr$label),
                           predicted = pred_ab$labels),
           ba = data.frame(target = as.character(atlases[[a]]$expr$label),
                           predicted = pred_ba$labels))
  }
  saveRDS(transfer, file = transfer_file)
}
```

```{r, eval=have_singler}
best_match = function(df) {
  df %>%
    dplyr::count(target, predicted, name = "n") %>%
    group_by(target) %>%
    mutate(frac = n / sum(n)) %>%
    slice_max(n, n = 1, with_ties = FALSE) %>%
    ungroup()
}

reciprocal = purrr::map_dfr(names(transfer), function(nm) {
  tr = transfer[[nm]]
  m_ab = best_match(tr$ab) %>%
    dplyr::rename(b_cluster = target, a_best = predicted,
                  n_ab = n, frac_ab = frac)
  m_ba = best_match(tr$ba) %>%
    dplyr::rename(a_cluster = target, b_best = predicted,
                  n_ba = n, frac_ba = frac)

  m_ab %>%
    inner_join(m_ba, by = c("b_cluster" = "b_best", "a_best" = "a_cluster")) %>%
    mutate(pair = nm, atlas_a = tr$a, atlas_b = tr$b) %>%
    dplyr::rename(a_cluster = a_best) %>%
    dplyr::select(pair, atlas_a, a_cluster, atlas_b, b_cluster,
                  frac_ab, frac_ba)
})

write.csv(reciprocal,
          file = paste0(results_dir, "/reciprocal_best_matches.csv"),
          row.names = FALSE)

datatable(reciprocal %>% mutate(across(where(is.numeric), ~round(., 3))),
          filter = "top", rownames = FALSE)
```

# 5. Did the transformer correction invent anything?

The check to read first. Every transformer cluster is compared to the male and
female atlases, which contain the same cells with no correction applied.

A transformer cluster is **supported** when it has a reciprocal best match in
the male atlas, the female atlas, or both. A transformer cluster with no
reciprocal match in either, despite being built from exactly those cells, is a
candidate artifact of the merge: `fastMNN` grouped cells that neither library
groups on its own.

The converse is also reported. A male or female cluster that survives into the
transformer atlas is a population the correction preserved; one that does not
was dissolved by it.

```{r, eval=have_singler}
tr_support = reciprocal %>%
  filter((atlas_a == "Transformer" & atlas_b %in% c("Male", "Female")) |
         (atlas_b == "Transformer" & atlas_a %in% c("Male", "Female"))) %>%
  mutate(tr_cluster  = ifelse(atlas_a == "Transformer", a_cluster, b_cluster),
         other_atlas = ifelse(atlas_a == "Transformer", atlas_b, atlas_a),
         other_clust = ifelse(atlas_a == "Transformer", b_cluster, a_cluster))

all_tr = levels(atlases[["Transformer"]]$clust$label)

tr_verdict = data.frame(tr_cluster = all_tr, stringsAsFactors = FALSE) %>%
  left_join(tr_support %>%
              group_by(tr_cluster) %>%
              summarize(matched_in = paste(sort(unique(other_atlas)),
                                           collapse = ", "),
                        matches    = paste(paste0(other_atlas, ":", other_clust),
                                           collapse = "; "),
                        .groups = "drop"),
            by = "tr_cluster") %>%
  mutate(matched_in = ifelse(is.na(matched_in), "neither", matched_in),
         supported  = matched_in != "neither") %>%
  left_join(comp_df %>% filter(atlas == "Transformer") %>%
              tidyr::pivot_wider(id_cols = label, names_from = library,
                                 values_from = n, values_fill = 0) %>%
              mutate(label = as.character(label)),
            by = c("tr_cluster" = "label"))

write.csv(tr_verdict,
          file = paste0(results_dir, "/transformer_cluster_support.csv"),
          row.names = FALSE)

datatable(tr_verdict, filter = "top", rownames = FALSE)

n_unsupported = sum(!tr_verdict$supported)
```

`r if (!have_singler) "Not evaluated, since SingleR is unavailable." else if (n_unsupported == 0) "**Every transformer cluster has a reciprocal match in the male atlas, the female atlas, or both.** The correction did not create a population that neither library supports on its own, so the transformer atlas can be presented as an integrated object." else paste0("**", n_unsupported, " transformer cluster(s) have no reciprocal match in either the male or the female atlas: ", paste(tr_verdict$tr_cluster[!tr_verdict$supported], collapse = ", "), ".** These are built from male and female cells but are not recovered when either library is clustered alone, which is the signature of a correction-induced grouping. Inspect their library composition in the table above: a cluster drawn from both libraries in roughly equal parts, with no counterpart in either, should not be reported as a cell population.")`

# 6. Shared clusters across atlases

The final table. A cluster pair is called shared when both methods agree: a
reciprocal best match from label transfer, and a centroid correlation at or
above the split-half ceiling reported in section 3. Requiring both means a
pair carried by one method alone is reported but not called.

```{r}
cor_lookup = function(a, b, ca, cb) {
  key = paste(a, b, sep = " vs ")
  if (key %in% names(cor_list)) {
    m = cor_list[[key]]
    if (ca %in% rownames(m) && cb %in% colnames(m)) return(m[ca, cb])
  }
  key = paste(b, a, sep = " vs ")
  if (key %in% names(cor_list)) {
    m = cor_list[[key]]
    if (cb %in% rownames(m) && ca %in% colnames(m)) return(m[cb, ca])
  }
  NA_real_
}
```

```{r, eval=have_singler}
shared = reciprocal %>%
  rowwise() %>%
  mutate(rho = cor_lookup(atlas_a, atlas_b, a_cluster, b_cluster)) %>%
  ungroup() %>%
  # only the transformer atlas shares cells with male and female. Female
  # against male is between different animals and so is independent.
  mutate(independent = !((atlas_a == "Transformer" &
                          atlas_b %in% c("Male", "Female")) |
                         (atlas_b == "Transformer" &
                          atlas_a %in% c("Male", "Female"))),
         cor_supports = !is.na(rho) & rho >= ceiling_med,
         call = dplyr::case_when(
           cor_supports  ~ "shared",
           is.na(rho)    ~ "label transfer only, correlation unavailable",
           TRUE          ~ "label transfer only, correlation below ceiling")) %>%
  arrange(desc(independent), pair, desc(rho))

write.csv(shared, file = paste0(results_dir, "/shared_clusters.csv"),
          row.names = FALSE)

datatable(shared %>% mutate(across(where(is.numeric), ~round(., 3))),
          filter = "top", rownames = FALSE)
```

`r if (!have_singler) "Not evaluated, since SingleR is unavailable. Use the centroid correlation heatmaps in section 3 as a single-method comparison until it is installed." else paste0("Of ", nrow(shared), " reciprocal best matches, ", sum(shared$call == "shared"), " are also supported by centroid correlation and are called shared. ", sum(shared$independent), " of the reciprocal matches are between disjoint sets of cells, meaning they involve the larval atlas; the remainder involve male or female against transformer and share cells, so they measure the integration rather than replication of a population.")`

## Larval populations with no transformer counterpart, and the reverse

```{r, eval=have_singler}
lt = shared %>% filter(pair == "Larva vs Transformer", call == "shared")

larva_unmatched = setdiff(levels(atlases[["Larva"]]$clust$label),
                          lt$a_cluster)
trans_unmatched = setdiff(levels(atlases[["Transformer"]]$clust$label),
                          lt$b_cluster)

data.frame(
  atlas   = c("Larva", "Transformer"),
  n_total = c(nlevels(atlases[["Larva"]]$clust$label),
              nlevels(atlases[["Transformer"]]$clust$label)),
  n_shared = c(length(unique(lt$a_cluster)), length(unique(lt$b_cluster))),
  unmatched = c(paste(larva_unmatched, collapse = ", "),
                paste(trans_unmatched, collapse = ", "))) %>%
  datatable(rownames = FALSE)
```

A cluster with no counterpart is a stage-restricted population, which is a
result rather than a failure. Oocytes and spermatocytes should appear here on
the transformer side, since larvae have undifferentiated gonads. A somatic
cluster appearing here is more interesting and should be examined directly.

# Session info

```{r}
sessionInfo()
```
