---
title: "10_four_atlases"
author: "Stephanie Hickey"
date: "`r Sys.Date()`"
output:
  html_document:
    code_folding: hide
    toc: true
    toc_float:
      toc_collapsed: true
params:
    project_dir: /mnt/ufs18/rs-013/bioinformaticsCore/projects/chong_davidson/BCC125_sea_lamprey_scRNA
    project_name: BCC125_sea_lamprey_scRNA
    lab: Weiming Li
    n_workers: 8
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

```{r, echo=FALSE, results="asis"}
cat("\n")
cat("#### Lab:", params$lab, " \n")
cat("#### Project:", params$project_name, " \n")
```

```{r}
library(tidyverse)
library(SingleCellExperiment)
library(scater)
library(scran)
library(patchwork)
library(RColorBrewer)
library(readxl)
library(batchelor)
library(DT)
library(rtracklayer)
library(BiocSingular)
source(paste0(params$project_dir, "/src/scRNA_seq_functions.R"))
```

```{r}
results_dir     = paste0(params$project_dir, "/results/20260812_four_atlases")
Robject_out_dir = paste0(results_dir, "/Robjects")
if (!dir.exists(Robject_out_dir)) dir.create(Robject_out_dir, recursive = TRUE)
data_dir        = paste0(params$project_dir, "/data")

# doublet-filtered per-sample objects, produced by
# 01_all_cells_clustering_annotation_no_doublets.Rmd
no_dbl_dir = paste0(params$project_dir, "/results/20260803_no_doublets")
```

```{r}
# Parallel backend for the expensive steps, the SingleR runs above all.
#
# On SLURM the allocation is what matters, not the node's core count, so
# SLURM_CPUS_PER_TASK wins over the n_workers parameter when it is set.
# Requesting more workers than the job was allocated oversubscribes the node
# and usually runs slower than serial.
slurm_cpus = suppressWarnings(as.integer(Sys.getenv("SLURM_CPUS_PER_TASK")))

n_workers = if (!is.na(slurm_cpus) && slurm_cpus > 0) slurm_cpus else
  as.integer(params$n_workers)
n_workers = max(1, min(n_workers, parallel::detectCores(), na.rm = TRUE))

bp = if (n_workers > 1) BiocParallel::MulticoreParam(n_workers) else
  BiocParallel::SerialParam()
```

Parallel workers in use: **`r n_workers`**`r if (!is.na(slurm_cpus) && slurm_cpus > 0) ", taken from SLURM_CPUS_PER_TASK" else ", taken from the n_workers parameter"`.

`MulticoreParam` forks, so it will not work on Windows; set `n_workers: 1` there
and it falls back to `SerialParam`. If the machine also has a threaded BLAS,
workers and BLAS threads can oversubscribe each other, in which case setting
`OMP_NUM_THREADS=1` before starting R is usually faster than tuning either
alone.

# Purpose

Four atlases are built from the doublet-filtered cells, clustered independently,
and then compared to establish which populations are shared between them.

```{r, echo=FALSE}
datatable(
  data.frame(
    atlas      = c("Larva", "Transformer", "Female", "Male"),
    libraries  = c("UND1, UND2", "TF, TM", "TF", "TM"),
    stage      = c("larva", "transformer", "transformer", "transformer"),
    correction = c("none", "fastMNN", "none, single library",
                   "none, single library"),
    stringsAsFactors = FALSE),
  rownames = FALSE,
  options  = list(dom = "t", ordering = FALSE))
```

Building the stages separately removes the reason the integrated atlas was
contested. Nothing here requires a correction to span the stage boundary,
because no atlas spans it.

## Design decisions, and what they cost

**The larval atlas is not corrected.** UND1 and UND2 interleave at 0.9 to 1.0
uncorrected in every cell type, so there is no batch effect to remove and
applying `fastMNN` would only risk removing real structure. `use_mnn` is a
parameter of the atlas builder, so this can be flipped and rerun if the mixing
diagnostic below disagrees.

**The transformer atlas is corrected, and this is the one place a correction
carries risk.** `fastMNN` assumes the libraries being merged share cell
populations. TF and TM do not fully: oocytes exist only in TF and spermatocytes
only in TM. Mutual nearest neighbor pairs cannot exist for a population with no
counterpart, and forcing an alignment between unmatched populations is exactly
how a correction manufactures a cluster that is not real.

The male and female atlases are the control for this. They contain the same
cells, uncorrected and unmerged, so a transformer cluster that fuses a
male-only and a female-only population will have no reciprocal match in either.
Section 5 reports that comparison directly, and it is the check to read first.

**Male and female cells are subsets of the transformer atlas.** Those three
comparisons therefore share cells and are not independent evidence. They
measure whether the integration preserved the within-library structure. Only
comparisons involving the larval atlas are between disjoint sets of cells, and
only those speak to whether a population recurs across animals and stages.

**Neither comparison uses absolute expression.** Normalization is computed
within each atlas, so expression levels are not comparable between atlases.
`SingleR` ranks genes within each cell, and the marker overlap uses only which
genes are markers, not how highly they are expressed. No cross-atlas claim
below rests on a difference in absolute level.

**The two methods are not fully independent, and the document does not claim
they are.** `SingleR` scores a cell by Spearman correlation against the
reference over marker genes, so it is a correlation method. The marker overlap
asks a different question, whether two clusters are distinguished by the same
genes, which is set membership rather than correlation of levels. They share
less than two correlation methods would, but they are both computed from the
same marker structure, so agreement between them is corroboration rather than
independent confirmation.

# Load the doublet-filtered samples

```{r}
all.sce = readRDS(paste0(no_dbl_dir, "/all.sce.Rds"))

library_meta = data.frame(
  batch   = c("UND1_C00004336", "UND2_C00004362", "TF_C00004337", "TM_C00004374"),
  library = c("Undifferentiated.1", "Undifferentiated.2", "Female", "Male"),
  stage   = c("larva", "larva", "transformer", "transformer"),
  stringsAsFactors = FALSE)

observed   = sort(names(all.sce))
unexpected = setdiff(observed, library_meta$batch)
missing    = setdiff(library_meta$batch, observed)
```

* Samples present: `r paste(observed, collapse = ", ")`
* In the objects but not in the mapping: `r if (length(unexpected) > 0) paste(unexpected, collapse = ", ") else "none"`
* In the mapping but not in the objects: `r if (length(missing) > 0) paste(missing, collapse = ", ") else "none"`

```{r}
if (length(unexpected) > 0 || length(missing) > 0) {
  stop("Sample identifiers do not match library_meta. Resolve before proceeding ",
       "rather than silently dropping cells.")
}

atlas_def = list(
  Larva       = list(batches = c("UND1_C00004336", "UND2_C00004362"), use_mnn = FALSE),
  Transformer = list(batches = c("TF_C00004337", "TM_C00004374"),     use_mnn = TRUE),
  Female      = list(batches = "TF_C00004337",                        use_mnn = FALSE),
  Male        = list(batches = "TM_C00004374",                        use_mnn = FALSE)
)

datatable(library_meta, rownames = FALSE)
```

# 1. Build the four atlases

Each atlas is normalized across its own libraries with `multiBatchNorm`, so
expression is comparable within an atlas. Highly variable genes are selected on
the combined variance model. Clustering is Louvain on a shared nearest neighbor
graph with `k = 30`, matching `01`, so cluster granularity is comparable to the
existing analysis.

Two objects are kept per atlas: the clustering object, which for a corrected
atlas holds the `fastMNN` output, and an uncorrected expression object holding
`logcounts` for every gene. All quantification below uses the second.

```{r}
build_atlas = function(sce_list, batches, use_mnn, seed = 1, k = 30,
                       n_hvg = 5000) {

  sub      = sce_list[batches]
  universe = Reduce(intersect, lapply(sub, rownames))
  sub      = lapply(sub, "[", i = universe, )

  # strip per-sample extras so cbind does not depend on which sample carries
  # which optional colData field
  sub = lapply(sub, function(x) {
    colData(x)     = colData(x)[, "sample", drop = FALSE]
    reducedDims(x) = list()
    rowData(x)     = rowData(x)[, character(0)]
    x
  })

  dec = lapply(sub, modelGeneVar)

  if (length(sub) > 1) {
    normed = do.call(multiBatchNorm, sub)
    hvg    = getTopHVGs(do.call(combineVar, dec), n = n_hvg)
  } else {
    normed = sub
    hvg    = getTopHVGs(dec[[1]], n = n_hvg)
  }

  expr        = do.call(cbind, normed)
  expr$batch  = expr$sample

  if (use_mnn && length(normed) > 1) {
    set.seed(seed)
    clust = do.call(fastMNN, c(normed,
              list(subset.row = hvg, BSPARAM = RandomParam())))
    dimred = "corrected"
  } else {
    set.seed(seed)
    clust  = runPCA(expr, subset_row = hvg, ncomponents = 50,
                    BSPARAM = RandomParam())
    dimred = "PCA"
  }

  # cbind preserves order, so the two objects stay aligned. Assert it anyway.
  stopifnot(identical(colnames(clust), colnames(expr)))

  set.seed(seed)
  g = buildSNNGraph(clust, use.dimred = dimred, k = k)
  set.seed(seed)
  colLabels(clust) = factor(igraph::cluster_louvain(g)$membership)

  set.seed(seed)
  clust = runTSNE(clust, dimred = dimred)
  set.seed(seed)
  clust = runUMAP(clust, dimred = dimred)

  clust$batch  = expr$batch
  expr$label   = clust$label

  list(clust = clust, expr = expr, dimred = dimred, hvg = hvg,
       corrected = use_mnn && length(normed) > 1)
}

# cache; the chunk that computes also writes, so a rerun never deadlocks
atlas_file = paste0(Robject_out_dir, "/atlases.Rds")

if (file.exists(atlas_file)) {
  atlases = readRDS(atlas_file)
} else {
  atlases = lapply(names(atlas_def), function(nm) {
    build_atlas(all.sce,
                batches = atlas_def[[nm]]$batches,
                use_mnn = atlas_def[[nm]]$use_mnn)
  })
  names(atlases) = names(atlas_def)
  saveRDS(atlases, file = atlas_file)
}

atlas_names = names(atlases)

# every unordered pair of atlases, used by both comparison methods below
atlas_pairs = t(combn(atlas_names, 2))
```

```{r}
atlas_summary = purrr::map_dfr(atlas_names, function(nm) {
  a = atlases[[nm]]
  data.frame(atlas      = nm,
             libraries  = paste(atlas_def[[nm]]$batches, collapse = ", "),
             corrected  = a$corrected,
             nCells     = ncol(a$clust),
             nClusters  = nlevels(a$clust$label),
             nHVG       = length(a$hvg))
})

write.csv(atlas_summary, file = paste0(results_dir, "/atlas_summary.csv"),
          row.names = FALSE)
datatable(atlas_summary, rownames = FALSE)
```

## Cluster composition by library

For the two multi-library atlases, a cluster drawn overwhelmingly from one
library is either a genuinely library-restricted population or a failure of the
merge. For the transformer atlas the distinction is the whole question, and it
is resolved in section 5, not here.

```{r, fig.width=9, fig.height=4}
comp_df = purrr::map_dfr(atlas_names, function(nm) {
  as.data.frame(colData(atlases[[nm]]$clust)) %>%
    dplyr::count(label, batch, name = "n") %>%
    mutate(atlas = nm)
}) %>%
  left_join(library_meta %>% dplyr::select(batch, library), by = "batch")

write.csv(comp_df, file = paste0(results_dir, "/cells_per_cluster_per_library.csv"),
          row.names = FALSE)

ggplot(comp_df %>% filter(atlas %in% c("Larva", "Transformer")),
       aes(fill = library, y = label, x = n)) +
  geom_bar(position = "fill", stat = "identity") +
  facet_wrap(~ atlas, scales = "free_y") +
  xlab("Proportion of cells") + ylab("Cluster")

ggsave(file = paste0(results_dir, "/cluster_composition_by_library.png"),
       width = 9, height = 4)
```

```{r}
datatable(comp_df %>%
            tidyr::pivot_wider(id_cols = c(atlas, label),
                               names_from = library, values_from = n,
                               values_fill = 0),
          filter = "top", rownames = FALSE)
```

## Embeddings

```{r, fig.width=11, fig.height=10}
p = lapply(atlas_names, function(nm) {
  plotTSNE(atlases[[nm]]$clust, colour_by = "label", text_by = "label",
           text_colour = "red", point_size = .05) +
    ggtitle(paste0(nm, if (atlases[[nm]]$corrected) " (fastMNN)" else " (uncorrected)")) +
    guides(color = guide_legend(override.aes = list(size = 4)))
})
wrap_plots(p, ncol = 2)

ggsave(file = paste0(results_dir, "/four_atlas_tsne.png"), width = 11, height = 10)
```

# 2. Marker genes per atlas

Markers are computed within each atlas, blocked on library where an atlas has
more than one, so no marker result depends on the correction. Ordering is set
explicitly by FDR then p-value rather than relying on the order `findMarkers`
returns, and the per-gene effect size is the mean across the pairwise
`logFC.*` columns. `summary.logFC` is not used: under `pval.type = "some"` it
comes from whichever pairwise comparison supplied the summary p-value, which
varies per gene and per run.

```{r}
marker_effect = function(df) {
  lf = grep("^logFC\\.", colnames(df), value = TRUE)
  if (length(lf) == 0) return(rep(NA_real_, nrow(df)))
  rowMeans(as.matrix(as.data.frame(df[, lf, drop = FALSE])), na.rm = TRUE)
}

marker_file = paste0(Robject_out_dir, "/atlas_markers.Rds")

if (file.exists(marker_file)) {
  atlas_markers = readRDS(marker_file)
} else {
  atlas_markers = lapply(atlas_names, function(nm) {
    e     = atlases[[nm]]$expr
    blk   = if (length(unique(e$batch)) > 1) e$batch else NULL
    fm    = findMarkers(e, e$label, test.type = "t", block = blk,
                        direction = "up", pval.type = "some",
                        BPPARAM = bp)
    lapply(fm, function(df) {
      df$meanLogFC = marker_effect(df)
      df[order(df$FDR, df$p.value), ]
    })
  })
  names(atlas_markers) = atlas_names
  saveRDS(atlas_markers, file = marker_file)
}

top_markers = function(nm, lab, n = 100) {
  df = atlas_markers[[nm]][[as.character(lab)]]
  if (is.null(df) || nrow(df) == 0) return(character(0))
  rownames(df)[seq_len(min(n, nrow(df)))]
}
```

```{r}
marker_tbl = purrr::map_dfr(atlas_names, function(nm) {
  purrr::map_dfr(names(atlas_markers[[nm]]), function(lab) {
    df = as.data.frame(atlas_markers[[nm]][[lab]])
    df$Gene = rownames(df)
    df %>%
      filter(FDR < .01) %>%
      dplyr::select(Gene, p.value, FDR, meanLogFC) %>%
      mutate(atlas = nm, cluster = lab)
  })
})

write.csv(marker_tbl, file = paste0(results_dir, "/atlas_cluster_markers.csv"),
          row.names = FALSE)

datatable(marker_tbl %>% group_by(atlas, cluster) %>% slice_head(n = 20),
          filter = "top", rownames = FALSE)
```

## Overlap with curated cell type markers

The hypergeometric overlap from `01`, run once per atlas: rows are curated cell
types, columns are that atlas's clusters, color is the overlap significance and
the number in each tile is the count of shared genes.

The curated list is `canonical_markers_long_2026-08-12.csv`, the repaired long
form, not the S1 sheet. Three of the four erythrocyte entries and the oocyte
`ZAR1` entry in S1 resolve against nothing, so a heatmap built on S1 silently
tests fewer genes than it reports.

Two cautions on reading these. The overlap is against each cluster's top 100
enriched genes, so it measures relative enrichment, not expression; the bubble
plots below measure expression and the two can disagree. And **the four
heatmaps are not comparable to each other**: a cluster competing against 11
others yields a different marker list than one competing against 15, so overlap
counts are biased by cluster number. Compare rows within a heatmap, not tiles
between heatmaps.

```{r}
gene_description = read.csv(paste0(no_dbl_dir, "/gene_descritption.csv"),
                            stringsAsFactors = FALSE)

canonical = read.csv(paste0(data_dir, "/canonical_markers_long_2026-08-12.csv"),
                     stringsAsFactors = FALSE) %>%
  mutate(symbol = trimws(symbol), loc_id = trimws(loc_id)) %>%
  filter(symbol != "")

known_markers = canonical %>%
  dplyr::select(Gene = loc_id, CellType = CellTypeMarkerSet) %>%
  distinct()

# how many marker genes per cluster feed every overlap test in this document
top_n_markers = 100

datatable(canonical %>% dplyr::count(CellTypeMarkerSet, name = "n_markers"),
          rownames = FALSE)
```

```{r, fig.width=8, fig.height=6, results="asis"}
overlap_plots = list()

for (nm in atlas_names) {
  test_genes = marker_tbl %>%
    filter(atlas == nm) %>%
    group_by(cluster) %>%
    slice_head(n = top_n_markers) %>%
    ungroup() %>%
    dplyr::select(Gene, cluster) %>%
    mutate(cluster = paste0("cluster", cluster))

  # nrow is the gene universe the hypergeometric needs, stated explicitly
  overlap_plots[[nm]] = plot_marker_overlap_heatmap(
    known_markers    = known_markers,
    test_genes       = as.data.frame(test_genes),
    background_n     = nrow(atlases[[nm]]$expr),
    gene_description = gene_description,
    overlaps_file    = paste0(results_dir, "/", nm, "_marker_gene_overlaps.csv"),
    pvals_file       = paste0(results_dir, "/", nm, "_marker_overlap_pvalues.csv"),
    plot_file        = paste0(results_dir, "/", nm, "_marker_overlap_heatmap.png"))
}
```

```{r, fig.width=8, fig.height=5}
overlap_plots[["Larva"]] + ggtitle("Larva")
```

```{r, fig.width=8, fig.height=5}
overlap_plots[["Transformer"]] + ggtitle("Transformer")
```

```{r, fig.width=8, fig.height=5}
overlap_plots[["Female"]] + ggtitle("Female")
```

```{r, fig.width=8, fig.height=5}
overlap_plots[["Male"]] + ggtitle("Male")
```

## Canonical marker expression per cluster

The same curated genes measured as expression rather than enrichment, in the
style of the dot plot in `06`. Dot size is the percentage of cells in the
cluster expressing the gene and color is the mean log expression, computed on
uncorrected values within each atlas.

This is the plot to annotate from. A cluster that scores a significant overlap
above but shows no expression here is enriched for a marker it barely
transcribes, which is how cluster 7 in the original analysis came to be called
germline.

```{r}
canon_bubble = function(nm) {
  e   = atlases[[nm]]$expr
  ids = intersect(unique(canonical$loc_id), rownames(e))
  if (length(ids) == 0) return(NULL)

  m   = as.matrix(logcounts(e)[ids, , drop = FALSE])
  lab = as.character(e$label)

  purrr::map_dfr(sort(unique(lab)), function(k) {
    sub = m[, lab == k, drop = FALSE]
    data.frame(atlas          = nm,
               cluster        = k,
               loc_id         = ids,
               n_cells        = ncol(sub),
               pct_expressing = 100 * rowMeans(sub > 0),
               mean_logexpr   = rowMeans(sub),
               stringsAsFactors = FALSE)
  }) %>%
    left_join(canonical, by = "loc_id", relationship = "many-to-many")
}

bubble_df = purrr::map_dfr(atlas_names, canon_bubble) %>%
  mutate(cluster = factor(cluster, levels = sort(unique(as.numeric(cluster)))))

write.csv(bubble_df,
          file = paste0(results_dir, "/canonical_marker_expression_per_cluster.csv"),
          row.names = FALSE)
```

```{r}
plot_bubble = function(nm) {
  ggplot(bubble_df %>% filter(atlas == nm),
         aes(x = cluster, y = symbol,
             size = pct_expressing, color = mean_logexpr)) +
    geom_point() +
    scale_color_viridis_c() +
    scale_size_continuous(range = c(0.5, 5)) +
    facet_grid(CellTypeMarkerSet ~ ., scales = "free_y", space = "free_y") +
    theme(strip.text.y = element_text(angle = 0, size = 6),
          axis.text.y  = element_text(size = 6),
          axis.text.x  = element_text(size = 7)) +
    xlab("Cluster") + ylab(NULL) + ggtitle(nm) +
    labs(size = "% expressing", color = "mean logexpr")
}

for (nm in atlas_names) {
  ggsave(plot_bubble(nm),
         file   = paste0(results_dir, "/", nm, "_canonical_marker_bubble.png"),
         width  = 9, height = 11)
}
```

```{r, fig.width=9, fig.height=11}
plot_bubble("Larva")
```

```{r, fig.width=9, fig.height=11}
plot_bubble("Transformer")
```

```{r, fig.width=9, fig.height=11}
plot_bubble("Female")
```

```{r, fig.width=9, fig.height=11}
plot_bubble("Male")
```

## Best curated cell type per cluster

The single strongest overlap from the heatmaps above, one row per cluster, used to name clusters in sections 5 and 7 rather than leaving them numbered. A cluster whose best overlap misses FDR 0.05 is left `unassigned`.

```{r}
# best curated cell type per cluster, from the same overlap test as section 2,
# so a population can be named rather than just numbered
celltype_file = paste0(Robject_out_dir, "/cluster_celltype_calls.Rds")

if (file.exists(celltype_file)) {
  cluster_celltype = readRDS(celltype_file)
} else {
  cluster_celltype = purrr::map_dfr(atlas_names, function(nm) {
    tg = marker_tbl %>%
      filter(atlas == nm) %>%
      group_by(cluster) %>% slice_head(n = top_n_markers) %>% ungroup() %>%
      dplyr::select(Gene, Group = cluster) %>%
      as.data.frame()

    phyperOverlapSets(known_markers, tg, nrow(atlases[[nm]]$expr))$pval.scores %>%
      dplyr::rename(CellType = Group1, cluster = Group2) %>%
      mutate(FDR = p.adjust(Pval, "BH"), atlas = nm) %>%
      group_by(cluster) %>% slice_min(FDR, n = 1, with_ties = FALSE) %>%
      ungroup() %>%
      mutate(CellType = ifelse(FDR < .05, as.character(CellType), "unassigned"),
             cluster  = as.character(cluster)) %>%
      dplyr::select(atlas, cluster, CellType, celltype_FDR = FDR, nSharedGenes)
  })
  saveRDS(cluster_celltype, file = celltype_file)
}

datatable(cluster_celltype %>%
            mutate(celltype_FDR = signif(celltype_FDR, 3)),
          filter = "top", rownames = FALSE)
```

# 3. Label transfer

The first of two comparison methods. `SingleR` is trained on one
atlas's clusters and used to classify the other atlas's cells, in both
directions. The reference is downsampled to at most 500 cells per cluster so
that large clusters do not dominate.

A cluster pair is called a **reciprocal best match** when each is the other's
most frequently assigned label. A one-directional match is reported but is
weaker: a small cluster can be the best match for a large one without the
reverse holding.

```{r}
have_singler = requireNamespace("SingleR", quietly = TRUE)
```

`r if (have_singler) "SingleR is available; label transfer runs below." else "**SingleR is not installed, so this section is skipped.** Install it with BiocManager::install('SingleR') and rerun. The marker overlap in section 4 does not depend on it and still runs, but section 6 cannot call a cluster pair shared without a reciprocal best match, so install it before relying on the results."`

```{r, eval=have_singler}
downsample_ref = function(nm, max_per = 500) {
  e = atlases[[nm]]$expr
  set.seed(1)
  idx = unlist(lapply(split(seq_len(ncol(e)), e$label), function(i)
    if (length(i) > max_per) sample(i, max_per) else i))
  e[, sort(idx)]
}

# SingleR always returns a label in $labels, even when nothing in the reference
# is a plausible match, so a population with no counterpart is silently pushed
# onto whichever reference cluster is least dissimilar. $pruned.labels is NA
# where the score is too low relative to the score distribution, which is the
# honest answer for a sex-restricted population. NA is kept as an explicit
# category rather than dropped, so it is visible in the plots.
prune_to_char = function(pred) {
  p = pred$pruned.labels
  ifelse(is.na(p), "unassigned", as.character(p))
}

transfer_file = paste0(Robject_out_dir, "/label_transfer.Rds")

if (file.exists(transfer_file)) {
  transfer = readRDS(transfer_file)
} else {
  transfer = list()
  for (i in seq_len(nrow(atlas_pairs))) {
    a = atlas_pairs[i, 1]; b = atlas_pairs[i, 2]
    g = intersect(rownames(atlases[[a]]$expr), rownames(atlases[[b]]$expr))

    ref_a = downsample_ref(a); ref_b = downsample_ref(b)

    # the assay types are named explicitly: these objects carry counts as the
    # first assay, so relying on positional defaults would classify on raw
    # counts without erroring
    pred_ab = SingleR::SingleR(test = atlases[[b]]$expr[g, ],
                               ref  = ref_a[g, ],
                               labels = ref_a$label, de.method = "wilcox",
                               assay.type.test = "logcounts",
                               assay.type.ref  = "logcounts",
                               BPPARAM = bp)
    pred_ba = SingleR::SingleR(test = atlases[[a]]$expr[g, ],
                               ref  = ref_b[g, ],
                               labels = ref_b$label, de.method = "wilcox",
                               assay.type.test = "logcounts",
                               assay.type.ref  = "logcounts",
                               BPPARAM = bp)

    transfer[[paste(a, b, sep = " vs ")]] =
      list(a = a, b = b,
           ab = data.frame(target        = as.character(atlases[[b]]$expr$label),
                           predicted     = prune_to_char(pred_ab),
                           predicted_raw = as.character(pred_ab$labels),
                           stringsAsFactors = FALSE),
           ba = data.frame(target        = as.character(atlases[[a]]$expr$label),
                           predicted     = prune_to_char(pred_ba),
                           predicted_raw = as.character(pred_ba$labels),
                           stringsAsFactors = FALSE))
  }
  saveRDS(transfer, file = transfer_file)
}
```

```{r, eval=have_singler}
# the best assigned label, with frac still taken over every cell in the cluster
# including the unassigned ones. A cluster that is 90% unassigned therefore
# reports its best match at frac 0.1, which is the honest figure. A cluster
# that is entirely unassigned drops out and can have no reciprocal match.
best_match = function(df) {
  totals = df %>% dplyr::count(target, name = "n_total")
  df %>%
    dplyr::count(target, predicted, name = "n") %>%
    left_join(totals, by = "target") %>%
    mutate(frac = n / n_total) %>%
    filter(predicted != "unassigned") %>%
    group_by(target) %>%
    slice_max(n, n = 1, with_ties = FALSE) %>%
    ungroup() %>%
    dplyr::select(target, predicted, n, frac)
}

unassigned_frac = function(df) {
  df %>%
    group_by(target) %>%
    summarize(frac_unassigned = mean(predicted == "unassigned"),
              .groups = "drop")
}

reciprocal = purrr::map_dfr(names(transfer), function(nm) {
  tr = transfer[[nm]]
  m_ab = best_match(tr$ab) %>%
    dplyr::rename(b_cluster = target, a_best = predicted,
                  n_ab = n, frac_ab = frac)
  m_ba = best_match(tr$ba) %>%
    dplyr::rename(a_cluster = target, b_best = predicted,
                  n_ba = n, frac_ba = frac)

  u_ab = unassigned_frac(tr$ab) %>%
    dplyr::rename(b_cluster = target, unassigned_b = frac_unassigned)
  u_ba = unassigned_frac(tr$ba) %>%
    dplyr::rename(a_cluster = target, unassigned_a = frac_unassigned)

  m_ab %>%
    inner_join(m_ba, by = c("b_cluster" = "b_best", "a_best" = "a_cluster")) %>%
    dplyr::rename(a_cluster = a_best) %>%
    left_join(u_ab, by = "b_cluster") %>%
    left_join(u_ba, by = "a_cluster") %>%
    mutate(pair = nm, atlas_a = tr$a, atlas_b = tr$b) %>%
    dplyr::select(pair, atlas_a, a_cluster, atlas_b, b_cluster,
                  frac_ab, frac_ba, unassigned_a, unassigned_b)
})

write.csv(reciprocal,
          file = paste0(results_dir, "/reciprocal_best_matches.csv"),
          row.names = FALSE)

datatable(reciprocal %>% mutate(across(where(is.numeric), ~round(., 3))),
          filter = "top", rownames = FALSE)
```

## Where the cells actually went

The classification itself, as a heatmap. Each row is a cluster in the atlas
being classified and each column is the label it was given. Color is the
fraction of that cluster's cells receiving the label, so every row sums to one.

This is the plainest view of the comparison and needs no threshold to read. A
row concentrated in one column is a cluster with a clear counterpart. A row
spread across several columns is a cluster the other atlas does not resolve the
same way.

**The `unassigned` column is the one to read first.** `SingleR` returns a label
for every cell in `$labels` whether or not the reference contains anything
suitable, so a population with no counterpart is pushed onto whichever
reference cluster is least dissimilar. Female germ cell clusters classified
against a male reference have no correct answer available, and without pruning
they pile onto one arbitrary male cluster and produce a bright column that
looks like a strong match for several unrelated populations at once.

These panels use `$pruned.labels` instead, which is `NA` where the assignment
score is too low relative to the score distribution, and the `NA` values are
kept as an explicit `unassigned` category. A row that is mostly unassigned is
a population with no counterpart in the other atlas, which for a sex-restricted
or stage-restricted cell type is the correct result rather than a failure.

```{r, eval=have_singler}
transfer_long = purrr::map_dfr(names(transfer), function(nm) {
  tr = transfer[[nm]]
  frac_of = function(d, lab) {
    d %>%
      dplyr::count(target, predicted, name = "n") %>%
      group_by(target) %>%
      mutate(frac = n / sum(n)) %>%
      ungroup() %>%
      mutate(direction = lab)
  }
  bind_rows(
    frac_of(tr$ab, paste0(tr$b, " cells, labeled by ", tr$a)),
    frac_of(tr$ba, paste0(tr$a, " cells, labeled by ", tr$b))
  ) %>% mutate(pair = nm)
})

write.csv(transfer_long,
          file = paste0(results_dir, "/label_transfer_assignments.csv"),
          row.names = FALSE)
```

```{r, eval=have_singler, fig.width=11, fig.height=14}
# numeric cluster ids sorted numerically, with "unassigned" pinned last.
# as.numeric alone would turn it into NA and drop the column silently.
cluster_levels = function(x) {
  num = suppressWarnings(sort(unique(as.numeric(x[x != "unassigned"]))))
  c(as.character(num), if (any(x == "unassigned")) "unassigned")
}

p_transfer = ggplot(transfer_long %>%
                      mutate(target    = factor(target,
                                                levels = cluster_levels(target)),
                             predicted = factor(predicted,
                                                levels = cluster_levels(predicted))),
                    aes(x = predicted, y = target, fill = frac)) +
  geom_tile() +
  facet_wrap(~ direction, scales = "free", ncol = 3) +
  scale_fill_viridis_c(limits = c(0, 1)) +
  theme(axis.text = element_text(size = 6),
        axis.text.x = element_text(angle = 90, hjust = 1, vjust = .5),
        strip.text = element_text(size = 7)) +
  xlab("label assigned") + ylab("cluster being classified") +
  labs(fill = "fraction of cells")

p_transfer
ggsave(p_transfer, file = paste0(results_dir, "/label_transfer_heatmap.png"),
       width = 11, height = 14)
```

# 4. Marker gene overlap

The second method. For every pair of clusters from different atlases, the
hypergeometric overlap between their marker gene lists, each computed within
its own atlas. This asks whether two clusters are distinguished from their
neighbors by the same genes, which is a question about gene identity rather
than about expression levels, so it is not a restatement of the label transfer.

Marker lists are restricted to genes present in both atlases before the top
`r top_n_markers` are taken, since a gene missing from one atlas could never overlap and
would only dilute the list. The gene universe is the number of shared genes.

**The known weakness, and it is your own.** `findMarkers` results depend on how
many clusters a cluster is competing against, so an atlas with 12 clusters
produces different marker lists than one with 16. Overlap counts are therefore
not calibrated between atlas pairs of different sizes. Compare within a panel,
and treat this as corroboration of the label transfer rather than as a
criterion that stands on its own.

```{r}
overlap_file = paste0(Robject_out_dir, "/cross_atlas_marker_overlap.Rds")

if (file.exists(overlap_file)) {
  cross_overlap = readRDS(overlap_file)
} else {
  cross_overlap = list()

  for (i in seq_len(nrow(atlas_pairs))) {
    a = atlas_pairs[i, 1]
    b = atlas_pairs[i, 2]

    shared_rn = intersect(rownames(atlases[[a]]$expr),
                          rownames(atlases[[b]]$expr))

    # marker_tbl is already ordered by FDR then p-value from section 2, so
    # slice_head takes the genuine top N rather than an arbitrary N
    tops = function(nm) {
      marker_tbl %>%
        filter(atlas == nm, Gene %in% shared_rn) %>%
        group_by(cluster) %>%
        slice_head(n = top_n_markers) %>%
        ungroup() %>%
        dplyr::select(Gene, Group = cluster) %>%
        as.data.frame()
    }

    ol = phyperOverlapSets(tops(a), tops(b), length(shared_rn))

    cross_overlap[[paste(a, b, sep = " vs ")]] =
      ol$pval.scores %>%
      dplyr::rename(a_cluster = Group1, b_cluster = Group2,
                    n_markers_a = nGroup1, n_markers_b = nGroup2) %>%
      mutate(FDR     = p.adjust(Pval, "BH"),
             pair    = paste(a, b, sep = " vs "),
             atlas_a = a,
             atlas_b = b)
  }
  saveRDS(cross_overlap, file = overlap_file)
}

overlap_tbl = bind_rows(cross_overlap) %>%
  mutate(a_cluster = as.character(a_cluster),
         b_cluster = as.character(b_cluster))

write.csv(overlap_tbl,
          file = paste0(results_dir, "/cross_atlas_marker_overlap.csv"),
          row.names = FALSE)

datatable(overlap_tbl %>%
            filter(FDR < .01) %>%
            arrange(FDR) %>%
            mutate(Pval = signif(Pval, 3), FDR = signif(FDR, 3),
                   Enrichment = round(Enrichment, 2)),
          filter = "top", rownames = FALSE)
```

```{r, fig.width=11, fig.height=8}
overlap_plot_df = overlap_tbl %>%
  mutate(logFDR    = -log10(FDR),
         a_cluster = factor(a_cluster,
                      levels = sort(unique(as.numeric(a_cluster)))),
         b_cluster = factor(b_cluster,
                      levels = sort(unique(as.numeric(b_cluster)))))

# an FDR that underflows to zero would otherwise blow out the scale
cap = max(overlap_plot_df$logFDR[is.finite(overlap_plot_df$logFDR)],
          na.rm = TRUE)
overlap_plot_df$logFDR[!is.finite(overlap_plot_df$logFDR)] = cap

p_overlap = ggplot(overlap_plot_df,
                   aes(x = b_cluster, y = a_cluster, fill = logFDR)) +
  geom_tile() +
  geom_text(aes(label = nSharedGenes), size = 1.6) +
  facet_wrap(~ pair, scales = "free") +
  scale_fill_gradientn(
    colors = c("white", "#fd9668", "#f1605d", "#de4968", "#9e2f7f")) +
  theme(axis.text = element_text(size = 6),
        strip.text = element_text(size = 8)) +
  xlab("second atlas cluster") + ylab("first atlas cluster") +
  labs(fill = "-log10(FDR)")

p_overlap
ggsave(p_overlap, file = paste0(results_dir, "/cross_atlas_marker_overlap.png"),
       width = 11, height = 8)
```

Numbers in the tiles are shared genes out of `r top_n_markers`.

# 5. Did the transformer correction invent anything?

The check to read first, and the only section that does not need `SingleR`.

The male and female atlases contain **the same cells** as the transformer
atlas, not similar cells, so there is no need to infer a correspondence. Every
cell can be looked up in both clusterings by its barcode and the two labels
cross-tabulated directly. That is exact where label transfer is inferential,
and it yields cell counts, so the question of whether a population was simply
too small to resolve becomes a measurement rather than an argument.

## The three explanations this has to separate

A transformer cluster that does not reappear when the libraries are clustered
separately has more than one possible cause, and an earlier version of this
section wrongly treated them as one.

1. **Real but under-resolved.** The population exists in both TF and TM at
   numbers too small for Louvain at `k = 30` to split it out of a larger
   neighbor. Pooling raises both the cell count and the connectivity of the
   nearest neighbor graph, so it separates only in the combined data. This is a
   genuine result, not an artifact.
2. **Single sex.** Nearly all the cells come from one library. The cluster is
   real but it is a female or a male population, and the label should say so.
3. **Correction induced.** `fastMNN` grouped cells that do not cohere in either
   library on its own.

They are separated by where the cells go when the libraries are clustered
separately. For a transformer cluster, take its TM cells and ask which male
cluster each one landed in. If most land in a single male cluster, the cells
cohere in the male data and explanation 1 or 2 applies; whether that male
cluster is much larger tells you whether the population was absorbed rather
than resolved. If the cells scatter across many male clusters, they do not
cohere without the correction, which is explanation 3. Then the same for TF
cells against the female atlas.

```{r}
tf_batch = library_meta$batch[library_meta$library == "Female"]
tm_batch = library_meta$batch[library_meta$library == "Male"]

cell_labels = purrr::map_dfr(atlas_names, function(nm)
  data.frame(barcode = colnames(atlases[[nm]]$expr),
             atlas   = nm,
             cluster = as.character(atlases[[nm]]$expr$label),
             batch   = as.character(atlases[[nm]]$expr$batch),
             stringsAsFactors = FALSE))

# the transformer atlas must be exactly the male and female cells, or the
# cross-tabulation below is comparing different cell sets
tr_bc  = cell_labels$barcode[cell_labels$atlas == "Transformer"]
sub_bc = cell_labels$barcode[cell_labels$atlas %in% c("Male", "Female")]
stopifnot(setequal(tr_bc, sub_bc))

crosstab_with = function(sub_atlas) {
  cell_labels %>%
    filter(atlas == "Transformer") %>%
    dplyr::select(barcode, tr_cluster = cluster) %>%
    inner_join(cell_labels %>%
                 filter(atlas == sub_atlas) %>%
                 dplyr::select(barcode, sub_cluster = cluster),
               by = "barcode") %>%
    dplyr::count(tr_cluster, sub_cluster, name = "n") %>%
    group_by(tr_cluster) %>%
    mutate(frac_of_tr = n / sum(n)) %>%
    ungroup() %>%
    group_by(sub_cluster) %>%
    mutate(frac_of_sub = n / sum(n)) %>%
    ungroup() %>%
    mutate(sub_atlas = sub_atlas)
}

crosstab = bind_rows(crosstab_with("Male"), crosstab_with("Female"))

write.csv(crosstab,
          file = paste0(results_dir, "/transformer_vs_sex_crosstab.csv"),
          row.names = FALSE)
```

```{r, fig.width=10, fig.height=5}
p_xt = ggplot(crosstab %>%
                mutate(tr_cluster  = factor(tr_cluster,
                         levels = sort(unique(as.numeric(tr_cluster)))),
                       sub_cluster = factor(sub_cluster,
                         levels = sort(unique(as.numeric(sub_cluster))))),
              aes(x = sub_cluster, y = tr_cluster, fill = frac_of_tr)) +
  geom_tile() +
  facet_wrap(~ sub_atlas, scales = "free_x") +
  scale_fill_viridis_c(limits = c(0, 1)) +
  theme(axis.text = element_text(size = 6)) +
  xlab("cluster in the single-library atlas") +
  ylab("transformer cluster") +
  labs(fill = "fraction of the\ntransformer cluster's\ncells from that library")

p_xt
ggsave(p_xt, file = paste0(results_dir, "/transformer_vs_sex_crosstab.png"),
       width = 10, height = 5)
```

## Do the two halves of each cluster look alike?

A second, independent check on the same question. A coherent population
expresses its own marker genes in its TF cells and in its TM cells alike. A
grouping forced by the correction shows the markers carried by one library's
cells and not the other.

```{r}
detect_pct    = 10
min_cells_lib = 10

marker_coherence = function(k, n_markers = 50) {
  e  = atlases[["Transformer"]]$expr
  ix = which(as.character(e$label) == k)
  if (length(ix) == 0) return(NULL)

  sub = e[, ix]
  gl  = intersect(top_markers("Transformer", k, n_markers), rownames(sub))
  if (length(gl) == 0) return(NULL)

  m   = as.matrix(logcounts(sub)[gl, , drop = FALSE])
  lib = as.character(sub$batch)

  pct_in = function(b) {
    cols = which(lib == b)
    if (length(cols) < min_cells_lib) return(rep(NA_real_, length(gl)))
    100 * rowMeans(m[, cols, drop = FALSE] > 0)
  }

  data.frame(tr_cluster = k, gene = gl,
             pct_TF = pct_in(tf_batch), pct_TM = pct_in(tm_batch),
             stringsAsFactors = FALSE)
}

coherence = purrr::map_dfr(levels(atlases[["Transformer"]]$clust$label),
                           marker_coherence)

coherence_summary = coherence %>%
  group_by(tr_cluster) %>%
  summarize(n_markers_tested = dplyr::n(),
            # NA when a library has too few cells in the cluster to judge,
            # which is a different result from the markers disagreeing
            pct_markers_in_both = if (all(is.na(pct_TF)) || all(is.na(pct_TM)))
              NA_real_ else
              100 * mean(pct_TF >= detect_pct & pct_TM >= detect_pct),
            .groups = "drop")
```

## Verdict

The two columns that decide it read the cross-tabulation from opposite sides,
and the difference between them is the whole test.

**`concentration_Male`** is read from the transformer cluster's side: of this
cluster's **male** cells, what fraction landed in one male cluster. High means
the cells hold together in the male atlas.

**`share_of_sub_cluster_Male`** is read from the male cluster's side: of that
**male cluster's** cells, what fraction are in this transformer cluster. Low
means this cluster is only a small piece of a larger male cluster.

So a cluster with `concentration_Male = 0.95` and
`share_of_sub_cluster_Male = 0.10` means nearly all its male cells sit together
in one male cluster, but they make up only a tenth of it. The male atlas has
these cells and folded them into something bigger rather than resolving them,
which is a real population the split lacks the power to separate. A cluster
with both values high is one the male atlas resolves on its own.

```{r, echo=FALSE}
datatable(
  data.frame(
    column = c("n_TF, n_TM", "best_sub_cluster_Male",
               "concentration_Male", "share_of_sub_cluster_Male",
               "coheres_male, coheres_female", "absorbed", "both_libraries",
               "n_markers_tested", "pct_markers_in_both", "verdict",
               "singleR_match_in"),
    meaning = c(
      "cells in this transformer cluster from the female and male libraries",
      "the male cluster most of this cluster's male cells landed in",
      "fraction of this cluster's male cells in that one male cluster; high means they cohere in the male atlas",
      "fraction of that male cluster's cells that are in this transformer cluster; low means this cluster was absorbed into something larger",
      "TRUE when concentration reaches conc_min, so the cells cohere in that library",
      "TRUE when share_of_sub_cluster is low on either side, so the population was absorbed rather than resolved",
      "TRUE when both libraries contribute at least min_cells_lib cells",
      "markers used for the TF against TM comparison within this cluster",
      "percent of those markers detected in its female cells and its male cells alike; NA when a library has too few cells to judge",
      "the four-way call; see the text below the table",
      "which sub-atlases gave a reciprocal SingleR match. Reported only, it does not drive the verdict"),
    stringsAsFactors = FALSE),
  rownames = FALSE, options = list(dom = "t", ordering = FALSE))
```

The Female columns mirror the Male ones exactly.

```{r}
conc_min = 0.5

best_sub = crosstab %>%
  group_by(sub_atlas, tr_cluster) %>%
  slice_max(frac_of_tr, n = 1, with_ties = FALSE) %>%
  ungroup() %>%
  dplyr::select(sub_atlas, tr_cluster, best_sub_cluster = sub_cluster,
                n_cells = n, concentration = frac_of_tr,
                share_of_sub_cluster = frac_of_sub)

lib_counts = cell_labels %>%
  filter(atlas == "Transformer") %>%
  dplyr::count(cluster, batch, name = "n") %>%
  tidyr::pivot_wider(id_cols = cluster, names_from = batch, values_from = n,
                     values_fill = 0) %>%
  dplyr::rename(tr_cluster = cluster, n_TF = all_of(tf_batch),
                n_TM = all_of(tm_batch))

wide_sub = best_sub %>%
  tidyr::pivot_wider(id_cols = tr_cluster, names_from = sub_atlas,
                     values_from = c(best_sub_cluster, concentration,
                                     share_of_sub_cluster))

tr_verdict = lib_counts %>%
  left_join(wide_sub, by = "tr_cluster") %>%
  left_join(coherence_summary, by = "tr_cluster") %>%
  mutate(
    both_libraries = n_TF >= min_cells_lib & n_TM >= min_cells_lib,
    coheres_male   = !is.na(concentration_Male)   & concentration_Male   >= conc_min,
    coheres_female = !is.na(concentration_Female) & concentration_Female >= conc_min,
    absorbed = (!is.na(share_of_sub_cluster_Male)   &
                  share_of_sub_cluster_Male   < conc_min) |
               (!is.na(share_of_sub_cluster_Female) &
                  share_of_sub_cluster_Female < conc_min),
    verdict = dplyr::case_when(
      !both_libraries                        ~ "single sex population",
      coheres_male & coheres_female & absorbed
        ~ "real, under-resolved when split",
      coheres_male & coheres_female          ~ "recovered in both sub-atlases",
      coheres_male | coheres_female          ~ "coheres in one library only",
      TRUE                                   ~ "correction induced, dispersed")) %>%
  arrange(verdict, tr_cluster)

if (have_singler) {
  tr_verdict = tr_verdict %>%
    left_join(reciprocal %>%
                filter((atlas_a == "Transformer" &
                          atlas_b %in% c("Male", "Female")) |
                       (atlas_b == "Transformer" &
                          atlas_a %in% c("Male", "Female"))) %>%
                mutate(tr_cluster = ifelse(atlas_a == "Transformer",
                                           a_cluster, b_cluster),
                       other = ifelse(atlas_a == "Transformer",
                                      atlas_b, atlas_a)) %>%
                group_by(tr_cluster) %>%
                summarize(singleR_match_in = paste(sort(unique(other)),
                                                   collapse = ", "),
                          .groups = "drop"),
              by = "tr_cluster") %>%
    mutate(singleR_match_in = ifelse(is.na(singleR_match_in), "neither",
                                     singleR_match_in))
}

write.csv(tr_verdict,
          file = paste0(results_dir, "/transformer_cluster_support.csv"),
          row.names = FALSE)

datatable(tr_verdict %>% mutate(across(where(is.numeric), ~round(., 3))),
          filter = "top", rownames = FALSE)

n_induced = sum(tr_verdict$verdict == "correction induced, dispersed")
n_under   = sum(tr_verdict$verdict == "real, under-resolved when split")
n_onesex  = sum(tr_verdict$verdict == "single sex population")
```

`r if (n_induced == 0) paste0("**No transformer cluster is dispersed across the single-library clusterings.** Every cluster's cells cohere in at least one library on its own, so the correction did not manufacture a population. Of the ", nrow(tr_verdict), " clusters, ", n_under, " separate only in the combined data despite cohering in both libraries, which is a real population that neither library has the numbers to resolve alone, and ", n_onesex, " are restricted to one library and should be labeled by sex.") else paste0("**", n_induced, " transformer cluster(s) have cells that scatter across many clusters in both single-library atlases: ", paste(tr_verdict$tr_cluster[tr_verdict$verdict == "correction induced, dispersed"], collapse = ", "), ".** These do not cohere without the correction and should not be reported as cell populations. Separately, ", n_under, " cluster(s) do cohere in both libraries but are absorbed into larger clusters when the libraries are clustered alone, which is a real population the split lacks the power to resolve rather than an artifact, and ", n_onesex, " are single sex.")`

Read `pct_markers_in_both` alongside the verdict. A cluster called real should
express most of its own markers in its TF cells and its TM cells alike. A high
concentration with low marker agreement between the two halves is worth
inspecting directly before the cluster is reported.

## Which populations does the merge pair up?

The compression from the single-library atlases to the transformer atlas forces
some male and female clusters together. Whether that is correct depends
entirely on **which** ones, so they are named here rather than left to be
eyeballed off the heatmap.

A transformer cluster qualifies when both libraries contribute at least
`r 0.2` of its cells. For each, the table gives the dominant male cluster and
the dominant female cluster it draws from, and the curated cell type assigned
to each of those. A pairing that joins a cell type to its counterpart in the
other sex is the merge working as intended. A pairing that joins two different
cell types is a cost, and the cluster should not be reported as a single
population.

```{r}
pair_share = 0.2

celltype_for = function(a) {
  cluster_celltype %>%
    filter(atlas == a) %>%
    dplyr::select(cluster, CellType)
}

merged_pairs = lib_counts %>%
  mutate(total    = n_TF + n_TM,
         share_TF = n_TF / total,
         share_TM = n_TM / total) %>%
  filter(share_TF >= pair_share, share_TM >= pair_share) %>%
  left_join(best_sub %>% filter(sub_atlas == "Male") %>%
              dplyr::select(tr_cluster, male_cluster = best_sub_cluster,
                            male_conc = concentration),
            by = "tr_cluster") %>%
  left_join(best_sub %>% filter(sub_atlas == "Female") %>%
              dplyr::select(tr_cluster, female_cluster = best_sub_cluster,
                            female_conc = concentration),
            by = "tr_cluster") %>%
  left_join(celltype_for("Male") %>%
              dplyr::rename(male_cluster = cluster, male_CellType = CellType),
            by = "male_cluster") %>%
  left_join(celltype_for("Female") %>%
              dplyr::rename(female_cluster = cluster, female_CellType = CellType),
            by = "female_cluster") %>%
  mutate(pairing = dplyr::case_when(
    is.na(male_CellType) | is.na(female_CellType)   ~ "no dominant cluster on one side",
    male_CellType == "unassigned" |
      female_CellType == "unassigned"               ~ "one side unassigned, cannot judge",
    male_CellType == female_CellType                ~ "same cell type, merge is correct",
    TRUE                                            ~ "different cell types, merge is a cost")) %>%
  arrange(pairing, tr_cluster)

write.csv(merged_pairs,
          file = paste0(results_dir, "/transformer_merged_pairings.csv"),
          row.names = FALSE)

datatable(merged_pairs %>% mutate(across(where(is.numeric), ~round(., 3))),
          filter = "top", rownames = FALSE)

n_pair_ok   = sum(merged_pairs$pairing == "same cell type, merge is correct")
n_pair_bad  = sum(merged_pairs$pairing == "different cell types, merge is a cost")
n_pair_unk  = sum(merged_pairs$pairing %in%
                    c("one side unassigned, cannot judge",
                      "no dominant cluster on one side"))
```

`r paste0(nrow(merged_pairs), " transformer cluster(s) draw at least ", pair_share, " of their cells from each library. Of those, ", n_pair_ok, " join a cell type to the same cell type in the other sex, ", n_pair_bad, " join two different cell types, and ", n_pair_unk, " cannot be judged because at least one side has no curated assignment. ", if (n_pair_bad == 0) "No merge pairs unrelated cell types, so the compression is doing what it should." else "The pairings that join different cell types are the concrete cost of merging and should be named in the manuscript rather than reported as single populations.", if (n_pair_unk > 0) " The unjudgeable pairings depend on the curated marker list rather than on the merge, so they are a limitation of the annotation, not evidence either way." else "")`

## The converse: single-library clusters the merge redistributes

A male or female cluster whose cells scatter across several transformer
clusters has two possible explanations, and the scattering alone does not
distinguish them:

- **Pulled apart.** The correction split a coherent population arbitrarily.
  That is a cost of merging.
- **Resolved more finely.** The single-library cluster was a coarse grouping
  and the transformer atlas separates genuine subtypes inside it. That is a
  benefit of merging.

The column is therefore named `redistributed` rather than dissolved, and which
of the two applies is decided by the test below rather than assumed.

One caution on cluster counts before reading the table, and it concerns a
different event from the one measured here. If the transformer atlas has fewer
clusters than the male and female atlases have between them, some
single-library clusters must **share** a transformer cluster by arithmetic
alone. That is merging, and it is not what `redistributed` measures:
`redistributed` fires only when a cluster's cells scatter so widely that no
single transformer cluster holds a majority of them. Two clusters can each map
cleanly and entirely into one shared transformer cluster without either being
flagged. Nothing about the cluster counts forces scattering.

A compression is also expected rather than alarming. A cell type shared between
the sexes is counted twice across the two single-library atlases, once in each,
so it should collapse to one cluster when they are merged. If the sexes share
`S` cell types, the transformer atlas should hold roughly the male count plus
the female count minus `S`, and the forced sharing is that collapse happening.
The subsection after next names the pairings so you can check whether each one
joins a cell type to its counterpart or to something unrelated.

```{r}
n_tr_clusters   = nlevels(atlases[["Transformer"]]$clust$label)
n_sub_clusters  = nlevels(atlases[["Male"]]$clust$label) +
                  nlevels(atlases[["Female"]]$clust$label)

redistributed = crosstab %>%
  group_by(sub_atlas, sub_cluster) %>%
  summarize(n_cells        = sum(n),
            top_tr_cluster = tr_cluster[which.max(frac_of_sub)],
            largest_share  = max(frac_of_sub),
            n_tr_clusters  = dplyr::n_distinct(tr_cluster[frac_of_sub > .1]),
            .groups = "drop") %>%
  mutate(redistributed = largest_share < conc_min) %>%
  arrange(largest_share)
```

The transformer atlas has `r n_tr_clusters` clusters against `r n_sub_clusters`
across the male and female atlases combined, so `r if (n_tr_clusters >= n_sub_clusters) "no merging is forced by cluster number alone" else paste0("at least ", n_sub_clusters - n_tr_clusters, " single-library cluster(s) must share a transformer cluster with another regardless of the biology")`.

## Is a redistributed cluster split into something real?

The decisive test, and it is clean: take one single-library cluster's cells,
group them by which transformer cluster they landed in, and test for
differential expression **between those groups using only that library's
cells**. Every cell comes from one animal and one library, so there is nothing
to block on and no correction is involved. Nothing about this test depends on
the thing being questioned.

If the pieces differ from each other, the transformer atlas found substructure
the single-library clustering missed. If they do not, the population was
partitioned arbitrarily.

A negative result is only meaningful with enough cells to detect a difference.
A split whose smallest piece is small is reported as underpowered rather than
as arbitrary, since absence of evidence at low `n` is not evidence of absence.

```{r, echo=FALSE}
datatable(
  data.frame(
    column = c("n_cells", "top_tr_cluster", "largest_share", "n_tr_clusters",
               "redistributed", "n_groups", "smallest_group",
               "min_sig_genes", "max_sig_genes", "split_verdict"),
    meaning = c(
      "cells in this single-library cluster",
      "the transformer cluster that took the most of them",
      "fraction of this cluster's cells that went to that one transformer cluster",
      "how many transformer clusters took more than 10% of it",
      "TRUE when largest_share is below conc_min, so no transformer cluster holds a majority",
      "pieces large enough to test, at min_cells_group each",
      "cells in the smallest piece tested; small values make a null result uninformative",
      "significant genes for the least distinct piece. This is the criterion: every piece must be distinguishable",
      "significant genes for the most distinct piece, shown for context",
      "whether the split found real subtypes, partitioned arbitrarily, or cannot be decided"),
    stringsAsFactors = FALSE),
  rownames = FALSE, options = list(dom = "t", ordering = FALSE))
```

```{r}
min_cells_group = 20   # a piece smaller than this is not tested at all
min_power_group = 50   # below this, a null result is called underpowered
min_sig_cut     = 10   # significant genes needed to call a piece distinct
                       # named min_sig_cut, not min_sig_genes, because the
                       # results table already has a min_sig_genes column and
                       # inside mutate the column would shadow the constant

redist_file = paste0(Robject_out_dir, "/redistribution_de.Rds")

redistribution_de = function(sub_atlas, sub_cl) {
  e   = atlases[[sub_atlas]]$expr
  key = cell_labels %>%
    filter(atlas == "Transformer") %>%
    dplyr::select(barcode, tr_cluster = cluster)

  ix = which(as.character(e$label) == sub_cl)
  if (length(ix) == 0) return(NULL)

  sub = e[, ix]
  grp = key$tr_cluster[match(colnames(sub), key$barcode)]

  keep_grp = names(which(table(grp) >= min_cells_group))
  if (length(keep_grp) < 2) return(NULL)

  cols = which(grp %in% keep_grp)
  sub  = sub[, cols]
  grp  = droplevels(factor(grp[cols]))

  # no block argument: every cell here is from a single library by construction
  fm    = findMarkers(sub, grp, test.type = "t", direction = "up",
                      pval.type = "some")
  n_sig = vapply(fm, function(d) sum(d$FDR < .01, na.rm = TRUE), numeric(1))

  data.frame(sub_atlas      = sub_atlas,
             sub_cluster    = sub_cl,
             n_groups       = nlevels(grp),
             groups         = paste(levels(grp), collapse = ","),
             n_cells_tested = ncol(sub),
             smallest_group = min(table(grp)),
             min_sig_genes  = min(n_sig),
             max_sig_genes  = max(n_sig),
             stringsAsFactors = FALSE)
}

if (file.exists(redist_file)) {
  redist_de = readRDS(redist_file)
} else {
  targets = redistributed %>% filter(redistributed)
  redist_de = if (nrow(targets) == 0) NULL else
    purrr::map_dfr(seq_len(nrow(targets)), function(i)
      redistribution_de(targets$sub_atlas[i], targets$sub_cluster[i]))
  saveRDS(redist_de, file = redist_file)
}

redistributed = redistributed %>%
  left_join(if (is.null(redist_de)) tibble::tibble(sub_atlas = character(),
                                                   sub_cluster = character())
            else redist_de,
            by = c("sub_atlas", "sub_cluster")) %>%
  mutate(split_verdict = dplyr::case_when(
    !redistributed                   ~ "not redistributed",
    is.na(min_sig_genes)             ~ "too few cells to test",
    min_sig_genes >= min_sig_cut     ~ "real substructure, merge resolves more",
    smallest_group < min_power_group ~ "underpowered, cannot decide",
    TRUE                             ~ "arbitrary partition, merge costs a population"))

write.csv(redistributed,
          file = paste0(results_dir, "/sex_clusters_redistributed.csv"),
          row.names = FALSE)

datatable(redistributed %>% mutate(across(where(is.numeric), ~round(., 3))),
          filter = "top", rownames = FALSE)
```

## Balance sheet: combine the transformers, or report them separately?

Merging is worth it when it reveals more than it destroys. The accounting below
is in **cells rather than clusters**, for two reasons.

Cells weight by consequence. A 2,000-cell population partitioned arbitrarily
matters far more than a 40-cell one, and a count of clusters treats them
identically.

Cells are also a common currency, which a cluster count is not. The three
findings in this section are counted in different units: transformer clusters
in one case, single-library clusters in another. Adding those together produces
a number that does not mean anything. Every percentage below is instead a share
of the same denominator, the total transformer cells, so the three ledgers are
directly comparable even though each describes a different partition of them.

For that reason the ledgers are reported separately and never summed. The same
cell is counted once in each, since it belongs to exactly one transformer
cluster and exactly one male or female cluster.

```{r}
n_tr_cells = ncol(atlases[["Transformer"]]$expr)

# ledger 1: every transformer cell, by what the cross-tabulation concluded
# about the transformer cluster it sits in. Exhaustive, sums to 100 percent.
ledger_tr = tr_verdict %>%
  mutate(n_cells = n_TF + n_TM) %>%
  group_by(outcome = verdict) %>%
  summarize(n_clusters = dplyr::n(), n_cells = sum(n_cells), .groups = "drop") %>%
  mutate(ledger = "transformer clusters",
         pct_cells = 100 * n_cells / n_tr_cells)

# ledger 2: every cell, by what happened to the single-library cluster it came
# from. Also exhaustive, also sums to 100 percent.
ledger_sub = redistributed %>%
  group_by(outcome = split_verdict) %>%
  summarize(n_clusters = dplyr::n(), n_cells = sum(n_cells), .groups = "drop") %>%
  mutate(ledger = "male and female clusters",
         pct_cells = 100 * n_cells / n_tr_cells)

# ledger 3: only the transformer clusters both libraries feed substantially,
# so this one does not sum to 100 percent and is not meant to.
ledger_pair = merged_pairs %>%
  group_by(outcome = pairing) %>%
  summarize(n_clusters = dplyr::n(), n_cells = sum(total), .groups = "drop") %>%
  mutate(ledger = "merged pairings",
         pct_cells = 100 * n_cells / n_tr_cells)

balance = bind_rows(ledger_tr, ledger_sub, ledger_pair) %>%
  dplyr::select(ledger, outcome, n_clusters, n_cells, pct_cells) %>%
  arrange(ledger, desc(pct_cells))

write.csv(balance, file = paste0(results_dir, "/merge_balance_sheet.csv"),
          row.names = FALSE)

datatable(balance %>% mutate(pct_cells = round(pct_cells, 1)),
          rownames = FALSE)
```

```{r, fig.width=9, fig.height=4}
p_balance = ggplot(balance, aes(x = pct_cells, y = reorder(outcome, pct_cells))) +
  geom_col(fill = "#9ecae1") +
  geom_text(aes(label = paste0(round(pct_cells, 1), "%  (", n_clusters, ")")),
            hjust = -0.1, size = 2.6) +
  facet_wrap(~ ledger, scales = "free_y", ncol = 1) +
  scale_x_continuous(expand = expansion(mult = c(0, .25))) +
  theme(axis.text.y = element_text(size = 7), strip.text = element_text(size = 8)) +
  xlab("percent of transformer cells") + ylab(NULL)

p_balance
ggsave(p_balance, file = paste0(results_dir, "/merge_balance_sheet.png"),
       width = 9, height = 4)
```

Bars are labeled with the cell percentage and, in brackets, the number of
clusters, so the two readings stay visible together.

```{r}
# a cost worth changing the presentation over, as a share of all cells. Not
# derived from anything; it is a judgment call and is stated so it can be argued with.
cost_threshold = 5

pct_of = function(tbl, x) {
  v = tbl$pct_cells[tbl$outcome == x]
  if (length(v) == 0) 0 else sum(v)
}

pct_induced   = pct_of(ledger_tr,   "correction induced, dispersed")
pct_underres  = pct_of(ledger_tr,   "real, under-resolved when split")
pct_arbitrary = pct_of(ledger_sub,  "arbitrary partition, merge costs a population")
pct_substruct = pct_of(ledger_sub,  "real substructure, merge resolves more")
pct_badpair   = pct_of(ledger_pair, "different cell types, merge is a cost")
pct_goodpair  = pct_of(ledger_pair, "same cell type, merge is correct")
pct_undecided = pct_of(ledger_sub,  "underpowered, cannot decide") +
                pct_of(ledger_sub,  "too few cells to test")

worst_cost = max(pct_induced, pct_arbitrary, pct_badpair)
```

`r paste0("**Costs.** ", round(pct_induced, 1), "% of transformer cells sit in clusters that cohere in neither library on its own, ", round(pct_arbitrary, 1), "% belong to single-library populations the merge partitions arbitrarily, and ", round(pct_badpair, 1), "% sit in clusters that pair two different cell types. **Gains.** ", round(pct_underres, 1), "% sit in populations that separate only in the combined data despite cohering in both libraries, ", round(pct_substruct, 1), "% belong to single-library clusters the merge splits into genuinely distinct subtypes, and ", round(pct_goodpair, 1), "% sit in clusters where the merge correctly joins a cell type to its counterpart in the other sex. ", round(pct_undecided, 1), "% cannot be decided at this cell number.")`

`r if (worst_cost < cost_threshold) paste0("No individual cost exceeds ", cost_threshold, "% of cells. **The integrated transformer atlas is defensible as the primary object**, with the male and female atlases reported alongside it for the sex-restricted populations, which are real results rather than merge failures.") else if (pct_badpair >= cost_threshold) paste0("**", round(pct_badpair, 1), "% of cells sit in clusters that merge two different cell types**, which is the most serious of the three costs because those clusters would be reported as single populations that are not. Either report the male and female atlases separately, or split those specific clusters by library and say so in the text.") else if (pct_induced >= cost_threshold) paste0("**", round(pct_induced, 1), "% of cells sit in clusters that cohere in neither library**, so the correction is generating structure rather than aligning it. Reporting the male and female atlases separately is better supported.") else paste0("**", round(pct_arbitrary, 1), "% of cells belong to single-library populations the merge partitions arbitrarily.** Those populations are visible in the sex atlases and lost in the integrated one, which argues for reporting the sex atlases as primary.")`

Two limits on this. The `r cost_threshold`% threshold is a judgment call, not a
derived quantity, so read the percentages rather than the verdict if any of
them sit near it. And the undecided share is undecided because of cell numbers,
so it can only be resolved by more cells, not by more analysis.

This balance is about the transformer libraries only. It says nothing about
whether larva and transformer should be integrated with each other, which is
the separate question the four-atlas design already answers by not doing it.

# 6. Shared clusters across atlases

The final table. A cluster pair is called shared when it is a reciprocal best
match under label transfer **and** its marker lists overlap significantly. The
label transfer proposes the pairing and the overlap corroborates it, so a pair
carried by one method alone is reported but not called.

`methods_agree` records something slightly stronger: whether the marker overlap
independently picks the same partner, meaning the pair is each other's top
overlap as well as each other's best transferred label. A pair can clear the
significance threshold without being the top overlap, which usually means
several clusters in the other atlas are plausible and the population is real
while the specific correspondence is not settled.

```{r, echo=FALSE}
datatable(
  data.frame(
    column = c("a_cluster, b_cluster", "frac_ab", "frac_ba",
               "unassigned_a, unassigned_b", "nSharedGenes", "FDR",
               "Enrichment", "methods_agree", "independent", "call"),
    meaning = c(
      "the matched pair, one cluster from each atlas",
      "fraction of b_cluster's cells that SingleR labeled a_cluster, counting unassigned cells in the denominator",
      "the same in the other direction",
      "fraction of each cluster's cells SingleR could not confidently assign at all; high values mean the population has no counterpart",
      "genes shared between the two clusters' top marker lists",
      "significance of that overlap, BH adjusted within the atlas pair",
      "log2 enrichment of the overlap over what chance would give",
      "TRUE when the marker overlap independently picks the same partner in both directions, which is stronger than passing the FDR cutoff",
      "TRUE when the two atlases are disjoint sets of cells. FALSE for male or female against transformer, which share cells",
      "shared, or which method carried it alone"),
    stringsAsFactors = FALSE),
  rownames = FALSE, options = list(dom = "t", ordering = FALSE))
```

```{r, eval=have_singler}
fdr_cut = 0.01

# does the overlap independently choose the same partner, in both directions
best_overlap_a = overlap_tbl %>%
  group_by(pair, a_cluster) %>% slice_min(FDR, n = 1, with_ties = FALSE) %>%
  ungroup() %>% dplyr::select(pair, a_cluster, top_overlap_b = b_cluster)

best_overlap_b = overlap_tbl %>%
  group_by(pair, b_cluster) %>% slice_min(FDR, n = 1, with_ties = FALSE) %>%
  ungroup() %>% dplyr::select(pair, b_cluster, top_overlap_a = a_cluster)

shared = reciprocal %>%
  mutate(a_cluster = as.character(a_cluster),
         b_cluster = as.character(b_cluster)) %>%
  left_join(overlap_tbl %>%
              dplyr::select(pair, a_cluster, b_cluster, nSharedGenes,
                            Pval, FDR, Enrichment),
            by = c("pair", "a_cluster", "b_cluster")) %>%
  left_join(best_overlap_a, by = c("pair", "a_cluster")) %>%
  left_join(best_overlap_b, by = c("pair", "b_cluster")) %>%
  # only the transformer atlas shares cells with male and female. Female
  # against male is between different animals and so is independent.
  mutate(independent = !((atlas_a == "Transformer" &
                          atlas_b %in% c("Male", "Female")) |
                         (atlas_b == "Transformer" &
                          atlas_a %in% c("Male", "Female"))),
         overlap_sig   = !is.na(FDR) & FDR < fdr_cut,
         methods_agree = !is.na(top_overlap_b) & !is.na(top_overlap_a) &
                         top_overlap_b == b_cluster &
                         top_overlap_a == a_cluster,
         call = dplyr::case_when(
           is.na(FDR)  ~ "label transfer only, overlap unavailable",
           overlap_sig ~ "shared",
           TRUE        ~ "label transfer only, overlap not significant")) %>%
  arrange(desc(independent), pair, FDR)

write.csv(shared, file = paste0(results_dir, "/shared_clusters.csv"),
          row.names = FALSE)

datatable(shared %>%
            mutate(Pval = signif(Pval, 3), FDR = signif(FDR, 3),
                   across(where(is.numeric), ~round(., 3))),
          filter = "top", rownames = FALSE)
```

The same table as a picture. Each point is one reciprocal best match from label
transfer, placed at the significance of its marker overlap. The dashed line is
the `r 0.01` FDR threshold.

**How to read it.** A point to the right of the dashed line is a shared
population, and the further right the stronger the marker evidence. A filled
circle means the marker overlap independently picked the same partner; a hollow
triangle means it did not, so the population is real but which cluster it
corresponds to is not settled. Panels involving male or female against
transformer share cells, so they show whether the merge preserved structure
rather than whether a population recurs.

```{r, eval=have_singler, fig.width=11, fig.height=9}
plot_df = shared %>%
  mutate(match_label   = paste0(a_cluster, " to ", b_cluster),
         methods_agree = ifelse(is.na(methods_agree), FALSE, methods_agree),
         logFDR        = -log10(FDR))

# an FDR that underflows to zero would otherwise sit at infinity
fdr_cap = suppressWarnings(max(plot_df$logFDR[is.finite(plot_df$logFDR)],
                               na.rm = TRUE))
if (!is.finite(fdr_cap)) fdr_cap = 1
plot_df$logFDR[!is.finite(plot_df$logFDR)] = fdr_cap

p_shared = ggplot(plot_df,
                  aes(x = logFDR, y = reorder(match_label, logFDR))) +
  geom_vline(xintercept = -log10(fdr_cut),
             linetype = "dashed", color = "grey40") +
  geom_point(aes(color = call, shape = methods_agree), size = 2.5) +
  facet_wrap(~ pair, scales = "free_y") +
  scale_shape_manual(values = c(`FALSE` = 2, `TRUE` = 16),
                     name = "overlap picks the same partner") +
  scale_color_brewer(palette = "Set1") +
  theme(axis.text.y = element_text(size = 6),
        strip.text  = element_text(size = 8),
        legend.position = "bottom",
        legend.box = "vertical") +
  xlab("-log10(FDR) of marker overlap") + ylab(NULL) +
  guides(color = guide_legend(ncol = 1))

p_shared
ggsave(p_shared, file = paste0(results_dir, "/shared_clusters_summary.png"),
       width = 11, height = 9)
```

`r if (!have_singler) "Not evaluated, since SingleR is unavailable. The marker overlap heatmap in section 4 is the only comparison the document can produce until it is installed, and it proposes no pairings on its own." else paste0("Of ", nrow(shared), " reciprocal best matches, ", sum(shared$call == "shared"), " also have a significant marker overlap and are called shared. ", sum(shared$methods_agree, na.rm = TRUE), " of those are the top overlap partner in both directions as well, which is the strongest form of agreement available here. ", sum(shared$independent), " of the reciprocal matches are between disjoint sets of cells: those involving the larval atlas, and female against male, which come from different animals. The remainder are male or female against transformer, which share cells and therefore measure the integration rather than replication of a population.")`

## Larval populations with no transformer counterpart, and the reverse

```{r, eval=have_singler}
lt = shared %>% filter(pair == "Larva vs Transformer", call == "shared")

larva_unmatched = setdiff(levels(atlases[["Larva"]]$clust$label),
                          lt$a_cluster)
trans_unmatched = setdiff(levels(atlases[["Transformer"]]$clust$label),
                          lt$b_cluster)

data.frame(
  atlas   = c("Larva", "Transformer"),
  n_total = c(nlevels(atlases[["Larva"]]$clust$label),
              nlevels(atlases[["Transformer"]]$clust$label)),
  n_shared = c(length(unique(lt$a_cluster)), length(unique(lt$b_cluster))),
  unmatched = c(paste(larva_unmatched, collapse = ", "),
                paste(trans_unmatched, collapse = ", "))) %>%
  datatable(rownames = FALSE)
```

A cluster with no counterpart is a stage-restricted population, which is a
result rather than a failure. Oocytes and spermatocytes should appear here on
the transformer side, since larvae have undifferentiated gonads. A somatic
cluster appearing here is more interesting and should be examined directly.

# 7. How many atlases is each population found in?

Sections 5 and 6 work in pairs. This section links those pairs into groups, so
a population can be counted across all four atlases at once.

Every cluster in every atlas is a node, every shared call from section 6 is an
edge, and a **population** is a connected component of that graph. The number
of atlases a population spans is then just the number of distinct atlases among
its member clusters.

**Three things to hold in mind when reading this.**

Connected components assume transitivity: if larva 3 matches transformer 5, and
transformer 5 matches male 2, the three are grouped even though larva 3 against
male 2 was never itself called shared. `fully_linked` reports whether every
cross-atlas pair inside a population was independently called, so a chained
group is visible rather than hidden.

A population can contain two clusters from the same atlas. That is a split, one
atlas resolving into two clusters what another resolves into one, and it is
reported rather than collapsed.

**`n_atlases` overcounts sampling.** The transformer atlas is built from the
male and female cells, so a population present in male, female and transformer
spans three atlases but only two animals. `n_independent` counts only the
larval, female and male atlases, which are disjoint sets of cells, and is the
number to use when asking how reproducible a population is.

```{r, echo=FALSE}
datatable(
  data.frame(
    column = c("population", "n_atlases", "n_independent", "atlases",
               "n_clusters", "split", "CellType", "members",
               "n_possible", "n_actual", "fully_linked"),
    meaning = c(
      "an arbitrary identifier for the connected component",
      "how many of the four atlases it appears in",
      "how many disjoint cell sets it appears in, counting only larva, female and male. Transformer is excluded because it is built from the male and female cells, so it adds no new sampling. This is the number to use for reproducibility",
      "which atlases, by name",
      "member clusters, which can exceed n_atlases when one atlas splits the population",
      "TRUE when some atlas resolves this population into more than one cluster",
      "most frequent curated cell type among the member clusters",
      "the member clusters, as atlas:cluster",
      "cross-atlas cluster pairs inside this population",
      "how many of those were independently called shared",
      "TRUE when every cross-atlas pair was called, rather than the population being chained together through a third cluster"),
    stringsAsFactors = FALSE),
  rownames = FALSE, options = list(dom = "t", ordering = FALSE))
```

```{r, eval=have_singler}
nodes = purrr::map_dfr(atlas_names, function(nm)
  data.frame(name    = paste0(nm, ":", levels(atlases[[nm]]$clust$label)),
             atlas   = nm,
             cluster = levels(atlases[[nm]]$clust$label),
             stringsAsFactors = FALSE)) %>%
  left_join(cluster_celltype, by = c("atlas", "cluster"))

edges = shared %>%
  filter(call == "shared") %>%
  transmute(from = paste0(atlas_a, ":", a_cluster),
            to   = paste0(atlas_b, ":", b_cluster))

stopifnot(all(c(edges$from, edges$to) %in% nodes$name))

g    = igraph::graph_from_data_frame(edges, directed = FALSE, vertices = nodes)
comp = igraph::components(g)
nodes$population = as.integer(comp$membership[nodes$name])

# how many of the cross-atlas pairs inside a population were actually called
link_counts = nodes %>%
  group_by(population) %>%
  summarize(n_possible = sum(outer(atlas, atlas, "!=")) / 2, .groups = "drop")

edge_counts = edges %>%
  mutate(population = nodes$population[match(from, nodes$name)]) %>%
  dplyr::count(population, name = "n_actual")

# the most frequent assigned cell type among a population's clusters. Written
# as a function because names(sort(table(x)))[1] returns NULL rather than NA
# when every cluster is unassigned, and summarize cannot combine NULL with the
# character results from the other groups. Ties break on the first name, which
# is alphabetical and therefore at least deterministic.
consensus_celltype = function(x) {
  x = x[!is.na(x) & x != "unassigned"]
  if (length(x) == 0) return("unassigned")
  tab = table(x)
  names(tab)[which.max(tab)]
}

pop_summary = nodes %>%
  group_by(population) %>%
  summarize(n_atlases     = dplyr::n_distinct(atlas),
            n_independent = dplyr::n_distinct(atlas[atlas != "Transformer"]),
            atlases       = paste(sort(unique(atlas)), collapse = ", "),
            n_clusters    = dplyr::n(),
            split         = dplyr::n() > dplyr::n_distinct(atlas),
            CellType      = consensus_celltype(CellType),
            members       = paste(sort(name), collapse = "; "),
            .groups = "drop") %>%
  left_join(link_counts, by = "population") %>%
  left_join(edge_counts, by = "population") %>%
  mutate(n_actual     = ifelse(is.na(n_actual), 0, n_actual),
         fully_linked = n_possible == 0 | n_actual == n_possible,
         CellType     = ifelse(is.na(CellType), "unassigned", CellType)) %>%
  arrange(desc(n_independent), desc(n_atlases), CellType)

write.csv(pop_summary, file = paste0(results_dir, "/populations_across_atlases.csv"),
          row.names = FALSE)

datatable(pop_summary, filter = "top", rownames = FALSE)
```

## Counts

```{r, eval=have_singler}
count_tbl = pop_summary %>%
  dplyr::count(n_independent, n_atlases, name = "n_populations") %>%
  arrange(desc(n_independent), desc(n_atlases))

datatable(count_tbl, rownames = FALSE)
```

`r if (!have_singler) "Not evaluated, since SingleR is unavailable." else paste0("There are ", nrow(pop_summary), " populations in total. ", sum(pop_summary$n_independent == 3), " are present in all three disjoint cell sets, larva, female and male, which is the strongest reproducibility available here. ", sum(pop_summary$n_independent == 2), " are present in two, and ", sum(pop_summary$n_independent == 1), " appear in one atlas only and are therefore either stage restricted, sex restricted, or not reproducible. ", sum(!pop_summary$fully_linked), " population(s) are chained rather than fully linked, meaning at least one cross-atlas pair inside them was never independently called shared.")`

## Which atlases each population appears in

```{r, eval=have_singler, fig.width=9, fig.height=10}
pop_atlas = nodes %>%
  group_by(population, atlas) %>%
  summarize(clusters = paste(sort(cluster), collapse = ","), .groups = "drop") %>%
  tidyr::complete(population, atlas = atlas_names) %>%
  left_join(pop_summary %>%
              dplyr::select(population, n_independent, n_atlases, CellType),
            by = "population") %>%
  mutate(atlas     = factor(atlas, levels = atlas_names),
         pop_label = paste0(population, ". ", CellType))

pop_order = pop_summary %>%
  arrange(n_independent, n_atlases, desc(CellType)) %>%
  mutate(pop_label = paste0(population, ". ", CellType)) %>%
  pull(pop_label)

p_pop = ggplot(pop_atlas %>%
                 mutate(pop_label = factor(pop_label, levels = pop_order)),
               aes(x = atlas, y = pop_label, fill = !is.na(clusters))) +
  geom_tile(color = "white") +
  geom_text(aes(label = clusters), size = 2.2) +
  scale_fill_manual(values = c(`TRUE` = "#9ecae1", `FALSE` = "grey95"),
                    guide = "none") +
  theme(axis.text.y = element_text(size = 6),
        axis.text.x = element_text(angle = 30, hjust = 1)) +
  xlab(NULL) + ylab(NULL)

p_pop
ggsave(p_pop, file = paste0(results_dir, "/populations_across_atlases.png"),
       width = 9, height = 10)
```

Numbers in the tiles are cluster identifiers in that atlas. Two numbers in one
tile means that atlas splits the population into two clusters.

# Session info

```{r}
sessionInfo()
```
