---
title: "11_no_doublets_one_atlas"
author: "Stephanie Hickey"
date: "`r Sys.Date()`"
output:
  html_document:
    code_folding: hide
    toc: true
    toc_float:
      toc_collapsed: true
params:
    project_dir: /mnt/ufs18/rs-013/bioinformaticsCore/projects/chong_davidson/BCC125_sea_lamprey_scRNA
    project_name: BCC125_sea_lamprey_scRNA
    lab: Weiming Li
    no_dbl_date: "20260813"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

```{r, echo=FALSE, results="asis"}
cat("\n")
cat("#### Lab:", params$lab, " \n")
cat("#### Project:", params$project_name, " \n")
```

```{r}
library(tidyverse)
library(SingleCellExperiment)
library(scater)
library(scran)
library(patchwork)
library(RColorBrewer)
library(DT)
source(paste0(params$project_dir, "/src/scRNA_seq_functions.R"))
```

```{r}
no_dbl_dir  = paste0(params$project_dir, "/results/", params$no_dbl_date, "_no_doublets")

results_dir     = paste0(params$project_dir, "/results/", params$no_dbl_date,
                         "_no_doublets_one_atlas")
Robject_out_dir = paste0(results_dir, "/Robjects")
if (!dir.exists(Robject_out_dir)) dir.create(Robject_out_dir, recursive = TRUE)
data_dir    = paste0(params$project_dir, "/data")
```

# Purpose

**Do the clusters of the doublet-filtered atlas name the same populations in
larval and in transformer cells, or should the stages be described separately?**

The test is whether each cluster **finds itself** across the stage boundary
using its own data-driven markers. Marker genes are computed from larval cells
alone and, separately, from transformer cells alone. If the genes that define
cluster 8 among larvae are the genes that define cluster 8 among transformers,
and they match cluster 8 better than they match any other cluster, then cluster 8
is the same population on both sides.

**Nothing here uses the curated marker list.** An earlier version of this
document tested whether each cluster expressed its assigned canonical markers.
That route runs every claim through mammalian orthology and the BLAST
re-annotation of LOC identifiers, which is the weakest link in this project and
the source of the withdrawn primordial germ cell annotation. The curated markers
also do not reliably separate cell types in this tissue, so a cluster failing
that test was as likely to mean the marker was uninformative as that the cluster
was incoherent. Using each cluster's own markers removes that dependency
entirely: no gene symbol, no ortholog call, and no cell type name enters the
computation.

**Why this survives the confound.** Larvae were collected in July and
transformers in November, so stage, collection date and technical batch are one
variable, and nothing computed from this dataset can say how much of the
difference between them is developmental. This test does not try. It asks
whether the populations correspond, not what makes them differ. A cluster that
matches itself across the boundary is the same population at both stages
whatever the July-to-November difference consists of.

**Everything is read against a reference computed the same way.** Undifferentiated.1
against Undifferentiated.2 is two animals of the same stage, collected on the
same date, so it measures what correspondence looks like when only replicate
variation separates the two sides. That is the ceiling. The stage comparison is
reported relative to it, never against 100%.

# Load

```{r}
merged.sce   = readRDS(paste0(no_dbl_dir, "/merged.sce.Rds"))
combined.sce = readRDS(paste0(no_dbl_dir, "/combined.sce.Rds"))

# combined.sce carries the uncorrected logcounts and merged.sce the clustering.
# All quantification below is on the uncorrected values; the correction only
# ever affected clustering and embedding.
stopifnot(identical(colnames(combined.sce), colnames(merged.sce)))
```

```{r}
library_meta = data.frame(
  batch   = c("UND1_C00004336", "UND2_C00004362", "TF_C00004337", "TM_C00004374"),
  library = c("Undifferentiated.1", "Undifferentiated.2", "Female", "Male"),
  stage   = c("larva", "larva", "transformer", "transformer"),
  stringsAsFactors = FALSE)

observed   = sort(unique(as.character(colData(merged.sce)$batch)))
unexpected = setdiff(observed, library_meta$batch)

if (length(unexpected) > 0) {
  stop("Library identifier(s) in the object but not in library_meta: ",
       paste(unexpected, collapse = ", "),
       ". Add them rather than silently dropping cells.")
}

cell_key = data.frame(barcode = colnames(combined.sce),
                      label   = as.character(colData(merged.sce)$label),
                      batch   = as.character(colData(merged.sce)$batch),
                      stringsAsFactors = FALSE) %>%
  left_join(library_meta, by = "batch")

all_clusters = as.character(sort(as.integer(unique(cell_key$label))))

datatable(library_meta, rownames = FALSE, options = list(dom = "t"))
```

`r paste0("This run has ", dplyr::n_distinct(cell_key$label), " clusters over ", format(nrow(cell_key), big.mark = ","), " cells. Cluster numbering is an output of the clustering and does not correspond to any other run's numbering.")`

# 1. Which clusters can be compared at all?

A cluster needs enough cells on both sides of a comparison to have a marker
profile there. Below the threshold it is not tested, which is a separate outcome
from failing.

```{r}
min_cells_side = 20

side_counts = cell_key %>%
  dplyr::count(label, library, name = "n") %>%
  tidyr::pivot_wider(id_cols = label, names_from = library,
                     values_from = n, values_fill = 0) %>%
  mutate(larva       = Undifferentiated.1 + Undifferentiated.2,
         transformer = Female + Male) %>%
  arrange(as.integer(label))

write.csv(side_counts, file = paste0(results_dir, "/cells_per_cluster_per_side.csv"),
          row.names = FALSE)

datatable(side_counts, rownames = FALSE, options = list(dom = "t"))
```

```{r}
# Core set: clusters testable in the replicate comparison AND in the stage
# comparison. Both are computed on this identical background, because marker
# lists built against different competing cluster sets are not comparable, and
# the whole point is to read the stage result against the replicate one.
core_clusters = side_counts %>%
  filter(Undifferentiated.1 >= min_cells_side,
         Undifferentiated.2 >= min_cells_side,
         transformer        >= min_cells_side) %>%
  pull(label)

# Sex comparison has its own eligible set, since a cluster confined to one sex
# cannot appear in it. Reported separately for that reason.
sex_clusters = side_counts %>%
  filter(Female >= min_cells_side, Male >= min_cells_side) %>%
  pull(label)
```

`r paste0(length(core_clusters), " of ", nrow(side_counts), " clusters have at least ", min_cells_side, " cells in Undifferentiated.1, in Undifferentiated.2 and in the transformers, and carry the replicate and stage comparisons. ", length(sex_clusters), " have at least ", min_cells_side, " cells in both transformer libraries and carry the sex comparison.")`

# 2. Does each cluster find itself across the boundary?

Three comparisons. The first is the reference and the second is the question.

* **Undifferentiated.1 against Undifferentiated.2.** Two animals, same stage,
  same collection date. This is what correspondence looks like when nothing but
  replicate variation separates the sides.
* **Larvae against transformers.** The stage contrast, which is also the
  collection date and batch contrast. Those cannot be separated.
* **Female against Male.** Two animals, same collection date, differing in sex.
  Read alongside the others: a cluster that separates the sexes as strongly as
  it separates the stages is telling you the axis is not stage.

Cell numbers are matched per cluster across every side of every comparison, at
that cluster's own depth, since marker rankings sharpen with cell number and an
unmatched comparison would measure sample size. Library is deliberately **not**
blocked, because library is the variable being measured. The procedure repeats
over independent draws and the similarity matrices are averaged before any
verdict is read off them.

```{r}
n_top      = 50
lfc_thresh = 1
n_draws    = 10

top_markers_for = function(marker_list, lab, n = n_top) {
  if (is.null(marker_list) || !lab %in% names(marker_list)) return(character(0))
  df = marker_list[[lab]]
  if (nrow(df) == 0) return(character(0))

  # sorted explicitly rather than trusting the order findMarkers returns.
  # Nothing would error if that order changed; the top 50 would silently
  # become an arbitrary 50 and every figure here would be wrong.
  ord = if (all(c("FDR", "p.value") %in% colnames(df))) {
    order(df$FDR, df$p.value)
  } else if ("FDR" %in% colnames(df)) {
    order(df$FDR)
  } else seq_len(nrow(df))
  rownames(df)[ord][seq_len(min(n, nrow(df)))]
}

# lfc is set because with thousands of cells a log-fold change of 0.02 clears
# FDR 0.01, and an unthresholded ranking measures sample size rather than
# effect. Both sides of every comparison use the same threshold.
markers_for_cells = function(cells, clusters_keep) {
  lab = factor(cell_key$label[cells], levels = clusters_keep)
  findMarkers(combined.sce[, cells], lab, test.type = "t",
              direction = "up", lfc = lfc_thresh, pval.type = "some")
}

# cluster i of side A against cluster j of side B, for every pair. The diagonal
# is a cluster against itself across the boundary; the off diagonal is what it
# has to beat.
similarity_matrix = function(mA, mB, clusters) {
  tA = lapply(clusters, function(cl) top_markers_for(mA, cl))
  tB = lapply(clusters, function(cl) top_markers_for(mB, cl))
  names(tA) = clusters; names(tB) = clusters

  sim = matrix(NA_real_, length(clusters), length(clusters),
               dimnames = list(clusters, clusters))
  for (i in clusters) for (j in clusters) {
    d = min(length(tA[[i]]), length(tB[[j]]))
    sim[i, j] = if (d == 0) NA_real_ else
      100 * length(intersect(tA[[i]], tB[[j]])) / d
  }
  sim
}
```

```{r}
# one comparison: subsample both sides to matched depth, compute markers on each
# side over the same cluster set, average the similarity matrices over draws
run_comparison = function(sideA_libs, sideB_libs, clusters_keep, seed = 1) {

  inA = which(cell_key$library %in% sideA_libs)
  inB = which(cell_key$library %in% sideB_libs)

  n_by_cluster = sapply(clusters_keep, function(cl)
    min(sum(cell_key$label[inA] == cl), sum(cell_key$label[inB] == cl)))
  names(n_by_cluster) = clusters_keep

  set.seed(seed)
  acc = NULL
  for (d in seq_len(n_draws)) {
    draw_side = function(idx) {
      unlist(lapply(clusters_keep, function(cl) {
        pool = idx[cell_key$label[idx] == cl]
        # sampled by position, never by value: sample(x, n) with x of length 1
        # draws from seq_len(x) instead of from x, which would silently return
        # cell indices that have nothing to do with this cluster
        pool[sample(length(pool), n_by_cluster[[cl]])]
      }))
    }
    mA = markers_for_cells(draw_side(inA), clusters_keep)
    mB = markers_for_cells(draw_side(inB), clusters_keep)
    sim = similarity_matrix(mA, mB, clusters_keep)
    acc = if (is.null(acc)) sim else acc + sim
  }
  list(sim = acc / n_draws, n_by_cluster = n_by_cluster)
}
```

```{r}
comparison_file = paste0(Robject_out_dir, "/stage_correspondence_",
                         n_top, "_lfc", lfc_thresh, "_", n_draws, "draws.Rds")

if (file.exists(comparison_file)) {
  comparisons = readRDS(comparison_file)
} else {
  comparisons = list(
    replicate = run_comparison("Undifferentiated.1", "Undifferentiated.2",
                               core_clusters),
    stage     = run_comparison(c("Undifferentiated.1", "Undifferentiated.2"),
                               c("Female", "Male"), core_clusters),
    sex       = run_comparison("Female", "Male", sex_clusters))
  saveRDS(comparisons, file = comparison_file)
}
```

## Reciprocal self-match

For each cluster, the strongest match on the other side is found in both
directions. A cluster **corresponds** when it is its own best match both ways,
which is stricter than being the best match in one direction, since a diffuse
cluster can be many clusters' best partner without any of them being its own.

The **margin** is how far the self-match sits above the strongest competitor. A
cluster that matches itself by one percentage point has not established
anything, and the margin is the number to read rather than the raw similarity.

```{r}
reciprocal_from_sim = function(sim) {
  cls = rownames(sim)
  do.call(rbind, lapply(cls, function(i) {
    row = sim[i, ]; col = sim[, i]
    off = row[names(row) != i]
    fwd = names(which.max(row)); rev = names(which.max(col))
    # a cluster with no measurable competitor has no margin to report, and
    # max() of an all-NA vector would hand back -Inf rather than saying so
    best_off = if (all(is.na(off))) NA_real_ else max(off, na.rm = TRUE)
    data.frame(
      cluster         = i,
      self_similarity = sim[i, i],
      best_forward    = if (length(fwd)) fwd else NA_character_,
      best_reverse    = if (length(rev)) rev else NA_character_,
      reciprocal      = identical(fwd, i) && identical(rev, i),
      margin          = sim[i, i] - best_off,
      stringsAsFactors = FALSE)
  }))
}

recip = bind_rows(lapply(names(comparisons), function(nm)
  reciprocal_from_sim(comparisons[[nm]]$sim) %>%
    mutate(comparison = nm,
           n_per_side = comparisons[[nm]]$n_by_cluster[cluster]))) %>%
  dplyr::select(comparison, cluster, n_per_side, self_similarity,
                best_forward, best_reverse, reciprocal, margin)

write.csv(recip, file = paste0(results_dir, "/SuppTable_reciprocal_matches.csv"),
          row.names = FALSE)

datatable(recip %>% mutate(across(where(is.numeric), ~round(., 1))),
          filter = "top", rownames = FALSE)
```

```{r}
summary_by_comparison = recip %>%
  group_by(comparison) %>%
  summarize(n_clusters      = dplyr::n(),
            n_reciprocal    = sum(reciprocal),
            pct_reciprocal  = 100 * mean(reciprocal),
            median_self     = median(self_similarity, na.rm = TRUE),
            median_margin   = median(margin, na.rm = TRUE),
            .groups = "drop")

datatable(summary_by_comparison %>% mutate(across(where(is.numeric), ~round(., 1))),
          rownames = FALSE, options = list(dom = "t", ordering = FALSE))
```

```{r}
rep_row   = summary_by_comparison %>% filter(comparison == "replicate")
stage_row = summary_by_comparison %>% filter(comparison == "stage")
sex_row   = summary_by_comparison %>% filter(comparison == "sex")

fails = recip %>% filter(comparison == "stage", !reciprocal)
```

`r if (nrow(rep_row) == 0 || nrow(stage_row) == 0) "One of the two comparisons could not be run, so no reference is available." else paste0("**Reference.** Between the two larval libraries, ", rep_row$n_reciprocal, " of ", rep_row$n_clusters, " clusters are reciprocal self-matches, median similarity ", round(rep_row$median_self, 1), "% and median margin ", round(rep_row$median_margin, 1), " points. That is what correspondence looks like when only replicate animals separate the sides, and it is the ceiling for this dataset rather than 100%. **Stage.** Across the stage boundary, ", stage_row$n_reciprocal, " of ", stage_row$n_clusters, " clusters are reciprocal self-matches, median similarity ", round(stage_row$median_self, 1), "% and median margin ", round(stage_row$median_margin, 1), " points.")`

`r if (nrow(sex_row) == 0) "" else paste0("**Sex.** Between the two transformer libraries, ", sex_row$n_reciprocal, " of ", sex_row$n_clusters, " clusters are reciprocal self-matches. This comparison runs on its own cluster set, since a cluster confined to one sex cannot appear in it, so the count is not directly comparable with the other two. The median margin, ", round(sex_row$median_margin, 1), " points, is comparable.")`

## The similarity matrices

The diagonal is each cluster against itself across the boundary. Off-diagonal
brightness means a cluster's markers resemble a different cluster on the other
side, which is what a failure of correspondence looks like.

```{r, fig.width=13, fig.height=5}
sim_long = bind_rows(lapply(names(comparisons), function(nm) {
  s = comparisons[[nm]]$sim
  as.data.frame(as.table(s)) %>%
    setNames(c("side_A", "side_B", "similarity")) %>%
    mutate(comparison = nm)
})) %>%
  mutate(side_A = factor(side_A, levels = all_clusters),
         side_B = factor(side_B, levels = all_clusters))

p_sim = ggplot(sim_long, aes(x = side_B, y = side_A, fill = similarity)) +
  geom_tile() +
  scale_fill_viridis_c(name = "top 50\noverlap %") +
  facet_wrap(~ comparison, nrow = 1, scales = "free") +
  xlab("cluster, second side") + ylab("cluster, first side") +
  theme(axis.text = element_text(size = 7))

p_sim
ggsave(p_sim, file = paste0(results_dir, "/cross_boundary_similarity.png"),
       width = 13, height = 5)
```

```{r, fig.width=9, fig.height=5}
p_margin = recip %>%
  mutate(cluster = factor(cluster, levels = all_clusters)) %>%
  ggplot(aes(x = cluster, y = margin, fill = reciprocal)) +
  geom_col(position = position_dodge()) +
  facet_wrap(~ comparison, ncol = 1) +
  scale_fill_manual(values = c(`TRUE` = "#7fbf7b", `FALSE` = "#d6604d"),
                    name = "reciprocal\nself-match") +
  geom_hline(yintercept = 0, linewidth = .3) +
  ylab("margin over strongest competitor, points") + xlab("cluster")

p_margin
ggsave(p_margin, file = paste0(results_dir, "/self_match_margin.png"),
       width = 9, height = 5)
```

# 3. Verdict

```{r}
stage_holds = nrow(stage_row) == 1 && nrow(rep_row) == 1 &&
              stage_row$n_reciprocal == stage_row$n_clusters
stage_mixed = nrow(stage_row) == 1 && stage_row$n_reciprocal > 0 &&
              stage_row$n_reciprocal < stage_row$n_clusters
```

`r if (nrow(stage_row) == 0) "**This document cannot decide.** The stage comparison did not run." else if (stage_holds) paste0("**Do not split the atlas.** Every one of the ", stage_row$n_clusters, " comparable clusters finds itself across the stage boundary by its own markers, in both directions. Those clusters are the same populations in larval and in transformer cells. The claim holds whatever the larva-to-transformer contrast is made of, which is why it survives the confound between stage, collection date and batch. Report one integrated atlas and state the confound as a limitation the design cannot resolve.") else if (stage_mixed) paste0("**Integrate, but name the exceptions.** ", stage_row$n_reciprocal, " of ", stage_row$n_clusters, " comparable clusters find themselves across the stage boundary. The ", nrow(fails), " that do not are: ", paste(fails$cluster, collapse = ", "), ". Report the integrated atlas, name those clusters in the text, and either describe them per stage or merge them with whichever cluster they match instead.") else "**Report the stages separately.** No comparable cluster finds itself across the stage boundary by its own markers, so the integrated clustering groups cells the marker evidence does not support grouping."`

**Read the stage result against the replicate result, not against 100%.** Marker
rankings shuffle between random halves of identical data, so some loss of
correspondence is expected between any two groups of cells. The replicate
comparison measures that loss with nothing but animal-to-animal variation
driving it, and the stage comparison should be judged by how far it falls below
that, not by its absolute value.

## What this document does not answer

* **What the stage difference is made of.** Stage, collection date and technical
  batch are one variable in this design and no analysis of this dataset can
  separate them. This test does not attempt to; it asks whether the populations
  correspond, not why they differ.
* **Whether the cell type names are right.** Correspondence says cluster 8 is
  the same population at both stages. It says nothing about what cluster 8 is.
  That question needs the annotation, which is settled elsewhere.
* **Independent evidence for the stage-associated expression patterns.** That
  requires an external dataset or staining, not this dataset.

# Session info

```{r}
sessionInfo()
```
