---
title: "11_no_doublets_one_atlas"
author: "Stephanie Hickey"
date: "`r Sys.Date()`"
output:
  html_document:
    code_folding: hide
    toc: true
    toc_float:
      toc_collapsed: true
params:
    project_dir: /mnt/ufs18/rs-013/bioinformaticsCore/projects/chong_davidson/BCC125_sea_lamprey_scRNA
    project_name: BCC125_sea_lamprey_scRNA
    lab: Weiming Li
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

```{r, echo=FALSE, results="asis"}
cat("\n")
cat("#### Lab:", params$lab, " \n")
cat("#### Project:", params$project_name, " \n")
```

```{r}
library(tidyverse)
library(SingleCellExperiment)
library(scater)
library(scran)
library(patchwork)
library(RColorBrewer)
library(DT)
source(paste0(params$project_dir, "/src/scRNA_seq_functions.R"))
```

```{r}
results_dir     = paste0(params$project_dir, "/results/20260812_no_doublets_one_atlas")
Robject_out_dir = paste0(results_dir, "/Robjects")
if (!dir.exists(Robject_out_dir)) dir.create(Robject_out_dir, recursive = TRUE)
data_dir    = paste0(params$project_dir, "/data")
no_dbl_dir  = paste0(params$project_dir, "/results/20260803_no_doublets")
```

# Purpose

**Should the four doublet-filtered libraries be reported as one integrated
atlas, or as separate atlases?**

Two different questions decide that, and they are easy to confuse. This
document runs both, on the doublet-filtered clustering, and nothing else.

**Question 1: how large is the batch effect?** Section 2 measures it. Marker
genes are computed separately within each library and the agreement between
libraries is compared. The comparison that matters is against a **replicate
floor**: Undifferentiated.1 against Undifferentiated.2 are the same stage, same
collection date and differ only by animal and library preparation, so whatever
disagreement appears there is what "no batch effect" looks like in this data.
Larvae against transformers is then read relative to that floor rather than
against zero.

**Question 2: does cell type identity survive it?** Section 3 measures that,
using the curated markers. A cluster is one cell type across the merge if it
expresses the canonical markers for its annotated type in larval cells and in
transformer cells **separately**.

**Why both are needed, and why either alone would mislead.** The two can
disagree, and the combination is what determines the answer:

| batch effect | identity holds | what it means |
| --- | --- | --- |
| small | yes | integrate freely, correction barely matters |
| **large** | **yes** | the cells are the same types but heavily remodeled between stages. One atlas is defensible **with** correction, and the remodeling is a result to report, not a problem to remove |
| large | no | the clusters do not name the same things at both stages. Report the stages separately |
| small | no | unlikely; would suggest the annotation rather than the batch is at fault |

Question 1 on its own cannot decide anything, because a cell type that remodels
its transcriptome between life stages will show large discordance while
remaining the same cell type. That is biology, not a batch artifact, and this
design cannot separate the two since stage is perfectly confounded with
collection date and batch. Question 2 is what distinguishes them.

**What decides nothing.** A cluster present in only one stage. That is a
stage-restricted population, a result rather than a failure, and it is counted
in its own category rather than folded into either the passes or the failures.

Everything else in `06`, the MST node composition, the pseudotime distributions
and the pseudotime marker trends, is trajectory machinery and is not here.

# Load

```{r}
merged.sce   = readRDS(paste0(no_dbl_dir, "/merged.sce.Rds"))
combined.sce = readRDS(paste0(no_dbl_dir, "/combined.sce.Rds"))

# combined.sce carries the uncorrected logcounts and merged.sce the clustering.
# All quantification below is on the uncorrected values; the correction only
# ever affected clustering and embedding.
stopifnot(identical(colnames(combined.sce), colnames(merged.sce)))

gene_description = read.csv(paste0(no_dbl_dir, "/gene_descritption.csv"),
                            stringsAsFactors = FALSE)
```

```{r}
library_meta = data.frame(
  batch   = c("UND1_C00004336", "UND2_C00004362", "TF_C00004337", "TM_C00004374"),
  library = c("Undifferentiated.1", "Undifferentiated.2", "Female", "Male"),
  stage   = c("larva", "larva", "transformer", "transformer"),
  stringsAsFactors = FALSE)

observed   = sort(unique(as.character(colData(merged.sce)$batch)))
unexpected = setdiff(observed, library_meta$batch)

if (length(unexpected) > 0) {
  stop("Library identifier(s) in the object but not in library_meta: ",
       paste(unexpected, collapse = ", "),
       ". Add them rather than silently dropping cells.")
}

cell_key = data.frame(barcode = colnames(combined.sce),
                      label   = as.character(colData(merged.sce)$label),
                      batch   = as.character(colData(merged.sce)$batch),
                      stringsAsFactors = FALSE) %>%
  left_join(library_meta, by = "batch")

datatable(library_meta, rownames = FALSE)
```

# Annotations and their marker sets

The doublet-filtered clustering yields 20 clusters and its numbering does not
match `01_all_cells_clustering_annotation.Rmd`. These labels are taken from
`01_all_cells_clustering_annotation_no_doublets.Rmd`.

```{r}
anno = c(
  "1. Fibroblasts (migrating)",
  "2. Mesonephric cells",
  "3. Granulosa cells (migrating)",
  "4. Granulosa cells",
  "5. Pre-granulosa cells (migrating)",
  "6. Endothelial cells",
  "7. Podocytes (mixed)",
  "8. Oogonia/Oocytes",
  "9. Erythrocytes",
  "10. Leukocytes",
  "11. Hemopoietic stem cells (1)",
  "12. Hemopoietic stem cells (2)",
  "13. Unresolved (globin-high, mixed origin)",
  "14. Macrophages",
  "15. Pre-granulosa cells (mitotic)",
  "16. Immune cells (migrating, mitotic)",
  "17. Spermatocytes (cytokinetic)",
  "18. Male germ cells (migrating)",
  "19. Spermatocytes (apoptotic)",
  "20. Spermatocytes (GPCR-activated)")

stopifnot(length(anno) == nlevels(colData(merged.sce)$label))

cluster_to_set = tribble(
  ~label, ~marker_set,
  "1",  "Fibroblasts",
  "2",  "Mesonephric cells",
  "3",  "Granulosa cells",
  "4",  "Granulosa cells",
  "5",  "Pre-granulosa cells",
  "6",  "Endothelial cells",
  "7",  "Mesenephric podocytes",
  "8",  "Oocyte",
  "9",  "Erythrocytes",
  "10", NA_character_,
  "11", "Hemopoietic stem cells",
  "12", "Hemopoietic stem cells",
  "13", NA_character_,
  "14", "Macrophages",
  "15", "Pre-granulosa cells",
  "16", NA_character_,
  "17", "Spermatocytes",
  "18", "Germ cells (pan-germline)",
  "19", "Spermatocytes",
  "20", "Spermatocytes") %>%
  mutate(CellType = anno[as.integer(label)],
         testable = !is.na(marker_set))

datatable(cluster_to_set, rownames = FALSE)

untestable = cluster_to_set %>% filter(!testable)
```

`r if (nrow(untestable) == 0) "Every cluster has a corresponding marker set." else paste0(nrow(untestable), " cluster(s) have no marker set and cannot be tested: ", paste(untestable$CellType, collapse = "; "), ". The immune clusters have none because lamprey lymphocytes are identified by variable lymphocyte receptors rather than immunoglobulin-based receptors, and the VLR loci are not present in this genome annotation under a recognizable name. Cluster 13 has none because it carries no cell type assignment to test.")`

## Canonical markers

The repaired long-format list. Three of the four erythrocyte entries and the
oocyte `ZAR1` entry in the original S1 sheet resolve against nothing, so a check
run on that sheet silently tests fewer genes than it reports.

```{r}
canonical = read.csv(paste0(data_dir, "/canonical_markers_long_2026-08-12.csv"),
                     stringsAsFactors = FALSE) %>%
  mutate(symbol = trimws(symbol), loc_id = trimws(loc_id)) %>%
  filter(symbol != "")

resolve_canonical = function(symbol, loc_id, rn, gd) {
  if (!is.na(loc_id) && loc_id %in% rn) return(loc_id)
  if (symbol %in% rn) return(symbol)
  g = gd$Gene[gd$WangGeneID == symbol]
  g = g[g %in% rn]
  if (length(g) > 0) return(g[1])
  NA_character_
}

rn_all = rownames(combined.sce)

canonical = canonical %>%
  rowwise() %>%
  mutate(gene_id = resolve_canonical(symbol, loc_id, rn_all, gene_description)) %>%
  ungroup()

canon_resolved   = canonical %>% filter(!is.na(gene_id))
canon_unresolved = canonical %>% filter(is.na(gene_id))

datatable(canon_resolved %>%
            dplyr::count(CellTypeMarkerSet, name = "n_markers"),
          rownames = FALSE)
```

* Canonical markers listed: `r nrow(canonical)`
* Resolved in `rownames(combined.sce)`: `r nrow(canon_resolved)`
* Not found, and therefore excluded: `r if (nrow(canon_unresolved) == 0) "none" else paste(unique(canon_unresolved$symbol), collapse = ", ")`

# 1. Which clusters are present in both stages?

Before any marker is tested, how many cells each cluster has in each stage. A
cluster below the threshold in a stage cannot be tested there, and that is a
separate outcome from failing.

```{r}
min_cells_stage = 10

stage_cells = cell_key %>%
  dplyr::count(label, stage, name = "n_cells") %>%
  tidyr::complete(label, stage = c("larva", "transformer"),
                  fill = list(n_cells = 0)) %>%
  tidyr::pivot_wider(id_cols = label, names_from = stage,
                     values_from = n_cells, values_fill = 0) %>%
  dplyr::rename(n_larva = larva, n_transformer = transformer) %>%
  mutate(CellType = anno[as.integer(label)],
         in_both  = n_larva >= min_cells_stage &
                    n_transformer >= min_cells_stage) %>%
  arrange(as.integer(label))

write.csv(stage_cells, file = paste0(results_dir, "/cells_per_cluster_per_stage.csv"),
          row.names = FALSE)

datatable(stage_cells, rownames = FALSE)
```

`r paste0(sum(stage_cells$in_both), " of ", nrow(stage_cells), " clusters have at least ", min_cells_stage, " cells in both stages. The remaining ", sum(!stage_cells$in_both), " are stage restricted and cannot speak to whether an integrated atlas is defensible, in either direction.")`

```{r, fig.width=8, fig.height=5}
p_stage = cell_key %>%
  dplyr::count(label, stage, name = "n") %>%
  mutate(label = factor(label, levels = as.character(1:length(anno)))) %>%
  ggplot(aes(x = n, y = label, fill = stage)) +
  geom_col(position = "fill") +
  scale_fill_brewer(palette = "Set2") +
  xlab("proportion of cells") + ylab("cluster")

p_stage
ggsave(p_stage, file = paste0(results_dir, "/cluster_composition_by_stage.png"),
       width = 8, height = 5)
```

# 2. How large is the batch effect?

Marker genes are computed separately within each single library and the
agreement between libraries is measured, with two metrics used throughout:

* **Top 50 overlap**, the percentage of a cluster's fifty strongest markers
  shared between two groups. Easy to read, but it depends on an arbitrary
  cutoff, and marker rankings shuffle even between random halves of identical
  data, so a raw percentage means nothing without a reference.
* **Spearman rank correlation across all tested genes.** Uses all the
  information rather than a cutoff and is far less sensitive to cell number.

Three comparisons, two of which are references:

* **Undifferentiated.1 versus Undifferentiated.2.** Same stage, same collection
  date, no sex difference. Disagreement here is produced by replicate animals
  and library preparation alone, so it is the **replicate floor**. Everything
  else is reported relative to it.
* **Female versus Male.** Same stage, same collection date, different sex. A
  second floor, measured in the November batch, so the two floors together show
  whether replicate variation is stable across collections. Any cluster
  reaching this comparison has cells in both transformer libraries, which
  excludes sex-specific populations. The caveat: these differ by animal *and*
  by sex, so any residual sex effect in shared somatic types is included,
  making this an upper bound on replicate variation rather than a clean floor.
* **Larvae pooled versus transformers pooled.** The stage contrast, which is
  also the collection date contrast. The two cannot be separated in this
  design.

Cell numbers are matched across every group and comparison, since otherwise
disagreement tracks statistical power rather than biology. Each cluster is
subsampled to its own depth, the smallest count available for it in any single
library, and the whole procedure repeats over independent draws. Library is
deliberately **not** blocked here, because library is the variable being
measured.

```{r}
top_markers_for = function(marker_list, lab, n = 50) {
  if (is.null(marker_list) || !lab %in% names(marker_list)) return(character(0))
  df = marker_list[[lab]]
  if (nrow(df) == 0) return(character(0))

  # sorted explicitly rather than trusting the order findMarkers returns.
  # Nothing would error if that order changed; the top 50 would silently
  # become an arbitrary 50 and every figure here would be wrong.
  ord = if (all(c("FDR", "p.value") %in% colnames(df))) {
    order(df$FDR, df$p.value)
  } else if ("FDR" %in% colnames(df)) {
    order(df$FDR)
  } else seq_len(nrow(df))
  rownames(df)[ord][seq_len(min(n, nrow(df)))]
}

# per-gene effect size: the mean across every pairwise comparison. summary.logFC
# is not used because under pval.type = "some" it comes from whichever pairwise
# comparison supplied the summary p-value, which varies per gene and per run, so
# correlating two runs of it compares different contrasts.
marker_effect = function(df) {
  lf = grep("^logFC\\.", colnames(df), value = TRUE)
  if (length(lf) == 0) return(NULL)
  v = rowMeans(as.matrix(as.data.frame(df[, lf, drop = FALSE])), na.rm = TRUE)
  names(v) = rownames(df)
  v
}

top_overlap_pct = function(m1, m2, lab, n = 50) {
  t1 = top_markers_for(m1, lab, n); t2 = top_markers_for(m2, lab, n)
  denom = min(length(t1), length(t2))
  if (denom == 0) return(NA_real_)
  100 * length(intersect(t1, t2)) / denom
}

stat_rank_cor = function(m1, m2, lab) {
  if (is.null(m1) || is.null(m2)) return(NA_real_)
  if (!lab %in% names(m1) || !lab %in% names(m2)) return(NA_real_)
  e1 = marker_effect(m1[[lab]]); e2 = marker_effect(m2[[lab]])
  if (is.null(e1) || is.null(e2)) return(NA_real_)
  g = intersect(names(e1), names(e2))
  if (length(g) < 10) return(NA_real_)
  suppressWarnings(cor(e1[g], e2[g], method = "spearman"))
}
```

```{r}
library_discordance = function(sce, labels, library_vec, min_cells = 20,
                               n_draws = 10, n_top = 50, seed = 1) {
  labels      = as.character(labels)
  library_vec = as.character(library_vec)

  singles = c("Undifferentiated.1", "Undifferentiated.2", "Female", "Male")
  group_defs = list(
    "Undifferentiated.1"  = "Undifferentiated.1",
    "Undifferentiated.2"  = "Undifferentiated.2",
    "Female"              = "Female",
    "Male"                = "Male",
    "larvae_pooled"       = c("Undifferentiated.1", "Undifferentiated.2"),
    "transformers_pooled" = c("Female", "Male"))

  comparisons = list(
    UND1_vs_UND2           = c("Undifferentiated.1", "Undifferentiated.2"),
    Female_vs_Male         = c("Female", "Male"),
    larvae_vs_transformers = c("larvae_pooled", "transformers_pooled"))

  clusters_all = sort(unique(labels))
  cnt = sapply(singles, function(lb)
    sapply(clusters_all, function(cl) sum(labels == cl & library_vec == lb)))
  if (is.null(dim(cnt))) cnt = matrix(cnt, nrow = length(clusters_all),
                                      dimnames = list(clusters_all, singles))
  rownames(cnt) = clusters_all

  eligibility = as.data.frame(cnt)
  eligibility$label = clusters_all
  eligibility$min_across_libraries = apply(cnt, 1, min)
  eligibility$eligible = eligibility$min_across_libraries >= min_cells
  rownames(eligibility) = NULL

  eligible = clusters_all[eligibility$eligible]
  if (length(eligible) < 2) {
    warning("Fewer than two clusters have enough cells in every library.")
    return(list(results = NULL, eligibility = eligibility,
                n_by_cluster = integer(0), eligible = eligible))
  }

  # each cluster matched at its own depth: the smallest count available for it
  # in any single library. What the comparison needs is equal power between the
  # groups being compared, and that holds cluster by cluster. A single global n
  # would force every cluster down to the sparsest one for no gain, since
  # clusters are never compared against each other here.
  n_by_cluster = apply(cnt[eligible, , drop = FALSE], 1, min)
  names(n_by_cluster) = eligible

  set.seed(seed)
  res = list()

  for (d in seq_len(n_draws)) {
    markers_by_group = list()
    for (gname in names(group_defs)) {
      libs = group_defs[[gname]]
      idx  = integer(0); ok = TRUE
      for (cl in eligible) {
        pool = which(labels == cl & library_vec %in% libs)
        if (length(pool) < n_by_cluster[[cl]]) { ok = FALSE; break }
        idx = c(idx, pool[sample(length(pool), n_by_cluster[[cl]])])
      }
      if (!ok) next
      markers_by_group[[gname]] = findMarkers(
        sce[, idx], droplevels(factor(labels[idx])),
        test.type = "t", direction = "up", pval.type = "some")
    }

    for (cmp in names(comparisons)) {
      g1 = comparisons[[cmp]][1]; g2 = comparisons[[cmp]][2]
      if (is.null(markers_by_group[[g1]]) ||
          is.null(markers_by_group[[g2]])) next
      res[[length(res) + 1]] = data.frame(
        comparison  = cmp, draw = d, label = eligible,
        n_per_group = as.integer(n_by_cluster[eligible]),
        overlap_pct = vapply(eligible, function(cl)
          top_overlap_pct(markers_by_group[[g1]], markers_by_group[[g2]],
                          cl, n_top), numeric(1)),
        rank_cor    = vapply(eligible, function(cl)
          stat_rank_cor(markers_by_group[[g1]], markers_by_group[[g2]], cl),
          numeric(1)),
        stringsAsFactors = FALSE)
    }
  }

  out = do.call(rbind, res)
  if (!is.null(out)) rownames(out) = NULL
  list(results = out, eligibility = eligibility,
       n_by_cluster = n_by_cluster, eligible = eligible)
}
```

```{r}
discordance_file = paste0(Robject_out_dir, "/library_discordance.Rds")

if (file.exists(discordance_file)) {
  discordance = readRDS(discordance_file)
} else {
  discordance = library_discordance(combined.sce,
                                    colData(merged.sce)$label,
                                    cell_key$library)
  saveRDS(discordance, file = discordance_file)
}
```

## Cells used, and clusters excluded

```{r}
write.csv(discordance$eligibility,
          file = paste0(results_dir, "/SuppTable_discordance_eligibility.csv"),
          row.names = FALSE)

datatable(discordance$eligibility, filter = "top", rownames = FALSE)

disc_excluded = discordance$eligibility %>% filter(!eligible)
n_low_power   = if (length(discordance$n_by_cluster) == 0) 0 else
  sum(discordance$n_by_cluster < 50)
```

`r if (length(discordance$n_by_cluster) == 0) "**Fewer than two clusters have enough cells in all four libraries, so this analysis cannot run.** That is a result rather than an error, and it would mean the clusters are too library-restricted for a within-library marker comparison to be possible at all." else paste0(length(discordance$eligible), " cluster(s) are eligible, matched at ", min(discordance$n_by_cluster), " to ", max(discordance$n_by_cluster), " cells per cluster per group (median ", round(median(discordance$n_by_cluster)), ").")`

`r if (nrow(disc_excluded) == 0) "No cluster was excluded." else paste0(nrow(disc_excluded), " cluster(s) were excluded for having too few cells in at least one library. A cluster confined to one sex, such as the oocyte and spermatocyte clusters, cannot appear here at all, since the Female versus Male comparison requires cells of that cluster in both transformer libraries. Their absence is structural and says nothing about the batch effect.")`

`r if (n_low_power == 0) "" else paste0(n_low_power, " cluster(s) are matched at fewer than 50 cells per group. Marker rankings are noisy at that depth, so their values carry little information in either direction and should not be read as evidence of agreement or disagreement.")`

## Discordance per cluster

Both metrics measure agreement, so **lower values mean greater discordance**.

```{r}
disc_results = discordance$results
if (is.null(disc_results)) {
  disc_results = data.frame(
    comparison = character(0), draw = integer(0), label = character(0),
    n_per_group = integer(0), overlap_pct = numeric(0),
    rank_cor = numeric(0), stringsAsFactors = FALSE)
}

disc_per_cluster = disc_results %>%
  group_by(label, comparison) %>%
  summarize(n_per_group    = dplyr::first(n_per_group),
            n_draws        = dplyr::n(),
            overlap_median = median(overlap_pct, na.rm = TRUE),
            overlap_min    = min(overlap_pct, na.rm = TRUE),
            overlap_max    = max(overlap_pct, na.rm = TRUE),
            rankcor_median = median(rank_cor, na.rm = TRUE),
            rankcor_min    = min(rank_cor, na.rm = TRUE),
            rankcor_max    = max(rank_cor, na.rm = TRUE),
            .groups = "drop")

write.csv(disc_per_cluster,
          file = paste0(results_dir, "/SuppTable_library_discordance.csv"),
          row.names = FALSE)

datatable(disc_per_cluster %>% mutate(across(where(is.numeric), ~round(., 3))),
          filter = "top", rownames = FALSE)
```

```{r, fig.width=10, fig.height=7}
cmp_cols = c(UND1_vs_UND2 = "#009292", Female_vs_Male = "#006ddb",
             larvae_vs_transformers = "#9e2f7f")

disc_long = disc_per_cluster %>%
  tidyr::pivot_longer(
    c(overlap_median, overlap_min, overlap_max,
      rankcor_median, rankcor_min, rankcor_max),
    names_to = c("metric", "stat"), names_sep = "_", values_to = "value") %>%
  tidyr::pivot_wider(names_from = stat, values_from = value) %>%
  mutate(metric = dplyr::recode(metric,
                                overlap = "Top 50 marker overlap (%)",
                                rankcor = "Spearman rank correlation"),
         label  = factor(label, levels = as.character(sort(as.integer(
           unique(disc_per_cluster$label))))))

p_disc = ggplot(disc_long, aes(x = label, y = median, color = comparison)) +
  geom_linerange(aes(ymin = min, ymax = max),
                 position = position_dodge(width = .6), linewidth = .6) +
  geom_point(position = position_dodge(width = .6), size = 2) +
  scale_color_manual(values = cmp_cols) +
  facet_wrap(~ metric, ncol = 1, scales = "free_y") +
  xlab("cluster") + ylab(NULL) +
  labs(caption = paste0("Points are medians across draws, bars the range. ",
                        "UND1 vs UND2 is the replicate floor. ",
                        "Lower means more discordant."))

p_disc
ggsave(p_disc, file = paste0(results_dir, "/library_discordance_by_cluster.png"),
       width = 10, height = 7)
```

## Is the stage contrast worse than the replicate floor?

Each cluster is compared against **its own** floor rather than against a pooled
average. Clusters differ enormously in how reproducible their markers are: a
cluster with sharply distinctive markers scores high in every comparison and a
diffuse one scores low in every comparison. Pooling lets that between-cluster
spread swamp the difference being measured, in the same way an unpaired test
lets subject-to-subject variation swamp a treatment effect.

```{r}
paired = disc_per_cluster %>%
  dplyr::select(label, comparison, n_per_group,
                overlap_median, rankcor_median) %>%
  tidyr::pivot_longer(c(overlap_median, rankcor_median),
                      names_to = "metric", values_to = "value") %>%
  tidyr::pivot_wider(id_cols = c(label, n_per_group, metric),
                     names_from = comparison, values_from = value)

paired_test = if (!"UND1_vs_UND2" %in% colnames(paired) || nrow(paired) == 0) {
  data.frame(metric = character(0), comparison = character(0),
             n_clusters = integer(0), median_floor = numeric(0),
             median_comparison = numeric(0), median_ratio = numeric(0),
             n_below_floor = integer(0), wilcox_p = numeric(0),
             exceeds_floor = logical(0), stringsAsFactors = FALSE)
} else {
  paired %>%
    filter(!is.na(UND1_vs_UND2)) %>%
    tidyr::pivot_longer(any_of(c("Female_vs_Male", "larvae_vs_transformers")),
                        names_to = "comparison", values_to = "value") %>%
    filter(!is.na(value)) %>%
    group_by(metric, comparison) %>%
    summarize(
      n_clusters        = dplyr::n(),
      median_floor      = median(UND1_vs_UND2),
      median_comparison = median(value),
      median_ratio      = median(value / UND1_vs_UND2, na.rm = TRUE),
      n_below_floor     = sum(value < UND1_vs_UND2),
      # a signed rank test needs a handful of pairs to say anything; below
      # that it reports NA rather than a meaningless p-value
      wilcox_p          = if (dplyr::n() >= 5)
        suppressWarnings(wilcox.test(value, UND1_vs_UND2,
                                     paired = TRUE)$p.value) else NA_real_,
      .groups = "drop") %>%
    mutate(exceeds_floor = median_ratio < 1 & n_below_floor > n_clusters / 2)
}

write.csv(paired_test,
          file = paste0(results_dir, "/SuppTable_discordance_paired_test.csv"),
          row.names = FALSE)

datatable(paired_test %>% mutate(across(where(is.numeric), ~signif(., 3))),
          rownames = FALSE)
```

`exceeds_floor` is TRUE when a comparison agrees **less** well than the
replicate floor: the median per-cluster ratio is below 1 and more than half the
clusters individually fall below their own floor.

```{r}
lt_row = paired_test %>%
  filter(comparison == "larvae_vs_transformers", metric == "overlap_median")
fm_row = paired_test %>%
  filter(comparison == "Female_vs_Male", metric == "overlap_median")

batch_effect_large = nrow(lt_row) == 1 && isTRUE(lt_row$exceeds_floor)
```

`r if (nrow(lt_row) == 0) "The stage contrast could not be tested, so the size of the batch effect is not established here." else paste0("**Stage contrast.** Median top 50 overlap is ", round(lt_row$median_comparison, 1), "% against a replicate floor of ", round(lt_row$median_floor, 1), "%, a ratio of ", round(lt_row$median_ratio, 2), ", with ", lt_row$n_below_floor, " of ", lt_row$n_clusters, " clusters below their own floor", if (!is.na(lt_row$wilcox_p)) paste0(" (Wilcoxon signed rank p = ", signif(lt_row$wilcox_p, 2), ")") else "", ". ", if (isTRUE(lt_row$exceeds_floor)) "**The stage contrast is substantially more discordant than replicate libraries of the same stage.** Expression is heavily remodeled across the stage boundary, so an uncorrected integration would cluster largely by stage." else "**The stage contrast is not more discordant than the replicate floor**, so there is little to correct for.")`

`r if (nrow(fm_row) == 0) "" else paste0("**Second floor.** Female versus Male gives a median overlap of ", round(fm_row$median_comparison, 1), "%, a ratio of ", round(fm_row$median_ratio, 2), " to the July floor. If the two floors are close, both the residual sex effect in shared somatic types and any difference in batch noise between collections are small, which makes the July floor a trustworthy reference for the stage contrast.")`

**This section cannot tell you whether the discordance is batch or biology.**
Stage is perfectly confounded with collection date and library, so a large stage
contrast is equally consistent with a technical batch effect and with genuine
developmental remodeling. That is why section 3 exists: the curated markers ask
whether the cells are still the same types, which is a question the confound
does not touch.

# 3. Canonical marker check

Each cluster is scored only against the markers of its own assigned type, in
each stage separately, on uncorrected depth-normalized values.

```{r}
canon_ids = unique(canon_resolved$gene_id)
canon_mat = as.matrix(logcounts(combined.sce)[canon_ids, , drop = FALSE])

canon_long = t(canon_mat) %>%
  as.data.frame() %>%
  tibble::rownames_to_column("barcode") %>%
  tidyr::pivot_longer(-barcode, names_to = "gene_id", values_to = "logexpr") %>%
  left_join(cell_key, by = "barcode")

canon_scored = canon_long %>%
  inner_join(canon_resolved %>%
               dplyr::select(gene_id, symbol, marker_set = CellTypeMarkerSet),
             by = "gene_id", relationship = "many-to-many") %>%
  inner_join(cluster_to_set %>% filter(testable) %>%
               dplyr::select(label, marker_set, CellType),
             by = c("label", "marker_set"))

canon_summary = canon_scored %>%
  group_by(CellType, label, marker_set, symbol, stage) %>%
  summarize(n_cells        = dplyr::n(),
            mean_logexpr   = mean(logexpr),
            pct_expressing = 100 * mean(logexpr > 0),
            .groups = "drop")

write.csv(canon_summary,
          file = paste0(results_dir, "/SuppTable_canonical_markers_by_stage.csv"),
          row.names = FALSE)

datatable(canon_summary %>% mutate(across(where(is.numeric), ~round(., 2))),
          filter = "top", rownames = FALSE)
```

```{r, fig.width=11, fig.height=9}
p_canon = ggplot(canon_summary,
                 aes(x = stage, y = symbol,
                     size = pct_expressing, color = mean_logexpr)) +
  geom_point() +
  scale_color_viridis_c() +
  scale_size_continuous(range = c(0.5, 5)) +
  facet_wrap(~ paste0(label, ". ", marker_set), scales = "free_y") +
  theme(axis.text.y = element_text(size = 6),
        axis.text.x = element_text(angle = 45, hjust = 1)) +
  xlab(NULL) + ylab(NULL) +
  labs(size = "% expressing", color = "mean logexpr")

p_canon
ggsave(p_canon, file = paste0(results_dir, "/canonical_markers_by_stage.png"),
       width = 11, height = 9)
```

## Verdict

A marker counts as expressed in a stage if it is detected in at least
`detect_pct`% of that cluster's cells from that stage, and identity holds in a
stage if at least `marker_frac`% of the assigned markers are expressed. A
cluster with fewer than `r min_cells_stage` cells in a stage is recorded as not
sampled there, which is its own outcome and is never folded into the passes or
the failures.

```{r}
detect_pct  = 10
marker_frac = 50

# every testable cluster crossed with both stages, so a cluster absent from a
# stage appears as a zero rather than as a missing row. Without the cross it
# would drop out of pivot_wider as NA, be excluded from the pass count by
# na.rm and from the failure list by filter(), and so vanish from the summary
# while still counting in its denominator.
cluster_stage_cells =
  cell_key %>%
  inner_join(cluster_to_set %>% filter(testable) %>%
               dplyr::select(label, marker_set, CellType), by = "label") %>%
  dplyr::count(CellType, label, marker_set, stage, name = "n_cells") %>%
  tidyr::complete(tidyr::nesting(CellType, label, marker_set),
                  stage = c("larva", "transformer"),
                  fill = list(n_cells = 0))

score_stage = function(detect, frac) {
  by_stage = canon_summary %>%
    mutate(expressed = pct_expressing >= detect) %>%
    group_by(CellType, label, marker_set, stage) %>%
    summarize(n_markers   = dplyr::n(),
              n_expressed = sum(expressed),
              pct_markers_expressed = 100 * mean(expressed),
              .groups = "drop")

  cluster_stage_cells %>%
    left_join(by_stage, by = c("CellType", "label", "marker_set", "stage")) %>%
    mutate(status = dplyr::case_when(
      n_cells < min_cells_stage    ~ "not sampled",
      is.na(pct_markers_expressed) ~ "not sampled",
      pct_markers_expressed >= frac ~ "holds",
      TRUE                          ~ "fails")) %>%
    tidyr::pivot_wider(id_cols = c(CellType, label, marker_set),
                       names_from  = stage,
                       values_from = c(n_cells, n_markers, n_expressed,
                                       pct_markers_expressed, status)) %>%
    mutate(verdict = dplyr::case_when(
      status_larva == "holds" & status_transformer == "holds"
        ~ "holds in both stages",
      status_larva == "fails" | status_transformer == "fails"
        ~ "fails in at least one stage",
      status_larva == "not sampled" & status_transformer == "not sampled"
        ~ "not sampled in either stage",
      TRUE
        ~ "present in one stage only"))
}

canon_verdict = score_stage(detect_pct, marker_frac) %>%
  arrange(verdict, as.integer(label))

write.csv(canon_verdict,
          file = paste0(results_dir, "/SuppTable_canonical_identity_verdict.csv"),
          row.names = FALSE)

datatable(canon_verdict %>% mutate(across(where(is.numeric), ~round(., 1))),
          filter = "top", rownames = FALSE)
```

```{r}
n_tested     = nrow(canon_verdict)
n_holds      = sum(canon_verdict$verdict == "holds in both stages")
n_fails      = sum(canon_verdict$verdict == "fails in at least one stage")
n_one_stage  = sum(canon_verdict$verdict == "present in one stage only")
n_unsampled  = sum(canon_verdict$verdict == "not sampled in either stage")
n_comparable = n_holds + n_fails

fails_clusters     = canon_verdict %>% filter(verdict == "fails in at least one stage")
one_stage_clusters = canon_verdict %>% filter(verdict == "present in one stage only")

# no cluster may fall outside the four outcomes
stopifnot(n_holds + n_fails + n_one_stage + n_unsampled == n_tested)
```

Of the `r n_tested` clusters with a marker set, `r n_comparable` have at least
`r min_cells_stage` cells in both stages and can be compared across the merge
at all. Of those, **`r n_holds` express at least half their assigned canonical
markers in both stages** and `r n_fails` do not.

`r if (n_one_stage == 0) "Every testable cluster is present in both stages." else paste0(n_one_stage, " cluster(s) are present in one stage only: ", paste(one_stage_clusters$CellType, collapse = "; "), ". A cluster that exists in one stage cannot agree or disagree with itself across the merge, so these are neither evidence for nor against an integrated atlas and are reported rather than counted.")`

`r if (n_comparable == 0) "**No cluster has enough cells in both stages to be compared, so this test cannot decide the question.**" else if (n_fails == 0) "**Cell type identity is preserved across the stages for every cluster that could be compared.** A single integrated atlas of the doublet-filtered data is defensible: those clusters name the same cell types at both stages. The claim extends only to the clusters present in both stages; the stage-restricted ones are carried on their within-stage markers alone." else paste0("**", n_fails, " cluster(s) do not express half their assigned markers in both stages: ", paste(fails_clusters$CellType, collapse = "; "), ".** For these the annotation does not hold across the merge, and grouping the stages into one cluster asserts an identity the canonical markers do not support. If this affects most of the comparable clusters, describing the stages separately is the more honest presentation.")`

# 4. Does the verdict survive the thresholds?

Both cutoffs are conventions. A conclusion that flips between reasonable
settings is not a conclusion, so the grid below repeats the whole test across a
range of both and reports how many clusters hold in each cell.

```{r}
grid = expand.grid(detect = c(5, 10, 20),
                   frac   = c(33, 50, 67))

sensitivity = purrr::map_dfr(seq_len(nrow(grid)), function(i) {
  v = score_stage(grid$detect[i], grid$frac[i])
  data.frame(detect_pct  = grid$detect[i],
             marker_frac = grid$frac[i],
             holds       = sum(v$verdict == "holds in both stages"),
             fails       = sum(v$verdict == "fails in at least one stage"),
             one_stage   = sum(v$verdict == "present in one stage only"),
             comparable  = sum(v$verdict %in% c("holds in both stages",
                                                "fails in at least one stage")))
}) %>%
  mutate(pct_holding = ifelse(comparable == 0, NA_real_,
                              100 * holds / comparable))

write.csv(sensitivity, file = paste0(results_dir, "/threshold_sensitivity.csv"),
          row.names = FALSE)

datatable(sensitivity %>% mutate(pct_holding = round(pct_holding, 1)),
          rownames = FALSE)
```

```{r, fig.width=6, fig.height=4}
p_sens = ggplot(sensitivity,
                aes(x = factor(detect_pct), y = factor(marker_frac),
                    fill = pct_holding)) +
  geom_tile() +
  geom_text(aes(label = paste0(holds, "/", comparable)), size = 3) +
  scale_fill_viridis_c(limits = c(0, 100)) +
  xlab("detection threshold, % of cells") +
  ylab("markers required, % of set") +
  labs(fill = "% of comparable\nclusters holding")

p_sens
ggsave(p_sens, file = paste0(results_dir, "/threshold_sensitivity.png"),
       width = 6, height = 4)
```

`r paste0("Across the nine settings, the share of comparable clusters holding in both stages ranges from ", round(min(sensitivity$pct_holding, na.rm = TRUE), 1), "% to ", round(max(sensitivity$pct_holding, na.rm = TRUE), 1), "%. ", if (all(sensitivity$fails == 0, na.rm = TRUE)) "No cluster fails at any setting, so the conclusion does not depend on either cutoff." else if (all(sensitivity$holds == 0, na.rm = TRUE)) "No cluster holds at any setting, so the conclusion does not depend on either cutoff." else "The result moves with the cutoffs, so the headline verdict above should be reported together with this grid rather than on its own.")`

# 5. Per library, not just per stage

A cluster can hold in the transformer stage on the strength of one library.
Since TF and TM are different animals and different sexes, the same test run per
library is stricter, and a cluster that holds in every library it is sampled in
is on firmer ground than one that holds only in aggregate.

```{r}
canon_lib = canon_scored %>%
  group_by(CellType, label, marker_set, symbol, library) %>%
  summarize(pct_expressing = 100 * mean(logexpr > 0),
            n_cells        = dplyr::n(),
            .groups = "drop") %>%
  mutate(expressed = pct_expressing >= detect_pct) %>%
  group_by(CellType, label, marker_set, library) %>%
  summarize(n_cells     = dplyr::first(n_cells),
            n_markers   = dplyr::n(),
            pct_markers = 100 * mean(expressed),
            .groups = "drop") %>%
  mutate(status = dplyr::case_when(
    n_cells < min_cells_stage ~ "not sampled",
    pct_markers >= marker_frac ~ "holds",
    TRUE ~ "fails"))

lib_verdict = canon_lib %>%
  group_by(CellType, label) %>%
  summarize(libraries_sampled = sum(status != "not sampled"),
            libraries_holding = sum(status == "holds"),
            .groups = "drop") %>%
  mutate(holds_everywhere = libraries_sampled > 0 &
                            libraries_holding == libraries_sampled) %>%
  arrange(as.integer(label))

write.csv(lib_verdict, file = paste0(results_dir, "/canonical_identity_by_library.csv"),
          row.names = FALSE)

datatable(lib_verdict, rownames = FALSE)
```

```{r, fig.width=8, fig.height=6}
p_lib = ggplot(canon_lib,
               aes(x = library, y = paste0(label, ". ", marker_set),
                   fill = status)) +
  geom_tile(color = "white") +
  geom_text(aes(label = round(pct_markers)), size = 2.5) +
  scale_fill_manual(values = c(holds = "#7fbf7b", fails = "#d6604d",
                               `not sampled` = "grey90")) +
  theme(axis.text = element_text(size = 7),
        axis.text.x = element_text(angle = 30, hjust = 1)) +
  xlab(NULL) + ylab(NULL) +
  labs(fill = NULL)

p_lib
ggsave(p_lib, file = paste0(results_dir, "/canonical_identity_by_library.png"),
       width = 8, height = 6)
```

`r paste0(sum(lib_verdict$holds_everywhere), " of ", nrow(lib_verdict), " testable clusters express their assigned markers in every library in which they are sampled. Numbers in the tiles are the percentage of the assigned marker set expressed in that library.")`

# 6. Summary

```{r}
summary_tbl = data.frame(
  question = c(
    "Q1. stage contrast more discordant than the replicate floor",
    "Q1. clusters eligible for the discordance test",
    "Q2. clusters with a curated marker set",
    "Q2. comparable across stages",
    "Q2. identity holds in both stages",
    "Q2. identity fails in at least one stage",
    "Q2. present in one stage only",
    "Q2. holds in every library sampled"),
  answer = c(
    if (nrow(lt_row) == 0) "not testable" else
      if (batch_effect_large) "yes, batch effect is large" else "no",
    as.character(length(discordance$eligible)),
    as.character(n_tested), as.character(n_comparable),
    as.character(n_holds), as.character(n_fails), as.character(n_one_stage),
    as.character(sum(lib_verdict$holds_everywhere))))

datatable(summary_tbl, rownames = FALSE)
```

## The recommendation

```{r}
identity_holds = n_comparable > 0 && n_fails == 0
identity_mixed = n_comparable > 0 && n_fails > 0 && n_fails < n_holds
identity_fails = n_comparable > 0 && n_fails >= n_holds
```

`r if (n_comparable == 0) "**This document cannot decide.** No cluster has enough cells in both stages for the identity test to run, so neither presentation is supported over the other by this evidence." else if (batch_effect_large && identity_holds) "**Do not ignore the batch effect, and do not split the atlas.** The stage contrast is far more discordant than replicate libraries of the same stage, so an uncorrected integration would cluster largely by stage rather than by cell type. But every cluster that can be compared still expresses its canonical markers in both stages, so the clusters do name the same cell types at both. One integrated atlas **with** correction is the supported presentation, and the large discordance is itself a result to report: these cell types are extensively remodeled between larva and transformer. Since stage is perfectly confounded with collection date and library, the manuscript should say that the remodeling cannot be separated from a technical batch effect, and that the identity test is what licenses the integration." else if (batch_effect_large && identity_fails) "**Report the stages separately.** The stage contrast is far more discordant than the replicate floor, and the curated markers do not hold across it for most comparable clusters, so the integrated clustering is asserting identities the markers do not support." else if (batch_effect_large && identity_mixed) "**Integrate with correction, but name the exceptions.** The stage contrast is much more discordant than the replicate floor, and most comparable clusters keep their identity across it, but a minority do not. Report the integrated atlas, name those clusters in the text, and either split them by stage or relabel them." else if (!batch_effect_large && identity_holds) "**The batch effect is not larger than replicate variation, and identity holds everywhere it can be tested.** One integrated atlas is defensible and the correction is doing little work. Worth confirming that the correction is not removing real signal, since there is little here for it to remove." else "**The batch effect is not larger than replicate variation, yet identity fails for some clusters.** That points at the annotation rather than the batch: a cluster whose curated markers do not hold across stages, in the absence of a large technical difference, is more likely mislabeled than mis-integrated. Revisit those annotations before deciding on the presentation."`

## What this document does not answer

* Whether the **transformer libraries** should be merged with each other. That
  is answered in `10_four_atlases.Rmd`.
* Whether the stages should be **plotted** together. That is a question about
  embedding, not identity.
* Whether the discordance is technical or biological. It cannot be answered
  with this design, because stage, collection date and library are one
  variable. The identity test sidesteps the confound rather than resolving it.

# Session info

```{r}
sessionInfo()
```
