---
title: "11_no_doublets_one_atlas"
author: "Stephanie Hickey"
date: "`r Sys.Date()`"
output:
  html_document:
    code_folding: hide
    toc: true
    toc_float:
      toc_collapsed: true
params:
    project_dir: /mnt/ufs18/rs-013/bioinformaticsCore/projects/chong_davidson/BCC125_sea_lamprey_scRNA
    project_name: BCC125_sea_lamprey_scRNA
    lab: Weiming Li
    no_dbl_date: "20260813"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

```{r, echo=FALSE, results="asis"}
cat("\n")
cat("#### Lab:", params$lab, " \n")
cat("#### Project:", params$project_name, " \n")
```

```{r}
library(tidyverse)
library(SingleCellExperiment)
library(scater)
library(scran)
library(patchwork)
library(RColorBrewer)
library(DT)
source(paste0(params$project_dir, "/src/scRNA_seq_functions.R"))
```

```{r}
no_dbl_dir  = paste0(params$project_dir, "/results/", params$no_dbl_date, "_no_doublets")

results_dir     = paste0(params$project_dir, "/results/", params$no_dbl_date,
                         "_no_doublets_one_atlas")
Robject_out_dir = paste0(results_dir, "/Robjects")
if (!dir.exists(Robject_out_dir)) dir.create(Robject_out_dir, recursive = TRUE)
data_dir    = paste0(params$project_dir, "/data")
```

# Purpose

**Do the clusters of the doublet-filtered atlas name the same populations in
larval and in transformer cells, or should the stages be described separately?**

The test is whether each cluster **finds itself** across the stage boundary
using its own data-driven markers. Marker genes are computed from larval cells
alone and, separately, from transformer cells alone. If the genes that define
cluster 8 among larvae are the genes that define cluster 8 among transformers,
and they match cluster 8 better than they match any other cluster, that is evidence that
cluster 8 is the same population on both sides.

**Why this survives the confound.** Larvae were collected in July and
transformers in November, so stage, collection date and technical batch are one
variable, and nothing computed from this dataset can say how much of the
difference between them is developmental. This test does not try. It asks
whether the populations correspond, not what makes them differ. 

**Everything is read against a reference computed the same way.** Undifferentiated.1
against Undifferentiated.2 is two animals of the same stage, collected on the
same date, so it measures what correspondence looks like when only replicate
variation separates the two sides. That is the ceiling. The stage comparison is
reported relative to it, never against 100%.

# Load

```{r}
merged.sce   = readRDS(paste0(no_dbl_dir, "/merged.sce.Rds"))
combined.sce = readRDS(paste0(no_dbl_dir, "/combined.sce.Rds"))

# combined.sce carries the uncorrected logcounts and merged.sce the clustering.
# All quantification below is on the uncorrected values; the correction only
# ever affected clustering and embedding.
stopifnot(identical(colnames(combined.sce), colnames(merged.sce)))
```

```{r}
library_meta = data.frame(
  batch   = c("UND1_C00004336", "UND2_C00004362", "TF_C00004337", "TM_C00004374"),
  library = c("Undifferentiated.1", "Undifferentiated.2", "Female", "Male"),
  stage   = c("larva", "larva", "transformer", "transformer"),
  stringsAsFactors = FALSE)

observed   = sort(unique(as.character(colData(merged.sce)$batch)))
unexpected = setdiff(observed, library_meta$batch)

if (length(unexpected) > 0) {
  stop("Library identifier(s) in the object but not in library_meta: ",
       paste(unexpected, collapse = ", "),
       ". Add them rather than silently dropping cells.")
}

cell_key = data.frame(barcode = colnames(combined.sce),
                      label   = as.character(colData(merged.sce)$label),
                      batch   = as.character(colData(merged.sce)$batch),
                      stringsAsFactors = FALSE) %>%
  left_join(library_meta, by = "batch")

all_clusters = as.character(sort(as.integer(unique(cell_key$label))))

datatable(library_meta, rownames = FALSE, options = list(dom = "t"))
```

`r paste0("This run has ", dplyr::n_distinct(cell_key$label), " clusters over ", format(nrow(cell_key), big.mark = ","), " cells. Cluster numbering is an output of the clustering and does not correspond to any other run's numbering.")`

# 1. Which clusters can be compared at all?

A cluster needs enough cells on both sides of a comparison to have a marker
profile there. Below the threshold it is not tested, which is a separate outcome
from failing.

```{r}
min_cells_side = 20

side_counts = cell_key %>%
  dplyr::count(label, library, name = "n") %>%
  tidyr::pivot_wider(id_cols = label, names_from = library,
                     values_from = n, values_fill = 0) %>%
  mutate(larva       = Undifferentiated.1 + Undifferentiated.2,
         transformer = Female + Male) %>%
  arrange(as.integer(label))

write.csv(side_counts, file = paste0(results_dir, "/cells_per_cluster_per_side.csv"),
          row.names = FALSE)

# paging = FALSE because dom = "t" hides the pagination control without
# disabling pagination, so a table longer than ten rows silently loses the rest
datatable(side_counts, rownames = FALSE,
          options = list(dom = "t", paging = FALSE))
```

```{r}
# Core set: clusters testable in the replicate comparison AND in the stage
# comparison. Both are computed on this identical background, because marker
# lists built against different competing cluster sets are not comparable, and
# the whole point is to read the stage result against the replicate one.
core_clusters = side_counts %>%
  filter(Undifferentiated.1 >= min_cells_side,
         Undifferentiated.2 >= min_cells_side,
         transformer        >= min_cells_side) %>%
  pull(label)

# Sex comparison has its own eligible set, since a cluster confined to one sex
# cannot appear in it. Reported separately for that reason.
sex_clusters = side_counts %>%
  filter(Female >= min_cells_side, Male >= min_cells_side) %>%
  pull(label)

# One depth per cluster, shared by the replicate and stage comparisons.
#
# This matters more than it looks. Subsampling each comparison to its own
# minimum sounds right and is not: the floor for a transformer-poor cluster then
# gets measured on many times more cells than the stage test it is supposed to
# calibrate, and the two margins are no longer comparable. The mismatch runs
# both ways, since transformer-heavy clusters have the opposite problem.
#
# larva is UND1 + UND2 so it never binds, and is left out.
core_n = side_counts %>%
  filter(label %in% core_clusters) %>%
  transmute(label,
            n = pmin(Undifferentiated.1, Undifferentiated.2, transformer))
core_n = setNames(core_n$n, core_n$label)

# the sex comparison is on its own scale and is not read against the floor
sex_n = side_counts %>%
  filter(label %in% sex_clusters) %>%
  transmute(label, n = pmin(Female, Male))
sex_n = setNames(sex_n$n, sex_n$label)
```

`r paste0(length(core_clusters), " of ", nrow(side_counts), " clusters have at least ", min_cells_side, " cells in Undifferentiated.1, in Undifferentiated.2 and in the transformers, and carry the replicate and stage comparisons. ", length(sex_clusters), " have at least ", min_cells_side, " cells in both transformer libraries and carry the sex comparison.")`

# 2. Does each cluster find itself across the boundary?

Three comparisons. The first is the reference and the second is the question.

* **Undifferentiated.1 against Undifferentiated.2.** Two animals, same stage,
  same collection date. This is what correspondence looks like when nothing but
  replicate variation separates the sides.
* **Larvae against transformers.** The stage contrast, which is also the
  collection date and batch contrast. Those cannot be separated.
* **Female against Male.** Two animals, same collection date, differing in sex.
  Read alongside the others: a cluster that separates the sexes as strongly as
  it separates the stages is telling you the axis is not stage.

Cell numbers are matched per cluster across every side of every comparison, at
that cluster's own depth, since marker rankings sharpen with cell number and an
unmatched comparison would measure sample size. **The replicate and stage
comparisons share one depth per cluster**, so the floor is measured on the same
number of cells as the test it calibrates. Library is deliberately **not**
blocked, because library is the variable being measured. The procedure repeats
over independent draws and the similarity matrices are averaged before any
verdict is read off them.

```{r}
n_top      = 50
lfc_thresh = 1
n_draws    = 10

top_markers_for = function(marker_list, lab, n = n_top) {
  if (is.null(marker_list) || !lab %in% names(marker_list)) return(character(0))
  df = marker_list[[lab]]
  if (nrow(df) == 0) return(character(0))

  # sorted explicitly rather than trusting the order findMarkers returns.
  # Nothing would error if that order changed; the top 50 would silently
  # become an arbitrary 50 and every figure here would be wrong.
  ord = if (all(c("FDR", "p.value") %in% colnames(df))) {
    order(df$FDR, df$p.value)
  } else if ("FDR" %in% colnames(df)) {
    order(df$FDR)
  } else seq_len(nrow(df))
  rownames(df)[ord][seq_len(min(n, nrow(df)))]
}

# lfc is set because with thousands of cells a log-fold change of 0.02 clears
# FDR 0.01, and an unthresholded ranking measures sample size rather than
# effect. Both sides of every comparison use the same threshold.
markers_for_cells = function(cells, clusters_keep) {
  lab = factor(cell_key$label[cells], levels = clusters_keep)
  findMarkers(combined.sce[, cells], lab, test.type = "t",
              direction = "up", lfc = lfc_thresh, pval.type = "some")
}

# cluster i of side A against cluster j of side B, for every pair. The diagonal
# is a cluster against itself across the boundary; the off diagonal is what it
# has to beat.
similarity_matrix = function(mA, mB, clusters) {
  tA = lapply(clusters, function(cl) top_markers_for(mA, cl))
  tB = lapply(clusters, function(cl) top_markers_for(mB, cl))
  names(tA) = clusters; names(tB) = clusters

  sim = matrix(NA_real_, length(clusters), length(clusters),
               dimnames = list(clusters, clusters))
  for (i in clusters) for (j in clusters) {
    d = min(length(tA[[i]]), length(tB[[j]]))
    sim[i, j] = if (d == 0) NA_real_ else
      100 * length(intersect(tA[[i]], tB[[j]])) / d
  }
  sim
}
```

```{r}
# one comparison: subsample both sides to matched depth, compute markers on each
# side over the same cluster set, average the similarity matrices over draws
run_comparison = function(sideA_libs, sideB_libs, clusters_keep, n_by_cluster,
                          seed = 1) {

  inA = which(cell_key$library %in% sideA_libs)
  inB = which(cell_key$library %in% sideB_libs)

  # depth is supplied rather than derived, so two comparisons can be run at the
  # same depth and their margins compared
  stopifnot(all(clusters_keep %in% names(n_by_cluster)))
  n_by_cluster = n_by_cluster[clusters_keep]

  avail = sapply(clusters_keep, function(cl)
    min(sum(cell_key$label[inA] == cl), sum(cell_key$label[inB] == cl)))
  if (any(avail < n_by_cluster))
    stop("requested depth exceeds available cells for cluster(s): ",
         paste(clusters_keep[avail < n_by_cluster], collapse = ", "))

  set.seed(seed)
  acc = NULL
  for (d in seq_len(n_draws)) {
    draw_side = function(idx) {
      unlist(lapply(clusters_keep, function(cl) {
        pool = idx[cell_key$label[idx] == cl]
        # sampled by position, never by value: sample(x, n) with x of length 1
        # draws from seq_len(x) instead of from x, which would silently return
        # cell indices that have nothing to do with this cluster
        pool[sample(length(pool), n_by_cluster[[cl]])]
      }))
    }
    mA = markers_for_cells(draw_side(inA), clusters_keep)
    mB = markers_for_cells(draw_side(inB), clusters_keep)
    sim = similarity_matrix(mA, mB, clusters_keep)
    acc = if (is.null(acc)) sim else acc + sim
  }
  list(sim = acc / n_draws, n_by_cluster = n_by_cluster)
}
```

```{r}
# The depth scheme is part of the cache key, not just the tuning parameters.
# An earlier version subsampled each comparison to its own minimum, which put
# the floor and the stage test on different cell counts; results from that
# scheme are not comparable with these and must not be silently reused.
depth_tag = paste0("commonN", sum(core_n))

comparison_file = paste0(Robject_out_dir, "/stage_correspondence_", depth_tag,
                         "_", n_top, "_lfc", lfc_thresh, "_", n_draws, "draws.Rds")

if (file.exists(comparison_file)) {
  comparisons = readRDS(comparison_file)
} else {
  comparisons = list(
    replicate = run_comparison("Undifferentiated.1", "Undifferentiated.2",
                               core_clusters, core_n),
    stage     = run_comparison(c("Undifferentiated.1", "Undifferentiated.2"),
                               c("Female", "Male"), core_clusters, core_n),
    sex       = run_comparison("Female", "Male", sex_clusters, sex_n))
  saveRDS(comparisons, file = comparison_file)
}
```

## Reciprocal self-match

For each cluster, the strongest match on the other side is found in both
directions. A cluster **corresponds** when it is its own best match both ways,
which is stricter than being the best match in one direction, since a diffuse
cluster can be many clusters' best partner without any of them being its own.

The **margin** is how far the self-match sits above the strongest competitor. A
cluster that matches itself by one percentage point has not established
anything, and the margin is the number to read rather than the raw similarity.

```{r}
reciprocal_from_sim = function(sim) {
  cls = rownames(sim)
  do.call(rbind, lapply(cls, function(i) {
    row = sim[i, ]; col = sim[, i]
    off = row[names(row) != i]
    fwd = names(which.max(row)); rev = names(which.max(col))
    # a cluster with no measurable competitor has no margin to report, and
    # max() of an all-NA vector would hand back -Inf rather than saying so
    best_off = if (all(is.na(off))) NA_real_ else max(off, na.rm = TRUE)
    data.frame(
      cluster         = i,
      self_similarity = sim[i, i],
      best_forward    = if (length(fwd)) fwd else NA_character_,
      best_reverse    = if (length(rev)) rev else NA_character_,
      reciprocal      = identical(fwd, i) && identical(rev, i),
      margin          = sim[i, i] - best_off,
      stringsAsFactors = FALSE)
  }))
}

recip = bind_rows(lapply(names(comparisons), function(nm)
  reciprocal_from_sim(comparisons[[nm]]$sim) %>%
    mutate(comparison = nm,
           n_per_side = comparisons[[nm]]$n_by_cluster[cluster]))) %>%
  dplyr::select(comparison, cluster, n_per_side, self_similarity,
                best_forward, best_reverse, reciprocal, margin)

write.csv(recip, file = paste0(results_dir, "/SuppTable_reciprocal_matches.csv"),
          row.names = FALSE)

datatable(recip %>% mutate(across(where(is.numeric), ~round(., 1))),
          filter = "top", rownames = FALSE)
```

```{r}
summary_by_comparison = recip %>%
  group_by(comparison) %>%
  summarize(n_clusters      = dplyr::n(),
            n_reciprocal    = sum(reciprocal),
            pct_reciprocal  = 100 * mean(reciprocal),
            median_self     = median(self_similarity, na.rm = TRUE),
            median_margin   = median(margin, na.rm = TRUE),
            .groups = "drop")

datatable(summary_by_comparison %>% mutate(across(where(is.numeric), ~round(., 1))),
          rownames = FALSE, options = list(dom = "t", ordering = FALSE))
```

## Which stage results can be read at all

A cluster that does not correspond to itself **between two larval animals**, at
the depth its stage test runs at, has not shown that it is resolvable at that
depth by any comparison. Its stage result carries no information, and calling it
a failure would be reading noise as disagreement.

Because both comparisons now run at the same depth, the replicate result is
exactly the right control: it asks whether this cluster is recoverable at this
cell count when only replicate variation separates the sides.

```{r}
stage_tbl = recip %>% filter(comparison == "stage") %>%
  dplyr::select(cluster, n_per_side, self_stage = self_similarity,
                margin_stage = margin, reciprocal_stage = reciprocal,
                best_forward, best_reverse)

rep_tbl = recip %>% filter(comparison == "replicate") %>%
  dplyr::select(cluster, self_rep = self_similarity,
                margin_rep = margin, reciprocal_rep = reciprocal)

outcome_tbl = stage_tbl %>%
  left_join(rep_tbl, by = "cluster") %>%
  left_join(side_counts %>% dplyr::select(cluster = label, larva, transformer),
            by = "cluster") %>%
  mutate(outcome = dplyr::case_when(
    !reciprocal_rep  ~ "untestable at this depth",
    reciprocal_stage ~ "corresponds across stages",
    TRUE             ~ "fails across stages"),
    # which side's marker list broke the match. A cluster whose larval markers
    # still pick it out but whose transformer markers do not is usually short of
    # transformer cells; a cluster that fails in both directions is not.
    direction = dplyr::case_when(
      reciprocal_stage                                    ~ "both directions agree",
      best_forward == cluster & best_reverse != cluster   ~ "transformer side fails",
      best_forward != cluster & best_reverse == cluster   ~ "larval side fails",
      best_forward == best_reverse                        ~ "both point to the same other cluster",
      TRUE                                                ~ "both fail, and disagree")) %>%
  dplyr::select(cluster, larva, transformer, n_per_side,
                self_rep, margin_rep, reciprocal_rep,
                self_stage, margin_stage, reciprocal_stage,
                best_forward, best_reverse, direction, outcome) %>%
  arrange(match(outcome, c("fails across stages", "corresponds across stages",
                           "untestable at this depth")),
          as.integer(cluster))

write.csv(outcome_tbl, file = paste0(results_dir, "/SuppTable_stage_outcomes.csv"),
          row.names = FALSE)

datatable(outcome_tbl %>% mutate(across(where(is.numeric), ~round(., 1))),
          filter = "top", rownames = FALSE)
```

```{r}
n_corresponds = sum(outcome_tbl$outcome == "corresponds across stages")
n_fails       = sum(outcome_tbl$outcome == "fails across stages")
n_untestable  = sum(outcome_tbl$outcome == "untestable at this depth")
n_readable    = n_corresponds + n_fails

fails_tbl      = outcome_tbl %>% filter(outcome == "fails across stages")
untestable_tbl = outcome_tbl %>% filter(outcome == "untestable at this depth")
```

`r paste0("Of ", nrow(outcome_tbl), " clusters, ", n_readable, " are resolvable between replicate animals at their own stage-test depth and so give a readable stage result. ", n_corresponds, " correspond across the stage boundary and ", n_fails, " do not.")`

`r if (n_untestable == 0) "Every cluster is resolvable at its own depth." else paste0("**", n_untestable, " cluster(s) are untestable at this depth: ", paste(untestable_tbl$cluster, collapse = ", "), ".** These fail to find themselves even between the two larval animals, so their stage result reflects how few cells they have on their sparser side rather than anything about stage. They are reported and not counted either way. Look at their larva and transformer counts: a cluster with a few dozen cells on one side cannot support a stable marker list on that side, and reciprocity fails in whichever direction depends on it.")`

`r if (n_fails == 0) "" else paste0("The ", n_fails, " genuine failure(s), cluster(s) ", paste(fails_tbl$cluster, collapse = ", "), ", are the ones to examine individually. A cluster that is well populated on both sides, resolves cleanly between replicate animals, and then does not resolve across the stage boundary is describing something real about that population rather than a shortage of cells. Check its library composition before interpreting it: a cluster whose two sides are one transformer library against the larval libraries is separated by animal, sex, stage and batch at once, and this design cannot say which.")`

**Read the `direction` column before deciding what a failure means.** A
reciprocity failure is one of three different situations:

* **One side fails.** The larval markers still pick the cluster out but the
  transformer markers pick out something else, or the reverse. The side that
  fails is usually the sparser one, and this is the mildest case: the cluster is
  probably fine and the marker list on the thin side is unstable.
* **Both fail and disagree**, pointing at two different clusters. Nothing about
  this cluster is stable across the boundary. This is the strongest failure.
* **Both point at the same other cluster.** Rare, and the only case where the
  cluster is plausibly the same population as that other one under a different
  label.

Every non-corresponding cluster, with where each side points:

```{r}
fail_detail = outcome_tbl %>%
  filter(outcome != "corresponds across stages") %>%
  dplyr::select(cluster, outcome, larva, transformer, n_per_side,
                best_forward, best_reverse, direction,
                self_stage, margin_stage, margin_rep) %>%
  arrange(match(outcome, c("fails across stages", "untestable at this depth")),
          as.integer(cluster))

write.csv(fail_detail,
          file = paste0(results_dir, "/SuppTable_non_corresponding_clusters.csv"),
          row.names = FALSE)

datatable(fail_detail %>% mutate(across(where(is.numeric), ~round(., 1))),
          rownames = FALSE,
          options = list(dom = "t", paging = FALSE, ordering = FALSE))
```

`r if (nrow(fail_detail) == 0) "Every cluster corresponds across the stage boundary, so this table is empty." else paste0("Read ", "**best_forward**", " as the transformer cluster that the larval side of this cluster most resembles, and ", "**best_reverse**", " as the larval cluster that its transformer side most resembles. Where those two disagree there is no single partner, which is why none of these can be merged into anything.")`

```{r}
rep_row   = summary_by_comparison %>% filter(comparison == "replicate")
stage_row = summary_by_comparison %>% filter(comparison == "stage")
sex_row   = summary_by_comparison %>% filter(comparison == "sex")

fails = recip %>% filter(comparison == "stage", !reciprocal)
```

`r if (nrow(rep_row) == 0 || nrow(stage_row) == 0) "One of the two comparisons could not be run, so no reference is available." else paste0("**Reference.** Between the two larval libraries, ", rep_row$n_reciprocal, " of ", rep_row$n_clusters, " clusters are reciprocal self-matches, median similarity ", round(rep_row$median_self, 1), "% and median margin ", round(rep_row$median_margin, 1), " points. That is what correspondence looks like when only replicate animals separate the sides, and it is the ceiling for this dataset rather than 100%. **Stage.** Across the stage boundary, ", stage_row$n_reciprocal, " of ", stage_row$n_clusters, " clusters are reciprocal self-matches, median similarity ", round(stage_row$median_self, 1), "% and median margin ", round(stage_row$median_margin, 1), " points.")`

`r if (nrow(sex_row) == 0) "" else paste0("**Sex.** Between the two transformer libraries, ", sex_row$n_reciprocal, " of ", sex_row$n_clusters, " clusters are reciprocal self-matches. This comparison runs on its own cluster set, since a cluster confined to one sex cannot appear in it, so the count is not directly comparable with the other two. The median margin, ", round(sex_row$median_margin, 1), " points, is comparable.")`

## The similarity matrices

The diagonal is each cluster against itself across the boundary. Off-diagonal
brightness means a cluster's markers resemble a different cluster on the other
side, which is what a failure of correspondence looks like.

```{r, fig.width=13, fig.height=5}
sim_long = bind_rows(lapply(names(comparisons), function(nm) {
  s = comparisons[[nm]]$sim
  as.data.frame(as.table(s)) %>%
    setNames(c("side_A", "side_B", "similarity")) %>%
    mutate(comparison = nm)
})) %>%
  mutate(side_A = factor(side_A, levels = all_clusters),
         side_B = factor(side_B, levels = all_clusters))

p_sim = ggplot(sim_long, aes(x = side_B, y = side_A, fill = similarity)) +
  geom_tile() +
  scale_fill_viridis_c(name = "top 50\noverlap %") +
  facet_wrap(~ comparison, nrow = 1, scales = "free") +
  xlab("cluster, second side") + ylab("cluster, first side") +
  theme(axis.text = element_text(size = 7))

p_sim
ggsave(p_sim, file = paste0(results_dir, "/cross_boundary_similarity.png"),
       width = 13, height = 5)
```

```{r, fig.width=9, fig.height=5}
p_margin = recip %>%
  mutate(cluster = factor(cluster, levels = all_clusters)) %>%
  ggplot(aes(x = cluster, y = margin, fill = reciprocal)) +
  geom_col(position = position_dodge()) +
  facet_wrap(~ comparison, ncol = 1) +
  scale_fill_manual(values = c(`TRUE` = "#7fbf7b", `FALSE` = "#d6604d"),
                    name = "reciprocal\nself-match") +
  geom_hline(yintercept = 0, linewidth = .3) +
  ylab("margin over strongest competitor, points") + xlab("cluster")

p_margin
ggsave(p_margin, file = paste0(results_dir, "/self_match_margin.png"),
       width = 9, height = 5)
```

# 3. Did the correction over-integrate?

Section 2 asks whether the clusters correspond. This asks the opposite question
about the same clusters: whether `fastMNN` pulled larval and transformer cells
together so hard that any apparent correspondence is an artifact of the
correction rather than a property of the cells.

For each cluster, each cell's `k` nearest neighbors in the **corrected** space
are counted, and the fraction belonging to the other library group is divided by
the fraction expected under random interleaving. A value of 1 means the two
groups mix as freely as chance allows; 0 means they occupy the cluster without
ever being each other's neighbors. Undifferentiated.1 against Undifferentiated.2
is the reference again, since two animals of the same stage are what full mixing
looks like in this dataset.

This is the one place in the document that reads the corrected space, because
over-integration is a property of the correction and cannot be seen anywhere
else.

```{r}
mix_k       = c(10, 20, 50)
mix_primary = 20

mix_comparisons = list(
  floor_UND1_UND2 = list("Undifferentiated.1", "Undifferentiated.2"),
  larva_vs_transf = list(c("Undifferentiated.1", "Undifferentiated.2"),
                         c("Female", "Male")),
  TF_vs_TM        = list("Female", "Male"))

mixing = bind_rows(lapply(mix_k, function(kk)
  library_mixing_by_cluster(merged.sce,
                            clusters    = cell_key$label,
                            library_vec = cell_key$library,
                            comparisons = mix_comparisons,
                            use.dimred  = "corrected",
                            k           = kk) %>%
    mutate(k = kk))) %>%
  mutate(ratio_to_floor = larva_vs_transf / floor_UND1_UND2)

write.csv(mixing, file = paste0(results_dir, "/SuppTable_library_mixing.csv"),
          row.names = FALSE)

datatable(mixing %>% filter(k == mix_primary) %>%
            dplyr::select(-k) %>%
            mutate(across(where(is.numeric), ~round(., 3))) %>%
            arrange(ratio_to_floor),
          rownames = FALSE, options = list(dom = "t", paging = FALSE))
```

```{r}
mix_p     = mixing %>% filter(k == mix_primary)
n_mixed   = sum(mix_p$ratio_to_floor >= 0.9, na.rm = TRUE)
n_meas    = sum(!is.na(mix_p$ratio_to_floor))
med_floor = median(mix_p$floor_UND1_UND2, na.rm = TRUE)
rng       = range(mix_p$ratio_to_floor, na.rm = TRUE)

# is the sex contrast milder than the stage contrast where both are measurable?
both_meas = mix_p %>% filter(!is.na(TF_vs_TM), !is.na(larva_vs_transf))
n_sex_higher = sum(both_meas$TF_vs_TM > both_meas$larva_vs_transf)
```

`r paste0("At k = ", mix_primary, ", the replicate reference has a median of ", round(med_floor, 2), ", so two same-stage animals interleave close to freely. Larval against transformer cells range from ", round(rng[1], 2), " to ", round(rng[2], 2), " of that reference across ", n_meas, " measurable clusters, with ", n_mixed, " reaching 0.9 of the floor.")`

`r if (n_mixed == 0) "**No cluster mixes across the stage boundary at anything like the rate replicate animals do.** Larval and transformer cells share every cluster without sharing neighbourhoods. Over-integration would look like the opposite, so this is direct evidence that the correction did not erase the stage difference; it brought the two groups close enough to fall in the same communities and no closer." else "Some clusters approach the replicate floor, so the correction has brought those groups substantially together. Read their correspondence results in section 2 with that in mind, since a cluster that both mixes freely and corresponds is harder to distinguish from a forced alignment."`

`r if (nrow(both_meas) == 0) "" else paste0("Where both are measurable, the sex contrast is milder than the stage contrast in ", n_sex_higher, " of ", nrow(both_meas), " clusters. Two transformers of different sexes, collected on the same date, interleave better than same-sex animals collected months apart do. That places the dominant axis on the July-to-November contrast rather than on sex or on animal-to-animal variation, without saying what that contrast is made of.")`

```{r, fig.width=9, fig.height=5}
p_mix = mixing %>%
  mutate(cluster = factor(cluster, levels = all_clusters)) %>%
  tidyr::pivot_longer(c(floor_UND1_UND2, larva_vs_transf, TF_vs_TM),
                      names_to = "comparison", values_to = "mixing") %>%
  ggplot(aes(x = cluster, y = mixing, colour = comparison,
             shape = factor(k), group = interaction(comparison, k))) +
  geom_hline(yintercept = 1, linetype = "dashed", linewidth = .3) +
  geom_point(position = position_dodge(width = .5), size = 1.8) +
  scale_colour_manual(values = c(floor_UND1_UND2 = "#009292",
                                 larva_vs_transf = "#9e2f7f",
                                 TF_vs_TM        = "#e69f00")) +
  ylab("observed / expected cross-group neighbours") +
  labs(shape = "k",
       caption = "1 is free mixing. The replicate pair is the reference, not 1.")

p_mix
ggsave(p_mix, file = paste0(results_dir, "/library_mixing_by_cluster.png"),
       width = 9, height = 5)
```

## Mixing and correspondence are different questions

They can disagree, and the combination is what the atlas decision rests on.

**Low mixing with correspondence holding is the expected result for one cell
type sampled at two stages.** The cells sit apart because a great deal of their
transcriptome has shifted, and the cluster still corresponds because the genes
that distinguish it *from the other clusters* are the same genes on both sides.
Those are not in tension. A larval and a transformer fibroblast can differ in
hundreds of genes while what makes each one a fibroblast rather than an
endothelial cell is unchanged.

**Low mixing with correspondence failing is the case to worry about**, and those
clusters are named in section 2 rather than folded into the headline.

So the two sections answer the two halves of the reviewer's concern. Section 2
uses no corrected values at all, computing markers within each stage on
uncorrected data, so a cluster that corresponds does so on evidence that does
not pass through `fastMNN`. This section shows the correction left the stage
difference intact, so correspondence cannot be dismissed as an artifact of
having removed it.

# 4. Verdict

The verdict is taken over the **readable** clusters only, the ones that resolve
between replicate animals at their own stage-test depth. Clusters that are
untestable at their depth are excluded from both the numerator and the
denominator rather than counted as failures.

```{r}
stage_holds = n_readable > 0 && n_fails == 0
stage_mixed = n_readable > 0 && n_fails > 0 && n_fails < n_corresponds
```

`r if (n_readable == 0) "**This document cannot decide.** No cluster resolves between replicate animals at the depth its stage test runs at, so there is no readable stage result. That is a statement about cell numbers, not about stage." else if (stage_holds) paste0("**Do not split the atlas.** All ", n_readable, " readable clusters find themselves across the stage boundary by their own markers, in both directions. Those clusters are the same populations in larval and in transformer cells. The claim holds whatever the larva-to-transformer contrast is made of, which is why it survives the confound between stage, collection date and batch. Report one integrated atlas and state the confound as a limitation the design cannot resolve.") else if (stage_mixed) paste0("**Integrate, but name the exceptions.** ", n_corresponds, " of ", n_readable, " readable clusters find themselves across the stage boundary. The ", n_fails, " that do not are cluster(s) ", paste(fails_tbl$cluster, collapse = ", "), ". Report the integrated atlas and name those clusters in the text as populations whose cross-stage identity this analysis does not resolve. Do not merge them into anything: a reciprocity failure means the larval and transformer sides point at different partners, so there is no single cluster to merge with. Merging is only indicated in the rare case where both directions agree on the same other cluster, which the direction column above reports. Give their library composition alongside them, since a cluster whose two sides are one library against another is separated by animal, sex, stage and batch at once.") else paste0("**Report the stages separately.** ", n_fails, " of ", n_readable, " readable clusters fail to find themselves across the stage boundary, so the integrated clustering groups cells the marker evidence does not support grouping.")`

**Read the stage result against the replicate result, not against 100%.** Marker
rankings shuffle between random halves of identical data, so some loss of
correspondence is expected between any two groups of cells. The replicate
comparison measures that loss with nothing but animal-to-animal variation
driving it, and the stage comparison should be judged by how far it falls below
that, not by its absolute value.

## What this document does not answer

* **What the stage difference is made of.** Stage, collection date and technical
  batch are one variable in this design and no analysis of this dataset can
  separate them. This test does not attempt to; it asks whether the populations
  correspond, not why they differ.
* **Whether the cell type names are right.** Correspondence says cluster 8 is
  the same population at both stages. It says nothing about what cluster 8 is.
  That question needs the annotation, which is settled elsewhere.
* **Independent evidence for the stage-associated expression patterns.** That
  requires an external dataset or staining, not this dataset.

# Session info

```{r}
sessionInfo()
```
