---
title: "12_doublets_versus_intermediates"
author: "Stephanie Hickey"
date: "`r Sys.Date()`"
output:
  html_document:
    code_folding: hide
    toc: true
    toc_float:
      toc_collapsed: true
params:
    project_dir: /mnt/ufs18/rs-013/bioinformaticsCore/projects/chong_davidson/BCC125_sea_lamprey_scRNA
    project_name: BCC125_sea_lamprey_scRNA
    lab: Weiming Li
    dbr: 0.08
    dbr_sd: 0.02
    prop_random: 0.5
    umi_split: 1.5
    k: 30
    n_hvg: 5000
    n_pcs: 50
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

```{r, echo=FALSE, results="asis"}
cat("\n")
cat("#### Lab:", params$lab, " \n")
cat("#### Project:", params$project_name, " \n")
```

```{r}
library(tidyverse)
library(SingleCellExperiment)
library(scater)
library(scran)
library(patchwork)
library(RColorBrewer)
library(DT)
library(scDblFinder)
source(paste0(params$project_dir, "/src/scRNA_seq_functions.R"))
```

```{r}
results_dir     = paste0(params$project_dir, "/results/20260813_doublet_recalibration")
Robject_out_dir = paste0(results_dir, "/Robjects")
if (!dir.exists(Robject_out_dir)) dir.create(Robject_out_dir, recursive = TRUE)

Robject_dir  = paste0(params$project_dir, "/results/20260312/Robjects")

```

# Purpose

**Is doublet filtering removing real transitional cells?**

The worry is structural rather than a tuning problem. A doublet is two cells'
RNA in one droplet, so it reads as a mixture of two cell types. A genuine
transitional cell also reads as a mixture of two cell types. Any method scoring
"looks like a mixture" will flag both, and no amount of care with the same
method separates them.

`scDblFinder`'s own documentation makes this concrete. Describing how artificial
doublets are generated: *"A proportion of the doublets will simply use the sum of
counts of the composing cells, while the rest will undergo a library size
adjustment and poisson resampling."*

The second kind is a mixture of two profiles rescaled to one cell's worth of
RNA. That is also a description of a transitional cell. Those artificial
doublets exist so that real doublets without elevated RNA content are still
caught, and the cost is that a cell which is genuinely intermediate at normal
depth matches them equally well.

This document does two things.

**Section 1 calls doublets at the platform rate**, against the unfiltered data
as baseline.

**Sections 2 to 6 test whether the flagged cells behave like collisions or like
biology**, using four properties that separate them.

## Why the default rate is wrong here

`scDblFinder` documents its `dbr` argument as: *"The expected doublet rate. By
default this is assumed to be 1% per thousand cells captured (so 4% among 4000
thousand cells), which is appropriate for 10x datasets."*

At roughly 16,700 cells per library that default expects about **16.7%**
doublets.

The PIPseq T20 3' v4.0 user guide, for the protocol actually used here, states
that 40,000 cells added per reaction gives recovery above 20,000 cells and a
**multiplet rate below 8%**. Cell concentration was adjusted to 4 x 10^6
cells/ml with 10 ul loaded, which is 40,000 cells, so the standard protocol
applies.

`dbr` sets where the cutoff falls on the doublet score, so expecting twice as
many doublets as the chemistry produces pushes the cutoff down and flags more
cells than warranted. The rate is therefore set from the platform specification
rather than left at the default.

**Two caveats on the replacement value.** Under 8% is an upper bound rather than
a point estimate, so `r params$dbr` is conservative and `dbr.sd` sets the
tolerance around it; both are document parameters and worth varying. And the
manufacturer figure comes from well-behaved cell suspensions, while dissociated
gonad may genuinely multiplet more, so the recalibrated rate is a better
estimate rather than ground truth.

# Load

```{r}
all.sce      = readRDS(paste0(Robject_dir, "/all.sce.Rds"))
merged.sce   = readRDS(paste0(Robject_dir, "/merged.sce.Rds"))
combined.sce = readRDS(paste0(Robject_dir, "/combined.sce.Rds"))

stopifnot(identical(colnames(combined.sce), colnames(merged.sce)))

anno = c(
  "1. Fibroblasts (migrating)", "2. Mesonephric cells",
  "3. Granulosa cells (migrating)", "4. Granulosa cells",
  "5. Oogonia/Oocytes", "6. Endothelial cells",
  "7. Primordial germ cells (migrating)", "8. Erythrocytes",
  "9. Pre-granulosa cells (migrating)", "10. Leukocytes",
  "11. Hemopoietic stem cells", "12. Podocytes", "13. Macrophages",
  "14. Pre-granulosa cells (mitotic)", "15. Immune cells (migrating, mitotic)",
  "16. Spermatocytes (cytokinetic)", "17. Male germ cells (migrating)",
  "18. Spermatocytes (apoptotic)", "19. Spermatocytes (GPCR-activated)")

stopifnot(length(anno) == nlevels(colData(merged.sce)$label))

cell_key = data.frame(
  barcode = colnames(combined.sce),
  batch   = as.character(colData(merged.sce)$batch),
  label   = as.character(colData(merged.sce)$label),
  stringsAsFactors = FALSE) %>%
  mutate(CellType = anno[as.integer(label)])

cells_per_library = cell_key %>% dplyr::count(batch, name = "n_cells")
datatable(cells_per_library, rownames = FALSE)
```

`r paste0("Default dbr would be ", paste(round(cells_per_library$n_cells / 1000, 1), collapse = ", "), "% for these libraries, against a platform specification below 8%.")`

# 1. Call doublets at the platform rate

`scDblFinder` is run once per library at the platform rate, with `clusters`
supplied so that each call carries a putative origin, which sections 3 and 4
need.

## What is supplied as `clusters`

A per-library clustering computed here: log-normalise, model variance, top
`r params$n_hvg` highly variable genes, PCA to `r params$n_pcs` components,
shared nearest neighbour graph at k = `r params$k`, Louvain. Groups under 10 cells
are pooled so their cells are still scored.

These settings match `01`, so cluster granularity is comparable to the rest of
the project. All three are document parameters.

A clustering derived from data that still contains doublets can give a doublet
population its own cluster, which the simulation would then treat as a source.
`propRandom` sets the share of artificial doublets built from random cell pairs
rather than cluster combinations, and is raised above its default here so that a
proportion of them never depend on the clustering being right.

```{r}
dbl_file = paste0(Robject_out_dir, "/scdbl_recalibrated.Rds")

if (file.exists(dbl_file)) {
  dbl = readRDS(dbl_file)
} else {
  dbl = lapply(names(all.sce), function(s) {
    x = all.sce[[s]]

    # populations for the simulation, computed per library
    set.seed(1)
    xc  = logNormCounts(x)
    dec = modelGeneVar(xc)
    hvg = getTopHVGs(dec, n = params$n_hvg)
    xc  = runPCA(xc, subset_row = hvg, ncomponents = params$n_pcs)
    g   = buildSNNGraph(xc, use.dimred = "PCA", k = params$k)
    loc = as.character(igraph::cluster_louvain(g)$membership)

    # a group too small to seed simulation is pooled rather than dropped
    tab = table(loc)
    loc[loc %in% names(tab)[tab < 10]] = "small"

    set.seed(100)
    d_platform = scDblFinder(x, dbr = params$dbr, dbr.sd = params$dbr_sd,
                             clusters = loc, propRandom = params$prop_random)

    data.frame(
      sample          = s,
      barcode         = colnames(x),
      local_cluster   = paste(s, loc, sep = ":"),
      sum             = colData(x)$sum,
      detected        = colData(x)$detected,
      class_platform  = as.character(d_platform$scDblFinder.class),
      score_platform  = d_platform$scDblFinder.score,
      origin          = as.character(d_platform$scDblFinder.mostLikelyOrigin),
      origin_ambig    = d_platform$scDblFinder.originAmbiguous,
      stringsAsFactors = FALSE)
  })
  dbl = do.call(rbind, dbl)
  saveRDS(dbl, file = dbl_file)
}

# the annotated cluster is attached afterwards and used only for reporting
# which populations the flagged cells belong to. It plays no part in the
# doublet calls or in the parent identification.
dbl = dbl %>%
  left_join(cell_key %>% dplyr::select(barcode, label, CellType), by = "barcode") %>%
  mutate(flagged = class_platform == "doublet") %>%
  group_by(local_cluster) %>%
  mutate(umi_ratio_local = sum / median(sum[!flagged])) %>%
  ungroup()

write.csv(dbl, file = paste0(results_dir, "/doublet_calls_both_rates.csv"),
          row.names = FALSE)
```

```{r}
rate_tbl = dbl %>%
  group_by(sample) %>%
  summarize(n_cells   = dplyr::n(),
            n_flagged = sum(flagged),
            pct_flagged = 100 * mean(flagged),
            .groups = "drop")

rate_tbl = bind_rows(rate_tbl,
  dbl %>% summarize(sample = "all", n_cells = dplyr::n(),
                    n_flagged = sum(flagged),
                    pct_flagged = 100 * mean(flagged)))

write.csv(rate_tbl, file = paste0(results_dir, "/doublet_rate_by_calibration.csv"),
          row.names = FALSE)

datatable(rate_tbl %>% mutate(across(where(is.numeric), ~round(., 2))),
          rownames = FALSE, options = list(dom = "t", ordering = FALSE))

n_flagged = sum(dbl$flagged)
```

`r paste0("Filtering at the platform rate would remove ", n_flagged, " cells, ", round(100 * n_flagged / nrow(dbl), 2), "% of the dataset. Those cells are the subject of the rest of this document.")`

**Loading was constant across libraries**, 4 x 10^6 cells/ml and 10 ul into
every reaction, so the expected collision rate is the same in each. Per-library
rates that differ substantially are therefore not explained by loading, and the
excess in a high library is worth attention.

```{r, fig.width=8, fig.height=4}
p_rate = rate_tbl %>%
  filter(sample != "all") %>%
  ggplot(aes(x = sample, y = pct_flagged)) +
  geom_col(fill = "#9ecae1") +
  geom_hline(yintercept = 8, linetype = "dashed", colour = "grey30") +
  theme(axis.text.x = element_text(angle = 20, hjust = 1)) +
  xlab(NULL) + ylab("percent of cells called doublet") +
  labs(caption = "Dashed line is the PIPseq specified multiplet rate ceiling.")

p_rate
ggsave(p_rate, file = paste0(results_dir, "/doublet_rate_by_calibration.png"),
       width = 8, height = 4)
```

# 2. Where do the flagged cells sit?

If the flagged cells are scattered across clusters, filtering thins every
population slightly and removes none of them. If they concentrate in a few
clusters, those clusters are what the concern is about, and the tests below
apply to them rather than to everything.

```{r}
concentration = dbl %>%
  group_by(label, CellType) %>%
  summarize(n_cells          = dplyr::n(),
            n_flagged        = sum(flagged),
            pct_of_cluster   = 100 * mean(flagged),
            .groups = "drop") %>%
  mutate(pct_of_all_flagged = 100 * n_flagged / sum(n_flagged)) %>%
  arrange(desc(pct_of_cluster))

write.csv(concentration,
          file = paste0(results_dir, "/flagged_cells_by_cluster.csv"),
          row.names = FALSE)

datatable(concentration %>% mutate(across(where(is.numeric), ~round(., 2))),
          filter = "top", rownames = FALSE)

# The tests below need the per-library clusters, since that is what
# mostLikelyOrigin refers to. Each is labelled with the annotated cell type its
# cells mostly belong to, so the output is readable.
local_focus = dbl %>%
  group_by(sample, local_cluster) %>%
  summarize(n_cells   = dplyr::n(),
            n_flagged = sum(flagged),
            pct_flagged = 100 * mean(flagged),
            CellType  = names(sort(table(CellType), decreasing = TRUE))[1],
            .groups = "drop") %>%
  arrange(desc(pct_flagged))

datatable(local_focus %>% mutate(across(where(is.numeric), ~round(., 2))),
          filter = "top", rownames = FALSE)

# local clusters losing a material share of their cells
focus = local_focus %>% filter(pct_flagged >= 5, n_flagged >= 20)
```

`r if (nrow(focus) == 0) "No cluster loses 5% or more of its cells to recalibration. The change is spread thinly, which means the default rate was not selectively removing any one population, and the concern does not have a specific target in this dataset." else paste0(nrow(focus), " per-library cluster(s) have at least 5% of their cells flagged, mostly annotated as: ", paste(unique(focus$CellType), collapse = "; "), ". Tests 3 and 4 run on these. Test 1 groups the flagged cells differently, by putative origin rather than by cluster, for the reason given there.")`

```{r}
# scDblFinder documents mostLikelyOrigin only as "the presumed cluster
# combination", without specifying the separator. Rather than assume one, split
# on any non-alphanumeric run and keep the pieces that are known cluster
# labels. Longest labels are matched first so "1" cannot be taken from "10".
known_labels = sort(unique(sub("^[^:]*:", "", dbl$local_cluster)))
# these are curated cell type names, not cluster numbers

parse_origin = function(x) {
  if (is.na(x)) return(character(0))
  parts = strsplit(x, "[^[:alnum:]]+")[[1]]
  parts = parts[nzchar(parts)]
  parts[parts %in% known_labels]
}

# origins refer to the per-library clusters computed above, so they are mapped
# back to annotated clusters for interpretation
local_to_anno = dbl %>%
  dplyr::count(local_cluster, CellType, name = "n") %>%
  group_by(local_cluster) %>%
  slice_max(n, n = 1, with_ties = FALSE) %>%
  ungroup() %>%
  dplyr::rename(dominant_CellType = CellType)

datatable(local_to_anno, filter = "top", rownames = FALSE)

origin_examples = dbl %>%
  filter(!is.na(origin)) %>%
  dplyr::count(origin, name = "n") %>%
  arrange(desc(n)) %>%
  head(10) %>%
  rowwise() %>%
  mutate(parsed = paste(parse_origin(origin), collapse = " + "),
         n_parsed = length(parse_origin(origin))) %>%
  ungroup()

datatable(origin_examples, rownames = FALSE,
          options = list(dom = "t", ordering = FALSE))
```

**Check this table before trusting sections 3 and 4.** `parsed` should contain
two cluster labels for every row. If `n_parsed` is not 2, the separator is
something the split does not handle and `parse_origin` needs adjusting; the two
tests below depend on it.

# 3. Test 1: can these cells express what neither parent does?

The decisive test, and the only one that can falsify the doublet explanation
outright.

A doublet is two real cells' RNA added together. It cannot express a gene above
the level of **both** contributing cell types, because it contains nothing that
was not in one of them. A transitional cell can, and often does, since
transitions involve genes transiently high at the intermediate state that are
low at both endpoints.

**The group being tested is an origin group, not a cluster.** Flagged cells are
grouped by their `mostLikelyOrigin`, so every cell in a group is attributed to
the same parent pair and is tested against its own parents. Grouping by cluster
would force a single modal origin onto cells whose origins differ, and test some
of them against the wrong reference. Cells whose origin is flagged ambiguous by
`scDblFinder` are excluded rather than assigned a parentage they do not have.

A doublet is a sum of two cells, so after normalisation its expression for any
gene is a weighted average of the two parents, weighted by their library sizes.
A weighted average cannot exceed the larger of the two values being averaged. So
a gene significantly above **both** parents cannot come from mixing them, at any
mixing proportion.

That is why this test is independent of the score. `scDblFinder` measures
closeness to simulated mixtures; this measures violation of an upper bound, and
a cell can sit on top of the simulated mixtures while still breaking it.

`pval.type = "all"` is what enforces "both". It takes the maximum across the
pairwise contrasts, so a gene counts only if it is up against parentA and
parentB. The default, `"any"`, would return genes above either parent, which
every mixture has.

**Each origin group is split on RNA content and tested twice.** A flagged cell
carrying more than `r params$umi_split` times the unflagged median of its own
cluster was called on a criterion independent of profile and is plausibly a
collision. A flagged cell at or below that was called on profile alone, which is
what both a library-size adjusted artificial doublet and a transitional cell
look like.

**The `low` rows are the ones the question is about.** Genes above both parents
there are cells with one cell's worth of RNA expressing something no mixture of
those parents could produce. The `high` rows are reported alongside as a
comparison: if they show the same thing, the split is not separating anything
and the result is about the origin groups rather than about RNA content.

```{r}
# a lookup from local cluster to the annotated cell type its cells mostly are
local_type = dbl %>%
  group_by(local_cluster) %>%
  summarize(CellType = names(sort(table(CellType), decreasing = TRUE))[1],
            .groups = "drop") %>%
  tibble::deframe()

# Flagged cells are grouped by their putative origin rather than by the cluster
# they sit in. Every cell in an origin group is attributed to the same parent
# pair, so all of them are tested against their own parents. Grouping by cluster
# instead would force one modal origin onto cells whose origins differ.
origin_groups = dbl %>%
  filter(flagged, !is.na(origin), !origin_ambig) %>%
  dplyr::count(sample, origin, name = "n_flagged") %>%
  rowwise() %>%
  mutate(parts = list(parse_origin(origin))) %>%
  ungroup() %>%
  filter(lengths(parts) == 2) %>%
  mutate(pa_lc = paste(sample, vapply(parts, `[`, character(1), 1), sep = ":"),
         pb_lc = paste(sample, vapply(parts, `[`, character(1), 2), sep = ":"),
         parentA = unname(local_type[pa_lc]),
         parentB = unname(local_type[pb_lc])) %>%
  dplyr::select(-parts) %>%
  arrange(desc(n_flagged))

write.csv(origin_groups,
          file = paste0(results_dir, "/origin_groups.csv"), row.names = FALSE)

datatable(origin_groups %>% dplyr::select(sample, origin, parentA, parentB,
                                          n_flagged),
          filter = "top", rownames = FALSE)

# origin groups large enough to test
min_origin_cells = 20
focus_origin = origin_groups %>% filter(n_flagged >= min_origin_cells)
```

`r if (nrow(focus_origin) == 0) paste0("No origin group has at least ", min_origin_cells, " flagged cells with an unambiguous origin, so this test cannot run. Ambiguous origins are excluded, so a large ambiguous fraction is one reason this can be empty.") else paste0(nrow(focus_origin), " origin group(s) have at least ", min_origin_cells, " flagged cells. The commonest pairs their cells are attributed to: ", paste(head(paste0(focus_origin$parentA, " + ", focus_origin$parentB), 5), collapse = "; "), ".")`

```{r}
novel_expression = function(i, subset = c("low", "high"),
                            min_cells = 20, fdr = 0.01) {
  subset = match.arg(subset)
  row    = focus_origin[i, ]

  cells = dbl %>% filter(flagged, sample == row$sample, origin == row$origin,
                         !origin_ambig)

  # Split on RNA content relative to the unflagged median of the cell's own
  # cluster. A flagged cell at 2x was called on a criterion independent of
  # profile and is plausibly a collision. A flagged cell at 1x was called on
  # profile alone, which is what both a library-size adjusted artificial
  # doublet and a transitional cell look like. The low subset is the one the
  # question is about; pooling the two dilutes it.
  cells = if (subset == "low") {
    cells %>% filter(umi_ratio_local <= params$umi_split)
  } else {
    cells %>% filter(umi_ratio_local > params$umi_split)
  }
  if (nrow(cells) < min_cells) return(NULL)

  test_bc = cells$barcode
  pa_bc   = setdiff(dbl$barcode[dbl$local_cluster == row$pa_lc], test_bc)
  pb_bc   = setdiff(dbl$barcode[dbl$local_cluster == row$pb_lc], test_bc)
  if (length(pa_bc) < min_cells || length(pb_bc) < min_cells) return(NULL)

  keep = c(test_bc, pa_bc, pb_bc)
  grp  = factor(c(rep("flagged", length(test_bc)),
                  rep("parentA", length(pa_bc)),
                  rep("parentB", length(pb_bc))),
                levels = c("flagged", "parentA", "parentB"))

  sub = combined.sce[, match(keep, colnames(combined.sce))]

  # pval.type = "all" takes the maximum across the pairwise contrasts, so a
  # gene counts only if it is up against parentA AND parentB. The default,
  # "any", would return genes above either parent, which every mixture has.
  fm = findMarkers(sub, grp, test.type = "t", direction = "up",
                   pval.type = "all")

  res = as.data.frame(fm[["flagged"]])
  res$Gene = rownames(res)

  data.frame(
    sample        = row$sample,
    subset        = subset,
    origin        = row$origin,
    parentA       = row$parentA,
    parentB       = row$parentB,
    n_tested      = length(test_bc),
    n_parentA     = length(pa_bc),
    n_parentB     = length(pb_bc),
    n_above_both  = sum(res$FDR < fdr, na.rm = TRUE),
    top_genes     = paste(head(res$Gene[order(res$FDR)], 10), collapse = ", "),
    stringsAsFactors = FALSE)
}

novel_file = paste0(Robject_out_dir, "/novel_expression.Rds")

if (file.exists(novel_file)) {
  novel = readRDS(novel_file)
} else {
  novel = if (nrow(focus_origin) == 0) NULL else
    bind_rows(
      purrr::map_dfr(seq_len(nrow(focus_origin)), novel_expression,
                     subset = "low"),
      purrr::map_dfr(seq_len(nrow(focus_origin)), novel_expression,
                     subset = "high"))
  saveRDS(novel, file = novel_file)
}

if (!is.null(novel)) {
  write.csv(novel, file = paste0(results_dir, "/novel_expression_test.csv"),
            row.names = FALSE)
  datatable(novel, rownames = FALSE)
}
```

`r if (is.null(novel) || nrow(novel) == 0) "No cluster had enough flagged cells with an unambiguous origin to run this test." else paste0("Genes expressed above **both** putative parents, at FDR < 0.01: ", paste(paste0(novel$parentA, " + ", novel$parentB, ": ", novel$n_above_both), collapse = "; "), ". A count above zero means those cells contain expression that no mixture of the two parent types could produce, which is evidence they are not doublets. A count of zero is consistent with a mixture, though it does not prove one, since a transitional state can also lie between its endpoints without exceeding either.")`

# 4. Test 2: are they commoner than collision allows?

Doublets form by random co-encapsulation, so the rate for a pair of types is
proportional to their abundances. A biological population has no reason to obey
that.

For clusters i and j, the expected share of doublets is proportional to
`2 * p_i * p_j`, where `p` is the proportion of cells in that cluster. The
observed share comes from `mostLikelyOrigin`.

```{r}
# Abundances are per library over the per-library clusters, because that is
# what origins refer to and because collisions happen within a reaction.
p_cluster = dbl %>%
  dplyr::count(sample, local_cluster, name = "n") %>%
  group_by(sample) %>%
  mutate(p = n / sum(n)) %>%
  ungroup() %>%
  mutate(cl = sub("^[^:]*:", "", local_cluster))

expected_pairs = p_cluster %>%
  dplyr::select(sample, i = cl, p_i = p) %>%
  inner_join(p_cluster %>% dplyr::select(sample, j = cl, p_j = p),
             by = "sample", relationship = "many-to-many") %>%
  filter(i < j) %>%
  mutate(w = 2 * p_i * p_j) %>%
  group_by(sample) %>%
  mutate(expected_share = w / sum(w)) %>%
  ungroup()

origin_counts = dbl %>%
  filter(flagged, !is.na(origin)) %>%
  dplyr::count(sample, origin, name = "observed")

collision = origin_counts %>%
  rowwise() %>%
  mutate(parts = list(sort(parse_origin(origin)))) %>%
  ungroup() %>%
  filter(lengths(parts) == 2) %>%
  mutate(i = vapply(parts, function(x) x[1], character(1)),
         j = vapply(parts, function(x) x[2], character(1))) %>%
  dplyr::select(-parts) %>%
  left_join(expected_pairs %>% dplyr::select(sample, i, j, expected_share),
            by = c("sample", "i", "j")) %>%
  filter(!is.na(expected_share)) %>%
  group_by(sample) %>%
  mutate(observed_share = observed / sum(observed)) %>%
  ungroup() %>%
  mutate(enrichment = observed_share / expected_share,
         typeA = unname(local_type[paste(sample, i, sep = ":")]),
         typeB = unname(local_type[paste(sample, j, sep = ":")])) %>%
  dplyr::select(sample, i, j, typeA, typeB, observed, observed_share,
                expected_share, enrichment) %>%
  arrange(desc(enrichment))

write.csv(collision, file = paste0(results_dir, "/collision_enrichment.csv"),
          row.names = FALSE)

datatable(collision %>% mutate(across(where(is.numeric), ~signif(., 3))),
          filter = "top", rownames = FALSE)
```

An enrichment near 1 means the pair occurs about as often as random collision
predicts. Values far above 1 mean the pair is commoner than collision can
account for, which is evidence for a real population rather than a coincidence
of loading.

`scDblFinder` also ships `doubletPairwiseEnrichment()` and
`clusterStickiness()`, which test this with their own null. They are worth
running alongside the arithmetic above; the calculation here is kept explicit so
the expectation is visible rather than delegated.

# 5. Test 3: do they carry two cells' worth of RNA?

**This test is only informative in one direction, and the documentation says
why.** Part of the artificial doublet set is raw sums of two cells, carrying
roughly twice the library size, so library size is part of what the score
responds to. The rest are library-size adjusted and Poisson resampled, so they
look like mixtures at one cell's worth of RNA.

That asymmetry sets what the ratios below can support. A ratio near 2 is
consistent with collisions. **A ratio near 1 is not evidence the cells are
real**, because it is equally what a doublet flagged through the adjusted
simulations looks like, and what a transitional cell looks like. The test can
corroborate the doublet explanation and cannot refute it.

How much of the score is library size in this dataset is still worth measuring,
since it sets how much of the flagged set could have been caught on RNA content
alone.

```{r, fig.width=8, fig.height=4}
lib_vs_score = dbl %>%
  group_by(sample) %>%
  summarize(rho = suppressWarnings(cor(sum, score_platform, method = "spearman")),
            .groups = "drop")

datatable(lib_vs_score %>% mutate(rho = round(rho, 3)),
          rownames = FALSE, options = list(dom = "t", ordering = FALSE))

p_libscore = ggplot(dbl, aes(x = sum, y = score_platform)) +
  geom_point(size = .2, alpha = .2) +
  scale_x_log10() +
  facet_wrap(~ sample, nrow = 1) +
  xlab("total counts") + ylab("doublet score")

p_libscore
ggsave(p_libscore, file = paste0(results_dir, "/score_vs_library_size.png"),
       width = 8, height = 4)
```

`r paste0("Spearman correlation between total counts and doublet score: ", paste(paste0(lib_vs_score$sample, " ", round(lib_vs_score$rho, 2)), collapse = "; "), ". A strong correlation means much of the flagged set could have been picked out on RNA content alone. A weak one means most of it was flagged on profile, which is the case where the distinction from a transitional cell is hardest and test 1 carries the weight.")`

Oocytes have the highest median UMI in this dataset and a below-average doublet
rate, which points toward the score not tracking RNA content.

## Is 2x attainable at this depth?

The expectation that a doublet carries twice the UMIs assumes recovery is
roughly linear in input molecules. Shallow sequencing makes the ratio noisy.
Saturation compresses it, because once the same molecules are being resequenced
a cell with twice the mRNA no longer returns twice the UMIs.

Both are checkable here. If the UMI distribution has no mass near twice the
median, the test cannot detect a doublet signature however many doublets are
present, and a low ratio below would carry no information.

```{r, fig.width=9, fig.height=4}
depth_headroom = dbl %>%
  group_by(sample) %>%
  summarize(median_umi        = median(sum),
            med_unflagged     = median(sum[!flagged]),
            med_flagged       = median(sum[flagged]),
            flagged_vs_unflag = median(sum[flagged]) / median(sum[!flagged]),
            p99_ratio         = quantile(umi_ratio_local, .99),
            pct_over_2x       = 100 * mean(umi_ratio_local > 2),
            pct_flagged_over_2x = 100 * mean(umi_ratio_local[flagged] > 2),
            .groups = "drop")

write.csv(depth_headroom, file = paste0(results_dir, "/depth_headroom.csv"),
          row.names = FALSE)

datatable(depth_headroom %>% mutate(across(where(is.numeric), ~round(., 2))),
          rownames = FALSE, options = list(dom = "t", ordering = FALSE))

# The reference is the unflagged median of each cell's own local cluster, not
# the library median. These distributions are right-skewed, so the library
# median sits well above the mode and comparing against twice it understates
# the flagged cells.
p_depth = ggplot(dbl, aes(x = umi_ratio_local, fill = flagged)) +
  geom_histogram(bins = 60, position = "identity", alpha = .55) +
  geom_vline(xintercept = 2, linetype = "dashed", colour = "grey30") +
  geom_vline(xintercept = params$umi_split, linetype = "dotted",
             colour = "grey30") +
  scale_x_log10() +
  scale_fill_brewer(palette = "Set1", direction = -1) +
  facet_wrap(~ sample, nrow = 1) +
  xlab("total counts, relative to the unflagged median of the same cluster") +
  ylab("cells") +
  labs(fill = "flagged",
       caption = paste0("Dashed line is 2x, what a collision should reach. ",
                        "Dotted line at ", params$umi_split,
                        "x is where test 1 splits the flagged cells."))

p_depth
ggsave(p_depth, file = paste0(results_dir, "/depth_headroom.png"),
       width = 9, height = 4)
```

`r paste0("Cells above twice their library median: ", paste(paste0(depth_headroom$sample, " ", round(depth_headroom$pct_over_2x, 1), "%"), collapse = "; "), ".")`

**How to read this before reading the ratios.** If a few percent of cells sit at
or beyond twice the median, the distribution has the range needed and the ratios
below mean what they say. If almost nothing does, the library is compressed,
whether by saturation or by a narrow range of cell sizes, and a ratio near 1 is
then uninformative rather than evidence against collisions.

Sequencing saturation from the PIPseq run would settle the mechanism directly if
those per-library metrics are to hand. The distribution above answers the
practical question without them.

## The ratios


A true doublet holds roughly twice the transcripts of a singlet of the same
type. A transitional cell holds one cell's worth. Comparison is within cluster,
because cell types differ in RNA content and a global comparison would only
recover that difference.

```{r}
umi_ratio = dbl %>%
  group_by(local_cluster, CellType) %>%
  summarize(
    n_singlet      = sum(!flagged),
    n_flagged      = sum(flagged),
    med_singlet    = median(sum[!flagged]),
    med_flagged    = if (sum(flagged) >= 10) median(sum[flagged]) else NA_real_,
    .groups = "drop") %>%
  mutate(ratio_flagged = med_flagged / med_singlet) %>%
  arrange(desc(n_flagged))

write.csv(umi_ratio, file = paste0(results_dir, "/umi_ratio_by_cluster.csv"),
          row.names = FALSE)

datatable(umi_ratio %>% mutate(across(where(is.numeric), ~round(., 2))),
          filter = "top", rownames = FALSE)
```

`ratio_doublet` near 2 is what collisions look like. **`ratio_flagged` is the
column that matters**: it is the flagged cells against the singlets of
their own cluster. Near 1 means they carry one cell's worth of RNA and were
being removed for looking like mixtures rather than for being them. Near 2 means
the default rate was catching genuine collisions and recalibration is losing
them.

# 6. Test 4: coherence, and where reproducibility does not apply

Doublets arise from many independent random pairings, so they should not form a
tight group with markers of their own. A real transitional population should.

```{r}
coherence = function(lc, min_cells = 20, fdr = 0.01) {
  # compared within the per-library cluster, so the flagged cells are set
  # against the cells that would be retained from the same population
  bc_r = dbl$barcode[dbl$flagged  & dbl$local_cluster == lc]
  bc_s = dbl$barcode[!dbl$flagged & dbl$local_cluster == lc]
  if (length(bc_r) < min_cells || length(bc_s) < min_cells) return(NULL)

  keep = c(bc_r, bc_s)
  grp  = factor(c(rep("flagged", length(bc_r)),
                  rep("singlet", length(bc_s))))
  sub  = combined.sce[, match(keep, colnames(combined.sce))]

  fm  = findMarkers(sub, grp, test.type = "t", direction = "up",
                    pval.type = "any")
  res = as.data.frame(fm[["flagged"]])

  data.frame(local_cluster = lc,
             CellType      = unname(local_type[lc]),
             n_flagged     = length(bc_r),
             n_own_markers = sum(res$FDR < fdr, na.rm = TRUE),
             stringsAsFactors = FALSE)
}

coh = if (nrow(focus) == 0) NULL else
  purrr::map_dfr(focus$local_cluster, coherence)

if (!is.null(coh)) {
  write.csv(coh, file = paste0(results_dir, "/coherence_test.csv"),
            row.names = FALSE)
  datatable(coh, rownames = FALSE)
}
```

`r if (is.null(coh) || nrow(coh) == 0) "No cluster had enough flagged cells to test coherence." else "A high count of markers distinguishing the flagged cells from the singlets of their own cluster means they are a distinct group rather than a random scatter of mixtures. A low count means they look like their neighbours plus noise."`

## Reproducibility, and why it applies to only one comparison

Cross-library reproducibility sounds like the obvious check and mostly is not
available here.

**Undifferentiated.1 against Undifferentiated.2 is the only replicate pair.**
Same stage, same collection, same sex-labile status, differing by animal and
library preparation. A real population present in one should appear in the
other.

**Everywhere else it does not apply.** A transitional state occurring during
metamorphosis exists only in TF and TM. A female-specific one exists only in TF.
Absence from the larval libraries is then biology, not evidence of doublethood,
and for a population confined to a single library there is no replicate at all.
For those, tests 1 and 2 carry the weight.

Even between the larval libraries the test is weak, since collisions occur in
both. Presence in both discriminates nothing; only the rate relative to
collision does, which is test 2.

```{r}
larval = c("UND1_C00004336", "UND2_C00004362")

repro = dbl %>%
  filter(sample %in% larval) %>%
  group_by(label, CellType, sample) %>%
  summarize(pct_flagged = 100 * mean(flagged), .groups = "drop") %>%
  tidyr::pivot_wider(id_cols = c(label, CellType), names_from = sample,
                     values_from = pct_flagged) %>%
  filter(!is.na(.data[[larval[1]]]), !is.na(.data[[larval[2]]]))

datatable(repro %>% mutate(across(where(is.numeric), ~round(., 2))),
          filter = "top", rownames = FALSE)
```

# 7. Summary

```{r}
summary_tbl = data.frame(
  question = c(
    "cells in the unfiltered dataset",
    "cells flagged at the PIPseq rate",
    "clusters losing 5% or more of their cells to filtering"),
  value = c(
    as.character(nrow(dbl)),
    paste0(n_flagged, " (", round(100 * n_flagged / nrow(dbl), 2), "%)"),
    as.character(nrow(focus))))

datatable(summary_tbl, rownames = FALSE,
          options = list(dom = "t", ordering = FALSE))
```

**How to read the four tests together.** Test 1 is the only one that can falsify
the doublet explanation, and the reason is worth stating because it is easy to
build a test that cannot. `scDblFinder` scores a cell by its similarity to
artificial doublets built from pairs of real cells, so any test asking "does
this look like a mixture of two cells" restates the criterion that flagged it
and will always agree. Test 1 asks something different: a sum of two cells
cannot exceed the higher of its two contributors in any gene. That is a
constraint rather than a similarity, and a cell can sit on top of the simulated
mixtures while still breaking it.

Test 2 is strong and independent, since collision rates come from loading rather
than from expression. Test 3 can corroborate the doublet
explanation but cannot refute it, because part of the artificial doublet set is
library-size adjusted, so a normal-depth mixture is something the score is built
to flag. A heterotypic doublet of a small cell and a large one need not reach
twice the median either, and the 2x expectation only holds if the depth allows
it, which the headroom check establishes. Test 4 is supporting.

**What none of them can do.** None proves a cell is transitional. They establish
whether the doublet explanation survives, and it is possible for it to survive
while the cells are still real. A negative result across all four means the
cells behave like collisions in every respect measurable here, which is the
strongest available reason to keep filtering them.

# Session info

```{r}
sessionInfo()
```
