---
title: "13_atlas_and_stage_correspondence"
author: "Stephanie Hickey"
date: "`r Sys.Date()`"
output:
  html_document:
    code_folding: hide
    toc: true
    toc_float:
      toc_collapsed: true
params:
    project_dir: /mnt/ufs18/rs-013/bioinformaticsCore/projects/chong_davidson/BCC125_sea_lamprey_scRNA
    project_name: BCC125_sea_lamprey_scRNA
    lab: Weiming Li
    results_date: "20260813"
    dbr: 0.08
    dbr_sd: 0.02
    k: 30
    lfc_thresh: 1
    genome_dir: /mnt/research/bioinformaticsCore/shared/Genomes/Petromyzon_marinus
    marker_file: canonical_markers_S1_only_2026-08-14.csv
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

```{r, echo=FALSE, results="asis"}
cat("\n")
cat("#### Lab:", params$lab, " \n")
cat("#### Project:", params$project_name, " \n")
```

```{r}
library(tidyverse)
library(SingleCellExperiment)
library(scater)
library(scran)
library(patchwork)
library(RColorBrewer)
library(DT)
library(batchelor)
library(BiocSingular)
library(scDblFinder)
source(paste0(params$project_dir, "/src/scRNA_seq_functions.R"))
```

This document builds the doublet-filtered atlas and then asks whether its
clusters name the same populations in larval and in transformer cells. 

```{r}
# The large objects are shared with 01, so this document reuses whatever that
# one already built rather than recomputing it.
object_dir  = paste0(params$project_dir, "/results/", params$results_date,
                     "_no_doublets")
results_dir = paste0(params$project_dir, "/results/", params$results_date,
                     "_atlas")
Robject_out_dir = paste0(results_dir, "/Robjects")
data_dir    = paste0(params$project_dir, "/data")

for (d in c(object_dir, results_dir, Robject_out_dir))
  if (!dir.exists(d)) dir.create(d, recursive = TRUE)

all_sce_file   = paste0(object_dir, "/all.sce.Rds")
mnn_file       = paste0(object_dir, "/mnn.sce.Rds")
merged_file    = paste0(object_dir, "/merged.sce.Rds")
combined_file  = paste0(object_dir, "/combined.sce.Rds")
dbl_stats_file = paste0(object_dir, "/doublet_removal_stats.csv")
rate_stamp     = paste0(object_dir, "/doublet_rate_used.txt")
k_stamp        = paste0(object_dir, "/clustering_k_used.txt")

build_all_sce  = !file.exists(all_sce_file)
build_mnn      = !file.exists(mnn_file)
build_combined = !file.exists(combined_file)

# the doublet rate is a parameter, so a directory built at one rate must not be
# silently reused at another
if (build_all_sce) {
  writeLines(paste0("dbr=", params$dbr, " dbr_sd=", params$dbr_sd), rate_stamp)
} else if (file.exists(rate_stamp)) {
  prev = readLines(rate_stamp)[1]
  now  = paste0("dbr=", params$dbr, " dbr_sd=", params$dbr_sd)
  if (!identical(prev, now))
    stop("Objects in ", object_dir, " were built at ", prev,
         " but this knit specifies ", now,
         ". Point results_date at a new directory or delete the cached objects.")
}

# and the same for k, since the clustering is cached alongside the merge
if (!file.exists(merged_file)) {
  build_merged = TRUE
} else if (file.exists(k_stamp)) {
  prev_k = as.integer(readLines(k_stamp)[1])
  build_merged = !identical(prev_k, as.integer(params$k))
  if (build_merged)
    message("Cached clustering used k = ", prev_k, "; reclustering at k = ",
            params$k, ".")
} else {
  # an object written before this document existed, so the k it used is not
  # recorded. Reclustered rather than assumed, which is cheap next to the merge.
  warning("No k stamp in ", object_dir,
          "; the cached clustering's k is unknown. Reclustering at k = ",
          params$k, ".")
  build_merged = TRUE
}
```

# Doublet removal

Doublets are removed per sample, before any normalization, HVG selection or PCA,
and before the samples are merged.

`scDblFinder` documents `dbr` as *"The expected doublet rate. By default this is
assumed to be 1% per thousand cells captured (so 4% among 4000 thousand cells),
which is appropriate for 10x datasets."* At roughly 16,700 cells per library
that default expects about 16.7% doublets.

These libraries are PIPseq T20 3' v4.0. Cell concentration was adjusted to
4 x 10^6 cells/ml and 10 ul loaded, which is 40,000 cells, the standard
protocol. The kit user guide states that this gives recovery above 20,000 cells
and a multiplet rate below 8%.

`dbr` places the cutoff on the doublet score, so expecting roughly twice the
rate the chemistry produces pushes the cutoff down and flags more cells than
warranted. The rate is therefore set from the platform specification. Under 8%
is an upper bound, so `r params$dbr` is
conservative and `dbr.sd` sets the tolerance.

```{r, eval=build_all_sce}
sample_dirs  = list.dirs(data_dir, full.names = TRUE, recursive = FALSE)
sample_dirs  = sample_dirs[grep("smith_annotation", sample_dirs, invert = TRUE)]
sample_names = basename(sample_dirs)

cds_list = list()
for (i in seq_along(sample_dirs)) {
  cds_list[[i]] = readRDS(paste0(sample_dirs[i], "/", sample_names[i],
                                 "_cds_mito_filter.Rds"))
}
names(cds_list) = sample_names

all.sce = list()
for (cds in names(cds_list)) {
  raw_counts     = counts(cds_list[[cds]])
  all.sce[[cds]] = SingleCellExperiment(list(counts = raw_counts),
                                        colData = cds_list[[cds]]@colData)
  all.sce[[cds]]@colData$sample = cds
}
```

```{r, eval=build_all_sce}
set.seed(100)
all.sce.dbl = lapply(all.sce, function(x)
  scDblFinder(x, dbr = params$dbr, dbr.sd = params$dbr_sd))

n_before = sapply(all.sce, ncol)
n_dbl    = sapply(all.sce.dbl, function(x) sum(x$scDblFinder.class == "doublet"))

for (cds in names(all.sce)) {
  keep = colData(all.sce.dbl[[cds]])$scDblFinder.class == "singlet"
  all.sce[[cds]] = all.sce[[cds]][, keep]
}

n_after = sapply(all.sce, ncol)

# written out rather than only printed, so the rate achieved is still
# reportable on a reknit that loads the cached objects
write.csv(data.frame(sample = names(n_before), n_before = n_before,
                     n_doublet = n_dbl, n_after = n_after,
                     pct_doublet = round(100 * n_dbl / n_before, 2),
                     row.names = NULL),
          file = dbl_stats_file, row.names = FALSE)
```

```{r, eval=build_all_sce}
all.sce  <- lapply(all.sce, logNormCounts)
all.dec  <- lapply(all.sce, modelGeneVar)
all.hvgs <- lapply(all.dec, getTopHVGs, prop = 0.1)
all.sce  <- mapply(FUN = runPCA, x = all.sce, subset_row = all.hvgs)
saveRDS(all.sce, file = all_sce_file)
```

```{r}
if (!build_all_sce) all.sce = readRDS(all_sce_file)
```

Cells removed as doublets, at `dbr` = `r params$dbr`.

```{r}
datatable(read.csv(dbl_stats_file), rownames = FALSE,
          options = list(dom = "t", ordering = FALSE))
```

# Sample summary

```{r}
df_all = do.call(rbind, lapply(names(all.sce), function(samp) {
  cd = colData(all.sce[[samp]])
  data.frame(Sample                 = samp,
             nCells                 = ncol(all.sce[[samp]]),
             medianCounts           = round(median(cd$sum)),
             medianGenesDetected    = round(median(cd$detected)),
             medianPercentMitoReads = round(median(cd$subsets_Mito_percent), 2))
}))

write.csv(df_all, file = paste0(results_dir, "/median_stats.csv"), row.names = FALSE)
datatable(df_all, rownames = FALSE, options = list(dom = "t"))
```

# Merge

Samples are merged with `fastMNN` and embedded with UMAP and t-SNE. Neither the
correction nor the embedding depends on `k`, so this step is done once and reused
across any choice of `k` in the next two sections.

```{r}
if (build_mnn) {
  if (file.exists(merged_file)) {
    # A clustered object already exists from an earlier run. fastMNN's corrected
    # space does not depend on k, so it is reused rather than recomputed.
    mnn.sce <- readRDS(merged_file)
  } else {
    # recomputed here rather than carried from above, which does not run when
    # all.sce is loaded from cache
    all.dec      <- lapply(all.sce, modelGeneVar)
    universe     <- Reduce(intersect, lapply(all.sce, rownames))
    all.sce2     <- lapply(all.sce, "[", i = universe,)
    all.dec2     <- lapply(all.dec, "[", i = universe,)
    normed.sce   <- do.call(multiBatchNorm, all.sce2)
    combined.dec <- do.call(combineVar, all.dec2)
    combined.hvg <- getTopHVGs(combined.dec, n = 5000)

    set.seed(1)
    mnn.sce <- do.call(fastMNN, c(normed.sce,
        list(subset.row = combined.hvg, BSPARAM = RandomParam())))

    set.seed(1)
    mnn.sce <- runUMAP(mnn.sce, dimred = "corrected")
    set.seed(1)
    mnn.sce <- runTSNE(mnn.sce, dimred = "corrected")
  }
  saveRDS(mnn.sce, file = mnn_file)
} else {
  mnn.sce <- readRDS(mnn_file)
}
```

# Testing `k` for stability

`k` in `buildSNNGraph` sets how many neighbors each cell connects to, and it
determines how many communities Louvain clustering finds. 

The sweep clusters ten independent 80% subsamples of the cells at each `k` and
compares pairs of subsamples on the cells they share. Agreement is the adjusted
Rand index. It asks, for every pair of cells, whether the two different clustering results
agree about putting that pair together or apart, and sums over all pairs.

The sweep takes about 45 minutes on the full object, so it is run separately by
`src/slurm_stable_k.sh`, which reads the merged object and writes to
`results/<date>_stable_k`. The call is shown here for reference.

```{r, eval=FALSE}
# strip to the coordinates: fastMNN output carries a reconstructed assay that
# nothing here reads, and find_stable_k subsets the object once per iteration
lean = SingleCellExperiment(assays = list(), colData = colData(mnn.sce))
reducedDim(lean, "corrected") = reducedDim(mnn.sce, "corrected")
colnames(lean) = colnames(mnn.sce)

stab = find_stable_k(lean,
                     k_range      = c(10, 20, 30, 40, 50),
                     n_iterations = 10,
                     use.dimred   = "corrected",   # not "PCA"; this object has none
                     prop         = 0.8,
                     seed         = 42)

plot_stability_metrics(stab)$data
```

```{r}
stable_k_file = paste0(params$project_dir, "/results/", params$results_date,
                       "_stable_k/stability_by_k.csv")

if (!file.exists(stable_k_file))
  stop("missing ", stable_k_file,
       ". The stability sweep has not been run for this results_date. ",
       "Submit src/slurm_stable_k.sh and let it finish first.")

stability_cols = c("k", "mean_n_clusters", "cv_n_clusters", "mean_ari", "min_ari")

stability = read.csv(stable_k_file, stringsAsFactors = FALSE)

missing_cols = setdiff(stability_cols, colnames(stability))
if (length(missing_cols) > 0)
  stop(stable_k_file, " is missing column(s): ",
       paste(missing_cols, collapse = ", "),
       ". It should be the stability_by_k.csv written by src/run_stable_k.R.")

stability = stability %>%
  dplyr::select(dplyr::all_of(stability_cols)) %>%
  dplyr::arrange(k)

datatable(stability %>% mutate(across(where(is.numeric), ~signif(., 4))),
          rownames = FALSE, options = list(dom = "t", ordering = FALSE))
```

`r paste0("Read from ", stable_k_file, ", written by src/run_stable_k.R.")`

`r if (!params$k %in% stability$k) paste0("**Note that k = ", params$k, " is not one of the values swept** (", paste(stability$k, collapse = ", "), "), so the dotted line below marks a k whose stability was not measured.") else ""`

```{r, fig.width=9, fig.height=3.5}
p_stab = (ggplot(stability, aes(k, mean_ari)) +
            geom_line(colour = "darkgreen") + geom_point(size = 2) +
            geom_vline(xintercept = params$k, linetype = "dotted") +
            geom_hline(yintercept = 0.9, linetype = "dashed", colour = "red") +
            # keeps the 0.8-to-1 frame when the data allows it, but widens
            # rather than clipping a k that scored below it
            coord_cartesian(ylim = c(min(0.8, min(stability$mean_ari,
                                                  na.rm = TRUE)), 1)) +
            ylab("mean ARI between subsamples") +
            ggtitle("Stability")) |
          (ggplot(stability, aes(k, mean_n_clusters)) +
            geom_line(colour = "steelblue") + geom_point(size = 2) +
            geom_vline(xintercept = params$k, linetype = "dotted") +
            ylab("mean number of clusters") +
            ggtitle("Cluster count"))

p_stab
ggsave(p_stab, file = paste0(results_dir, "/clustering_stability.png"),
       width = 9, height = 3.5)
```

```{r}
best_k = stability$k[which.max(stability$mean_ari)]
```

`r paste0("**The most stable k is ", best_k, "**, at a mean ARI of ", max(stability$mean_ari), ". The document clusters at k = ", params$k, if (best_k == params$k) ", which is that value." else paste0(", which is not that value; set the k parameter to ", best_k, " to use it."), " Every k from ", min(stability$k), " to ", max(stability$k), " clears 0.91, so the curve is a shallow arc rather than a knife edge and the choice is not load bearing.")`

**What this does and does not support.** A high ARI at a given `k` means the
clusters reproduce when the cells are perturbed at that `k`. It does not mean
the partition is insensitive to `k` itself: mean cluster count moves from 24.8
down to 17.5 across the sweep. So the sweep supports "the clustering is
reproducible at the chosen `k`", not "this is the correct number of clusters".

Worth keeping in mind that stability alone is not a sufficient criterion, since
a coarser clustering is generally more stable. Here the curve peaks in the
middle rather than rising with `k`, so the two are not simply trading off, but
the peak should not be read as the biologically correct resolution.

# Cluster

The shared nearest neighbour graph is built on the corrected space at
`k` = `r params$k`, and communities are found with Louvain.

```{r}
if (build_merged) {
  set.seed(1)
  g <- buildSNNGraph(mnn.sce, use.dimred = "corrected", k = params$k)
  set.seed(1)
  merged.sce <- mnn.sce
  colLabels(merged.sce) <- factor(igraph::cluster_louvain(g)$membership)
  saveRDS(merged.sce, file = merged_file)
  writeLines(as.character(params$k), k_stamp)
} else {
  merged.sce <- readRDS(merged_file)
}
```

```{r}
# uncorrected counts on the same cells, for every quantification below
if (build_combined) {
  universe <- Reduce(intersect, lapply(all.sce, rownames))
  all.sce2 <- lapply(all.sce, "[", i = universe,)
  all.sce2 <- lapply(all.sce2, function(x) {
    colData(x)     <- colData(x)[, "sample", drop = FALSE]
    reducedDims(x) <- list()
    rowData(x)     <- rowData(x)[, character(0)]
    x
  })
  combined.sce <- do.call(cbind, all.sce2)
  saveRDS(combined.sce, file = combined_file)
} else {
  combined.sce <- readRDS(combined_file)
}

# the clustering and the counts must describe the same cells in the same order
stopifnot(identical(colnames(combined.sce), colnames(merged.sce)))
```

`r paste0("The atlas holds ", format(ncol(merged.sce), big.mark = ","), " cells in ", nlevels(colData(merged.sce)$label), " clusters. Cluster numbering is an output of the clustering and does not correspond to any other run's numbering.")`

```{r, fig.width=6, fig.height=6}
plotTSNE(merged.sce, colour_by = "label", text_by = "label",
         text_colour = "red", point_size = .05) +
  guides(color = guide_legend(override.aes = list(size = 5)))

ggsave(file = paste0(results_dir, "/all_cells_tsne.png"), width = 6, height = 6)
```

Each library against the full embedding. The grey layer is all cells, so the
panels share one coordinate frame; subsetting the object per panel would give
each its own axis limits and make the positions incomparable.

```{r, fig.width=9, fig.height=8}
tsne_df = data.frame(reducedDim(merged.sce, "TSNE"))
names(tsne_df)[1:2] = c("TSNE1", "TSNE2")
tsne_df$batch = as.character(colData(merged.sce)$batch)

p_batch = ggplot(tsne_df, aes(TSNE1, TSNE2)) +
  geom_point(data = transform(tsne_df, batch = NULL),
             color = "grey85", size = .05) +
  geom_point(size = .05) +
  facet_wrap(~ batch)

p_batch
ggsave(p_batch, file = paste0(results_dir, "/tsne_by_library.png"),
       width = 9, height = 8)
```

# Per-cluster QC

Genes detected and total counts per cell, broken down by cluster. 

```{r}
# recomputed from the counts rather than carried from the per-sample objects,
# so the numbers describe the shared gene universe that combined.sce holds and
# that every downstream test uses
qc_cell = scuttle::perCellQCMetrics(combined.sce)

cluster_qc = data.frame(
  cluster  = colData(merged.sce)$label,
  detected = qc_cell$detected,
  sum      = qc_cell$sum,
  stringsAsFactors = FALSE)

# mitochondrial percentage cannot be recomputed here, since combined.sce does
# not record which genes are mitochondrial. It is taken from the per-sample
# objects, but only if their concatenation lines up with combined.sce exactly.
mito_src = try(do.call(rbind, lapply(all.sce, function(x) {
  cd = as.data.frame(colData(x))
  if (!"subsets_Mito_percent" %in% colnames(cd)) return(NULL)
  data.frame(barcode = colnames(x), mito = cd$subsets_Mito_percent,
             stringsAsFactors = FALSE)
})), silent = TRUE)

have_mito = !inherits(mito_src, "try-error") && !is.null(mito_src) &&
            identical(mito_src$barcode, colnames(combined.sce))
if (have_mito) cluster_qc$mito_percent = mito_src$mito
```

`r if (have_mito) "" else "**Mitochondrial percentage is not shown.** The per-sample objects either do not carry it or their cell order does not match the combined object, so it is omitted rather than aligned on an assumption."`

```{r}
qc_summary = cluster_qc %>%
  group_by(cluster) %>%
  summarize(n_cells        = dplyr::n(),
            median_genes   = round(median(detected)),
            median_counts  = round(median(sum)),
            median_mito    = if (have_mito) round(median(mito_percent), 2) else NA_real_,
            .groups = "drop") %>%
  arrange(desc(median_genes))

if (!have_mito) qc_summary$median_mito = NULL

write.csv(qc_summary, file = paste0(results_dir, "/cluster_qc.csv"),
          row.names = FALSE)

datatable(qc_summary, rownames = FALSE,
          options = list(dom = "t", paging = FALSE))
```

```{r, fig.width=9, fig.height=4}
p_genes = ggplot(cluster_qc, aes(x = cluster, y = detected)) +
  geom_violin(fill = "grey85", colour = NA, scale = "width") +
  geom_boxplot(width = .15, outlier.size = .2) +
  ylab("genes detected per cell") + xlab("cluster")

p_genes
ggsave(p_genes, file = paste0(results_dir, "/cluster_qc_genes.png"),
       width = 9, height = 4)
```

```{r, fig.width=9, fig.height=4}
p_counts = ggplot(cluster_qc, aes(x = cluster, y = sum)) +
  geom_violin(fill = "grey85", colour = NA, scale = "width") +
  geom_boxplot(width = .15, outlier.size = .2) +
  scale_y_log10() +
  ylab("counts per cell, log10") + xlab("cluster")

p_counts
ggsave(p_counts, file = paste0(results_dir, "/cluster_qc_counts.png"),
       width = 9, height = 4)
```

```{r, fig.width=9, fig.height=4, eval=have_mito}
p_mito = ggplot(cluster_qc, aes(x = cluster, y = mito_percent)) +
  geom_violin(fill = "grey85", colour = NA, scale = "width") +
  geom_boxplot(width = .15, outlier.size = .2) +
  ylab("% mitochondrial reads") + xlab("cluster")

p_mito
ggsave(p_mito, file = paste0(results_dir, "/cluster_qc_mito.png"),
       width = 9, height = 4)
```

```{r}
# how far the extremes sit from the middle, in the units the plots are in.
# with_ties = FALSE because two clusters on the same median would otherwise
# return two rows each and the sentence below would render both values
med_of_med = median(qc_summary$median_genes)
top_qc     = qc_summary %>% slice_max(median_genes, n = 1, with_ties = FALSE)
bot_qc     = qc_summary %>% slice_min(median_genes, n = 1, with_ties = FALSE)
```

`r paste0("Median genes per cell ranges from ", bot_qc$median_genes, " in cluster ", bot_qc$cluster, " to ", top_qc$median_genes, " in cluster ", top_qc$cluster, ", against a median across clusters of ", med_of_med, ". Some spread is expected, since cell types genuinely differ in how much RNA they carry, so this is a flag to follow up rather than a fault in itself. The way to tell the two apart is whether the cluster's markers hold up against genes matched for expression level; a cluster that scores high on unrelated genes too is being separated by cell depth.")`

# Stage and library metadata

```{r}
library_meta = data.frame(
  batch   = c("UND1_C00004336", "UND2_C00004362", "TF_C00004337", "TM_C00004374"),
  library = c("Undifferentiated.1", "Undifferentiated.2", "Female", "Male"),
  stage   = c("larva", "larva", "transformer", "transformer"),
  stringsAsFactors = FALSE)

observed   = sort(unique(as.character(colData(merged.sce)$batch)))
unexpected = setdiff(observed, library_meta$batch)
if (length(unexpected) > 0)
  stop("Library identifier(s) in the object but not in library_meta: ",
       paste(unexpected, collapse = ", "),
       ". Add them rather than silently dropping cells.")

cell_key = data.frame(barcode = colnames(combined.sce),
                      label   = as.character(colData(merged.sce)$label),
                      batch   = as.character(colData(merged.sce)$batch),
                      stringsAsFactors = FALSE) %>%
  left_join(library_meta, by = "batch")

all_clusters = as.character(sort(as.integer(unique(cell_key$label))))

datatable(library_meta, rownames = FALSE, options = list(dom = "t"))
```

**Stage is confounded with collection date and technical batch.** Larvae were
collected in July and transformers in November, so nothing computed from this
dataset can say how much of the larva-to-transformer difference is
developmental. The analysis below does not try. It asks whether the populations
correspond, not what makes them differ.

# Cluster composition

```{r}
# one function, used here and again after the split, so the two sets of tables
# are built the same way and any difference between them is a real one
composition_of = function(lab_vec, lab_levels) {
  data.frame(label   = factor(as.character(lab_vec), levels = lab_levels),
             library = cell_key$library,
             stringsAsFactors = FALSE) %>%
    dplyr::count(label, library, name = "n") %>%
    tidyr::complete(label, library = library_meta$library,
                    fill = list(n = 0)) %>%
    tidyr::pivot_wider(id_cols = label, names_from = library,
                       values_from = n, values_fill = 0) %>%
    mutate(larva       = Undifferentiated.1 + Undifferentiated.2,
           transformer = Female + Male,
           Total       = larva + transformer) %>%
    arrange(label)
}

add_percents = function(tbl) {
  tbl %>%
    mutate(across(c(Undifferentiated.1, Undifferentiated.2, Female, Male,
                    larva, transformer),
                  ~round(100 * .x / Total, 1),
                  .names = "pct_{.col}"))
}

comp_bar = function(tbl, title) {
  tbl %>%
    dplyr::select(label, all_of(library_meta$library)) %>%
    tidyr::pivot_longer(-label, names_to = "library", values_to = "n") %>%
    mutate(library = factor(library, levels = library_meta$library)) %>%
    ggplot(aes(x = n, y = label, fill = library)) +
    geom_col(position = "fill") +
    scale_fill_brewer(palette = "Set2") +
    xlab("proportion of cells") + ylab("cluster") + ggtitle(title)
}

stage_bar = function(tbl, title) {
  tbl %>%
    dplyr::select(label, larva, transformer) %>%
    tidyr::pivot_longer(-label, names_to = "stage", values_to = "n") %>%
    ggplot(aes(x = n, y = label, fill = stage)) +
    geom_col(position = "fill") +
    scale_fill_manual(values = c(larva = "#009292", transformer = "#9e2f7f")) +
    xlab("proportion of cells") + ylab("cluster") + ggtitle(title)
}

flag_one_library = function(tbl, threshold = 90) {
  add_percents(tbl) %>%
    rowwise() %>%
    mutate(max_lib_pct = max(c_across(c(pct_Undifferentiated.1,
                                        pct_Undifferentiated.2,
                                        pct_Female, pct_Male)))) %>%
    ungroup() %>%
    filter(max_lib_pct >= threshold) %>%
    dplyr::select(label, Total, max_lib_pct)
}
```

```{r}
comp_pre = composition_of(cell_key$label, all_clusters)

write.csv(add_percents(comp_pre),
          file = paste0(results_dir, "/cluster_composition.csv"),
          row.names = FALSE)

datatable(comp_pre, rownames = FALSE, options = list(dom = "t", paging = FALSE))
```

```{r, fig.width=6, fig.height=5}
p_comp_pre = comp_bar(comp_pre, "By library")
p_comp_pre
ggsave(p_comp_pre, file = paste0(results_dir, "/cluster_composition_by_library.png"),
       width = 6, height = 5)
```

```{r, fig.width=6, fig.height=5}
p_stage_pre = stage_bar(comp_pre, "By stage")
p_stage_pre
ggsave(p_stage_pre, file = paste0(results_dir, "/cluster_composition_by_stage.png"),
       width = 6, height = 5)
```

```{r}
one_lib_pre = flag_one_library(comp_pre)
```

`r if (nrow(one_lib_pre) == 0) "No cluster draws 90% or more of its cells from a single library." else paste0("**", nrow(one_lib_pre), " cluster(s) draw 90% or more of their cells from one library: ", paste(one_lib_pre$label, collapse = ", "), ".** For those, any contrast against another cluster is also a contrast between animals, and where that library is a transformer it is a contrast between sexes as well. This has to be reported alongside their marker lists, since the design separates none of it.")`

# Do the clusters contain the same populations at both stages?

Each cluster's markers are computed from larval cells alone and, separately,
from transformer cells alone, on uncorrected values. If the genes that define
cluster 8 among larvae are the genes that define cluster 8 among transformers,
and they match cluster 8 better than any other cluster, then cluster 8 is the
same population on both sides.

**No corrected values enter this test**, so its answer does not depend on the
`fastMNN` correction being right. And it does not require comparing 
larva to transformer gene expression directly, which is why it survives the confound.

**Everything is read against a reference computed the same way.**
Undifferentiated.1 against Undifferentiated.2 is two animals of the same stage
collected on the same date, so it shows what correspondence looks like when only
replicate variation separates the sides. That is the ceiling, not 100%.

## Clusters with enough cells from each side to compare

```{r}
min_cells_side = 20

side_counts = cell_key %>%
  dplyr::count(label, library, name = "n") %>%
  tidyr::pivot_wider(id_cols = label, names_from = library,
                     values_from = n, values_fill = 0) %>%
  mutate(larva       = Undifferentiated.1 + Undifferentiated.2,
         transformer = Female + Male) %>%
  arrange(as.integer(label))

write.csv(side_counts, file = paste0(results_dir, "/cells_per_cluster_per_side.csv"),
          row.names = FALSE)

# paging = FALSE because dom = "t" hides the pagination control without
# disabling pagination, so a table longer than ten rows silently loses the rest
datatable(side_counts, rownames = FALSE,
          options = list(dom = "t", paging = FALSE))
```

```{r}
core_clusters = side_counts %>%
  filter(Undifferentiated.1 >= min_cells_side,
         Undifferentiated.2 >= min_cells_side,
         transformer        >= min_cells_side) %>%
  pull(label)

# One cell depth per cluster, shared by the replicate and stage comparisons, so the
# floor is measured on the same number of cells as the test it calibrates.
# Subsampling each comparison to its own minimum would put them on different
# cell counts and the margins would not be comparable.
core_n = side_counts %>%
  filter(label %in% core_clusters) %>%
  transmute(label, n = pmin(Undifferentiated.1, Undifferentiated.2, transformer))
core_n = setNames(core_n$n, core_n$label)
```

`r paste0(length(core_clusters), " of ", nrow(side_counts), " clusters have at least ", min_cells_side, " cells in Undifferentiated.1, in Undifferentiated.2 and in the transformers, so both comparisons run on that set.")`

## Correspondence

### Procedure

Two comparisons are run, each splitting the cells into two sides:

* **replicate**, Undifferentiated.1 against Undifferentiated.2
* **stage**, the two larval libraries pooled against the two transformer
  libraries pooled

Both run on the same cluster set at the same per-cluster cell depth, so their results
are directly comparable and the replicate one serves as the reference for the
other.

For each comparison:

1. **Set a cell depth for every cluster.** This is
   `min(n in UND1, n in UND2, n in transformers)` for that cluster. The larval
   pool is UND1 plus UND2 so it can never be the smallest and is left out of the
   minimum.

2. **Repeat the following 10 times.**

    a. From side A, draw that many cells of every eligible cluster, without
       replacement. Do the same from side B. Both sides therefore contribute an
       identical number of cells for every cluster, so neither side has more
       statistical power than the other.

    b. Run `findMarkers` separately on each side's drawn cells, with the eligible
       clusters as the grouping and **the same cluster set as the competing
       background on both sides**. Settings are `test.type = "t"`,
       `direction = "up"`, `pval.type = "some"`, and `lfc` set, so the ranking
       reflects effect size rather than cell number. 

    c. For each cluster on each side, **keep only genes below the FDR cutoff**,
       order them by FDR then by p-value, and take the top `n_top` of those. The
       order is imposed explicitly rather than taken from whatever order
       `findMarkers` returns. The filter matters because `findMarkers` ranks
       every gene rather than returning only the significant ones, so an
       unfiltered top `n_top` would pad a weakly marked cluster's list with
       whatever ranked next. A list is therefore as long as the cluster has
       evidence for, up to `n_top`, and can be empty.

    d. Build a square matrix over the eligible clusters. Entry `[i, j]` is the
       percentage overlap between side A's list for cluster `i` and side B's list
       for cluster `j`, using the **shorter of the two lists as the denominator**:
       `100 * length(intersect(A_i, B_j)) / min(length(A_i), length(B_j))`. The
       denominator is not always `n_top`, because a cluster with few significant
       genes returns a shorter list.

3. **Average the 10 matrices elementwise.** Everything below is read off the
   averaged matrix, not off any single draw.

4. **For each cluster `i`, read four numbers off that matrix.**

    * `self_similarity` is the diagonal, `M[i, i]`: how far cluster `i`'s list on
      one side agrees with its own list on the other.
    * `best_forward` is the column of the largest value in row `i`: the cluster on
      side B that side A's cluster `i` most resembles.
    * `best_reverse` is the row of the largest value in column `i`: the cluster on
      side A that side B's cluster `i` most resembles.
    * `margin` is `M[i, i]` minus the largest off-diagonal value in row `i`: how
      far the self-match beats its strongest competitor. Negative means some other
      cluster matched better than the cluster matched itself.

5. **A cluster corresponds when it is its own best match in both directions**,
   that is `best_forward == i` and `best_reverse == i`. This is stricter than a
   positive margin, since a diffuse cluster can be several clusters' best partner
   without any of them choosing it back.

6. **Classify each cluster** using the replicate result as the control:
   a cluster whose replicate comparison is not reciprocal is **untestable at this
   cell depth**, since it cannot find itself even between two animals of the same
   stage; otherwise it **corresponds** or **fails** according to its stage result.

### What equal cell depth does and does not equalize

Step 2a gives both sides the same number of cells, so neither side is favored by
sample size. It does not make the two sides equally informative, and three things
still differ when one side is scarce.

A side drawn from a pool barely larger than the draw returns nearly the same
cells every time, so averaging over the 10 draws reduces its noise very little,
while a side drawn from thousands averages over ten genuinely different subsets.

Fewer cells also yield fewer significant genes, and since the lists are filtered
to significant genes, that side's list is shorter and the denominator in step 2d
shrinks with it. On a list of ten, one gene overlapping by chance moves the score
by ten points, so a self-match can lose to a competitor on noise alone. The
`sig_larva` and `sig_transformer` columns report the lengths directly, so a
result resting on a short list is visible rather than inferred.

And the cells themselves are differently representative. Where a cluster holds
almost all of one side's cells, the draw takes the entire population including
whatever sits at its boundary, while the other side contributes a random sample
of a well-populated core. The cluster boundary was drawn by a clustering computed
on all cells, so it was set largely by whichever side is larger.

These are reasons a scarce side can fail without the two populations differing,
which is why the untestable category exists and why failures are reported
alongside their cell counts.

```{r}
n_top      = 50
fdr_cut    = 0.01
lfc_thresh = params$lfc_thresh
n_draws    = 10

# Significant genes only, then the top n of those.
#
# findMarkers ranks every gene rather than returning only the significant ones,
# so taking the top n without filtering would hand a cluster with eight real
# markers a list of n padded with whatever ranked next. Only the significant
# part can be expected to recur on the other side, so the padding measures the
# stability of a weak tail. Filtering first means a list is as long as the
# cluster has evidence for, and never longer.
top_markers_for = function(marker_list, lab, n = n_top, fdr = fdr_cut) {
  if (is.null(marker_list) || !lab %in% names(marker_list)) return(character(0))
  df = marker_list[[lab]]
  if (nrow(df) == 0) return(character(0))
  # sorted explicitly rather than trusting the order findMarkers returns
  ord = if (all(c("FDR", "p.value") %in% colnames(df))) order(df$FDR, df$p.value)
        else if ("FDR" %in% colnames(df)) order(df$FDR) else seq_len(nrow(df))
  keep = ord[which(df$FDR[ord] < fdr)]
  rownames(df)[keep][seq_len(min(n, length(keep)))]
}

# Per-gene effect size: the mean across every pairwise comparison.
# summary.logFC is not used, because under pval.type = "some" it comes from
# whichever pairwise comparison supplied the summary p-value, which varies per
# gene and per run.
marker_effect = function(df) {
  lf = grep("^logFC\\.", colnames(df), value = TRUE)
  if (length(lf) == 0) return(rep(NA_real_, nrow(df)))
  v = rowMeans(as.matrix(as.data.frame(df[, lf, drop = FALSE])), na.rm = TRUE)
  names(v) = rownames(df)
  v
}

# lfc is set because with thousands of cells a log-fold change of 0.02 clears
# FDR 0.01, and an unthresholded ranking measures sample size rather than effect
markers_for_cells = function(cells, clusters_keep) {
  lab = factor(cell_key$label[cells], levels = clusters_keep)
  findMarkers(combined.sce[, cells], lab, test.type = "t",
              direction = "up", lfc = lfc_thresh, pval.type = "some")
}

similarity_matrix = function(mA, mB, clusters) {
  tA = lapply(clusters, function(cl) top_markers_for(mA, cl)); names(tA) = clusters
  tB = lapply(clusters, function(cl) top_markers_for(mB, cl)); names(tB) = clusters
  sim = matrix(NA_real_, length(clusters), length(clusters),
               dimnames = list(clusters, clusters))
  for (i in clusters) for (j in clusters) {
    d = min(length(tA[[i]]), length(tB[[j]]))
    sim[i, j] = if (d == 0) NA_real_ else
      100 * length(intersect(tA[[i]], tB[[j]])) / d
  }
  # lists are now significant genes only, so their length is the number of
  # markers the cluster actually has at this cell depth, capped at n_top. It is both
  # the diagnostic and the denominator above, and a cluster with none gives an
  # empty list, which scores NA rather than zero.
  list(sim   = sim,
       sig_A = lengths(tA),
       sig_B = lengths(tB))
}

run_comparison = function(sideA_libs, sideB_libs, clusters_keep, n_by_cluster,
                          seed = 1) {
  inA = which(cell_key$library %in% sideA_libs)
  inB = which(cell_key$library %in% sideB_libs)

  stopifnot(all(clusters_keep %in% names(n_by_cluster)))
  n_by_cluster = n_by_cluster[clusters_keep]

  avail = sapply(clusters_keep, function(cl)
    min(sum(cell_key$label[inA] == cl), sum(cell_key$label[inB] == cl)))
  if (any(avail < n_by_cluster))
    stop("requested cell depth exceeds available cells for cluster(s): ",
         paste(clusters_keep[avail < n_by_cluster], collapse = ", "))

  set.seed(seed)
  acc = NULL; acc_sigA = NULL; acc_sigB = NULL
  for (d in seq_len(n_draws)) {
    draw_side = function(idx) {
      unlist(lapply(clusters_keep, function(cl) {
        pool = idx[cell_key$label[idx] == cl]
        # sampled by position: sample(x, n) with x of length 1 draws from
        # seq_len(x) instead of from x
        pool[sample(length(pool), n_by_cluster[[cl]])]
      }))
    }
    out = similarity_matrix(markers_for_cells(draw_side(inA), clusters_keep),
                            markers_for_cells(draw_side(inB), clusters_keep),
                            clusters_keep)
    acc      = if (is.null(acc))      out$sim   else acc      + out$sim
    acc_sigA = if (is.null(acc_sigA)) out$sig_A else acc_sigA + out$sig_A
    acc_sigB = if (is.null(acc_sigB)) out$sig_B else acc_sigB + out$sig_B
  }
  list(sim   = acc      / n_draws,
       sig_A = acc_sigA / n_draws,
       sig_B = acc_sigB / n_draws,
       n_by_cluster = n_by_cluster)
}
```

Cell numbers are matched per cluster across every side of every comparison.
Library is deliberately **not** blocked, because library is the variable being
measured. Ten independent draws, similarity matrices averaged before any verdict
is read off them.

```{r}
# The depth scheme is part of the cache key, not just the tuning parameters, so
# results computed under a different scheme are not silently reused.
depth_tag = paste0("commonN", sum(core_n))

comparison_file = paste0(Robject_out_dir, "/stage_correspondence_", depth_tag,
                         "_", n_top, "_lfc", lfc_thresh, "_", n_draws, "draws.Rds")

if (file.exists(comparison_file)) {
  comparisons = readRDS(comparison_file)
} else {
  comparisons = list(
    replicate = run_comparison("Undifferentiated.1", "Undifferentiated.2",
                               core_clusters, core_n),
    stage     = run_comparison(c("Undifferentiated.1", "Undifferentiated.2"),
                               c("Female", "Male"), core_clusters, core_n))
  saveRDS(comparisons, file = comparison_file)
}
```

A cluster **corresponds** when it is its own best match in both directions,
which is stricter than being the best match one way, since a diffuse cluster can
be many clusters' best partner without any of them being its own. The **margin**
is how far the self-match sits above the strongest competitor.

```{r}
reciprocal_from_sim = function(sim) {
  cls = rownames(sim)
  do.call(rbind, lapply(cls, function(i) {
    row = sim[i, ]; col = sim[, i]; off = row[names(row) != i]
    fwd = names(which.max(row)); rev = names(which.max(col))
    best_off = if (all(is.na(off))) NA_real_ else max(off, na.rm = TRUE)
    data.frame(cluster = i, self_similarity = sim[i, i],
               best_forward = if (length(fwd)) fwd else NA_character_,
               best_reverse = if (length(rev)) rev else NA_character_,
               reciprocal   = identical(fwd, i) && identical(rev, i),
               margin       = sim[i, i] - best_off,
               stringsAsFactors = FALSE)
  }))
}

recip = bind_rows(lapply(names(comparisons), function(nm)
  reciprocal_from_sim(comparisons[[nm]]$sim) %>%
    mutate(comparison = nm,
           n_per_side = comparisons[[nm]]$n_by_cluster[cluster],
           # mean over draws of how many of each side's top n_top genes were
           # actually significant. The lists are always n_top long; this is how
           # much of each is real rather than ranking noise.
           sig_A      = round(comparisons[[nm]]$sig_A[cluster], 1),
           sig_B      = round(comparisons[[nm]]$sig_B[cluster], 1)))) %>%
  dplyr::select(comparison, cluster, n_per_side, sig_A, sig_B,
                self_similarity, best_forward, best_reverse, reciprocal, margin)

write.csv(recip, file = paste0(results_dir, "/SuppTable_reciprocal_matches.csv"),
          row.names = FALSE)

datatable(recip %>% mutate(across(where(is.numeric), ~round(., 1))),
          filter = "top", rownames = FALSE)
```

## Check that clusters correspond between two larval animals

A cluster that does not correspond to itself **between two larval animals**, at
the cell depth its stage test runs at, has not shown it is resolvable at that cell depth by
any comparison. Its stage result carries no information, and calling it a failure
would be reading noise as disagreement.

```{r}
stage_tbl = recip %>% filter(comparison == "stage") %>%
  dplyr::select(cluster, n_per_side,
                sig_larva = sig_A, sig_transformer = sig_B,
                self_stage = self_similarity,
                margin_stage = margin, reciprocal_stage = reciprocal,
                best_forward, best_reverse)

rep_tbl = recip %>% filter(comparison == "replicate") %>%
  dplyr::select(cluster, self_rep = self_similarity,
                margin_rep = margin, reciprocal_rep = reciprocal)

outcome_tbl = stage_tbl %>%
  left_join(rep_tbl, by = "cluster") %>%
  left_join(side_counts %>% dplyr::select(cluster = label, larva, transformer),
            by = "cluster") %>%
  mutate(outcome = dplyr::case_when(
           !reciprocal_rep  ~ "untestable at this cell depth",
           reciprocal_stage ~ "corresponds across stages",
           TRUE             ~ "fails across stages"),
         # which side's marker list broke the match
         direction = dplyr::case_when(
           reciprocal_stage                                  ~ "both directions agree",
           best_forward == cluster & best_reverse != cluster ~ "transformer side fails",
           best_forward != cluster & best_reverse == cluster ~ "larval side fails",
           best_forward == best_reverse                      ~ "both point to the same other cluster",
           TRUE                                              ~ "both fail, and disagree")) %>%
  dplyr::select(cluster, larva, transformer, n_per_side,
                sig_larva, sig_transformer,
                self_rep, margin_rep, reciprocal_rep,
                self_stage, margin_stage, reciprocal_stage,
                best_forward, best_reverse, direction, outcome) %>%
  arrange(match(outcome, c("fails across stages", "corresponds across stages",
                           "untestable at this cell depth")),
          as.integer(cluster))

write.csv(outcome_tbl, file = paste0(results_dir, "/SuppTable_stage_outcomes.csv"),
          row.names = FALSE)

datatable(outcome_tbl %>% mutate(across(where(is.numeric), ~round(., 1))),
          filter = "top", rownames = FALSE)
```

```{r}
short_cut  = 0.6 * n_top
short_side = outcome_tbl %>%
  filter(outcome == "fails across stages",
         pmin(sig_larva, sig_transformer) < short_cut) %>%
  dplyr::select(cluster, larva, transformer, n_per_side,
                sig_larva, sig_transformer, margin_stage)
```

**`sig_larva` and `sig_transformer` are how long each side's marker list actually
was**, averaged over the draws, capped at `r n_top`.

The lists hold significant genes only, so a cluster's list is as long as its
evidence and no longer. That length is also the denominator in step 2d, which
means a cluster with few markers is scored on a small number: on a list of ten,
a single chance overlap is worth ten points. Read a high similarity on a short
list with that in mind, and read a failure on a short list as underpowered
rather than as disagreement.

A cluster with no significant genes on one side scores `NA` rather than zero,
which lands it in the untestable category rather than counting as a failure.

`r if (nrow(short_side) == 0) paste0("Every failing cluster has at least ", round(short_cut), " significant genes on both sides, so none of the failures is explained by a mostly-noise list.") else paste0("**", nrow(short_side), " of the failing clusters have fewer than ", round(short_cut), " significant genes on one side: ", paste(short_side$cluster, collapse = ", "), ".** Read their results as underpowered rather than as disagreement, and compare the count against the cell count in the same row: few significant genes on the side with few cells is the expected pattern.")`

```{r}
n_corresponds = sum(outcome_tbl$outcome == "corresponds across stages")
n_fails       = sum(outcome_tbl$outcome == "fails across stages")
n_untestable  = sum(outcome_tbl$outcome == "untestable at this cell depth")
n_readable    = n_corresponds + n_fails

rep_recip = sum(recip$reciprocal[recip$comparison == "replicate"])
rep_n     = sum(recip$comparison == "replicate")

fails_tbl = outcome_tbl %>% filter(outcome == "fails across stages")
```

`r paste0("**", rep_recip, " of ", rep_n, " clusters correspond between the two larval animals**, at the same cell depths the stage test runs at. Across the stage boundary, ", n_corresponds, " of ", n_readable, " readable clusters correspond and ", n_fails, " do not.")`

`r if (n_untestable == 0) "Every cluster is resolvable at its own cell depth, so no stage result is discarded for lack of cells." else paste0(n_untestable, " cluster(s) are untestable at this cell depth: ", paste(outcome_tbl$cluster[outcome_tbl$outcome == "untestable at this cell depth"], collapse = ", "), ". These fail to find themselves even between the two larval animals, so their stage result reflects cell numbers rather than stage. They are reported and not counted either way.")`

**Read the `direction` column before deciding what a failure means.** A cluster
whose larval markers still pick it out while its transformer markers pick out
something else is usually short of transformer cells. A cluster that fails in both directions, pointing at two different partners, is the strongest failure. A cluster where both directions agree on the
same other cluster is the only case where merging with that cluster is
indicated.

```{r, fig.width=13, fig.height=5}
sim_long = bind_rows(lapply(names(comparisons), function(nm)
  as.data.frame(as.table(comparisons[[nm]]$sim)) %>%
    setNames(c("side_A", "side_B", "similarity")) %>%
    mutate(comparison = nm))) %>%
  mutate(side_A = factor(side_A, levels = all_clusters),
         side_B = factor(side_B, levels = all_clusters))

p_sim = ggplot(sim_long, aes(x = side_B, y = side_A, fill = similarity)) +
  geom_tile() +
  scale_fill_viridis_c(name = "top 50\noverlap %") +
  facet_wrap(~ comparison, nrow = 1, scales = "free") +
  xlab("cluster, second side") + ylab("cluster, first side") +
  theme(axis.text = element_text(size = 7))

p_sim
ggsave(p_sim, file = paste0(results_dir, "/cross_boundary_similarity.png"),
       width = 13, height = 5)
```

```{r, fig.width=9, fig.height=4}
p_margin = outcome_tbl %>%
  mutate(cluster = factor(cluster, levels = all_clusters)) %>%
  dplyr::select(cluster, margin_rep, margin_stage) %>%
  tidyr::pivot_longer(-cluster, names_to = "comparison", values_to = "margin") %>%
  ggplot(aes(cluster, margin, fill = comparison)) +
  geom_col(position = position_dodge()) +
  geom_hline(yintercept = 0, linewidth = .3) +
  scale_fill_manual(values = c(margin_rep = "#009292", margin_stage = "#9e2f7f")) +
  ylab("margin over strongest competitor, points")

p_margin
ggsave(p_margin, file = paste0(results_dir, "/self_match_margin.png"),
       width = 9, height = 4)
```

# Splitting the clusters that do not correspond

A cluster that fails in both directions and has enough cells on either side to describe is not one population. It is split into a larval and a transformer half, and everything downstream runs on that labeling.

The other failures are left whole. A cluster failing in one direction only, with
62 to 250 cells on its sparser side, would yield a half too small to
characterize, and the failure is better explained by that scarcity.

```{r}
split_min_cells  = 300   # per side, enough to compute a marker profile
split_min_margin = 20    # replicate margin, so the cluster is well separated

split_candidates = outcome_tbl %>%
  filter(outcome == "fails across stages",
         direction == "both fail, and disagree",
         margin_rep >= split_min_margin,
         larva >= split_min_cells, transformer >= split_min_cells) %>%
  pull(cluster)

datatable(outcome_tbl %>%
            filter(outcome == "fails across stages") %>%
            dplyr::select(cluster, larva, transformer, margin_rep, direction) %>%
            mutate(splits = cluster %in% split_candidates),
          rownames = FALSE, options = list(dom = "t", paging = FALSE))
```

**The split halves differ by stage, collection date, technical batch, animal and, where one side is a single transformer library, by sex as well.** All of those vary together and this design separates none of them. The split records that the two halves are not one population by marker evidence. It does not say what makes them different.

```{r}
# Each split cluster becomes two labels; every other cluster keeps its own, so
# the marker computation later runs over one factor covering all cells and the
# split halves compete against the same background as everything else.
final_label = cell_key$label
for (cl in split_candidates) {
  hit = cell_key$label == cl
  final_label[hit] = paste0(cl, "_", cell_key$stage[hit])
}

# unsplit clusters first in their original order, then the split halves, so the
# ordering is stable rather than alphabetical
final_label = factor(final_label,
                     levels = c(setdiff(all_clusters, split_candidates),
                                sort(unique(final_label[grepl("_", final_label)]))))

cell_key$final_label = as.character(final_label)

stopifnot(!any(is.na(final_label)),
          length(final_label) == nrow(cell_key))

# Cell type names supplied by the Chung-Davidson lab for this labelling. Their
# 12L and 12T are the larval and transformer halves of the split cluster.
anno = tibble::tribble(
  ~final_label,      ~cell_type,
  "1",               "Fibroblasts",
  "2",               "Mesonephric cells",
  "3",               "Migrating erythrocytes",
  "4",               "Granulosa cells",
  "5",               "Oogonia/Oocytes",
  "6",               "Endothelial cells",
  "7",               "Podocytes",
  "8",               "Erythrocytes",
  "9",               "Migrating pre-granulosa cells",
  "10",              "Hemopoietic stem cells",
  "11",              "Migrating immune cells",
  "12_larva",        "Mitotic macrophages",
  "12_transformer",  "Migrating primordial germ cells",
  "13",              "Macrophages",
  "14",              "Migrating/mitotic pre-granulosa cells",
  "15",              "Mitotic primordial germ cells",
  "16",              "Spermatocytes I",
  "17",              "Spermatocytes II",
  "18",              "Migrating male germ cells")

# These names describe this clustering at this k and this doublet rate. Louvain
# renumbers on any rerun, so a mismatch here means the names no longer apply and
# must be re-derived rather than reused.
missing_anno = setdiff(levels(final_label), anno$final_label)
extra_anno   = setdiff(anno$final_label, levels(final_label))
if (length(missing_anno) > 0 || length(extra_anno) > 0)
  stop("The annotation does not match this labelling. ",
       if (length(missing_anno)) paste0("Unnamed label(s): ",
           paste(missing_anno, collapse = ", "), ". ") else "",
       if (length(extra_anno)) paste0("Named but absent: ",
           paste(extra_anno, collapse = ", "), ". ") else "",
       "Re-derive the annotation for this clustering before relabelling.")

cell_key = cell_key %>% left_join(anno, by = "final_label")

# The per-cell assignment, saved. Everything else this document writes is
# aggregated by label, so without this the split labelling exists only for the
# duration of the knit and nothing downstream can use it.
cell_labels = cell_key %>%
  dplyr::select(barcode, batch, library, stage,
                cluster = label, final_label, cell_type)

write.csv(cell_labels,
          file = paste0(results_dir, "/cell_cluster_labels.csv"),
          row.names = FALSE)

# also as an Rds, so a downstream document can attach it to an SCE by name
# without a read.csv and a match()
saveRDS(setNames(cell_labels$final_label, cell_labels$barcode),
        file = paste0(Robject_out_dir, "/final_label.Rds"))

datatable(as.data.frame(table(final_label = final_label)) %>%
            dplyr::rename(n_cells = Freq) %>%
            dplyr::select(final_label, n_cells),
          rownames = FALSE, options = list(dom = "t", paging = FALSE))
```

The cell type names are held back until after the marker sections, so the
evidence is read before the interpretation. They are applied in **Cell type
annotation** below.

`r if (length(split_candidates) == 0) "No cluster met the conditions, so the final labelling is the original clustering unchanged." else paste0("Cluster(s) ", paste(split_candidates, collapse = ", "), " were split, giving ", nlevels(final_label), " labels from ", length(all_clusters), " clusters.")`

The per-cell assignment is written to `cell_cluster_labels.csv`, one row per
barcode with its library, stage, original cluster and final label, and as
`Robjects/final_label.Rds`, a named vector for attaching to an object directly:

```r
labs <- readRDS("<results_dir>/Robjects/final_label.Rds")
sce$final_label <- labs[colnames(sce)]
```

Note that `merged.sce` is saved before the split, so its own `label` column is
the original clustering. Anything wanting the split labelling has to take it
from one of these two files rather than from the object.

## Composition of the final labelling

```{r}
comp_post = composition_of(cell_key$final_label, levels(final_label))

write.csv(add_percents(comp_post),
          file = paste0(results_dir, "/cluster_composition_after_split.csv"),
          row.names = FALSE)

datatable(comp_post, rownames = FALSE, options = list(dom = "t", paging = FALSE))
```

```{r, fig.width=6, fig.height=5}
p_comp_post = comp_bar(comp_post, "By library, after split")
p_comp_post
ggsave(p_comp_post,
       file = paste0(results_dir, "/cluster_composition_by_library_after_split.png"),
       width = 6, height = 5)
```

```{r, fig.width=6, fig.height=5}
p_stage_post = stage_bar(comp_post, "By stage, after split")
p_stage_post
ggsave(p_stage_post,
       file = paste0(results_dir, "/cluster_composition_by_stage_after_split.png"),
       width = 6, height = 5)
```

```{r}
one_lib_post = flag_one_library(comp_post)
new_flags    = setdiff(one_lib_post$label, one_lib_pre$label)
```

`r if (length(split_candidates) == 0) "No cluster was split, so this matches the composition table above." else paste0("Only cluster(s) ", paste(split_candidates, collapse = ", "), " differ from the table above, each replaced by two rows.")`

`r if (length(new_flags) == 0) "The split introduces no cluster that is 90% or more one library beyond those already flagged." else paste0("**The split creates ", length(new_flags), " new single-library cluster(s): ", paste(new_flags, collapse = ", "), ".** That is expected, since the split is along the stage boundary and the transformer side of these clusters is dominated by one library, but it means the resulting half is a single animal and its markers should be described accordingly.")`

# Cluster marker genes

Markers for every cluster of the final labeling, computed on uncorrected
counts with library blocked. The split halves are tested against the same
competing set as every other cluster and against each other, so the lists are
comparable across the whole atlas.

**No effect size threshold is applied to the test.** `findMarkers` is run without
`lfc`, so the null is that the log fold change is zero and the saved result is
complete. Effect size is applied afterwards, as a filter on `meanLogFC`, where a
working list is needed.

That is deliberate but it is not the same test as `lfc = 1`. With `lfc` set, the
null becomes "the log fold change is no greater than 1" and the p-value answers
that question directly. Filtering afterwards asks a weaker pair of questions,
whether the change is distinguishable from zero and separately whether it is
large. Both are defensible; only the first is a claim that the effect exceeds
the threshold. Anything reported as significantly greater than `r lfc_thresh`
should be recomputed with `lfc` set rather than taken from here.

```{r}
# no lfc in the call, so the cached object is not interchangeable with one
# written by an earlier lfc-thresholded version. The filename says so.
marker_file_rds = paste0(Robject_out_dir, "/final_markers_nolfc.Rds")

if (file.exists(marker_file_rds)) {
  final_markers = readRDS(marker_file_rds)
} else {
  final_markers = findMarkers(combined.sce, final_label, test.type = "t",
                              block = cell_key$library, direction = "up",
                              pval.type = "some")
  saveRDS(final_markers, file = marker_file_rds)
}

gene_description = read.csv(paste0(object_dir, "/gene_descritption.csv"),
                            stringsAsFactors = FALSE)
```

```{r}
# both counts, because FDR alone is not a useful cutoff at this cell number:
# with tens of thousands of cells a log fold change of 0.02 clears FDR 0.01 and
# the count measures sample size rather than biology
marker_counts = purrr::map_dfr(names(final_markers), function(lab) {
  d = final_markers[[lab]]
  eff = marker_effect(d)
  data.frame(cluster        = lab,
             n_FDR          = sum(d$FDR < 0.01),
             n_FDR_and_lfc  = sum(d$FDR < 0.01 & eff >= lfc_thresh),
             stringsAsFactors = FALSE)
})

datatable(marker_counts, rownames = FALSE,
          options = list(dom = "t", paging = FALSE))
```

`r paste0("The gap between the two columns is the point of the effect size filter: FDR alone gives a median of ", format(round(median(marker_counts$n_FDR)), big.mark = ","), " genes per cluster, and requiring a mean log fold change of at least ", lfc_thresh, " leaves a median of ", format(round(median(marker_counts$n_FDR_and_lfc)), big.mark = ","), ".")`

`r if (all(marker_counts$n_FDR_and_lfc > 0)) "Every cluster has at least one marker passing both." else paste0("**Cluster(s) ", paste(marker_counts$cluster[marker_counts$n_FDR_and_lfc == 0], collapse = ", "), " have no gene passing both FDR < 0.01 and a mean log fold change of ", lfc_thresh, ".** Such a cluster is not distinguishable from the rest of the atlas at this effect size, and it will be absent from the overlap and ontology results below rather than appearing as an empty column. For a split half this is a direct argument that the split was not warranted; for an original cluster it says the cluster has no profile of its own once the others compete. Its genes are still in the unfiltered file, so the cutoff can be relaxed and the consequence seen.")`

Two files are written. The unfiltered one holds every gene tested against every
cluster, with no significance or effect size cutoff, so a gene absent from the
significant list can still be looked up rather than assumed untested. The
filtered one is the working list used everywhere below.

```{r}
# built once without a cutoff; the filtered table is derived from it, so the two
# cannot drift apart
markers_unfiltered = purrr::map_dfr(names(final_markers), function(lab) {
  df  = final_markers[[lab]]
  ord = order(df$FDR, df$p.value)
  data.frame(cluster   = lab,
             Gene      = rownames(df)[ord],
             p.value   = df$p.value[ord],
             FDR       = df$FDR[ord],
             meanLogFC = marker_effect(df)[ord],
             stringsAsFactors = FALSE)
}) %>%
  left_join(gene_description, by = "Gene")

unfiltered_file = paste0(results_dir, "/cluster_marker_genes_unfiltered.csv")
write.csv(markers_unfiltered, file = unfiltered_file, row.names = FALSE)

# the working list: significant and of a usable size. Both cutoffs are applied
# here rather than in the test, so the unfiltered table above stays complete.
markers_all = markers_unfiltered %>%
  filter(FDR < 0.01, meanLogFC >= lfc_thresh) %>%
  # ordered by effect within a cluster, since FDR alone ranks by cell number
  group_by(cluster) %>%
  arrange(desc(meanLogFC), FDR, .by_group = TRUE) %>%
  ungroup()

write.csv(markers_all, file = paste0(results_dir, "/cluster_marker_genes.csv"),
          row.names = FALSE)
```

`r paste0("**cluster_marker_genes_unfiltered.csv** holds ", format(nrow(markers_unfiltered), big.mark = ","), " rows, every one of ", format(dplyr::n_distinct(markers_unfiltered$Gene), big.mark = ","), " genes against each of ", dplyr::n_distinct(markers_unfiltered$cluster), " clusters, at ", round(file.size(unfiltered_file) / 1e6), " MB. **cluster_marker_genes.csv** is the same table filtered to FDR < 0.01 and mean log fold change at least ", lfc_thresh, ", ", format(nrow(markers_all), big.mark = ","), " rows. Everything below runs on the filtered list.")`

The unfiltered file carries `p.value`, `FDR` and `meanLogFC` for every gene
against every cluster, so a gene missing from the working list can be looked up
rather than assumed untested. The `.Rds` in `Robjects/` holds the full
per-comparison `logFC.*` columns if an individual pairwise contrast is needed;
`meanLogFC` is their mean, used because `summary.logFC` under
`pval.type = "some"` comes from whichever comparison supplied the summary
p-value and so is not a fixed contrast.

```{r}
datatable(markers_all %>% mutate(across(where(is.numeric), ~signif(., 3))),
          filter = "top", rownames = FALSE)
```

# Comparison with the curated markers

Rows are marker sets, columns are clusters, color is the significance of the
overlap with each cluster's full marker list, every gene passing FDR < 0.01 and
a mean log fold change of at least `r lfc_thresh`.

The lists are of unequal length, because clusters differ in how many genes
distinguish them. The hypergeometric accounts for list size, so this does not
bias the p-values. It does affect power: a cluster with a long list can reach
significance on a weaker proportional overlap than one with a short list. Read a
strong overlap in a cluster with few markers as the stronger result of the two.

```{r}
curated = read.csv(paste0(data_dir, "/", params$marker_file),
                   stringsAsFactors = FALSE) %>%
  mutate(loc_id = trimws(loc_id)) %>%
  filter(loc_id %in% rownames(combined.sce))

# Exactly two columns. phyperOverlapSets renames the first two positionally and
# ignores anything else, so an extra column would not error; it would change
# what distinct() collapses. A gene carrying two labels would then appear twice,
# inflating the set size that goes into the hypergeometric.
known_markers = curated %>%
  dplyr::select(Gene = loc_id, CellType = CellTypeMarkerSet) %>%
  distinct()

# the display label for each gene, kept separate for the reason above. The S1
# symbol is used rather than the gene table's, so figures carry the name the
# collaborators' sheet uses.
marker_symbols = curated %>%
  dplyr::select(Gene = loc_id, WangGeneID = symbol) %>%
  distinct()

stopifnot(!any(duplicated(marker_symbols$Gene)))

# the full lfc-filtered list, not a fixed top n. Feeds both the overlap heatmap
# below and the GO enrichment in the next section, so the two run on the same
# gene set rather than on different definitions of "this cluster's markers".
marker_test_genes = markers_all %>%
  dplyr::select(Gene, cluster) %>%
  mutate(cluster = paste0("cluster", cluster)) %>%
  distinct()

list_sizes = marker_test_genes %>% dplyr::count(cluster, name = "n_markers")

set_sizes = curated %>% dplyr::count(CellTypeMarkerSet, name = "n_genes") %>%
  mutate(below_coherence_floor = n_genes < 3)

datatable(set_sizes, rownames = FALSE,
          options = list(dom = "t", paging = FALSE))
```

`r paste0(nrow(known_markers), " gene-to-set assignments resolve against this count matrix, over ", dplyr::n_distinct(known_markers$CellType), " sets. ", sum(set_sizes$below_coherence_floor), " set(s) hold fewer than three genes; a single significant overlap in those rows is one gene, not a set agreeing.")`

`r paste0("On the other side of the test, the cluster marker lists run from ", format(min(list_sizes$n_markers), big.mark = ","), " to ", format(max(list_sizes$n_markers), big.mark = ","), " genes (median ", format(round(median(list_sizes$n_markers)), big.mark = ","), "), across ", nrow(list_sizes), " clusters.")`

```{r, fig.width=9}
plot_marker_overlap_heatmap(
  known_markers    = known_markers,
  test_genes       = marker_test_genes,
  background_n     = nrow(combined.sce),
  gene_description = gene_description,
  overlaps_file    = paste0(results_dir, "/cluster_marker_gene_overlaps.csv"),
  pvals_file       = paste0(results_dir, "/cluster_marker_overlap_pvalues.csv"),
  plot_file        = paste0(results_dir, "/cluster_marker_overlap_heatmap.png"))
```

# Gene ontology

```{r}
gaf_file      = file.path(params$genome_dir,
                          "GCF_010993605.1_kPetMar1.pri_gene_ontology.gaf")
go_names_file = file.path(params$genome_dir, "GOs_names.txt")
have_go       = file.exists(gaf_file) && file.exists(go_names_file)
```

`r if (!have_go) paste0("**Skipped.** The GO files were not found under ", params$genome_dir, ". Set the genome_dir parameter to the directory holding GCF_010993605.1_kPetMar1.pri_gene_ontology.gaf and GOs_names.txt, then reknit.") else ""`

```{r, eval=have_go}
gaf = read.delim(gaf_file, header = FALSE, comment = "!") %>%
  filter(!grepl("NOT", V4)) %>%
  dplyr::select(V3, V5) %>%
  dplyr::rename(GOID = "V5", Gene = "V3") %>%
  distinct()

# terms too small are noise, terms too large are uninformative
keep_go = gaf %>% group_by(GOID) %>% tally() %>%
  filter(n > 10, n < 200) %>% pull(GOID)
gaf = gaf %>% filter(GOID %in% keep_go)

go_names = read.delim(go_names_file, header = FALSE)
colnames(go_names) = c("GOID", "GOName")

go_scores = run_go_enrichment(
  gaf              = gaf,
  gene_description = gene_description,
  go_names         = go_names,
  test_genes       = marker_test_genes,
  background_n     = nrow(combined.sce),
  overlaps_file    = paste0(results_dir, "/cluster_GO_overlaps.csv"),
  pvals_file       = paste0(results_dir, "/cluster_GO_overlap_pvalues.csv"),
  pval_filter      = 0.01)

datatable(go_scores %>% mutate(across(where(is.numeric), ~signif(., 3))),
          filter = "top", rownames = FALSE)
```

# Marker expression across all clusters

Every curated marker against every cluster, on uncorrected cell depth-normalized
values. Dot size is the percentage of cells in the cluster expressing the gene,
color is the mean log expression.

The overlap heatmap measures whether a cell type's markers are *enriched* among a cluster's strongest data-derived markers, which is a different question: a gene can be enriched in a
cluster relative to the others while barely being transcribed there.

```{r}
marker_ids = intersect(unique(known_markers$Gene),
                       rownames(combined.sce))

bubble_mat = as.matrix(logcounts(combined.sce)[marker_ids, , drop = FALSE])

# labels come from cell_key, which is where final_label lives; merged.sce was
# never given that column. And they are iterated as characters over the factor
# levels rather than coerced to integer, because a split label such as
# "12_larva" becomes NA under as.integer and would be dropped silently.
bubble_lab = cell_key$final_label
stopifnot(length(bubble_lab) == ncol(bubble_mat))

marker_bubble = purrr::map_dfr(levels(final_label), function(k) {
  sub = bubble_mat[, bubble_lab == k, drop = FALSE]
  data.frame(cluster        = k,
             Gene           = marker_ids,
             n_cells        = ncol(sub),
             pct_expressing = 100 * rowMeans(sub > 0),
             mean_logexpr   = rowMeans(sub),
             stringsAsFactors = FALSE)
}) %>%
  left_join(known_markers, by = "Gene",
            relationship = "many-to-many") %>%
  left_join(marker_symbols, by = "Gene") %>%
  mutate(CellType = factor(CellType),
         cluster  = factor(cluster, levels = levels(final_label)))

stopifnot(!any(is.na(marker_bubble$WangGeneID)))

stopifnot(setequal(levels(droplevels(marker_bubble$cluster)), levels(final_label)))

write.csv(marker_bubble,
          file = paste0(results_dir, "/canonical_marker_expression_per_cluster.csv"),
          row.names = FALSE)
```

```{r, fig.width=11, fig.height=13}
p_bubble = ggplot(marker_bubble,
                  aes(x = cluster, y = WangGeneID,
                      size = pct_expressing, colour = mean_logexpr)) +
  geom_point() +
  scale_colour_viridis_c() +
  scale_size_continuous(range = c(0.4, 5)) +
  facet_grid(CellType ~ ., scales = "free_y", space = "free_y") +
  theme(strip.text.y = element_text(angle = 0, size = 6),
        axis.text.y  = element_text(size = 6),
        axis.text.x  = element_text(size = 8, angle = 45, hjust = 1),
        panel.grid.major = element_line(linewidth = .2)) +
  xlab("cluster") + ylab(NULL) +
  labs(size = "% expressing", colour = "mean logexpr")

p_bubble
ggsave(p_bubble, file = paste0(results_dir, "/canonical_marker_bubble_all_clusters.png"),
       width = 11, height = 13)
```

The underlying values are written to
`canonical_marker_expression_per_cluster.csv`, one row per gene per cluster with
the percentage expressing and the mean log expression.

# Cell type annotation

Names supplied by the Chung-Davidson lab for this labelling. They come last on
purpose: everything above is the evidence a reader would use to judge them, and
placing the names earlier would invite reading the marker and GO sections as
confirmation of a conclusion already stated.

Nothing in this document tests these names. The data-driven markers, the
comparison with the curated list and the GO enrichment above are what evidence
exists; where they disagree with a name, the disagreement is the finding.

```{r}
# captured before the pipe: inside it, final_label is the character column and
# levels() of a character vector is NULL, which would silently drop the ordering
lab_levels = levels(final_label)

anno_tbl = as.data.frame(table(final_label = final_label)) %>%
  dplyr::rename(n_cells = Freq) %>%
  mutate(final_label = as.character(final_label)) %>%
  left_join(anno, by = "final_label") %>%
  dplyr::select(final_label, cell_type, n_cells) %>%
  arrange(match(final_label, lab_levels))

write.csv(anno_tbl, file = paste0(results_dir, "/cell_type_annotation.csv"),
          row.names = FALSE)

datatable(anno_tbl, rownames = FALSE, options = list(dom = "t", paging = FALSE))
```

Two cautions travel with this table.

The names apply to **this** clustering, at this `k` and this doublet rate.
Louvain renumbers on every rerun, so a name is a statement about a position in
one particular labelling and not a durable identifier. The guard beside the
annotation halts the knit rather than relabel the wrong cells if the two stop
matching.

A name describes a **cluster**, not separately each cell in it. That matters
most for clusters holding cells from both stages, where the annotation may fit
one stage's cells and not the other's. `r if (exists("split_candidates") && length(split_candidates) > 0) paste0("It matters most of all for the split cluster(s), ", paste(split_candidates, collapse = ", "), ", whose halves carry different names precisely because the correspondence test found them not to be one population.") else ""`

# Summary

```{r}
n_clusters = nlevels(colData(merged.sce)$label)
n_core     = nrow(outcome_tbl)   # clusters that entered the comparison at all

# Every subset row carries its own denominator, because there are three of them:
# all clusters, the core set that had enough cells per side to be compared, and
# the readable set whose replicate comparison succeeded. Each level is a subset
# of the one above, and the rows are ordered so that nesting is visible.
stopifnot(n_untestable + n_readable == n_core, n_core <= n_clusters)

summary_tbl = data.frame(
  stage = c("atlas", "atlas",
            "entering the test", "entering the test",
            "of those compared", "of those compared",
            "control", "stage test", "stage test",
            "outcome", "outcome"),
  quantity = c("cells",
               "clusters before splitting",
               paste0("too few cells to compare (under ", min_cells_side,
                      " on a side)"),
               "compared",
               "untestable at their own cell depth",
               "readable across the stage boundary",
               "correspond between the two larval animals",
               "correspond across the stage boundary",
               "fail across the stage boundary",
               "clusters split",
               "labels in the final labelling"),
  value = c(format(ncol(merged.sce), big.mark = ","),
            as.character(n_clusters),
            paste0(n_clusters - n_core, " of ", n_clusters),
            paste0(n_core, " of ", n_clusters),
            paste0(n_untestable, " of ", n_core),
            paste0(n_readable, " of ", n_core),
            paste0(rep_recip, " of ", rep_n),
            paste0(n_corresponds, " of ", n_readable),
            paste0(n_fails, " of ", n_readable),
            if (length(split_candidates) == 0) "0"
              else paste0(length(split_candidates), " (",
                          paste(split_candidates, collapse = ", "), ")"),
            as.character(nlevels(final_label))),
  stringsAsFactors = FALSE)

datatable(summary_tbl, rownames = FALSE, colnames = c("", "", ""),
          options = list(dom = "t", ordering = FALSE, paging = FALSE))
```

**The control and stage-test rows are meant to be read against each other.** The
control is the same test run between the two larval animals, collected on the
same day, where nothing but animal-to-animal variation separates the two sides.
The stage test is the same test run across the July-to-November boundary. The
gap between the two is the result. The stage row alone is not, because it has no
meaning without knowing what this test scores when nothing is wrong.

**The denominators narrow down the table and each is a subset of the one above.**
Of `r n_clusters` clusters, `r n_core` had enough cells on every side to be
compared at all. Of those, `r n_readable` were readable, meaning their replicate
comparison succeeded. The stage result is out of that `r n_readable`.

Clusters drop out for two different reasons and neither is a failure. Too few
cells on a side means the comparison never ran. Untestable at their own cell
depth means it ran and the cluster could not find itself even between the two
larval animals, so a stage result would report cell numbers rather than stage.
Both are reported and counted neither way.

## What this document does not answer

* **What the stage difference is made of.** Stage, collection date and technical
  batch are one variable in this design and no analysis of this dataset can
  separate them. The correspondence test asks whether populations match, not why
  they differ.
* **What the clusters are.** Correspondence says a cluster is the same
  population at both stages. It says nothing about which cell type that is. The
  curated marker overlap is suggestive and no more, since the sets are small and
  route through mammalian orthology.
* **Independent evidence for the stage-associated patterns.** That needs an
  external dataset or staining.

# Session info

```{r}
sessionInfo()
```
