---
title: "14_germ_cell_trajectory"
author: "Stephanie Hickey"
date: "`r Sys.Date()`"
output:
  html_document:
    code_folding: hide
    toc: true
    toc_float:
      toc_collapsed: true
params:
    project_dir: /mnt/ufs18/rs-013/bioinformaticsCore/projects/chong_davidson/BCC125_sea_lamprey_scRNA
    project_name: BCC125_sea_lamprey_scRNA
    lab: Weiming Li
    results_date: "20260813"
    germ_clusters: "5,12_transformer,15,16,17,18"
    root_node: "PGC.migrating"
    min_library_cells: 50
    genome_dir: /mnt/research/bioinformaticsCore/shared/Genomes/Petromyzon_marinus
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

```{r, echo=FALSE, results="asis"}
cat("\n")
cat("#### Lab:", params$lab, " \n")
cat("#### Project:", params$project_name, " \n")
```

```{r}
library(tidyverse)
library(SingleCellExperiment)
library(scater)
library(scran)
library(patchwork)
library(RColorBrewer)
library(readxl)
library(batchelor)
library(DT)
library(TSCAN)
library(rtracklayer)
library(BiocSingular)
library(ggraph)
library(tidygraph)
library(scatterpie)
source(paste0(params$project_dir, "/src/scRNA_seq_functions.R"))
```

Pseudotime over the germ cell clusters, following the structure of
`03_trajectory_analysis_no_doublets.Rmd` but built on the labelling and
annotation from `13_atlas_and_stage_correspondence.Rmd` rather than on the
sex-cell subclustering in `02`.

# What this document can and cannot establish

**The ordering it produces cannot be shown to be developmental from this dataset.**
Larvae were collected in July and transformers in November, so stage, collection
date, technical batch and animal vary together. A trajectory orders cells along
the largest axis of variation in the data, and here that axis coincides with the
library boundary. The ordering is therefore equally consistent with germ cell
maturation and with the difference between collections.

This is a property of the design, not of the algorithm. Running a different
trajectory method recovers the same axis, so nothing below should be read as
evidence that the axis is developmental.

**What it is for.** The genes that change along the ordering are the output that
matters. Those are to be tested against independently collected bulk RNA-seq,
where the July-to-November batch does not exist, and that comparison is what can
distinguish maturation from batch. This document produces the candidate list;
it does not adjudicate it.

The library composition of every node is reported below and drawn onto the
trajectory graph, so the extent of the confound is visible on the figure rather
than only in this paragraph.

```{r}
object_dir  = paste0(params$project_dir, "/results/", params$results_date,
                     "_no_doublets")
atlas_dir   = paste0(params$project_dir, "/results/", params$results_date,
                     "_atlas")
results_dir = paste0(params$project_dir, "/results/", params$results_date,
                     "_germ_trajectory")
Robject_out_dir = paste0(results_dir, "/Robjects")
if (!dir.exists(Robject_out_dir)) dir.create(Robject_out_dir, recursive = TRUE)
data_dir    = paste0(params$project_dir, "/data")
```

```{r}
label_file = paste0(atlas_dir, "/cell_cluster_labels.csv")
if (!file.exists(label_file))
  stop("missing ", label_file,
       ". Run 13_atlas_and_stage_correspondence.Rmd first.")

cell_labels = read.csv(label_file, stringsAsFactors = FALSE)
all.sce     = readRDS(paste0(object_dir, "/all.sce.Rds"))

library_meta = data.frame(
  batch   = c("UND1_C00004336", "UND2_C00004362", "TF_C00004337", "TM_C00004374"),
  library = c("Undifferentiated.1", "Undifferentiated.2", "Female", "Male"),
  stage   = c("larva", "larva", "transformer", "transformer"),
  stringsAsFactors = FALSE)

lib_cols = c("Female" = "#FF6DB6", "Male" = "#006ddb",
             "Undifferentiated.1" = "#009292", "Undifferentiated.2" = "#b66ddf")
```

# The germ cells

```{r}
germ_clusters = trimws(strsplit(params$germ_clusters, ",")[[1]])

absent = setdiff(germ_clusters, cell_labels$final_label)
if (length(absent) > 0)
  stop("cluster(s) ", paste(absent, collapse = ", "), " are not in ", label_file,
       ". Labels present: ",
       paste(sort(unique(cell_labels$final_label)), collapse = ", "),
       ". Cluster numbering changes on every rerun of the clustering, so set ",
       "the germ_clusters parameter from the current labelling.")

# short node names, since these become trajectory path names and the annotated
# cell type names contain spaces and slashes
node_names = tibble::tribble(
  ~final_label,     ~node,
  "5",              "Oogonia.Oocytes",
  "12_transformer", "PGC.migrating",
  "15",             "PGC.mitotic",
  "16",             "Spermatocytes.I",
  "17",             "Spermatocytes.II",
  "18",             "MGC.migrating")

unnamed = setdiff(germ_clusters, node_names$final_label)
if (length(unnamed) > 0)
  stop("no short node name for cluster(s): ", paste(unnamed, collapse = ", "),
       ". Add them to node_names.")

germ_key = cell_labels %>%
  filter(final_label %in% germ_clusters) %>%
  left_join(node_names, by = "final_label")

germ_cells = germ_key$barcode
```

```{r}
germ_tab = germ_key %>%
  dplyr::count(final_label, node, cell_type, library, name = "n") %>%
  tidyr::pivot_wider(id_cols = c(final_label, node, cell_type),
                     names_from = library, values_from = n, values_fill = 0) %>%
  mutate(Total = rowSums(across(any_of(library_meta$library)))) %>%
  arrange(match(final_label, germ_clusters))

write.csv(germ_tab, file = paste0(results_dir, "/germ_cluster_composition.csv"),
          row.names = FALSE)

datatable(germ_tab, rownames = FALSE, options = list(dom = "t", paging = FALSE))
```

```{r}
lib_totals = germ_key %>% dplyr::count(library, name = "n") %>%
  mutate(pct = round(100 * n / sum(n), 1)) %>%
  arrange(desc(n))

top_lib = lib_totals$library[1]
top_pct = lib_totals$pct[1]

datatable(lib_totals, rownames = FALSE, options = list(dom = "t"))
```

`r paste0("**", format(length(germ_cells), big.mark = ","), " germ cells across ", length(germ_clusters), " clusters, of which ", top_pct, "% come from a single library, ", top_lib, ".** Read every result below against that number. A trajectory over a set this dominated by one animal will place that animal's cells along most of the ordering whether or not the ordering is developmental.")`

# Build the germ cell object

```{r}
germ_file = paste0(Robject_out_dir, "/merged.germ.sce.Rds")

if (file.exists(germ_file)) {
  merged.germ.sce = readRDS(germ_file)
} else {
  germ_list_all = lapply(all.sce, function(x) {
    keep = intersect(colnames(x), germ_cells)
    if (length(keep) == 0) return(NULL)
    x[, keep]
  })
  germ_list_all = germ_list_all[!vapply(germ_list_all, is.null, logical(1))]

  n_by_lib = vapply(germ_list_all, ncol, integer(1))
  keep_lib = names(n_by_lib)[n_by_lib >= params$min_library_cells]
  germ_list = germ_list_all[keep_lib]

  germ_dec  = lapply(germ_list, modelGeneVar)
  universe  = Reduce(intersect, lapply(germ_list, rownames))
  germ_list2 = lapply(germ_list, "[", i = universe,)
  germ_dec2  = lapply(germ_dec,  "[", i = universe,)

  normed    = do.call(multiBatchNorm, germ_list2)
  comb_dec  = do.call(combineVar, germ_dec2)
  comb_hvg  = getTopHVGs(comb_dec, n = 5000)

  set.seed(1)
  merged.germ.sce = do.call(fastMNN, c(normed,
      list(subset.row = comb_hvg, BSPARAM = RandomParam())))

  set.seed(1)
  merged.germ.sce = runUMAP(merged.germ.sce, dimred = "corrected")
  set.seed(1)
  merged.germ.sce = runTSNE(merged.germ.sce, dimred = "corrected")

  saveRDS(merged.germ.sce, file = germ_file)
}

# node labels attached by barcode, never by position
node_lookup = setNames(germ_key$node, germ_key$barcode)
colData(merged.germ.sce)$node = factor(node_lookup[colnames(merged.germ.sce)],
                                       levels = node_names$node[
                                         node_names$final_label %in% germ_clusters])
stopifnot(!any(is.na(colData(merged.germ.sce)$node)))

colData(merged.germ.sce)$library =
  library_meta$library[match(as.character(colData(merged.germ.sce)$batch),
                             library_meta$batch)]
```

```{r}
# the uncorrected counts on the same cells, for the pseudotime gene test
combined_file = paste0(Robject_out_dir, "/combined.germ.sce.Rds")

if (file.exists(combined_file)) {
  combined.germ.sce = readRDS(combined_file)
} else {
  germ_list = lapply(all.sce, function(x) {
    keep = intersect(colnames(x), colnames(merged.germ.sce))
    if (length(keep) == 0) return(NULL)
    x[, keep]
  })
  germ_list = germ_list[!vapply(germ_list, is.null, logical(1))]
  universe  = Reduce(intersect, lapply(germ_list, rownames))
  germ_list = lapply(germ_list, function(x) {
    x = x[universe, ]
    colData(x)     = colData(x)[, "sample", drop = FALSE]
    reducedDims(x) = list()
    rowData(x)     = rowData(x)[, character(0)]
    x
  })
  combined.germ.sce = do.call(cbind, germ_list)
  saveRDS(combined.germ.sce, file = combined_file)
}

# align to the corrected object rather than assuming a shared order
combined.germ.sce = combined.germ.sce[, colnames(merged.germ.sce)]
stopifnot(identical(colnames(combined.germ.sce), colnames(merged.germ.sce)))

colData(combined.germ.sce)$node = colData(merged.germ.sce)$node
```

`r paste0("The merged germ object holds ", format(ncol(merged.germ.sce), big.mark = ","), " cells over ", nlevels(colData(merged.germ.sce)$node), " nodes.")`

```{r, fig.width=7, fig.height=6}
p_nodes = plotTSNE(merged.germ.sce, colour_by = "node", point_size = .4) +
  guides(color = guide_legend(override.aes = list(size = 4)))
p_libs  = plotTSNE(merged.germ.sce, colour_by = "library", point_size = .4) +
  scale_colour_manual(values = lib_cols) +
  guides(color = guide_legend(override.aes = list(size = 4)))

p_nodes / p_libs
ggsave(file = paste0(results_dir, "/germ_tsne_node_and_library.png"),
       width = 7, height = 10)
```

The two panels are the whole problem in one figure. If the node panel and the
library panel show the same partition, the trajectory below is tracing the
libraries.

# Gene annotation

```{r}
gene_description = read.csv(paste0(object_dir, "/gene_descritption.csv"),
                            stringsAsFactors = FALSE)

gaf_file      = file.path(params$genome_dir,
                          "GCF_010993605.1_kPetMar1.pri_gene_ontology.gaf")
go_names_file = file.path(params$genome_dir, "GOs_names.txt")
have_go       = file.exists(gaf_file) && file.exists(go_names_file)
```

```{r, eval=have_go}
gaf = read.delim(gaf_file, header = FALSE, comment = "!") %>%
  filter(!grepl("NOT", V4)) %>%
  dplyr::select(V3, V5) %>%
  dplyr::rename(GOID = "V5", Gene = "V3") %>%
  distinct()

keep_go = gaf %>% group_by(GOID) %>% tally() %>%
  filter(n > 10, n < 200) %>% pull(GOID)
gaf = gaf %>% filter(GOID %in% keep_go)

go_names = read.delim(go_names_file, header = FALSE)
colnames(go_names) = c("GOID", "GOName")
```

# Trajectory

## In t-SNE space

```{r, fig.width=7, fig.height=6}
set.seed(1)
by.cluster = aggregateAcrossCells(merged.germ.sce,
                                  ids = colData(merged.germ.sce)$node,
                                  use.assay.type = "reconstructed")

centroids = reducedDim(by.cluster, "corrected")
mst       = createClusterMST(centroids, clusters = NULL)
line.data = reportEdges(by.cluster, mst = mst, clusters = NULL,
                        use.dimred = "TSNE")

plotTSNE(merged.germ.sce, colour_by = "node", point_size = .5) +
  geom_line(data = line.data,
            mapping = aes(x = TSNE1, y = TSNE2, group = edge)) +
  guides(color = guide_legend(override.aes = list(size = 5)))

ggsave(file = paste0(results_dir, "/germ_trajectory_tsne.png"),
       width = 7, height = 6)
```

## In batch-corrected expression space

The trajectory is computed in corrected space, so this view is the one it was
actually fitted in.

```{r, fig.width=7, fig.height=6}
line.data_corrected = reportEdges(by.cluster, mst = mst, clusters = NULL,
                                  use.dimred = "corrected")

plotReducedDim(merged.germ.sce, colour_by = "node", dimred = "corrected",
               point_size = .4) +
  geom_line(data = line.data_corrected,
            mapping = aes(x = dim1, y = dim2, group = edge)) +
  guides(color = guide_legend(override.aes = list(size = 5)))

ggsave(file = paste0(results_dir, "/germ_trajectory_corrected.png"),
       width = 7, height = 6)
```

## Pseudotime

```{r}
if (!params$root_node %in% levels(colData(merged.germ.sce)$node))
  stop("root_node '", params$root_node, "' is not one of the nodes: ",
       paste(levels(colData(merged.germ.sce)$node), collapse = ", "))
```

`r paste0("The ordering is rooted at **", params$root_node, "**, chosen as the least mature of the annotated germ populations. The root is an assumption, not a result: the same graph rooted elsewhere gives a different ordering, and nothing in the data fixes it.")`

```{r}
pseudo_file = paste0(Robject_out_dir, "/germ.pseudo.Rds")

if (file.exists(pseudo_file)) {
  pseudo = readRDS(pseudo_file)
} else {
  set.seed(1)
  pseudo = quickPseudotime(merged.germ.sce,
                           clusters   = colData(merged.germ.sce)$node,
                           use.dimred = "corrected",
                           outgroup   = FALSE,
                           start      = params$root_node)
  saveRDS(pseudo, file = pseudo_file)
}

common.pseudo = averagePseudotime(pseudo$ordering)
merged.germ.sce$Pseudotime   = common.pseudo
combined.germ.sce$Pseudotime = common.pseudo
```

## Trajectory node graph

Pie charts give the library composition of each node, and the lower panel colors
nodes by mean pseudotime. **These two panels together are the diagnostic.** If
the pies change composition along the same direction pseudotime increases, the
ordering and the library boundary are the same axis.

```{r, fig.width=7, fig.height=7}
meta_df = as.data.frame(colData(merged.germ.sce))

sample_tally = meta_df %>%
  mutate(node = as.factor(node), library = as.factor(library)) %>%
  group_by(node, library, .drop = FALSE) %>%
  tally()

gr = as_tbl_graph(pseudo$mst)

node_meta = sample_tally %>%
  tidyr::pivot_wider(names_from = library, values_from = n, values_fill = 0) %>%
  ungroup()
for (l in names(lib_cols)) if (!l %in% names(node_meta)) node_meta[[l]] = 0
node_meta = node_meta %>%
  mutate(totalCells = rowSums(across(all_of(names(lib_cols))))) %>%
  dplyr::rename(name = "node") %>%
  mutate(name = as.character(name))

mean_pseudo = meta_df %>%
  group_by(node) %>%
  summarise(mean_pseudo = mean(Pseudotime, na.rm = TRUE), .groups = "drop") %>%
  dplyr::rename(name = "node") %>%
  mutate(name = as.character(name))

gr = gr %>%
  activate(nodes) %>%
  left_join(node_meta, by = "name") %>%
  arrange(desc(totalCells)) %>%
  left_join(mean_pseudo, by = "name")

layout_data = create_layout(gr, layout = "stress")

gr_plot =
  ggraph(layout_data) +
  geom_edge_link() +
  geom_scatterpie(aes(x, y), cols = names(lib_cols), data = layout_data,
                  colour = NA, pie_scale = 1.5) +
  scale_fill_manual(name = "Library", values = lib_cols) +
  guides(fill = guide_legend(keywidth = .8, keyheight = .8)) +
  geom_node_text(aes(label = name), color = "black", size = 2.6) +
  theme_void(base_size = 8) +
  theme(legend.position = "left", legend.title = element_text(size = 8),
        legend.text = element_text(size = 8),
        plot.tag = element_text(size = 12)) +
  labs(tag = "A")

gr_pseudo =
  ggraph(layout_data) +
  geom_edge_link() +
  geom_node_point(aes(color = mean_pseudo), size = 9) +
  scale_color_viridis_c(option = "viridis") +
  geom_node_text(aes(label = name), vjust = .75, size = 2.6, color = "red") +
  scale_x_continuous(expand = expansion(mult = 0.15)) +
  scale_y_continuous(expand = expansion(mult = 0.15)) +
  theme_void(base_size = 8) +
  theme(legend.position = "left", legend.title = element_text(size = 8),
        legend.text = element_text(size = 8),
        plot.tag = element_text(size = 12)) +
  labs(tag = "B")

(gr_plot + coord_fixed()) / (gr_pseudo + coord_fixed())

ggsave(file = paste0(results_dir, "/germ_pseudo_mst.pdf"), width = 7, height = 7)
```

```{r}
# how far pseudotime is explained by library alone
pseudo_lib_r2 = summary(lm(meta_df$Pseudotime ~ meta_df$library))$r.squared
pseudo_stage_r2 = {
  st = library_meta$stage[match(meta_df$library, library_meta$library)]
  if (dplyr::n_distinct(st) > 1)
    summary(lm(meta_df$Pseudotime ~ st))$r.squared else NA_real_
}
```

`r paste0("**Library of origin explains ", round(100 * pseudo_lib_r2, 1), "% of the variance in pseudotime", if (!is.na(pseudo_stage_r2)) paste0(", and stage alone explains ", round(100 * pseudo_stage_r2, 1), "%") else "", ".** This is the number to quote when describing the ordering. It does not say the ordering is wrong; it says how much of it is indistinguishable from the collection structure.")`

# Pseudotime-associated genes

Genes changing along each path, on **uncorrected** counts. All significant genes
at FDR < 0.01 are written to `germ_pseudotime_de_genes.csv`.

```{r}
pseudo_mat  = pathStat(pseudo$ordering)
test_pseudo = TSCAN::testPseudotime(combined.germ.sce, pseudotime = pseudo_mat)

test_pseudo_filt = list()
for (i in seq_along(test_pseudo)) {
  test_pseudo[[i]]$SYMBOL = rownames(combined.germ.sce)
  test_pseudo[[i]]$Path   = names(test_pseudo)[i]
  test_pseudo_filt[[i]] = as.data.frame(test_pseudo[[i]]) %>%
    filter(FDR < .01) %>%
    arrange(desc(abs(logFC)))
}

symbol_description = gene_description %>% dplyr::rename(SYMBOL = "Gene")

test_pseudo_filt_df = do.call(rbind, test_pseudo_filt) %>%
  left_join(symbol_description, by = "SYMBOL") %>%
  dplyr::select(Path, SYMBOL, WangGeneID, description, logFC, p.value, FDR)

write.csv(test_pseudo_filt_df,
          file = paste0(results_dir, "/germ_pseudotime_de_genes.csv"),
          row.names = FALSE)

test_pseudo_dir = test_pseudo_filt_df %>%
  mutate(Path = ifelse(logFC > 0, paste0(Path, ".Upregulated"),
                                  paste0(Path, ".Downregulated")))

test_pseudo100 = test_pseudo_dir %>%
  group_by(Path) %>%
  slice_min(order_by = FDR, n = 100, with_ties = FALSE) %>%
  ungroup() %>%
  mutate(logFC = round(logFC, 2),
         p.value = formatC(signif(p.value, 2)),
         FDR = formatC(signif(FDR, 2)))

datatable(test_pseudo100, filter = "top", rownames = FALSE)
```

`r paste0(format(nrow(test_pseudo_filt_df), big.mark = ","), " gene-path results at FDR < 0.01, over ", dplyr::n_distinct(test_pseudo_filt_df$Path), " path(s). **This list is the output to carry forward.** It is the input to the external bulk comparison, which is the only place the developmental question can be settled.")`

## Trajectory heatmaps

```{r}
colData(combined.germ.sce)$pseudotime_TSCAN = averagePseudotime(pseudo$ordering)

top_by_path = test_pseudo_dir %>%
  group_by(Path) %>%
  arrange(FDR, desc(abs(logFC)), .by_group = TRUE) %>%
  slice_head(n = 25) %>%
  summarise(SYMBOL = list(SYMBOL), .groups = "drop") %>%
  tibble::deframe()

p_list = list(); n_cells_list = list()
for (p in names(top_by_path)) {
  gene_ids = top_by_path[[p]]
  gene_labels = gene_description %>% filter(Gene %in% gene_ids) %>% pull(WangGeneID)

  path_name = sub("\\.(Upregulated|Downregulated)$", "", p)
  cell_subset = as.data.frame(pseudo_mat) %>%
    tibble::rownames_to_column("Barcode") %>%
    filter(!is.na(.data[[path_name]])) %>%
    pull(Barcode)

  n_cells_list[[p]] = length(cell_subset)
  p_list[[p]] = plot_pseudotime_heatmap(combined.germ.sce, gene_ids, gene_labels,
                                        cell_subset,
                                        pseudotime_column = "pseudotime_TSCAN")
}

make_heatmap_pair = function(p_list, path_name, n_cells_list = NULL) {
  up_name = paste0(path_name, ".Upregulated")
  down_name = paste0(path_name, ".Downregulated")
  if (!all(c(up_name, down_name) %in% names(p_list))) return(NULL)

  lab = function(nm) if (!is.null(n_cells_list))
    paste0(nm, " (n=", n_cells_list[[nm]], ")") else nm
  ttl = function(nm) ggplot() + labs(title = lab(nm)) + theme_void() +
    theme(plot.title = element_text(size = 8, face = "bold"),
          plot.margin = margin(t = 10))

  col1 = (ttl(up_name)   / p_list[[up_name]][[1]]   / p_list[[up_name]][[2]]) +
    plot_layout(heights = c(0.02, 0.025, 1))
  col2 = (ttl(down_name) / p_list[[down_name]][[1]] / p_list[[down_name]][[2]]) +
    plot_layout(heights = c(0.02, 0.025, 1))

  (col1 | col2) + plot_layout(guides = "collect") &
    theme(legend.position = "right")
}

path_names = unique(sub("\\.(Upregulated|Downregulated)$", "", names(p_list)))
```

```{r, fig.width=7, fig.height=4, results="asis"}
for (pn in path_names) {
  hp = make_heatmap_pair(p_list, pn, n_cells_list)
  if (is.null(hp)) next
  cat("\n\n### ", pn, " path\n\n", sep = "")
  print(hp)
  ggsave(hp, file = paste0(results_dir, "/", gsub("[^A-Za-z0-9]+", "_", pn),
                           "_heatmaps.png"), width = 8, height = 5)
}
```

# Gene ontology of pseudotime-associated genes

```{r, eval=have_go}
go_scores = run_go_enrichment(
  gaf              = gaf,
  gene_description = gene_description,
  go_names         = go_names,
  test_genes       = test_pseudo_dir %>% dplyr::select(SYMBOL, Path),
  background_n     = nrow(combined.germ.sce),
  group2_label     = "Path",
  ngroup2_label    = "nPathDEGenes",
  overlaps_file    = paste0(results_dir, "/germ_pseudotime_GO_overlaps.csv"),
  pvals_file       = paste0(results_dir, "/germ_pseudotime_GO_pvalues.csv"),
  pval_filter      = 0.05)

datatable(go_scores %>% mutate(across(where(is.numeric), ~signif(., 3))),
          filter = "top", rownames = FALSE)
```

`r if (!have_go) paste0("**GO skipped.** The annotation files were not found under ", params$genome_dir, ".") else ""`

# Export

```{r}
germ_out = data.frame(
  barcode    = colnames(merged.germ.sce),
  node       = as.character(colData(merged.germ.sce)$node),
  library    = colData(merged.germ.sce)$library,
  Pseudotime = merged.germ.sce$Pseudotime,
  stringsAsFactors = FALSE) %>%
  left_join(germ_key %>% dplyr::select(barcode, final_label, cell_type, stage),
            by = "barcode")

germ_tsne = reducedDim(merged.germ.sce, "TSNE")
germ_out$germ.TSNE1 = germ_tsne[, 1]
germ_out$germ.TSNE2 = germ_tsne[, 2]

write.csv(germ_out, file = paste0(results_dir, "/germ_cell_pseudotime.csv"),
          row.names = FALSE)

datatable(head(germ_out, 50), rownames = FALSE)
```

# Summary

```{r}
summary_tbl = data.frame(
  question = c("germ cells",
               "clusters used",
               "libraries contributing",
               "share from the largest single library",
               "root node",
               "variance in pseudotime explained by library",
               "genes at FDR < 0.01 along a path"),
  answer = c(format(ncol(merged.germ.sce), big.mark = ","),
             paste(germ_clusters, collapse = ", "),
             as.character(dplyr::n_distinct(germ_out$library)),
             paste0(top_pct, "% (", top_lib, ")"),
             params$root_node,
             paste0(round(100 * pseudo_lib_r2, 1), "%"),
             format(nrow(test_pseudo_filt_df), big.mark = ",")))

datatable(summary_tbl, rownames = FALSE,
          options = list(dom = "t", ordering = FALSE, paging = FALSE))
```

**What to do with this.** The gene list is a hypothesis about germ cell
maturation that this dataset cannot test, because the maturation axis and the
collection axis are the same axis here. Testing it means asking whether the same
genes move the same way in an independently collected series, which is what the
Yasmin comparison is for. Until that is done, the ordering should be described as
a transcriptional axis across the germ cell clusters rather than as
developmental time.

# Session info

```{r}
sessionInfo()
```
