---
title: "16_bulk_validation"
author: "Stephanie Hickey"
date: "`r Sys.Date()`"
output:
  html_document:
    code_folding: hide
    toc: true
    toc_float:
      toc_collapsed: true
params:
    project_dir: /mnt/ufs18/rs-013/bioinformaticsCore/projects/chong_davidson/BCC125_sea_lamprey_scRNA
    project_name: BCC125_sea_lamprey_scRNA
    lab: Weiming Li
    results_date: "20260813"
    genome_dir: /mnt/research/bioinformaticsCore/shared/Genomes/Petromyzon_marinus
    nfcore_dir: /mnt/research/bioinformaticsCore/projects/chong_davidson/BCC125_sea_lamprey_scRNA/results/yasmin_nfcore
    group_column: group_suppT1labels
    top_n_list: "100,250,500"
    n_boot: 2000
    n_perm_labels: 200
    min_pseudobulk_count: 10
    go_pval: 0.01
    germ_clusters: "5,15,16,17,18"
    germ_marker_lineage: all
    n_signature_genes: 100
    cores: 1
---

```{r setup, include=FALSE}
# html_document renders inline figures at knitr's 72 dpi doubled by
# fig.retina, so 144, while every ggsave here writes at 300. Rendering inline
# at 300 as well puts the figure in the document at the resolution of the file
# on disk. fig.retina is set to 1 so the two are not multiplied.
# Figures are vector by default. A raster figure in html_document is rendered
# at a fixed pixel size and then scaled to the page, so it is only ever as sharp
# as that size; svg has no resolution to run out of. The chunk that draws one
# mark per gene sets dev = "png" individually, because an svg of twenty thousand
# points is tens of megabytes where the png is a couple of hundred kilobytes.
# dpi applies to those, and fig.retina is 1 so it is not doubled again.
# svglite if it is installed, otherwise the grDevices svg device, which needs
# the cairo capability enabled below
vec_dev = if (requireNamespace("svglite", quietly = TRUE)) "svglite" else "svg"
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      dev = vec_dev, dpi = 300, fig.retina = 1)

# Compute nodes have no X11, and the default png device asks for it. Cairo
# renders off screen.
if (isTRUE(capabilities("cairo"))) options(bitmapType = "cairo")
```

```{r, echo=FALSE, results="asis"}
cat("\n")
cat("#### Lab:", params$lab, " \n")
cat("#### Project:", params$project_name, " \n")
```

```{r}
library(tidyverse)
library(SingleCellExperiment)
library(edgeR)
library(patchwork)
library(ggrepel)
library(DT)
source(paste0(params$project_dir, "/src/scRNA_seq_functions.R"))
```

Independent evidence for the stage-associated patterns. Answers the reviewer's request for
"independent evidence supporting the stage-associated patterns, such as
agreement with staining or previous bulk RNA-seq studies."

# Scope

**What agreement shows.** Whether the larva-to-transformer expression changes we
report reproduce in animals collected, dissected, prepared and sequenced by a
different group at a different time. Our July-to-November technical batch does
not exist in Yasmin's data. A pattern driven by that batch has no reason to
appear there.

**What it does not show.** That the changes are developmental. Yasmin's series
has its own collection structure and may carry its own confounds. Agreement
shows the pattern is not specific to our batch. That is what the reviewer asks
for. The claim goes no further.

**Bulk is a mixture.** A gene changing in bulk can reflect a shift in which cell
types are present rather than a change within any cell type. Pseudobulk shares
that property. The two sides are on the same footing. The composition section
measures how much of the agreement composition accounts for.

**The larval endpoint is matched by stage rather than by size.** Our larvae are
112 mm and sex-labile by our own staging. Every larval-stage animal in Yasmin's
series is smaller, the largest being 99 mm. Their 74–99 mm animals are already
classed as past the point where ovarian differentiation completes. The two series
draw the larval boundary at different sizes. The transformer endpoints match on
stage and on size.

```{r}
object_dir  = paste0(params$project_dir, "/results/", params$results_date,
                     "_no_doublets")
atlas_dir   = paste0(params$project_dir, "/results/", params$results_date,
                     "_atlas")
results_dir = paste0(params$project_dir, "/results/", params$results_date,
                     "_bulk_validation")
Robject_out_dir = paste0(results_dir, "/Robjects")
if (!dir.exists(Robject_out_dir)) dir.create(Robject_out_dir, recursive = TRUE)
data_dir    = paste0(params$project_dir, "/data")
```

```{r}
# The Yasmin reads were quantified with nf-core/rnaseq against
# GCF_010993605.1_kPetMar1.pri. Point nfcore_dir at that run's output directory.
# Either the top level or star_salmon/ itself is accepted.
#
# Length-scaled counts come first in the candidate list. They carry the
# transcript-length correction that makes gene-level Salmon estimates comparable
# across samples, and they are the right edgeR input without offsets. Plain
# gene_counts is the fallback.
nfcore_candidates = c(
  "star_salmon/salmon.merged.gene_counts_length_scaled.tsv",
  "salmon/salmon.merged.gene_counts_length_scaled.tsv",
  "star_salmon/salmon.merged.gene_counts.tsv",
  "salmon/salmon.merged.gene_counts.tsv",
  "salmon.merged.gene_counts_length_scaled.tsv",
  "salmon.merged.gene_counts.tsv")

nfcore_dir  = as.character(params$nfcore_dir)
count_path  = NA_character_

if (nzchar(nfcore_dir)) {
  hits = file.path(nfcore_dir, nfcore_candidates)
  hits = hits[file.exists(hits)]
  if (length(hits) > 0) count_path = hits[1]
}

have_bulk    = !is.na(count_path)
length_scaled = have_bulk && grepl("length_scaled", count_path)
```

`r if (!have_bulk && !nzchar(nfcore_dir)) "**The nfcore_dir parameter is not set.** Everything below is skipped. Set it to the nf-core/rnaseq output directory for the Yasmin data, either the run's top level or its star_salmon subdirectory." else ""`

`r if (!have_bulk && nzchar(nfcore_dir)) paste0("**No gene count matrix was found under ", nfcore_dir, ".** Everything below is skipped. Looked for: ", paste(nfcore_candidates, collapse = ", "), ". Found instead: ", paste(head(list.files(nfcore_dir), 20), collapse = ", "), ".") else ""`

`r if (have_bulk) paste0("Counts read from ", basename(count_path), if (length_scaled) ". The length-scaled matrix is the right edgeR input without offsets." else ". **This is the unscaled gene_counts matrix.** The transcript-length correction is absent. Use salmon.merged.gene_counts_length_scaled.tsv if the run produced one.") else ""`

# The Yasmin series

```{r, eval=have_bulk}
sample_meta = read.csv(paste0(data_dir, "/yasmin_PRJNA749754_sample_groups.csv"),
                       stringsAsFactors = FALSE)

# nf-core/rnaseq gene matrix: gene_id, gene_name, then one column per sample.
x = read.delim(count_path, check.names = FALSE, stringsAsFactors = FALSE)

id_cols = intersect(c("gene_id", "gene_name"), colnames(x))
if (!"gene_id" %in% id_cols)
  stop("no gene_id column in ", count_path, ". Columns found: ",
       paste(head(colnames(x), 10), collapse = ", "))

yasmin_counts = as.matrix(x[, setdiff(colnames(x), id_cols), drop = FALSE])
rownames(yasmin_counts) = x$gene_id

# Salmon estimates are non-integer. edgeR wants counts, and the length-scaled
# matrix is already on the count scale, so rounding is the only step needed.
yasmin_counts = round(yasmin_counts)
storage.mode(yasmin_counts) = "integer"

# nf-core sample names come from the samplesheet the run used. They may be run
# accessions, the paper's own animal names, or the two glued together, as in
# SRR15275627_F_JUV. Read the column names in all three forms and take the
# form and sample sheet column that resolve the most samples.
raw_ids = colnames(yasmin_counts)
has_acc = grepl("^[DES]RR[0-9]+", raw_ids)

id_forms = list(
  "as given"      = raw_ids,
  "run accession" = ifelse(has_acc, sub("^([DES]RR[0-9]+).*$", "\\1", raw_ids),
                           NA_character_),
  "name after the accession" =
                    ifelse(grepl("^[DES]RR[0-9]+_", raw_ids),
                           sub("^[DES]RR[0-9]+_", "", raw_ids), NA_character_))

id_candidates = c("run", "sample_name", "library_name", "suppT1_id")
id_candidates = intersect(id_candidates, colnames(sample_meta))

match_grid = expand.grid(form = names(id_forms), column = id_candidates,
                         stringsAsFactors = FALSE)
match_grid$n_matched = mapply(function(f, cl)
  sum(id_forms[[f]] %in% sample_meta[[cl]], na.rm = TRUE),
  match_grid$form, match_grid$column)

best          = match_grid[which.max(match_grid$n_matched), ]
sample_ids    = id_forms[[best$form]]
sample_id_col = best$column

if (best$n_matched < ncol(yasmin_counts))
  stop("could not match all ", ncol(yasmin_counts), " count columns to the ",
       "sample sheet. The best pairing was the '", best$form, "' form of the ",
       "column names against sample sheet column '", sample_id_col,
       "', matching ", best$n_matched, ". Unmatched: ",
       paste(raw_ids[!(sample_ids %in% sample_meta[[sample_id_col]])],
             collapse = ", "),
       ". Add the nf-core sample names to ",
       "data/yasmin_PRJNA749754_sample_groups.csv as a new column.")

absent = setdiff(sample_meta[[sample_id_col]], sample_ids)
if (length(absent) > 0)
  stop(length(absent), " sample(s) in the sample sheet are missing from the ",
       "count matrix: ", paste(absent, collapse = ", "),
       ". The nf-core run is incomplete, or its samplesheet omitted them.")

sample_meta = sample_meta[match(sample_ids, sample_meta[[sample_id_col]]), ]
stopifnot(identical(sample_meta[[sample_id_col]], sample_ids))

# Everything downstream selects columns with sample_meta$run, so put the count
# matrix on run accessions whatever form the nf-core names took.
stopifnot(!anyDuplicated(sample_meta$run))
colnames(yasmin_counts) = sample_meta$run

id_note = paste0("Count columns were matched to the sample sheet by their ",
                 best$form, ", against the `", sample_id_col, "` column. All ",
                 ncol(yasmin_counts), " samples resolved.")
```

`r if (have_bulk) id_note else ""`

```{r, eval=have_bulk}
# a numeric developmental position per animal, parsed from the recorded life
# stage rather than from the group label. The mid female group spans stage 1 to
# post-metamorphosis, so the label does not place an animal.
stage_score = function(s) {
  s = tolower(s)
  dplyr::case_when(
    grepl("post-metamorphosis", s)             ~ 8,
    grepl("metamorphosing stage", s)           ~ suppressWarnings(
                                                   as.numeric(sub(".*stage ([0-9]+).*",
                                                                  "\\1", s))),
    grepl("larval", s)                         ~ 0,
    grepl("upstream migrant", s)               ~ 10,
    TRUE                                       ~ NA_real_)
}

sample_meta$stage_score = stage_score(sample_meta$suppT1_life_stage)
stopifnot(!any(is.na(sample_meta$stage_score)))

# Which group assignment to use. The sample sheet carries two, because the paper
# and its Supplementary Table 1 split early from mid males differently: the text
# and count tables cut at metamorphic stage 5, Supplementary Table 1 cuts at
# stage 7. They disagree for M_M5 and M_M6 and agree everywhere else, so no
# female group and no female result depends on this choice.
group_col = as.character(params$group_column)
stopifnot(group_col %in% colnames(sample_meta))
sample_meta$group_used = sample_meta[[group_col]]

series_tab = sample_meta %>%
  dplyr::group_by(group_used) %>%
  dplyr::summarise(n = dplyr::n(),
                   length_mm = paste0(min(suppT1_length_mm), "-",
                                      max(suppT1_length_mm)),
                   stages = paste(sort(unique(suppT1_life_stage)), collapse = "; "),
                   .groups = "drop")

datatable(series_tab, rownames = FALSE, options = list(dom = "t", paging = FALSE))
```

`r if (have_bulk) paste0("Groups are taken from the **", group_col, "** column. ", if (group_col == "group_suppT1labels") "Supplementary Table 1 labels its own rows M_E1 to M_E5 and M_M1 to M_M4, putting metamorphic stages 1 to 6 in early males and stage 7 with post-metamorphosis in mid males. This matches how the main text describes mid males, \"metamorphosing stage 7 and immediately post-metamorphosis\". It does not reproduce the n of 4 early and 6 mid printed in the text, which the group_textcounts column does. Mid males here are the four animals at stage 7 or beyond, which is the same rule used to pick the stage-matched mid females, so the two sexes are selected alike." else "This column reproduces the group sizes printed in the main text by cutting early from mid males at metamorphic stage 5. Supplementary Table 1 labels those rows differently, cutting at stage 7, which is what group_suppT1labels holds.") else ""`

## Stage correspondence

Our animals are 112 mm sex-labile larvae and 131 mm newly transformed
transformers. The matches below use the
per-animal records in Supplementary Table 1 rather than the group labels.

```{r, eval=have_bulk}
# The animals within each mid group that are staged like our transformers, kept
# for the correspondence table. The endpoint contrasts use the whole group.
stage_matched = function(lbl) sample_meta$sample_name[
  sample_meta$group_used == lbl & sample_meta$stage_score >= 7]

len_range = function(lbl) {
  L = sample_meta$suppT1_length_mm[sample_meta$group_used == lbl]
  paste0(min(L), "-", max(L), " mm")
}

match_tab = tibble::tribble(
  ~ours,            ~yasmin,                                     ~basis,
  "larvae, 112 mm", paste0("undifferentiated larvae, ",
                           len_range("UD undifferentiated larvae")),
  "both pre-differentiation by description; ours are sex-labile by our staging; the two series set the larval size boundary differently",
  "TM, 131 mm",     paste0("all mid males, ", len_range("MM mid males")),
  paste0("stage 7 and post-metamorphosis by the Supplementary Table 1 labels; closest staged: ",
         paste(stage_matched("MM mid males"), collapse = ", ")),
  "TF, 131 mm",     paste0("all mid females, ", len_range("MF mid females")),
  paste0("group spans stage 1 to post-metamorphosis, used whole; closest staged: ",
         paste(stage_matched("MF mid females"), collapse = ", ")))

datatable(match_tab, rownames = FALSE, options = list(dom = "t", paging = FALSE))
```

# Pseudobulk

Raw counts are summed across all cells within each library, giving four samples.
Corrected and per-cell normalized values are not used. A bulk library measures
summed raw counts.

```{r, eval=have_bulk}
pseudo_file = paste0(Robject_out_dir, "/pseudobulk.Rds")

if (file.exists(pseudo_file)) {
  pseudobulk = readRDS(pseudo_file)
} else {
  all.sce = readRDS(paste0(object_dir, "/all.sce.Rds"))
  universe = Reduce(intersect, lapply(all.sce, rownames))
  pseudobulk = vapply(all.sce, function(x)
    Matrix::rowSums(counts(x[universe, ])), numeric(length(universe)))
  rownames(pseudobulk) = universe
  saveRDS(pseudobulk, file = pseudo_file)
}

library_meta = data.frame(
  batch   = c("UND1_C00004336", "UND2_C00004362", "TF_C00004337", "TM_C00004374"),
  library = c("Undifferentiated.1", "Undifferentiated.2", "Female", "Male"),
  stage   = c("larva", "larva", "transformer", "transformer"),
  stringsAsFactors = FALSE)

colnames(pseudobulk) = library_meta$library[
  match(colnames(pseudobulk), library_meta$batch)]
stopifnot(!any(is.na(colnames(pseudobulk))))
```

`r if (have_bulk) paste0("Pseudobulk over ", format(nrow(pseudobulk), big.mark = ","), " genes and ", ncol(pseudobulk), " libraries.") else ""`

## Shared gene set

```{r, eval=have_bulk}
shared = intersect(rownames(pseudobulk), rownames(yasmin_counts))

overlap_tab = data.frame(
  set = c("our single-cell genes", "Yasmin quantified genes", "shared"),
  n   = c(nrow(pseudobulk), nrow(yasmin_counts), length(shared)))

datatable(overlap_tab, rownames = FALSE, options = list(dom = "t"))

if (length(shared) < 0.5 * nrow(pseudobulk))
  stop("only ", length(shared), " of ", nrow(pseudobulk), " genes are shared. ",
       "Both sides should be NCBI kPetMar1 gene ids, so a low overlap means the ",
       "nf-core run used a different annotation from the single-cell counts. ",
       "Diagnose before continuing.")
```

`r if (have_bulk) paste0("**", format(length(shared), big.mark = ","), " genes are shared.** Everything below runs on that set. Genes are matched by identifier. Both sides come from the same assembly, so no ortholog assignment is involved.") else ""`

# Yasmin stage effects

```{r, eval=have_bulk}
y = DGEList(counts = yasmin_counts[shared, ], samples = sample_meta)
keep = filterByExpr(y, group = sample_meta$group_used)
y = y[keep, , keep.lib.sizes = FALSE]
y = calcNormFactors(y)

tested = rownames(y)
```

`r if (have_bulk) paste0(format(length(tested), big.mark = ","), " genes pass filterByExpr in the Yasmin series and carry through to the comparison.") else ""`

```{r, eval=have_bulk}
# Two contrast shapes serve the whole document: a two-group test and an ordered
# trend. Both return log fold changes, p-values and FDR per gene.
run_two_group = function(dge, run_a, run_b, label) {
  sub = dge[, c(run_a, run_b)]
  grp = factor(c(rep("early", length(run_a)), rep("late", length(run_b))),
               levels = c("early", "late"))
  des = model.matrix(~grp)
  sub = estimateDisp(sub, des)
  fit = glmQLFit(sub, des)
  res = glmQLFTest(fit, coef = 2)
  topTags(res, n = Inf, sort.by = "none")$table %>%
    tibble::rownames_to_column("Gene") %>%
    dplyr::select(Gene, logFC, PValue, FDR) %>%
    dplyr::rename_with(~paste0(label, "_", .x), -Gene)
}

run_trend = function(dge, runs, score, label) {
  sub = dge[, runs]
  des = model.matrix(~score)
  sub = estimateDisp(sub, des)
  fit = glmQLFit(sub, des)
  res = glmQLFTest(fit, coef = 2)
  topTags(res, n = Inf, sort.by = "none")$table %>%
    tibble::rownames_to_column("Gene") %>%
    dplyr::select(Gene, logFC, PValue, FDR) %>%
    dplyr::rename_with(~paste0(label, "_", .x), -Gene)
}

# Spearman with a bootstrap CI. cor.test gives a p-value but no interval, and
# the interval is the quantity worth reporting.
spearman_ci = function(x, y, n_boot = params$n_boot, seed = 1) {
  ok = is.finite(x) & is.finite(y)
  x = x[ok]; y = y[ok]
  r = suppressWarnings(cor(x, y, method = "spearman"))
  set.seed(seed)
  bs = replicate(n_boot, {
    i = sample(length(x), replace = TRUE)
    suppressWarnings(cor(x[i], y[i], method = "spearman"))
  })
  c(rho = r, lo = unname(quantile(bs, .025, na.rm = TRUE)),
    hi = unname(quantile(bs, .975, na.rm = TRUE)), n = length(x))
}

g = sample_meta$group_used
runs_ud = sample_meta$run[g == "UD undifferentiated larvae"]
runs_pm = sample_meta$run[g == "PM prospective males"]
runs_em = sample_meta$run[g == "EM early males"]
runs_mm = sample_meta$run[g == "MM mid males"]
runs_ef = sample_meta$run[g == "EF early females"]
runs_mf = sample_meta$run[g == "MF mid females"]
```

```{r, eval=have_bulk}
# Two endpoints and two trends.
#
# The endpoints are larvae against the whole mid group, one per sex, so the two
# sexes are contrasted the same way and on the largest groups available.
#
# The trends order the animals by group and then, inside each group, by the
# metamorphic stage recorded for that animal in Supplementary Table 1. The
# within-group offset is scaled to sit inside the gap to the next group, so
# groups never overlap.
#
# The offset uses the finest developmental information the paper reports. It
# changes nothing in the groups whose animals share a stage: undifferentiated
# larvae, prospective males and early females are all recorded as larval, so
# they stay flat. It orders early males (stages 1, 1, 3, 4, 5, 6), mid females
# (1, 1, 3, 4, 7, 8) and, barely, mid males (7, 7, 8, 8).
stage_offset = function(runs) {
  st = sample_meta$stage_score[match(runs, sample_meta$run)]
  (rank(st, ties.method = "min") - 1) / length(runs)
}

make_score = function(group_runs, mode) {
  unlist(lapply(seq_along(group_runs), function(i) {
    runs = group_runs[[i]]
    (i - 1) + if (mode == "stage") stage_offset(runs) else rep(0, length(runs))
  }))
}

male_end = run_two_group(y, runs_ud, runs_mm, "male_end")
fem_end  = run_two_group(y, runs_ud, runs_mf, "fem_end")

male_groups = list(runs_ud, runs_pm, runs_em, runs_mm)
male_runs   = unlist(male_groups)
male_group_index = make_score(male_groups, "group")
male_score       = make_score(male_groups, "stage")
male_trend       = run_trend(y, male_runs, male_score, "male_trend")

fem_groups = list(runs_ud, runs_ef, runs_mf)
fem_runs   = unlist(fem_groups)
fem_group_index = make_score(fem_groups, "group")
fem_score       = make_score(fem_groups, "stage")
fem_trend       = run_trend(y, fem_runs, fem_score, "fem_trend")

yasmin_res = Reduce(function(a, b) dplyr::full_join(a, b, by = "Gene"),
                    list(male_end, fem_end, male_trend, fem_trend))
```

## Trend scores

Each trend regresses expression on a single ordered covariate. The covariate
assigns a number to every animal.

The groups supply the coarse order, the one Yasmin give: undifferentiated larvae,
then prospective males, early males and mid males on the male axis;
undifferentiated larvae, early females and mid females on the female axis. That
is the `group_index` column below.

Within a group the animals are ordered by their own recorded life stage.
Supplementary Table 1 gives one for every animal. It is finer than the group
labels. Mid females span metamorphic stage 1 to post-metamorphosis, and early
males span stages 1 to 6. Scoring by group alone would discard per-animal
information the paper reports.

The offset is built so it can only order animals inside a group, never move one
past another group:

1. Parse the recorded life stage to a number. Larval is 0, metamorphosing stage
   *N* is *N*, post-metamorphosis is 8, upstream migrant is 10. `M_M1b`, recorded
   as "Larval stage/Metamorphosing stage 1", is read as stage 1.
2. Rank the animals within their group by that number, ties taking the lower
   rank.
3. Subtract one and divide by the group size, giving an offset in [0, 1).
4. Add it to the group integer.

Because the offset is always below 1, group *k* occupies [*k*, *k*+1) and the
groups stay in order. Where every animal in a group shares a stage the offset is
0 for all of them, so the two scorings agree there. That is the case for the
undifferentiated larvae, the prospective males and the early females, which are
all recorded simply as larval.

```{r, eval=have_bulk}
score_tab = dplyr::bind_rows(
  data.frame(axis = "male", run = male_runs,
             group_index = male_group_index, trend_score = male_score),
  data.frame(axis = "female", run = fem_runs,
             group_index = fem_group_index, trend_score = fem_score)) %>%
  dplyr::mutate(
    group  = sample_meta$group_used[match(run, sample_meta$run)],
    sample = sample_meta$sample_name[match(run, sample_meta$run)],
    length_mm  = sample_meta$suppT1_length_mm[match(run, sample_meta$run)],
    life_stage = sample_meta$suppT1_life_stage[match(run, sample_meta$run)]) %>%
  dplyr::arrange(axis, trend_score) %>%
  dplyr::select(axis, group, sample, length_mm, life_stage,
                group_index, trend_score)

write.csv(score_tab, file = paste0(results_dir, "/trend_scores.csv"),
          row.names = FALSE)

datatable(score_tab %>% mutate(across(where(is.numeric), ~signif(., 3))),
          rownames = FALSE, options = list(dom = "t", pageLength = 30))
```

```{r, eval=have_bulk}
contrast_tab = data.frame(
  contrast = c("male endpoint", "female endpoint",
               "male trend", "female trend"),
  design   = c(paste0("UD (n=", length(runs_ud), ") vs all mid males (n=",
                      length(runs_mm), ")"),
               paste0("UD (n=", length(runs_ud), ") vs all mid females (n=",
                      length(runs_mf), ")"),
               paste0("UD, PM, EM, MM ordered (n=", length(male_runs), ")"),
               paste0("UD, EF, MF ordered (n=", length(fem_runs), ")")),
  n_FDR_05 = c(sum(yasmin_res$male_end_FDR   < 0.05, na.rm = TRUE),
               sum(yasmin_res$fem_end_FDR    < 0.05, na.rm = TRUE),
               sum(yasmin_res$male_trend_FDR < 0.05, na.rm = TRUE),
               sum(yasmin_res$fem_trend_FDR  < 0.05, na.rm = TRUE)))

datatable(contrast_tab, rownames = FALSE, options = list(dom = "t", paging = FALSE))
```

The endpoints are the smaller designs. Both use the two undifferentiated larvae
as the early side. The correlation below uses effect sizes alone. Noise in the
Yasmin estimates pulls agreement toward zero. Low power therefore makes that test
conservative rather than invalid. Per-gene significance comes from the trends.

# Our stage effect

Two larval libraries against one transformer library per sex is not a formal
test. The log fold change is used as an effect size. No p-values are reported
from it.

Nothing on the Yasmin side requires a gene to be measurable here. Their
`filterByExpr` runs on their 28 samples. The shared set is an intersection of
identifiers. A gene with three counts in one of our libraries and none in the
others produces a large fold change on almost no evidence. Such a gene sits near
the top of a list ranked by absolute fold change. `min_pseudobulk_count` marks
the genes that carry enough counts here to be worth ranking. Both the marked and
unmarked sets are carried through the comparison. The effect of the filter is
therefore visible rather than assumed.

```{r, eval=have_bulk}
pb = pseudobulk[shared, ][tested, ]

p = DGEList(counts = pb)
p = calcNormFactors(p)
plog = cpm(p, log = TRUE, prior.count = 1)

larva_mean = rowMeans(plog[, c("Undifferentiated.1", "Undifferentiated.2")])

min_pb = as.integer(params$min_pseudobulk_count)

ours = data.frame(
  Gene          = rownames(plog),
  ours_male     = plog[, "Male"]   - larva_mean,
  ours_female   = plog[, "Female"] - larva_mean,
  max_count     = apply(pb, 1, max),
  well_measured = apply(pb, 1, max) >= min_pb,
  stringsAsFactors = FALSE)
```

`r if (have_bulk) paste0(format(sum(ours$well_measured), big.mark = ","), " of ", format(nrow(ours), big.mark = ","), " genes reach ", min_pb, " counts in at least one of our four libraries.") else ""`

# Diagnostics

This section runs two checks that decide how the comparison below should be
read. Neither of them uses the Yasmin data.

## Do the two larval libraries agree with each other?

UND1 and UND2 are the only replication in our data. The larva-to-transformer
difference has to be larger than the difference between two libraries of the
same stage, or there is no effect for any external dataset to reproduce.

The ratio of the two spreads has a known value when there is no stage effect at
all. With independent library noise of size sigma, the larval difference has
spread sigma times root two, and a transformer library against the mean of two
larval libraries has spread sigma times root one and a half. The ratio is
therefore root three quarters, or 0.866, under the null. Read the observed ratio
against that number rather than against 1.

```{r, eval=have_bulk}
und_diff = plog[, "Undifferentiated.1"] - plog[, "Undifferentiated.2"]
null_ratio = sqrt(0.75)

rep_tab = data.frame(
  quantity = c("UND1 against UND2, Spearman on log-CPM",
               "SD of the UND1 minus UND2 difference",
               "SD of our male stage effect",
               "SD of our female stage effect",
               "male effect SD divided by larval difference SD",
               "female effect SD divided by larval difference SD",
               "the same ratio when there is no stage effect"),
  value = c(
    suppressWarnings(cor(plog[, "Undifferentiated.1"],
                         plog[, "Undifferentiated.2"], method = "spearman")),
    sd(und_diff),
    sd(ours$ours_male),
    sd(ours$ours_female),
    sd(ours$ours_male)   / sd(und_diff),
    sd(ours$ours_female) / sd(und_diff),
    null_ratio))

datatable(rep_tab %>% mutate(across(where(is.numeric), ~signif(., 3))),
          rownames = FALSE, options = list(dom = "t", ordering = FALSE,
                                           paging = FALSE))
```

`r if (have_bulk) { r_m = sd(ours$ours_male) / sd(und_diff); excess = r_m / null_ratio; if (excess < 1.25) paste0("**The stage effect is barely larger than library noise.** The male ratio is ", signif(r_m, 3), " against ", signif(null_ratio, 3), " expected with no stage effect at all, an excess of ", signif(excess, 3), " fold. Library-level variation accounts for most of what is being called a stage effect, and a failure to reproduce in Yasmin would follow from that alone.") else paste0("The male ratio is ", signif(r_m, 3), " against ", signif(null_ratio, 3), " expected with no stage effect, an excess of ", signif(excess, 3), " fold. The stage effect is larger than library noise.") } else ""`

## What is in the top genes?

```{r, eval=have_bulk}
top_probe = as.integer(trimws(strsplit(as.character(params$top_n_list),
                                       ",")[[1]]))[1]

top_counts = dplyr::bind_rows(lapply(c("ours_male", "ours_female"),
  function(cl) {
    idx = order(abs(ours[[cl]]), decreasing = TRUE)[seq_len(top_probe)]
    data.frame(
      set = paste0("top ", top_probe, ", ", sub("ours_", "", cl)),
      median_max_count = median(ours$max_count[idx]),
      pct_below_cutoff = 100 * mean(!ours$well_measured[idx]))
  }))

top_counts = dplyr::bind_rows(
  top_counts,
  data.frame(set = "all genes",
             median_max_count = median(ours$max_count),
             pct_below_cutoff = 100 * mean(!ours$well_measured)))

datatable(top_counts %>% mutate(across(where(is.numeric), ~signif(., 3))),
          rownames = FALSE, options = list(dom = "t", paging = FALSE))
```

`r if (have_bulk) { pc = top_counts$pct_below_cutoff[1]; if (pc > 25) paste0("**", signif(pc, 3), "% of the top male genes fall below the count cutoff.** The unfiltered ranking is largely a low-count list, and the filtered results below are the ones to read.") else paste0(signif(pc, 3), "% of the top male genes fall below the count cutoff, so the unfiltered ranking is not dominated by low-count genes.") } else ""`

# The comparison

```{r, eval=have_bulk}
cmp = ours %>% dplyr::inner_join(yasmin_res, by = "Gene")

# The Wang identifier, so figures and the export carry the name the
# collaborators' sheets use. The table is written by 01, one row per NCBI id.
gd_file = paste0(object_dir, "/gene_descritption.csv")
have_gd = file.exists(gd_file)

if (have_gd) {
  gene_description = read.csv(gd_file, stringsAsFactors = FALSE) %>%
    dplyr::distinct(Gene, .keep_all = TRUE)
  cmp = cmp %>%
    dplyr::left_join(gene_description %>% dplyr::select(Gene, WangGeneID),
                     by = "Gene")
} else {
  gene_description = data.frame(Gene = character(), WangGeneID = character(),
                                description = character())
  cmp$WangGeneID = NA_character_
}
cmp$label = ifelse(is.na(cmp$WangGeneID), cmp$Gene, cmp$WangGeneID)

top_ns = as.integer(trimws(strsplit(as.character(params$top_n_list), ",")[[1]]))

# genome-wide, then restricted to our strongest stage genes.
#
# Relabelled contrasts drawn from the same 28 animals give a genome-wide
# correlation centered at zero, so the genome-wide value is a result in its own
# right rather than a baseline. What the two numbers compare is concentration:
# whether agreement is stronger among the genes we nominate than among genes in
# general. That is a statement about specificity.
#
# Sign agreement is scored genome-wide for the same purpose. Whether agreement
# exceeds chance is left to the permutation.
# rank_on picks which dataset selects the genes. "ours" is the original
# direction. "yasmin" ranks on their side, where the contrast is replicated and
# thousands of genes reach FDR 0.05, and asks only whether our four libraries
# follow. keep restricts to the genes that carry enough counts here to rank.
agreement_for = function(ours_col, yas_col, axis, keep, pool,
                         rank_on = "ours") {
  d    = cmp[keep, , drop = FALSE]
  rank_vec = if (rank_on == "ours") d[[ours_col]] else d[[yas_col]]

  rows = list()
  gw = spearman_ci(d[[ours_col]], d[[yas_col]])

  gw_agree = sign(d[[ours_col]]) == sign(d[[yas_col]])
  gw_agree = gw_agree[is.finite(gw_agree)]

  rows[[1]] = data.frame(
    axis = axis, pool = pool, ranked_on = rank_on, set = "genome-wide",
    n = gw[["n"]], rho = gw[["rho"]], lo = gw[["lo"]], hi = gw[["hi"]],
    pct_same_sign = 100 * mean(gw_agree))

  ord = order(abs(rank_vec), decreasing = TRUE)
  for (n in top_ns) {
    idx = ord[seq_len(min(n, nrow(d)))]
    s = spearman_ci(d[[ours_col]][idx], d[[yas_col]][idx])
    agree = sign(d[[ours_col]][idx]) == sign(d[[yas_col]][idx])
    agree = agree[is.finite(agree)]
    rows[[length(rows) + 1]] = data.frame(
      axis = axis, pool = pool, ranked_on = rank_on, set = paste0("top ", n),
      n = length(idx), rho = s[["rho"]], lo = s[["lo"]], hi = s[["hi"]],
      pct_same_sign = 100 * mean(agree))
  }
  dplyr::bind_rows(rows)
}

axis_cols = list(
  "male, endpoint"   = c(ours = "ours_male",   yas = "male_end_logFC"),
  "female, endpoint" = c(ours = "ours_female", yas = "fem_end_logFC"),
  "male, trend"      = c(ours = "ours_male",   yas = "male_trend_logFC"),
  "female, trend"    = c(ours = "ours_female", yas = "fem_trend_logFC"))

gene_pools = list("all genes"     = rep(TRUE, nrow(cmp)),
                  "well measured" = cmp$well_measured)

grid = expand.grid(axis = names(axis_cols), pool = names(gene_pools),
                   rank_on = c("ours", "yasmin"), stringsAsFactors = FALSE)

agreement = dplyr::bind_rows(Map(function(ax, pl, ro) {
  cols = axis_cols[[ax]]
  agreement_for(cols[["ours"]], cols[["yas"]], ax, gene_pools[[pl]], pl, ro)
}, grid$axis, grid$pool, grid$rank_on))

write.csv(agreement, file = paste0(results_dir, "/bulk_agreement.csv"),
          row.names = FALSE)

datatable(agreement %>% mutate(across(where(is.numeric), ~signif(., 3))),
          rownames = FALSE, filter = "top",
          options = list(pageLength = 20))
```

The table now carries every combination of contrast, gene pool and ranking side.
`ranked_on = "ours"` is the original comparison. `ranked_on = "yasmin"` is the
reciprocal one: their contrast picks the genes and our four libraries only have
to follow. The reciprocal direction is the better powered of the two, since the
selection happens on the replicated side.

`r if (have_bulk) { say = function(ax) { a = agreement[agreement$axis == ax & agreement$pool == "well measured", ]; ao = a[a$ranked_on == "ours", ]; ay = a[a$ranked_on == "yasmin", ]; paste0(ax, ": genome-wide ", signif(ao$rho[ao$set == "genome-wide"], 3), ", our top ", top_ns[1], " ", signif(ao$rho[ao$set == paste0("top ", top_ns[1])], 3), ", their top ", top_ns[1], " ", signif(ay$rho[ay$set == paste0("top ", top_ns[1])], 3)) }; paste0("Spearman rho among well measured genes, for each contrast. ", paste(vapply(names(axis_cols), say, character(1)), collapse = ". "), ". The gap between the genome-wide value and the top-gene values says whether agreement concentrates in the genes each side nominates. The permutation below says whether any of it exceeds chance.") } else ""`

## Reading the numbers

**`rho` and `pct_same_sign`** are descriptive. They say how far the two datasets
agree, genome-wide and across the top genes, and the bootstrap interval on `rho`
says how precisely each is estimated. Neither is a test on its own, because the
top-gene set is picked using one of the two variables being correlated.

The genome-wide and top-gene values together say whether agreement concentrates
in the genes each side nominates, which is a question about specificity.

Two p-values appear. They test different things.

**`p_perm`**, in the permutation section below, is the one that says whether
agreement exceeds chance. It is the fraction of random relabelings of the same 28
animals that agree at least as well as Yasmin's real stage contrast does. It
assumes nothing about independence between genes. It also holds the
gene-selection effect fixed. The same selection rule is applied inside every
permutation.
Read the `ranked_on = "yasmin"` rows alongside the `ours` rows. Their contrasts
are replicated and ours are not. A list picked on their side therefore tests the
shared axis with the selection happening on the better powered side.

**`p_rand`**, in the composition section, asks whether marker genes carry more of
the agreement than a random gene set of the same size. It is about which genes
carry the signal, not about whether signal exists.

Both are one-sided, in the direction of agreement. Disagreement appears as a
p-value near 1 rather than as a small one.

## A label permutation null

The two numbers above cannot be read as a test on their own. The top-gene set is
picked by the absolute value of our own log fold change, which is one of the two
variables being correlated. Selecting large effects widens the spread of that
variable and lowers its relative measurement error, which raises the correlation
by itself. Selecting on a noisy estimate also picks up genes whose effect we
overstated. Those regress toward the mean in Yasmin, which lowers the
correlation. Both
act at once.

The permutation holds all of that fixed. Group labels are reassigned at random
among the 28 animals. The same contrast shape is refitted. The agreement
statistics are recomputed against our unchanged effect sizes and our unchanged
top-gene list. Whatever the selection does to the observed value, it does to
every null value in the same way. What is left is whether Yasmin's stage
contrast agrees with our genes more than an arbitrary contrast of the same shape
drawn from the same animals.

Labels are drawn from the whole series rather than shuffled within the contrast.
Shuffling within it has too few distinct splits to resolve anything: 2 against 6
gives 28 arrangements and 2 against 2 gives 6, so the smallest p-value reachable
on the female endpoint would be 0.17.

All four contrasts get a null, on the same terms. For the endpoints a random set
of animals is split into groups of the observed sizes. For the trends a random
set of the same size receives the observed score vector. The two trends carry
more animals than the endpoints do. They are the better powered half of this
table on both sex axes.

```{r, eval=have_bulk}
n_perm_labels = as.integer(params$n_perm_labels)
perm_cores    = max(1L, as.integer(params$cores))

# One set of refits serves every summary. The fit does not depend on the gene
# pool or on which side ranks the genes, so each permutation is scored for all
# four combinations rather than refitted four times.
#
# When Yasmin ranks the genes, the null re-selects inside each permutation from
# the permuted contrast. Reusing the real contrast's gene list would carry the
# real labels into the null.
perm_combos = expand.grid(pool = names(gene_pools),
                          rank_on = c("ours", "yasmin"),
                          stringsAsFactors = FALSE)

# The draws are generated here, in one stream, rather than inside the workers,
# so the result does not depend on how many cores are used. The stream is the
# same one the earlier serial version consumed — one sample() call per
# permutation, in order — so a cache written by that version stays valid.
all_runs = sample_meta$run

draw_runs = function(n_draw, n_perm, seed) {
  set.seed(seed)
  replicate(n_perm, sample(all_runs, n_draw), simplify = FALSE)
}

# scoring for one permutation: every pool and ranking side from the one fit
score_draw = function(lfc_all, ours_col, top_n) {
  dplyr::bind_rows(lapply(seq_len(nrow(perm_combos)), function(k) {
    keep = gene_pools[[perm_combos$pool[k]]]
    ov   = cmp[[ours_col]][keep]
    lv   = lfc_all[keep]
    idx  = if (perm_combos$rank_on[k] == "ours")
             order(abs(ov), decreasing = TRUE)[seq_len(min(top_n, length(ov)))]
           else
             order(abs(lv), decreasing = TRUE)[seq_len(min(top_n, length(lv)))]
    agree = sign(ov[idx]) == sign(lv[idx])
    data.frame(
      pool = perm_combos$pool[k], rank_on = perm_combos$rank_on[k],
      rho_gw   = suppressWarnings(cor(ov, lv, method = "spearman",
                                      use = "complete.obs")),
      rho_top  = suppressWarnings(cor(ov[idx], lv[idx], method = "spearman",
                                      use = "complete.obs")),
      pct_sign = 100 * mean(agree, na.rm = TRUE))
  }))
}

# The fits are independent, so they fan out across cores. mclapply forks, which
# is why the job script pins the BLAS threads to one: eight R processes each
# spawning threads would oversubscribe the node.
perm_null = function(draws, fit_draw, ours_col, top_n, cores = 1) {
  work = function(d) {
    tab = try(fit_draw(d), silent = TRUE)
    if (inherits(tab, "try-error")) return(NULL)
    score_draw(tab$perm_logFC[match(cmp$Gene, tab$Gene)], ours_col, top_n)
  }
  res = if (cores > 1) parallel::mclapply(draws, work, mc.cores = cores)
        else lapply(draws, work)
  failed = vapply(res, function(x) is.null(x) || inherits(x, "try-error"),
                  logical(1))
  if (any(failed)) message(sum(failed), " of ", length(draws),
                           " permutations failed to fit and are dropped")
  dplyr::bind_rows(res[!failed])
}

two_group_perm = function(n_a, n_b, ours_col, seed) {
  perm_null(draw_runs(n_a + n_b, n_perm_labels, seed),
            function(d) run_two_group(y, d[seq_len(n_a)],
                                      d[n_a + seq_len(n_b)], "perm"),
            ours_col, top_ns[1], perm_cores)
}

trend_perm = function(score_vec, ours_col, seed) {
  perm_null(draw_runs(length(score_vec), n_perm_labels, seed),
            function(d) run_trend(y, d, score_vec, "perm"),
            ours_col, top_ns[1], perm_cores)
}

# cached on a signature, since the permutation is the slow step in the document
perm_file = paste0(Robject_out_dir, "/label_permutation_null.Rds")
perm_sig  = list(n_perm  = n_perm_labels,
                 top_n   = top_ns[1],
                 n_genes = nrow(cmp),
                 n_well  = sum(cmp$well_measured),
                 min_pb  = min_pb,
                 combos  = perm_combos,
                 scores  = list(male = male_score, female = fem_score),
                 runs    = sort(sample_meta$run),
                 sizes   = c(length(runs_ud), length(runs_mm),
                             length(runs_mf), length(male_runs),
                             length(fem_runs)),
                 shapes  = c("endpoints", "trends"))

cached = if (file.exists(perm_file)) readRDS(perm_file) else NULL

if (!is.null(cached) && identical(cached$signature, perm_sig)) {
  perm_res = cached$perm_res
} else {
  perm_res = list(
    "male, endpoint"   = two_group_perm(length(runs_ud), length(runs_mm),
                                        "ours_male", 1),
    "female, endpoint" = two_group_perm(length(runs_ud), length(runs_mf),
                                        "ours_female", 2),
    "male, trend"      = trend_perm(male_score, "ours_male", 3),
    "female, trend"    = trend_perm(fem_score, "ours_female", 4))

  saveRDS(list(signature = perm_sig, perm_res = perm_res), file = perm_file)
}
```

```{r, eval=have_bulk}
# one-sided: the question is whether the observed statistic sits above the null
emp_p = function(null_vals, obs) {
  ok = is.finite(null_vals)
  (1 + sum(null_vals[ok] >= obs)) / (1 + sum(ok))
}

perm_tab = dplyr::bind_rows(lapply(names(perm_res), function(ax) {
  nd = perm_res[[ax]]
  dplyr::bind_rows(lapply(split(nd, list(nd$pool, nd$rank_on), drop = TRUE),
    function(g) {
      pl  = g$pool[1]
      ro  = g$rank_on[1]
      obs = agreement[agreement$axis == ax & agreement$pool == pl &
                        agreement$ranked_on == ro, ]
      o   = c(obs$rho[obs$set == "genome-wide"],
              obs$rho[obs$set == paste0("top ", top_ns[1])],
              obs$pct_same_sign[obs$set == paste0("top ", top_ns[1])])
      nl  = list(g$rho_gw, g$rho_top, g$pct_sign)

      out = data.frame(
        axis        = ax,
        pool        = pl,
        ranked_on   = ro,
        statistic   = c("rho, genome-wide", paste0("rho, top ", top_ns[1]),
                        paste0("% same sign, top ", top_ns[1])),
        observed    = o,
        null_median = vapply(nl, median, numeric(1), na.rm = TRUE),
        null_95th   = vapply(nl,
                             function(v) unname(quantile(v, .95, na.rm = TRUE)),
                             numeric(1)),
        p_perm      = mapply(emp_p, nl, o),
        n_perm      = vapply(nl, function(v) sum(is.finite(v)), integer(1)))

      # the genome-wide row does not depend on which side ranks the genes
      if (ro == "yasmin") out = out[out$statistic != "rho, genome-wide", ]
      out
    }))
}))

write.csv(perm_tab, file = paste0(results_dir, "/bulk_permutation_null.csv"),
          row.names = FALSE)

datatable(perm_tab %>% mutate(across(where(is.numeric), ~signif(., 3))),
          rownames = FALSE, filter = "top", options = list(pageLength = 20))
```

`r if (have_bulk) paste0("The smallest p-value reachable with ", n_perm_labels, " permutations is ", signif(1 / (1 + n_perm_labels), 2), ". A statistic at that value is reported as a floor rather than as its true p-value.") else ""`

```{r, eval=have_bulk, fig.width=15, fig.height=7}
# the well measured pool only, since the unfiltered ranking is the one under
# suspicion
perm_long = dplyr::bind_rows(lapply(names(perm_res), function(ax) {
  nd = perm_res[[ax]][perm_res[[ax]]$pool == "well measured", ]
  data.frame(axis = ax, ranked_on = nd$rank_on, rho_top = nd$rho_top)
}))

obs_long = dplyr::bind_rows(lapply(names(perm_res), function(ax)
  dplyr::bind_rows(lapply(c("ours", "yasmin"), function(ro) {
    obs = agreement[agreement$axis == ax & agreement$pool == "well measured" &
                      agreement$ranked_on == ro, ]
    data.frame(axis = ax, ranked_on = ro,
               observed = obs$rho[obs$set == paste0("top ", top_ns[1])])
  }))))

p_perm = ggplot(perm_long, aes(x = rho_top)) +
  geom_histogram(bins = 40, fill = "grey75", colour = NA) +
  geom_vline(data = obs_long, aes(xintercept = observed),
             colour = "#c1272d", linewidth = .7) +
  facet_grid(ranked_on ~ axis, scales = "free_y") +
  labs(x = paste0("Spearman rho across the top ", top_ns[1], " genes"),
       y = "permutations",
       title = "Observed agreement against random contrasts of the same shape",
       subtitle = "well measured genes; rows give the side that ranked the genes") +
  theme_bw()

p_perm

ggsave(p_perm, file = paste0(results_dir, "/bulk_permutation_null.png"),
       width = 15, height = 7, dpi = 300)
```

## The trend figure

The scatters take our top 100 genes on each sex's axis and plot our effect
against Yasmin's trend coefficient, with points colored by whether the two
datasets put the gene in the same direction. Each panel carries its Spearman
correlation and the permutation p-value for that correlation. The paired plot
underneath takes the strongest concordant genes and shows both effect sizes side
by side, so a gene that rises in both series reads as two points on the same
side of zero. The agreement and permutation tables above also report the
selection made on Yasmin's side, which this figure leaves out.

```{r, eval=have_bulk, fig.width=12, fig.height=10}
trend_axes  = c("male, trend", "female, trend")
# our effect picks the genes. The Yasmin-ranked selection is in the tables.
rank_sides  = "ours"
top_set_lab = paste0("top ", top_ns[1])

panel_grid = expand.grid(axis = trend_axes, ranked_on = rank_sides,
                         stringsAsFactors = FALSE)

trend_top = dplyr::bind_rows(Map(function(ax, ro) {
  cols = axis_cols[[ax]]
  d = cmp[cmp$well_measured, ] %>%
    dplyr::select(Gene, label, ours = dplyr::all_of(cols[["ours"]]),
                  yasmin = dplyr::all_of(cols[["yas"]])) %>%
    dplyr::filter(is.finite(ours), is.finite(yasmin))
  rank_vec = if (ro == "ours") d$ours else d$yasmin
  d = d[order(abs(rank_vec), decreasing = TRUE)[
          seq_len(min(top_ns[1], nrow(d)))], , drop = FALSE]
  d$axis      = ax
  d$ranked_on = ro
  d$agree     = sign(d$ours) == sign(d$yasmin)
  d
}, panel_grid$axis, panel_grid$ranked_on))

trend_top$axis      = factor(trend_top$axis, levels = trend_axes)
trend_top$ranked_on = factor(trend_top$ranked_on, levels = rank_sides,
                             labels = "genes picked by our effect")

# rho from the agreement table, p from the permutation, so the panel carries the
# same numbers the tables report rather than a second calculation of them
rho_tab = agreement %>%
  dplyr::filter(pool == "well measured", set == top_set_lab) %>%
  dplyr::select(axis, ranked_on, rho)

p_tab = perm_tab %>%
  dplyr::filter(pool == "well measured",
                statistic == paste0("rho, ", top_set_lab)) %>%
  dplyr::select(axis, ranked_on, p_perm)

panel_ann = trend_top %>%
  dplyr::group_by(axis, ranked_on) %>%
  dplyr::summarise(pct = 100 * mean(agree), .groups = "drop") %>%
  dplyr::mutate(rank_key = ifelse(grepl("our effect", ranked_on),
                                  "ours", "yasmin")) %>%
  dplyr::left_join(rho_tab, by = c("axis" = "axis", "rank_key" = "ranked_on")) %>%
  dplyr::left_join(p_tab,   by = c("axis" = "axis", "rank_key" = "ranked_on")) %>%
  dplyr::mutate(txt = paste0("rho = ", signif(rho, 2),
                             "   p = ", signif(p_perm, 2), "\n",
                             round(pct), "% same direction"))

# the joins above coerce the factor columns to character, which would let ggplot
# order the facets alphabetically instead of by the levels set on trend_top
panel_ann$axis      = factor(panel_ann$axis, levels = levels(trend_top$axis))
panel_ann$ranked_on = factor(panel_ann$ranked_on,
                             levels = levels(trend_top$ranked_on))

# Okabe-Ito blue and vermillion. A red-green pair is the one contrast a red-green
# colorblind reader cannot make, and it is the default reach for agree/disagree.
agree_cols = c(`TRUE` = "#0072B2", `FALSE` = "#D55E00")

p_top = ggplot(trend_top, aes(x = ours, y = yasmin, colour = agree)) +
  geom_hline(yintercept = 0, colour = "grey70") +
  geom_vline(xintercept = 0, colour = "grey70") +
  geom_point(size = 1.6, alpha = .85) +
  geom_text_repel(data = trend_top %>%
                    dplyr::group_by(axis, ranked_on) %>%
                    dplyr::slice_max(abs(ours), n = 10, with_ties = FALSE),
                  aes(label = label), size = 2.4, max.overlaps = 22,
                  show.legend = FALSE) +
  geom_label(data = panel_ann, aes(x = -Inf, y = Inf, label = txt),
             hjust = -0.05, vjust = 1.1, inherit.aes = FALSE, size = 3.1,
             label.size = 0, fill = "white", alpha = .75, lineheight = 1.05) +
  scale_colour_manual(values = agree_cols, name = NULL,
                      labels = c(`TRUE` = "same direction",
                                 `FALSE` = "opposite")) +
  facet_wrap(~ axis, scales = "free", nrow = 1) +
  labs(x = "our pseudobulk log2FC, larva to transformer",
       y = "Yasmin trend coefficient",
       title = paste0("Our top ", top_ns[1],
                      " genes against the Yasmin trend"),
       subtitle = "p is the label permutation p-value for the correlation shown") +
  theme_bw() + theme(legend.position = "top")

# the strongest concordant genes, both effect sizes side by side, from the
# list picked by our effect
paired = trend_top %>%
  dplyr::filter(agree, grepl("our effect", ranked_on)) %>%
  dplyr::group_by(axis) %>%
  dplyr::slice_max(abs(ours), n = 20, with_ties = FALSE) %>%
  dplyr::ungroup() %>%
  dplyr::mutate(direction = ifelse(ours > 0, "up in transformers",
                                   "down in transformers"))

paired_long = paired %>%
  tidyr::pivot_longer(c(ours, yasmin), names_to = "dataset",
                      values_to = "effect") %>%
  dplyr::mutate(dataset = dplyr::recode(dataset, ours = "ours",
                                        yasmin = "Yasmin"))

p_paired = ggplot(paired_long,
                  aes(x = effect, y = reorder(label, effect))) +
  geom_vline(xintercept = 0, colour = "grey70") +
  geom_line(aes(group = label), colour = "grey65", linewidth = .4) +
  geom_point(aes(shape = dataset, fill = direction), size = 2.2,
             colour = "grey25", stroke = .3) +
  scale_shape_manual(values = c(ours = 21, Yasmin = 24), name = NULL) +
  scale_fill_manual(values = c(`up in transformers` = "#0072B2",
                               `down in transformers` = "#E69F00"),
                    name = NULL) +
  facet_wrap(~axis, scales = "free") +
  labs(x = "effect size in each dataset", y = NULL,
       title = "Strongest concordant genes, both datasets") +
  theme_bw() + theme(legend.position = "top",
                     axis.text.y = element_text(size = 7))

p_trend = p_top / p_paired + patchwork::plot_layout(heights = c(1, 1.2)) +
  patchwork::plot_annotation(tag_levels = "A")

p_trend

ggsave(p_trend, file = paste0(results_dir, "/bulk_trend_agreement.png"),
       width = 12, height = 10, dpi = 300)
```

`r if (have_bulk) paste0("Panel A colors each of our top ", top_ns[1], " genes by whether the two datasets agree on direction. The correlation and its permutation p-value are printed in each panel and are the same values the agreement and permutation tables carry. Panel B takes the concordant genes with the largest effects and plots both estimates, so agreement reads as two markers on the same side of zero.") else ""`

## Genome-wide scatter

```{r, eval=have_bulk, fig.width=11, fig.height=11, dev="png", dpi=600}
# One point per measurable gene in each of four panels, so this one stays
# raster. The rest of the figures in this document are vector.
scatter_for = function(ours_col, yas_col, title, n_label = 15) {
  d = cmp %>%
    dplyr::select(Gene, label, ours = dplyr::all_of(ours_col),
                  yasmin = dplyr::all_of(yas_col)) %>%
    dplyr::filter(is.finite(ours), is.finite(yasmin))
  lab = d %>% dplyr::slice_max(abs(ours), n = n_label, with_ties = FALSE)

  ggplot(d, aes(x = ours, y = yasmin)) +
    geom_hline(yintercept = 0, colour = "grey80") +
    geom_vline(xintercept = 0, colour = "grey80") +
    geom_point(alpha = .2, size = .5) +
    geom_point(data = lab, colour = "#c1272d", size = 1.2) +
    geom_text_repel(data = lab, aes(label = label), size = 2.5,
                    max.overlaps = 20) +
    geom_smooth(method = "lm", se = FALSE, linewidth = .5, colour = "#004488") +
    labs(x = "our pseudobulk log2FC, larva to transformer",
         y = "Yasmin log2FC", title = title) +
    theme_bw()
}

p_male  = scatter_for("ours_male",   "male_end_logFC",   "Male endpoint")
p_fem   = scatter_for("ours_female", "fem_end_logFC",    "Female endpoint")
p_mtr   = scatter_for("ours_male",   "male_trend_logFC", "Male trend")
p_ftr   = scatter_for("ours_female", "fem_trend_logFC",  "Female trend")

p_scatter = (p_male + p_fem) / (p_mtr + p_ftr)
p_scatter

ggsave(p_scatter, file = paste0(results_dir, "/bulk_agreement_scatter.png"),
       width = 11, height = 11, dpi = 300)
```

Every gene is plotted here, not only the top hundred, with the fifteen strongest
labeled. The line is an ordinary least squares fit shown for orientation.

# Top gene lists

The hundred most up- and most down-regulated genes from each side, taken against
the genes measurable in both datasets. Our lists are the same for both contrasts
on a sex, since our effect does not depend on which Yasmin contrast it is
compared with. Yasmin's are built per contrast. The GO section below tests these
lists, and the overlap below intersects them.

```{r, eval=have_bulk}
# The background is the genes measurable on both sides, which is the set the
# lists were drawn from. Up and down are taken separately. A term enriched among
# rising genes says something different from one enriched among falling genes.
go_pool = cmp[cmp$well_measured, , drop = FALSE]

top_dir = function(v, direction, n = top_ns[1]) {
  ok = is.finite(v)
  idx = if (direction == "up") which(ok & v > 0) else which(ok & v < 0)
  idx[order(abs(v[idx]), decreasing = TRUE)][seq_len(min(n, length(idx)))]
}

# Each set carries the source, sex and direction it came from, so terms can be
# matched across sources later without parsing the label back apart.
set_defs = c(
  list(list(label = "ours, male",   col = "ours_male",
            source = "ours", sex = "male",   contrast = NA_character_),
       list(label = "ours, female", col = "ours_female",
            source = "ours", sex = "female", contrast = NA_character_)),
  lapply(names(axis_cols), function(ax) {
    parts = trimws(strsplit(ax, ",")[[1]])
    list(label = paste0("Yasmin, ", ax), col = axis_cols[[ax]][["yas"]],
         source = "Yasmin", sex = parts[1], contrast = parts[2])
  }))

set_rows = lapply(set_defs, function(d)
  lapply(c("up", "down"), function(dir) {
    idx = top_dir(go_pool[[d$col]], dir)
    list(genes = data.frame(Gene = go_pool$Gene[idx],
                            set  = paste0(d$label, ", ", dir),
                            stringsAsFactors = FALSE),
         key   = data.frame(set = paste0(d$label, ", ", dir),
                            source = d$source, sex = d$sex,
                            contrast = d$contrast, direction = dir,
                            stringsAsFactors = FALSE))
  }))

set_rows = unlist(set_rows, recursive = FALSE)
go_sets  = dplyr::bind_rows(lapply(set_rows, `[[`, "genes"))
set_key  = dplyr::bind_rows(lapply(set_rows, `[[`, "key"))

set_sizes = set_key %>%
  dplyr::left_join(go_sets %>% dplyr::count(set, name = "n_genes"), by = "set")

datatable(set_sizes, rownames = FALSE,
          options = list(dom = "t", pageLength = 20))
```
## Overlap between the lists

A gene in our top hundred and in Yasmin's top hundred, on the same sex and in the
same direction, is a stricter kind of agreement than the sign agreement reported
above. Sign agreement asks only whether Yasmin moves our genes the same way,
however weakly. This asks whether they are among her strongest too.

The expected overlap is the product of the two list sizes over the pool they were
drawn from, and the p-value is hypergeometric on that pool. Both treat the lists
as independent draws, which they are not: the two datasets share the biology
under test. Read the p-value as a reference point rather than as a test of
reproducibility.

```{r, eval=have_bulk}
all_lists = split(go_sets$Gene, go_sets$set)
pool_n    = nrow(go_pool)

our_keys = set_key %>% dplyr::filter(source == "ours")
yas_keys = set_key %>% dplyr::filter(source == "Yasmin")

overlap_tab = dplyr::bind_rows(lapply(our_keys$set, function(a)
  dplyr::bind_rows(lapply(yas_keys$set, function(b) {
    ga = all_lists[[a]]
    gb = all_lists[[b]]
    k  = length(intersect(ga, gb))
    data.frame(
      ours = a, yasmin = b, n_ours = length(ga), n_yasmin = length(gb),
      overlap = k,
      expected = length(ga) * length(gb) / pool_n,
      # P(X >= k) under the hypergeometric, so lower.tail on k - 1
      p_hyper = stats::phyper(k - 1, length(gb), pool_n - length(gb),
                              length(ga), lower.tail = FALSE),
      stringsAsFactors = FALSE)
  })))) %>%
  dplyr::left_join(our_keys %>% dplyr::select(ours = set, sex_ours = sex,
                                              dir_ours = direction),
                   by = "ours") %>%
  dplyr::left_join(yas_keys %>% dplyr::select(yasmin = set, sex_yas = sex,
                                              dir_yas = direction,
                                              contrast_yas = contrast),
                   by = "yasmin") %>%
  dplyr::mutate(matched = sex_ours == sex_yas & dir_ours == dir_yas,
                fold = overlap / expected,
                fdr = p.adjust(p_hyper, "BH")) %>%
  dplyr::arrange(dplyr::desc(matched), p_hyper)

write.csv(overlap_tab,
          file = paste0(results_dir, "/top_gene_list_overlap.csv"),
          row.names = FALSE)

# the genes themselves, for the pairs the section is about
overlap_genes = dplyr::bind_rows(lapply(which(overlap_tab$matched), function(i) {
  g = intersect(all_lists[[overlap_tab$ours[i]]],
                all_lists[[overlap_tab$yasmin[i]]])
  if (length(g) == 0) return(NULL)
  # our effect does not depend on the contrast, only on the sex
  ours_col = paste0("ours_", overlap_tab$sex_ours[i])
  idx      = match(g, cmp$Gene)
  data.frame(ours = overlap_tab$ours[i], yasmin = overlap_tab$yasmin[i],
             Gene = g,
             label = cmp$label[idx],
             ours_effect = cmp[[ours_col]][idx],
             stringsAsFactors = FALSE)
}))

write.csv(overlap_genes,
          file = paste0(results_dir, "/top_gene_list_overlap_genes.csv"),
          row.names = FALSE)

datatable(overlap_tab %>%
            dplyr::select(ours, yasmin, matched, n_ours, n_yasmin, overlap,
                          expected, fold, p_hyper, fdr) %>%
            mutate(across(where(is.numeric), ~signif(., 3))),
          rownames = FALSE, filter = "top", options = list(pageLength = 16))
```

`r if (!have_bulk) "" else { m = overlap_tab %>% dplyr::filter(matched); paste0("Of the ", nrow(overlap_tab), " pairings, ", nrow(m), " match on sex and direction. Their overlaps run from ", min(m$overlap), " to ", max(m$overlap), " genes against an expectation of about ", signif(mean(m$expected), 2), ", and ", sum(m$fdr < 0.05), " of them reach FDR < 0.05. The intersecting genes are in top_gene_list_overlap_genes.csv.") }`

```{r, eval=have_bulk, fig.width=12, fig.height=8}
# Hand-built rather than UpSetR or ComplexUpset, so the document takes no new
# dependency and the figure comes out as vector like the rest. Set sizes go in
# the y axis labels instead of a third aligned panel.
upset_contrast = "trend"   # the Yasmin contrast the figure shows
n_int          = 20        # intersections drawn, largest first

fig_sets = set_key %>%
  dplyr::filter(source == "ours" | contrast == upset_contrast) %>%
  dplyr::arrange(source, sex, direction)

set_lists = split(go_sets$Gene, go_sets$set)[fig_sets$set]
all_genes = sort(unique(unlist(set_lists)))

# one column per set, in fig_sets order, so the bit strings below line up with it
mem = vapply(set_lists, function(g) all_genes %in% g,
             logical(length(all_genes)))

int_all = data.frame(pattern = apply(mem, 1, function(r)
                       paste(as.integer(r), collapse = "")),
                     stringsAsFactors = FALSE) %>%
  dplyr::count(pattern, name = "size") %>%
  dplyr::arrange(dplyr::desc(size))

# only the largest are drawn, so the drawn columns need not cover every gene
int_sizes = int_all %>%
  dplyr::slice_head(n = n_int) %>%
  dplyr::mutate(x = dplyr::row_number())

bits = strsplit(int_sizes$pattern, "")
mem_long = dplyr::bind_rows(lapply(seq_len(nrow(int_sizes)), function(i)
  data.frame(x = int_sizes$x[i], set = fig_sets$set,
             member = bits[[i]] == "1", stringsAsFactors = FALSE)))

set_lab = setNames(paste0(fig_sets$set, "  (",
                          vapply(set_lists, length, integer(1)), ")"),
                   fig_sets$set)

x_scale = scale_x_continuous(breaks = int_sizes$x,
                             expand = ggplot2::expansion(add = .6))

p_ubar = ggplot(int_sizes, aes(x = x, y = size)) +
  geom_col(fill = "#0072B2", width = .7) +
  geom_text(aes(label = size), vjust = -0.35, size = 2.8) +
  x_scale +
  scale_y_continuous(expand = ggplot2::expansion(mult = c(0, .14))) +
  labs(x = NULL, y = "genes in the intersection",
       title = paste0("Top ", top_ns[1],
                      " gene lists, ours and the Yasmin ", upset_contrast),
       subtitle = paste0("the ", n_int,
                         " largest intersections; set sizes in brackets")) +
  theme_bw(base_size = 10) +
  theme(axis.text.x = element_blank(), axis.ticks.x = element_blank(),
        panel.grid.major.x = element_blank())

p_umat = ggplot(mem_long, aes(x = x, y = set)) +
  geom_point(colour = "grey88", size = 3) +
  geom_line(data = dplyr::filter(mem_long, member), aes(group = x),
            colour = "#0072B2", linewidth = .9) +
  geom_point(data = dplyr::filter(mem_long, member),
             colour = "#0072B2", size = 3) +
  x_scale +
  scale_y_discrete(limits = rev(fig_sets$set),
                   labels = function(z) unname(set_lab[z])) +
  labs(x = NULL, y = NULL) +
  theme_bw(base_size = 10) +
  theme(axis.text.x = element_blank(), axis.ticks.x = element_blank(),
        panel.grid.major.x = element_blank())

p_upset = p_ubar / p_umat + patchwork::plot_layout(heights = c(1, 1.15))

p_upset

ggsave(p_upset, file = paste0(results_dir, "/top_gene_list_upset.png"),
       width = 12, height = 8, dpi = 300)
```

`r if (!have_bulk) "" else paste0("Each column is the set of genes sharing one membership pattern, so the columns are disjoint rather than nested. ", format(length(all_genes), big.mark = ","), " genes appear in at least one of the ", nrow(fig_sets), " lists, spread over ", nrow(int_all), " patterns; the ", nrow(int_sizes), " drawn here hold ", format(sum(int_sizes$size), big.mark = ","), " of them. A column with one filled dot is genes unique to that list. The columns worth reading are the two-dot ones joining a list of ours to Yasmin's on the same sex and in the same direction.")`


# Gene ontology

The lists built above are tested here for GO enrichment against the genes
measurable in both datasets.

Terms are kept when they carry more than 10 and fewer than 200 genes. Tiny terms
are noise and very large ones restate the whole ontology.

```{r}
gaf_file      = file.path(params$genome_dir,
                          "GCF_010993605.1_kPetMar1.pri_gene_ontology.gaf")
go_names_file = file.path(params$genome_dir, "GOs_names.txt")
have_go       = have_bulk && file.exists(gaf_file) && file.exists(go_names_file)
```

`r if (have_bulk && !have_go) paste0("**GO enrichment is skipped.** The GO files were not found under ", params$genome_dir, ".") else ""`

```{r, eval=have_go}
gaf = read.delim(gaf_file, header = FALSE, comment = "!") %>%
  dplyr::filter(!grepl("NOT", V4)) %>%
  dplyr::select(V3, V5) %>%
  dplyr::rename(GOID = "V5", Gene = "V3") %>%
  dplyr::distinct()

keep_go = gaf %>%
  dplyr::count(GOID) %>%
  dplyr::filter(n > 10, n < 200) %>%
  dplyr::pull(GOID)
gaf = gaf %>% dplyr::filter(GOID %in% keep_go)

go_names = read.delim(go_names_file, header = FALSE)
colnames(go_names) = c("GOID", "GOName")
```


```{r, eval=have_go}
go_res = run_go_enrichment(
  gaf              = gaf,
  gene_description = gene_description,
  go_names         = go_names,
  test_genes       = go_sets,
  background_n     = nrow(go_pool),
  group2_label     = "set",
  ngroup2_label    = "nSetGenes",
  overlaps_file    = paste0(results_dir, "/top_gene_GO_overlaps.csv"),
  pvals_file       = paste0(results_dir, "/top_gene_GO_pvalues.csv"),
  pval_filter      = params$go_pval,
  require_shared   = TRUE)

# Mark the terms that come up on both sides of the same comparison. A partner is
# a set from the other source with the same sex and the same direction, so
# "Yasmin, male, trend, up" pairs with "ours, male, up". Our sets are not split
# by contrast, so one of ours can partner both Yasmin contrasts on its sex.
go_res = go_res %>% dplyr::left_join(set_key, by = "set")

partners = go_res %>%
  dplyr::select(GOID, set, source, sex, direction) %>%
  dplyr::inner_join(go_res %>% dplyr::select(GOID, set, source, sex, direction),
                    by = c("GOID", "sex", "direction"),
                    suffix = c("", "_partner"),
                    relationship = "many-to-many") %>%
  dplyr::filter(source != source_partner) %>%
  dplyr::group_by(GOID, set) %>%
  dplyr::summarise(shared_with = paste(sort(unique(set_partner)),
                                       collapse = "; "), .groups = "drop")

go_res = go_res %>%
  dplyr::left_join(partners, by = c("GOID", "set")) %>%
  dplyr::mutate(shared = !is.na(shared_with),
                shared_with = ifelse(is.na(shared_with), "", shared_with)) %>%
  dplyr::relocate(shared, shared_with, .after = set)

write.csv(go_res, file = paste0(results_dir, "/top_gene_GO_pvalues.csv"),
          row.names = FALSE)

datatable(go_res %>% mutate(across(where(is.numeric), ~signif(., 3))),
          filter = "top", rownames = FALSE)
```

`r if (have_go && exists("go_res")) paste0(format(nrow(go_res), big.mark = ","), " term-by-set results pass p < ", params$go_pval, " across ", dplyr::n_distinct(go_sets$set), " gene sets. The full table is in top_gene_GO_pvalues.csv and the genes behind each term in top_gene_GO_overlaps.csv. Read the FDR column rather than Pval where a term is named individually. The FDR is computed within each set.") else ""`

## Terms shared with the Yasmin trend

A term counts as shared when it is enriched in our top genes and in Yasmin's
trend genes, on the same sex and in the same direction. The endpoint contrasts
are left out here, so a term that pairs only with Yasmin's endpoint does not
appear. Our gene sets are not split by contrast, so one of our terms can pair
with the trend and the endpoint both; the `shared_with` column of the table
above names which.

```{r, eval=have_go}
go_shared_trend = go_res %>%
  dplyr::filter(source == "ours") %>%
  dplyr::select(GOID, GOName, sex, direction, n_genes_ours = nSharedGenes,
                p_ours = Pval, fdr_ours = FDR) %>%
  dplyr::inner_join(
    go_res %>%
      dplyr::filter(source == "Yasmin", contrast == "trend") %>%
      dplyr::select(GOID, sex, direction, n_genes_yasmin = nSharedGenes,
                    p_yasmin = Pval, fdr_yasmin = FDR),
    by = c("GOID", "sex", "direction")) %>%
  dplyr::arrange(sex, direction, pmax(p_ours, p_yasmin))

write.csv(go_shared_trend,
          file = paste0(results_dir, "/shared_trend_GO_terms.csv"),
          row.names = FALSE)

datatable(go_shared_trend %>% mutate(across(where(is.numeric), ~signif(., 3))),
          rownames = FALSE, filter = "top", options = list(pageLength = 15))
```

`r if (!have_go) "" else if (nrow(go_shared_trend) == 0) "**No term is enriched on both sides of the same sex in the same direction.**" else { cnt = go_shared_trend %>% dplyr::count(sex, direction); paste0(nrow(go_shared_trend), " term pairs are enriched on both sides, across ", dplyr::n_distinct(go_shared_trend$GOID), " distinct terms: ", paste0(cnt$n, " ", cnt$sex, " ", cnt$direction, collapse = "; "), ".") }`

```{r, eval=have_go && nrow(go_shared_trend) > 0, fig.width=12, fig.height=9}
# The weaker of the two p-values orders the terms, so a term high on the axis is
# one that both sides support rather than one side strongly and the other barely.
n_show = 12

shared_plot = go_shared_trend %>%
  dplyr::group_by(sex, direction) %>%
  dplyr::slice_min(pmax(p_ours, p_yasmin), n = n_show, with_ties = FALSE) %>%
  dplyr::ungroup() %>%
  dplyr::mutate(panel = paste0(sex, ", ", direction),
                ykey  = paste(panel, GOID, sep = "|"))

# one ordering per panel, so the y axis is sorted inside each facet rather than
# across all of them
lev = shared_plot %>%
  dplyr::arrange(panel, dplyr::desc(pmax(p_ours, p_yasmin))) %>%
  dplyr::pull(ykey)
shared_plot$ykey = factor(shared_plot$ykey, levels = lev)

shared_long = shared_plot %>%
  tidyr::pivot_longer(c(p_ours, p_yasmin), names_to = "side",
                      values_to = "p") %>%
  dplyr::mutate(side = dplyr::recode(side, p_ours = "ours",
                                     p_yasmin = "Yasmin trend"),
                neglog = -log10(p))

p_go_shared = ggplot(shared_long, aes(x = neglog, y = ykey)) +
  geom_line(aes(group = ykey), colour = "grey65", linewidth = .5) +
  geom_point(aes(colour = side), size = 2.6) +
  scale_colour_manual(values = c(ours = "#0072B2",
                                 `Yasmin trend` = "#E69F00"),
                      name = NULL) +
  scale_y_discrete(labels = function(x)
    shared_plot$GOName[match(x, as.character(shared_plot$ykey))]) +
  facet_wrap(~ panel, scales = "free_y", ncol = 2) +
  labs(x = expression(-log[10](italic(p))), y = NULL,
       title = "GO terms enriched in both datasets, same sex and direction",
       subtitle = paste0("up to ", n_show,
                         " terms per panel, ordered by the weaker side")) +
  theme_bw(base_size = 10) +
  theme(legend.position = "top",
        axis.text.y = element_text(size = 7),
        panel.grid.major.y = element_line(colour = "grey92"))

p_go_shared

ggsave(p_go_shared, file = paste0(results_dir, "/shared_trend_GO_terms.png"),
       width = 12, height = 9, dpi = 300)
```

`r if (!have_go || nrow(go_shared_trend) == 0) "" else paste0("Each term carries a point from each dataset. A term whose two points sit close together is supported about equally on both sides. One with a long connector is carried mostly by the dataset further to the right. Every term shown passes p < ", params$go_pval, " on both sides, which is what being shared means here.")`


# Sex specificity

A germ cell signature scored across the Yasmin samples asks whether their series
carries the germ cell states we describe. A signature built from a male germ
cell state should score higher in their males than in their females, and a
female one should do the reverse. Where it does not, the signature is reading
germ cell content rather than germ cell identity.

Two marker sources are scored side by side, because they measure different
things. *Atlas markers*, from `13_atlas_and_stage_correspondence.Rmd`, compare
each cluster against every other cluster in the atlas. Most of those clusters
are somatic, so a gene expressed across germ cells and absent from soma marks
every germ cluster at once. *Pseudotime markers*, from
`17_germ_pseudotime_tscan.Rmd`, compare node against node within the germ
compartment alone, where the shared germ cell program is already divided out.

```{r}
marker_sources = list(
  pseudotime = paste0(params$project_dir, "/results/", params$results_date,
                      "_pseudotime_tscan_",
                      as.character(params$germ_marker_lineage),
                      "/subcluster_marker_genes.csv"),
  atlas      = paste0(atlas_dir, "/cluster_marker_genes.csv"))

src_found       = vapply(marker_sources, file.exists, logical(1))
have_cl_markers = have_bulk && any(src_found)
```

`r if (have_bulk && !have_cl_markers) paste0("**This section is skipped.** Neither marker file was found: ", paste(unlist(marker_sources), collapse = "; "), ".") else ""`

`r if (have_cl_markers && !all(src_found)) paste0("**Only the ", paste(names(src_found)[src_found], collapse = " and "), " source was found.** Missing: ", paste(unlist(marker_sources)[!src_found], collapse = "; "), ".") else ""`

```{r, eval=have_cl_markers}
n_sig = as.integer(params$n_signature_genes)

read_markers = function(src) {
  tab = read.csv(marker_sources[[src]], stringsAsFactors = FALSE)
  if (src == "pseudotime") {
    stopifnot(all(c("Gene", "node", "meanLogFC") %in% colnames(tab)))
    tab$group = as.character(tab$node)
  } else {
    stopifnot(all(c("Gene", "cluster", "meanLogFC") %in% colnames(tab)))
    germ_clusters = trimws(strsplit(as.character(params$germ_clusters),
                                    ",")[[1]])
    missing_cl = setdiff(germ_clusters, unique(as.character(tab$cluster)))
    if (length(missing_cl) > 0)
      stop("germ_clusters not present in ", marker_sources[[src]], ": ",
           paste(missing_cl, collapse = ", "), ". Found: ",
           paste(sort(unique(as.character(tab$cluster))), collapse = ", "))
    tab = tab[as.character(tab$cluster) %in% germ_clusters, , drop = FALSE]
    tab$group = as.character(tab$cluster)
  }
  tab$source = src
  tab %>% dplyr::select(source, group, Gene, meanLogFC)
}

germ_markers = dplyr::bind_rows(lapply(names(src_found)[src_found],
                                       read_markers))

# the signature is the strongest markers that are also measurable in the bulk
sig = germ_markers %>%
  dplyr::filter(Gene %in% rownames(y)) %>%
  dplyr::group_by(source, group) %>%
  dplyr::arrange(dplyr::desc(meanLogFC), .by_group = TRUE) %>%
  dplyr::slice_head(n = n_sig) %>%
  dplyr::ungroup()
```

`r if (have_cl_markers) paste0("Signatures are the top ", n_sig, " markers per group by mean log fold change, kept only where the gene is measurable in the Yasmin series.") else ""`

```{r, eval=have_cl_markers}
# Each gene is standardized across the 28 samples first, so a signature score
# averages comparable quantities rather than raw abundances.
ylog    = cpm(y, log = TRUE, prior.count = 1)
gene_sd = apply(ylog, 1, sd)
zmat    = t(scale(t(ylog[gene_sd > 0, , drop = FALSE])))

score_mats = lapply(split(sig, sig$source), function(d) {
  sl = split(d$Gene, d$group)
  m  = vapply(sl, function(gs) {
    gs = intersect(gs, rownames(zmat))
    colMeans(zmat[gs, , drop = FALSE], na.rm = TRUE)
  }, numeric(ncol(zmat)))
  rownames(m) = colnames(zmat)
  m
})
```

```{r, eval=have_cl_markers}
sex_tab = dplyr::bind_rows(lapply(names(score_mats), function(src)
  dplyr::bind_rows(lapply(colnames(score_mats[[src]]), function(gp)
    data.frame(source = src, group = gp,
               mid_males   = mean(score_mats[[src]][runs_mm, gp]),
               mid_females = mean(score_mats[[src]][runs_mf, gp]),
               difference  = mean(score_mats[[src]][runs_mm, gp]) -
                             mean(score_mats[[src]][runs_mf, gp]))))))

write.csv(sex_tab, file = paste0(results_dir, "/germ_signature_sex.csv"),
          row.names = FALSE)

datatable(sex_tab %>% mutate(across(where(is.numeric), ~signif(., 3))),
          rownames = FALSE, filter = "top", options = list(pageLength = 20))
```

`r if (have_cl_markers) "Mid males against mid females, mean score in each. A signature that separates the sexes is measuring a germ cell state. One that does not is measuring how much germ cell content the gonad has, which every germ signature shares." else ""`

```{r, eval=have_cl_markers, fig.width=11, fig.height=7}
# One ordering per marker source, since the two sources name different groups
# and a global sort would interleave them.
sex_plot = sex_tab %>%
  dplyr::arrange(source, difference) %>%
  dplyr::mutate(ykey = factor(paste(source, group, sep = "|"),
                              levels = paste(source, group, sep = "|")))

sex_long = sex_plot %>%
  tidyr::pivot_longer(c(mid_males, mid_females), names_to = "sex",
                      values_to = "score") %>%
  dplyr::mutate(sex = dplyr::recode(sex, mid_males = "mid males",
                                    mid_females = "mid females"))

p_sex = ggplot(sex_long, aes(x = score, y = ykey)) +
  geom_vline(xintercept = 0, colour = "grey70") +
  geom_line(aes(group = ykey), colour = "grey65", linewidth = .5) +
  geom_point(aes(colour = sex), size = 2.6) +
  scale_colour_manual(values = c(`mid males` = "#0072B2",
                                 `mid females` = "#CC79A7"),
                      name = NULL) +
  scale_y_discrete(labels = function(x) sub("^[^|]*\\|", "", x)) +
  facet_wrap(~ source, scales = "free_y", ncol = 2) +
  labs(x = "mean signature score", y = NULL,
       title = "Germ signature scores in Yasmin's mid males and mid females",
       subtitle = "groups ordered by the male-minus-female difference") +
  theme_bw(base_size = 10) +
  theme(legend.position = "top",
        panel.grid.major.y = element_line(colour = "grey92"))

p_sex

ggsave(p_sex, file = paste0(results_dir, "/germ_signature_sex.png"),
       width = 11, height = 7, dpi = 300)
```

`r if (!have_cl_markers) "" else { w = sex_tab %>% dplyr::filter(difference > 0); paste0("Scores are averages of genes standardized across all 28 samples, so zero is the series mean rather than no expression. ", nrow(w), " of ", nrow(sex_tab), " signatures score higher in the mid males. A signature whose two points nearly coincide is not reading identity, whatever it was built from.") }`


# Full gene table

```{r, eval=have_bulk}
out = cmp %>%
  dplyr::select(Gene, WangGeneID, ours_male, ours_female,
                dplyr::starts_with("male_"), dplyr::starts_with("fem_"))

write.csv(out, file = paste0(results_dir, "/bulk_vs_pseudobulk_per_gene.csv"),
          row.names = FALSE)

datatable(head(out %>% mutate(across(where(is.numeric), ~signif(., 3))), 100),
          rownames = FALSE, filter = "top")
```

`r if (have_bulk && exists("have_gd")) { if (!have_gd) paste0("**WangGeneID is not shown.** ", gd_file, " was not found, so the export carries NCBI gene ids only.") else { n_na = sum(is.na(out$WangGeneID)); if (n_na > 0) paste0(format(n_na, big.mark = ","), " of ", format(nrow(out), big.mark = ","), " genes have no Wang identifier and are left blank.") else "Every gene in the export carries a Wang identifier." } } else ""`

```{r, results="asis"}
# Nothing is written when the bulk matrix is absent, so the section does not
# print a table of nothing.
if (have_bulk) cat("# Export\n\n## Everything written by this document\n")
```

```{r, eval=have_bulk}
file_notes = c(
  "trend_scores.csv"               = "The ordered covariate every Yasmin animal takes on its sex's trend axis, with the group, length and recorded life stage it was built from.",
  "bulk_agreement.csv"             = "Spearman rho and percent sign agreement between our stage effect and Yasmin's, for every axis, gene pool and ranking side, with bootstrap intervals.",
  "bulk_permutation_null.csv"      = "Each agreement statistic against its label-permutation null: the observed value, the null median and 95th percentile, and the empirical p-value.",
  "bulk_permutation_null.png"      = "The null distribution of rho across the top genes, with the observed value marked.",
  "bulk_trend_agreement.png"       = "The strongest concordant genes, our effect against Yasmin's, and the same genes with both effect sizes side by side.",
  "bulk_agreement_scatter.png"     = "Our log fold change against Yasmin's for every measurable gene, one panel per axis.",
  "top_gene_GO_pvalues.csv"        = "GO enrichment of the top stage genes from both sides, up and down taken separately, one row per term and gene set.",
  "top_gene_GO_overlaps.csv"       = "The genes behind each enriched term.",
  "top_gene_list_overlap.csv"      = "Every pairing of one of our top gene lists with one of Yasmin's: the two list sizes, the overlap, the overlap expected by chance and a hypergeometric p-value, with the same-sex same-direction pairings flagged.",
  "top_gene_list_overlap_genes.csv" = "The genes in each same-sex same-direction overlap, with our effect size for each.",
  "top_gene_list_upset.png"        = "Intersections of our top gene lists with Yasmin's trend lists, largest first, as an UpSet plot.",
  "shared_trend_GO_terms.csv"      = "Terms enriched in our top genes and in Yasmin's trend genes on the same sex in the same direction, with each side's p-value, FDR and gene count.",
  "shared_trend_GO_terms.png"      = "The strongest of those shared terms, each with a point per dataset, faceted by sex and direction.",
  "germ_signature_sex.csv"         = "Mean germ signature score in Yasmin's mid males and mid females, per signature and marker source, with the difference between them.",
  "germ_signature_sex.png"         = "The same scores as paired points per signature, ordered by the male-minus-female difference.",
  "bulk_vs_pseudobulk_per_gene.csv" = "One row per gene: our male and female stage effects, every Yasmin contrast's log fold change, and the Wang identifier.")

in_dir = sort(setdiff(list.files(results_dir), "Robjects"))

# The table describes what this document writes. The results directory also
# holds whatever earlier runs left behind, and those are named below rather
# than given rows with no description.
on_disk = intersect(names(file_notes), in_dir)
unknown = setdiff(in_dir, names(file_notes))
missing = setdiff(names(file_notes), in_dir)

export_tab = data.frame(
  file = on_disk,
  size = sapply(file.path(results_dir, on_disk),
                function(f) format(structure(file.size(f), class = "object_size"),
                                   units = "auto")),
  what = unname(file_notes[on_disk]),
  stringsAsFactors = FALSE)

datatable(export_tab, rownames = FALSE, escape = FALSE,
          options = list(pageLength = 20, dom = "tp",
                         columnDefs = list(list(width = "260px", targets = 0))))
```

`r if (!have_bulk) "" else if (length(unknown) > 0) paste0(length(unknown), " file(s) in the directory are not written by this document and are left out of the table: ", paste(unknown, collapse = ", "), ". Either an earlier version wrote them, or they are new and need a line in file_notes.") else ""`

`r if (!have_bulk) "" else if (length(missing) > 0) paste0("Described but not written on this run: ", paste(missing, collapse = ", "), ". A section that was skipped writes nothing.") else ""`

`r if (!have_bulk) "" else paste0("Two cached intermediates sit in ", results_dir, "/Robjects and are not listed: pseudobulk.Rds, the per-library count sums, and label_permutation_null.Rds, the permutation refits. Both are reused across runs and rebuilt when the inputs change.")`


# Summary

```{r, eval=have_bulk}
# The headline numbers come from the well measured pool. Both ranking directions
# are shown, since they answer different questions.
top_set = paste0("top ", top_ns[1])

row_for = function(ax, set, ro = "ours")
  agreement[agreement$axis == ax & agreement$pool == "well measured" &
              agreement$ranked_on == ro & agreement$set == set, ]

perm_for = function(ax, stat, ro = "ours")
  perm_tab$p_perm[perm_tab$axis == ax & perm_tab$pool == "well measured" &
                    perm_tab$ranked_on == ro & perm_tab$statistic == stat]

axis_block = function(ax, label) {
  ours_row = row_for(ax, top_set, "ours")
  yas_row  = row_for(ax, top_set, "yasmin")
  gw_row   = row_for(ax, "genome-wide", "ours")
  data.frame(
    question = paste0(label, c(", rho genome-wide",
                               ", rho on our top genes",
                               ", permutation p, our ranking",
                               ", rho on Yasmin's top genes",
                               ", permutation p, their ranking",
                               ", % same sign on their top genes")),
    answer = c(signif(gw_row$rho, 3),
               signif(ours_row$rho, 3),
               signif(perm_for(ax, paste0("rho, ", top_set), "ours"), 3),
               signif(yas_row$rho, 3),
               signif(perm_for(ax, paste0("rho, ", top_set), "yasmin"), 3),
               paste0(signif(yas_row$pct_same_sign, 3), "%")))
}

summary_tbl = dplyr::bind_rows(
  data.frame(question = c("shared genes tested",
                          paste0("genes at or above ", min_pb, " counts here")),
             answer = c(format(nrow(cmp), big.mark = ","),
                        format(sum(cmp$well_measured), big.mark = ","))),
  axis_block("male, endpoint",               "male endpoint"),
  axis_block("female, endpoint",             "female endpoint"),
  axis_block("male, trend",                  "male trend"),
  axis_block("female, trend",                "female trend"),
  data.frame(question = c("animals in the male endpoint",
                          "animals in the female endpoint",
                          "animals in each male trend",
                          "animals in each female trend"),
             answer = as.character(c(length(runs_ud) + length(runs_mm),
                                     length(runs_ud) + length(runs_mf),
                                     length(male_runs),
                                     length(fem_runs)))))

datatable(summary_tbl, rownames = FALSE,
          options = list(dom = "tp", ordering = FALSE, pageLength = 22))
```

## What this does not answer

Agreement shows the pattern reproduces in a series without our July-to-November
batch. It does not separate stage from batch within our data.

Both endpoints use the whole mid group. Neither is stage-matched to our
transformers animal by animal. The mid male group is stage 7 and
post-metamorphosis throughout. The mid female group spans stage 1 to
post-metamorphosis, which makes its late endpoint broader than ours. Read each
contrast on its own size rather than treating one sex as primary.

The larval endpoint is matched by stage description rather than by size. Our
larvae are sex-labile by our own staging and larger than every larval animal in
the comparison series. The two series set the larval boundary at different sizes.

The permutation asks whether Yasmin's stage contrast agrees with our genes more
than an arbitrary contrast drawn from the same animals. It does not rule out a
confound shared by both studies. Anything that moves gonad composition the same
way in both series would pass it.

# Session info

```{r}
sessionInfo()
```
