---
title: "16_bulk_validation"
author: "Stephanie Hickey"
date: "`r Sys.Date()`"
output:
  html_document:
    code_folding: hide
    toc: true
    toc_float:
      toc_collapsed: true
params:
    project_dir: /mnt/ufs18/rs-013/bioinformaticsCore/projects/chong_davidson/BCC125_sea_lamprey_scRNA
    project_name: BCC125_sea_lamprey_scRNA
    lab: Weiming Li
    results_date: "20260813"
    genome_dir: /mnt/research/bioinformaticsCore/shared/Genomes/Petromyzon_marinus
    nfcore_dir: /mnt/research/bioinformaticsCore/projects/chong_davidson/BCC125_sea_lamprey_scRNA/results/yasmin_nfcore
    group_column: group_suppT1labels
    top_n_list: "100,250,500"
    n_boot: 2000
    n_perm_labels: 200
    min_pseudobulk_count: 10
    go_pval: 0.01
    germ_clusters: "5,15,16,17,18"
    germ_marker_lineage: all
    germ_marker_suffix: "_recluster_remove_IEG"
    n_signature_genes: 100
    cores: 1
---

```{r setup, include=FALSE}
# html_document renders inline figures at knitr's 72 dpi doubled by
# fig.retina, so 144, while every ggsave here writes at 300. Rendering inline
# at 300 as well puts the figure in the document at the resolution of the file
# on disk. fig.retina is set to 1 so the two are not multiplied.
# Figures are vector by default. A raster figure in html_document is rendered
# at a fixed pixel size and then scaled to the page, so it is only ever as sharp
# as that size; svg has no resolution to run out of. The chunk that draws one
# mark per gene sets dev = "png" individually, because an svg of twenty thousand
# points is tens of megabytes where the png is a couple of hundred kilobytes.
# dpi applies to those, and fig.retina is 1 so it is not doubled again.
# svglite if it is installed, otherwise the grDevices svg device, which needs
# the cairo capability enabled below
vec_dev = if (requireNamespace("svglite", quietly = TRUE)) "svglite" else "svg"
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      dev = vec_dev, dpi = 300, fig.retina = 1)

# Compute nodes have no X11, and the default png device asks for it. Cairo
# renders off screen.
if (isTRUE(capabilities("cairo"))) options(bitmapType = "cairo")
```

```{r, echo=FALSE, results="asis"}
cat("\n")
cat("#### Lab:", params$lab, " \n")
cat("#### Project:", params$project_name, " \n")
```

```{r}
library(tidyverse)
library(SingleCellExperiment)
library(edgeR)
library(patchwork)
library(ggrepel)
library(DT)
source(paste0(params$project_dir, "/src/scRNA_seq_functions.R"))
```

Independent evidence for the stage-associated patterns. Answers the reviewer's request for
"independent evidence supporting the stage-associated patterns, such as
agreement with staining or previous bulk RNA-seq studies."

# Scope

**What agreement shows.** Whether the larva-to-transformer expression changes we
report reproduce in animals collected, dissected, prepared and sequenced by a
different group at a different time. Our July-to-November technical batch does
not exist in Yasmin et al data. A pattern driven by that batch has no reason to
appear there.

**What it does not show.** That the changes are developmental. Yasmin et al series
has its own collection structure and may carry its own confounds. Agreement
shows the pattern is not specific to our batch. That is what the reviewer asks
for. The claim goes no further.

**Bulk is a mixture.** A gene changing in bulk can reflect a shift in which cell
types are present rather than a change within any cell type. Pseudobulk shares
that property. The two sides are on the same footing. The composition section
measures how much of the agreement composition accounts for.

**The larval endpoint is matched by stage rather than by size.** Our larvae are
112 mm and sex-labile by our own staging. Every larval-stage animal in Yasmin et al
series is smaller, the largest being 99 mm. Their 74–99 mm animals are already
classed as past the point where ovarian differentiation completes. The two series
draw the larval boundary at different sizes. The transformer endpoints match on
stage and on size.

```{r}
object_dir  = paste0(params$project_dir, "/results/", params$results_date,
                     "_no_doublets")
atlas_dir   = paste0(params$project_dir, "/results/", params$results_date,
                     "_atlas")
results_dir = paste0(params$project_dir, "/results/", params$results_date,
                     "_bulk_validation")
Robject_out_dir = paste0(results_dir, "/Robjects")
if (!dir.exists(Robject_out_dir)) dir.create(Robject_out_dir, recursive = TRUE)
data_dir    = paste0(params$project_dir, "/data")
```

```{r}
# The Yasmin et al reads were quantified with nf-core/rnaseq against
# GCF_010993605.1_kPetMar1.pri. Point nfcore_dir at that run's output directory.
# Either the top level or star_salmon/ itself is accepted.
#
# Length-scaled counts come first in the candidate list. They carry the
# transcript-length correction that makes gene-level Salmon estimates comparable
# across samples, and they are the right edgeR input without offsets. Plain
# gene_counts is the fallback.
nfcore_candidates = c(
  "star_salmon/salmon.merged.gene_counts_length_scaled.tsv",
  "salmon/salmon.merged.gene_counts_length_scaled.tsv",
  "star_salmon/salmon.merged.gene_counts.tsv",
  "salmon/salmon.merged.gene_counts.tsv",
  "salmon.merged.gene_counts_length_scaled.tsv",
  "salmon.merged.gene_counts.tsv")

nfcore_dir  = as.character(params$nfcore_dir)
count_path  = NA_character_

if (nzchar(nfcore_dir)) {
  hits = file.path(nfcore_dir, nfcore_candidates)
  hits = hits[file.exists(hits)]
  if (length(hits) > 0) count_path = hits[1]
}

have_bulk    = !is.na(count_path)
length_scaled = have_bulk && grepl("length_scaled", count_path)
```

`r if (!have_bulk && !nzchar(nfcore_dir)) "**The nfcore_dir parameter is not set.** Everything below is skipped. Set it to the nf-core/rnaseq output directory for the Yasmin et al data, either the run's top level or its star_salmon subdirectory." else ""`

`r if (!have_bulk && nzchar(nfcore_dir)) paste0("**No gene count matrix was found under ", nfcore_dir, ".** Everything below is skipped. Looked for: ", paste(nfcore_candidates, collapse = ", "), ". Found instead: ", paste(head(list.files(nfcore_dir), 20), collapse = ", "), ".") else ""`

`r if (have_bulk) paste0("Counts read from ", basename(count_path), if (length_scaled) ". The length-scaled matrix is the right edgeR input without offsets." else ". **This is the unscaled gene_counts matrix.** The transcript-length correction is absent. Use salmon.merged.gene_counts_length_scaled.tsv if the run produced one.") else ""`

# The Yasmin et al series

```{r, eval=have_bulk}
sample_meta = read.csv(paste0(data_dir, "/yasmin_PRJNA749754_sample_groups.csv"),
                       stringsAsFactors = FALSE)

# nf-core/rnaseq gene matrix: gene_id, gene_name, then one column per sample.
x = read.delim(count_path, check.names = FALSE, stringsAsFactors = FALSE)

id_cols = intersect(c("gene_id", "gene_name"), colnames(x))
if (!"gene_id" %in% id_cols)
  stop("no gene_id column in ", count_path, ". Columns found: ",
       paste(head(colnames(x), 10), collapse = ", "))

yasmin_counts = as.matrix(x[, setdiff(colnames(x), id_cols), drop = FALSE])
rownames(yasmin_counts) = x$gene_id

# Salmon estimates are non-integer. edgeR wants counts, and the length-scaled
# matrix is already on the count scale, so rounding is the only step needed.
yasmin_counts = round(yasmin_counts)
storage.mode(yasmin_counts) = "integer"

# nf-core sample names come from the samplesheet the run used. They may be run
# accessions, the paper's own animal names, or the two glued together, as in
# SRR15275627_F_JUV. Read the column names in all three forms and take the
# form and sample sheet column that resolve the most samples.
raw_ids = colnames(yasmin_counts)
has_acc = grepl("^[DES]RR[0-9]+", raw_ids)

id_forms = list(
  "as given"      = raw_ids,
  "run accession" = ifelse(has_acc, sub("^([DES]RR[0-9]+).*$", "\\1", raw_ids),
                           NA_character_),
  "name after the accession" =
                    ifelse(grepl("^[DES]RR[0-9]+_", raw_ids),
                           sub("^[DES]RR[0-9]+_", "", raw_ids), NA_character_))

id_candidates = c("run", "sample_name", "library_name", "suppT1_id")
id_candidates = intersect(id_candidates, colnames(sample_meta))

match_grid = expand.grid(form = names(id_forms), column = id_candidates,
                         stringsAsFactors = FALSE)
match_grid$n_matched = mapply(function(f, cl)
  sum(id_forms[[f]] %in% sample_meta[[cl]], na.rm = TRUE),
  match_grid$form, match_grid$column)

best          = match_grid[which.max(match_grid$n_matched), ]
sample_ids    = id_forms[[best$form]]
sample_id_col = best$column

if (best$n_matched < ncol(yasmin_counts))
  stop("could not match all ", ncol(yasmin_counts), " count columns to the ",
       "sample sheet. The best pairing was the '", best$form, "' form of the ",
       "column names against sample sheet column '", sample_id_col,
       "', matching ", best$n_matched, ". Unmatched: ",
       paste(raw_ids[!(sample_ids %in% sample_meta[[sample_id_col]])],
             collapse = ", "),
       ". Add the nf-core sample names to ",
       "data/yasmin_PRJNA749754_sample_groups.csv as a new column.")

absent = setdiff(sample_meta[[sample_id_col]], sample_ids)
if (length(absent) > 0)
  stop(length(absent), " sample(s) in the sample sheet are missing from the ",
       "count matrix: ", paste(absent, collapse = ", "),
       ". The nf-core run is incomplete, or its samplesheet omitted them.")

sample_meta = sample_meta[match(sample_ids, sample_meta[[sample_id_col]]), ]
stopifnot(identical(sample_meta[[sample_id_col]], sample_ids))

# Everything downstream selects columns with sample_meta$run, so put the count
# matrix on run accessions whatever form the nf-core names took.
stopifnot(!anyDuplicated(sample_meta$run))
colnames(yasmin_counts) = sample_meta$run

id_note = paste0("Count columns were matched to the sample sheet by their ",
                 best$form, ", against the `", sample_id_col, "` column. All ",
                 ncol(yasmin_counts), " samples resolved.")
```

`r if (have_bulk) id_note else ""`

```{r, eval=have_bulk}
# a numeric developmental position per animal, parsed from the recorded life
# stage rather than from the group label. The group label is too coarse on its
# own: the six mid females run from metamorphic stage 1 to post-metamorphosis
# and all carry the same label, so the label alone cannot order them.
stage_score = function(s) {
  s = tolower(s)
  dplyr::case_when(
    grepl("post-metamorphosis", s)             ~ 8,
    grepl("metamorphosing stage", s)           ~ suppressWarnings(
                                                   as.numeric(sub(".*stage ([0-9]+).*",
                                                                  "\\1", s))),
    grepl("larval", s)                         ~ 0,
    grepl("upstream migrant", s)               ~ 10,
    TRUE                                       ~ NA_real_)
}

sample_meta$stage_score = stage_score(sample_meta$suppT1_life_stage)
stopifnot(!any(is.na(sample_meta$stage_score)))

# Which group assignment to use. The sample sheet carries two, because the paper
# and its Supplementary Table 1 split early from mid males differently: the text
# and count tables cut at metamorphic stage 5, Supplementary Table 1 cuts at
# stage 7. They disagree for M_M5 and M_M6 and agree everywhere else, so no
# female group and no female result depends on this choice.
group_col = as.character(params$group_column)
stopifnot(group_col %in% colnames(sample_meta))
sample_meta$group_used = sample_meta[[group_col]]

series_tab = sample_meta %>%
  dplyr::group_by(group_used) %>%
  dplyr::summarise(n = dplyr::n(),
                   length_mm = paste0(min(suppT1_length_mm), "-",
                                      max(suppT1_length_mm)),
                   stages = paste(sort(unique(suppT1_life_stage)), collapse = "; "),
                   .groups = "drop")

datatable(series_tab, rownames = FALSE, options = list(dom = "t", paging = FALSE))
```

`r if (have_bulk) paste0("Groups are taken from the **", group_col, "** column. ", if (group_col == "group_suppT1labels") "Supplementary Table 1 labels its own rows M_E1 to M_E5 and M_M1 to M_M4, putting metamorphic stages 1 to 6 in early males and stage 7 with post-metamorphosis in mid males. This matches how the main text describes mid males, \"metamorphosing stage 7 and immediately post-metamorphosis\". It does not reproduce the n of 4 early and 6 mid printed in the text, which the group_textcounts column does. Mid males here are the four animals at stage 7 or beyond, which is the same rule used to pick the stage-matched mid females, so the two sexes are selected alike." else "This column reproduces the group sizes printed in the main text by cutting early from mid males at metamorphic stage 5. Supplementary Table 1 labels those rows differently, cutting at stage 7, which is what group_suppT1labels holds.") else ""`

## Stage correspondence

Our animals are 112 mm sex-labile larvae and 131 mm newly transformed
transformers. The matches below use the
per-animal records in Supplementary Table 1 rather than the group labels.

```{r, eval=have_bulk}
# The animals within each mid group that are staged like our transformers, kept
# for the correspondence table. The endpoint contrasts use the whole group.
stage_matched = function(lbl) sample_meta$sample_name[
  sample_meta$group_used == lbl & sample_meta$stage_score >= 7]

len_range = function(lbl) {
  L = sample_meta$suppT1_length_mm[sample_meta$group_used == lbl]
  paste0(min(L), "-", max(L), " mm")
}

match_tab = tibble::tribble(
  ~ours,            ~yasmin,                                     ~basis,
  "larvae, 112 mm", paste0("undifferentiated larvae, ",
                           len_range("UD undifferentiated larvae")),
  "both pre-differentiation by description; ours are sex-labile by our staging; the two series set the larval size boundary differently",
  "TM, 131 mm",     paste0("all mid males, ", len_range("MM mid males")),
  paste0("stage 7 and post-metamorphosis by the Supplementary Table 1 labels; closest staged: ",
         paste(stage_matched("MM mid males"), collapse = ", ")),
  "TF, 131 mm",     paste0("all mid females, ", len_range("MF mid females")),
  paste0("group spans stage 1 to post-metamorphosis, used whole; closest staged: ",
         paste(stage_matched("MF mid females"), collapse = ", ")))

datatable(match_tab, rownames = FALSE, options = list(dom = "t", paging = FALSE))
```

# Pseudobulk

Raw counts are summed across all cells within each library, giving four samples.
Corrected and per-cell normalized values are not used. A bulk library measures
summed raw counts.

```{r, eval=have_bulk}
pseudo_file = paste0(Robject_out_dir, "/pseudobulk.Rds")

if (file.exists(pseudo_file)) {
  pseudobulk = readRDS(pseudo_file)
} else {
  all.sce = readRDS(paste0(object_dir, "/all.sce.Rds"))
  universe = Reduce(intersect, lapply(all.sce, rownames))
  pseudobulk = vapply(all.sce, function(x)
    Matrix::rowSums(counts(x[universe, ])), numeric(length(universe)))
  rownames(pseudobulk) = universe
  saveRDS(pseudobulk, file = pseudo_file)
}

library_meta = data.frame(
  batch   = c("UND1_C00004336", "UND2_C00004362", "TF_C00004337", "TM_C00004374"),
  library = c("Undifferentiated.1", "Undifferentiated.2", "Female", "Male"),
  stage   = c("larva", "larva", "transformer", "transformer"),
  stringsAsFactors = FALSE)

colnames(pseudobulk) = library_meta$library[
  match(colnames(pseudobulk), library_meta$batch)]
stopifnot(!any(is.na(colnames(pseudobulk))))
```

`r if (have_bulk) paste0("Pseudobulk over ", format(nrow(pseudobulk), big.mark = ","), " genes and ", ncol(pseudobulk), " libraries.") else ""`

## Shared gene set

```{r, eval=have_bulk}
shared = intersect(rownames(pseudobulk), rownames(yasmin_counts))

overlap_tab = data.frame(
  set = c("our single-cell genes", "Yasmin et al quantified genes", "shared"),
  n   = c(nrow(pseudobulk), nrow(yasmin_counts), length(shared)))

datatable(overlap_tab, rownames = FALSE, options = list(dom = "t"))

if (length(shared) < 0.5 * nrow(pseudobulk))
  stop("only ", length(shared), " of ", nrow(pseudobulk), " genes are shared. ",
       "Both sides should be NCBI kPetMar1 gene ids, so a low overlap means the ",
       "nf-core run used a different annotation from the single-cell counts. ",
       "Diagnose before continuing.")
```

`r if (have_bulk) paste0("**", format(length(shared), big.mark = ","), " genes are shared.** Everything below runs on that set. Genes are matched by identifier. Both sides come from the same assembly, so no ortholog assignment is involved.") else ""`

# Yasmin et al stage effects

The counts are the nf-core/rnaseq gene matrix for the Yasmin et al reads, taken
length-scaled where that output exists and rounded, since Salmon estimates are
not integers. They are cut to the genes our pseudobulk also measures, then
filtered with `filterByExpr` on their own 28 samples under their own group
labels, so the filter is theirs rather than one imposed by our data, and
normalized with `calcNormFactors`.

```{r, eval=have_bulk}
y = DGEList(counts = yasmin_counts[shared, ], samples = sample_meta)
keep = filterByExpr(y, group = sample_meta$group_used)
y = y[keep, , keep.lib.sizes = FALSE]
y = calcNormFactors(y)

tested = rownames(y)
```

`r if (have_bulk) paste0(format(length(tested), big.mark = ","), " genes pass filterByExpr in the Yasmin et al series and carry through to the comparison.") else ""`

```{r, eval=have_bulk}
# Two contrast shapes serve the whole document: a two-group test and an ordered
# trend. Both return log fold changes, p-values and FDR per gene.
run_two_group = function(dge, run_a, run_b, label) {
  sub = dge[, c(run_a, run_b)]
  grp = factor(c(rep("early", length(run_a)), rep("late", length(run_b))),
               levels = c("early", "late"))
  des = model.matrix(~grp)
  sub = estimateDisp(sub, des)
  fit = glmQLFit(sub, des)
  res = glmQLFTest(fit, coef = 2)
  topTags(res, n = Inf, sort.by = "none")$table %>%
    tibble::rownames_to_column("Gene") %>%
    # The signed test statistic, which the enrichment section ranks on.
    # glmQLFTest reports an F with one numerator degree of freedom, so its root
    # is the absolute t-like statistic and logFC supplies the sign. It is carried
    # here rather than recomputed, because a p-value cannot be signed and the
    # fold change alone ignores how well each gene is measured.
    dplyr::mutate(stat = sign(logFC) * sqrt(.data$F)) %>%
    dplyr::select(Gene, logFC, stat, PValue, FDR) %>%
    dplyr::rename_with(~paste0(label, "_", .x), -Gene)
}

run_trend = function(dge, runs, score, label) {
  sub = dge[, runs]
  des = model.matrix(~score)
  sub = estimateDisp(sub, des)
  fit = glmQLFit(sub, des)
  res = glmQLFTest(fit, coef = 2)
  # Two columns, because they answer different questions and one of them is what
  # edgeR actually returned.
  #
  #   slope          edgeR's coefficient, renamed. On a continuous covariate it
  #                  is a slope, the log2 change for a step of 1 on the score,
  #                  not the fold change that run_two_group returns as logFC.
  #                  Kept so a rerun of glmQLFTest matches this file.
  #   series_log2FC  the same fit read across the span the score covers, which
  #                  is the log2 change from one end of the series to the other.
  #                  The model is linear in score, so this is a rescaling and
  #                  needs no refit. Everything downstream is rank or sign
  #                  based, so the two give identical statistics; this one is
  #                  the one that can be read as a size.
  span = diff(range(score))
  topTags(res, n = Inf, sort.by = "none")$table %>%
    tibble::rownames_to_column("Gene") %>%
    dplyr::mutate(stat = sign(logFC) * sqrt(.data$F)) %>%
    dplyr::select(Gene, logFC, stat, PValue, FDR) %>%
    dplyr::rename(slope = "logFC") %>%
    dplyr::mutate(series_log2FC = slope * span) %>%
    dplyr::relocate(series_log2FC, .after = slope) %>%
    dplyr::rename_with(~paste0(label, "_", .x), -Gene)
}

# Spearman with a bootstrap CI. cor.test gives a p-value but no interval, and
# the interval is the quantity worth reporting.
spearman_ci = function(x, y, n_boot = params$n_boot, seed = 1) {
  ok = is.finite(x) & is.finite(y)
  x = x[ok]; y = y[ok]
  r = suppressWarnings(cor(x, y, method = "spearman"))
  set.seed(seed)
  bs = replicate(n_boot, {
    i = sample(length(x), replace = TRUE)
    suppressWarnings(cor(x[i], y[i], method = "spearman"))
  })
  c(rho = r, lo = unname(quantile(bs, .025, na.rm = TRUE)),
    hi = unname(quantile(bs, .975, na.rm = TRUE)), n = length(x))
}

g = sample_meta$group_used
runs_ud = sample_meta$run[g == "UD undifferentiated larvae"]
runs_pm = sample_meta$run[g == "PM prospective males"]
runs_em = sample_meta$run[g == "EM early males"]
runs_mm = sample_meta$run[g == "MM mid males"]
runs_ef = sample_meta$run[g == "EF early females"]
runs_mf = sample_meta$run[g == "MF mid females"]
```

Each endpoint is an ordinary two-group test: the undifferentiated larvae against
one mid group, fitted on just those libraries. The design is a two-level factor,
dispersions are estimated within that subset, and the quasi-likelihood F test is
taken on the second coefficient, so a positive log fold change means higher in
the mid group than in the larvae. That matches the sign of our own effect, which
runs from larva to transformer.

`r if (have_bulk) paste0("The male endpoint is ", length(runs_ud), " undifferentiated larvae against ", length(runs_mm), " mid males, and the female endpoint the same larvae against ", length(runs_mf), " mid females.") else ""` Both
endpoints share that larval side, so the two are not independent of each other,
and each dispersion estimate comes from its own handful of libraries rather than
from all 28. The late side is the whole mid group rather than animals matched to
our transformers by stage.


Each trend is a regression rather than a contrast between two groups. Every
animal on a sex's axis enters at once carrying its own trend score, and
expression is regressed on that score, so the intermediate groups contribute
rather than being set aside.

The design matrix holds an intercept and one column, the score, so the model has
a single coefficient beyond the intercept and that coefficient is the slope.
Dispersions are estimated on those libraries, and the quasi-likelihood F test is
taken on the slope. A positive slope means the gene rises along the ordering.

The slope is reported two ways, as the coefficient itself and as the change
across the span the score covers; the scatter section below says which is which
and why.


```{r, eval=have_bulk}
# Two endpoints and two trends.
#
# The endpoints are larvae against the whole mid group, one per sex, so the two
# sexes are contrasted the same way and on the largest groups available.
#
# The trends order the animals by group and then, inside each group, by the
# metamorphic stage recorded for that animal in Supplementary Table 1. The
# within-group offset is scaled to sit inside the gap to the next group, so
# groups never overlap.
#
# The offset uses the finest developmental information the paper reports. It
# changes nothing in the groups whose animals share a stage: undifferentiated
# larvae, prospective males and early females are all recorded as larval, so
# they stay flat. It orders early males (stages 1, 1, 3, 4, 5, 6), mid females
# (1, 1, 3, 4, 7, 8) and, barely, mid males (7, 7, 8, 8).
stage_offset = function(runs) {
  st = sample_meta$stage_score[match(runs, sample_meta$run)]
  (rank(st, ties.method = "min") - 1) / length(runs)
}

make_score = function(group_runs, mode) {
  unlist(lapply(seq_along(group_runs), function(i) {
    runs = group_runs[[i]]
    (i - 1) + if (mode == "stage") stage_offset(runs) else rep(0, length(runs))
  }))
}

male_end = run_two_group(y, runs_ud, runs_mm, "male_end")
fem_end  = run_two_group(y, runs_ud, runs_mf, "fem_end")

male_groups = list(runs_ud, runs_pm, runs_em, runs_mm)
male_runs   = unlist(male_groups)
male_group_index = make_score(male_groups, "group")
male_score       = make_score(male_groups, "stage")
male_trend       = run_trend(y, male_runs, male_score, "male_trend")

fem_groups = list(runs_ud, runs_ef, runs_mf)
fem_runs   = unlist(fem_groups)
fem_group_index = make_score(fem_groups, "group")
fem_score       = make_score(fem_groups, "stage")
fem_trend       = run_trend(y, fem_runs, fem_score, "fem_trend")

yasmin_res = Reduce(function(a, b) dplyr::full_join(a, b, by = "Gene"),
                    list(male_end, fem_end, male_trend, fem_trend))
```

`r if (have_bulk) paste0("The male axis carries ", length(male_runs), " animals across ", length(male_groups), " groups and the female axis ", length(fem_runs), " across ", length(fem_groups), ".") else ""`

## Trend scores

Each trend regresses expression on a single ordered covariate. The covariate
assigns a number to every animal.

The groups supply the coarse order, the one Yasmin et al give: undifferentiated larvae,
then prospective males, early males and mid males on the male axis;
undifferentiated larvae, early females and mid females on the female axis. That
is the `group_index` column below.

Within a group the animals are ordered by their own recorded life stage.
Supplementary Table 1 gives one for every animal. It is finer than the group
labels. Mid females span metamorphic stage 1 to post-metamorphosis, and early
males span stages 1 to 6. Scoring by group alone would discard per-animal
information the paper reports.

The offset is built so it can only order animals inside a group, never move one
past another group:

1. Parse the recorded life stage to a number. Larval is 0, metamorphosing stage
   *N* is *N*, post-metamorphosis is 8, upstream migrant is 10. `M_M1b`, recorded
   as "Larval stage/Metamorphosing stage 1", is read as stage 1.
2. Rank the animals within their group by that number, ties taking the lower
   rank.
3. Subtract one and divide by the group size, giving an offset in [0, 1).
4. Add it to the group integer.

Because the offset is always below 1, group *k* occupies [*k*, *k*+1) and the
groups stay in order. Where every animal in a group shares a stage the offset is
0 for all of them, so the two scorings agree there. That is the case for the
undifferentiated larvae, the prospective males and the early females, which are
all recorded simply as larval.

```{r, eval=have_bulk}
score_tab = dplyr::bind_rows(
  data.frame(axis = "male", run = male_runs,
             group_index = male_group_index, trend_score = male_score),
  data.frame(axis = "female", run = fem_runs,
             group_index = fem_group_index, trend_score = fem_score)) %>%
  dplyr::mutate(
    group  = sample_meta$group_used[match(run, sample_meta$run)],
    sample = sample_meta$sample_name[match(run, sample_meta$run)],
    length_mm  = sample_meta$suppT1_length_mm[match(run, sample_meta$run)],
    life_stage = sample_meta$suppT1_life_stage[match(run, sample_meta$run)]) %>%
  dplyr::arrange(axis, trend_score) %>%
  dplyr::select(axis, group, sample, length_mm, life_stage,
                group_index, trend_score)

write.csv(score_tab, file = paste0(results_dir, "/trend_scores.csv"),
          row.names = FALSE)

datatable(score_tab %>% mutate(across(where(is.numeric), ~signif(., 3))),
          rownames = FALSE, options = list(dom = "t", pageLength = 30))
```

```{r, eval=have_bulk}
contrast_tab = data.frame(
  contrast = c("male endpoint", "female endpoint",
               "male trend", "female trend"),
  design   = c(paste0("UD (n=", length(runs_ud), ") vs all mid males (n=",
                      length(runs_mm), ")"),
               paste0("UD (n=", length(runs_ud), ") vs all mid females (n=",
                      length(runs_mf), ")"),
               paste0("UD, PM, EM, MM ordered (n=", length(male_runs), ")"),
               paste0("UD, EF, MF ordered (n=", length(fem_runs), ")")),
  n_FDR_05 = c(sum(yasmin_res$male_end_FDR   < 0.05, na.rm = TRUE),
               sum(yasmin_res$fem_end_FDR    < 0.05, na.rm = TRUE),
               sum(yasmin_res$male_trend_FDR < 0.05, na.rm = TRUE),
               sum(yasmin_res$fem_trend_FDR  < 0.05, na.rm = TRUE)))

datatable(contrast_tab, rownames = FALSE, options = list(dom = "t", paging = FALSE))
```

The endpoints are the smaller designs. Both use the two undifferentiated larvae
as the early side. The correlation below uses effect sizes alone. Noise in the
Yasmin et al estimates pulls agreement toward zero. Low power therefore makes that test
conservative rather than invalid. Per-gene significance comes from the trends.

# Our stage effect

Our effect is a pseudobulk log ratio. Counts are summed over every cell of a
library, giving one column per library and four in all: two undifferentiated
larval, one male transformer, one female transformer. Those columns are
normalized with `calcNormFactors` and put on the log2 CPM scale with a prior
count of one. The two larval columns are averaged, and that average is
subtracted from each transformer column, so our male and female effects are each
one transformer library against the mean of the same two larval libraries.

Two larval libraries against one transformer library per sex is not a formal
test. The log fold change is used as an effect size. No p-values are reported
from it.

```{r, eval=have_bulk}
pb = pseudobulk[shared, ][tested, ]

p = DGEList(counts = pb)
p = calcNormFactors(p)
plog = cpm(p, log = TRUE, prior.count = 1)

larva_mean = rowMeans(plog[, c("Undifferentiated.1", "Undifferentiated.2")])

min_pb = as.integer(params$min_pseudobulk_count)

ours = data.frame(
  Gene          = rownames(plog),
  ours_male     = plog[, "Male"]   - larva_mean,
  ours_female   = plog[, "Female"] - larva_mean,
  max_count     = apply(pb, 1, max),
  well_measured = apply(pb, 1, max) >= min_pb,
  stringsAsFactors = FALSE)

top_ns = as.integer(trimws(strsplit(as.character(params$top_n_list), ",")[[1]]))

# The n strongest genes in each direction, and the two together.
#
# Selecting the n strongest by absolute effect gives a list whose direction
# balance is whatever the data happens to carry, and here that is lopsided: the
# largest magnitudes are overwhelmingly downward on both sides, so a top list
# picked that way is almost entirely down genes. Two things go wrong with it.
# Sign agreement then averages the up and down genes, which behave oppositely,
# and the average can land below chance while both halves sit above it. rho
# loses the range it depends on, because the list crowds into one tail.
#
# Taking n each way fixes the balance by construction and keeps both tails, so
# rho is computed across the full spread of effects. Restricting to a single
# direction instead is worse than either: with the range cut to one tail, rho
# measures magnitude ranking within it and goes to nothing.
top_dir = function(v, direction, n = top_ns[1]) {
  ok = is.finite(v)
  idx = if (direction == "up") which(ok & v > 0) else which(ok & v < 0)
  idx[order(abs(v[idx]), decreasing = TRUE)][seq_len(min(n, length(idx)))]
}
top_both = function(v, n = top_ns[1])
  sort(union(top_dir(v, "up", n), top_dir(v, "down", n)))

# the label the agreement and permutation tables share for that set
top_set_lab = paste0("top ", top_ns[1], " each way")
```

`r if (have_bulk) paste0(format(sum(ours$well_measured), big.mark = ","), " of ", format(nrow(ours), big.mark = ","), " genes reach ", min_pb, " counts in at least one of our four libraries.") else ""`

# Diagnostics

This section checks what our own top gene list is made of before the
comparison below reads anything into it. It does not use the Yasmin et al data.

## What is in the top genes?

```{r, eval=have_bulk}
top_probe = as.integer(trimws(strsplit(as.character(params$top_n_list),
                                       ",")[[1]]))[1]

top_counts = dplyr::bind_rows(lapply(c("ours_male", "ours_female"),
  function(cl) {
    idx = top_both(ours[[cl]], top_probe)
    data.frame(
      set = paste0("top ", top_probe, ", ", sub("ours_", "", cl)),
      median_max_count = median(ours$max_count[idx]),
      pct_below_cutoff = 100 * mean(!ours$well_measured[idx]))
  }))

top_counts = dplyr::bind_rows(
  top_counts,
  data.frame(set = "all genes",
             median_max_count = median(ours$max_count),
             pct_below_cutoff = 100 * mean(!ours$well_measured)))

datatable(top_counts %>% mutate(across(where(is.numeric), ~signif(., 3))),
          rownames = FALSE, options = list(dom = "t", paging = FALSE))
```

`r if (have_bulk) { pick = function(s) top_counts$pct_below_cutoff[top_counts$set == paste0("top ", top_probe, ", ", s)]; pm = pick("male"); pf = pick("female"); hi = c("male", "female")[c(pm > 25, pf > 25)]; nums = paste0(signif(pm, 3), "% of the top male genes and ", signif(pf, 3), "% of the top female genes fall below the count cutoff. "); if (length(hi) == 0) paste0(nums, "Neither unfiltered ranking is dominated by low-count genes.") else paste0("**The top ", paste(hi, collapse = " and "), if (length(hi) == 2) " lists are" else " list is", " largely low-count.** ", nums, "The filtered results below are the ones to read.") } else ""`

# Effect size agreement

Every gene carries an effect size on both sides. This section asks whether the
two sets of numbers move together. It reports the Spearman correlation across
genes, the share of genes the two datasets push the same way, and a null for
each statistic built by relabeling the animals.

```{r, eval=have_bulk}
cmp = ours %>% dplyr::inner_join(yasmin_res, by = "Gene")

# The Wang identifier, so figures and the export carry the name the
# collaborators' sheets use. The table is written by 01, one row per NCBI id.
gd_file = paste0(object_dir, "/gene_descritption.csv")
have_gd = file.exists(gd_file)

if (have_gd) {
  gene_description = read.csv(gd_file, stringsAsFactors = FALSE) %>%
    dplyr::distinct(Gene, .keep_all = TRUE)
  cmp = cmp %>%
    dplyr::left_join(gene_description %>% dplyr::select(Gene, WangGeneID),
                     by = "Gene")
} else {
  gene_description = data.frame(Gene = character(), WangGeneID = character(),
                                description = character())
  cmp$WangGeneID = NA_character_
}
cmp$label = ifelse(is.na(cmp$WangGeneID), cmp$Gene, cmp$WangGeneID)

# genome-wide, then restricted to our strongest stage genes.
#
# Relabelled contrasts drawn from the same 28 animals give a genome-wide
# correlation centered at zero, so the genome-wide value is a result in its own
# right rather than a baseline. What the two numbers compare is concentration:
# whether agreement is stronger among the genes we nominate than among genes in
# general. That is a statement about specificity.
#
# Sign agreement is scored genome-wide for the same purpose. Whether agreement
# exceeds chance is left to the permutation.
# rank_on picks which dataset selects the genes. "ours" is the original
# direction. "yasmin" ranks on their side, where the contrast is replicated and
# thousands of genes reach FDR 0.05, and asks only whether our four libraries
# follow. keep restricts to the genes that carry enough counts here to rank.
agreement_for = function(ours_col, yas_col, axis, keep, pool,
                         rank_on = "ours") {
  d    = cmp[keep, , drop = FALSE]
  rank_vec = if (rank_on == "ours") d[[ours_col]] else d[[yas_col]]

  rows = list()
  gw = spearman_ci(d[[ours_col]], d[[yas_col]])

  gw_agree = sign(d[[ours_col]]) == sign(d[[yas_col]])
  gw_agree = gw_agree[is.finite(gw_agree)]

  rows[[1]] = data.frame(
    axis = axis, pool = pool, ranked_on = rank_on, set = "genome-wide",
    n = gw[["n"]], rho = gw[["rho"]], lo = gw[["lo"]], hi = gw[["hi"]],
    pct_same_sign = 100 * mean(gw_agree))

  for (n in top_ns) {
    idx = top_both(rank_vec, n)
    s = spearman_ci(d[[ours_col]][idx], d[[yas_col]][idx])
    agree = sign(d[[ours_col]][idx]) == sign(d[[yas_col]][idx])
    agree = agree[is.finite(agree)]
    rows[[length(rows) + 1]] = data.frame(
      axis = axis, pool = pool, ranked_on = rank_on,
      set = paste0("top ", n, " each way"),
      n = length(idx), rho = s[["rho"]], lo = s[["lo"]], hi = s[["hi"]],
      pct_same_sign = 100 * mean(agree))
  }
  dplyr::bind_rows(rows)
}

# yas is the effect size, for the scatters and the correlations. stat is the
# signed test statistic, for the enrichment, which needs a ranking that accounts
# for how well each gene is measured.
axis_cols = list(
  "male, endpoint"   = c(ours = "ours_male",   yas = "male_end_logFC",
                         stat = "male_end_stat"),
  "female, endpoint" = c(ours = "ours_female", yas = "fem_end_logFC",
                         stat = "fem_end_stat"),
  "male, trend"      = c(ours = "ours_male",   yas = "male_trend_series_log2FC",
                         stat = "male_trend_stat"),
  "female, trend"    = c(ours = "ours_female", yas = "fem_trend_series_log2FC",
                         stat = "fem_trend_stat"))

# One pool. Running both and crossing every statistic by it changed nothing
# worth reading: on this data the two pools agree to three decimal places on
# rho, agree exactly on most permutation p-values, and never disagree about
# whether a statistic clears 0.05. The filter still decides which genes the top
# lists, the GO background and the trend figure are drawn from, which is where
# it does its work.
gene_pools = list("well measured" = cmp$well_measured)

grid = expand.grid(axis = names(axis_cols), pool = names(gene_pools),
                   rank_on = c("ours", "yasmin"), stringsAsFactors = FALSE)

agreement = dplyr::bind_rows(Map(function(ax, pl, ro) {
  cols = axis_cols[[ax]]
  agreement_for(cols[["ours"]], cols[["yas"]], ax, gene_pools[[pl]], pl, ro)
}, grid$axis, grid$pool, grid$rank_on))

write.csv(agreement, file = paste0(results_dir, "/bulk_agreement.csv"),
          row.names = FALSE)

datatable(agreement %>% dplyr::select(-pool) %>%
            mutate(across(where(is.numeric), ~signif(., 3))),
          rownames = FALSE, filter = "top",
          options = list(pageLength = 20))
```

```{r, eval=have_bulk}
datatable(data.frame(
  column = c("<code>axis</code>", "<code>ranked_on</code>", "<code>set</code>", "<code>n</code>", "<code>rho</code>", "<code>lo</code>, <code>hi</code>", "<code>pct_same_sign</code>"),
  holds  = c("which Yasmin et al contrast, as sex and shape",
            "which dataset picked the gene list, <code>ours</code> or <code>yasmin</code>",
            "<code>genome-wide</code> for every gene in the pool, or the top list",
            "genes the row is computed on",
            "Spearman correlation between our effect and that of Yasmin et al",
            "bootstrap interval on <code>rho</code>",
            "percent of those genes both datasets move the same way"),
  stringsAsFactors = FALSE),
  rownames = FALSE, escape = FALSE,
  colnames = c("column", "what it holds"),
  options = list(dom = "t", ordering = FALSE, paging = FALSE))
```

The table carries every contrast crossed with both ranking sides.
`ranked_on = "ours"` is the original comparison. `ranked_on = "yasmin"` is the
reciprocal one: their contrast picks the genes and our four libraries only have
to follow. The reciprocal direction is the better powered of the two, since the
selection happens on the replicated side.

## A label permutation null

The numbers above need something to be measured against. A rho of 0.25 says
little on its own, because it is not obvious what rho an unrelated contrast
would have produced on the same genes.

The null answers that by making unrelated contrasts. Animals are put into groups
at random, the contrast is refitted on those scrambled labels, and the agreement
statistics are recomputed against our effect sizes, which never change. Doing
that many times shows the range the statistic covers when the grouping carries
no developmental information. `p_perm` is the share of those random contrasts
that matched our genes as well as the real one did, or better.

Our gene list is held fixed the whole time. Picking the top genes on our own
effect inflates the correlation, but it inflates the null by the same amount, so
the two stay comparable.

Animals are drawn from the whole 28-animal series rather than shuffled inside
one contrast, because a small contrast has too few rearrangements to give a
useful p-value. `r if (have_bulk) paste0("Reshuffling the male endpoint within itself allows only ", choose(length(runs_ud) + length(runs_mm), length(runs_ud)), " splits and the female endpoint ", choose(length(runs_ud) + length(runs_mf), length(runs_ud)), ", putting a floor of ", signif(1 / choose(length(runs_ud) + length(runs_mm), length(runs_ud)), 2), " and ", signif(1 / choose(length(runs_ud) + length(runs_mf), length(runs_ud)), 2), " under the p-value.") else ""`

Every contrast gets a null on the same terms. An endpoint takes a random set of
animals and splits it into groups of the observed sizes. A trend takes a random
set of the same size and gives it the observed score vector.

The trends use more animals than the endpoints, which is not the same as having
more power here. Contrasting the two extreme groups gives a larger effect than
fitting a line through the intermediate ones, and in this data the
endpoint contrasts reach FDR 0.05 for more genes than the trends do on both axes.

```{r, eval=have_bulk}
n_perm_labels = as.integer(params$n_perm_labels)
perm_cores    = max(1L, as.integer(params$cores))

# One set of refits serves every summary. The fit does not depend on the gene
# pool or on which side ranks the genes, so each permutation is scored for all
# four combinations rather than refitted four times.
#
# When Yasmin et al ranks the genes, the null re-selects inside each permutation from
# the permuted contrast. Reusing the real contrast's gene list would carry the
# real labels into the null.
perm_combos = expand.grid(pool = names(gene_pools),
                          rank_on = c("ours", "yasmin"),
                          stringsAsFactors = FALSE)

# The draws are generated here, in one stream, rather than inside the workers,
# so the result does not depend on how many cores are used. The stream is the
# same one the earlier serial version consumed — one sample() call per
# permutation, in order — so a cache written by that version stays valid.
all_runs = sample_meta$run

draw_runs = function(n_draw, n_perm, seed) {
  set.seed(seed)
  replicate(n_perm, sample(all_runs, n_draw), simplify = FALSE)
}

# scoring for one permutation: every pool and ranking side from the one fit
score_draw = function(lfc_all, ours_col, top_n) {
  dplyr::bind_rows(lapply(seq_len(nrow(perm_combos)), function(k) {
    keep = gene_pools[[perm_combos$pool[k]]]
    ov   = cmp[[ours_col]][keep]
    lv   = lfc_all[keep]
    idx  = if (perm_combos$rank_on[k] == "ours") top_both(ov, top_n)
           else top_both(lv, top_n)
    agree = sign(ov[idx]) == sign(lv[idx])
    data.frame(
      pool = perm_combos$pool[k], rank_on = perm_combos$rank_on[k],
      rho_gw   = suppressWarnings(cor(ov, lv, method = "spearman",
                                      use = "complete.obs")),
      rho_top  = suppressWarnings(cor(ov[idx], lv[idx], method = "spearman",
                                      use = "complete.obs")),
      pct_sign = 100 * mean(agree, na.rm = TRUE))
  }))
}

# The fits are independent, so they fan out across cores. mclapply forks, which
# is why the job script pins the BLAS threads to one: eight R processes each
# spawning threads would oversubscribe the node.
# eff_col differs by fit shape: run_two_group names its effect logFC and
# run_trend names its effect slope, so the caller says which to score.
perm_null = function(draws, fit_draw, ours_col, top_n, cores = 1,
                     eff_col = "perm_logFC") {
  work = function(d) {
    tab = try(fit_draw(d), silent = TRUE)
    if (inherits(tab, "try-error")) return(NULL)
    stopifnot(eff_col %in% names(tab))
    score_draw(tab[[eff_col]][match(cmp$Gene, tab$Gene)], ours_col, top_n)
  }
  res = if (cores > 1) parallel::mclapply(draws, work, mc.cores = cores)
        else lapply(draws, work)
  failed = vapply(res, function(x) is.null(x) || inherits(x, "try-error"),
                  logical(1))
  if (any(failed)) message(sum(failed), " of ", length(draws),
                           " permutations failed to fit and are dropped")
  dplyr::bind_rows(res[!failed])
}

two_group_perm = function(n_a, n_b, ours_col, seed) {
  perm_null(draw_runs(n_a + n_b, n_perm_labels, seed),
            function(d) run_two_group(y, d[seq_len(n_a)],
                                      d[n_a + seq_len(n_b)], "perm"),
            ours_col, top_ns[1], perm_cores)
}

trend_perm = function(score_vec, ours_col, seed) {
  perm_null(draw_runs(length(score_vec), n_perm_labels, seed),
            function(d) run_trend(y, d, score_vec, "perm"),
            ours_col, top_ns[1], perm_cores,
            eff_col = "perm_series_log2FC")
}

# cached on a signature, since the permutation is the slow step in the document
perm_file = paste0(Robject_out_dir, "/label_permutation_null.Rds")
perm_sig  = list(n_perm  = n_perm_labels,
                 top_n   = top_ns[1],
                 # which genes each permutation is scored on. Without this a
                 # change to the selection rule reuses a null built on a
                 # different gene set, and nothing says so.
                 top_rule = "top_both, n each way",
                 stat    = "spearman rho and percent same sign",
                 n_genes = nrow(cmp),
                 n_well  = sum(cmp$well_measured),
                 min_pb  = min_pb,
                 combos  = perm_combos,
                 scores  = list(male = male_score, female = fem_score),
                 runs    = sort(sample_meta$run),
                 sizes   = c(length(runs_ud), length(runs_mm),
                             length(runs_mf), length(male_runs),
                             length(fem_runs)),
                 shapes  = c("endpoints", "trends"))

cached = if (file.exists(perm_file)) readRDS(perm_file) else NULL

if (!is.null(cached) && identical(cached$signature, perm_sig)) {
  perm_res = cached$perm_res
} else {
  perm_res = list(
    "male, endpoint"   = two_group_perm(length(runs_ud), length(runs_mm),
                                        "ours_male", 1),
    "female, endpoint" = two_group_perm(length(runs_ud), length(runs_mf),
                                        "ours_female", 2),
    "male, trend"      = trend_perm(male_score, "ours_male", 3),
    "female, trend"    = trend_perm(fem_score, "ours_female", 4))

  saveRDS(list(signature = perm_sig, perm_res = perm_res), file = perm_file)
}
```

```{r, eval=have_bulk}
# one-sided: the question is whether the observed statistic sits above the null
emp_p = function(null_vals, obs) {
  ok = is.finite(null_vals)
  (1 + sum(null_vals[ok] >= obs)) / (1 + sum(ok))
}

perm_tab = dplyr::bind_rows(lapply(names(perm_res), function(ax) {
  nd = perm_res[[ax]]
  dplyr::bind_rows(lapply(split(nd, list(nd$pool, nd$rank_on), drop = TRUE),
    function(g) {
      pl  = g$pool[1]
      ro  = g$rank_on[1]
      obs = agreement[agreement$axis == ax & agreement$pool == pl &
                        agreement$ranked_on == ro, ]
      o   = c(obs$rho[obs$set == "genome-wide"],
              obs$rho[obs$set == top_set_lab],
              obs$pct_same_sign[obs$set == top_set_lab])
      nl  = list(g$rho_gw, g$rho_top, g$pct_sign)

      out = data.frame(
        axis        = ax,
        pool        = pl,
        ranked_on   = ro,
        statistic   = c("rho, genome-wide",
                        paste0("rho, ", top_set_lab),
                        paste0("% same sign, ", top_set_lab)),
        observed    = o,
        null_median = vapply(nl, median, numeric(1), na.rm = TRUE),
        null_95th   = vapply(nl,
                             function(v) unname(quantile(v, .95, na.rm = TRUE)),
                             numeric(1)),
        p_perm      = mapply(emp_p, nl, o),
        n_perm      = vapply(nl, function(v) sum(is.finite(v)), integer(1)))

      # the genome-wide row does not depend on which side ranks the genes
      if (ro == "yasmin") out = out[out$statistic != "rho, genome-wide", ]
      out
    }))
}))

write.csv(perm_tab, file = paste0(results_dir, "/bulk_permutation_null.csv"),
          row.names = FALSE)

datatable(perm_tab %>% dplyr::select(-pool) %>%
            mutate(across(where(is.numeric), ~signif(., 3))),
          rownames = FALSE, filter = "top", options = list(pageLength = 20))
```

```{r, eval=have_bulk}
datatable(data.frame(
  column = c("<code>axis</code>, <code>ranked_on</code>", "<code>statistic</code>", "<code>observed</code>", "<code>null_median</code>", "<code>null_95th</code>", "<code>p_perm</code>", "<code>n_perm</code>"),
  holds  = c("as in the agreement table above",
            "which of the three agreement measures the row tests",
            "its value on Yasmin et al real contrast",
            "its median over the relabeled contrasts",
            "its 95th percentile over them, the value to beat",
            "fraction of relabelings reaching the observed value or better",
            "relabelings that fitted"),
  stringsAsFactors = FALSE),
  rownames = FALSE, escape = FALSE,
  colnames = c("column", "what it holds"),
  options = list(dom = "t", ordering = FALSE, paging = FALSE))
```

`r if (have_bulk) paste0("The smallest p-value reachable with ", n_perm_labels, " permutations is ", signif(1 / (1 + n_perm_labels), 2), ". A statistic at that value is reported as a floor rather than as its true p-value.") else ""`

```{r, eval=have_bulk, fig.width=15, fig.height=7}
# the well measured pool only, since the unfiltered ranking is the one under
# suspicion
perm_long = dplyr::bind_rows(lapply(names(perm_res), function(ax) {
  nd = perm_res[[ax]][perm_res[[ax]]$pool == "well measured", ]
  data.frame(axis = ax, ranked_on = nd$rank_on, rho_top = nd$rho_top)
}))

obs_long = dplyr::bind_rows(lapply(names(perm_res), function(ax)
  dplyr::bind_rows(lapply(c("ours", "yasmin"), function(ro) {
    obs = agreement[agreement$axis == ax & agreement$pool == "well measured" &
                      agreement$ranked_on == ro, ]
    data.frame(axis = ax, ranked_on = ro,
               observed = obs$rho[obs$set == top_set_lab])
  }))))

p_perm = ggplot(perm_long, aes(x = rho_top)) +
  geom_histogram(bins = 40, fill = "grey75", colour = NA) +
  geom_vline(data = obs_long, aes(xintercept = observed),
             colour = "#c1272d", linewidth = .7) +
  facet_grid(ranked_on ~ axis, scales = "free_y") +
  labs(x = paste0("Spearman rho across the ", top_set_lab, " genes"),
       y = "permutations",
       title = "Observed agreement against random contrasts of the same shape",
       subtitle = "well measured genes; rows give the side that ranked the genes") +
  theme_bw()

p_perm

ggsave(p_perm, file = paste0(results_dir, "/bulk_permutation_null.png"),
       width = 15, height = 7, dpi = 300)
```

## Agreement among our top genes

The scatters take our top 100 genes each way on a sex's axis and plot our effect
against that of Yasmin et al, with points colored by whether the two datasets put the gene in
the same direction. Each panel carries its Spearman correlation and the
permutation p-value for that correlation.

The endpoints and the trends get their own figure, because their y axes are
different quantities. Putting them on shared panels would invite reading one
against the other.

The endpoint panels compare two groups, so their y value is an ordinary log fold
change: 1 means the gene sits twice as high in the mid group as in the larvae.

The trend panels come from a fitted line. Each animal carries a trend score, a
number that puts it in developmental order: the undifferentiated larvae at 0,
the next group at 1, the one after at 2, with the animals inside a group spread
over the fraction between. Expression is regressed on that score, which gives a
slope. The slope on its own says the change for a step of 1, and a step of 1 is
not a fixed amount of development, so the number cannot be read without also
knowing how far the score runs.

The y value is therefore the slope read across the span the score covers, which
is the log2 change from the first animal in the series to the last. `r if (!have_bulk) "" else paste0("The male score runs 0 to ", signif(max(male_score), 3), " and the female score 0 to ", signif(max(fem_score), 3), ".")`
A value of 1 means the gene doubles over the whole series. Both figures are then
log2 fold changes, and what still differs is the comparison: the endpoints
contrast two groups of animals directly, while the trends read the change off a
line through every animal.

Three columns in the export cover this, because the slope and the rescaling
answer different questions. `male_end_logFC` and `fem_end_logFC` are fold
changes between two groups. `male_trend_slope` and `fem_trend_slope` are the
coefficients edgeR returned, kept so that refitting reproduces this file.
`male_trend_series_log2FC` and `fem_trend_series_log2FC` are those slopes across
the span, which is what the figures and every statistic use.

The rescaling changes no result. Spearman rho, percent same sign, the top-gene
rankings and the permutation are all ranks or signs, so multiplying one column
by a positive constant leaves every one of them identical.

```{r, eval=have_bulk}
# our effect picks the genes. The selection made on the Yasmin et al side is in
  # the tables.
rank_sides = "ours"

# Okabe-Ito blue and vermillion. A red-green pair is the one contrast a red-green
# colorblind reader cannot make, and it is the default reach for agree/disagree.
agree_cols = c(`TRUE` = "#0072B2", `FALSE` = "#D55E00")

# Built once and called per contrast shape, so the two figures cannot drift apart.
# The size arguments carry the defaults this section draws at. The composite
# calls the same function again with smaller type, because at journal width a
# facet is about three inches and the annotation set in ten point runs off it.
agreement_figure = function(axes, ylab, what, base_size = 11, pt_size = 1.6,
                            lab_size = 2.4, ann_size = 3.1, n_lab = 10) {
  panel_grid = expand.grid(axis = axes, ranked_on = rank_sides,
                           stringsAsFactors = FALSE)

  top = dplyr::bind_rows(Map(function(ax, ro) {
    cols = axis_cols[[ax]]
    d = cmp[cmp$well_measured, ] %>%
      dplyr::select(Gene, label, ours = dplyr::all_of(cols[["ours"]]),
                    yasmin = dplyr::all_of(cols[["yas"]])) %>%
      dplyr::filter(is.finite(ours), is.finite(yasmin))
    rank_vec = if (ro == "ours") d$ours else d$yasmin
    # the same balanced set the agreement table scores, so the rho and p printed
    # on each panel describe the points that are drawn
    d = d[top_both(rank_vec, top_ns[1]), , drop = FALSE]
    d$axis      = ax
    d$ranked_on = ro
    d$agree     = sign(d$ours) == sign(d$yasmin)
    d
  }, panel_grid$axis, panel_grid$ranked_on))

  top$axis      = factor(top$axis, levels = axes)
  top$ranked_on = factor(top$ranked_on, levels = rank_sides,
                         labels = "genes picked by our effect")

  # rho from the agreement table, so the panel carries the number the table
  # reports rather than a second calculation of it
  rho_tab = agreement %>%
    dplyr::filter(set == top_set_lab) %>%
    dplyr::select(axis, ranked_on, rho)

  # Direction agreement, tested per direction against the rate in the Yasmin et al et
  # al contrast as a whole. Those contrasts are not balanced, so the share of
  # genes moving each way there is the background the matching half of our set
  # has to beat, and it is not 50 percent. Up and down are kept apart because the two backgrounds
  # are complements and a combined count would cancel them.
  pool = vapply(axes, function(ax) {
    v = cmp[cmp$well_measured, ][[axis_cols[[ax]][["yas"]]]]
    c(n = sum(is.finite(v)), up = sum(v > 0, na.rm = TRUE))
  }, numeric(2))

  dir_ann = top %>%
    dplyr::mutate(our_dir = ifelse(ours > 0, "up", "down")) %>%
    dplyr::group_by(axis, ranked_on, our_dir) %>%
    dplyr::summarise(n = dplyr::n(), k = sum(agree), .groups = "drop") %>%
    dplyr::mutate(
      N = pool["n", as.character(axis)],
      K = ifelse(our_dir == "up", pool["up", as.character(axis)],
                 N - pool["up", as.character(axis)]),
      # Hypergeometric: our n genes are a draw from the pool without
      # replacement, and K of the pool move the way we do, so K/N is the
      # background. One sided, in the direction of agreement, so a half of the
      # set that departs the other way reads as no evidence rather than as a
      # small p-value.
      p_dir = stats::phyper(k - 1, K, N - K, n, lower.tail = FALSE)) %>%
    dplyr::arrange(axis, dplyr::desc(our_dir))

  panel_ann = dir_ann %>%
    dplyr::group_by(axis, ranked_on) %>%
    dplyr::summarise(dir_txt = paste0(our_dir, " ", k, "/", n, " agree, ",
                                      round(100 * K / N), "% by chance, ",
                                      "p = ", signif(p_dir, 2),
                                      collapse = "\n"), .groups = "drop") %>%
    dplyr::mutate(rank_key = ifelse(grepl("our effect", ranked_on),
                                    "ours", "yasmin")) %>%
    dplyr::left_join(rho_tab, by = c("axis" = "axis", "rank_key" = "ranked_on")) %>%
    dplyr::mutate(txt = paste0("rho = ", signif(rho, 2), "\n", dir_txt))

  # the joins above coerce the factor columns to character, which would let
  # ggplot order the facets alphabetically instead of by the levels set on top
  panel_ann$axis      = factor(panel_ann$axis, levels = levels(top$axis))
  panel_ann$ranked_on = factor(panel_ann$ranked_on, levels = levels(top$ranked_on))

  p_top = ggplot(top, aes(x = ours, y = yasmin, colour = agree)) +
    geom_hline(yintercept = 0, colour = "grey70") +
    geom_vline(xintercept = 0, colour = "grey70") +
    geom_point(size = pt_size, alpha = .85) +
    geom_text_repel(data = top %>%
                      dplyr::group_by(axis, ranked_on) %>%
                      dplyr::slice_max(abs(ours), n = n_lab, with_ties = FALSE),
                    # the labels inherit the agree colour otherwise, which puts
                    # text in the same two colours as the points it sits among.
                    # A dark label on a white halo reads over either colour.
                    aes(label = label), colour = "grey15",
                    segment.colour = "grey65", bg.color = "white", bg.r = .12,
                    size = lab_size, max.overlaps = 22, show.legend = FALSE) +
    geom_label(data = panel_ann, aes(x = -Inf, y = Inf, label = txt),
               hjust = -0.05, vjust = 1.1, inherit.aes = FALSE, size = ann_size,
               label.size = 0, fill = "white", alpha = .75, lineheight = 1.05) +
    scale_colour_manual(values = agree_cols, name = NULL,
                        breaks = c("TRUE", "FALSE"),
                        labels = c(`TRUE` = "same direction",
                                   `FALSE` = "opposite")) +
    facet_wrap(~ axis, scales = "free", nrow = 1) +
    labs(x = "our pseudobulk log2FC, larva to transformer", y = ylab,
         title = paste0("Our top ", top_ns[1], " genes each way against the ",
                        what),
         subtitle = paste0("p is hypergeometric and one sided, our genes ",
                           "moving the same way against the rate among all genes")) +
    theme_bw(base_size = base_size) + theme(legend.position = "top")

  p_top
}
```

```{r, eval=have_bulk, fig.width=12, fig.height=5.5}
p_end = agreement_figure(c("male, endpoint", "female, endpoint"),
                         "Yasmin et al log2FC, larvae to mid group",
                         "Yasmin et al endpoint")
p_end
ggsave(p_end, file = paste0(results_dir, "/bulk_endpoint_agreement.png"),
       width = 12, height = 5.5, dpi = 300)
```

```{r, eval=have_bulk, fig.width=12, fig.height=5.5}
p_trend = agreement_figure(c("male, trend", "female, trend"),
                           "Yasmin et al log2FC across the series",
                           "Yasmin et al trend")
p_trend
ggsave(p_trend, file = paste0(results_dir, "/bulk_trend_agreement.png"),
       width = 12, height = 5.5, dpi = 300)
```

`r if (have_bulk) paste0("Points are our ", top_set_lab, " genes, colored by whether the two datasets agree on direction. Each panel prints the rho from the agreement table, then one line per direction: how many of our hundred Yasmin et al move the same way, how many would do so by chance given the share of all genes moving that way in that contrast, and a hypergeometric p-value. The chance share is not 50 percent, because their contrasts are not balanced. The test is one sided in the direction of agreement, so a half of the set that departs the other way returns a p-value near 1.") else ""`


## Genome-wide scatter

Every gene measurable on both sides, our effect against that of Yasmin et al, with the
fifteen largest of ours labeled in each panel.

The y axes carry the same two quantities described above: a log fold change
between two groups in the endpoint row, and the fitted change across the whole
series in the trend row. All four panels here sit in one figure, so the rows are
not comparable with each other even though both are on a log2 scale.

```{r, eval=have_bulk, fig.width=11, fig.height=11, dev="png", dpi=600}
# One point per measurable gene in each of four panels, so this one stays
# raster. The rest of the figures in this document are vector.
# Both rows are log2 fold changes, but of different comparisons, so the label
# says which rather than leaving them looking interchangeable.
yl_end   = "Yasmin et al log2FC, larvae to mid group"
yl_trend = "Yasmin et al log2FC across the series"

scatter_for = function(ours_col, yas_col, title, ylab, n_label = 15) {
  d = cmp %>%
    dplyr::select(Gene, label, ours = dplyr::all_of(ours_col),
                  yasmin = dplyr::all_of(yas_col)) %>%
    dplyr::filter(is.finite(ours), is.finite(yasmin))
  lab = d %>% dplyr::slice_max(abs(ours), n = n_label, with_ties = FALSE)

  ggplot(d, aes(x = ours, y = yasmin)) +
    geom_hline(yintercept = 0, colour = "grey80") +
    geom_vline(xintercept = 0, colour = "grey80") +
    geom_point(alpha = .2, size = .5) +
    geom_point(data = lab, colour = "#c1272d", size = 1.2) +
    geom_text_repel(data = lab, aes(label = label), colour = "grey15",
                    segment.colour = "grey65", bg.color = "white", bg.r = .12,
                    size = 2.5, max.overlaps = 20) +
    geom_smooth(method = "lm", se = FALSE, linewidth = .5, colour = "#004488") +
    labs(x = "our pseudobulk log2FC, larva to transformer",
         y = ylab, title = title) +
    theme_bw()
}

p_male  = scatter_for("ours_male",   "male_end_logFC",   "Male endpoint",   yl_end)
p_fem   = scatter_for("ours_female", "fem_end_logFC",    "Female endpoint", yl_end)
p_mtr   = scatter_for("ours_male",   "male_trend_series_log2FC",
                      "Male trend",   yl_trend)
p_ftr   = scatter_for("ours_female", "fem_trend_series_log2FC",
                      "Female trend", yl_trend)

p_scatter = (p_male + p_fem) / (p_mtr + p_ftr)
p_scatter

ggsave(p_scatter, file = paste0(results_dir, "/bulk_agreement_scatter.png"),
       width = 11, height = 11, dpi = 300)
```

Every gene is plotted here, not only the top hundred, with the fifteen strongest
labeled. The line is an ordinary least squares fit shown for orientation.

# Top gene lists

The hundred most up- and most down-regulated genes from each side, taken against
the genes measurable in both datasets. Our lists are the same for both contrasts
on a sex, since our effect does not depend on which Yasmin et al contrast it is
compared with. The Yasmin et al lists are built per contrast. The GO section below tests these
lists, and the overlap below intersects them.

```{r, eval=have_bulk}
# The background is the genes measurable on both sides, which is the set the
# lists were drawn from. Up and down are taken separately. A term enriched among
# rising genes says something different from one enriched among falling genes.
go_pool = cmp[cmp$well_measured, , drop = FALSE]

# Each set carries the source, sex and direction it came from, so terms can be
# matched across sources later without parsing the label back apart.
set_defs = c(
  list(list(label = "ours, male",   col = "ours_male",
            source = "ours", sex = "male",   contrast = NA_character_),
       list(label = "ours, female", col = "ours_female",
            source = "ours", sex = "female", contrast = NA_character_)),
  lapply(names(axis_cols), function(ax) {
    parts = trimws(strsplit(ax, ",")[[1]])
    list(label = paste0("Yasmin et al, ", ax), col = axis_cols[[ax]][["yas"]],
         source = "Yasmin et al", sex = parts[1], contrast = parts[2])
  }))

set_rows = lapply(set_defs, function(d)
  lapply(c("up", "down"), function(dir) {
    idx = top_dir(go_pool[[d$col]], dir)
    list(genes = data.frame(Gene = go_pool$Gene[idx],
                            set  = paste0(d$label, ", ", dir),
                            stringsAsFactors = FALSE),
         # col is carried so the overlap below can recover which genes each
         # side moves in each direction, not just which made its top hundred
         key   = data.frame(set = paste0(d$label, ", ", dir),
                            source = d$source, sex = d$sex,
                            contrast = d$contrast, direction = dir,
                            col = d$col,
                            stringsAsFactors = FALSE))
  }))

set_rows = unlist(set_rows, recursive = FALSE)
go_sets  = dplyr::bind_rows(lapply(set_rows, `[[`, "genes"))
set_key  = dplyr::bind_rows(lapply(set_rows, `[[`, "key"))

```

`r if (!have_bulk) "" else { n_per = go_sets %>% dplyr::count(set, name = "n"); short = n_per %>% dplyr::filter(n < top_ns[1]); paste0(nrow(n_per), " lists, ", if (nrow(short) == 0) paste0("each holding ", top_ns[1], " genes.") else paste0("holding ", top_ns[1], " genes each except ", paste0(short$set, " (", short$n, ")", collapse = "; "), ", where fewer genes move in that direction.")) }`

## Enrichment of our top genes in the Yasmin et al ranking

The Yasmin et al genes are ranked once, from most up to most down, and the
question is whether our list sits toward one end of that ranking rather than
spread through it.

Our top hundred up and our top hundred down are separate gene sets, and each is
tested on its own. A set of genes we call up should land at the top of the
ranking, giving a positive enrichment score, and a set we call down should land
at the bottom, giving a negative one.

The ranking is the signed test statistic, the root of the quasi-likelihood F
with the sign of the fold change.

```{r}
have_fgsea = have_bulk && requireNamespace("fgsea", quietly = TRUE)
```

`r if (have_bulk && !have_fgsea) "**Enrichment is skipped.** The fgsea package is not installed." else ""`

```{r, eval=have_fgsea}
our_sets = split(go_sets$Gene, go_sets$set)[
  set_key$set[set_key$source == "ours"]]

# fgsea once per contrast, reused for both the table and the leading edge
set.seed(1)
gsea_raw = lapply(names(axis_cols), function(ax) {
  sex = trimws(strsplit(ax, ",")[[1]])[1]
  # only this sex's sets, since our effect is computed per sex
  use = our_sets[set_key$set[set_key$source == "ours" & set_key$sex == sex]]

  rank_vec = go_pool[[axis_cols[[ax]][["stat"]]]]
  names(rank_vec) = go_pool$Gene
  rank_vec = rank_vec[is.finite(rank_vec)]

  # BPPARAM rather than nproc: with nproc the multilevel step draws a progress
  # bar straight to the console, which lands in the knitted output and is not
  # caught by the chunk's message or warning options.
  list(ax = ax, rank_vec = rank_vec,
       res = fgsea::fgsea(use, rank_vec, minSize = 10,
                          BPPARAM = BiocParallel::SerialParam(progressbar = FALSE)))
})

gsea_tab = dplyr::bind_rows(lapply(gsea_raw, function(z) {
  r = z$res
  data.frame(contrast = z$ax, set = r$pathway, size = r$size,
             ES = r$ES, NES = r$NES, p = r$pval, fdr = r$padj,
             leading_edge = vapply(r$leadingEdge, length, integer(1)),
             stringsAsFactors = FALSE) %>%
    dplyr::mutate(direction = ifelse(grepl("up$", set), "up", "down"),
                  as_expected = (direction == "up" & NES > 0) |
                                (direction == "down" & NES < 0))
}))

write.csv(gsea_tab, file = paste0(results_dir, "/top_gene_enrichment.csv"),
          row.names = FALSE)

# the leading edge genes, which are the ones carrying each enrichment
gsea_genes = dplyr::bind_rows(lapply(gsea_raw, function(z) {
  r = z$res
  dplyr::bind_rows(lapply(seq_len(nrow(r)), function(i) {
    g = r$leadingEdge[[i]]
    if (length(g) == 0) return(NULL)
    j = match(g, go_pool$Gene)
    data.frame(contrast = z$ax, set = r$pathway[i], Gene = g,
               label = go_pool$label[j],
               ours = go_pool[[axis_cols[[z$ax]][["ours"]]]][j],
               yasmin_stat = unname(z$rank_vec[g]),
               stringsAsFactors = FALSE)
  }))
}))

write.csv(gsea_genes,
          file = paste0(results_dir, "/top_gene_enrichment_leading_edge.csv"),
          row.names = FALSE)

datatable(gsea_tab %>%
            dplyr::select(contrast, set, size, NES, p, fdr, leading_edge,
                          as_expected) %>%
            mutate(across(where(is.numeric), ~signif(., 3))),
          rownames = FALSE, filter = "top", options = list(pageLength = 10))
```

```{r, eval=have_fgsea}
datatable(data.frame(
  column = c("<code>contrast</code>", "<code>set</code>", "<code>size</code>", "<code>ES</code>", "<code>NES</code>", "<code>p</code>, <code>fdr</code>", "<code>leading_edge</code>", "<code>as_expected</code>"),
  holds  = c("which Yasmin et al contrast supplied the ranking",
            "which of our lists was tested",
            "genes of the set found in the ranking",
            "enrichment score, how far toward one end the set sits",
            "that score scaled by its null, so sets are comparable",
            "permutation p-value and its Benjamini-Hochberg adjustment",
            "genes of the set before the score peaks, which carry the signal",
            "an up set with a positive score, or a down set with a negative one"),
  stringsAsFactors = FALSE),
  rownames = FALSE, escape = FALSE,
  colnames = c("column", "what it holds"),
  options = list(dom = "t", ordering = FALSE, paging = FALSE))
```

`r if (!have_fgsea) "" else { ok = gsea_tab %>% dplyr::filter(as_expected, fdr < 0.05); up = gsea_tab %>% dplyr::filter(direction == "up"); dn = gsea_tab %>% dplyr::filter(direction == "down"); paste0(nrow(ok), " of the ", nrow(gsea_tab), " tests enrich in the expected direction at FDR < 0.05. The up sets score ", paste(range(signif(up$NES, 2)), collapse = " to "), " and the down sets ", paste(range(signif(dn$NES, 2)), collapse = " to "), ". A set that enriches the wrong way, or not at all, is listed with as_expected FALSE. The genes behind each score are in top_gene_enrichment_leading_edge.csv.") }`

```{r, eval=have_fgsea, fig.width=10, fig.height=7}
# The running enrichment score. The ranked genes are laid out left to right from
# up to most down. The walk steps up at each gene of our set, in proportion to
# its statistic, and down at every other gene, so a set concentrated at one end
# drives the line away from zero and a set spread evenly leaves it flat. The
# peak is the enrichment score fgsea reports, and this reproduces it exactly.
running_es = function(stat, genes) {
  s   = sort(stat, decreasing = TRUE)
  hit = names(s) %in% genes
  w   = abs(s[hit])
  inc = numeric(length(s))
  inc[hit]  = w / sum(w)
  inc[!hit] = -1 / sum(!hit)
  cumsum(inc)
}

dir_of   = function(x) factor(ifelse(grepl("up$", x), "up", "down"),
                              levels = c("up", "down"))
dir_cols = c(up = "#0072B2", down = "#E69F00")

curve_df = dplyr::bind_rows(lapply(gsea_raw, function(z)
  dplyr::bind_rows(lapply(z$res$pathway, function(st) {
    es = running_es(z$rank_vec, our_sets[[st]])
    # thinned for drawing, since an svg of seventeen thousand points is large
    # and the line is smooth; the peak and the ends are kept exactly
    step = max(1L, as.integer(length(es) / 600))
    keep = sort(unique(c(1L, seq(1L, length(es), by = step),
                         which.max(abs(es)), length(es))))
    data.frame(contrast = z$ax, set = st, rank = keep, es = es[keep],
               stringsAsFactors = FALSE)
  }))))

tick_df = dplyr::bind_rows(lapply(gsea_raw, function(z) {
  ord = names(sort(z$rank_vec, decreasing = TRUE))
  dplyr::bind_rows(lapply(z$res$pathway, function(st)
    data.frame(contrast = z$ax, set = st,
               rank = match(our_sets[[st]], ord),
               stringsAsFactors = FALSE)))
})) %>% dplyr::filter(!is.na(rank))

peak_df = curve_df %>%
  dplyr::group_by(contrast, set) %>%
  dplyr::slice_max(abs(es), n = 1, with_ties = FALSE) %>%
  dplyr::ungroup()

# the barcode sits below the curves, one row per set, scaled to the data so it
# never overlaps the lines
lo = min(curve_df$es); hi = max(curve_df$es); sp = hi - lo
bar = tick_df %>%
  dplyr::mutate(direction = dir_of(set),
                y0 = ifelse(direction == "up", lo - .10 * sp, lo - .22 * sp),
                y1 = ifelse(direction == "up", lo - .18 * sp, lo - .30 * sp))

lab = gsea_tab %>%
  dplyr::mutate(direction = dir_of(set),
                txt = paste0(direction, ": NES ", signif(NES, 2),
                             ", FDR ", signif(fdr, 2)),
                y = ifelse(direction == "up", hi + .22 * sp, hi + .10 * sp),
                contrast = factor(contrast, levels = names(axis_cols)))

# Type and mark sizes are arguments so the same figure can be drawn again at
# journal size for the composite. The defaults are the sizes this section uses.
gsea_figure = function(base_size = 10, pt_size = 2.2, txt_size = 2.9,
                       line_w = .8) {
  ggplot(curve_df %>%
           dplyr::mutate(direction = dir_of(set),
                         contrast = factor(contrast,
                                           levels = names(axis_cols))),
         aes(x = rank, y = es, colour = direction)) +
    geom_hline(yintercept = 0, colour = "grey75", linewidth = .4) +
    geom_line(aes(group = set), linewidth = line_w) +
    geom_point(data = peak_df %>%
                 dplyr::mutate(direction = dir_of(set),
                               contrast = factor(contrast,
                                                 levels = names(axis_cols))),
               size = pt_size) +
    geom_segment(data = bar %>%
                   dplyr::mutate(contrast = factor(contrast,
                                                   levels = names(axis_cols))),
                 aes(x = rank, xend = rank, y = y0, yend = y1,
                     colour = direction),
                 linewidth = .25, alpha = .55, inherit.aes = FALSE) +
    geom_text(data = lab, aes(x = 0, y = y, label = txt), hjust = 0,
              size = txt_size, show.legend = FALSE) +
    scale_colour_manual(values = dir_cols, name = "our genes",
                        breaks = c("up", "down"),
                        labels = c(up = "up in transformers",
                                   down = "down in transformers")) +
    scale_x_continuous(labels = function(x) format(x, big.mark = ","),
                       expand = ggplot2::expansion(mult = c(.01, .02))) +
    facet_wrap(~ contrast, ncol = 2) +
    labs(x = "Yasmin et al genes, ranked from most up to most down",
         y = "running enrichment score",
         title = "Our top genes against the Yasmin et al ranking",
         subtitle = paste0("each line walks the ranking and steps up at our ",
                           "genes; ticks mark where they fall")) +
    theme_bw(base_size = base_size) +
    theme(legend.position = "bottom",
          panel.grid.minor = element_blank(),
          strip.background = element_rect(fill = "grey95", colour = NA))
}

p_gsea = gsea_figure()
p_gsea

ggsave(p_gsea, file = paste0(results_dir, "/top_gene_enrichment.png"),
       width = 10, height = 7, dpi = 300)
```

## Overlap between the lists

The same lists, counted rather than ranked. A gene in our top hundred and in
theirs is a stricter statement than enrichment, which asks only that our genes
sit toward one end of the ranking.

```{r, eval=have_bulk, fig.width=12, fig.height=8}
# Hand-built rather than UpSetR or ComplexUpset, so the document takes no new
# dependency and the figure comes out as vector like the rest. Set sizes go in
# the y axis labels instead of a third aligned panel.
upset_contrast = "trend"   # the Yasmin et al contrast the figure shows
n_int          = 20        # intersections drawn, largest first

fig_keys = set_key %>%
  dplyr::filter(source == "ours" | contrast == upset_contrast) %>%
  dplyr::arrange(source, sex, direction)

all_lists = split(go_sets$Gene, go_sets$set)
fig_sets  = fig_keys$set
set_lists = all_lists[fig_sets]
all_genes = sort(unique(unlist(set_lists)))

# one column per set, in fig_sets order, so the bit strings below line up with it
mem = vapply(set_lists, function(g) all_genes %in% g,
             logical(length(all_genes)))

int_all = data.frame(pattern = apply(mem, 1, function(r)
                       paste(as.integer(r), collapse = "")),
                     stringsAsFactors = FALSE) %>%
  dplyr::count(pattern, name = "size") %>%
  dplyr::arrange(dplyr::desc(size))

# only the largest are drawn, so the drawn columns need not cover every gene
int_sizes = int_all %>%
  dplyr::slice_head(n = n_int) %>%
  dplyr::mutate(x = dplyr::row_number())

bits = strsplit(int_sizes$pattern, "")
mem_long = dplyr::bind_rows(lapply(seq_len(nrow(int_sizes)), function(i)
  data.frame(x = int_sizes$x[i], set = fig_sets,
             member = bits[[i]] == "1", stringsAsFactors = FALSE)))

set_lab = setNames(paste0(fig_sets, "  (",
                          vapply(set_lists, length, integer(1)), ")"),
                   fig_sets)

x_scale = scale_x_continuous(breaks = int_sizes$x,
                             expand = ggplot2::expansion(add = .6))

p_ubar = ggplot(int_sizes, aes(x = x, y = size)) +
  geom_col(fill = "#0072B2", width = .7) +
  geom_text(aes(label = size), vjust = -0.35, size = 2.8) +
  x_scale +
  scale_y_continuous(expand = ggplot2::expansion(mult = c(0, .14))) +
  labs(x = NULL, y = "genes in the intersection",
       title = paste0("Top ", top_ns[1], " genes each way, ours and the Yasmin et al ",
                      upset_contrast),
       subtitle = paste0("the ", nrow(int_sizes),
                         " largest intersections; set sizes in brackets")) +
  theme_bw(base_size = 10) +
  theme(axis.text.x = element_blank(), axis.ticks.x = element_blank(),
        panel.grid.major.x = element_blank())

p_umat = ggplot(mem_long, aes(x = x, y = set)) +
  geom_point(colour = "grey88", size = 3) +
  geom_line(data = dplyr::filter(mem_long, member), aes(group = x),
            colour = "#0072B2", linewidth = .9) +
  geom_point(data = dplyr::filter(mem_long, member),
             colour = "#0072B2", size = 3) +
  x_scale +
  scale_y_discrete(limits = rev(fig_sets),
                   labels = function(z) unname(set_lab[z])) +
  labs(x = NULL, y = NULL) +
  theme_bw(base_size = 10) +
  theme(axis.text.x = element_blank(), axis.ticks.x = element_blank(),
        panel.grid.major.x = element_blank())

p_upset = p_ubar / p_umat + patchwork::plot_layout(heights = c(1, 1.15))

p_upset

ggsave(p_upset, file = paste0(results_dir, "/top_gene_list_upset.png"),
       width = 12, height = 8, dpi = 300)
```

`r if (!have_bulk) "" else paste0("Each column is the set of genes sharing one membership pattern, so the columns are disjoint rather than nested. ", format(length(all_genes), big.mark = ","), " genes appear in at least one of the ", length(fig_sets), " lists, spread over ", nrow(int_all), " patterns; the ", nrow(int_sizes), " drawn here hold ", format(sum(int_sizes$size), big.mark = ","), " of them. A column with one filled dot is genes unique to that list. The columns worth reading are the two-dot ones joining one of our lists to a Yasmin et al list on the same sex and in the same direction.")`



# Gene ontology

The lists built above are tested here for GO enrichment against the genes
measurable in both datasets.

Terms are kept when they carry more than 10 and fewer than 200 genes. Tiny terms
are noise and very large ones restate the whole ontology.

```{r}
gaf_file      = file.path(params$genome_dir,
                          "GCF_010993605.1_kPetMar1.pri_gene_ontology.gaf")
go_names_file = file.path(params$genome_dir, "GOs_names.txt")
have_go       = have_bulk && file.exists(gaf_file) && file.exists(go_names_file)
```

`r if (have_bulk && !have_go) paste0("**GO enrichment is skipped.** The GO files were not found under ", params$genome_dir, ".") else ""`

```{r, eval=have_go}
gaf = read.delim(gaf_file, header = FALSE, comment = "!") %>%
  dplyr::filter(!grepl("NOT", V4)) %>%
  dplyr::select(V3, V5) %>%
  dplyr::rename(GOID = "V5", Gene = "V3") %>%
  dplyr::distinct()

keep_go = gaf %>%
  dplyr::count(GOID) %>%
  dplyr::filter(n > 10, n < 200) %>%
  dplyr::pull(GOID)
gaf = gaf %>% dplyr::filter(GOID %in% keep_go)

go_names = read.delim(go_names_file, header = FALSE)
colnames(go_names) = c("GOID", "GOName")
```


```{r, eval=have_go}
go_res = run_go_enrichment(
  gaf              = gaf,
  gene_description = gene_description,
  go_names         = go_names,
  test_genes       = go_sets,
  background_n     = nrow(go_pool),
  group2_label     = "set",
  ngroup2_label    = "nSetGenes",
  overlaps_file    = paste0(results_dir, "/top_gene_GO_overlaps.csv"),
  pvals_file       = paste0(results_dir, "/top_gene_GO_pvalues.csv"),
  pval_filter      = params$go_pval,
  require_shared   = TRUE)

# Mark the terms that come up on both sides of the same comparison. A partner is
# a set from the other source with the same sex and the same direction, so
# "Yasmin et al, male, trend, up" pairs with "ours, male, up". Our sets are not split
# by contrast, so one of ours can partner both Yasmin et al contrasts on its sex.
go_res = go_res %>% dplyr::left_join(set_key, by = "set")

partners = go_res %>%
  dplyr::select(GOID, set, source, sex, direction) %>%
  dplyr::inner_join(go_res %>% dplyr::select(GOID, set, source, sex, direction),
                    by = c("GOID", "sex", "direction"),
                    suffix = c("", "_partner"),
                    relationship = "many-to-many") %>%
  dplyr::filter(source != source_partner) %>%
  dplyr::group_by(GOID, set) %>%
  dplyr::summarise(shared_with = paste(sort(unique(set_partner)),
                                       collapse = "; "), .groups = "drop")

go_res = go_res %>%
  dplyr::left_join(partners, by = c("GOID", "set")) %>%
  dplyr::mutate(shared = !is.na(shared_with),
                shared_with = ifelse(is.na(shared_with), "", shared_with)) %>%
  dplyr::relocate(shared, shared_with, .after = set)

write.csv(go_res, file = paste0(results_dir, "/top_gene_GO_pvalues.csv"),
          row.names = FALSE)

datatable(go_res %>% mutate(across(where(is.numeric), ~signif(., 3))),
          filter = "top", rownames = FALSE)
```

`r if (have_go && exists("go_res")) paste0(format(nrow(go_res), big.mark = ","), " term-by-set results pass p < ", params$go_pval, " across ", dplyr::n_distinct(go_sets$set), " gene sets. The full table is in top_gene_GO_pvalues.csv and the genes behind each term in top_gene_GO_overlaps.csv. Read the FDR column rather than Pval where a term is named individually. The FDR is computed within each set.") else ""`

## Terms shared with the Yasmin et al trend

A term counts as shared when it is enriched in our top genes and in Yasmin et al
trend genes, on the same sex and in the same direction. The endpoint contrasts
are left out here, so a term that pairs only with Yasmin et al endpoint does not
appear. Our gene sets are not split by contrast, so one of our terms can pair
with the trend and the endpoint both; the `shared_with` column of the table
above names which.

```{r, eval=have_go}
go_shared_trend = go_res %>%
  dplyr::filter(source == "ours") %>%
  dplyr::select(GOID, GOName, sex, direction, n_genes_ours = nSharedGenes,
                p_ours = Pval, fdr_ours = FDR) %>%
  dplyr::inner_join(
    go_res %>%
      dplyr::filter(source == "Yasmin et al", contrast == "trend") %>%
      dplyr::select(GOID, sex, direction, n_genes_yasmin = nSharedGenes,
                    p_yasmin = Pval, fdr_yasmin = FDR),
    by = c("GOID", "sex", "direction")) %>%
  dplyr::arrange(sex, direction, pmax(p_ours, p_yasmin))

write.csv(go_shared_trend,
          file = paste0(results_dir, "/shared_trend_GO_terms.csv"),
          row.names = FALSE)

datatable(go_shared_trend %>% mutate(across(where(is.numeric), ~signif(., 3))),
          rownames = FALSE, filter = "top", options = list(pageLength = 15))
```

`r if (!have_go) "" else if (nrow(go_shared_trend) == 0) "**No term is enriched on both sides of the same sex in the same direction.**" else { cnt = go_shared_trend %>% dplyr::count(sex, direction); paste0(nrow(go_shared_trend), " term pairs are enriched on both sides, across ", dplyr::n_distinct(go_shared_trend$GOID), " distinct terms: ", paste0(cnt$n, " ", cnt$sex, " ", cnt$direction, collapse = "; "), ".") }`

```{r, eval=have_go && nrow(go_shared_trend) > 0, fig.width=12, fig.height=9}
# The weaker of the two p-values orders the terms, so a term high on the axis is
# one that both sides support rather than one side strongly and the other barely.
n_show = 12

shared_plot = go_shared_trend %>%
  dplyr::group_by(sex, direction) %>%
  dplyr::slice_min(pmax(p_ours, p_yasmin), n = n_show, with_ties = FALSE) %>%
  dplyr::ungroup() %>%
  dplyr::mutate(panel = paste0(sex, ", ", direction),
                ykey  = paste(panel, GOID, sep = "|"))

# one ordering per panel, so the y axis is sorted inside each facet rather than
# across all of them
lev = shared_plot %>%
  dplyr::arrange(panel, dplyr::desc(pmax(p_ours, p_yasmin))) %>%
  dplyr::pull(ykey)
shared_plot$ykey = factor(shared_plot$ykey, levels = lev)

shared_long = shared_plot %>%
  tidyr::pivot_longer(c(p_ours, p_yasmin), names_to = "side",
                      values_to = "p") %>%
  dplyr::mutate(side = dplyr::recode(side, p_ours = "ours",
                                     p_yasmin = "Yasmin et al trend"),
                neglog = -log10(p))

p_go_shared = ggplot(shared_long, aes(x = neglog, y = ykey)) +
  geom_line(aes(group = ykey), colour = "grey65", linewidth = .5) +
  geom_point(aes(colour = side), size = 2.6) +
  scale_colour_manual(values = c(ours = "#0072B2",
                                 `Yasmin et al trend` = "#E69F00"),
                      name = NULL) +
  scale_y_discrete(labels = function(x)
    shared_plot$GOName[match(x, as.character(shared_plot$ykey))]) +
  facet_wrap(~ panel, scales = "free_y", ncol = 2) +
  labs(x = expression(-log[10](italic(p))), y = NULL,
       title = "GO terms enriched in both datasets, same sex and direction",
       subtitle = paste0("up to ", n_show,
                         " terms per panel, ordered by the weaker side")) +
  theme_bw(base_size = 10) +
  theme(legend.position = "top",
        axis.text.y = element_text(size = 7),
        panel.grid.major.y = element_line(colour = "grey92"))

p_go_shared

ggsave(p_go_shared, file = paste0(results_dir, "/shared_trend_GO_terms.png"),
       width = 12, height = 9, dpi = 300)
```

`r if (!have_go || nrow(go_shared_trend) == 0) "" else paste0("Each term carries a point from each dataset. A term whose two points sit close together is supported about equally on both sides. One with a long connector is carried mostly by the dataset further to the right. Every term shown passes p < ", params$go_pval, " on both sides, which is what being shared means here.")`


# Sex specificity

A germ cell signature scored across the Yasmin et al samples asks whether their series
carries the germ cell states we describe. A signature built from a male germ
cell state should score higher in their males than in their females, and a
female one should do the reverse. Where it does not, the signature is reading
germ cell content rather than germ cell identity.

Two marker sources are scored side by side, because they measure different
things. *Atlas markers*, from `13_atlas_and_stage_correspondence.Rmd`, compare
each cluster against every other cluster in the atlas. Most of those clusters
are somatic, so a gene expressed across germ cells and absent from soma marks
every germ cluster at once. *Germ subcluster markers*, from
`17_germ_pseudotime_tscan.Rmd`, compare subcluster against subcluster within the
germ compartment alone, where the shared germ cell program is already divided
out. Neither set is pseudotime-associated; both are markers of a group of cells
against other groups.

```{r}
marker_sources = list(
  `germ subclusters` = paste0(params$project_dir, "/results/",
                              params$results_date, "_pseudotime_tscan_",
                              as.character(params$germ_marker_lineage),
                              as.character(params$germ_marker_suffix),
                              "/subcluster_marker_genes.csv"),
  atlas              = paste0(atlas_dir, "/cluster_marker_genes.csv"))

src_found       = vapply(marker_sources, file.exists, logical(1))
have_cl_markers = have_bulk && any(src_found)
```

`r if (have_bulk && !have_cl_markers) paste0("**This section is skipped.** Neither marker file was found: ", paste(unlist(marker_sources), collapse = "; "), ".") else ""`

`r if (have_cl_markers && !all(src_found)) paste0("**Only the ", paste(names(src_found)[src_found], collapse = " and "), " source was found.** Missing: ", paste(unlist(marker_sources)[!src_found], collapse = "; "), ".") else ""`

```{r, eval=have_cl_markers}
n_sig = as.integer(params$n_signature_genes)

read_markers = function(src) {
  # The id column is read as character or 5.1 and 5.10 both come back as 5.1.
  # read.csv converts a column of 5.1, 5.2, 15, 16 to numeric without complaint,
  # and the collision is silent.
  id_col = if (src == "germ subclusters") "node" else "cluster"
  tab = read.csv(marker_sources[[src]], stringsAsFactors = FALSE,
                 colClasses = setNames("character", id_col))
  if (src == "germ subclusters") {
    stopifnot(all(c("Gene", "node", "meanLogFC") %in% colnames(tab)))
    tab$group = as.character(tab$node)
  } else {
    stopifnot(all(c("Gene", "cluster", "meanLogFC") %in% colnames(tab)))
    germ_clusters = trimws(strsplit(as.character(params$germ_clusters),
                                    ",")[[1]])
    missing_cl = setdiff(germ_clusters, unique(as.character(tab$cluster)))
    if (length(missing_cl) > 0)
      stop("germ_clusters not present in ", marker_sources[[src]], ": ",
           paste(missing_cl, collapse = ", "), ". Found: ",
           paste(sort(unique(as.character(tab$cluster))), collapse = ", "))
    tab = tab[as.character(tab$cluster) %in% germ_clusters, , drop = FALSE]
    tab$group = as.character(tab$cluster)
  }
  tab$source = src
  tab %>% dplyr::select(source, group, Gene, meanLogFC)
}

germ_markers = dplyr::bind_rows(lapply(names(src_found)[src_found],
                                       read_markers))

# the signature is the strongest markers that are also measurable in the bulk
sig = germ_markers %>%
  dplyr::filter(Gene %in% rownames(y)) %>%
  dplyr::group_by(source, group) %>%
  dplyr::arrange(dplyr::desc(meanLogFC), .by_group = TRUE) %>%
  dplyr::slice_head(n = n_sig) %>%
  dplyr::ungroup()
```

`r if (have_cl_markers) paste0("Signatures are the top ", n_sig, " markers per group by mean log fold change, kept only where the gene is measurable in the Yasmin et al series.", if ("germ subclusters" %in% names(src_found)[src_found]) paste0(" The germ subclusters come from ", basename(dirname(marker_sources[["germ subclusters"]])), ", so the groups are that run's subclusters and anything it removed is absent: ", paste(sort(unique(germ_markers$group[germ_markers$source == "germ subclusters"])), collapse = ", "), ".") else "") else ""`

```{r, eval=have_cl_markers}
# Each gene is standardized across the 28 samples first, so a signature score
# averages comparable quantities rather than raw abundances.
ylog    = cpm(y, log = TRUE, prior.count = 1)
gene_sd = apply(ylog, 1, sd)
zmat    = t(scale(t(ylog[gene_sd > 0, , drop = FALSE])))

score_mats = lapply(split(sig, sig$source), function(d) {
  sl = split(d$Gene, d$group)
  m  = vapply(sl, function(gs) {
    gs = intersect(gs, rownames(zmat))
    colMeans(zmat[gs, , drop = FALSE], na.rm = TRUE)
  }, numeric(ncol(zmat)))
  rownames(m) = colnames(zmat)
  m
})
```

```{r, eval=have_cl_markers}
# Mean pseudotime per node, from the same run the germ subcluster markers come
# from, so the nodes line up one for one. 17 writes one pseudotime per cell,
# averaged over the paths out of the root, which is the value it reports
# everywhere else. The atlas clusters have no pseudotime and stay empty.
pt_file = paste0(dirname(marker_sources[["germ subclusters"]]),
                 "/pseudotime_per_cell.csv")
want_pt = "germ subclusters" %in% names(src_found)[src_found]

node_pt = NULL
if (want_pt && file.exists(pt_file)) {
  pt_cells = read.csv(pt_file, stringsAsFactors = FALSE,
                      colClasses = c(node = "character"))
  stopifnot(all(c("node", "pseudotime") %in% colnames(pt_cells)))
  node_pt = pt_cells %>%
    dplyr::group_by(group = .data$node) %>%
    dplyr::summarise(mean_pseudotime = mean(.data$pseudotime, na.rm = TRUE),
                     n_cells = dplyr::n(), .groups = "drop") %>%
    dplyr::mutate(source = "germ subclusters")
}
have_pt = !is.null(node_pt)
```

`r if (!have_cl_markers || !want_pt || have_pt) "" else paste0("**Pseudotime is left out.** ", pt_file, " was not found, so the subcluster panel is ordered by the sex difference and the pseudotime panel is dropped.")`

```{r, eval=have_cl_markers}
sex_tab = dplyr::bind_rows(lapply(names(score_mats), function(src)
  dplyr::bind_rows(lapply(colnames(score_mats[[src]]), function(gp)
    data.frame(source = src, group = gp,
               mid_males   = mean(score_mats[[src]][runs_mm, gp]),
               mid_females = mean(score_mats[[src]][runs_mf, gp]),
               difference  = mean(score_mats[[src]][runs_mm, gp]) -
                             mean(score_mats[[src]][runs_mf, gp]))))))

if (have_pt)
  sex_tab = sex_tab %>% dplyr::left_join(node_pt, by = c("source", "group"))

write.csv(sex_tab, file = paste0(results_dir, "/germ_signature_sex.csv"),
          row.names = FALSE)

datatable(sex_tab %>% mutate(across(where(is.numeric), ~signif(., 3))),
          rownames = FALSE, filter = "top", options = list(pageLength = 20))
```

`r if (have_cl_markers) "Mid males against mid females, mean score in each. A signature that separates the sexes is measuring a germ cell state. One that does not is measuring how much germ cell content the gonad has, which every germ signature shares." else ""`

```{r, eval=have_cl_markers}
# Spearman because twelve or so nodes cannot support a parametric correlation,
# and because the question is monotone rather than linear. Both the signed
# difference and its size are reported: the first asks whether later nodes score
# more male, the second whether they separate the sexes at all.
pt_cor = if (!have_pt) NULL else {
  d = sex_tab %>% dplyr::filter(source == "germ subclusters",
                                is.finite(mean_pseudotime))
  ct = function(y) suppressWarnings(
    stats::cor.test(d$mean_pseudotime, y, method = "spearman"))
  list(n = nrow(d), signed = ct(d$difference), size = ct(abs(d$difference)))
}
```

```{r, eval=have_cl_markers, fig.width=9, fig.height=6}
# One ordering per marker source, since the two sources name different groups
# and a global sort would interleave them.
sex_long = sex_tab %>%
  tidyr::pivot_longer(c(mid_males, mid_females), names_to = "sex",
                      values_to = "score") %>%
  dplyr::mutate(sex = factor(dplyr::recode(sex, mid_males = "mid males",
                                           mid_females = "mid females"),
                             levels = c("mid males", "mid females")))

# Rows run in pseudotime where there is one, so the dumbbell panel and the
# scatter beside it are read in the same order. The atlas groups have no
# pseudotime and fall back to the difference.
sex_long$ord = if (have_pt)
  dplyr::coalesce(sex_long$mean_pseudotime, sex_long$difference) else
  sex_long$difference

# One plot per source rather than two facets. The sources hold different numbers
# of groups, so facets of equal height leave the smaller one mostly empty; as
# separate plots the heights follow the group counts and the rows line up at the
# same spacing. It also lets each panel carry a tag.
sex_panel = function(src) {
  d = sex_long %>% dplyr::filter(source == src)
  ggplot(d, aes(x = score, y = reorder(group, ord))) +
    geom_vline(xintercept = 0, colour = "grey70") +
    geom_line(aes(group = group), colour = "grey65", linewidth = .4) +
    geom_point(aes(colour = sex), size = 2) +
    scale_colour_manual(values = c(`mid males` = "#0072B2",
                                   `mid females` = "#CC79A7"), name = NULL) +
    labs(x = "mean signature score", y = NULL, title = src) +
    theme_bw(base_size = 9) +
    theme(plot.title = element_text(size = 9, face = "plain"),
          panel.grid.major.y = element_line(colour = "grey92"),
          panel.grid.minor.x = element_blank())
}

pt_panel = function() {
  d = sex_tab %>% dplyr::filter(source == "germ subclusters",
                                is.finite(mean_pseudotime))
  rho = unname(pt_cor$signed$estimate)
  ggplot(d, aes(x = mean_pseudotime, y = difference)) +
    geom_hline(yintercept = 0, colour = "grey70") +
    geom_smooth(method = "lm", formula = y ~ x, se = FALSE,
                colour = "grey45", linewidth = .4, linetype = "22") +
    geom_point(size = 2, colour = "#009E73") +
    ggrepel::geom_text_repel(aes(label = group), size = 2.5, colour = "grey15",
                             segment.colour = "grey65", bg.color = "white",
                             bg.r = .12, min.segment.length = 0, seed = 1) +
    labs(x = "mean pseudotime", y = "male minus female score",
         title = "germ subclusters, by pseudotime",
         subtitle = sprintf("Spearman rho = %.2f, p = %.3f, n = %d nodes",
                            rho, pt_cor$signed$p.value, pt_cor$n)) +
    theme_bw(base_size = 9) +
    theme(plot.title = element_text(size = 9, face = "plain"),
          plot.subtitle = element_text(size = 8, colour = "grey30"),
          panel.grid.minor = element_blank())
}

srcs = unique(sex_tab$source)
dumb = setNames(lapply(srcs, sex_panel), srcs)
n_by = vapply(srcs, function(x) sum(sex_tab$source == x), integer(1))

# The scatter sits beside the subcluster dumbbell it is drawn from, so the two
# read as one statement about the subclusters. The atlas keeps the row above.
if (have_pt) {
  row_pt = dumb[["germ subclusters"]] | pt_panel()
  p_sex  = if ("atlas" %in% srcs) dumb[["atlas"]] / row_pt else row_pt
  hts    = if ("atlas" %in% srcs)
    unname(n_by[c("atlas", "germ subclusters")]) else 1
  # the scatter is last in the tag order, which is third only if the atlas drew
  pt_tag = if ("atlas" %in% srcs) "C" else "B"
} else {
  p_sex = patchwork::wrap_plots(dumb, ncol = 1)
  hts   = unname(n_by)
}

p_sex = p_sex +
  patchwork::plot_layout(heights = hts, guides = "collect") +
  patchwork::plot_annotation(
    tag_levels = "A",
    title = "Germ signature scores in Yasmin et al mid males and mid females",
    subtitle = if (have_pt)
      "subclusters ordered by mean pseudotime, atlas clusters by the male-minus-female difference" else
      "groups ordered by the male-minus-female difference",
    theme = theme(plot.title = element_text(size = 11),
                  plot.subtitle = element_text(size = 9))) &
  # collected to the foot of the figure; at the top it opens a gap between the
  # first panel's title and its own axes
  theme(legend.position = "bottom")

p_sex

ggsave(p_sex, file = paste0(results_dir, "/germ_signature_sex.png"),
       width = 9, height = 6, dpi = 300)
```

`r if (!have_cl_markers) "" else { n_m = sum(sex_tab$difference > 0); n_f = sum(sex_tab$difference < 0); paste0("Scores are averages of genes standardized across all 28 samples, so zero is the series mean rather than no expression. Of ", nrow(sex_tab), " signatures, ", n_m, " score higher in the mid males and ", n_f, " higher in the mid females. A signature whose two points nearly coincide is not reading identity, whatever it was built from.") }`

`r if (!have_cl_markers || !have_pt) "" else sprintf("Panel %s sets each subcluster's sex difference against its mean pseudotime, taken over the cells in that node. Spearman rho is %.2f, p = %.3f, across %d nodes. On the size of the difference rather than its sign, rho is %.2f, p = %.3f. Pseudotime is the average over the paths out of the root, so a node on a branch carries a blended value.", pt_tag, unname(pt_cor$signed$estimate), pt_cor$signed$p.value, pt_cor$n, unname(pt_cor$size$estimate), pt_cor$size$p.value)`


# Summary figure

The two comparisons that ask whether our signal reproduces, in one figure, drawn
at the size it will be printed. The panels come from the same two functions the
sections above call, with the type and mark sizes set smaller, so the numbers
are the ones those sections report. The germ signature scores are a different
question and stay in their own figure.

```{r}
# Drawn at final size rather than drawn large and reduced, because reduction is
# what takes type below a journal's minimum. 180 mm is the full width most
# journals set a wide figure at, and 225 mm is about what a page leaves above
# the caption.
jrnl_w = 180 / 25.4
jrnl_h = 225 / 25.4
# base_size 8 puts axis labels at 6.4 pt and the smallest text, the gene names,
# at 5.7 pt. Journal minima are usually 5 to 7 pt at final size.
jrnl_base = 8
```

```{r, eval=have_bulk && have_fgsea, fig.width=jrnl_w, fig.height=jrnl_h}
# titles are rewritten for the composite and the per-figure subtitles dropped,
# since the caption below carries what they said
panel = function(p, ttl) p + labs(title = ttl, subtitle = NULL) +
  theme(plot.title = element_text(size = jrnl_base, face = "plain"))

# All three stacked rather than the scatters set side by side. Each scatter is
# two facets wide and carries a three line annotation inside the panel, so half
# the page width clips that text. Stacking also keeps the enrichment panel from
# spreading across the whole width, which flattens the curves.
gsea_j = gsea_figure(base_size = jrnl_base, pt_size = 1.3, txt_size = 2.1,
                     line_w = .5)
# five gene labels rather than ten; at this panel size ten collide whatever the
# repel does with them
end_j   = agreement_figure(c("male, endpoint", "female, endpoint"),
                           "Yasmin et al log2FC, larvae to mid group",
                           "Yasmin et al endpoint",
                           base_size = jrnl_base, pt_size = .9, lab_size = 2.0,
                           ann_size = 2.2, n_lab = 5)
trend_j = agreement_figure(c("male, trend", "female, trend"),
                           "Yasmin et al log2FC across the series",
                           "Yasmin et al trend",
                           base_size = jrnl_base, pt_size = .9, lab_size = 2.0,
                           ann_size = 2.2, n_lab = 5)

composite =
  panel(gsea_j, "Our top genes in the Yasmin et al ranking") /
  # the two scatters carry the same legend, so the block collects it into one
  ((panel(end_j,   "Effect sizes, endpoint contrasts") /
    panel(trend_j, "Effect sizes, trend contrasts")) +
     patchwork::plot_layout(guides = "collect") &
     theme(legend.position = "bottom")) +
  # the enrichment panel is two facet rows, the scatters one each
  patchwork::plot_layout(heights = c(3.5, 5.36)) +
  patchwork::plot_annotation(tag_levels = "A") &
  # the default margins leave a band of white under each legend, which on a
  # page is width the panels could have had
  theme(plot.tag = element_text(size = jrnl_base + 2, face = "bold"),
        plot.margin = margin(2, 2, 2, 2),
        legend.margin = margin(0, 0, 0, 0),
        legend.box.spacing = unit(4, "pt"))

composite

# pdf is the version to submit: vector, no resolution to lose at typesetting
ggsave(composite, file = paste0(results_dir, "/bulk_validation_summary.pdf"),
       width = jrnl_w, height = jrnl_h)
ggsave(composite, file = paste0(results_dir, "/bulk_validation_summary.png"),
       width = jrnl_w, height = jrnl_h, dpi = 600)
```

`r if (!(have_bulk && have_fgsea)) "" else paste0("**A** walks the Yasmin et al ranking from most up to most down and steps up at each of our genes, so a set concentrated at one end drives the line away from zero. **B** and **C** plot our ", top_set_lab, " genes against the Yasmin et al effect sizes, colored by whether the two datasets agree on direction, with the counts and their hypergeometric p-values printed per panel. A asks whether our strongest genes concentrate in theirs; B and C ask whether the effect sizes themselves agree. The figure is drawn at 180 by 225 mm, so the type is at printed size and nothing is reduced at typesetting. B and C name their five strongest genes rather than the ten the section figures name, which is as many as the panels hold at this width.")`


# Full gene table

```{r, eval=have_bulk}
out = cmp %>%
  dplyr::select(Gene, WangGeneID, ours_male, ours_female,
                dplyr::starts_with("male_"), dplyr::starts_with("fem_"))

write.csv(out, file = paste0(results_dir, "/bulk_vs_pseudobulk_per_gene.csv"),
          row.names = FALSE)

datatable(head(out %>% mutate(across(where(is.numeric), ~signif(., 3))), 100),
          rownames = FALSE, filter = "top")
```

`r if (have_bulk && exists("have_gd")) { if (!have_gd) paste0("**WangGeneID is not shown.** ", gd_file, " was not found, so the export carries NCBI gene ids only.") else { n_na = sum(is.na(out$WangGeneID)); if (n_na > 0) paste0(format(n_na, big.mark = ","), " of ", format(nrow(out), big.mark = ","), " genes have no Wang identifier and are left blank.") else "Every gene in the export carries a Wang identifier." } } else ""`

```{r, results="asis"}
# Nothing is written when the bulk matrix is absent, so the section does not
# print a table of nothing.
if (have_bulk) cat("# Export\n\n## Everything written by this document\n")
```

```{r, eval=have_bulk}
file_notes = c(
  "trend_scores.csv"               = "The ordered covariate every Yasmin et al animal takes on its sex's trend axis, with the group, length and recorded life stage it was built from.",
  "bulk_agreement.csv"             = "Spearman rho and percent sign agreement between our stage effect and that of Yasmin et al, for every contrast and ranking side, genome-wide and on the top genes, with bootstrap intervals on rho.",
  "bulk_permutation_null.csv"      = "Each agreement statistic against its label-permutation null: the observed value, the null median and 95th percentile, and the empirical p-value.",
  "bulk_permutation_null.png"      = "The null distribution of rho across the top genes, with the observed value marked.",
  "bulk_endpoint_agreement.png"    = "Our top genes, our effect against the Yasmin et al endpoint contrasts, colored by whether the two datasets agree on direction.",
  "bulk_trend_agreement.png"       = "The same figure against Yasmin et al trend contrasts.",
  "bulk_agreement_scatter.png"     = "Our log fold change against that of Yasmin et al for every measurable gene, one panel per axis.",
  "top_gene_GO_pvalues.csv"        = "GO enrichment of the top stage genes from both sides, up and down taken separately, one row per term and gene set.",
  "top_gene_GO_overlaps.csv"       = "The genes behind each enriched term.",
  "top_gene_enrichment.csv"        = "Each of our top gene lists tested for enrichment in each Yasmin et al contrast's ranking, with the enrichment score, its normalized value, the permutation p-value and FDR, and whether the direction came out as expected.",
  "top_gene_enrichment_leading_edge.csv" = "The genes carrying each enrichment, with our effect size and the Yasmin et al test statistic.",
  "top_gene_enrichment.png"        = "Normalized enrichment score per set and contrast, with the up and down sets colored and significance marked.",
  "top_gene_list_upset.png"        = "Intersections of our top gene lists with Yasmin et al trend lists, largest first, as an UpSet plot.",
  "shared_trend_GO_terms.csv"      = "Terms enriched in our top genes and in Yasmin et al trend genes on the same sex in the same direction, with each side's p-value, FDR and gene count.",
  "shared_trend_GO_terms.png"      = "The strongest of those shared terms, each with a point per dataset, faceted by sex and direction.",
  "germ_signature_sex.csv"         = "Mean germ signature score in Yasmin et al mid males and mid females, per signature and marker source, with the difference between them. The germ subcluster rows also carry that node's mean pseudotime and its cell count; the atlas rows have neither.",
  "bulk_validation_summary.png"    = "The enrichment curves and the effect size scatters in one figure, drawn at 180 by 225 mm and written at 600 dpi.",
  "bulk_validation_summary.pdf"    = "The same figure as vector at the same size, which is the one to submit.",
  "germ_signature_sex.png"         = "The same scores as paired points per signature, with the subcluster sex difference set against mean pseudotime beside them.",
  "bulk_vs_pseudobulk_per_gene.csv" = "One row per gene: our male and female stage effects, each Yasmin et al contrast's coefficient, and the Wang identifier. The endpoint columns are log fold changes between two groups. Each Yasmin et al trend carries two columns: *_trend_slope, the coefficient edgeR returned for the regression on the trend score, and *_trend_series_log2FC, that slope multiplied by the span the score covers, which is the log2 change across the whole series and the column every figure and statistic here uses. edgeR names its coefficient logFC for any design, so the slope is renamed to keep it apart from the endpoint columns, which are fold changes between two groups.")

in_dir = sort(setdiff(list.files(results_dir), "Robjects"))

# The table describes what this document writes. The results directory also
# holds whatever earlier runs left behind, and those are named below rather
# than given rows with no description.
on_disk = intersect(names(file_notes), in_dir)
unknown = setdiff(in_dir, names(file_notes))
missing = setdiff(names(file_notes), in_dir)

export_tab = data.frame(
  file = on_disk,
  size = sapply(file.path(results_dir, on_disk),
                function(f) format(structure(file.size(f), class = "object_size"),
                                   units = "auto")),
  what = unname(file_notes[on_disk]),
  stringsAsFactors = FALSE)

datatable(export_tab, rownames = FALSE, escape = FALSE,
          options = list(pageLength = 20, dom = "tp",
                         columnDefs = list(list(width = "260px", targets = 0))))
```

`r if (!have_bulk) "" else if (length(unknown) > 0) paste0(length(unknown), " file(s) in the directory are not written by this document and are left out of the table: ", paste(unknown, collapse = ", "), ". Either an earlier version wrote them, or they are new and need a line in file_notes.") else ""`

`r if (!have_bulk) "" else if (length(missing) > 0) paste0("Described but not written on this run: ", paste(missing, collapse = ", "), ". A section that was skipped writes nothing.") else ""`

`r if (!have_bulk) "" else paste0("Two cached intermediates sit in ", results_dir, "/Robjects and are not listed: pseudobulk.Rds, the per-library count sums, and label_permutation_null.Rds, the permutation refits. Both are reused across runs and rebuilt when the inputs change.")`


# Summary

```{r, eval=have_bulk}
# The headline numbers come from the well measured pool. Both ranking directions
# are shown, since they answer different questions.
top_set = paste0("top ", top_ns[1], " each way")

row_for = function(ax, set, ro = "ours")
  agreement[agreement$axis == ax & agreement$pool == "well measured" &
              agreement$ranked_on == ro & agreement$set == set, ]

perm_for = function(ax, stat, ro = "ours")
  perm_tab$p_perm[perm_tab$axis == ax & perm_tab$pool == "well measured" &
                    perm_tab$ranked_on == ro & perm_tab$statistic == stat]

axis_block = function(ax, label) {
  ours_row = row_for(ax, top_set, "ours")
  yas_row  = row_for(ax, top_set, "yasmin")
  gw_row   = row_for(ax, "genome-wide", "ours")
  data.frame(
    question = paste0(label, c(", rho genome-wide",
                               ", rho on our top genes",
                               ", permutation p, our ranking",
                               ", rho on the Yasmin et al top genes",
                               ", permutation p, their ranking",
                               ", % same sign on their top genes")),
    answer = c(signif(gw_row$rho, 3),
               signif(ours_row$rho, 3),
               signif(perm_for(ax, paste0("rho, ", top_set), "ours"), 3),
               signif(yas_row$rho, 3),
               signif(perm_for(ax, paste0("rho, ", top_set), "yasmin"), 3),
               paste0(signif(yas_row$pct_same_sign, 3), "%")))
}

summary_tbl = dplyr::bind_rows(
  data.frame(question = c("shared genes tested",
                          paste0("genes at or above ", min_pb, " counts here")),
             answer = c(format(nrow(cmp), big.mark = ","),
                        format(sum(cmp$well_measured), big.mark = ","))),
  axis_block("male, endpoint",               "male endpoint"),
  axis_block("female, endpoint",             "female endpoint"),
  axis_block("male, trend",                  "male trend"),
  axis_block("female, trend",                "female trend"),
  data.frame(question = c("animals in the male endpoint",
                          "animals in the female endpoint",
                          "animals in each male trend",
                          "animals in each female trend"),
             answer = as.character(c(length(runs_ud) + length(runs_mm),
                                     length(runs_ud) + length(runs_mf),
                                     length(male_runs),
                                     length(fem_runs)))))

datatable(summary_tbl, rownames = FALSE,
          options = list(dom = "tp", ordering = FALSE, pageLength = 22))
```

## What this does not answer

Agreement shows the pattern reproduces in a series without our July-to-November
batch. It does not separate stage from batch within our data.

Both endpoints use the whole mid group. Neither is stage-matched to our
transformers animal by animal. The mid male group is stage 7 and
post-metamorphosis throughout. The mid female group spans stage 1 to
post-metamorphosis, which makes its late endpoint broader than ours. Read each
contrast on its own size rather than treating one sex as primary.

The larval endpoint is matched by stage description rather than by size. Our
larvae are sex-labile by our own staging and larger than every larval animal in
the comparison series. The two series set the larval boundary at different sizes.

The permutation asks whether Yasmin et al stage contrast agrees with our genes more
than an arbitrary contrast drawn from the same animals. It does not rule out a
confound shared by both studies. Anything that moves gonad composition the same
way in both series would pass it.

# Session info

```{r}
sessionInfo()
```
