---
title: "16_bulk_validation"
author: "Stephanie Hickey"
date: "`r Sys.Date()`"
output:
  html_document:
    code_folding: hide
    toc: true
    toc_float:
      toc_collapsed: true
params:
    project_dir: /mnt/ufs18/rs-013/bioinformaticsCore/projects/chong_davidson/BCC125_sea_lamprey_scRNA
    project_name: BCC125_sea_lamprey_scRNA
    lab: Weiming Li
    results_date: "20260813"
    genome_dir: /mnt/research/bioinformaticsCore/shared/Genomes/Petromyzon_marinus
    nfcore_dir: /mnt/research/bioinformaticsCore/projects/chong_davidson/BCC125_sea_lamprey_scRNA/results/yasmin_nfcore
    marker_file: canonical_markers_S1_only_2026-08-14.csv
    group_column: group_suppT1labels
    top_n_list: "100,250,500"
    n_boot: 2000
    n_perm_labels: 200
    n_rand_sets: 2000
    min_pseudobulk_count: 10
    go_pval: 0.01
    cores: 1
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

```{r, echo=FALSE, results="asis"}
cat("\n")
cat("#### Lab:", params$lab, " \n")
cat("#### Project:", params$project_name, " \n")
```

```{r}
library(tidyverse)
library(SingleCellExperiment)
library(edgeR)
library(patchwork)
library(ggrepel)
library(DT)
source(paste0(params$project_dir, "/src/scRNA_seq_functions.R"))
```

Independent evidence for the stage-associated patterns. Answers the reviewer's request for
"independent evidence supporting the stage-associated patterns, such as
agreement with staining or previous bulk RNA-seq studies."

# Scope

**What agreement shows.** Whether the larva-to-transformer expression changes we
report reproduce in animals collected, dissected, prepared and sequenced by a
different group at a different time. Our July-to-November technical batch does
not exist in Yasmin's data. A pattern driven by that batch has no reason to
appear there.

**What it does not show.** That the changes are developmental. Yasmin's series
has its own collection structure and may carry its own confounds. Agreement
shows the pattern is not specific to our batch. That is what the reviewer asks
for. The claim goes no further.

**Bulk is a mixture.** A gene changing in bulk can reflect a shift in which cell
types are present rather than a change within any cell type. Pseudobulk shares
that property. The two sides are on the same footing. The composition section
measures how much of the agreement composition accounts for.

**The larval endpoint is matched by stage rather than by size.** Our larvae are
112 mm and sex-labile by our own staging. Every larval-stage animal in Yasmin's
series is smaller, the largest being 99 mm. Their 74–99 mm animals are already
classed as past the point where ovarian differentiation completes. The two series
draw the larval boundary at different sizes. The transformer endpoints match on
stage and on size.

```{r}
object_dir  = paste0(params$project_dir, "/results/", params$results_date,
                     "_no_doublets")
atlas_dir   = paste0(params$project_dir, "/results/", params$results_date,
                     "_atlas")
results_dir = paste0(params$project_dir, "/results/", params$results_date,
                     "_bulk_validation")
Robject_out_dir = paste0(results_dir, "/Robjects")
if (!dir.exists(Robject_out_dir)) dir.create(Robject_out_dir, recursive = TRUE)
data_dir    = paste0(params$project_dir, "/data")
```

```{r}
# The Yasmin reads were quantified with nf-core/rnaseq against
# GCF_010993605.1_kPetMar1.pri. Point nfcore_dir at that run's output directory.
# Either the top level or star_salmon/ itself is accepted.
#
# Length-scaled counts come first in the candidate list. They carry the
# transcript-length correction that makes gene-level Salmon estimates comparable
# across samples, and they are the right edgeR input without offsets. Plain
# gene_counts is the fallback.
nfcore_candidates = c(
  "star_salmon/salmon.merged.gene_counts_length_scaled.tsv",
  "salmon/salmon.merged.gene_counts_length_scaled.tsv",
  "star_salmon/salmon.merged.gene_counts.tsv",
  "salmon/salmon.merged.gene_counts.tsv",
  "salmon.merged.gene_counts_length_scaled.tsv",
  "salmon.merged.gene_counts.tsv")

nfcore_dir  = as.character(params$nfcore_dir)
count_path  = NA_character_

if (nzchar(nfcore_dir)) {
  hits = file.path(nfcore_dir, nfcore_candidates)
  hits = hits[file.exists(hits)]
  if (length(hits) > 0) count_path = hits[1]
}

have_bulk    = !is.na(count_path)
length_scaled = have_bulk && grepl("length_scaled", count_path)
```

`r if (!have_bulk && !nzchar(nfcore_dir)) "**The nfcore_dir parameter is not set.** Everything below is skipped. Set it to the nf-core/rnaseq output directory for the Yasmin data, either the run's top level or its star_salmon subdirectory." else ""`

`r if (!have_bulk && nzchar(nfcore_dir)) paste0("**No gene count matrix was found under ", nfcore_dir, ".** Everything below is skipped. Looked for: ", paste(nfcore_candidates, collapse = ", "), ". Found instead: ", paste(head(list.files(nfcore_dir), 20), collapse = ", "), ".") else ""`

`r if (have_bulk) paste0("Counts read from ", basename(count_path), if (length_scaled) ". The length-scaled matrix is the right edgeR input without offsets." else ". **This is the unscaled gene_counts matrix.** The transcript-length correction is absent. Use salmon.merged.gene_counts_length_scaled.tsv if the run produced one.") else ""`

# The Yasmin series

```{r, eval=have_bulk}
sample_meta = read.csv(paste0(data_dir, "/yasmin_PRJNA749754_sample_groups.csv"),
                       stringsAsFactors = FALSE)

# nf-core/rnaseq gene matrix: gene_id, gene_name, then one column per sample.
x = read.delim(count_path, check.names = FALSE, stringsAsFactors = FALSE)

id_cols = intersect(c("gene_id", "gene_name"), colnames(x))
if (!"gene_id" %in% id_cols)
  stop("no gene_id column in ", count_path, ". Columns found: ",
       paste(head(colnames(x), 10), collapse = ", "))

yasmin_counts = as.matrix(x[, setdiff(colnames(x), id_cols), drop = FALSE])
rownames(yasmin_counts) = x$gene_id

# Salmon estimates are non-integer. edgeR wants counts, and the length-scaled
# matrix is already on the count scale, so rounding is the only step needed.
yasmin_counts = round(yasmin_counts)
storage.mode(yasmin_counts) = "integer"

# nf-core sample names come from the samplesheet the run used. They may be run
# accessions, the paper's own animal names, or the two glued together, as in
# SRR15275627_F_JUV. Read the column names in all three forms and take the
# form and sample sheet column that resolve the most samples.
raw_ids = colnames(yasmin_counts)
has_acc = grepl("^[DES]RR[0-9]+", raw_ids)

id_forms = list(
  "as given"      = raw_ids,
  "run accession" = ifelse(has_acc, sub("^([DES]RR[0-9]+).*$", "\\1", raw_ids),
                           NA_character_),
  "name after the accession" =
                    ifelse(grepl("^[DES]RR[0-9]+_", raw_ids),
                           sub("^[DES]RR[0-9]+_", "", raw_ids), NA_character_))

id_candidates = c("run", "sample_name", "library_name", "suppT1_id")
id_candidates = intersect(id_candidates, colnames(sample_meta))

match_grid = expand.grid(form = names(id_forms), column = id_candidates,
                         stringsAsFactors = FALSE)
match_grid$n_matched = mapply(function(f, cl)
  sum(id_forms[[f]] %in% sample_meta[[cl]], na.rm = TRUE),
  match_grid$form, match_grid$column)

best          = match_grid[which.max(match_grid$n_matched), ]
sample_ids    = id_forms[[best$form]]
sample_id_col = best$column

if (best$n_matched < ncol(yasmin_counts))
  stop("could not match all ", ncol(yasmin_counts), " count columns to the ",
       "sample sheet. The best pairing was the '", best$form, "' form of the ",
       "column names against sample sheet column '", sample_id_col,
       "', matching ", best$n_matched, ". Unmatched: ",
       paste(raw_ids[!(sample_ids %in% sample_meta[[sample_id_col]])],
             collapse = ", "),
       ". Add the nf-core sample names to ",
       "data/yasmin_PRJNA749754_sample_groups.csv as a new column.")

absent = setdiff(sample_meta[[sample_id_col]], sample_ids)
if (length(absent) > 0)
  stop(length(absent), " sample(s) in the sample sheet are missing from the ",
       "count matrix: ", paste(absent, collapse = ", "),
       ". The nf-core run is incomplete, or its samplesheet omitted them.")

sample_meta = sample_meta[match(sample_ids, sample_meta[[sample_id_col]]), ]
stopifnot(identical(sample_meta[[sample_id_col]], sample_ids))

# Everything downstream selects columns with sample_meta$run, so put the count
# matrix on run accessions whatever form the nf-core names took.
stopifnot(!anyDuplicated(sample_meta$run))
colnames(yasmin_counts) = sample_meta$run

id_note = paste0("Count columns were matched to the sample sheet by their ",
                 best$form, ", against the `", sample_id_col, "` column. All ",
                 ncol(yasmin_counts), " samples resolved.")
```

`r if (have_bulk) id_note else ""`

```{r, eval=have_bulk}
# a numeric developmental position per animal, parsed from the recorded life
# stage rather than from the group label. The mid female group spans stage 1 to
# post-metamorphosis, so the label does not place an animal.
stage_score = function(s) {
  s = tolower(s)
  dplyr::case_when(
    grepl("post-metamorphosis", s)             ~ 8,
    grepl("metamorphosing stage", s)           ~ suppressWarnings(
                                                   as.numeric(sub(".*stage ([0-9]+).*",
                                                                  "\\1", s))),
    grepl("larval", s)                         ~ 0,
    grepl("upstream migrant", s)               ~ 10,
    TRUE                                       ~ NA_real_)
}

sample_meta$stage_score = stage_score(sample_meta$suppT1_life_stage)
stopifnot(!any(is.na(sample_meta$stage_score)))

# Which group assignment to use. The sample sheet carries two, because the paper
# and its Supplementary Table 1 split early from mid males differently: the text
# and count tables cut at metamorphic stage 5, Supplementary Table 1 cuts at
# stage 7. They disagree for M_M5 and M_M6 and agree everywhere else, so no
# female group and no female result depends on this choice.
group_col = as.character(params$group_column)
stopifnot(group_col %in% colnames(sample_meta))
sample_meta$group_used = sample_meta[[group_col]]

series_tab = sample_meta %>%
  dplyr::group_by(group_used) %>%
  dplyr::summarise(n = dplyr::n(),
                   length_mm = paste0(min(suppT1_length_mm), "-",
                                      max(suppT1_length_mm)),
                   stages = paste(sort(unique(suppT1_life_stage)), collapse = "; "),
                   .groups = "drop")

datatable(series_tab, rownames = FALSE, options = list(dom = "t", paging = FALSE))
```

`r if (have_bulk) paste0("Groups are taken from the **", group_col, "** column. ", if (group_col == "group_suppT1labels") "Supplementary Table 1 labels its own rows M_E1 to M_E5 and M_M1 to M_M4, putting metamorphic stages 1 to 6 in early males and stage 7 with post-metamorphosis in mid males. This matches how the main text describes mid males, \"metamorphosing stage 7 and immediately post-metamorphosis\". It does not reproduce the n of 4 early and 6 mid printed in the text, which the group_textcounts column does. Mid males here are the four animals at stage 7 or beyond, which is the same rule used to pick the stage-matched mid females, so the two sexes are selected alike." else "This column reproduces the group sizes printed in the main text by cutting early from mid males at metamorphic stage 5. Supplementary Table 1 labels those rows differently, cutting at stage 7, which is what group_suppT1labels holds.") else ""`

## Stage correspondence

Our animals are 112 mm sex-labile larvae and 131 mm newly transformed
transformers. The matches below use the
per-animal records in Supplementary Table 1 rather than the group labels.

```{r, eval=have_bulk}
# The animals within each mid group that are staged like our transformers, kept
# for the correspondence table. The endpoint contrasts use the whole group.
stage_matched = function(lbl) sample_meta$sample_name[
  sample_meta$group_used == lbl & sample_meta$stage_score >= 7]

len_range = function(lbl) {
  L = sample_meta$suppT1_length_mm[sample_meta$group_used == lbl]
  paste0(min(L), "-", max(L), " mm")
}

match_tab = tibble::tribble(
  ~ours,            ~yasmin,                                     ~basis,
  "larvae, 112 mm", paste0("undifferentiated larvae, ",
                           len_range("UD undifferentiated larvae")),
  "both pre-differentiation by description; ours are sex-labile by our staging; the two series set the larval size boundary differently",
  "TM, 131 mm",     paste0("all mid males, ", len_range("MM mid males")),
  paste0("stage 7 and post-metamorphosis by the Supplementary Table 1 labels; closest staged: ",
         paste(stage_matched("MM mid males"), collapse = ", ")),
  "TF, 131 mm",     paste0("all mid females, ", len_range("MF mid females")),
  paste0("group spans stage 1 to post-metamorphosis, used whole; closest staged: ",
         paste(stage_matched("MF mid females"), collapse = ", ")))

datatable(match_tab, rownames = FALSE, options = list(dom = "t", paging = FALSE))
```

# Pseudobulk

Raw counts are summed across all cells within each library, giving four samples.
Corrected and per-cell normalized values are not used. A bulk library measures
summed raw counts.

```{r, eval=have_bulk}
pseudo_file = paste0(Robject_out_dir, "/pseudobulk.Rds")

if (file.exists(pseudo_file)) {
  pseudobulk = readRDS(pseudo_file)
} else {
  all.sce = readRDS(paste0(object_dir, "/all.sce.Rds"))
  universe = Reduce(intersect, lapply(all.sce, rownames))
  pseudobulk = vapply(all.sce, function(x)
    Matrix::rowSums(counts(x[universe, ])), numeric(length(universe)))
  rownames(pseudobulk) = universe
  saveRDS(pseudobulk, file = pseudo_file)
}

library_meta = data.frame(
  batch   = c("UND1_C00004336", "UND2_C00004362", "TF_C00004337", "TM_C00004374"),
  library = c("Undifferentiated.1", "Undifferentiated.2", "Female", "Male"),
  stage   = c("larva", "larva", "transformer", "transformer"),
  stringsAsFactors = FALSE)

colnames(pseudobulk) = library_meta$library[
  match(colnames(pseudobulk), library_meta$batch)]
stopifnot(!any(is.na(colnames(pseudobulk))))
```

`r if (have_bulk) paste0("Pseudobulk over ", format(nrow(pseudobulk), big.mark = ","), " genes and ", ncol(pseudobulk), " libraries.") else ""`

## Shared gene set

```{r, eval=have_bulk}
shared = intersect(rownames(pseudobulk), rownames(yasmin_counts))

overlap_tab = data.frame(
  set = c("our single-cell genes", "Yasmin quantified genes", "shared"),
  n   = c(nrow(pseudobulk), nrow(yasmin_counts), length(shared)))

datatable(overlap_tab, rownames = FALSE, options = list(dom = "t"))

if (length(shared) < 0.5 * nrow(pseudobulk))
  stop("only ", length(shared), " of ", nrow(pseudobulk), " genes are shared. ",
       "Both sides should be NCBI kPetMar1 gene ids, so a low overlap means the ",
       "nf-core run used a different annotation from the single-cell counts. ",
       "Diagnose before continuing.")
```

`r if (have_bulk) paste0("**", format(length(shared), big.mark = ","), " genes are shared.** Everything below runs on that set. Genes are matched by identifier. Both sides come from the same assembly, so no ortholog assignment is involved.") else ""`

# Yasmin stage effects

```{r, eval=have_bulk}
y = DGEList(counts = yasmin_counts[shared, ], samples = sample_meta)
keep = filterByExpr(y, group = sample_meta$group_used)
y = y[keep, , keep.lib.sizes = FALSE]
y = calcNormFactors(y)

tested = rownames(y)
```

`r if (have_bulk) paste0(format(length(tested), big.mark = ","), " genes pass filterByExpr in the Yasmin series and carry through to the comparison.") else ""`

```{r, eval=have_bulk}
# Two contrast shapes serve the whole document: a two-group test and an ordered
# trend. Both return log fold changes, p-values and FDR per gene.
run_two_group = function(dge, run_a, run_b, label) {
  sub = dge[, c(run_a, run_b)]
  grp = factor(c(rep("early", length(run_a)), rep("late", length(run_b))),
               levels = c("early", "late"))
  des = model.matrix(~grp)
  sub = estimateDisp(sub, des)
  fit = glmQLFit(sub, des)
  res = glmQLFTest(fit, coef = 2)
  topTags(res, n = Inf, sort.by = "none")$table %>%
    tibble::rownames_to_column("Gene") %>%
    dplyr::select(Gene, logFC, PValue, FDR) %>%
    dplyr::rename_with(~paste0(label, "_", .x), -Gene)
}

run_trend = function(dge, runs, score, label) {
  sub = dge[, runs]
  des = model.matrix(~score)
  sub = estimateDisp(sub, des)
  fit = glmQLFit(sub, des)
  res = glmQLFTest(fit, coef = 2)
  topTags(res, n = Inf, sort.by = "none")$table %>%
    tibble::rownames_to_column("Gene") %>%
    dplyr::select(Gene, logFC, PValue, FDR) %>%
    dplyr::rename_with(~paste0(label, "_", .x), -Gene)
}

# Spearman with a bootstrap CI. cor.test gives a p-value but no interval, and
# the interval is the quantity worth reporting.
spearman_ci = function(x, y, n_boot = params$n_boot, seed = 1) {
  ok = is.finite(x) & is.finite(y)
  x = x[ok]; y = y[ok]
  r = suppressWarnings(cor(x, y, method = "spearman"))
  set.seed(seed)
  bs = replicate(n_boot, {
    i = sample(length(x), replace = TRUE)
    suppressWarnings(cor(x[i], y[i], method = "spearman"))
  })
  c(rho = r, lo = unname(quantile(bs, .025, na.rm = TRUE)),
    hi = unname(quantile(bs, .975, na.rm = TRUE)), n = length(x))
}

g = sample_meta$group_used
runs_ud = sample_meta$run[g == "UD undifferentiated larvae"]
runs_pm = sample_meta$run[g == "PM prospective males"]
runs_em = sample_meta$run[g == "EM early males"]
runs_mm = sample_meta$run[g == "MM mid males"]
runs_ef = sample_meta$run[g == "EF early females"]
runs_mf = sample_meta$run[g == "MF mid females"]
```

```{r, eval=have_bulk}
# Two endpoints and two trends.
#
# The endpoints are larvae against the whole mid group, one per sex, so the two
# sexes are contrasted the same way and on the largest groups available.
#
# The trends order the animals by group and then, inside each group, by the
# metamorphic stage recorded for that animal in Supplementary Table 1. The
# within-group offset is scaled to sit inside the gap to the next group, so
# groups never overlap.
#
# The offset uses the finest developmental information the paper reports. It
# changes nothing in the groups whose animals share a stage: undifferentiated
# larvae, prospective males and early females are all recorded as larval, so
# they stay flat. It orders early males (stages 1, 1, 3, 4, 5, 6), mid females
# (1, 1, 3, 4, 7, 8) and, barely, mid males (7, 7, 8, 8).
stage_offset = function(runs) {
  st = sample_meta$stage_score[match(runs, sample_meta$run)]
  (rank(st, ties.method = "min") - 1) / length(runs)
}

make_score = function(group_runs, mode) {
  unlist(lapply(seq_along(group_runs), function(i) {
    runs = group_runs[[i]]
    (i - 1) + if (mode == "stage") stage_offset(runs) else rep(0, length(runs))
  }))
}

male_end = run_two_group(y, runs_ud, runs_mm, "male_end")
fem_end  = run_two_group(y, runs_ud, runs_mf, "fem_end")

male_groups = list(runs_ud, runs_pm, runs_em, runs_mm)
male_runs   = unlist(male_groups)
male_group_index = make_score(male_groups, "group")
male_score       = make_score(male_groups, "stage")
male_trend       = run_trend(y, male_runs, male_score, "male_trend")

fem_groups = list(runs_ud, runs_ef, runs_mf)
fem_runs   = unlist(fem_groups)
fem_group_index = make_score(fem_groups, "group")
fem_score       = make_score(fem_groups, "stage")
fem_trend       = run_trend(y, fem_runs, fem_score, "fem_trend")

yasmin_res = Reduce(function(a, b) dplyr::full_join(a, b, by = "Gene"),
                    list(male_end, fem_end, male_trend, fem_trend))
```

## Trend scores

Each trend is a regression of expression on a single ordered covariate, so the
question is what number each animal gets.

The groups supply the coarse order, the one Yasmin give: undifferentiated larvae,
then prospective males, early males and mid males on the male axis;
undifferentiated larvae, early females and mid females on the female axis. That
is the `group_index` column below.

Within a group the animals are ordered by their own recorded life stage.
Supplementary Table 1 gives one for every animal and it is finer than the group
labels: mid females span metamorphic stage 1 to post-metamorphosis, and early
males span stages 1 to 6. Scoring by group alone would discard per-animal
information the paper reports.

The offset is built so it can only order animals inside a group, never move one
past another group:

1. Parse the recorded life stage to a number. Larval is 0, metamorphosing stage
   *N* is *N*, post-metamorphosis is 8, upstream migrant is 10. `M_M1b`, recorded
   as "Larval stage/Metamorphosing stage 1", is read as stage 1.
2. Rank the animals within their group by that number, ties taking the lower
   rank.
3. Subtract one and divide by the group size, giving an offset in [0, 1).
4. Add it to the group integer.

Because the offset is always below 1, group *k* occupies [*k*, *k*+1) and the
groups stay in order. Where every animal in a group shares a stage the offset is
0 for all of them, so the two scorings agree there. That is the case for the
undifferentiated larvae, the prospective males and the early females, which are
all recorded simply as larval.

```{r, eval=have_bulk}
score_tab = dplyr::bind_rows(
  data.frame(axis = "male", run = male_runs,
             group_index = male_group_index, trend_score = male_score),
  data.frame(axis = "female", run = fem_runs,
             group_index = fem_group_index, trend_score = fem_score)) %>%
  dplyr::mutate(
    group  = sample_meta$group_used[match(run, sample_meta$run)],
    sample = sample_meta$sample_name[match(run, sample_meta$run)],
    length_mm  = sample_meta$suppT1_length_mm[match(run, sample_meta$run)],
    life_stage = sample_meta$suppT1_life_stage[match(run, sample_meta$run)]) %>%
  dplyr::arrange(axis, trend_score) %>%
  dplyr::select(axis, group, sample, length_mm, life_stage,
                group_index, trend_score)

write.csv(score_tab, file = paste0(results_dir, "/trend_scores.csv"),
          row.names = FALSE)

datatable(score_tab %>% mutate(across(where(is.numeric), ~signif(., 3))),
          rownames = FALSE, options = list(dom = "t", pageLength = 30))
```

```{r, eval=have_bulk}
contrast_tab = data.frame(
  contrast = c("male endpoint", "female endpoint",
               "male trend", "female trend"),
  design   = c(paste0("UD (n=", length(runs_ud), ") vs all mid males (n=",
                      length(runs_mm), ")"),
               paste0("UD (n=", length(runs_ud), ") vs all mid females (n=",
                      length(runs_mf), ")"),
               paste0("UD, PM, EM, MM ordered (n=", length(male_runs), ")"),
               paste0("UD, EF, MF ordered (n=", length(fem_runs), ")")),
  n_FDR_05 = c(sum(yasmin_res$male_end_FDR   < 0.05, na.rm = TRUE),
               sum(yasmin_res$fem_end_FDR    < 0.05, na.rm = TRUE),
               sum(yasmin_res$male_trend_FDR < 0.05, na.rm = TRUE),
               sum(yasmin_res$fem_trend_FDR  < 0.05, na.rm = TRUE)))

datatable(contrast_tab, rownames = FALSE, options = list(dom = "t", paging = FALSE))
```

The endpoints are the smaller designs, and both rest on the two undifferentiated
larvae as the early side. The correlation below uses effect sizes alone. Noise in
the Yasmin estimates pulls agreement toward zero, so low power makes that test
conservative rather than invalid. Per-gene significance comes from the trends.

Each trend is reported twice, scored by group alone and by group with the
animals ranked by recorded metamorphic stage inside it. The two answer the same
question with different assumptions about what the group labels leave out, and
agreement between them is worth as much as either on its own.

# Our stage effect

Two larval libraries against one transformer library per sex is not a formal
test. The log fold change is used as an effect size. No p-values are reported
from it.

Nothing on the Yasmin side requires a gene to be measurable here. Their
`filterByExpr` runs on their 28 samples, and the shared set is an intersection of
identifiers. A gene with three counts in one of our libraries and none in the
others produces a large fold change on almost no evidence, and it will sit near
the top of a list ranked by absolute fold change. `min_pseudobulk_count` marks
the genes that carry enough counts here to be worth ranking. Both the marked and
unmarked sets are carried through the comparison, so the effect of the filter is
visible rather than assumed.

```{r, eval=have_bulk}
pb = pseudobulk[shared, ][tested, ]

p = DGEList(counts = pb)
p = calcNormFactors(p)
plog = cpm(p, log = TRUE, prior.count = 1)

larva_mean = rowMeans(plog[, c("Undifferentiated.1", "Undifferentiated.2")])

min_pb = as.integer(params$min_pseudobulk_count)

ours = data.frame(
  Gene          = rownames(plog),
  ours_male     = plog[, "Male"]   - larva_mean,
  ours_female   = plog[, "Female"] - larva_mean,
  max_count     = apply(pb, 1, max),
  well_measured = apply(pb, 1, max) >= min_pb,
  stringsAsFactors = FALSE)
```

`r if (have_bulk) paste0(format(sum(ours$well_measured), big.mark = ","), " of ", format(nrow(ours), big.mark = ","), " genes reach ", min_pb, " counts in at least one of our four libraries.") else ""`

# Diagnostics

Two checks that decide how the comparison should be read. Neither depends on
Yasmin.

## Do the two larval libraries agree with each other?

UND1 and UND2 are the only replication in our data. The larva-to-transformer
difference has to be larger than the difference between two libraries of the
same stage, or there is no effect for any external dataset to reproduce.

The ratio of the two spreads has a known value when there is no stage effect at
all. With independent library noise of size sigma, the larval difference has
spread sigma times root two, and a transformer library against the mean of two
larval libraries has spread sigma times root one and a half. The ratio is
therefore root three quarters, or 0.866, under the null. Read the observed ratio
against that number rather than against 1.

```{r, eval=have_bulk}
und_diff = plog[, "Undifferentiated.1"] - plog[, "Undifferentiated.2"]
null_ratio = sqrt(0.75)

rep_tab = data.frame(
  quantity = c("UND1 against UND2, Spearman on log-CPM",
               "SD of the UND1 minus UND2 difference",
               "SD of our male stage effect",
               "SD of our female stage effect",
               "male effect SD divided by larval difference SD",
               "female effect SD divided by larval difference SD",
               "the same ratio when there is no stage effect"),
  value = c(
    suppressWarnings(cor(plog[, "Undifferentiated.1"],
                         plog[, "Undifferentiated.2"], method = "spearman")),
    sd(und_diff),
    sd(ours$ours_male),
    sd(ours$ours_female),
    sd(ours$ours_male)   / sd(und_diff),
    sd(ours$ours_female) / sd(und_diff),
    null_ratio))

datatable(rep_tab %>% mutate(across(where(is.numeric), ~signif(., 3))),
          rownames = FALSE, options = list(dom = "t", ordering = FALSE,
                                           paging = FALSE))
```

`r if (have_bulk) { r_m = sd(ours$ours_male) / sd(und_diff); excess = r_m / null_ratio; if (excess < 1.25) paste0("**The stage effect is barely larger than library noise.** The male ratio is ", signif(r_m, 3), " against ", signif(null_ratio, 3), " expected with no stage effect at all, an excess of ", signif(excess, 3), " fold. Library-level variation accounts for most of what is being called a stage effect, and a failure to reproduce in Yasmin would follow from that alone.") else paste0("The male ratio is ", signif(r_m, 3), " against ", signif(null_ratio, 3), " expected with no stage effect, an excess of ", signif(excess, 3), " fold. The stage effect is larger than library noise.") } else ""`

## What is in the top genes?

```{r, eval=have_bulk}
top_probe = as.integer(trimws(strsplit(as.character(params$top_n_list),
                                       ",")[[1]]))[1]

top_counts = dplyr::bind_rows(lapply(c("ours_male", "ours_female"),
  function(cl) {
    idx = order(abs(ours[[cl]]), decreasing = TRUE)[seq_len(top_probe)]
    data.frame(
      set = paste0("top ", top_probe, ", ", sub("ours_", "", cl)),
      median_max_count = median(ours$max_count[idx]),
      pct_below_cutoff = 100 * mean(!ours$well_measured[idx]))
  }))

top_counts = dplyr::bind_rows(
  top_counts,
  data.frame(set = "all genes",
             median_max_count = median(ours$max_count),
             pct_below_cutoff = 100 * mean(!ours$well_measured)))

datatable(top_counts %>% mutate(across(where(is.numeric), ~signif(., 3))),
          rownames = FALSE, options = list(dom = "t", paging = FALSE))
```

`r if (have_bulk) { pc = top_counts$pct_below_cutoff[1]; if (pc > 25) paste0("**", signif(pc, 3), "% of the top male genes fall below the count cutoff.** The unfiltered ranking is largely a low-count list, and the filtered results below are the ones to read.") else paste0(signif(pc, 3), "% of the top male genes fall below the count cutoff, so the unfiltered ranking is not dominated by low-count genes.") } else ""`

# The comparison

```{r, eval=have_bulk}
cmp = ours %>% dplyr::inner_join(yasmin_res, by = "Gene")

# The Wang identifier, so figures and the export carry the name the
# collaborators' sheets use. The table is written by 01, one row per NCBI id.
gd_file = paste0(object_dir, "/gene_descritption.csv")
have_gd = file.exists(gd_file)

if (have_gd) {
  gene_description = read.csv(gd_file, stringsAsFactors = FALSE) %>%
    dplyr::distinct(Gene, .keep_all = TRUE)
  cmp = cmp %>%
    dplyr::left_join(gene_description %>% dplyr::select(Gene, WangGeneID),
                     by = "Gene")
} else {
  gene_description = data.frame(Gene = character(), WangGeneID = character(),
                                description = character())
  cmp$WangGeneID = NA_character_
}
cmp$label = ifelse(is.na(cmp$WangGeneID), cmp$Gene, cmp$WangGeneID)

top_ns = as.integer(trimws(strsplit(as.character(params$top_n_list), ",")[[1]]))

# genome-wide, then restricted to our strongest stage genes.
#
# Relabelled contrasts drawn from the same 28 animals give a genome-wide
# correlation centred at zero, so the genome-wide value is a result in its own
# right rather than a baseline. What the two numbers compare is concentration:
# whether agreement is stronger among the genes we nominate than among genes in
# general. That is a statement about specificity.
#
# Sign agreement is scored genome-wide for the same purpose. Whether agreement
# exceeds chance is left to the permutation.
# rank_on picks which dataset selects the genes. "ours" is the original
# direction. "yasmin" ranks on their side, where the contrast is replicated and
# thousands of genes reach FDR 0.05, and asks only whether our four libraries
# follow. keep restricts to the genes that carry enough counts here to rank.
agreement_for = function(ours_col, yas_col, axis, keep, pool,
                         rank_on = "ours") {
  d    = cmp[keep, , drop = FALSE]
  rank_vec = if (rank_on == "ours") d[[ours_col]] else d[[yas_col]]

  rows = list()
  gw = spearman_ci(d[[ours_col]], d[[yas_col]])

  gw_agree = sign(d[[ours_col]]) == sign(d[[yas_col]])
  gw_agree = gw_agree[is.finite(gw_agree)]

  rows[[1]] = data.frame(
    axis = axis, pool = pool, ranked_on = rank_on, set = "genome-wide",
    n = gw[["n"]], rho = gw[["rho"]], lo = gw[["lo"]], hi = gw[["hi"]],
    pct_same_sign = 100 * mean(gw_agree))

  ord = order(abs(rank_vec), decreasing = TRUE)
  for (n in top_ns) {
    idx = ord[seq_len(min(n, nrow(d)))]
    s = spearman_ci(d[[ours_col]][idx], d[[yas_col]][idx])
    agree = sign(d[[ours_col]][idx]) == sign(d[[yas_col]][idx])
    agree = agree[is.finite(agree)]
    rows[[length(rows) + 1]] = data.frame(
      axis = axis, pool = pool, ranked_on = rank_on, set = paste0("top ", n),
      n = length(idx), rho = s[["rho"]], lo = s[["lo"]], hi = s[["hi"]],
      pct_same_sign = 100 * mean(agree))
  }
  dplyr::bind_rows(rows)
}

axis_cols = list(
  "male, endpoint"   = c(ours = "ours_male",   yas = "male_end_logFC"),
  "female, endpoint" = c(ours = "ours_female", yas = "fem_end_logFC"),
  "male, trend"      = c(ours = "ours_male",   yas = "male_trend_logFC"),
  "female, trend"    = c(ours = "ours_female", yas = "fem_trend_logFC"))

gene_pools = list("all genes"     = rep(TRUE, nrow(cmp)),
                  "well measured" = cmp$well_measured)

grid = expand.grid(axis = names(axis_cols), pool = names(gene_pools),
                   rank_on = c("ours", "yasmin"), stringsAsFactors = FALSE)

agreement = dplyr::bind_rows(Map(function(ax, pl, ro) {
  cols = axis_cols[[ax]]
  agreement_for(cols[["ours"]], cols[["yas"]], ax, gene_pools[[pl]], pl, ro)
}, grid$axis, grid$pool, grid$rank_on))

write.csv(agreement, file = paste0(results_dir, "/bulk_agreement.csv"),
          row.names = FALSE)

datatable(agreement %>% mutate(across(where(is.numeric), ~signif(., 3))),
          rownames = FALSE, filter = "top",
          options = list(pageLength = 20))
```

The table now carries every combination of contrast, gene pool and ranking side.
`ranked_on = "ours"` is the original comparison. `ranked_on = "yasmin"` is the
reciprocal one: their contrast picks the genes and our four libraries only have
to follow. The reciprocal direction is the better powered of the two, since the
selection happens on the replicated side.

`r if (have_bulk) { say = function(ax) { a = agreement[agreement$axis == ax & agreement$pool == "well measured", ]; ao = a[a$ranked_on == "ours", ]; ay = a[a$ranked_on == "yasmin", ]; paste0(ax, ": genome-wide ", signif(ao$rho[ao$set == "genome-wide"], 3), ", our top ", top_ns[1], " ", signif(ao$rho[ao$set == paste0("top ", top_ns[1])], 3), ", their top ", top_ns[1], " ", signif(ay$rho[ay$set == paste0("top ", top_ns[1])], 3)) }; paste0("Spearman rho among well measured genes, for each contrast. ", paste(vapply(names(axis_cols), say, character(1)), collapse = ". "), ". The gap between the genome-wide value and the top-gene values says whether agreement concentrates in the genes each side nominates. The permutation below says whether any of it exceeds chance.") } else ""`

## Reading the numbers

**`rho` and `pct_same_sign`** are descriptive. They say how far the two datasets
agree, genome-wide and across the top genes, and the bootstrap interval on `rho`
says how precisely each is estimated. Neither is a test on its own, because the
top-gene set is picked using one of the two variables being correlated.

The genome-wide and top-gene values together say whether agreement concentrates
in the genes each side nominates, which is a question about specificity.

Two p-values appear, and they test different things.

**`p_perm`**, in the permutation section below, is the one that says whether
agreement exceeds chance. It is the fraction of random relabelings of the same 28
animals that agree at least as well as Yasmin's real stage contrast does. It
assumes nothing about independence between genes, and it holds the gene-selection
effect fixed, since the same selection rule is applied inside every permutation.
Read the `ranked_on = "yasmin"` rows alongside the `ours` rows: their contrasts
are replicated and ours are not, so a list picked on their side tests the shared
axis with the selection happening on the better powered side.

**`p_rand`**, in the composition section, asks whether marker genes carry more of
the agreement than a random gene set of the same size. It is about which genes
carry the signal, not about whether signal exists.

Both are one-sided, in the direction of agreement. Disagreement appears as a
p-value near 1 rather than as a small one.

## A label permutation null

The two numbers above cannot be read as a test on their own. The top-gene set is
picked by the absolute value of our own log fold change, which is one of the two
variables being correlated. Selecting large effects widens the spread of that
variable and lowers its relative measurement error, which raises the correlation
by itself. Selecting on a noisy estimate also picks up genes whose effect we
overstated, and those regress toward the mean in Yasmin, which lowers it. Both
act at once.

The permutation holds all of that fixed. Group labels are reassigned at random
among the 28 animals, the same contrast shape is refitted, and the agreement
statistics are recomputed against our unchanged effect sizes and our unchanged
top-gene list. Whatever the selection does to the observed value, it does to
every null value in the same way. What is left is whether Yasmin's stage
contrast agrees with our genes more than an arbitrary contrast of the same shape
drawn from the same animals.

Labels are drawn from the whole series rather than shuffled within the contrast.
Shuffling within it has too few distinct splits to resolve anything: 2 against 6
gives 28 arrangements and 2 against 2 gives 6, so the smallest p-value reachable
on the female endpoint would be 0.17.

All four contrasts get a null, on the same terms. For the endpoints a random set
of animals is split into groups of the observed sizes. For the trends a random
set of the same size receives the observed score vector. The two trends carry
more animals than the endpoints do, so they are the better powered half of this
table on both sex axes.

```{r, eval=have_bulk}
n_perm_labels = as.integer(params$n_perm_labels)
perm_cores    = max(1L, as.integer(params$cores))

# One set of refits serves every summary. The fit does not depend on the gene
# pool or on which side ranks the genes, so each permutation is scored for all
# four combinations rather than refitted four times.
#
# When Yasmin ranks the genes, the null re-selects inside each permutation from
# the permuted contrast. Reusing the real contrast's gene list would carry the
# real labels into the null.
perm_combos = expand.grid(pool = names(gene_pools),
                          rank_on = c("ours", "yasmin"),
                          stringsAsFactors = FALSE)

# The draws are generated here, in one stream, rather than inside the workers,
# so the result does not depend on how many cores are used. The stream is the
# same one the earlier serial version consumed — one sample() call per
# permutation, in order — so a cache written by that version stays valid.
all_runs = sample_meta$run

draw_runs = function(n_draw, n_perm, seed) {
  set.seed(seed)
  replicate(n_perm, sample(all_runs, n_draw), simplify = FALSE)
}

# scoring for one permutation: every pool and ranking side from the one fit
score_draw = function(lfc_all, ours_col, top_n) {
  dplyr::bind_rows(lapply(seq_len(nrow(perm_combos)), function(k) {
    keep = gene_pools[[perm_combos$pool[k]]]
    ov   = cmp[[ours_col]][keep]
    lv   = lfc_all[keep]
    idx  = if (perm_combos$rank_on[k] == "ours")
             order(abs(ov), decreasing = TRUE)[seq_len(min(top_n, length(ov)))]
           else
             order(abs(lv), decreasing = TRUE)[seq_len(min(top_n, length(lv)))]
    agree = sign(ov[idx]) == sign(lv[idx])
    data.frame(
      pool = perm_combos$pool[k], rank_on = perm_combos$rank_on[k],
      rho_gw   = suppressWarnings(cor(ov, lv, method = "spearman",
                                      use = "complete.obs")),
      rho_top  = suppressWarnings(cor(ov[idx], lv[idx], method = "spearman",
                                      use = "complete.obs")),
      pct_sign = 100 * mean(agree, na.rm = TRUE))
  }))
}

# The fits are independent, so they fan out across cores. mclapply forks, which
# is why the job script pins the BLAS threads to one: eight R processes each
# spawning threads would oversubscribe the node.
perm_null = function(draws, fit_draw, ours_col, top_n, cores = 1) {
  work = function(d) {
    tab = try(fit_draw(d), silent = TRUE)
    if (inherits(tab, "try-error")) return(NULL)
    score_draw(tab$perm_logFC[match(cmp$Gene, tab$Gene)], ours_col, top_n)
  }
  res = if (cores > 1) parallel::mclapply(draws, work, mc.cores = cores)
        else lapply(draws, work)
  failed = vapply(res, function(x) is.null(x) || inherits(x, "try-error"),
                  logical(1))
  if (any(failed)) message(sum(failed), " of ", length(draws),
                           " permutations failed to fit and are dropped")
  dplyr::bind_rows(res[!failed])
}

two_group_perm = function(n_a, n_b, ours_col, seed) {
  perm_null(draw_runs(n_a + n_b, n_perm_labels, seed),
            function(d) run_two_group(y, d[seq_len(n_a)],
                                      d[n_a + seq_len(n_b)], "perm"),
            ours_col, top_ns[1], perm_cores)
}

trend_perm = function(score_vec, ours_col, seed) {
  perm_null(draw_runs(length(score_vec), n_perm_labels, seed),
            function(d) run_trend(y, d, score_vec, "perm"),
            ours_col, top_ns[1], perm_cores)
}

# cached on a signature, since the permutation is the slow step in the document
perm_file = paste0(Robject_out_dir, "/label_permutation_null.Rds")
perm_sig  = list(n_perm  = n_perm_labels,
                 top_n   = top_ns[1],
                 n_genes = nrow(cmp),
                 n_well  = sum(cmp$well_measured),
                 min_pb  = min_pb,
                 combos  = perm_combos,
                 scores  = list(male = male_score, female = fem_score),
                 runs    = sort(sample_meta$run),
                 sizes   = c(length(runs_ud), length(runs_mm),
                             length(runs_mf), length(male_runs),
                             length(fem_runs)),
                 shapes  = c("endpoints", "trends"))

cached = if (file.exists(perm_file)) readRDS(perm_file) else NULL

if (!is.null(cached) && identical(cached$signature, perm_sig)) {
  perm_res = cached$perm_res
} else {
  perm_res = list(
    "male, endpoint"   = two_group_perm(length(runs_ud), length(runs_mm),
                                        "ours_male", 1),
    "female, endpoint" = two_group_perm(length(runs_ud), length(runs_mf),
                                        "ours_female", 2),
    "male, trend"      = trend_perm(male_score, "ours_male", 3),
    "female, trend"    = trend_perm(fem_score, "ours_female", 4))

  saveRDS(list(signature = perm_sig, perm_res = perm_res), file = perm_file)
}
```

```{r, eval=have_bulk}
# one-sided: the question is whether the observed statistic sits above the null
emp_p = function(null_vals, obs) {
  ok = is.finite(null_vals)
  (1 + sum(null_vals[ok] >= obs)) / (1 + sum(ok))
}

perm_tab = dplyr::bind_rows(lapply(names(perm_res), function(ax) {
  nd = perm_res[[ax]]
  dplyr::bind_rows(lapply(split(nd, list(nd$pool, nd$rank_on), drop = TRUE),
    function(g) {
      pl  = g$pool[1]
      ro  = g$rank_on[1]
      obs = agreement[agreement$axis == ax & agreement$pool == pl &
                        agreement$ranked_on == ro, ]
      o   = c(obs$rho[obs$set == "genome-wide"],
              obs$rho[obs$set == paste0("top ", top_ns[1])],
              obs$pct_same_sign[obs$set == paste0("top ", top_ns[1])])
      nl  = list(g$rho_gw, g$rho_top, g$pct_sign)

      out = data.frame(
        axis        = ax,
        pool        = pl,
        ranked_on   = ro,
        statistic   = c("rho, genome-wide", paste0("rho, top ", top_ns[1]),
                        paste0("% same sign, top ", top_ns[1])),
        observed    = o,
        null_median = vapply(nl, median, numeric(1), na.rm = TRUE),
        null_95th   = vapply(nl,
                             function(v) unname(quantile(v, .95, na.rm = TRUE)),
                             numeric(1)),
        p_perm      = mapply(emp_p, nl, o),
        n_perm      = vapply(nl, function(v) sum(is.finite(v)), integer(1)))

      # the genome-wide row does not depend on which side ranks the genes
      if (ro == "yasmin") out = out[out$statistic != "rho, genome-wide", ]
      out
    }))
}))

write.csv(perm_tab, file = paste0(results_dir, "/bulk_permutation_null.csv"),
          row.names = FALSE)

datatable(perm_tab %>% mutate(across(where(is.numeric), ~signif(., 3))),
          rownames = FALSE, filter = "top", options = list(pageLength = 20))
```

`r if (have_bulk) paste0("The smallest p-value reachable with ", n_perm_labels, " permutations is ", signif(1 / (1 + n_perm_labels), 2), ". A statistic at that value is reported as a floor rather than as its true p-value.") else ""`

```{r, eval=have_bulk, fig.width=15, fig.height=7}
# the well measured pool only, since the unfiltered ranking is the one under
# suspicion
perm_long = dplyr::bind_rows(lapply(names(perm_res), function(ax) {
  nd = perm_res[[ax]][perm_res[[ax]]$pool == "well measured", ]
  data.frame(axis = ax, ranked_on = nd$rank_on, rho_top = nd$rho_top)
}))

obs_long = dplyr::bind_rows(lapply(names(perm_res), function(ax)
  dplyr::bind_rows(lapply(c("ours", "yasmin"), function(ro) {
    obs = agreement[agreement$axis == ax & agreement$pool == "well measured" &
                      agreement$ranked_on == ro, ]
    data.frame(axis = ax, ranked_on = ro,
               observed = obs$rho[obs$set == paste0("top ", top_ns[1])])
  }))))

p_perm = ggplot(perm_long, aes(x = rho_top)) +
  geom_histogram(bins = 40, fill = "grey75", colour = NA) +
  geom_vline(data = obs_long, aes(xintercept = observed),
             colour = "#c1272d", linewidth = .7) +
  facet_grid(ranked_on ~ axis, scales = "free_y") +
  labs(x = paste0("Spearman rho across the top ", top_ns[1], " genes"),
       y = "permutations",
       title = "Observed agreement against random contrasts of the same shape",
       subtitle = "well measured genes; rows give the side that ranked the genes") +
  theme_bw()

p_perm

ggsave(p_perm, file = paste0(results_dir, "/bulk_permutation_null.png"),
       width = 15, height = 7, dpi = 300)
```

## The trend figure

Two panels per sex. The scatter shows the top 100 genes by absolute pseudobulk
log fold change, ours against Yasmin's, with points coloured by whether the two
datasets put the gene in the same direction. The paired plot underneath takes the
strongest concordant genes and shows both effect sizes side by side, so a gene
that rises in both series reads as two points on the same side of zero.

```{r, eval=have_bulk, fig.width=12, fig.height=10}
trend_axes = c("male, trend", "female, trend")

trend_top = dplyr::bind_rows(lapply(trend_axes, function(ax) {
  cols = axis_cols[[ax]]
  d = cmp[cmp$well_measured, ] %>%
    dplyr::select(Gene, label, ours = dplyr::all_of(cols[["ours"]]),
                  yasmin = dplyr::all_of(cols[["yas"]])) %>%
    dplyr::filter(is.finite(ours), is.finite(yasmin)) %>%
    dplyr::slice_max(abs(ours), n = top_ns[1], with_ties = FALSE)
  d$axis  = ax
  d$agree = sign(d$ours) == sign(d$yasmin)
  d
}))

trend_top$axis = factor(trend_top$axis, levels = trend_axes)

agree_cols = c(`TRUE` = "#1b7837", `FALSE` = "#b2182b")

pct_lab = trend_top %>%
  dplyr::group_by(axis) %>%
  dplyr::summarise(pct = 100 * mean(agree), .groups = "drop") %>%
  dplyr::mutate(txt = paste0(round(pct), "% in the same direction"))

p_top = ggplot(trend_top, aes(x = ours, y = yasmin, colour = agree)) +
  geom_hline(yintercept = 0, colour = "grey70") +
  geom_vline(xintercept = 0, colour = "grey70") +
  geom_point(size = 1.6, alpha = .85) +
  geom_text_repel(data = trend_top %>% dplyr::group_by(axis) %>%
                    dplyr::slice_max(abs(ours), n = 12, with_ties = FALSE),
                  aes(label = label), size = 2.5, max.overlaps = 25,
                  show.legend = FALSE) +
  geom_text(data = pct_lab, aes(x = -Inf, y = Inf, label = txt),
            hjust = -0.08, vjust = 1.6, inherit.aes = FALSE, size = 3.4) +
  scale_colour_manual(values = agree_cols, name = NULL,
                      labels = c(`TRUE` = "same direction",
                                 `FALSE` = "opposite")) +
  facet_wrap(~axis, scales = "free") +
  labs(x = "our pseudobulk log2FC, larva to transformer",
       y = "Yasmin trend coefficient",
       title = paste0("Top ", top_ns[1],
                      " stage genes, ours against the Yasmin trend")) +
  theme_bw() + theme(legend.position = "top")

# the strongest concordant genes, both effect sizes side by side
paired = trend_top %>%
  dplyr::filter(agree) %>%
  dplyr::group_by(axis) %>%
  dplyr::slice_max(abs(ours), n = 20, with_ties = FALSE) %>%
  dplyr::ungroup() %>%
  dplyr::mutate(direction = ifelse(ours > 0, "up in transformers",
                                   "down in transformers"))

paired_long = paired %>%
  tidyr::pivot_longer(c(ours, yasmin), names_to = "dataset",
                      values_to = "effect") %>%
  dplyr::mutate(dataset = dplyr::recode(dataset, ours = "ours",
                                        yasmin = "Yasmin"))

p_paired = ggplot(paired_long,
                  aes(x = effect, y = reorder(label, effect))) +
  geom_vline(xintercept = 0, colour = "grey70") +
  geom_line(aes(group = label), colour = "grey65", linewidth = .4) +
  geom_point(aes(shape = dataset, fill = direction), size = 2.2,
             colour = "grey25", stroke = .3) +
  scale_shape_manual(values = c(ours = 21, Yasmin = 24), name = NULL) +
  scale_fill_manual(values = c(`up in transformers` = "#2166ac",
                               `down in transformers` = "#d6604d"),
                    name = NULL) +
  facet_wrap(~axis, scales = "free") +
  labs(x = "effect size in each dataset", y = NULL,
       title = "Strongest concordant genes, both datasets") +
  theme_bw() + theme(legend.position = "top",
                     axis.text.y = element_text(size = 7))

p_trend = p_top / p_paired + patchwork::plot_layout(heights = c(1, 1.3)) +
  patchwork::plot_annotation(tag_levels = "A")

p_trend

ggsave(p_trend, file = paste0(results_dir, "/bulk_trend_agreement.png"),
       width = 12, height = 10, dpi = 300)
```

`r if (have_bulk) paste0("Panel A keeps the top ", top_ns[1], " genes by our effect size and colours each by whether the two datasets agree on direction. Panel B takes the concordant genes with the largest effects and plots both estimates, so agreement reads as two markers on the same side of zero. The permutation p-values for these contrasts are in the table above; the percentages shown here are the same quantity the permutation tests.") else ""`

```{r, eval=have_bulk, fig.width=11, fig.height=11}
scatter_for = function(ours_col, yas_col, title, n_label = 15) {
  d = cmp %>%
    dplyr::select(Gene, ours = dplyr::all_of(ours_col),
                  yasmin = dplyr::all_of(yas_col)) %>%
    dplyr::filter(is.finite(ours), is.finite(yasmin))
  lab = d %>% dplyr::slice_max(abs(ours), n = n_label, with_ties = FALSE)

  ggplot(d, aes(x = ours, y = yasmin)) +
    geom_hline(yintercept = 0, colour = "grey80") +
    geom_vline(xintercept = 0, colour = "grey80") +
    geom_point(alpha = .2, size = .5) +
    geom_point(data = lab, colour = "#c1272d", size = 1.2) +
    geom_text_repel(data = lab, aes(label = Gene), size = 2.5, max.overlaps = 20) +
    geom_smooth(method = "lm", se = FALSE, linewidth = .5, colour = "#004488") +
    labs(x = "our pseudobulk log2FC, larva to transformer",
         y = "Yasmin log2FC", title = title) +
    theme_bw()
}

p_male  = scatter_for("ours_male",   "male_end_logFC",       "Male endpoint")
p_fem   = scatter_for("ours_female", "fem_end_logFC",        "Female endpoint")
p_mtr   = scatter_for("ours_male",   "male_trend_logFC", "Male trend")
p_ftr   = scatter_for("ours_female", "fem_trend_logFC",  "Female trend")

p_scatter = (p_male + p_fem) / (p_mtr + p_ftr)
p_scatter

ggsave(p_scatter, file = paste0(results_dir, "/bulk_agreement_scatter.png"),
       width = 11, height = 11, dpi = 300)
```

# Gene ontology

The hundred most up- and most down-regulated genes from each side, tested for GO
enrichment against the genes measurable in both datasets. Our sets are the same
for both contrasts on a sex, since our effect does not depend on which Yasmin
contrast it is compared with. Yasmin's sets are per contrast.

Terms are kept when they carry more than 10 and fewer than 200 genes, so the
result is neither noise from tiny terms nor a restatement of the whole ontology.

```{r}
gaf_file      = file.path(params$genome_dir,
                          "GCF_010993605.1_kPetMar1.pri_gene_ontology.gaf")
go_names_file = file.path(params$genome_dir, "GOs_names.txt")
have_go       = have_bulk && file.exists(gaf_file) && file.exists(go_names_file)
```

`r if (have_bulk && !have_go) paste0("**GO enrichment is skipped.** The GO files were not found under ", params$genome_dir, ".") else ""`

```{r, eval=have_go}
gaf = read.delim(gaf_file, header = FALSE, comment = "!") %>%
  dplyr::filter(!grepl("NOT", V4)) %>%
  dplyr::select(V3, V5) %>%
  dplyr::rename(GOID = "V5", Gene = "V3") %>%
  dplyr::distinct()

keep_go = gaf %>%
  dplyr::count(GOID) %>%
  dplyr::filter(n > 10, n < 200) %>%
  dplyr::pull(GOID)
gaf = gaf %>% dplyr::filter(GOID %in% keep_go)

go_names = read.delim(go_names_file, header = FALSE)
colnames(go_names) = c("GOID", "GOName")
```

```{r, eval=have_go}
# The background is the genes measurable on both sides, which is the set the
# lists were drawn from. Up and down are taken separately: a term enriched among
# rising genes says something different from one enriched among falling genes,
# and pooling them hides both.
go_pool = cmp[cmp$well_measured, , drop = FALSE]

top_dir = function(v, direction, n = top_ns[1]) {
  ok = is.finite(v)
  idx = if (direction == "up") which(ok & v > 0) else which(ok & v < 0)
  idx[order(abs(v[idx]), decreasing = TRUE)][seq_len(min(n, length(idx)))]
}

set_defs = c(
  # ours, one pair per sex
  list(list(label = "ours, male",   col = "ours_male"),
       list(label = "ours, female", col = "ours_female")),
  # Yasmin, one pair per contrast
  lapply(names(axis_cols), function(ax)
    list(label = paste0("Yasmin, ", ax), col = axis_cols[[ax]][["yas"]])))

go_sets = dplyr::bind_rows(lapply(set_defs, function(d)
  dplyr::bind_rows(lapply(c("up", "down"), function(dir) {
    idx = top_dir(go_pool[[d$col]], dir)
    data.frame(Gene = go_pool$Gene[idx],
               set  = paste0(d$label, ", ", dir),
               stringsAsFactors = FALSE)
  }))))

set_sizes = go_sets %>% dplyr::count(set, name = "n_genes")

datatable(set_sizes, rownames = FALSE,
          options = list(dom = "t", pageLength = 20))
```

```{r, eval=have_go}
go_res = run_go_enrichment(
  gaf              = gaf,
  gene_description = gene_description,
  go_names         = go_names,
  test_genes       = go_sets,
  background_n     = nrow(go_pool),
  group2_label     = "set",
  ngroup2_label    = "nSetGenes",
  overlaps_file    = paste0(results_dir, "/top_gene_GO_overlaps.csv"),
  pvals_file       = paste0(results_dir, "/top_gene_GO_pvalues.csv"),
  pval_filter      = params$go_pval,
  require_shared   = TRUE)

datatable(go_res %>% mutate(across(where(is.numeric), ~signif(., 3))),
          filter = "top", rownames = FALSE)
```

`r if (have_go && exists("go_res")) paste0(format(nrow(go_res), big.mark = ","), " term-by-set results pass p < ", params$go_pval, " across ", dplyr::n_distinct(go_sets$set), " gene sets. The full table is in top_gene_GO_pvalues.csv and the genes behind each term in top_gene_GO_overlaps.csv. Read the FDR column rather than Pval where a term is named individually: the FDR is computed within each set.") else ""`

# Composition

A bulk change can arise because a cell type became more abundant rather than
because any cell changed. If agreement is concentrated in cell type marker
genes, the shared signal is largely compositional. Both datasets would be seeing
the same shift in which cell types are present. If agreement holds among
non-markers too, expression is changing within cell types in both datasets.

Either outcome supports the reviewer's point. The result decides how the claim
is worded.

This runs on all four contrasts. The composition question is about which genes
carry the agreement, so it applies to the female axis and to the trends on the
same terms as to the male endpoint.

The marker set is small, and comparing it against the fifteen thousand genes
outside it compares a noisy estimate against a precise one. Markers are also
highly expressed and cell type restricted by construction, so they are not a
random draw. The control for both is a size-matched null: gene sets of the same
size drawn at random, giving the range of correlations any set that size
produces.

This section uses the well measured genes. Low-count genes are noise on our side
whichever set they fall into, and markers are highly expressed, so leaving them
in would load the non-marker set with genes that cannot agree with anything.

```{r, eval=have_bulk}
marker_path = paste0(data_dir, "/", params$marker_file)
have_markers = file.exists(marker_path)

if (have_markers) {
  km = read.csv(marker_path, stringsAsFactors = FALSE)
  marker_genes = unique(c(km$loc_id, km$symbol))
  marker_genes = marker_genes[!is.na(marker_genes) & nzchar(marker_genes)]

  cmp_wm = cmp[cmp$well_measured, , drop = FALSE]
  cmp_wm$is_marker = cmp_wm$Gene %in% marker_genes

  comp_tab = dplyr::bind_rows(lapply(names(axis_cols), function(ax) {
    cols = axis_cols[[ax]]
    dplyr::bind_rows(lapply(c(TRUE, FALSE), function(m) {
      d = cmp_wm[cmp_wm$is_marker == m, ]
      s = spearman_ci(d[[cols[["ours"]]]], d[[cols[["yas"]]]])
      data.frame(axis = ax,
                 gene_set = if (m) "cell type markers" else "all other genes",
                 n = s[["n"]], rho = s[["rho"]], lo = s[["lo"]], hi = s[["hi"]])
    }))
  }))

  datatable(comp_tab %>% mutate(across(where(is.numeric), ~signif(., 3))),
            rownames = FALSE, options = list(dom = "t", paging = FALSE))
}
```

```{r, eval=have_bulk}
# the size-matched null: where does the marker correlation fall among random
# gene sets of the same size?
if (have_markers) {
  n_rand = as.integer(params$n_rand_sets)

  rand_tab = dplyr::bind_rows(lapply(names(axis_cols), function(ax) {
    cols = axis_cols[[ax]]
    x  = cmp_wm[[cols[["ours"]]]]
    yv = cmp_wm[[cols[["yas"]]]]
    ok = is.finite(x) & is.finite(yv)
    x  = x[ok]; yv = yv[ok]; mk = cmp_wm$is_marker[ok]

    obs = suppressWarnings(cor(x[mk], yv[mk], method = "spearman"))
    set.seed(1)
    nulls = replicate(n_rand, {
      i = sample(length(x), sum(mk))
      suppressWarnings(cor(x[i], yv[i], method = "spearman"))
    })

    data.frame(axis = ax, n_markers = sum(mk), marker_rho = obs,
               random_median = median(nulls, na.rm = TRUE),
               random_95th = unname(quantile(nulls, .95, na.rm = TRUE)),
               p_rand = (1 + sum(nulls >= obs, na.rm = TRUE)) /
                        (1 + sum(is.finite(nulls))))
  }))

  write.csv(rand_tab, file = paste0(results_dir, "/bulk_composition_null.csv"),
            row.names = FALSE)

  datatable(rand_tab %>% mutate(across(where(is.numeric), ~signif(., 3))),
            rownames = FALSE, options = list(dom = "t", paging = FALSE))
}
```

`r if (have_bulk && exists("have_markers") && !have_markers) paste0("**The composition check is skipped.** The marker file ", params$marker_file, " was not found in the data directory.") else ""`

`r if (have_bulk && exists("rand_tab")) { n_hi = sum(rand_tab$p_rand < 0.05, na.rm = TRUE); paste0("Marker genes agree more than a random set of ", rand_tab$n_markers[1], " genes on ", n_hi, " of the ", nrow(rand_tab), " contrasts. ", if (n_hi >= 3) "The shared signal is partly compositional on most contrasts. Describe it as a shift in gonad composition together with a shift in expression." else if (n_hi == 0) "Agreement is not concentrated in marker genes on any contrast. Expression appears to change within cell types in both datasets." else "The contrasts disagree, so the compositional contribution is not resolved. Report the table rather than a verdict.") } else ""`

# Export

```{r, eval=have_bulk}
out = cmp %>%
  dplyr::select(Gene, WangGeneID, ours_male, ours_female,
                dplyr::starts_with("male_"), dplyr::starts_with("fem_"))

write.csv(out, file = paste0(results_dir, "/bulk_vs_pseudobulk_per_gene.csv"),
          row.names = FALSE)

datatable(head(out %>% mutate(across(where(is.numeric), ~signif(., 3))), 100),
          rownames = FALSE, filter = "top")
```

`r if (have_bulk && exists("have_gd")) { if (!have_gd) paste0("**WangGeneID is not shown.** ", gd_file, " was not found, so the export carries NCBI gene ids only.") else { n_na = sum(is.na(out$WangGeneID)); if (n_na > 0) paste0(format(n_na, big.mark = ","), " of ", format(nrow(out), big.mark = ","), " genes have no Wang identifier and are left blank.") else "Every gene in the export carries a Wang identifier." } } else ""`

# Summary

```{r, eval=have_bulk}
# The headline numbers come from the well measured pool. Both ranking directions
# are shown, since they answer different questions.
top_set = paste0("top ", top_ns[1])

row_for = function(ax, set, ro = "ours")
  agreement[agreement$axis == ax & agreement$pool == "well measured" &
              agreement$ranked_on == ro & agreement$set == set, ]

perm_for = function(ax, stat, ro = "ours")
  perm_tab$p_perm[perm_tab$axis == ax & perm_tab$pool == "well measured" &
                    perm_tab$ranked_on == ro & perm_tab$statistic == stat]

axis_block = function(ax, label) {
  ours_row = row_for(ax, top_set, "ours")
  yas_row  = row_for(ax, top_set, "yasmin")
  gw_row   = row_for(ax, "genome-wide", "ours")
  data.frame(
    question = paste0(label, c(", rho genome-wide",
                               ", rho on our top genes",
                               ", permutation p, our ranking",
                               ", rho on Yasmin's top genes",
                               ", permutation p, their ranking",
                               ", % same sign on their top genes")),
    answer = c(signif(gw_row$rho, 3),
               signif(ours_row$rho, 3),
               signif(perm_for(ax, paste0("rho, ", top_set), "ours"), 3),
               signif(yas_row$rho, 3),
               signif(perm_for(ax, paste0("rho, ", top_set), "yasmin"), 3),
               paste0(signif(yas_row$pct_same_sign, 3), "%")))
}

summary_tbl = dplyr::bind_rows(
  data.frame(question = c("shared genes tested",
                          paste0("genes at or above ", min_pb, " counts here")),
             answer = c(format(nrow(cmp), big.mark = ","),
                        format(sum(cmp$well_measured), big.mark = ","))),
  axis_block("male, endpoint",               "male endpoint"),
  axis_block("female, endpoint",             "female endpoint"),
  axis_block("male, trend",                  "male trend"),
  axis_block("female, trend",                "female trend"),
  data.frame(question = c("animals in the male endpoint",
                          "animals in the female endpoint",
                          "animals in each male trend",
                          "animals in each female trend"),
             answer = as.character(c(length(runs_ud) + length(runs_mm),
                                     length(runs_ud) + length(runs_mf),
                                     length(male_runs),
                                     length(fem_runs)))))

datatable(summary_tbl, rownames = FALSE,
          options = list(dom = "tp", ordering = FALSE, pageLength = 22))
```

## What this does not answer

Agreement shows the pattern reproduces in a series without our July-to-November
batch. It does not separate stage from batch within our data.

Both endpoints use the whole mid group, so neither is stage-matched to our
transformers animal by animal. The mid male group is stage 7 and
post-metamorphosis throughout; the mid female group spans stage 1 to
post-metamorphosis, so its late endpoint is broader than ours. Read each
contrast on its own size rather than treating one sex as primary.

The larval endpoint is matched by stage description rather than by size. Our
larvae are sex-labile by our own staging and larger than every larval animal in
the comparison series. The two series set the larval boundary at different sizes.

The permutation asks whether Yasmin's stage contrast agrees with our genes more
than an arbitrary contrast drawn from the same animals. It does not rule out a
confound shared by both studies. Anything that moves gonad composition the same
way in both series would pass it.

# Session info

```{r}
sessionInfo()
```
