---
title: "16_bulk_validation"
author: "Stephanie Hickey"
date: "`r Sys.Date()`"
output:
  html_document:
    code_folding: hide
    toc: true
    toc_float:
      toc_collapsed: true
params:
    project_dir: /mnt/ufs18/rs-013/bioinformaticsCore/projects/chong_davidson/BCC125_sea_lamprey_scRNA
    project_name: BCC125_sea_lamprey_scRNA
    lab: Weiming Li
    results_date: "20260813"
    genome_dir: /mnt/research/bioinformaticsCore/shared/Genomes/Petromyzon_marinus
    nfcore_dir: /mnt/research/bioinformaticsCore/projects/chong_davidson/BCC125_sea_lamprey_scRNA/results/yasmin_nfcore
    marker_file: canonical_markers_S1_only_2026-08-14.csv
    top_n_list: "100,250,500"
    n_boot: 2000
    n_perm_labels: 200
    n_rand_sets: 2000
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

```{r, echo=FALSE, results="asis"}
cat("\n")
cat("#### Lab:", params$lab, " \n")
cat("#### Project:", params$project_name, " \n")
```

```{r}
library(tidyverse)
library(SingleCellExperiment)
library(edgeR)
library(patchwork)
library(ggrepel)
library(DT)
source(paste0(params$project_dir, "/src/scRNA_seq_functions.R"))
```

Independent evidence for the stage-associated patterns. Answers the reviewer's request for
"independent evidence supporting the stage-associated patterns, such as
agreement with staining or previous bulk RNA-seq studies."

# Scope

**What agreement shows.** Whether the larva-to-transformer expression changes we
report reproduce in animals collected, dissected, prepared and sequenced by a
different group at a different time. Our July-to-November technical batch does
not exist in Yasmin's data. A pattern driven by that batch has no reason to
appear there.

**What it does not show.** That the changes are developmental. Yasmin's series
has its own collection structure and may carry its own confounds. Agreement
shows the pattern is not specific to our batch. That is what the reviewer asks
for. The claim goes no further.

**Bulk is a mixture.** A gene changing in bulk can reflect a shift in which cell
types are present rather than a change within any cell type. Pseudobulk shares
that property. The two sides are on the same footing. The composition section
measures how much of the agreement composition accounts for.

**The larval endpoint is matched by stage rather than by size.** Our larvae are
112 mm and sex-labile by our own staging. Every larval-stage animal in Yasmin's
series is smaller, the largest being 99 mm. Their 74–99 mm animals are already
classed as past the point where ovarian differentiation completes. The two series
draw the larval boundary at different sizes. The transformer endpoints match on
stage and on size.

```{r}
object_dir  = paste0(params$project_dir, "/results/", params$results_date,
                     "_no_doublets")
atlas_dir   = paste0(params$project_dir, "/results/", params$results_date,
                     "_atlas")
results_dir = paste0(params$project_dir, "/results/", params$results_date,
                     "_bulk_validation")
Robject_out_dir = paste0(results_dir, "/Robjects")
if (!dir.exists(Robject_out_dir)) dir.create(Robject_out_dir, recursive = TRUE)
data_dir    = paste0(params$project_dir, "/data")
```

```{r}
# The Yasmin reads were quantified with nf-core/rnaseq against
# GCF_010993605.1_kPetMar1.pri. Point nfcore_dir at that run's output directory.
# Either the top level or star_salmon/ itself is accepted.
#
# Length-scaled counts come first in the candidate list. They carry the
# transcript-length correction that makes gene-level Salmon estimates comparable
# across samples, and they are the right edgeR input without offsets. Plain
# gene_counts is the fallback.
nfcore_candidates = c(
  "star_salmon/salmon.merged.gene_counts_length_scaled.tsv",
  "salmon/salmon.merged.gene_counts_length_scaled.tsv",
  "star_salmon/salmon.merged.gene_counts.tsv",
  "salmon/salmon.merged.gene_counts.tsv",
  "salmon.merged.gene_counts_length_scaled.tsv",
  "salmon.merged.gene_counts.tsv")

nfcore_dir  = as.character(params$nfcore_dir)
count_path  = NA_character_

if (nzchar(nfcore_dir)) {
  hits = file.path(nfcore_dir, nfcore_candidates)
  hits = hits[file.exists(hits)]
  if (length(hits) > 0) count_path = hits[1]
}

have_bulk    = !is.na(count_path)
length_scaled = have_bulk && grepl("length_scaled", count_path)
```

`r if (!have_bulk && !nzchar(nfcore_dir)) "**The nfcore_dir parameter is not set.** Everything below is skipped. Set it to the nf-core/rnaseq output directory for the Yasmin data, either the run's top level or its star_salmon subdirectory." else ""`

`r if (!have_bulk && nzchar(nfcore_dir)) paste0("**No gene count matrix was found under ", nfcore_dir, ".** Everything below is skipped. Looked for: ", paste(nfcore_candidates, collapse = ", "), ". Found instead: ", paste(head(list.files(nfcore_dir), 20), collapse = ", "), ".") else ""`

`r if (have_bulk) paste0("Counts read from ", basename(count_path), if (length_scaled) ". The length-scaled matrix is the right edgeR input without offsets." else ". **This is the unscaled gene_counts matrix.** The transcript-length correction is absent. Use salmon.merged.gene_counts_length_scaled.tsv if the run produced one.") else ""`

# The Yasmin series

```{r, eval=have_bulk}
sample_meta = read.csv(paste0(data_dir, "/yasmin_PRJNA749754_sample_groups.csv"),
                       stringsAsFactors = FALSE)

# nf-core/rnaseq gene matrix: gene_id, gene_name, then one column per sample.
x = read.delim(count_path, check.names = FALSE, stringsAsFactors = FALSE)

id_cols = intersect(c("gene_id", "gene_name"), colnames(x))
if (!"gene_id" %in% id_cols)
  stop("no gene_id column in ", count_path, ". Columns found: ",
       paste(head(colnames(x), 10), collapse = ", "))

yasmin_counts = as.matrix(x[, setdiff(colnames(x), id_cols), drop = FALSE])
rownames(yasmin_counts) = x$gene_id

# Salmon estimates are non-integer. edgeR wants counts, and the length-scaled
# matrix is already on the count scale, so rounding is the only step needed.
yasmin_counts = round(yasmin_counts)
storage.mode(yasmin_counts) = "integer"

# nf-core sample names come from the samplesheet the run used. They may be run
# accessions, the paper's own animal names, or the two glued together, as in
# SRR15275627_F_JUV. Read the column names in all three forms and take the
# form and sample sheet column that resolve the most samples.
raw_ids = colnames(yasmin_counts)
has_acc = grepl("^[DES]RR[0-9]+", raw_ids)

id_forms = list(
  "as given"      = raw_ids,
  "run accession" = ifelse(has_acc, sub("^([DES]RR[0-9]+).*$", "\\1", raw_ids),
                           NA_character_),
  "name after the accession" =
                    ifelse(grepl("^[DES]RR[0-9]+_", raw_ids),
                           sub("^[DES]RR[0-9]+_", "", raw_ids), NA_character_))

id_candidates = c("run", "sample_name", "library_name", "suppT1_id")
id_candidates = intersect(id_candidates, colnames(sample_meta))

match_grid = expand.grid(form = names(id_forms), column = id_candidates,
                         stringsAsFactors = FALSE)
match_grid$n_matched = mapply(function(f, cl)
  sum(id_forms[[f]] %in% sample_meta[[cl]], na.rm = TRUE),
  match_grid$form, match_grid$column)

best          = match_grid[which.max(match_grid$n_matched), ]
sample_ids    = id_forms[[best$form]]
sample_id_col = best$column

if (best$n_matched < ncol(yasmin_counts))
  stop("could not match all ", ncol(yasmin_counts), " count columns to the ",
       "sample sheet. The best pairing was the '", best$form, "' form of the ",
       "column names against sample sheet column '", sample_id_col,
       "', matching ", best$n_matched, ". Unmatched: ",
       paste(raw_ids[!(sample_ids %in% sample_meta[[sample_id_col]])],
             collapse = ", "),
       ". Add the nf-core sample names to ",
       "data/yasmin_PRJNA749754_sample_groups.csv as a new column.")

absent = setdiff(sample_meta[[sample_id_col]], sample_ids)
if (length(absent) > 0)
  stop(length(absent), " sample(s) in the sample sheet are missing from the ",
       "count matrix: ", paste(absent, collapse = ", "),
       ". The nf-core run is incomplete, or its samplesheet omitted them.")

sample_meta = sample_meta[match(sample_ids, sample_meta[[sample_id_col]]), ]
stopifnot(identical(sample_meta[[sample_id_col]], sample_ids))

# Everything downstream selects columns with sample_meta$run, so put the count
# matrix on run accessions whatever form the nf-core names took.
stopifnot(!anyDuplicated(sample_meta$run))
colnames(yasmin_counts) = sample_meta$run

id_note = paste0("Count columns were matched to the sample sheet by their ",
                 best$form, ", against the `", sample_id_col, "` column. All ",
                 ncol(yasmin_counts), " samples resolved.")
```

`r if (have_bulk) id_note else ""`

```{r, eval=have_bulk}
# a numeric developmental position per animal, parsed from the recorded life
# stage rather than from the group label. The mid female group spans stage 1 to
# post-metamorphosis, so the label does not place an animal.
stage_score = function(s) {
  s = tolower(s)
  dplyr::case_when(
    grepl("post-metamorphosis", s)             ~ 8,
    grepl("metamorphosing stage", s)           ~ suppressWarnings(
                                                   as.numeric(sub(".*stage ([0-9]+).*",
                                                                  "\\1", s))),
    grepl("larval", s)                         ~ 0,
    grepl("upstream migrant", s)               ~ 10,
    TRUE                                       ~ NA_real_)
}

sample_meta$stage_score = stage_score(sample_meta$suppT1_life_stage)
stopifnot(!any(is.na(sample_meta$stage_score)))

series_tab = sample_meta %>%
  dplyr::group_by(group_textcounts) %>%
  dplyr::summarise(n = dplyr::n(),
                   length_mm = paste0(min(suppT1_length_mm), "-",
                                      max(suppT1_length_mm)),
                   stages = paste(sort(unique(suppT1_life_stage)), collapse = "; "),
                   .groups = "drop")

datatable(series_tab, rownames = FALSE, options = list(dom = "t", paging = FALSE))
```

## Stage correspondence

Our animals are 112 mm sex-labile larvae and 131 mm newly transformed
transformers. The matches below use the
per-animal records in Supplementary Table 1 rather than the group labels.

```{r, eval=have_bulk}
# the two stage-matched female transformers, selected by recorded stage rather
# than by group. Mid females span stage 1 to post-metamorphosis.
tf_match_runs = sample_meta$run[
  sample_meta$group_textcounts == "MF mid females" &
  sample_meta$stage_score >= 7]

if (length(tf_match_runs) != 2)
  stop("expected 2 stage-matched mid females at stage 7 or post-metamorphosis, ",
       "found ", length(tf_match_runs),
       ". Check suppT1_life_stage in the sample sheet.")

match_tab = tibble::tribble(
  ~ours,             ~yasmin,                                          ~basis,
  "larvae, 112 mm",  "undifferentiated larvae, 57-66 mm",              "both pre-differentiation by description; ours are sex-labile by our staging; the two series set the larval size boundary differently",
  "TM, 131 mm",      "mid males, 114-133 mm, stage 5 to post-metamorphosis", "lab confirmed; lengths and stages agree",
  "TF, 131 mm",      paste0(paste(sample_meta$sample_name[
                                    match(tf_match_runs, sample_meta$run)],
                                  collapse = ", "), ", 129-130 mm"),
                                                                       "stage 7 and post-metamorphosis mid females; mirrors the male match")

datatable(match_tab, rownames = FALSE, options = list(dom = "t", paging = FALSE))
```

# Pseudobulk

Raw counts are summed across all cells within each library, giving four samples.
Corrected and per-cell normalized values are not used. A bulk library measures
summed raw counts.

```{r, eval=have_bulk}
pseudo_file = paste0(Robject_out_dir, "/pseudobulk.Rds")

if (file.exists(pseudo_file)) {
  pseudobulk = readRDS(pseudo_file)
} else {
  all.sce = readRDS(paste0(object_dir, "/all.sce.Rds"))
  universe = Reduce(intersect, lapply(all.sce, rownames))
  pseudobulk = vapply(all.sce, function(x)
    Matrix::rowSums(counts(x[universe, ])), numeric(length(universe)))
  rownames(pseudobulk) = universe
  saveRDS(pseudobulk, file = pseudo_file)
}

library_meta = data.frame(
  batch   = c("UND1_C00004336", "UND2_C00004362", "TF_C00004337", "TM_C00004374"),
  library = c("Undifferentiated.1", "Undifferentiated.2", "Female", "Male"),
  stage   = c("larva", "larva", "transformer", "transformer"),
  stringsAsFactors = FALSE)

colnames(pseudobulk) = library_meta$library[
  match(colnames(pseudobulk), library_meta$batch)]
stopifnot(!any(is.na(colnames(pseudobulk))))
```

`r if (have_bulk) paste0("Pseudobulk over ", format(nrow(pseudobulk), big.mark = ","), " genes and ", ncol(pseudobulk), " libraries.") else ""`

## Shared gene set

```{r, eval=have_bulk}
shared = intersect(rownames(pseudobulk), rownames(yasmin_counts))

overlap_tab = data.frame(
  set = c("our single-cell genes", "Yasmin quantified genes", "shared"),
  n   = c(nrow(pseudobulk), nrow(yasmin_counts), length(shared)))

datatable(overlap_tab, rownames = FALSE, options = list(dom = "t"))

if (length(shared) < 0.5 * nrow(pseudobulk))
  stop("only ", length(shared), " of ", nrow(pseudobulk), " genes are shared. ",
       "Both sides should be NCBI kPetMar1 gene ids, so a low overlap means the ",
       "nf-core run used a different annotation from the single-cell counts. ",
       "Diagnose before continuing.")
```

`r if (have_bulk) paste0("**", format(length(shared), big.mark = ","), " genes are shared.** Everything below runs on that set. Genes are matched by identifier. Both sides come from the same assembly, so no ortholog assignment is involved.") else ""`

# Yasmin stage effects

```{r, eval=have_bulk}
y = DGEList(counts = yasmin_counts[shared, ], samples = sample_meta)
keep = filterByExpr(y, group = sample_meta$group_textcounts)
y = y[keep, , keep.lib.sizes = FALSE]
y = calcNormFactors(y)

tested = rownames(y)
```

`r if (have_bulk) paste0(format(length(tested), big.mark = ","), " genes pass filterByExpr in the Yasmin series and carry through to the comparison.") else ""`

```{r, eval=have_bulk}
# Four contrasts: an endpoint-matched two-group test and an ordered trend on
# each sex axis. The endpoint tests supply effect sizes for the correlation. The
# trends supply per-gene significance for any gene named individually.
run_two_group = function(dge, run_a, run_b, label) {
  sub = dge[, c(run_a, run_b)]
  grp = factor(c(rep("early", length(run_a)), rep("late", length(run_b))),
               levels = c("early", "late"))
  des = model.matrix(~grp)
  sub = estimateDisp(sub, des)
  fit = glmQLFit(sub, des)
  res = glmQLFTest(fit, coef = 2)
  topTags(res, n = Inf, sort.by = "none")$table %>%
    tibble::rownames_to_column("Gene") %>%
    dplyr::select(Gene, logFC, PValue, FDR) %>%
    dplyr::rename_with(~paste0(label, "_", .x), -Gene)
}

run_trend = function(dge, runs, score, label) {
  sub = dge[, runs]
  des = model.matrix(~score)
  sub = estimateDisp(sub, des)
  fit = glmQLFit(sub, des)
  res = glmQLFTest(fit, coef = 2)
  topTags(res, n = Inf, sort.by = "none")$table %>%
    tibble::rownames_to_column("Gene") %>%
    dplyr::select(Gene, logFC, PValue, FDR) %>%
    dplyr::rename_with(~paste0(label, "_", .x), -Gene)
}

g = sample_meta$group_textcounts
runs_ud = sample_meta$run[g == "UD undifferentiated larvae"]
runs_pm = sample_meta$run[g == "PM prospective males"]
runs_em = sample_meta$run[g == "EM early males"]
runs_mm = sample_meta$run[g == "MM mid males"]
runs_ef = sample_meta$run[g == "EF early females"]
runs_mf = sample_meta$run[g == "MF mid females"]

# male endpoint and male ordered trend over the four ordered male-axis groups
male_end = run_two_group(y, runs_ud, runs_mm, "male_end")

male_runs  = c(runs_ud, runs_pm, runs_em, runs_mm)
male_score = c(rep(0, length(runs_ud)), rep(1, length(runs_pm)),
               rep(2, length(runs_em)), rep(3, length(runs_mm)))
male_trend = run_trend(y, male_runs, male_score, "male_trend")

# female endpoint uses the two stage-matched mid females, not the whole group
fem_end = run_two_group(y, runs_ud, tf_match_runs, "fem_end")

fem_runs  = c(runs_ud, runs_ef, runs_mf)
fem_score = c(rep(0, length(runs_ud)), rep(1, length(runs_ef)),
              2 + rank(sample_meta$stage_score[match(runs_mf, sample_meta$run)],
                       ties.method = "min") / length(runs_mf))
fem_trend = run_trend(y, fem_runs, fem_score, "fem_trend")

yasmin_res = Reduce(function(a, b) dplyr::full_join(a, b, by = "Gene"),
                    list(male_end, male_trend, fem_end, fem_trend))
```

```{r, eval=have_bulk}
contrast_tab = data.frame(
  contrast = c("male endpoint", "male trend", "female endpoint", "female trend"),
  design   = c(paste0("UD (n=", length(runs_ud), ") vs mid males (n=",
                      length(runs_mm), ")"),
               paste0("UD, PM, EM, MM ordered (n=", length(male_runs), ")"),
               paste0("UD (n=", length(runs_ud), ") vs stage-matched mid females (n=",
                      length(tf_match_runs), ")"),
               paste0("UD, EF, MF ordered (n=", length(fem_runs), ")")),
  n_FDR_05 = c(sum(yasmin_res$male_end_FDR   < 0.05, na.rm = TRUE),
               sum(yasmin_res$male_trend_FDR < 0.05, na.rm = TRUE),
               sum(yasmin_res$fem_end_FDR    < 0.05, na.rm = TRUE),
               sum(yasmin_res$fem_trend_FDR  < 0.05, na.rm = TRUE)))

datatable(contrast_tab, rownames = FALSE, options = list(dom = "t", paging = FALSE))
```

The endpoint contrasts have little power. The female one has the least, at 2
against 2. The correlation below uses effect sizes alone. Noise in the Yasmin
estimates pulls agreement toward zero, so low power makes that test
conservative rather than invalid. Per-gene significance comes from the trends.

# Our stage effect

Two larval libraries against one transformer library per sex is not a formal
test. The log fold change is used as an effect size. No p-values are reported
from it.

```{r, eval=have_bulk}
p = DGEList(counts = pseudobulk[shared, ][tested, ])
p = calcNormFactors(p)
plog = cpm(p, log = TRUE, prior.count = 1)

larva_mean = rowMeans(plog[, c("Undifferentiated.1", "Undifferentiated.2")])

ours = data.frame(
  Gene       = rownames(plog),
  ours_male  = plog[, "Male"]   - larva_mean,
  ours_female = plog[, "Female"] - larva_mean,
  stringsAsFactors = FALSE)
```

# The comparison

```{r, eval=have_bulk}
cmp = ours %>% dplyr::inner_join(yasmin_res, by = "Gene")

top_ns = as.integer(trimws(strsplit(as.character(params$top_n_list), ",")[[1]]))

# Spearman with a bootstrap CI. cor.test gives a p-value but no interval, and
# the interval is the quantity worth reporting.
spearman_ci = function(x, y, n_boot = params$n_boot, seed = 1) {
  ok = is.finite(x) & is.finite(y)
  x = x[ok]; y = y[ok]
  r = suppressWarnings(cor(x, y, method = "spearman"))
  set.seed(seed)
  bs = replicate(n_boot, {
    i = sample(length(x), replace = TRUE)
    suppressWarnings(cor(x[i], y[i], method = "spearman"))
  })
  c(rho = r, lo = unname(quantile(bs, .025, na.rm = TRUE)),
    hi = unname(quantile(bs, .975, na.rm = TRUE)), n = length(x))
}

# genome-wide, then restricted to our strongest stage genes. The genome-wide
# value is a floor rather than a result. Two gonad series compared on fold
# changes agree before any shared stage regulation is involved: both transitions
# move cell composition the same way, library-size normalization pushes the rest
# of the transcriptome down in both, and genes at the detection floor in the
# larval samples can only move up. The top-gene value is read against that floor.
#
# Sign agreement is scored genome-wide for the same reason. That rate, not 50%,
# is the honest null for the top-gene sets, so both binomial tests are reported.
agreement_for = function(ours_col, yas_col, axis) {
  rows = list()
  gw = spearman_ci(cmp[[ours_col]], cmp[[yas_col]])

  gw_agree = sign(cmp[[ours_col]]) == sign(cmp[[yas_col]])
  gw_agree = gw_agree[is.finite(gw_agree)]
  bg_rate  = mean(gw_agree)

  rows[[1]] = data.frame(
    axis = axis, set = "genome-wide", n = gw[["n"]],
    rho = gw[["rho"]], lo = gw[["lo"]], hi = gw[["hi"]],
    pct_same_sign = 100 * bg_rate,
    binom_p_vs_50 = binom.test(sum(gw_agree), length(gw_agree),
                               p = 0.5)$p.value,
    binom_p_vs_bg = NA_real_)

  # binom.test needs a probability strictly inside (0, 1)
  bg_p = min(max(bg_rate, 1e-6), 1 - 1e-6)

  ord = order(abs(cmp[[ours_col]]), decreasing = TRUE)
  for (n in top_ns) {
    idx = ord[seq_len(min(n, nrow(cmp)))]
    s = spearman_ci(cmp[[ours_col]][idx], cmp[[yas_col]][idx])
    agree = sign(cmp[[ours_col]][idx]) == sign(cmp[[yas_col]][idx])
    agree = agree[is.finite(agree)]
    rows[[length(rows) + 1]] = data.frame(
      axis = axis, set = paste0("top ", n), n = length(idx),
      rho = s[["rho"]], lo = s[["lo"]], hi = s[["hi"]],
      pct_same_sign = 100 * mean(agree),
      binom_p_vs_50 = binom.test(sum(agree), length(agree), p = 0.5)$p.value,
      binom_p_vs_bg = binom.test(sum(agree), length(agree), p = bg_p)$p.value)
  }
  dplyr::bind_rows(rows)
}

agreement = dplyr::bind_rows(
  agreement_for("ours_male",   "male_end_logFC", "male, endpoint"),
  agreement_for("ours_male",   "male_trend_logFC", "male, trend"),
  agreement_for("ours_female", "fem_end_logFC",  "female, endpoint"),
  agreement_for("ours_female", "fem_trend_logFC", "female, trend"))

write.csv(agreement, file = paste0(results_dir, "/bulk_agreement.csv"),
          row.names = FALSE)

datatable(agreement %>% mutate(across(where(is.numeric), ~signif(., 3))),
          rownames = FALSE, options = list(dom = "t", paging = FALSE))
```

`r if (have_bulk) { a = agreement[agreement$axis == "male, endpoint", ]; gw = a$rho[a$set == "genome-wide"]; tp = a$rho[a$set == paste0("top ", top_ns[1])]; sg = a$pct_same_sign[a$set == paste0("top ", top_ns[1])]; paste0("On the male axis the genome-wide correlation is ", signif(gw, 3), ". Across our top ", top_ns[1], " stage genes it is ", signif(tp, 3), ", with ", signif(sg, 3), "% sign agreement. Both values belong in the write-up. The genome-wide value is the floor that the top-gene value is read against. The permutation below turns that reading into a p-value.") } else ""`

## A label permutation null

The two numbers above cannot be read as a test on their own. The top-gene set is
picked by the absolute value of our own log fold change, which is one of the two
variables being correlated. Selecting large effects widens the spread of that
variable and lowers its relative measurement error, which raises the correlation
by itself. Selecting on a noisy estimate also picks up genes whose effect we
overstated, and those regress toward the mean in Yasmin, which lowers it. Both
act at once.

The permutation holds all of that fixed. Group labels are reassigned at random
among the 28 animals, the same contrast shape is refitted, and the agreement
statistics are recomputed against our unchanged effect sizes and our unchanged
top-gene list. Whatever the selection does to the observed value, it does to
every null value in the same way. What is left is whether Yasmin's stage
contrast agrees with our genes more than an arbitrary contrast of the same shape
drawn from the same animals.

Labels are drawn from the whole series rather than shuffled within the contrast.
Shuffling within it has too few distinct splits to resolve anything: 2 against 6
gives 28 arrangements and 2 against 2 gives 6, so the smallest p-value
reachable on the female axis would be 0.17.

```{r, eval=have_bulk}
n_perm_labels = as.integer(params$n_perm_labels)

perm_null = function(dge, pool_runs, n_a, n_b, ours_col, top_n, n_perm,
                     seed = 1) {
  ours_vec = cmp[[ours_col]]
  idx_top  = order(abs(ours_vec), decreasing = TRUE)[seq_len(min(top_n,
                                                                 nrow(cmp)))]
  set.seed(seed)
  res = vector("list", n_perm)
  for (i in seq_len(n_perm)) {
    pick = sample(pool_runs, n_a + n_b)
    tab  = try(run_two_group(dge, pick[seq_len(n_a)], pick[n_a + seq_len(n_b)],
                             "perm"), silent = TRUE)
    if (inherits(tab, "try-error")) next
    lfc   = tab$perm_logFC[match(cmp$Gene, tab$Gene)]
    agree = sign(ours_vec[idx_top]) == sign(lfc[idx_top])
    res[[i]] = data.frame(
      rho_gw   = suppressWarnings(cor(ours_vec, lfc, method = "spearman",
                                      use = "complete.obs")),
      rho_top  = suppressWarnings(cor(ours_vec[idx_top], lfc[idx_top],
                                      method = "spearman",
                                      use = "complete.obs")),
      pct_sign = 100 * mean(agree, na.rm = TRUE))
  }
  dplyr::bind_rows(res)
}

# cached on a signature, since the permutation is the slow step in the document
perm_file = paste0(Robject_out_dir, "/label_permutation_null.Rds")
perm_sig  = list(n_perm  = n_perm_labels,
                 top_n   = top_ns[1],
                 n_genes = nrow(cmp),
                 runs    = sort(sample_meta$run),
                 sizes   = c(length(runs_ud), length(runs_mm),
                             length(tf_match_runs)))

cached = if (file.exists(perm_file)) readRDS(perm_file) else NULL

if (!is.null(cached) && identical(cached$signature, perm_sig)) {
  perm_res = cached$perm_res
} else {
  perm_res = list(
    "male, endpoint"   = perm_null(y, sample_meta$run, length(runs_ud),
                                   length(runs_mm), "ours_male", top_ns[1],
                                   n_perm_labels, seed = 1),
    "female, endpoint" = perm_null(y, sample_meta$run, length(runs_ud),
                                   length(tf_match_runs), "ours_female",
                                   top_ns[1], n_perm_labels, seed = 2))
  saveRDS(list(signature = perm_sig, perm_res = perm_res), file = perm_file)
}
```

```{r, eval=have_bulk}
# one-sided: the question is whether the observed statistic sits above the null
emp_p = function(null_vals, obs) {
  ok = is.finite(null_vals)
  (1 + sum(null_vals[ok] >= obs)) / (1 + sum(ok))
}

perm_tab = dplyr::bind_rows(lapply(names(perm_res), function(ax) {
  nd  = perm_res[[ax]]
  obs = agreement[agreement$axis == ax, ]
  o   = c(obs$rho[obs$set == "genome-wide"],
          obs$rho[obs$set == paste0("top ", top_ns[1])],
          obs$pct_same_sign[obs$set == paste0("top ", top_ns[1])])
  nl  = list(nd$rho_gw, nd$rho_top, nd$pct_sign)

  data.frame(
    axis        = ax,
    statistic   = c("rho, genome-wide", paste0("rho, top ", top_ns[1]),
                    paste0("% same sign, top ", top_ns[1])),
    observed    = o,
    null_median = vapply(nl, median, numeric(1), na.rm = TRUE),
    null_95th   = vapply(nl, function(v) unname(quantile(v, .95, na.rm = TRUE)),
                         numeric(1)),
    p_perm      = mapply(emp_p, nl, o),
    n_perm      = vapply(nl, function(v) sum(is.finite(v)), integer(1)))
}))

write.csv(perm_tab, file = paste0(results_dir, "/bulk_permutation_null.csv"),
          row.names = FALSE)

datatable(perm_tab %>% mutate(across(where(is.numeric), ~signif(., 3))),
          rownames = FALSE, options = list(dom = "t", paging = FALSE))
```

`r if (have_bulk) paste0("The smallest p-value reachable with ", n_perm_labels, " permutations is ", signif(1 / (1 + n_perm_labels), 2), ". A statistic at that value is reported as a floor rather than as its true p-value.") else ""`

```{r, eval=have_bulk, fig.width=10, fig.height=4}
perm_long = dplyr::bind_rows(lapply(names(perm_res), function(ax)
  data.frame(axis = ax, rho_top = perm_res[[ax]]$rho_top)))

obs_long = dplyr::bind_rows(lapply(names(perm_res), function(ax) {
  obs = agreement[agreement$axis == ax, ]
  data.frame(axis = ax, observed = obs$rho[obs$set == paste0("top ", top_ns[1])])
}))

p_perm = ggplot(perm_long, aes(x = rho_top)) +
  geom_histogram(bins = 40, fill = "grey75", colour = NA) +
  geom_vline(data = obs_long, aes(xintercept = observed),
             colour = "#c1272d", linewidth = .7) +
  facet_wrap(~axis, scales = "free_y") +
  labs(x = paste0("Spearman rho across our top ", top_ns[1], " stage genes"),
       y = "permutations",
       title = "Observed agreement against random contrasts of the same shape") +
  theme_bw()

p_perm

ggsave(p_perm, file = paste0(results_dir, "/bulk_permutation_null.png"),
       width = 10, height = 4, dpi = 300)
```

```{r, eval=have_bulk, fig.width=11, fig.height=5.5}
scatter_for = function(ours_col, yas_col, title, n_label = 15) {
  d = cmp %>%
    dplyr::select(Gene, ours = dplyr::all_of(ours_col),
                  yasmin = dplyr::all_of(yas_col)) %>%
    dplyr::filter(is.finite(ours), is.finite(yasmin))
  lab = d %>% dplyr::slice_max(abs(ours), n = n_label, with_ties = FALSE)

  ggplot(d, aes(x = ours, y = yasmin)) +
    geom_hline(yintercept = 0, colour = "grey80") +
    geom_vline(xintercept = 0, colour = "grey80") +
    geom_point(alpha = .2, size = .5) +
    geom_point(data = lab, colour = "#c1272d", size = 1.2) +
    geom_text_repel(data = lab, aes(label = Gene), size = 2.5, max.overlaps = 20) +
    geom_smooth(method = "lm", se = FALSE, linewidth = .5, colour = "#004488") +
    labs(x = "our pseudobulk log2FC, larva to transformer",
         y = "Yasmin log2FC", title = title) +
    theme_bw()
}

p_male = scatter_for("ours_male", "male_end_logFC", "Male axis, endpoint-matched")
p_fem  = scatter_for("ours_female", "fem_end_logFC", "Female axis, endpoint-matched")

p_male + p_fem

ggsave(p_male + p_fem, file = paste0(results_dir, "/bulk_agreement_scatter.png"),
       width = 11, height = 5.5, dpi = 300)
```

# Composition

A bulk change can arise because a cell type became more abundant rather than
because any cell changed. If agreement is concentrated in cell type marker
genes, the shared signal is largely compositional. Both datasets would be seeing
the same shift in which cell types are present. If agreement holds among
non-markers too, expression is changing within cell types in both datasets.

Either outcome supports the reviewer's point. The result decides how the claim
is worded.

This runs on all four contrasts. The composition question is about which genes
carry the agreement, so it applies to the female axis and to the trends on the
same terms as to the male endpoint.

The marker set is small, and comparing it against the fifteen thousand genes
outside it compares a noisy estimate against a precise one. Markers are also
highly expressed and cell type restricted by construction, so they are not a
random draw. The control for both is a size-matched null: gene sets of the same
size drawn at random, giving the range of correlations any set that size
produces.

```{r, eval=have_bulk}
marker_path = paste0(data_dir, "/", params$marker_file)
have_markers = file.exists(marker_path)

axis_defs = list(
  "male, endpoint"   = c(ours = "ours_male",   yas = "male_end_logFC"),
  "male, trend"      = c(ours = "ours_male",   yas = "male_trend_logFC"),
  "female, endpoint" = c(ours = "ours_female", yas = "fem_end_logFC"),
  "female, trend"    = c(ours = "ours_female", yas = "fem_trend_logFC"))

if (have_markers) {
  km = read.csv(marker_path, stringsAsFactors = FALSE)
  marker_genes = unique(c(km$loc_id, km$symbol))
  marker_genes = marker_genes[!is.na(marker_genes) & nzchar(marker_genes)]

  cmp$is_marker = cmp$Gene %in% marker_genes

  comp_tab = dplyr::bind_rows(lapply(names(axis_defs), function(ax) {
    cols = axis_defs[[ax]]
    dplyr::bind_rows(lapply(c(TRUE, FALSE), function(m) {
      d = cmp[cmp$is_marker == m, ]
      s = spearman_ci(d[[cols[["ours"]]]], d[[cols[["yas"]]]])
      data.frame(axis = ax,
                 gene_set = if (m) "cell type markers" else "all other genes",
                 n = s[["n"]], rho = s[["rho"]], lo = s[["lo"]], hi = s[["hi"]])
    }))
  }))

  datatable(comp_tab %>% mutate(across(where(is.numeric), ~signif(., 3))),
            rownames = FALSE, options = list(dom = "t", paging = FALSE))
}
```

```{r, eval=have_bulk}
# the size-matched null: where does the marker correlation fall among random
# gene sets of the same size?
if (have_markers) {
  n_rand = as.integer(params$n_rand_sets)

  rand_tab = dplyr::bind_rows(lapply(names(axis_defs), function(ax) {
    cols = axis_defs[[ax]]
    x  = cmp[[cols[["ours"]]]]
    yv = cmp[[cols[["yas"]]]]
    ok = is.finite(x) & is.finite(yv)
    x  = x[ok]; yv = yv[ok]; mk = cmp$is_marker[ok]

    obs = suppressWarnings(cor(x[mk], yv[mk], method = "spearman"))
    set.seed(1)
    nulls = replicate(n_rand, {
      i = sample(length(x), sum(mk))
      suppressWarnings(cor(x[i], yv[i], method = "spearman"))
    })

    data.frame(axis = ax, n_markers = sum(mk), marker_rho = obs,
               random_median = median(nulls, na.rm = TRUE),
               random_95th = unname(quantile(nulls, .95, na.rm = TRUE)),
               p_rand = (1 + sum(nulls >= obs, na.rm = TRUE)) /
                        (1 + sum(is.finite(nulls))))
  }))

  write.csv(rand_tab, file = paste0(results_dir, "/bulk_composition_null.csv"),
            row.names = FALSE)

  datatable(rand_tab %>% mutate(across(where(is.numeric), ~signif(., 3))),
            rownames = FALSE, options = list(dom = "t", paging = FALSE))
}
```

`r if (have_bulk && exists("have_markers") && !have_markers) paste0("**The composition check is skipped.** The marker file ", params$marker_file, " was not found in the data directory.") else ""`

`r if (have_bulk && exists("rand_tab")) { n_hi = sum(rand_tab$p_rand < 0.05, na.rm = TRUE); paste0("Marker genes agree more than a random set of ", rand_tab$n_markers[1], " genes on ", n_hi, " of the ", nrow(rand_tab), " contrasts. ", if (n_hi >= 3) "The shared signal is partly compositional on most contrasts. Describe it as a shift in gonad composition together with a shift in expression." else if (n_hi == 0) "Agreement is not concentrated in marker genes on any contrast. Expression appears to change within cell types in both datasets." else "The contrasts disagree, so the compositional contribution is not resolved. Report the table rather than a verdict.") } else ""`

# Export

```{r, eval=have_bulk}
out = cmp %>%
  dplyr::select(Gene, ours_male, ours_female,
                dplyr::starts_with("male_"), dplyr::starts_with("fem_"))

write.csv(out, file = paste0(results_dir, "/bulk_vs_pseudobulk_per_gene.csv"),
          row.names = FALSE)

datatable(head(out %>% mutate(across(where(is.numeric), ~signif(., 3))), 100),
          rownames = FALSE, filter = "top")
```

# Summary

```{r, eval=have_bulk}
top_set = paste0("top ", top_ns[1])

row_for = function(ax, set) agreement[agreement$axis == ax &
                                        agreement$set == set, ]
male_row = row_for("male, endpoint", top_set)
male_gw  = row_for("male, endpoint", "genome-wide")
fem_row  = row_for("female, endpoint", top_set)
fem_gw   = row_for("female, endpoint", "genome-wide")

perm_for = function(ax, stat) perm_tab$p_perm[perm_tab$axis == ax &
                                                perm_tab$statistic == stat]

summary_tbl = data.frame(
  question = c("shared genes tested",
               "male axis, rho on top genes",
               "male axis, rho genome-wide (the floor)",
               "male axis, % same sign on top genes",
               "male axis, % same sign genome-wide (the null)",
               "male axis, permutation p for rho on top genes",
               "female axis, rho on top genes",
               "female axis, rho genome-wide (the floor)",
               "female axis, % same sign on top genes",
               "female axis, % same sign genome-wide (the null)",
               "female axis, permutation p for rho on top genes",
               "female animals at the matched endpoint"),
  answer = c(format(nrow(cmp), big.mark = ","),
             signif(male_row$rho, 3),
             signif(male_gw$rho, 3),
             paste0(signif(male_row$pct_same_sign, 3), "%"),
             paste0(signif(male_gw$pct_same_sign, 3), "%"),
             signif(perm_for("male, endpoint", paste0("rho, ", top_set)), 3),
             signif(fem_row$rho, 3),
             signif(fem_gw$rho, 3),
             paste0(signif(fem_row$pct_same_sign, 3), "%"),
             paste0(signif(fem_gw$pct_same_sign, 3), "%"),
             signif(perm_for("female, endpoint", paste0("rho, ", top_set)), 3),
             as.character(length(tf_match_runs))))

datatable(summary_tbl, rownames = FALSE,
          options = list(dom = "t", ordering = FALSE, paging = FALSE))
```

## What this does not answer

Agreement shows the pattern reproduces in a series without our July-to-November
batch. It does not separate stage from batch within our data.

The female result rests on two stage-matched animals. It supports the male axis
rather than replicating it independently.

The larval endpoint is matched by stage description rather than by size. Our
larvae are sex-labile by our own staging and larger than every larval animal in
the comparison series. The two series set the larval boundary at different sizes.

The permutation asks whether Yasmin's stage contrast agrees with our genes more
than an arbitrary contrast drawn from the same animals. It does not rule out a
confound shared by both studies. Anything that moves gonad composition the same
way in both series would pass it.

# Session info

```{r}
sessionInfo()
```
