---
title: "17_germ_pseudotime_tscan"
author: "Stephanie Hickey"
date: "`r Sys.Date()`"
output:
  html_document:
    code_folding: hide
    toc: true
    toc_float:
      toc_collapsed: true
params:
    project_dir: /mnt/ufs18/rs-013/bioinformaticsCore/projects/chong_davidson/BCC125_sea_lamprey_scRNA
    project_name: BCC125_sea_lamprey_scRNA
    lab: Weiming Li
    results_date: "20260813"
    lineage: "female"
    female_clusters: "5"
    male_clusters: "12_transformer,15,16,17,18"
    all_clusters: "5,12_transformer,15,16,17,18"
    female_root_rule: "min:Oocyte"
    male_root_rule: "max:Primordial germ cells"
    all_root_rule: "max:Primordial germ cells"
    min_detect_pct: 5
    marker_file: canonical_markers_S1_only_2026-08-14.csv
    split_by_subcluster: "5"
    sub_k: 0
    sub_k_range: "5,10,15,20,25,30,40,50"
    sub_n_iterations: 10
    root_node: ""
    min_library_cells: 50
    fdr_cut: 0.01
    lfc_thresh: 1
    go_pval: 0.05
    genome_dir: /mnt/research/bioinformaticsCore/shared/Genomes/Petromyzon_marinus
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

```{r, echo=FALSE, results="asis"}
cat("\n")
cat("#### Lab:", params$lab, " \n")
cat("#### Project:", params$project_name, " \n")
cat("#### Lineage:", params$lineage, " \n")
```

```{r}
library(tidyverse)
library(SingleCellExperiment)
library(scater)
library(scran)
library(batchelor)
library(BiocSingular)
library(TSCAN)
library(patchwork)
library(DT)
library(ggraph)
library(tidygraph)
library(scatterpie)
source(paste0(params$project_dir, "/src/scRNA_seq_functions.R"))
```

TSCAN pseudotime for the `r params$lineage` germ cell clusters. The lineage is
`female`, `male`, or `all`, which orders every germ cluster in one tree.
`18_germ_pseudotime_monocle3.Rmd` computes pseudotime for the same cells with
monocle3, and `knit_17_18_pseudotime.sh` knits both documents for each lineage.

Clusters named in `split_by_subcluster` are subclustered within this document
and enter the tree as their subclusters.

Stage and collection date vary together in this dataset, so the ordering is not
shown to be developmental. The library composition of each node and the variance
in pseudotime explained by library are reported with the result.

```{r}
lineage = as.character(params$lineage)
if (!lineage %in% c("female", "male", "all"))
  stop("lineage must be 'female', 'male' or 'all', not '", lineage, "'")

clusters = trimws(strsplit(as.character(switch(lineage,
  female = params$female_clusters,
  male   = params$male_clusters,
  all    = params$all_clusters)), ",")[[1]])
clusters = clusters[nzchar(clusters)]

object_dir      = paste0(params$project_dir, "/results/", params$results_date,
                         "_no_doublets")
atlas_dir       = paste0(params$project_dir, "/results/", params$results_date,
                         "_atlas")
results_dir     = paste0(params$project_dir, "/results/", params$results_date,
                         "_pseudotime_tscan_", lineage)
Robject_out_dir = paste0(results_dir, "/Robjects")
if (!dir.exists(Robject_out_dir)) dir.create(Robject_out_dir, recursive = TRUE)

label_file = paste0(atlas_dir, "/cell_cluster_labels.csv")
if (!file.exists(label_file))
  stop("missing ", label_file, ". Run 13_atlas_and_stage_correspondence.Rmd first.")

cell_labels      = read.csv(label_file, stringsAsFactors = FALSE)
all.sce          = readRDS(paste0(object_dir, "/all.sce.Rds"))
gene_description = read.csv(paste0(object_dir, "/gene_descritption.csv"),
                            stringsAsFactors = FALSE) %>%
  dplyr::distinct(Gene, .keep_all = TRUE)

lib_cols = c("Female" = "#FF6DB6", "Male" = "#006ddb",
             "Undifferentiated.1" = "#009292", "Undifferentiated.2" = "#b66ddf")
```

```{r}
gaf_file      = file.path(params$genome_dir,
                          "GCF_010993605.1_kPetMar1.pri_gene_ontology.gaf")
go_names_file = file.path(params$genome_dir, "GOs_names.txt")
have_go       = file.exists(gaf_file) && file.exists(go_names_file)
```

```{r, eval=have_go}
gaf = read.delim(gaf_file, header = FALSE, comment = "!") %>%
  dplyr::filter(!grepl("NOT", V4)) %>%
  dplyr::select(V3, V5) %>%
  dplyr::rename(GOID = "V5", Gene = "V3") %>%
  dplyr::distinct()

# terms too small are noise, terms too large are uninformative
keep_go = gaf %>%
  dplyr::count(GOID) %>%
  dplyr::filter(n > 10, n < 200) %>%
  dplyr::pull(GOID)
gaf = gaf %>% dplyr::filter(GOID %in% keep_go)

go_names = read.delim(go_names_file, header = FALSE)
colnames(go_names) = c("GOID", "GOName")
```

`r if (!have_go) paste0("**GO enrichment is skipped throughout.** The GO files were not found under ", params$genome_dir, ".") else ""`

# Cells

```{r}
absent = setdiff(clusters, cell_labels$final_label)
if (length(absent) > 0)
  stop("cluster(s) ", paste(absent, collapse = ", "), " are not in ", label_file,
       ". Labels present: ",
       paste(sort(unique(cell_labels$final_label)), collapse = ", "),
       ". Set the ", lineage, "_clusters parameter from the current labelling.")

lineage_key = cell_labels %>% dplyr::filter(final_label %in% clusters)

comp_tab = lineage_key %>%
  dplyr::count(final_label, cell_type, library, name = "n") %>%
  tidyr::pivot_wider(names_from = library, values_from = n, values_fill = 0) %>%
  dplyr::mutate(Total = rowSums(dplyr::across(dplyr::any_of(names(lib_cols))))) %>%
  dplyr::arrange(match(final_label, clusters))

write.csv(comp_tab, file = paste0(results_dir, "/cluster_composition.csv"),
          row.names = FALSE)

datatable(comp_tab, rownames = FALSE, options = list(dom = "t", paging = FALSE))
```

`r paste0(format(nrow(lineage_key), big.mark = ","), " cells from cluster(s) ", paste(clusters, collapse = ", "), ".")`

# Corrected space

Libraries contributing fewer than `r params$min_library_cells` cells are dropped.
With more than one library left, the space is `fastMNN` output computed on these
cells alone. With one library, it is PCA on that library.

```{r}
obj_file     = paste0(Robject_out_dir, "/lineage_objects.Rds")
obj_sig_file = paste0(Robject_out_dir, "/lineage_objects_signature.txt")
obj_sig      = paste(lineage, paste(sort(clusters), collapse = ","),
                     nrow(lineage_key), params$min_library_cells, sep = "|")

if (file.exists(obj_file) && file.exists(obj_sig_file) &&
    identical(readLines(obj_sig_file, warn = FALSE)[1], obj_sig)) {
  objs = readRDS(obj_file)
} else {
  objs = build_lineage_sce(all.sce, lineage_key$barcode,
                           min_library_cells = params$min_library_cells)
  saveRDS(objs, file = obj_file)
  writeLines(obj_sig, obj_sig_file)
}

merged   = objs$corrected
combined = objs$combined

# metadata attached by barcode, never by position
key_idx = match(colnames(merged), lineage_key$barcode)
stopifnot(!anyNA(key_idx))
merged$library     = lineage_key$library[key_idx]
merged$stage       = lineage_key$stage[key_idx]
merged$final_label = lineage_key$final_label[key_idx]
merged$cell_type   = lineage_key$cell_type[key_idx]

lib_tab = data.frame(batch   = names(objs$n_by_library),
                     n_cells = as.integer(objs$n_by_library),
                     kept    = names(objs$n_by_library) %in% objs$kept_batches,
                     stringsAsFactors = FALSE) %>%
  dplyr::left_join(dplyr::distinct(cell_labels, batch, library), by = "batch")

datatable(lib_tab, rownames = FALSE, options = list(dom = "t"))
```

`r n_kept = length(objs$kept_batches); paste0(format(ncol(merged), big.mark = ","), " cells from ", n_kept, " librar", if (n_kept == 1) "y" else "ies", " are kept.")`

# Nodes

TSCAN builds a minimum spanning tree over node centroids, so it needs at least
two nodes.

```{r}
split_clusters = trimws(strsplit(as.character(params$split_by_subcluster), ",")[[1]])
split_clusters = intersect(split_clusters[nzchar(split_clusters)], clusters)
have_split     = length(split_clusters) > 0
```

`r if (have_split) paste0("Cluster(s) ", paste(split_clusters, collapse = ", "), " are subclustered in this lineage's corrected space, using only that cluster's cells, and every other cluster is one node. k is the most reproducible value in a subsampling stability sweep that gives at least two clusters, unless the sub_k parameter is set. The space depends on the lineage, so cluster 5 subclusters from the female and all runs are computed separately and need not match.") else paste0("The ", length(clusters), " selected clusters are the nodes.")`

```{r, eval=have_split}
k_range = as.integer(trimws(strsplit(as.character(params$sub_k_range), ",")[[1]]))
stopifnot(length(k_range) > 0, !anyNA(k_range))

sub_assign = list()
stab_tabs  = list()
chosen_ks  = integer(0)

for (cl in split_clusters) {
  cl_cells = colnames(merged)[merged$final_label == cl]

  stab_file     = paste0(Robject_out_dir, "/subcluster_stability_", cl, ".Rds")
  stab_sig_file = paste0(Robject_out_dir, "/subcluster_stability_", cl, "_signature.txt")
  stab_sig      = paste(obj_sig, cl, length(cl_cells), paste(k_range, collapse = ","),
                        params$sub_n_iterations, sep = "|")

  if (file.exists(stab_file) && file.exists(stab_sig_file) &&
      identical(readLines(stab_sig_file, warn = FALSE)[1], stab_sig)) {
    stab = readRDS(stab_file)
  } else {
    # coordinates only, since find_stable_k subsets the object once per iteration
    # column names first: reducedDim<- checks its rownames against them
    lean = SingleCellExperiment(assays = list(), colData = colData(merged)[cl_cells, ])
    colnames(lean) = cl_cells
    reducedDim(lean, "corrected") = reducedDim(merged, "corrected")[cl_cells, , drop = FALSE]

    stab = find_stable_k(lean, k_range = k_range,
                         n_iterations = params$sub_n_iterations,
                         use.dimred = "corrected")
    saveRDS(stab, file = stab_file)
    writeLines(stab_sig, stab_sig_file)
  }

  tab = dplyr::bind_rows(lapply(stab, function(z)
    data.frame(cluster = cl, k = z$k, mean_n_clusters = z$mean_n_clusters,
               mean_ari = z$mean_ari, min_ari = z$min_ari)))

  eligible = tab %>% dplyr::filter(mean_n_clusters >= 2, !is.na(mean_ari))

  k_cl =
    if (as.integer(params$sub_k) > 0) {
      as.integer(params$sub_k)
    } else if (nrow(eligible) > 0) {
      eligible$k[which.max(eligible$mean_ari)]
    } else {
      NA_integer_
    }

  if (is.na(k_cl))
    stop("cluster ", cl, ": no k in the sweep gave two or more subclusters. ",
         "Set sub_k to force a value.")

  set.seed(1)
  snn = buildSNNGraph(merged[, cl_cells], k = k_cl, use.dimred = "corrected")
  sub_assign[[cl]] = setNames(igraph::cluster_louvain(snn)$membership, cl_cells)

  stab_tabs[[cl]] = tab
  chosen_ks[cl]   = k_cl
}

stab_tab = dplyr::bind_rows(stab_tabs)
write.csv(stab_tab, file = paste0(results_dir, "/subcluster_stability.csv"),
          row.names = FALSE)

datatable(stab_tab %>% mutate(across(where(is.numeric), ~signif(., 3))),
          rownames = FALSE, options = list(dom = "t", paging = FALSE))
```

`r if (have_split) paste0("Subclustered ", paste0("cluster ", names(chosen_ks), " at k = ", chosen_ks, collapse = "; "), if (as.integer(params$sub_k) > 0) ", set by the sub_k parameter." else ".") else ""`

```{r}
node_chr = merged$final_label

if (have_split) {
  for (cl in split_clusters) {
    idx = which(merged$final_label == cl)
    # matched by barcode, never by position
    node_chr[idx] = paste0(cl, ".", sub_assign[[cl]][colnames(merged)[idx]])
  }
}

# cluster order from the parameter, subclusters numerically within their cluster
node_levels = data.frame(parent = merged$final_label, node = node_chr,
                         stringsAsFactors = FALSE) %>%
  dplyr::distinct() %>%
  dplyr::mutate(sub_n = suppressWarnings(
                  as.integer(sub("^.*\\.([0-9]+)$", "\\1", node))),
                sub_n = ifelse(node == parent, 0L, sub_n)) %>%
  dplyr::arrange(match(parent, clusters), sub_n) %>%
  dplyr::pull(node)

if (length(node_levels) < 2)
  stop("only one node (", node_levels, "). TSCAN needs at least two; for a ",
       "single cluster, name it in split_by_subcluster.")

merged$node   = factor(node_chr, levels = node_levels)
combined$node = merged$node

node_tab = as.data.frame(colData(merged)) %>%
  dplyr::count(node, library, name = "n") %>%
  tidyr::pivot_wider(names_from = library, values_from = n, values_fill = 0) %>%
  dplyr::mutate(Total = rowSums(dplyr::across(dplyr::any_of(names(lib_cols)))))

write.csv(node_tab, file = paste0(results_dir, "/node_composition.csv"),
          row.names = FALSE)

datatable(node_tab, rownames = FALSE, options = list(dom = "t", paging = FALSE))
```

## Root

Every node is scored on the germ cell marker sets from the collaborators' S1
sheet: the mean log expression across each set's genes, and the mean percent of
cells detecting each gene. Unless `root_node` is set, the root is chosen by this
lineage's root rule, which names a set and whether the root is the node with its
highest (`max`) or lowest (`min`) mean expression. The scores are written to
`root_marker_set_scores.csv`.

```{r}
germ_sets = c("Primordial germ cells", "Migrating germ cells",
              "Spermatogonial stem cells", "Spermatocytes", "Oocyte")

set_genes = read.csv(paste0(params$project_dir, "/data/", params$marker_file),
                     stringsAsFactors = FALSE) %>%
  dplyr::mutate(loc_id = trimws(loc_id)) %>%
  dplyr::filter(CellTypeMarkerSet %in% germ_sets) %>%
  dplyr::transmute(set = CellTypeMarkerSet, gene = loc_id)

set_scores = score_marker_sets(combined, merged$node, set_genes) %>%
  dplyr::rename(node = group)

set_table = set_scores %>%
  dplyr::mutate(set          = factor(set, levels = germ_sets),
                node         = factor(node, levels = node_levels),
                mean_logexpr = signif(mean_logexpr, 3),
                pct_detected = round(pct_detected, 1)) %>%
  dplyr::arrange(set, node) %>%
  tidyr::pivot_wider(id_cols = c(node, n_cells), names_from = set,
                     values_from = c(mean_logexpr, pct_detected),
                     names_glue = "{set}: {.value}") %>%
  dplyr::arrange(node)

write.csv(set_table, file = paste0(results_dir, "/root_marker_set_scores.csv"),
          row.names = FALSE)

root_rule = switch(lineage,
                   female = params$female_root_rule,
                   male   = params$male_root_rule,
                   all    = params$all_root_rule)
rule_dir  = NA_character_
rule_set  = NA_character_
root_pct  = NA_real_
weak_root = FALSE
n_tied    = 1L

if (nzchar(params$root_node)) {
  root = as.character(params$root_node)
} else {
  rule_parts = regmatches(root_rule, regexec("^(max|min):(.+)$", root_rule))[[1]]
  if (length(rule_parts) != 3)
    stop(lineage, "_root_rule '", root_rule, "' must look like max:<set> or min:<set>")
  rule_dir = rule_parts[2]
  rule_set = trimws(rule_parts[3])
  if (!rule_set %in% germ_sets)
    stop("root rule set '", rule_set, "' is not one of: ",
         paste(germ_sets, collapse = ", "))

  cand = set_scores %>% dplyr::filter(set == rule_set, !is.na(mean_logexpr))
  if (nrow(cand) == 0)
    stop("no genes from the ", rule_set, " set are in the data, so the root ",
         "rule cannot be applied. Set root_node.")

  pick      = if (rule_dir == "max") which.max(cand$mean_logexpr) else which.min(cand$mean_logexpr)
  root      = as.character(cand$node[pick])
  root_pct  = cand$pct_detected[pick]
  weak_root = rule_dir == "max" && root_pct < params$min_detect_pct
  # which.max and which.min take the first of any tie, so report ties
  n_tied    = sum(cand$mean_logexpr == cand$mean_logexpr[pick])
}

if (!root %in% levels(merged$node))
  stop("root '", root, "' is not a node. Nodes: ",
       paste(levels(merged$node), collapse = ", "))

root_type = unique(lineage_key$cell_type[lineage_key$final_label == root])

datatable(set_table, rownames = FALSE,
          options = list(dom = "t", paging = FALSE, scrollX = TRUE))
```

`r paste0("The ordering is rooted at node **", root, "**", if (length(root_type) == 1) paste0(" (", root_type, ")") else "", if (nzchar(params$root_node)) ", set by the root_node parameter." else paste0(", the node with the ", if (rule_dir == "max") "highest" else "lowest", " mean expression of the ", rule_set, " set (", lineage, "_root_rule = ", root_rule, ")."), " The root is an assumption: rooting the same tree elsewhere gives a different ordering.")`

`r if (!is.na(rule_set)) paste0("The ", rule_set, " set chose the root, so its genes rise or fall along pseudotime partly by construction. Do not read their trend as a result.") else ""`

`r if (weak_root) paste0("**The ", rule_set, " genes are barely detected in any node.** The root node detects them in ", round(root_pct, 1), "% of cells on average, below the ", params$min_detect_pct, "% threshold, so the choice rests on noise rather than expression.") else ""`

`r if (n_tied > 1) paste0("**", n_tied, " nodes tie for the ", if (rule_dir == "max") "highest" else "lowest", " ", rule_set, " score.** The root is the first of them in node order, not a distinct choice.") else ""`

```{r, echo=FALSE, results="asis"}
do_markers = lineage %in% c("female", "male", "all")

if (do_markers) {
  cat("\n# Node markers\n\n",
      "Markers for each trajectory node",
      switch(lineage,
             female = ", meaning each cluster 5 subcluster,",
             male   = ", meaning each male germ cluster,",
             all    = ", meaning each cluster 5 subcluster and each other germ cluster,"),
      " from `findMarkers` on uncorrected log counts, comparing each node with ",
      "every other node in this lineage (`direction = \"up\"`, ",
      "`pval.type = \"some\"`). Library is not blocked: a node drawn mostly ",
      "from one library has no within-library comparison, so its markers include ",
      "library differences. Genes are kept at FDR < 0.01 with a mean log fold ",
      "change of at least ", params$lfc_thresh, " across the pairwise comparisons, ",
      "and written to `subcluster_marker_genes.csv`.\n\n", sep = "")
}
```

```{r, eval=do_markers}
# mean over every pairwise logFC; summary.logFC under pval.type = "some" comes
# from whichever comparison supplied the p-value, so it is not a fixed contrast
marker_effect = function(df) {
  lf = grep("^logFC\\.", colnames(df), value = TRUE)
  if (length(lf) == 0) return(rep(NA_real_, nrow(df)))
  rowMeans(as.matrix(as.data.frame(df[, lf, drop = FALSE])), na.rm = TRUE)
}

marker_file     = paste0(Robject_out_dir, "/subcluster_markers.Rds")
marker_sig_file = paste0(Robject_out_dir, "/subcluster_markers_signature.txt")
marker_sig      = paste(obj_sig, paste(node_levels, collapse = ","),
                        paste(as.integer(table(merged$node)), collapse = ","),
                        sep = "|")

if (file.exists(marker_file) && file.exists(marker_sig_file) &&
    identical(readLines(marker_sig_file, warn = FALSE)[1], marker_sig)) {
  sub_markers = readRDS(marker_file)
} else {
  sub_markers = findMarkers(combined, combined$node, test.type = "t",
                            direction = "up", pval.type = "some")
  saveRDS(sub_markers, file = marker_file)
  writeLines(marker_sig, marker_sig_file)
}

sub_marker_tab = purrr::map_dfr(names(sub_markers), function(nd) {
  df = sub_markers[[nd]]
  data.frame(node      = nd,
             Gene      = rownames(df),
             p.value   = df$p.value,
             FDR       = df$FDR,
             meanLogFC = marker_effect(df),
             stringsAsFactors = FALSE)
}) %>%
  dplyr::filter(FDR < 0.01, meanLogFC >= params$lfc_thresh) %>%
  dplyr::left_join(gene_description, by = "Gene") %>%
  dplyr::group_by(node) %>%
  dplyr::arrange(desc(meanLogFC), FDR, .by_group = TRUE) %>%
  dplyr::ungroup() %>%
  dplyr::select(node, Gene, WangGeneID, description, meanLogFC, p.value, FDR)

write.csv(sub_marker_tab, file = paste0(results_dir, "/subcluster_marker_genes.csv"),
          row.names = FALSE)

marker_counts = data.frame(node = node_levels) %>%
  dplyr::left_join(sub_marker_tab %>% dplyr::count(node, name = "n_markers"),
                   by = "node") %>%
  dplyr::mutate(n_markers = ifelse(is.na(n_markers), 0L, n_markers))

datatable(marker_counts, rownames = FALSE, options = list(dom = "t", paging = FALSE))
```

```{r, eval=do_markers}
datatable(sub_marker_tab %>% mutate(across(c(meanLogFC, p.value, FDR), ~signif(., 3))),
          filter = "top", rownames = FALSE)
```

```{r, echo=FALSE, results="asis"}
if (do_markers) {
  cat("\n## GO enrichment of node markers\n\n",
      "GO enrichment of each node's marker genes, against every gene in the ",
      "data. Terms with p < ", params$go_pval, " are shown and written to ",
      "`subcluster_marker_GO_pvalues.csv`.\n\n", sep = "")
  if (!have_go) cat("Skipped: the GO files were not found.\n\n")
  else if (nrow(sub_marker_tab) == 0) cat("Skipped: no node has marker genes at this cutoff.\n\n")
}
```

```{r, eval=do_markers && have_go && exists("sub_marker_tab") && nrow(sub_marker_tab) > 0}
marker_go = run_go_enrichment(
  gaf              = gaf,
  gene_description = gene_description,
  go_names         = go_names,
  test_genes       = sub_marker_tab %>% dplyr::select(Gene, node) %>% dplyr::distinct(),
  background_n     = nrow(combined),
  group2_label     = "node",
  ngroup2_label    = "nMarkerGenes",
  overlaps_file    = paste0(results_dir, "/subcluster_marker_GO_overlaps.csv"),
  pvals_file       = paste0(results_dir, "/subcluster_marker_GO_pvalues.csv"),
  pval_filter      = params$go_pval)

datatable(marker_go %>% mutate(across(where(is.numeric), ~signif(., 3))),
          filter = "top", rownames = FALSE)
```

# Trajectory

```{r}
pseudo_file     = paste0(Robject_out_dir, "/tscan_pseudo.Rds")
pseudo_sig_file = paste0(Robject_out_dir, "/tscan_pseudo_signature.txt")
pseudo_sig      = paste(obj_sig, root, paste(node_levels, collapse = ","),
                        paste(as.integer(table(merged$node)), collapse = ","),
                        sep = "|")

if (file.exists(pseudo_file) && file.exists(pseudo_sig_file) &&
    identical(readLines(pseudo_sig_file, warn = FALSE)[1], pseudo_sig)) {
  pseudo = readRDS(pseudo_file)
} else {
  set.seed(1)
  pseudo = quickPseudotime(merged,
                           clusters   = merged$node,
                           use.dimred = "corrected",
                           outgroup   = FALSE,
                           start      = root)
  saveRDS(pseudo, file = pseudo_file)
  writeLines(pseudo_sig, pseudo_sig_file)
}

merged$pseudotime   = averagePseudotime(pseudo$ordering)
combined$pseudotime = merged$pseudotime
n_paths = ncol(pseudo$ordering)
```

`r paste0(n_paths, " path", if (n_paths == 1) "" else "s", " from the root: ", paste(colnames(pseudo$ordering), collapse = ", "), ". The tree is fitted on node centroids in the corrected space and drawn below on the UMAP of that space.")`

```{r, fig.width=12, fig.height=5}
by_node = aggregateAcrossCells(
  merged, ids = merged$node,
  use.assay.type = if ("reconstructed" %in% assayNames(merged)) "reconstructed" else "logcounts")

edges_umap = reportEdges(by_node, mst = pseudo$mst, clusters = NULL,
                         use.dimred = "UMAP")

p_node = plotUMAP(merged, colour_by = "node", text_by = "node", point_size = .5) +
  geom_line(data = edges_umap, mapping = aes(x = UMAP1, y = UMAP2, group = edge)) +
  ggtitle("Nodes")
p_pt = plotUMAP(merged, colour_by = "pseudotime", point_size = .5) +
  geom_line(data = edges_umap, mapping = aes(x = UMAP1, y = UMAP2, group = edge)) +
  ggtitle("Pseudotime")

p_node | p_pt
ggsave(p_node | p_pt, file = paste0(results_dir, "/tscan_trajectory_umap.png"),
       width = 12, height = 5)
```

```{r, fig.width=12, fig.height=4}
pt_df = as.data.frame(colData(merged))

p_box_node = ggplot(pt_df, aes(x = node, y = pseudotime)) +
  geom_boxplot(outlier.size = .3) +
  xlab("node") + ylab("pseudotime")
p_box_lib = ggplot(pt_df, aes(x = library, y = pseudotime, fill = library)) +
  geom_boxplot(outlier.size = .3) +
  scale_fill_manual(values = lib_cols) +
  xlab(NULL) + ylab("pseudotime") +
  theme(legend.position = "none")

p_box_node | p_box_lib
ggsave(p_box_node | p_box_lib, file = paste0(results_dir, "/tscan_pseudotime_by_node.png"),
       width = 12, height = 4)
```

## Node graph

Pie charts give the library composition of each node, and the lower panel colors
nodes by mean pseudotime.

```{r, fig.width=7, fig.height=7}
sample_tally = pt_df %>%
  mutate(node = as.factor(node), library = as.factor(library)) %>%
  group_by(node, library, .drop = FALSE) %>%
  tally()

node_meta = sample_tally %>%
  tidyr::pivot_wider(names_from = library, values_from = n, values_fill = 0) %>%
  ungroup()
for (l in names(lib_cols)) if (!l %in% names(node_meta)) node_meta[[l]] = 0
node_meta = node_meta %>%
  mutate(totalCells = rowSums(across(all_of(names(lib_cols))))) %>%
  dplyr::rename(name = "node") %>%
  mutate(name = as.character(name))

mean_pseudo = pt_df %>%
  group_by(node) %>%
  summarise(mean_pseudo = mean(pseudotime, na.rm = TRUE), .groups = "drop") %>%
  dplyr::rename(name = "node") %>%
  mutate(name = as.character(name))

gr = as_tbl_graph(pseudo$mst) %>%
  activate(nodes) %>%
  left_join(node_meta, by = "name") %>%
  left_join(mean_pseudo, by = "name")

layout_data = create_layout(gr, layout = "stress")

gr_pie =
  ggraph(layout_data) +
  geom_edge_link() +
  geom_scatterpie(aes(x, y), cols = names(lib_cols), data = layout_data,
                  colour = NA, pie_scale = 1.5) +
  scale_fill_manual(name = "Library", values = lib_cols) +
  geom_node_text(aes(label = name), color = "black", size = 2.6) +
  theme_void(base_size = 8) +
  theme(legend.position = "left") +
  labs(tag = "A")

gr_pseudo =
  ggraph(layout_data) +
  geom_edge_link() +
  geom_node_point(aes(color = mean_pseudo), size = 9) +
  scale_color_viridis_c() +
  geom_node_text(aes(label = name), vjust = .75, size = 2.6, color = "red") +
  scale_x_continuous(expand = expansion(mult = 0.15)) +
  scale_y_continuous(expand = expansion(mult = 0.15)) +
  theme_void(base_size = 8) +
  theme(legend.position = "left") +
  labs(tag = "B")

(gr_pie + coord_fixed()) / (gr_pseudo + coord_fixed())
ggsave(file = paste0(results_dir, "/tscan_node_graph.pdf"), width = 7, height = 7)
```

## Library and stage

```{r}
lib_r2 = if (dplyr::n_distinct(pt_df$library) > 1)
  summary(lm(pseudotime ~ library, data = pt_df))$r.squared else NA_real_
stage_r2 = if (dplyr::n_distinct(pt_df$stage) > 1)
  summary(lm(pseudotime ~ stage, data = pt_df))$r.squared else NA_real_
```

`r if (is.na(lib_r2)) "All cells come from one library, so library cannot explain pseudotime." else paste0("Library explains ", round(100 * lib_r2, 1), "% of the variance in pseudotime", if (!is.na(stage_r2)) paste0(", and stage explains ", round(100 * stage_r2, 1), "%") else "", ".")`

# Pseudotime-associated genes

Genes changing along each path, tested on uncorrected log counts. All genes at
FDR < `r params$fdr_cut` are written to `tscan_pseudotime_genes.csv`.

```{r}
test_file     = paste0(Robject_out_dir, "/tscan_test_pseudotime.Rds")
test_sig_file = paste0(Robject_out_dir, "/tscan_test_pseudotime_signature.txt")

pseudo_mat = pathStat(pseudo$ordering)

if (file.exists(test_file) && file.exists(test_sig_file) &&
    identical(readLines(test_sig_file, warn = FALSE)[1], pseudo_sig)) {
  test_pseudo = readRDS(test_file)
} else {
  test_pseudo = TSCAN::testPseudotime(combined, pseudotime = pseudo_mat)
  saveRDS(test_pseudo, file = test_file)
  writeLines(pseudo_sig, test_sig_file)
}

path_genes = dplyr::bind_rows(lapply(seq_along(test_pseudo), function(i) {
  as.data.frame(test_pseudo[[i]]) %>%
    tibble::rownames_to_column("Gene") %>%
    dplyr::mutate(Path = names(test_pseudo)[i])
})) %>%
  dplyr::filter(FDR < params$fdr_cut) %>%
  dplyr::left_join(gene_description, by = "Gene") %>%
  dplyr::mutate(Direction = ifelse(logFC > 0, "up", "down")) %>%
  dplyr::select(Path, Direction, Gene, WangGeneID, description, logFC, p.value, FDR) %>%
  dplyr::arrange(Path, FDR)

write.csv(path_genes, file = paste0(results_dir, "/tscan_pseudotime_genes.csv"),
          row.names = FALSE)

datatable(path_genes %>% mutate(across(c(logFC, p.value, FDR), ~signif(., 3))),
          filter = "top", rownames = FALSE)
```

`r paste0(format(nrow(path_genes), big.mark = ","), " gene-path results at FDR < ", params$fdr_cut, ".")`

## Heatmaps

The 25 most significant genes in each direction along each path.

```{r}
top_by_path = path_genes %>%
  dplyr::mutate(set = paste0(Path, ".", Direction)) %>%
  dplyr::group_by(set) %>%
  dplyr::arrange(FDR, desc(abs(logFC)), .by_group = TRUE) %>%
  dplyr::slice_head(n = 25) %>%
  dplyr::summarise(Gene = list(Gene), .groups = "drop") %>%
  tibble::deframe()

# the heatmap's pseudotime bar uses a fixed 0 to 1 color scale
pt_range = range(combined$pseudotime, na.rm = TRUE)
combined$pseudotime_scaled = (combined$pseudotime - pt_range[1]) / diff(pt_range)

heatmaps = list()
for (s in names(top_by_path)) {
  gene_ids    = top_by_path[[s]]
  gene_labels = gene_description$WangGeneID[match(gene_ids, gene_description$Gene)]
  gene_labels = ifelse(is.na(gene_labels), gene_ids, gene_labels)

  path_name   = sub("\\.(up|down)$", "", s)
  cell_subset = rownames(pseudo_mat)[!is.na(pseudo_mat[, path_name])]

  heatmaps[[s]] = plot_pseudotime_heatmap(combined, gene_ids, gene_labels,
                                          cell_subset,
                                          pseudotime_column = "pseudotime_scaled")
}
```

```{r, fig.width=7, fig.height=5, results="asis"}
for (s in names(heatmaps)) {
  cat("\n\n### ", s, "\n\n", sep = "")
  hp = (heatmaps[[s]][[1]] / heatmaps[[s]][[2]]) + plot_layout(heights = c(0.03, 1))
  print(hp)
  ggsave(hp, file = paste0(results_dir, "/tscan_heatmap_",
                           gsub("[^A-Za-z0-9]+", "_", s), ".png"),
         width = 7, height = 5)
}
```

## GO enrichment of pseudotime-associated genes

GO enrichment for each path and direction, against every gene in the data. Terms
with p < `r params$go_pval` are shown and written to
`tscan_pseudotime_GO_pvalues.csv`.

`r if (!have_go) "Skipped: the GO files were not found." else if (nrow(path_genes) == 0) "Skipped: there are no pseudotime-associated genes to test." else ""`

```{r, eval=have_go && nrow(path_genes) > 0}
path_go = run_go_enrichment(
  gaf              = gaf,
  gene_description = gene_description,
  go_names         = go_names,
  test_genes       = path_genes %>%
                       dplyr::transmute(Gene, set = paste0(Path, ".", Direction)) %>%
                       dplyr::distinct(),
  background_n     = nrow(combined),
  group2_label     = "Path",
  ngroup2_label    = "nPathGenes",
  overlaps_file    = paste0(results_dir, "/tscan_pseudotime_GO_overlaps.csv"),
  pvals_file       = paste0(results_dir, "/tscan_pseudotime_GO_pvalues.csv"),
  pval_filter      = params$go_pval)

datatable(path_go %>% mutate(across(where(is.numeric), ~signif(., 3))),
          filter = "top", rownames = FALSE)
```

# Export

```{r}
cell_out = pt_df %>%
  tibble::rownames_to_column("barcode") %>%
  dplyr::select(barcode, library, stage, final_label, cell_type, node, pseudotime)

umap = reducedDim(merged, "UMAP")
cell_out$UMAP1 = umap[, 1]
cell_out$UMAP2 = umap[, 2]

write.csv(cell_out, file = paste0(results_dir, "/pseudotime_per_cell.csv"),
          row.names = FALSE)
```

Per-cell pseudotime is written to `pseudotime_per_cell.csv`.

# Summary

```{r}
summary_tbl = data.frame(
  quantity = c("lineage", "clusters", "cells", "libraries kept", "nodes", "root",
               "paths", "variance in pseudotime explained by library",
               paste0("gene-path results at FDR < ", params$fdr_cut)),
  value = c(lineage,
            paste(clusters, collapse = ", "),
            format(ncol(merged), big.mark = ","),
            paste(objs$kept_batches, collapse = ", "),
            as.character(nlevels(merged$node)),
            root,
            as.character(n_paths),
            if (is.na(lib_r2)) "one library" else paste0(round(100 * lib_r2, 1), "%"),
            format(nrow(path_genes), big.mark = ",")),
  stringsAsFactors = FALSE)

datatable(summary_tbl, rownames = FALSE, colnames = c("", ""),
          options = list(dom = "t", ordering = FALSE, paging = FALSE))
```

# Session info

```{r}
sessionInfo()
```
