---
title: "18_germ_pseudotime_monocle3"
author: "Stephanie Hickey"
date: "`r Sys.Date()`"
output:
  html_document:
    code_folding: hide
    toc: true
    toc_float:
      toc_collapsed: true
params:
    project_dir: /mnt/ufs18/rs-013/bioinformaticsCore/projects/chong_davidson/BCC125_sea_lamprey_scRNA
    project_name: BCC125_sea_lamprey_scRNA
    lab: Weiming Li
    results_date: "20260813"
    lineage: "female"
    female_clusters: "5"
    male_clusters: "12_transformer,15,16,17,18"
    male_root: "12_transformer"
    all_clusters: "5,12_transformer,15,16,17,18"
    all_root: "12_transformer"
    root_label: ""
    root_monocle_cluster: ""
    min_library_cells: 50
    use_partition: FALSE
    graph_q: 0.01
    cores: 4
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

```{r, echo=FALSE, results="asis"}
cat("\n")
cat("#### Lab:", params$lab, " \n")
cat("#### Project:", params$project_name, " \n")
cat("#### Lineage:", params$lineage, " \n")
```

```{r}
library(tidyverse)
library(SingleCellExperiment)
library(scater)
library(scran)
library(batchelor)
library(BiocSingular)
library(monocle3)
library(patchwork)
library(DT)
source(paste0(params$project_dir, "/src/scRNA_seq_functions.R"))
```

monocle3 pseudotime for the `r params$lineage` germ cell clusters. The lineage
is `female`, `male`, or `all`, which orders every germ cluster in one graph.
`17_germ_pseudotime_tscan.Rmd` computes TSCAN pseudotime for the same cells, and
`knit_17_18_pseudotime.sh` knits both documents for each lineage.

Both documents use the same cells and the same corrected space, built by
`build_lineage_sce` in `scRNA_seq_functions.R`. monocle3 learns its principal
graph on the UMAP of that space, and its genes are tested on uncorrected counts.

Stage and collection date vary together in this dataset, so the ordering is not
shown to be developmental. The variance in pseudotime explained by library is
reported with the result.

```{r}
lineage = as.character(params$lineage)
if (!lineage %in% c("female", "male", "all"))
  stop("lineage must be 'female', 'male' or 'all', not '", lineage, "'")

clusters = trimws(strsplit(as.character(switch(lineage,
  female = params$female_clusters,
  male   = params$male_clusters,
  all    = params$all_clusters)), ",")[[1]])
clusters = clusters[nzchar(clusters)]

object_dir      = paste0(params$project_dir, "/results/", params$results_date,
                         "_no_doublets")
atlas_dir       = paste0(params$project_dir, "/results/", params$results_date,
                         "_atlas")
results_dir     = paste0(params$project_dir, "/results/", params$results_date,
                         "_pseudotime_monocle3_", lineage)
tscan_file = paste0(params$project_dir, "/results/", params$results_date,
                    "_pseudotime_tscan_", lineage, "/pseudotime_per_cell.csv")
Robject_out_dir = paste0(results_dir, "/Robjects")
if (!dir.exists(Robject_out_dir)) dir.create(Robject_out_dir, recursive = TRUE)

label_file = paste0(atlas_dir, "/cell_cluster_labels.csv")
if (!file.exists(label_file))
  stop("missing ", label_file, ". Run 13_atlas_and_stage_correspondence.Rmd first.")

cell_labels      = read.csv(label_file, stringsAsFactors = FALSE)
all.sce          = readRDS(paste0(object_dir, "/all.sce.Rds"))
gene_description = read.csv(paste0(object_dir, "/gene_descritption.csv"),
                            stringsAsFactors = FALSE) %>%
  dplyr::distinct(Gene, .keep_all = TRUE)

lib_cols = c("Female" = "#FF6DB6", "Male" = "#006ddb",
             "Undifferentiated.1" = "#009292", "Undifferentiated.2" = "#b66ddf")

use_partition = isTRUE(as.logical(params$use_partition))
```

# Cells

```{r}
absent = setdiff(clusters, cell_labels$final_label)
if (length(absent) > 0)
  stop("cluster(s) ", paste(absent, collapse = ", "), " are not in ", label_file,
       ". Labels present: ",
       paste(sort(unique(cell_labels$final_label)), collapse = ", "),
       ". Set the ", lineage, "_clusters parameter from the current labelling.")

lineage_key = cell_labels %>% dplyr::filter(final_label %in% clusters)

comp_tab = lineage_key %>%
  dplyr::count(final_label, cell_type, library, name = "n") %>%
  tidyr::pivot_wider(names_from = library, values_from = n, values_fill = 0) %>%
  dplyr::mutate(Total = rowSums(dplyr::across(dplyr::any_of(names(lib_cols))))) %>%
  dplyr::arrange(match(final_label, clusters))

write.csv(comp_tab, file = paste0(results_dir, "/cluster_composition.csv"),
          row.names = FALSE)

datatable(comp_tab, rownames = FALSE, options = list(dom = "t", paging = FALSE))
```

`r paste0(format(nrow(lineage_key), big.mark = ","), " cells from cluster(s) ", paste(clusters, collapse = ", "), ".")`

# Corrected space

Libraries contributing fewer than `r params$min_library_cells` cells are dropped.
With more than one library left, the space is `fastMNN` output computed on these
cells alone. With one library, it is PCA on that library.

```{r}
obj_file     = paste0(Robject_out_dir, "/lineage_objects.Rds")
obj_sig_file = paste0(Robject_out_dir, "/lineage_objects_signature.txt")
obj_sig      = paste(lineage, paste(sort(clusters), collapse = ","),
                     nrow(lineage_key), params$min_library_cells, sep = "|")

if (file.exists(obj_file) && file.exists(obj_sig_file) &&
    identical(readLines(obj_sig_file, warn = FALSE)[1], obj_sig)) {
  objs = readRDS(obj_file)
} else {
  objs = build_lineage_sce(all.sce, lineage_key$barcode,
                           min_library_cells = params$min_library_cells)
  saveRDS(objs, file = obj_file)
  writeLines(obj_sig, obj_sig_file)
}

merged   = objs$corrected
combined = objs$combined

# metadata attached by barcode, never by position
key_idx = match(colnames(merged), lineage_key$barcode)
stopifnot(!anyNA(key_idx))
merged$library     = lineage_key$library[key_idx]
merged$stage       = lineage_key$stage[key_idx]
merged$final_label = lineage_key$final_label[key_idx]
merged$cell_type   = lineage_key$cell_type[key_idx]

lib_tab = data.frame(batch   = names(objs$n_by_library),
                     n_cells = as.integer(objs$n_by_library),
                     kept    = names(objs$n_by_library) %in% objs$kept_batches,
                     stringsAsFactors = FALSE) %>%
  dplyr::left_join(dplyr::distinct(cell_labels, batch, library), by = "batch")

datatable(lib_tab, rownames = FALSE, options = list(dom = "t"))
```

`r n_kept = length(objs$kept_batches); paste0(format(ncol(merged), big.mark = ","), " cells from ", n_kept, " librar", if (n_kept == 1) "y" else "ies", " are kept.")`

# Principal graph

The cell data set holds uncorrected counts. The corrected space and its UMAP are
placed in its reduced dimensions, and `cluster_cells` and `learn_graph` run on
that UMAP.

`r if (use_partition) "use_partition is TRUE, so a separate graph is learned for each partition. Cells in partitions that do not contain the root get infinite pseudotime." else "use_partition is FALSE, so one graph is learned across all cells and every cell receives a pseudotime."`

```{r}
graph_file     = paste0(Robject_out_dir, "/cds_learned.Rds")
graph_sig_file = paste0(Robject_out_dir, "/cds_learned_signature.txt")
graph_sig      = paste(obj_sig, use_partition, sep = "|")

if (file.exists(graph_file) && file.exists(graph_sig_file) &&
    identical(readLines(graph_sig_file, warn = FALSE)[1], graph_sig)) {
  cds = readRDS(graph_file)
} else {
  cell_md = as.data.frame(colData(merged))
  gene_md = data.frame(gene_short_name = rownames(combined),
                       row.names = rownames(combined))

  cds = new_cell_data_set(counts(combined),
                          cell_metadata = cell_md[colnames(combined), , drop = FALSE],
                          gene_metadata = gene_md)

  pca  = reducedDim(merged, "corrected"); rownames(pca)  = colnames(merged)
  umap = reducedDim(merged, "UMAP");      rownames(umap) = colnames(merged)
  reducedDim(cds, "PCA")  = pca[colnames(cds), , drop = FALSE]
  reducedDim(cds, "UMAP") = umap[colnames(cds), , drop = FALSE]

  cds = cluster_cells(cds, reduction_method = "UMAP", random_seed = 1)
  cds = learn_graph(cds, use_partition = use_partition)

  saveRDS(cds, file = graph_file)
  writeLines(graph_sig, graph_sig_file)
}

colData(cds)$monocle_cluster = as.character(monocle3::clusters(cds))

n_monocle_clusters = dplyr::n_distinct(colData(cds)$monocle_cluster)
n_partitions = dplyr::n_distinct(cds@clusters[["UMAP"]]$partitions)
```

`r paste0("cluster_cells finds ", n_monocle_clusters, " cluster", if (n_monocle_clusters == 1) "" else "s", " in ", n_partitions, " partition", if (n_partitions == 1) "" else "s", ".")`

`r if (n_partitions > 1 && !use_partition) paste0("These cells fall into ", n_partitions, " partitions, but one graph is learned across them because use_partition is FALSE. The edges joining separate partitions are placed by the method, not by continuity in the data.") else ""`

## Root

```{r}
zp_genes = gene_description %>%
  dplyr::filter(grepl("^ZP[0-9]", WangGeneID)) %>%
  dplyr::pull(Gene) %>%
  intersect(rownames(combined))

zp_cell = if (length(zp_genes) > 0) {
  colMeans(as.matrix(logcounts(combined)[zp_genes, colnames(cds), drop = FALSE]))
} else {
  rep(NA_real_, ncol(cds))
}

zp_by_cluster = data.frame(monocle_cluster = colData(cds)$monocle_cluster,
                           zp = zp_cell) %>%
  dplyr::group_by(monocle_cluster) %>%
  dplyr::summarise(n_cells = dplyr::n(), mean_ZP = mean(zp), .groups = "drop") %>%
  dplyr::arrange(as.integer(monocle_cluster))

if (nzchar(params$root_label)) {
  root_desc  = paste0("final label ", params$root_label, ", set by the root_label parameter")
  root_cells = colnames(cds)[colData(cds)$final_label == params$root_label]
} else if (nzchar(params$root_monocle_cluster)) {
  root_desc  = paste0("monocle3 cluster ", params$root_monocle_cluster,
                      ", set by the root_monocle_cluster parameter")
  root_cells = colnames(cds)[colData(cds)$monocle_cluster == params$root_monocle_cluster]
} else if (lineage %in% c("male", "all")) {
  root_param = if (lineage == "male") "male_root" else "all_root"
  root_value = as.character(params[[root_param]])
  root_desc  = paste0("final label ", root_value, ", set by the ", root_param, " parameter")
  root_cells = colnames(cds)[colData(cds)$final_label == root_value]
} else {
  if (length(zp_genes) == 0)
    stop("no ZP genes resolved, so the female root cannot be chosen ",
         "automatically. Set root_label or root_monocle_cluster.")
  low_zp     = as.character(zp_by_cluster$monocle_cluster[which.min(zp_by_cluster$mean_ZP)])
  root_desc  = paste0("monocle3 cluster ", low_zp,
                      ", the cluster with the lowest mean zona pellucida (ZP) expression")
  root_cells = colnames(cds)[colData(cds)$monocle_cluster == low_zp]
}

if (length(root_cells) == 0)
  stop("no cells match the root (", root_desc, ").")

# the principal graph vertex most of the root cells project onto
closest = cds@principal_graph_aux[["UMAP"]]$pr_graph_cell_proj_closest_vertex
closest = as.matrix(closest[colnames(cds), , drop = FALSE])
vertex_names = igraph::V(principal_graph(cds)[["UMAP"]])$name
root_pr_node = vertex_names[as.numeric(names(which.max(table(closest[root_cells, ]))))]

cds = order_cells(cds, reduction_method = "UMAP", root_pr_nodes = root_pr_node)

pt = pseudotime(cds)
colData(cds)$monocle_pseudotime = pt
n_infinite = sum(!is.finite(pt))

datatable(zp_by_cluster %>% mutate(mean_ZP = signif(mean_ZP, 3)),
          rownames = FALSE, options = list(dom = "t", paging = FALSE))
```

`r paste0("The ordering is rooted at principal graph vertex ", root_pr_node, ", the vertex most cells of ", root_desc, " project onto. The root is an assumption: rooting the same graph elsewhere gives a different ordering.")`

`r if (n_infinite == 0) "Every cell has a finite pseudotime." else paste0("**", n_infinite, " cells have infinite pseudotime**, because they sit in partitions not connected to the root. They are excluded from the correlations and gene heatmap below.")`

# Trajectory

```{r, fig.width=13, fig.height=4.5}
colour_label = if (lineage == "female") "monocle_cluster" else "final_label"

p_pt = plot_cells(cds, color_cells_by = "pseudotime", label_cell_groups = FALSE,
                  label_leaves = FALSE, label_branch_points = FALSE,
                  label_roots = TRUE, cell_size = 0.5) +
  ggtitle("Pseudotime")
p_cl = plot_cells(cds, color_cells_by = colour_label, label_cell_groups = TRUE,
                  label_leaves = FALSE, label_branch_points = FALSE,
                  cell_size = 0.5, group_label_size = 3) +
  ggtitle(colour_label)
p_lib = plot_cells(cds, color_cells_by = "library", label_cell_groups = FALSE,
                   label_leaves = FALSE, label_branch_points = FALSE,
                   cell_size = 0.5) +
  scale_color_manual(values = lib_cols) +
  ggtitle("Library")

p_pt | p_cl | p_lib
ggsave(p_pt | p_cl | p_lib, file = paste0(results_dir, "/monocle3_trajectory_umap.png"),
       width = 13, height = 4.5)
```

```{r, fig.width=12, fig.height=4}
pt_df = as.data.frame(colData(cds)) %>%
  tibble::rownames_to_column("barcode") %>%
  dplyr::filter(is.finite(monocle_pseudotime))

p_box_lab = ggplot(pt_df, aes(x = final_label, y = monocle_pseudotime)) +
  geom_boxplot(outlier.size = .3) +
  xlab("final label") + ylab("pseudotime")
p_box_lib = ggplot(pt_df, aes(x = library, y = monocle_pseudotime, fill = library)) +
  geom_boxplot(outlier.size = .3) +
  scale_fill_manual(values = lib_cols) +
  xlab(NULL) + ylab("pseudotime") +
  theme(legend.position = "none")

p_box_lab | p_box_lib
ggsave(p_box_lab | p_box_lib,
       file = paste0(results_dir, "/monocle3_pseudotime_by_label.png"),
       width = 12, height = 4)
```

## Library and stage

```{r}
lib_r2 = if (dplyr::n_distinct(pt_df$library) > 1)
  summary(lm(monocle_pseudotime ~ library, data = pt_df))$r.squared else NA_real_
stage_r2 = if (dplyr::n_distinct(pt_df$stage) > 1)
  summary(lm(monocle_pseudotime ~ stage, data = pt_df))$r.squared else NA_real_
```

`r if (is.na(lib_r2)) "All cells with a finite pseudotime come from one library, so library cannot explain pseudotime." else paste0("Library explains ", round(100 * lib_r2, 1), "% of the variance in pseudotime", if (!is.na(stage_r2)) paste0(", and stage explains ", round(100 * stage_r2, 1), "%") else "", ".")`

# Genes varying along the graph

`graph_test` with `neighbor_graph = "principal_graph"` tests each gene for
spatial autocorrelation over the principal graph (Moran's I). Genes at
q < `r params$graph_q` are written to `monocle3_graph_genes.csv`.

```{r}
gt_file     = paste0(Robject_out_dir, "/graph_test.Rds")
gt_sig_file = paste0(Robject_out_dir, "/graph_test_signature.txt")

if (file.exists(gt_file) && file.exists(gt_sig_file) &&
    identical(readLines(gt_sig_file, warn = FALSE)[1], graph_sig)) {
  gt = readRDS(gt_file)
} else {
  gt = graph_test(cds, neighbor_graph = "principal_graph",
                  cores = as.integer(params$cores))
  saveRDS(gt, file = gt_file)
  writeLines(graph_sig, gt_sig_file)
}

graph_genes = gt %>%
  tibble::rownames_to_column("Gene") %>%
  dplyr::filter(q_value < params$graph_q) %>%
  dplyr::left_join(gene_description, by = "Gene") %>%
  dplyr::select(Gene, WangGeneID, description, morans_I, p_value, q_value) %>%
  dplyr::arrange(desc(morans_I))

write.csv(graph_genes, file = paste0(results_dir, "/monocle3_graph_genes.csv"),
          row.names = FALSE)

datatable(graph_genes %>% mutate(across(c(morans_I, p_value, q_value), ~signif(., 3))),
          filter = "top", rownames = FALSE)
```

`r paste0(format(nrow(graph_genes), big.mark = ","), " genes at q < ", params$graph_q, ".")`

## Heatmap

The 50 genes with the highest Moran's I, ordered by pseudotime.

```{r, fig.width=7, fig.height=6}
if (nrow(graph_genes) > 0) {
  combined$monocle_pseudotime = pt[colnames(combined)]

  # the heatmap's pseudotime bar uses a fixed 0 to 1 color scale
  finite_pt = combined$monocle_pseudotime[is.finite(combined$monocle_pseudotime)]
  combined$pseudotime_scaled = (combined$monocle_pseudotime - min(finite_pt)) /
    diff(range(finite_pt))

  gene_ids    = head(graph_genes$Gene, 50)
  gene_labels = ifelse(is.na(graph_genes$WangGeneID[match(gene_ids, graph_genes$Gene)]),
                       gene_ids,
                       graph_genes$WangGeneID[match(gene_ids, graph_genes$Gene)])
  cell_subset = colnames(combined)[is.finite(combined$monocle_pseudotime)]

  hm = plot_pseudotime_heatmap(combined, gene_ids, gene_labels, cell_subset,
                               pseudotime_column = "pseudotime_scaled")
  hp = (hm[[1]] / hm[[2]]) + plot_layout(heights = c(0.03, 1))
  print(hp)
  ggsave(hp, file = paste0(results_dir, "/monocle3_heatmap.png"), width = 7, height = 6)
}
```

# Comparison with TSCAN

```{r}
have_tscan = file.exists(tscan_file)
```

`r if (!have_tscan) paste0("The TSCAN result for this lineage is not present at ", tscan_file, ". Knit 17_germ_pseudotime_tscan.Rmd first to compare the two orderings.") else ""`

```{r, eval=have_tscan, fig.width=5, fig.height=4.5}
tscan_pt = read.csv(tscan_file, stringsAsFactors = FALSE) %>%
  dplyr::select(barcode, tscan_pseudotime = pseudotime)

pt_compare = pt_df %>%
  dplyr::select(barcode, library, monocle_pseudotime) %>%
  dplyr::inner_join(tscan_pt, by = "barcode") %>%
  dplyr::filter(is.finite(tscan_pseudotime))

rho_methods = cor(pt_compare$monocle_pseudotime, pt_compare$tscan_pseudotime,
                  method = "spearman")

p_cmp = ggplot(pt_compare, aes(x = tscan_pseudotime, y = monocle_pseudotime,
                               colour = library)) +
  geom_point(size = .4, alpha = .5) +
  scale_colour_manual(values = lib_cols) +
  xlab("TSCAN pseudotime") + ylab("monocle3 pseudotime") +
  ggtitle(paste0("Spearman rho = ", signif(rho_methods, 3)))

p_cmp
ggsave(p_cmp, file = paste0(results_dir, "/monocle3_vs_tscan.png"), width = 5, height = 4.5)
```

`r if (have_tscan) paste0("Across ", format(nrow(pt_compare), big.mark = ","), " cells with a pseudotime from both methods, the Spearman correlation between the two orderings is ", signif(rho_methods, 3), ". Both orderings use the same cells and the same corrected space, so agreement between them does not address the stage confound.") else ""`

# Export

```{r}
umap = reducedDim(cds, "UMAP")

cell_out = as.data.frame(colData(cds)) %>%
  tibble::rownames_to_column("barcode") %>%
  dplyr::select(barcode, library, stage, final_label, cell_type,
                monocle_cluster, pseudotime = monocle_pseudotime)
cell_out$UMAP1 = umap[cell_out$barcode, 1]
cell_out$UMAP2 = umap[cell_out$barcode, 2]

write.csv(cell_out, file = paste0(results_dir, "/pseudotime_per_cell.csv"),
          row.names = FALSE)
```

Per-cell pseudotime is written to `pseudotime_per_cell.csv`.

# Summary

```{r}
summary_tbl = data.frame(
  quantity = c("lineage", "clusters", "cells", "libraries kept",
               "monocle3 clusters", "partitions", "use_partition", "root vertex",
               "cells with infinite pseudotime",
               "variance in pseudotime explained by library",
               paste0("genes at q < ", params$graph_q),
               "Spearman rho with TSCAN"),
  value = c(lineage,
            paste(clusters, collapse = ", "),
            format(ncol(cds), big.mark = ","),
            paste(objs$kept_batches, collapse = ", "),
            as.character(n_monocle_clusters),
            as.character(n_partitions),
            as.character(use_partition),
            root_pr_node,
            as.character(n_infinite),
            if (is.na(lib_r2)) "one library" else paste0(round(100 * lib_r2, 1), "%"),
            format(nrow(graph_genes), big.mark = ","),
            if (have_tscan) as.character(signif(rho_methods, 3)) else "not run"),
  stringsAsFactors = FALSE)

datatable(summary_tbl, rownames = FALSE, colnames = c("", ""),
          options = list(dom = "t", ordering = FALSE, paging = FALSE))
```

# Session info

```{r}
sessionInfo()
```
