---
title: "19_germ_pseudotime_fine_nodes"
author: "Stephanie Hickey"
date: "`r Sys.Date()`"
output:
  html_document:
    code_folding: hide
    toc: true
    toc_float:
      toc_collapsed: true
params:
    project_dir: /mnt/ufs18/rs-013/bioinformaticsCore/projects/chong_davidson/BCC125_sea_lamprey_scRNA
    project_name: BCC125_sea_lamprey_scRNA
    lab: Weiming Li
    results_date: "20260813"
    atlas_suffix: ""
    lineage: "all"
    female_clusters: "5"
    male_clusters: "15,16,17,18"
    all_clusters: "5,15,16,17,18"
    sub_k: 10
    min_node_cells: 30
    dist_method: "mnn"
    mnn_k: 50
    outgroup: FALSE
    root_rule: "max:Primordial germ cells"
    root_node: ""
    min_detect_pct: 5
    marker_file: canonical_markers_S1_only_2026-08-14.csv
    prev_nodes_file: ""
    node_annotation: "5.1=Migrating neural crest cells;5.2=Oogonia;5.3=Oocytes;5.4=Migrating female germ cells (FGC I);5.5=Granulosa;5.6=Migrating female germ cells (FGC II);5.7=Primordial germ cells (PGCs);5.8=Migrating PGCs;15=Spermatogonia;16=Spermatocytes I;17=Spermatocytes II;18=Spermatocytes III"
    min_library_cells: 50
    fdr_cut: 0.01
    go_pval: 0.05
    genome_dir: /mnt/research/bioinformaticsCore/shared/Genomes/Petromyzon_marinus
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

```{r, echo=FALSE, results="asis"}
cat("\n")
cat("#### Lab:", params$lab, " \n")
cat("#### Project:", params$project_name, " \n")
cat("#### Lineage:", params$lineage, " \n")
```

```{r}
library(tidyverse)
library(SingleCellExperiment)
library(scater)
library(scran)
library(batchelor)
library(BiocSingular)
library(TSCAN)
library(patchwork)
library(DT)
source(paste0(params$project_dir, "/src/scRNA_seq_functions.R"))
```

TSCAN pseudotime for the `r params$lineage` germ cells, with **every** cluster
subclustered finely and no cell removed.

The tree TSCAN fits is a minimum spanning tree over one centroid per node, so
the number of nodes is the number of points the trajectory is fitted to. `17`
gives it around a dozen, of which four are whole clusters holding thousands of
cells each. A tree on twelve points is coarse and moves whenever a node moves.

This document treats nodes as waypoints rather than cell types. Every cluster is
cut into small subclusters at a fixed `sub_k`, which gives the tree many more
points to follow. Nothing here needs a node to be interpretable, so nothing here
depends on the subclustering being stable or on any node carrying a label.

The collaborators' annotation is carried **per cell**, by barcode, from the run
named in `prev_nodes_file`. It is never attached to a node. Re-clustering
therefore cannot break it, and the interpretation is read off the composition of
the trajectory rather than off node names.

```{r}
lineage = as.character(params$lineage)
if (!lineage %in% c("female", "male", "all"))
  stop("lineage must be 'female', 'male' or 'all', not '", lineage, "'")

clusters = trimws(strsplit(as.character(switch(lineage,
  female = params$female_clusters,
  male   = params$male_clusters,
  all    = params$all_clusters)), ",")[[1]])
clusters = clusters[nzchar(clusters)]

object_dir      = paste0(params$project_dir, "/results/", params$results_date,
                         "_no_doublets")
atlas_dir       = paste0(params$project_dir, "/results/", params$results_date,
                         "_atlas", params$atlas_suffix)
results_dir     = paste0(params$project_dir, "/results/", params$results_date,
                         "_pseudotime_fine_", lineage, params$atlas_suffix)
Robject_out_dir = paste0(results_dir, "/Robjects")
if (!dir.exists(Robject_out_dir)) dir.create(Robject_out_dir, recursive = TRUE)

label_file = paste0(atlas_dir, "/cell_cluster_labels.csv")
if (!file.exists(label_file))
  stop("missing ", label_file, ". Run 13 first.")

cell_labels      = read.csv(label_file, stringsAsFactors = FALSE)
all.sce          = readRDS(paste0(object_dir, "/all.sce.Rds"))
gene_description = read.csv(paste0(object_dir, "/gene_descritption.csv"),
                            stringsAsFactors = FALSE) %>%
  dplyr::distinct(Gene, .keep_all = TRUE)

lib_cols = c("Female" = "#FF6DB6", "Male" = "#006ddb",
             "Undifferentiated.1" = "#009292", "Undifferentiated.2" = "#b66ddf")

tech     = technical_gene_ids(gene_description)
mito_ids = tech$mito
rrna_ids = tech$rrna
ribo_ids = tech$ribo
```

# Cells

```{r}
absent = setdiff(clusters, cell_labels$final_label)
if (length(absent) > 0)
  stop("cluster(s) ", paste(absent, collapse = ", "), " are not in ", label_file,
       ". Labels present: ",
       paste(sort(unique(cell_labels$final_label)), collapse = ", "))

lineage_key = cell_labels %>% dplyr::filter(final_label %in% clusters)

# the annotation, per cell, by barcode. Nothing is dropped on the strength of it.
parse_pairs = function(x) {
  x = trimws(x)
  if (!nzchar(x)) return(character(0))
  kv = strsplit(trimws(strsplit(x, ";")[[1]]), "=")
  setNames(vapply(kv, function(z) trimws(z[2]), character(1)),
           vapply(kv, function(z) trimws(z[1]), character(1)))
}

anno_map  = parse_pairs(params$node_annotation)
prev_file = as.character(params$prev_nodes_file)
have_prev = nzchar(prev_file) && file.exists(prev_file)

cell_anno = rep(NA_character_, nrow(lineage_key))
if (have_prev) {
  prev = read.csv(prev_file, stringsAsFactors = FALSE)
  stopifnot(all(c("barcode", "node") %in% colnames(prev)))
  old = prev$node[match(lineage_key$barcode, prev$barcode)]
  cell_anno = unname(anno_map[as.character(old)])
  cell_anno[is.na(cell_anno) & !is.na(old)] = as.character(old)[is.na(cell_anno) & !is.na(old)]
}
lineage_key$annotation = ifelse(is.na(cell_anno), "unannotated", cell_anno)

comp_tab = lineage_key %>%
  dplyr::count(final_label, annotation, name = "n") %>%
  tidyr::pivot_wider(names_from = annotation, values_from = n, values_fill = 0)

write.csv(comp_tab, file = paste0(results_dir, "/cluster_by_annotation.csv"),
          row.names = FALSE)

datatable(comp_tab, rownames = FALSE, options = list(dom = "t", scrollX = TRUE))
```

`r paste0(format(nrow(lineage_key), big.mark = ","), " cells from cluster(s) ", paste(clusters, collapse = ", "), ". No cell is removed, so 5.1 is present.", if (!have_prev) " **prev_nodes_file is not set**, so no annotation is carried and the trajectory cannot be read against it." else paste0(" ", format(sum(lineage_key$annotation != "unannotated"), big.mark = ","), " carry an annotation from ", basename(prev_file), "."))`

# Corrected space

```{r}
obj_file     = paste0(Robject_out_dir, "/lineage_objects.Rds")
obj_sig_file = paste0(Robject_out_dir, "/lineage_objects_signature.txt")
obj_sig      = paste(lineage, paste(sort(clusters), collapse = ","),
                     nrow(lineage_key), params$min_library_cells, sep = "|")

if (file.exists(obj_file) && file.exists(obj_sig_file) &&
    identical(readLines(obj_sig_file, warn = FALSE)[1], obj_sig)) {
  objs = readRDS(obj_file)
} else {
  objs = build_lineage_sce(all.sce, lineage_key$barcode,
                           min_library_cells = params$min_library_cells)
  saveRDS(objs, file = obj_file)
  writeLines(obj_sig, obj_sig_file)
}

merged   = objs$corrected
combined = objs$combined

key_idx = match(colnames(merged), lineage_key$barcode)
stopifnot(!anyNA(key_idx))
merged$library     = lineage_key$library[key_idx]
merged$stage       = lineage_key$stage[key_idx]
merged$final_label = lineage_key$final_label[key_idx]
merged$annotation  = lineage_key$annotation[key_idx]
```

`r paste0(format(ncol(merged), big.mark = ","), " cells from ", length(objs$kept_batches), " librar", if (length(objs$kept_batches) == 1) "y" else "ies", " are kept.")`

# Fine nodes

Every cluster is subclustered in the shared corrected space, using only that
cluster's cells, at a fixed `sub_k`. A smaller `sub_k` gives more and smaller
subclusters. There is no stability sweep, because nothing downstream asks a node
to be a reproducible cell type.

Subclusters below `min_node_cells` are merged into the nearest subcluster of the
same parent cluster, by centroid distance. A node of five cells gives a centroid
that is mostly noise, and the tree is fitted on centroids.

```{r}
set.seed(1)
sub_of = rep(NA_character_, ncol(merged))
names(sub_of) = colnames(merged)

for (cl in clusters) {
  cl_cells = colnames(merged)[merged$final_label == cl]
  if (length(cl_cells) < 2 * params$min_node_cells) {
    sub_of[cl_cells] = paste0(cl, ".1")
    next
  }
  set.seed(1)
  snn = buildSNNGraph(merged[, cl_cells], k = params$sub_k,
                      use.dimred = "corrected")
  mem = igraph::cluster_louvain(snn)$membership
  names(mem) = cl_cells

  # merge anything too small into the nearest surviving subcluster
  co = reducedDim(merged, "corrected")[cl_cells, , drop = FALSE]
  repeat {
    tab   = table(mem)
    small = names(tab)[tab < params$min_node_cells]
    if (length(small) == 0 || length(tab) <= 1) break
    cent = do.call(rbind, lapply(names(tab), function(g)
      colMeans(co[mem == g, , drop = FALSE])))
    rownames(cent) = names(tab)
    d = as.matrix(dist(cent))
    diag(d) = Inf
    s = small[which.min(tab[small])]
    mem[mem == s] = names(tab)[which.min(d[s, ])]
  }
  sub_of[cl_cells] = paste0(cl, ".", match(mem, sort(unique(mem))))
}

stopifnot(!anyNA(sub_of))

node_levels = unique(sub_of)
node_levels = node_levels[order(match(sub(
  "\\..*$", "", node_levels), clusters),
  as.integer(sub("^.*\\.", "", node_levels)))]
merged$node   = factor(sub_of, levels = node_levels)
combined$node = merged$node

node_tab = as.data.frame(colData(merged)) %>%
  dplyr::count(node, final_label, name = "n_cells") %>%
  dplyr::arrange(node)

write.csv(node_tab, file = paste0(results_dir, "/node_composition.csv"),
          row.names = FALSE)

datatable(node_tab, rownames = FALSE, options = list(pageLength = 15))
```

`r paste0(length(node_levels), " nodes from ", length(clusters), " clusters, at sub_k = ", params$sub_k, ", with a floor of ", params$min_node_cells, " cells. Median node size ", round(median(as.integer(table(merged$node)))), " cells, smallest ", min(as.integer(table(merged$node))), ", largest ", max(as.integer(table(merged$node))), ". ", if (length(node_levels) < 2) "**Fewer than two nodes, so no tree can be built.**" else "")`

```{r}
per_cluster = node_tab %>%
  dplyr::count(final_label, name = "n_nodes") %>%
  dplyr::arrange(match(final_label, clusters))
datatable(per_cluster, rownames = FALSE, options = list(dom = "t", paging = FALSE))
```

## What the nodes are made of

Nodes are not cell types, so the question is not what each one is but how the
annotation is distributed over them. A node drawn from one annotation is a
waypoint inside that population. A node mixing several is a waypoint on a
boundary, which is expected wherever the biology is continuous.

```{r, fig.width=12, fig.height=5}
anno_by_node = as.data.frame(colData(merged)) %>%
  dplyr::count(node, annotation, name = "n") %>%
  dplyr::group_by(node) %>%
  dplyr::mutate(frac = n / sum(n)) %>%
  dplyr::ungroup()

write.csv(anno_by_node, file = paste0(results_dir, "/node_by_annotation.csv"),
          row.names = FALSE)

p_anno = ggplot(anno_by_node, aes(x = node, y = frac, fill = annotation)) +
  geom_col() +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = .5, size = 7),
        legend.text = element_text(size = 7)) +
  labs(x = NULL, y = "fraction of the node", fill = NULL,
       title = "Annotation composition of each node")

p_anno
ggsave(p_anno, file = paste0(results_dir, "/node_by_annotation.png"),
       width = 12, height = 5, dpi = 300)
```

# Root

Every node is scored on the germ cell marker sets. The root is the node with the
highest mean expression of the set named in `root_rule`, unless `root_node` names
one directly. This does not consult the annotation, so the root does not depend
on any label being right.

```{r}
germ_sets = c("Primordial germ cells", "Migrating germ cells",
              "Spermatogonial stem cells", "Spermatocytes", "Oocyte")

set_genes = read.csv(paste0(params$project_dir, "/data/", params$marker_file),
                     stringsAsFactors = FALSE) %>%
  dplyr::mutate(loc_id = trimws(loc_id)) %>%
  dplyr::filter(CellTypeMarkerSet %in% germ_sets) %>%
  dplyr::transmute(set = CellTypeMarkerSet, gene = loc_id)

set_scores = score_marker_sets(combined, merged$node, set_genes) %>%
  dplyr::rename(node = group)

write.csv(set_scores, file = paste0(results_dir, "/node_marker_set_scores.csv"),
          row.names = FALSE)

root_req = as.character(params$root_node)
if (nzchar(root_req)) {
  if (!root_req %in% node_levels)
    stop("root_node '", root_req, "' is not a node. Nodes: ",
         paste(node_levels, collapse = ", "))
  root      = root_req
  rule_set  = NA_character_
  root_pct  = NA_real_
  weak_root = FALSE
} else {
  rp = regmatches(params$root_rule,
                  regexec("^(max|min):(.+)$", params$root_rule))[[1]]
  if (length(rp) != 3)
    stop("root_rule '", params$root_rule, "' must look like max:<set> or min:<set>")
  rule_dir = rp[2]
  rule_set = trimws(rp[3])
  if (!rule_set %in% germ_sets)
    stop("root rule set '", rule_set, "' is not one of: ",
         paste(germ_sets, collapse = ", "))

  cand = set_scores %>% dplyr::filter(set == rule_set, !is.na(mean_logexpr))
  if (nrow(cand) == 0)
    stop("no gene from the ", rule_set, " set is in the data; set root_node")
  pick      = if (rule_dir == "max") which.max(cand$mean_logexpr) else
                which.min(cand$mean_logexpr)
  root      = as.character(cand$node[pick])
  root_pct  = cand$pct_detected[pick]
  weak_root = rule_dir == "max" && root_pct < params$min_detect_pct
}
```

`r paste0("Rooted at **", root, "**", if (nzchar(root_req)) ", set by root_node." else paste0(", the node with the highest mean expression of the ", rule_set, " set, detected in ", round(root_pct, 1), " percent of its cells."), if (weak_root) paste0(" **That is below min_detect_pct (", params$min_detect_pct, "), so the root rests on weak evidence.**") else "", " The root is an assumption. Rooting elsewhere gives a different ordering.")`

# Trajectory

`dist_method` sets how the distance between two nodes is measured. `simple` is
plain Euclidean distance between centroids and is the TSCAN default. `mnn`, the
default here, instead finds the mutual nearest neighbor pairs between each pair
of nodes and takes the median distance between them. It therefore measures how
close two nodes are at their *boundaries* rather than at their centers, which is
the right question when nodes are broad and intermingle. `mnn_k` sets how many
neighbors the search considers.

MNN runs a cell-based neighbor search for every pair of nodes, so it costs more
time and memory than `simple`, and that cost grows with the number of nodes.

`outgroup` decides whether every node must join one tree. With `FALSE` the tree
is forced to be connected, so a node that belongs nowhere is still grafted on at
its nearest neighbor. With `TRUE`, an edge longer than 1.5 times the median edge
length is rerouted through an artificial outgroup and the tree breaks into
components. That is how a node detaches itself rather than being removed by hand.

```{r}
pseudo_file     = paste0(Robject_out_dir, "/tscan_pseudo.Rds")
pseudo_sig_file = paste0(Robject_out_dir, "/tscan_pseudo_signature.txt")
pseudo_sig      = paste(obj_sig, root, params$sub_k, params$min_node_cells,
                        params$dist_method, params$mnn_k, params$outgroup,
                        paste(node_levels, collapse = ","), sep = "|")

if (file.exists(pseudo_file) && file.exists(pseudo_sig_file) &&
    identical(readLines(pseudo_sig_file, warn = FALSE)[1], pseudo_sig)) {
  pseudo = readRDS(pseudo_file)
} else {
  set.seed(1)
  pseudo = quickPseudotime(merged,
                           clusters    = merged$node,
                           use.dimred  = "corrected",
                           dist.method = as.character(params$dist_method),
                           mnn.k       = params$mnn_k,
                           outgroup    = params$outgroup,
                           start       = root)
  saveRDS(pseudo, file = pseudo_file)
  writeLines(pseudo_sig, pseudo_sig_file)
}

merged$pseudotime   = averagePseudotime(pseudo$ordering)
combined$pseudotime = merged$pseudotime
n_paths = ncol(pseudo$ordering)
n_comp  = igraph::count_components(pseudo$mst)
```

`r paste0(n_paths, " path", if (n_paths == 1) "" else "s", " from the root. The tree has ", n_comp, " component", if (n_comp == 1) "" else "s", if (n_comp > 1) paste0(", so outgroup splitting separated ", n_comp - 1, " group(s) from the main tree.") else ".", " ", sum(is.na(merged$pseudotime)), " cells have no pseudotime.")`

## Edge confidence

`createClusterMST` records a `gain` on every edge: the extra distance the tree
would need if that edge were removed, divided by the edge's own length. A small
gain means the edge is arbitrary, because an almost equally short path exists
without it. This is the tree's own statement about which connections to trust.

```{r}
edge_tab = igraph::as_data_frame(pseudo$mst, what = "edges") %>%
  dplyr::rename(node_1 = from, node_2 = to) %>%
  dplyr::mutate(length = signif(weight, 3),
                gain   = signif(gain, 3)) %>%
  dplyr::select(node_1, node_2, length, gain) %>%
  dplyr::arrange(gain)

write.csv(edge_tab, file = paste0(results_dir, "/mst_edges.csv"),
          row.names = FALSE)

datatable(edge_tab, rownames = FALSE, options = list(pageLength = 15))
```

`r if (nrow(edge_tab) == 0) "" else paste0("The least confident edge is ", edge_tab$node_1[1], " to ", edge_tab$node_2[1], " at a normalized gain of ", edge_tab$gain[1], ". ", sum(edge_tab$gain < 1), " of ", nrow(edge_tab), " edges have a gain below 1, meaning removing them would lengthen the tree by less than their own length.")`

```{r, fig.width=13, fig.height=5}
by_node = aggregateAcrossCells(
  merged, ids = merged$node,
  use.assay.type = if ("reconstructed" %in% assayNames(merged))
    "reconstructed" else "logcounts")
edges_umap = reportEdges(by_node, mst = pseudo$mst, clusters = NULL,
                         use.dimred = "UMAP")

p_anno_umap = plotUMAP(merged, colour_by = "annotation", point_size = .4) +
  geom_line(data = edges_umap, aes(x = UMAP1, y = UMAP2, group = edge)) +
  ggtitle("Annotation, carried per cell")
p_pt_umap = plotUMAP(merged, colour_by = "pseudotime", point_size = .4) +
  geom_line(data = edges_umap, aes(x = UMAP1, y = UMAP2, group = edge)) +
  ggtitle("Pseudotime")

p_anno_umap | p_pt_umap
ggsave(p_anno_umap | p_pt_umap,
       file = paste0(results_dir, "/trajectory_umap.png"), width = 13, height = 5)
```

# What the ordering says

This is where the trajectory is interpreted, and it is done from the cells
rather than from node names. The annotation is the collaborators' expert call
carried by barcode, so its distribution along pseudotime is the claim.

```{r, fig.width=12, fig.height=5}
pt_df = as.data.frame(colData(merged))
pt_df$barcode = colnames(merged)

p_anno_pt = ggplot(pt_df %>% dplyr::filter(!is.na(pseudotime)),
                   aes(x = pseudotime, y = reorder(annotation, pseudotime,
                                                   FUN = median))) +
  geom_boxplot(outlier.size = .2) +
  theme_bw() + labs(x = "pseudotime", y = NULL,
                    title = "Annotation along the ordering")
p_lib_pt = ggplot(pt_df %>% dplyr::filter(!is.na(pseudotime)),
                  aes(x = pseudotime, y = library, fill = library)) +
  geom_boxplot(outlier.size = .2) +
  scale_fill_manual(values = lib_cols) +
  theme_bw() + theme(legend.position = "none") +
  labs(x = "pseudotime", y = NULL, title = "Library")

p_anno_pt | p_lib_pt
ggsave(p_anno_pt | p_lib_pt,
       file = paste0(results_dir, "/pseudotime_by_annotation.png"),
       width = 12, height = 5)
```

```{r}
anno_order = pt_df %>%
  dplyr::filter(!is.na(pseudotime)) %>%
  dplyr::group_by(annotation) %>%
  dplyr::summarise(n_cells = dplyr::n(),
                   median_pseudotime = signif(median(pseudotime), 3),
                   q1 = signif(quantile(pseudotime, .25), 3),
                   q3 = signif(quantile(pseudotime, .75), 3),
                   .groups = "drop") %>%
  dplyr::arrange(median_pseudotime)

write.csv(anno_order, file = paste0(results_dir, "/annotation_pseudotime.csv"),
          row.names = FALSE)

datatable(anno_order, rownames = FALSE, options = list(dom = "t", paging = FALSE))
```

## Marker sets along the ordering

Independent of both the node cuts and the annotation. If the ordering is
developmental, the germ marker sets should move along it.

```{r, fig.width=12, fig.height=4}
mk_long = dplyr::bind_rows(lapply(germ_sets, function(s) {
  g = intersect(unique(set_genes$gene[set_genes$set == s]), rownames(combined))
  if (length(g) == 0) return(NULL)
  m = as.matrix(logcounts(combined)[g, , drop = FALSE])
  data.frame(set = s, pseudotime = merged$pseudotime,
             score = colMeans(m), stringsAsFactors = FALSE)
})) %>% dplyr::filter(!is.na(pseudotime))

mk_stats = mk_long %>%
  dplyr::group_by(set) %>%
  dplyr::summarise(rho = signif(suppressWarnings(
                     cor(pseudotime, score, method = "spearman")), 3),
                   p = signif(suppressWarnings(
                     cor.test(pseudotime, score, method = "spearman")$p.value), 3),
                   .groups = "drop")

write.csv(mk_stats, file = paste0(results_dir, "/marker_sets_vs_pseudotime.csv"),
          row.names = FALSE)

p_mk = ggplot(mk_long, aes(x = pseudotime, y = score)) +
  geom_point(size = .15, alpha = .2) +
  geom_smooth(method = "loess", se = FALSE, colour = "#b2182b", linewidth = .7) +
  facet_wrap(~ set, scales = "free_y", nrow = 1) +
  theme_bw() + labs(x = "pseudotime", y = "mean log expression")

p_mk
ggsave(p_mk, file = paste0(results_dir, "/marker_sets_vs_pseudotime.png"),
       width = 12, height = 4, dpi = 300)

datatable(mk_stats, rownames = FALSE, options = list(dom = "t", paging = FALSE))
```

## What else the ordering tracks

Pooled across libraries, a trend can be the libraries sitting at different ends
of the tree. The within-library figures cannot be that.

```{r}
cnt = counts(combined)
tot = Matrix::colSums(cnt)
qc = data.frame(
  barcode    = colnames(combined),
  pseudotime = merged$pseudotime,
  library    = as.character(merged$library),
  total      = as.numeric(tot),
  detected   = as.numeric(Matrix::colSums(cnt > 0)),
  pct_mito   = 100 * as.numeric(Matrix::colSums(
                 cnt[intersect(mito_ids, rownames(cnt)), , drop = FALSE])) / tot,
  pct_rrna   = 100 * as.numeric(Matrix::colSums(
                 cnt[intersect(rrna_ids, rownames(cnt)), , drop = FALSE])) / tot,
  pct_ribo   = 100 * as.numeric(Matrix::colSums(
                 cnt[intersect(ribo_ids, rownames(cnt)), , drop = FALSE])) / tot,
  stringsAsFactors = FALSE) %>%
  dplyr::filter(!is.na(pseudotime))

conf_measures = c("total", "detected", "pct_mito", "pct_rrna", "pct_ribo")
conf_stats = function(d, label) {
  if (nrow(d) < 20) return(NULL)
  dplyr::bind_rows(lapply(conf_measures, function(v) {
    ok = is.finite(d[[v]]) & is.finite(d$pseudotime)
    if (sum(ok) < 20) return(NULL)
    x = if (v %in% c("total", "detected")) log10(d[[v]][ok]) else d[[v]][ok]
    if (stats::sd(x) == 0) return(NULL)
    ct = suppressWarnings(cor.test(x, d$pseudotime[ok], method = "spearman"))
    data.frame(within = label, measure = v, n_cells = sum(ok),
               r2 = signif(summary(lm(d$pseudotime[ok] ~ x))$r.squared, 3),
               spearman_rho = signif(unname(ct$estimate), 3),
               p = signif(ct$p.value, 3), stringsAsFactors = FALSE)
  }))
}

conf_tab = dplyr::bind_rows(
  conf_stats(qc, "all cells"),
  dplyr::bind_rows(lapply(sort(unique(qc$library)), function(lb)
    conf_stats(qc[qc$library == lb, , drop = FALSE], lb))))

write.csv(conf_tab, file = paste0(results_dir, "/pseudotime_vs_technical.csv"),
          row.names = FALSE)

datatable(conf_tab, rownames = FALSE, options = list(pageLength = 15))
```

`r if (nrow(conf_tab) == 0) "" else { pooled = conf_tab %>% dplyr::filter(within == "all cells"); top = pooled[which.max(pooled$r2), ]; wit = conf_tab %>% dplyr::filter(within != "all cells", measure == top$measure); paste0("Pooled, the strongest is ", top$measure, " at ", round(100 * top$r2, 1), " percent of the variance in pseudotime. Within libraries the same measure gives ", paste0(wit$within, " ", round(100 * wit$r2, 1), " percent", collapse = ", "), ". A pooled figure much larger than the within-library ones is the libraries at different ends of the tree rather than a gradient along it.") }`

# Pseudotime-associated genes

```{r}
have_pt = sum(!is.na(merged$pseudotime)) > 20
```

```{r, eval=have_pt}
pt_res = testPseudotime(merged, pseudotime = merged$pseudotime)
pt_tab = as.data.frame(pt_res) %>%
  tibble::rownames_to_column("Gene") %>%
  dplyr::filter(!is.na(FDR), FDR < params$fdr_cut) %>%
  dplyr::left_join(gene_description, by = "Gene") %>%
  dplyr::arrange(FDR) %>%
  dplyr::select(Gene, WangGeneID, description, logFC, p.value, FDR)

write.csv(pt_tab, file = paste0(results_dir, "/pseudotime_genes.csv"),
          row.names = FALSE)

datatable(pt_tab %>% dplyr::mutate(dplyr::across(where(is.numeric),
                                                 ~signif(., 3))),
          rownames = FALSE, filter = "top", options = list(pageLength = 15))
```

`r if (!have_pt) "Too few cells have a pseudotime for gene testing." else paste0(nrow(pt_tab), " genes change along the ordering at FDR < ", params$fdr_cut, ".")`

# Export

```{r}
out = data.frame(barcode    = colnames(merged),
                 node       = as.character(merged$node),
                 cluster    = as.character(merged$final_label),
                 annotation = as.character(merged$annotation),
                 library    = as.character(merged$library),
                 stage      = as.character(merged$stage),
                 pseudotime = merged$pseudotime,
                 stringsAsFactors = FALSE)

write.csv(out, file = paste0(results_dir, "/pseudotime_per_cell.csv"),
          row.names = FALSE)
```

`r paste0("Written to ", results_dir, ".")`

# Session info

```{r}
sessionInfo()
```
