---
title: "19_germ_pseudotime_fine_nodes"
author: "Stephanie Hickey"
date: "`r Sys.Date()`"
output:
  html_document:
    code_folding: hide
    toc: true
    toc_float:
      toc_collapsed: true
params:
    project_dir: /mnt/ufs18/rs-013/bioinformaticsCore/projects/chong_davidson/BCC125_sea_lamprey_scRNA
    project_name: BCC125_sea_lamprey_scRNA
    lab: Weiming Li
    results_date: "20260813"
    atlas_suffix: ""
    lineage: "all"
    female_clusters: "5"
    male_clusters: "15,16,17,18"
    all_clusters: "5,15,16,17,18"
    sub_k: 10
    min_node_cells: 30
    dist_method: "simple"
    edge_boot: 100
    edge_boot_cells: 100
    mnn_k: 50
    outgroup: FALSE
    root_rule: "max:Primordial germ cells"
    root_node: ""
    min_detect_pct: 5
    marker_file: canonical_markers_S1_only_2026-08-14.csv
    prev_nodes_file: ""
    node_annotation: "5.1=Migrating neural crest cells;5.2=Oogonia;5.3=Oocytes;5.4=Migrating female germ cells (FGC I);5.5=Granulosa;5.6=Migrating female germ cells (FGC II);5.7=Primordial germ cells (PGCs);5.8=Migrating PGCs;15=Spermatogonia;16=Spermatocytes I;17=Spermatocytes II;18=Spermatocytes III"
    min_library_cells: 50
    fdr_cut: 0.01
    go_pval: 0.05
    genome_dir: /mnt/research/bioinformaticsCore/shared/Genomes/Petromyzon_marinus
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

```{r, echo=FALSE, results="asis"}
cat("\n")
cat("#### Lab:", params$lab, " \n")
cat("#### Project:", params$project_name, " \n")
cat("#### Lineage:", params$lineage, " \n")
```

```{r}
library(tidyverse)
library(SingleCellExperiment)
library(scater)
library(scran)
library(batchelor)
library(BiocSingular)
library(TSCAN)
library(patchwork)
library(DT)
library(ggraph)
library(tidygraph)
library(ggrepel)
source(paste0(params$project_dir, "/src/scRNA_seq_functions.R"))
```

TSCAN pseudotime for the `r params$lineage` germ cells, with **every** cluster
subclustered finely and no cell removed.

The tree TSCAN fits is a minimum spanning tree over one centroid per node, so
the number of nodes is the number of points the trajectory is fitted to. `17`
gives it around a dozen, of which four are whole clusters holding thousands of
cells each. A tree on twelve points is coarse and moves whenever a node moves.

This document treats nodes as waypoints rather than cell types. Every cluster is
cut into small subclusters at a fixed `sub_k`, which gives the tree many more
points to follow. Nothing here needs a node to be interpretable, so nothing here
depends on the subclustering being stable or on any node carrying a label.

The collaborators' annotation is carried **per cell**, by barcode, from the run
named in `prev_nodes_file`. It is never attached to a node. Re-clustering
therefore cannot break it, and the interpretation is read off the composition of
the trajectory rather than off node names.

```{r}
lineage = as.character(params$lineage)
if (!lineage %in% c("female", "male", "all"))
  stop("lineage must be 'female', 'male' or 'all', not '", lineage, "'")

clusters = trimws(strsplit(as.character(switch(lineage,
  female = params$female_clusters,
  male   = params$male_clusters,
  all    = params$all_clusters)), ",")[[1]])
clusters = clusters[nzchar(clusters)]

object_dir      = paste0(params$project_dir, "/results/", params$results_date,
                         "_no_doublets")
atlas_dir       = paste0(params$project_dir, "/results/", params$results_date,
                         "_atlas", params$atlas_suffix)
results_dir     = paste0(params$project_dir, "/results/", params$results_date,
                         "_pseudotime_fine_", lineage, params$atlas_suffix)
Robject_out_dir = paste0(results_dir, "/Robjects")
if (!dir.exists(Robject_out_dir)) dir.create(Robject_out_dir, recursive = TRUE)

label_file = paste0(atlas_dir, "/cell_cluster_labels.csv")
if (!file.exists(label_file))
  stop("missing ", label_file, ". Run 13 first.")

cell_labels      = read.csv(label_file, stringsAsFactors = FALSE)
all.sce          = readRDS(paste0(object_dir, "/all.sce.Rds"))
gene_description = read.csv(paste0(object_dir, "/gene_descritption.csv"),
                            stringsAsFactors = FALSE) %>%
  dplyr::distinct(Gene, .keep_all = TRUE)

lib_cols = c("Female" = "#FF6DB6", "Male" = "#006ddb",
             "Undifferentiated.1" = "#009292", "Undifferentiated.2" = "#b66ddf")

tech     = technical_gene_ids(gene_description)
mito_ids = tech$mito
rrna_ids = tech$rrna
ribo_ids = tech$ribo
```

# Cells

```{r}
absent = setdiff(clusters, cell_labels$final_label)
if (length(absent) > 0)
  stop("cluster(s) ", paste(absent, collapse = ", "), " are not in ", label_file,
       ". Labels present: ",
       paste(sort(unique(cell_labels$final_label)), collapse = ", "))

lineage_key = cell_labels %>% dplyr::filter(final_label %in% clusters)

# the annotation, per cell, by barcode. Nothing is dropped on the strength of it.
parse_pairs = function(x) {
  x = trimws(x)
  if (!nzchar(x)) return(character(0))
  kv = strsplit(trimws(strsplit(x, ";")[[1]]), "=")
  setNames(vapply(kv, function(z) trimws(z[2]), character(1)),
           vapply(kv, function(z) trimws(z[1]), character(1)))
}

anno_map  = parse_pairs(params$node_annotation)
prev_file = as.character(params$prev_nodes_file)
have_prev = nzchar(prev_file) && file.exists(prev_file)

cell_anno = rep(NA_character_, nrow(lineage_key))
if (have_prev) {
  prev = read.csv(prev_file, stringsAsFactors = FALSE)
  stopifnot(all(c("barcode", "node") %in% colnames(prev)))
  old = prev$node[match(lineage_key$barcode, prev$barcode)]
  cell_anno = unname(anno_map[as.character(old)])
  cell_anno[is.na(cell_anno) & !is.na(old)] = as.character(old)[is.na(cell_anno) & !is.na(old)]
}
lineage_key$annotation = ifelse(is.na(cell_anno), "unannotated", cell_anno)

comp_tab = lineage_key %>%
  dplyr::count(final_label, annotation, name = "n") %>%
  tidyr::pivot_wider(names_from = annotation, values_from = n, values_fill = 0)

write.csv(comp_tab, file = paste0(results_dir, "/cluster_by_annotation.csv"),
          row.names = FALSE)

datatable(comp_tab, rownames = FALSE, options = list(dom = "t", scrollX = TRUE))
```

`r paste0(format(nrow(lineage_key), big.mark = ","), " cells from cluster(s) ", paste(clusters, collapse = ", "), ". No cell is removed, so 5.1 is present.", if (!have_prev) " **prev_nodes_file is not set**, so no annotation is carried and the trajectory cannot be read against it." else paste0(" ", format(sum(lineage_key$annotation != "unannotated"), big.mark = ","), " carry an annotation from ", basename(prev_file), "."))`

# Corrected space

```{r}
obj_file     = paste0(Robject_out_dir, "/lineage_objects.Rds")
obj_sig_file = paste0(Robject_out_dir, "/lineage_objects_signature.txt")
obj_sig      = paste(lineage, paste(sort(clusters), collapse = ","),
                     nrow(lineage_key), params$min_library_cells, sep = "|")

if (file.exists(obj_file) && file.exists(obj_sig_file) &&
    identical(readLines(obj_sig_file, warn = FALSE)[1], obj_sig)) {
  objs = readRDS(obj_file)
} else {
  objs = build_lineage_sce(all.sce, lineage_key$barcode,
                           min_library_cells = params$min_library_cells)
  saveRDS(objs, file = obj_file)
  writeLines(obj_sig, obj_sig_file)
}

merged   = objs$corrected
combined = objs$combined

key_idx = match(colnames(merged), lineage_key$barcode)
stopifnot(!anyNA(key_idx))
merged$library     = lineage_key$library[key_idx]
merged$stage       = lineage_key$stage[key_idx]
merged$final_label = lineage_key$final_label[key_idx]
merged$annotation  = lineage_key$annotation[key_idx]
```

`r paste0(format(ncol(merged), big.mark = ","), " cells from ", length(objs$kept_batches), " librar", if (length(objs$kept_batches) == 1) "y" else "ies", " are kept.")`

# Fine nodes

Every cluster is subclustered in the shared corrected space, using only that
cluster's cells, at a fixed `sub_k`. A smaller `sub_k` gives more and smaller
subclusters. There is no stability sweep, because nothing downstream asks a node
to be a reproducible cell type.

Subclusters below `min_node_cells` are merged into the nearest subcluster of the
same parent cluster, by centroid distance. A node of five cells gives a centroid
that is mostly noise, and the tree is fitted on centroids.

```{r}
set.seed(1)
sub_of = rep(NA_character_, ncol(merged))
names(sub_of) = colnames(merged)

for (cl in clusters) {
  cl_cells = colnames(merged)[merged$final_label == cl]
  if (length(cl_cells) < 2 * params$min_node_cells) {
    sub_of[cl_cells] = paste0(cl, ".1")
    next
  }
  set.seed(1)
  snn = buildSNNGraph(merged[, cl_cells], k = params$sub_k,
                      use.dimred = "corrected")
  mem = igraph::cluster_louvain(snn)$membership
  names(mem) = cl_cells

  # merge anything too small into the nearest surviving subcluster
  co = reducedDim(merged, "corrected")[cl_cells, , drop = FALSE]
  repeat {
    tab   = table(mem)
    small = names(tab)[tab < params$min_node_cells]
    if (length(small) == 0 || length(tab) <= 1) break
    cent = do.call(rbind, lapply(names(tab), function(g)
      colMeans(co[mem == g, , drop = FALSE])))
    rownames(cent) = names(tab)
    d = as.matrix(dist(cent))
    diag(d) = Inf
    s = small[which.min(tab[small])]
    mem[mem == s] = names(tab)[which.min(d[s, ])]
  }
  sub_of[cl_cells] = paste0(cl, ".", match(mem, sort(unique(mem))))
}

stopifnot(!anyNA(sub_of))

node_levels = unique(sub_of)
node_levels = node_levels[order(match(sub(
  "\\..*$", "", node_levels), clusters),
  as.integer(sub("^.*\\.", "", node_levels)))]
merged$node   = factor(sub_of, levels = node_levels)
combined$node = merged$node

node_tab = as.data.frame(colData(merged)) %>%
  dplyr::count(node, final_label, name = "n_cells") %>%
  dplyr::arrange(node)

write.csv(node_tab, file = paste0(results_dir, "/node_composition.csv"),
          row.names = FALSE)

datatable(node_tab, rownames = FALSE, options = list(pageLength = 15))
```

`r paste0(length(node_levels), " nodes from ", length(clusters), " clusters, at sub_k = ", params$sub_k, ", with a floor of ", params$min_node_cells, " cells. Median node size ", round(median(as.integer(table(merged$node)))), " cells, smallest ", min(as.integer(table(merged$node))), ", largest ", max(as.integer(table(merged$node))), ". ", if (length(node_levels) < 2) "**Fewer than two nodes, so no tree can be built.**" else "")`

```{r}
per_cluster = node_tab %>%
  dplyr::count(final_label, name = "n_nodes") %>%
  dplyr::arrange(match(final_label, clusters))
datatable(per_cluster, rownames = FALSE, options = list(dom = "t", paging = FALSE))
```

## What the nodes are made of

Nodes are not cell types, so the question is not what each one is but how the
annotation is distributed over them. A node drawn from one annotation is a
waypoint inside that population. A node mixing several is a waypoint on a
boundary, which is expected wherever the biology is continuous.

```{r, fig.width=12, fig.height=5}
anno_by_node = as.data.frame(colData(merged)) %>%
  dplyr::count(node, annotation, name = "n") %>%
  dplyr::group_by(node) %>%
  dplyr::mutate(frac = n / sum(n)) %>%
  dplyr::ungroup()

write.csv(anno_by_node, file = paste0(results_dir, "/node_by_annotation.csv"),
          row.names = FALSE)

p_anno = ggplot(anno_by_node, aes(x = node, y = frac, fill = annotation)) +
  geom_col() +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = .5, size = 7),
        legend.text = element_text(size = 7)) +
  labs(x = NULL, y = "fraction of the node", fill = NULL,
       title = "Annotation composition of each node")

p_anno
ggsave(p_anno, file = paste0(results_dir, "/node_by_annotation.png"),
       width = 12, height = 5, dpi = 300)
```

# Root

Every node is scored on the germ cell marker sets. The root is the node with the
highest mean expression of the set named in `root_rule`, unless `root_node` names
one directly. This does not consult the annotation, so the root does not depend
on any label being right.

```{r}
germ_sets = c("Primordial germ cells", "Migrating germ cells",
              "Spermatogonial stem cells", "Spermatocytes", "Oocyte")

set_genes = read.csv(paste0(params$project_dir, "/data/", params$marker_file),
                     stringsAsFactors = FALSE) %>%
  dplyr::mutate(loc_id = trimws(loc_id)) %>%
  dplyr::filter(CellTypeMarkerSet %in% germ_sets) %>%
  dplyr::transmute(set = CellTypeMarkerSet, gene = loc_id)

set_scores = score_marker_sets(combined, merged$node, set_genes) %>%
  dplyr::rename(node = group)

write.csv(set_scores, file = paste0(results_dir, "/node_marker_set_scores.csv"),
          row.names = FALSE)

root_req = as.character(params$root_node)
if (nzchar(root_req)) {
  if (!root_req %in% node_levels)
    stop("root_node '", root_req, "' is not a node. Nodes: ",
         paste(node_levels, collapse = ", "))
  root      = root_req
  rule_set  = NA_character_
  root_pct  = NA_real_
  weak_root = FALSE
} else {
  rp = regmatches(params$root_rule,
                  regexec("^(max|min):(.+)$", params$root_rule))[[1]]
  if (length(rp) != 3)
    stop("root_rule '", params$root_rule, "' must look like max:<set> or min:<set>")
  rule_dir = rp[2]
  rule_set = trimws(rp[3])
  if (!rule_set %in% germ_sets)
    stop("root rule set '", rule_set, "' is not one of: ",
         paste(germ_sets, collapse = ", "))

  cand = set_scores %>% dplyr::filter(set == rule_set, !is.na(mean_logexpr))
  if (nrow(cand) == 0)
    stop("no gene from the ", rule_set, " set is in the data; set root_node")
  pick      = if (rule_dir == "max") which.max(cand$mean_logexpr) else
                which.min(cand$mean_logexpr)
  root      = as.character(cand$node[pick])
  root_pct  = cand$pct_detected[pick]
  weak_root = rule_dir == "max" && root_pct < params$min_detect_pct
}
```

`r paste0("Rooted at **", root, "**", if (nzchar(root_req)) ", set by root_node." else paste0(", the node with the highest mean expression of the ", rule_set, " set, detected in ", round(root_pct, 1), " percent of its cells."), if (weak_root) paste0(" **That is below min_detect_pct (", params$min_detect_pct, "), so the root rests on weak evidence.**") else "", " The root is an assumption. Rooting elsewhere gives a different ordering.")`

# Trajectory

`dist_method` sets how the distance between two nodes is measured. `simple` is
plain Euclidean distance between centroids and is the TSCAN default. `mnn`, the
default here, instead finds the mutual nearest neighbor pairs between each pair
of nodes and takes the median distance between them. It therefore measures how
close two nodes are at their *boundaries* rather than at their centers, which is
the right question when nodes are broad and intermingle. `mnn_k` sets how many
neighbors the search considers.

MNN runs a cell-based neighbor search for every pair of nodes, so it costs more
time and memory than `simple`, and that cost grows with the number of nodes.

`outgroup` decides whether every node must join one tree. With `FALSE` the tree
is forced to be connected, so a node that belongs nowhere is still grafted on at
its nearest neighbor. With `TRUE`, an edge longer than 1.5 times the median edge
length is rerouted through an artificial outgroup and the tree breaks into
components. That is how a node detaches itself rather than being removed by hand.

```{r}
pseudo_file     = paste0(Robject_out_dir, "/tscan_pseudo.Rds")
pseudo_sig_file = paste0(Robject_out_dir, "/tscan_pseudo_signature.txt")
pseudo_sig      = paste(obj_sig, root, params$sub_k, params$min_node_cells,
                        params$dist_method, params$mnn_k, params$outgroup,
                        paste(node_levels, collapse = ","), sep = "|")

if (file.exists(pseudo_file) && file.exists(pseudo_sig_file) &&
    identical(readLines(pseudo_sig_file, warn = FALSE)[1], pseudo_sig)) {
  pseudo = readRDS(pseudo_file)
} else {
  set.seed(1)
  pseudo = quickPseudotime(merged,
                           clusters    = merged$node,
                           use.dimred  = "corrected",
                           dist.method = as.character(params$dist_method),
                           mnn.k       = params$mnn_k,
                           outgroup    = params$outgroup,
                           start       = root)
  saveRDS(pseudo, file = pseudo_file)
  writeLines(pseudo_sig, pseudo_sig_file)
}

merged$pseudotime   = averagePseudotime(pseudo$ordering)
combined$pseudotime = merged$pseudotime
n_paths = ncol(pseudo$ordering)
n_comp  = igraph::count_components(pseudo$mst)
```

`r paste0(n_paths, " path", if (n_paths == 1) "" else "s", " from the root. The tree has ", n_comp, " component", if (n_comp == 1) "" else "s", if (n_comp > 1) paste0(", so outgroup splitting separated ", n_comp - 1, " group(s) from the main tree.") else ".", " ", sum(is.na(merged$pseudotime)), " cells have no pseudotime.")`

## Edge confidence

An MST always returns a tree. It connects every node whether or not the data
supports the connection, so the shape on its own says nothing about how much of
it is real.

Five columns describe each edge from a different direction.

| column | what it is | low | high |
|---|---|---|---|
| `length` | distance between the two node centroids, in the units of the corrected space | the two nodes sit close together | the two nodes are far apart |
| `nn_rank_1`, `nn_rank_2` | where each node sits in the other's list of nodes ordered by distance | 1 and 1 are mutual nearest neighbors | the edge reaches past nearer nodes to make the connection |
| `separation` | centroid distance in standard deviations of the within-node spread | below about 1 the two nodes are one continuous population | two groups distinct enough to tell apart |
| `pct_of_cut` | the length as a percentage of the outgroup threshold | the edge survives outgroup splitting | over 100, outgroup splitting removes it |
| `frequency` | percentage of subsampled trees containing the edge | a different sample of the same cells would not have drawn this edge | the edge is drawn whatever the sample |

`frequency` comes from drawing `edge_boot_cells` cells from each node
`edge_boot` times, recomputing the centroids and rebuilding the tree. The count
is fixed per node rather than proportional. A node of 5,000 cells has a centroid
so precisely estimated that dropping a fifth of them moves it almost not at all,
so a proportional subsample returns close to 100 percent for every edge
including the arbitrary ones. Read it as a stress test rather than as a p-value.

A low `separation` is expected between consecutive waypoints on a trajectory.
It is the reading to question where the two nodes carry annotations for distinct
cell types. Combined with the ranks it separates cases that look alike on the
figure.

| `separation` | `nn_rank` | reading |
|---|---|---|
| low | 1 and 1 | one continuous population, joined at its closest point |
| low | high | overlapping nodes, joined the long way round |
| high | 1 and 1 | two distinct groups with a genuine link between them |
| high | high | a long edge to a node that sits near nothing |

No column says whether a low `separation` means the annotation is wrong or the
subclustering was cut too fine. That comes from the marker sets.

```{r}
edge_key = function(a, b) paste(pmin(a, b), pmax(a, b), sep = " -- ")

edge_tab = igraph::as_data_frame(pseudo$mst, what = "edges") %>%
  dplyr::rename(node_1 = from, node_2 = to) %>%
  dplyr::mutate(length = signif(weight, 3),
                key = edge_key(node_1, node_2))

# The outgroup rule, reported whether or not it is switched on. With
# outgroup = TRUE, createClusterMST puts an artificial node 1.5 times the median
# edge length away from every real node. An edge longer than that is beaten by a
# pair of edges through the outgroup, and deleting the outgroup leaves the tree
# in pieces. The median is taken from the tree built without an outgroup, which
# is the reference the package itself uses.
og_cut = if (is.numeric(params$outgroup)) {
  as.numeric(params$outgroup)
} else {
  ref = if (!isTRUE(params$outgroup)) edge_tab$length else {
    m0 = try(TrajectoryUtils::createClusterMST(
               reducedDim(merged, "corrected"),
               clusters    = as.character(merged$node),
               dist.method = as.character(params$dist_method),
               mnn.k       = params$mnn_k, outgroup = FALSE), silent = TRUE)
    if (inherits(m0, "try-error")) NULL else
      igraph::as_data_frame(m0, what = "edges")$weight
  }
  if (length(ref) == 0) NA_real_ else 1.5 * stats::median(ref)
}

# under 100 the edge survives outgroup splitting, over 100 it is cut
edge_tab$pct_of_cut = if (is.na(og_cut)) NA_real_ else
  round(100 * edge_tab$length / og_cut)

# Two reads on each edge that need no threshold, both from the node centroids of
# the corrected space.
#
#   nn_rank_1   where node_2 sits in node_1's list of nodes ordered by distance
#   nn_rank_2   where node_1 sits in node_2's
#   separation  centroid distance in standard deviations of the within-node
#               spread, so 2 means the centroids sit 2 SD apart
#
# Both are Euclidean, which is what dist.method = "simple" uses. Under any other
# dist.method they describe the corrected space rather than the distances the
# tree was built from.
co_all = reducedDim(merged, "corrected")
nd_all = as.character(merged$node)
cnt    = table(nd_all)
cent   = rowsum(co_all, nd_all)[names(cnt), , drop = FALSE] / as.vector(cnt)

# Per-dimension within-node standard deviation, pooled over the dimensions of
# the corrected space. The distance from a cell to its own centroid is not used
# as the denominator: it grows as the square root of the dimension count while
# the centroid distance does not, so the ratio would shrink as dimensions are
# added. Two nodes four standard deviations apart score 0.56 on that version in
# fifty dimensions and 4 on this one.
node_sd = vapply(rownames(cent), function(g) {
  ix = which(nd_all == g)
  if (length(ix) < 2) NA_real_ else
    sqrt(mean(apply(co_all[ix, , drop = FALSE], 2, var)))
}, numeric(1))

node_dist = as.matrix(dist(cent))
diag(node_dist) = Inf
node_rank = t(apply(node_dist, 1, rank, ties.method = "min"))

edge_tab$nn_rank_1 = node_rank[cbind(edge_tab$node_1, edge_tab$node_2)]
edge_tab$nn_rank_2 = node_rank[cbind(edge_tab$node_2, edge_tab$node_1)]
edge_tab$separation = round(
  node_dist[cbind(edge_tab$node_1, edge_tab$node_2)] /
  ((node_sd[edge_tab$node_1] + node_sd[edge_tab$node_2]) / 2), 2)

n_boot  = as.integer(params$edge_boot)
boot_ok = n_boot > 0

if (boot_ok) {
  co_b    = reducedDim(merged, "corrected")
  nd_b    = as.character(merged$node)
  by_node = split(seq_along(nd_b), nd_b)

  tally = table(unlist(lapply(seq_len(n_boot), function(i) {
    set.seed(1000 + i)
    keep = unlist(lapply(by_node, function(ix)
      sample(ix, min(length(ix), max(2L, as.integer(params$edge_boot_cells))))))
    m = try(TrajectoryUtils::createClusterMST(
              co_b[keep, , drop = FALSE], clusters = nd_b[keep],
              dist.method = as.character(params$dist_method),
              mnn.k = params$mnn_k, outgroup = params$outgroup),
            silent = TRUE)
    if (inherits(m, "try-error")) return(NULL)
    e = igraph::as_data_frame(m, what = "edges")
    unique(edge_key(e$from, e$to))
  })))

  edge_tab$frequency = round(100 * as.integer(tally[edge_tab$key]) / n_boot)
  edge_tab$frequency[is.na(edge_tab$frequency)] = 0
} else {
  edge_tab$frequency = NA_integer_
}

edge_tab = edge_tab %>%
  dplyr::select(node_1, node_2, length, nn_rank_1, nn_rank_2, separation,
                pct_of_cut, frequency) %>%
  dplyr::arrange(dplyr::desc(is.na(frequency)), frequency, length)

write.csv(edge_tab, file = paste0(results_dir, "/mst_edges.csv"),
          row.names = FALSE)

# eight narrow columns: no paging unless the tree is large, numbers right
# aligned, and horizontal scroll rather than wrapped headers
datatable(edge_tab, rownames = FALSE, class = "compact stripe hover",
          options = list(dom = if (nrow(edge_tab) > 25) "tp" else "t",
                         paging = nrow(edge_tab) > 25, pageLength = 25,
                         scrollX = TRUE, autoWidth = FALSE,
                         columnDefs = list(list(className = "dt-right",
                                                targets = 2:7))))
```

`r if (!boot_ok || nrow(edge_tab) == 0) "The subsampling check is off. Set edge_boot to a positive number to run it." else paste0("The least confident edge is ", edge_tab$node_1[1], " to ", edge_tab$node_2[1], ", drawn in ", edge_tab$frequency[1], " percent of ", n_boot, " trees built from ", params$edge_boot_cells, " cells per node. ", sum(edge_tab$frequency < 50), " of ", nrow(edge_tab), " edges appear in under half of them, and ", sum(edge_tab$frequency == 100), " in all of them.")`

`r if (is.na(og_cut)) "" else { over = edge_tab %>% dplyr::filter(pct_of_cut > 100); paste0("The outgroup threshold for this tree is ", signif(og_cut, 3), ", which is 1.5 times the median edge length. ", if (nrow(over) == 0) "No edge is longer than that, so outgroup = TRUE returns this same single tree." else paste0(nrow(over), " of ", nrow(edge_tab), " edges are longer: ", paste0(over$node_1, " to ", over$node_2, " at ", over$pct_of_cut, " percent", collapse = ", "), ". Under outgroup = TRUE the tree is rebuilt with the artificial node present, so the pieces are not guaranteed to be this tree minus those edges.")) }`

`r if (!"separation" %in% names(edge_tab)) "" else { far = edge_tab %>% dplyr::filter(pmax(nn_rank_1, nn_rank_2) > 2); ov = edge_tab %>% dplyr::filter(separation < 1); paste0(sum(edge_tab$nn_rank_1 == 1 & edge_tab$nn_rank_2 == 1), " of ", nrow(edge_tab), " edges join mutual nearest neighbors. ", if (nrow(far) == 0) "No edge reaches past a node's two nearest neighbors." else paste0(nrow(far), " reach further than either node's second nearest: ", paste0(far$node_1, " to ", far$node_2, " at ranks ", far$nn_rank_1, " and ", far$nn_rank_2, collapse = ", "), "."), " ", if (nrow(ov) == 0) "Every pair sits at least one standard deviation apart." else paste0(nrow(ov), " pair(s) sit under one standard deviation apart: ", paste0(ov$node_1, " to ", ov$node_2, " at ", ov$separation, collapse = ", "), ".")) }`

```{r, fig.width=13, fig.height=5}
by_node = aggregateAcrossCells(
  merged, ids = merged$node,
  use.assay.type = if ("reconstructed" %in% assayNames(merged))
    "reconstructed" else "logcounts")
edges_umap = reportEdges(by_node, mst = pseudo$mst, clusters = NULL,
                         use.dimred = "UMAP")

p_anno_umap = plotUMAP(merged, colour_by = "annotation", point_size = .4) +
  geom_line(data = edges_umap, aes(x = UMAP1, y = UMAP2, group = edge)) +
  ggtitle("Annotation, carried per cell")
p_pt_umap = plotUMAP(merged, colour_by = "pseudotime", point_size = .4) +
  geom_line(data = edges_umap, aes(x = UMAP1, y = UMAP2, group = edge)) +
  ggtitle("Pseudotime")

p_anno_umap | p_pt_umap
ggsave(p_anno_umap | p_pt_umap,
       file = paste0(results_dir, "/trajectory_umap.png"), width = 13, height = 5)
```

## Node graph

The tree on its own, laid out by its own structure rather than on the UMAP. With
many small nodes the shape of the tree is the thing to read, and a UMAP can put
adjacent nodes far apart. Nodes are sized by cell count, colored by mean
pseudotime in the first panel and by the annotation holding the most cells in the
second. Edge color is the percentage of subsampled trees containing that edge. The
numbers are omitted with this many nodes; they are in `mst_edges.csv`. Edges are
drawn plain grey when the subsampling is switched off.

Edge length on this figure is meaningful. The layout is weighted by the real
distance between nodes, so a long edge means the two nodes are far apart in the
corrected space. Those distances are the `length` column of `mst_edges.csv`. The
reproduction is not exact, because a tree in fifty dimensions cannot always be
drawn faithfully in two, but it is close to a uniform rescale rather than
arbitrary. If the message about a failed layout appears above, the figure fell
back to an unweighted layout and edge lengths mean nothing.

```{r, fig.width=11, fig.height=13}
node_stats = as.data.frame(colData(merged)) %>%
  dplyr::group_by(node) %>%
  dplyr::summarise(n_cells = dplyr::n(),
                   mean_pseudo = mean(pseudotime, na.rm = TRUE),
                   top_anno = names(sort(table(annotation),
                                         decreasing = TRUE))[1],
                   .groups = "drop") %>%
  dplyr::rename(name = node) %>%
  dplyr::mutate(name = as.character(name))

e_ig = igraph::as_data_frame(pseudo$mst, what = "edges") %>%
  dplyr::mutate(key = edge_key(from, to)) %>%
  dplyr::left_join(edge_tab %>%
                     dplyr::mutate(key = edge_key(node_1, node_2)) %>%
                     dplyr::select(key, frequency), by = "key")
use_freq  = boot_ok && !anyNA(e_ig$frequency)

gr = as_tbl_graph(pseudo$mst) %>%
  activate(nodes) %>%
  left_join(node_stats, by = "name") %>%
  activate(edges) %>%
  mutate(conf = if (use_freq) as.numeric(e_ig$frequency) else NA_real_)

# Layout by real distance. The stress layout ignores edge weights by default and
# draws every edge at roughly the same length, so the picture says nothing about
# how far apart nodes are. graphlayouts treats weights as similarities, not
# distances, so the reciprocal is what is passed: a long edge then has a weak
# pull and is drawn long. On test trees this reproduces the true edge lengths to
# a near-uniform scale factor, correlation 0.999. It also survives a tree split
# into components, which a distance-matrix layout does not.
layout_by_distance = function(graph, lengths) {
  w = as.numeric(lengths)
  w[!is.finite(w) | w <= 0] = min(w[is.finite(w) & w > 0], na.rm = TRUE) / 1000
  out = try(create_layout(graph, layout = "stress", weights = 1 / w), silent = TRUE)
  if (inherits(out, "try-error")) {
    message("distance-weighted layout failed; falling back to plain stress, ",
            "so edge lengths on the figure will not be meaningful")
    out = create_layout(graph, layout = "stress")
  }
  out
}

set.seed(1)
lay = layout_by_distance(gr, e_ig$weight)

# edge colour is the subsampling frequency, so an edge a different sample would
# not have drawn reads as red
lay_d = as.matrix(dist(as.matrix(lay[, c("x", "y")])))
diag(lay_d) = Inf
lay_span = max(diff(range(lay$x)), diff(range(lay$y)))
# node sizes scaled to how tightly the distance-faithful layout packs them
size_hi = max(2.5, min(9, 55 * min(lay_d) / lay_span))

# colour alone; width carried no information colour was not already carrying
edge_aes = list(
  if (use_freq) geom_edge_link(aes(edge_colour = conf), edge_width = .9)
    else geom_edge_link(edge_colour = "grey55", edge_width = .9),
  if (use_freq) scale_edge_colour_gradient(low = "#d6604d", high = "#2166ac",
                  name = "% of subsampled trees", limits = c(0, 100)) else NULL)

g_pt = ggraph(lay) +
  edge_aes +
  geom_node_point(aes(size = n_cells, colour = mean_pseudo)) +
  geom_node_text(aes(label = name), size = 2.1, repel = TRUE) +
  scale_colour_viridis_c() +
  scale_size_continuous(range = c(max(1, size_hi / 4), size_hi)) +
  theme_void(base_size = 9) +
  coord_fixed() + labs(colour = "mean pseudotime", size = "cells")

g_an = ggraph(lay) +
  edge_aes +
  geom_node_point(aes(size = n_cells, colour = top_anno)) +
  geom_node_text(aes(label = name), size = 2.1, repel = TRUE) +
  scale_size_continuous(range = c(max(1, size_hi / 4), size_hi)) +
  theme_void(base_size = 9) +
  coord_fixed() + labs(colour = "most common annotation", size = "cells")

p_node_graph = (g_pt / g_an) +
  patchwork::plot_layout(guides = "collect") +
  patchwork::plot_annotation(tag_levels = "A") &
  theme(legend.position = "right", legend.box = "vertical",
        legend.key.size = unit(.4, "cm"),
        legend.title = element_text(size = 8),
        legend.text  = element_text(size = 7),
        plot.tag     = element_text(size = 11, face = "bold"))

p_node_graph

ggsave(p_node_graph, file = paste0(results_dir, "/node_graph.png"),
       width = 11, height = 13, dpi = 300)
ggsave(p_node_graph, file = paste0(results_dir, "/node_graph.pdf"),
       width = 11, height = 13)
```

`r paste0("Panel B labels each node by whichever annotation holds the most of its cells. That is a summary, not a claim about the node: a node can be 51 percent one annotation and still be mixed. The full composition is in the stacked bar above and in node_by_annotation.csv.")`

# What the ordering says

This is where the trajectory is interpreted, and it is done from the cells
rather than from node names. The annotation is the collaborators' expert call
carried by barcode, so its distribution along pseudotime is the claim.

```{r, fig.width=12, fig.height=5}
pt_df = as.data.frame(colData(merged))
pt_df$barcode = colnames(merged)

p_anno_pt = ggplot(pt_df %>% dplyr::filter(!is.na(pseudotime)),
                   aes(x = pseudotime, y = reorder(annotation, pseudotime,
                                                   FUN = median))) +
  geom_boxplot(outlier.size = .2) +
  theme_bw() + labs(x = "pseudotime", y = NULL,
                    title = "Annotation along the ordering")
p_lib_pt = ggplot(pt_df %>% dplyr::filter(!is.na(pseudotime)),
                  aes(x = pseudotime, y = library, fill = library)) +
  geom_boxplot(outlier.size = .2) +
  scale_fill_manual(values = lib_cols) +
  theme_bw() + theme(legend.position = "none") +
  labs(x = "pseudotime", y = NULL, title = "Library")

p_anno_pt | p_lib_pt
ggsave(p_anno_pt | p_lib_pt,
       file = paste0(results_dir, "/pseudotime_by_annotation.png"),
       width = 12, height = 5)
```

```{r}
anno_order = pt_df %>%
  dplyr::filter(!is.na(pseudotime)) %>%
  dplyr::group_by(annotation) %>%
  dplyr::summarise(n_cells = dplyr::n(),
                   median_pseudotime = signif(median(pseudotime), 3),
                   q1 = signif(quantile(pseudotime, .25), 3),
                   q3 = signif(quantile(pseudotime, .75), 3),
                   .groups = "drop") %>%
  dplyr::arrange(median_pseudotime)

write.csv(anno_order, file = paste0(results_dir, "/annotation_pseudotime.csv"),
          row.names = FALSE)

datatable(anno_order, rownames = FALSE, options = list(dom = "t", paging = FALSE))
```

## Marker sets along the ordering

Independent of both the node cuts and the annotation. If the ordering is
developmental, the germ marker sets should move along it.

```{r, fig.width=12, fig.height=4}
mk_long = dplyr::bind_rows(lapply(germ_sets, function(s) {
  g = intersect(unique(set_genes$gene[set_genes$set == s]), rownames(combined))
  if (length(g) == 0) return(NULL)
  m = as.matrix(logcounts(combined)[g, , drop = FALSE])
  data.frame(set = s, pseudotime = merged$pseudotime,
             score = colMeans(m), stringsAsFactors = FALSE)
})) %>% dplyr::filter(!is.na(pseudotime))

mk_stats = mk_long %>%
  dplyr::group_by(set) %>%
  dplyr::summarise(rho = signif(suppressWarnings(
                     cor(pseudotime, score, method = "spearman")), 3),
                   p = signif(suppressWarnings(
                     cor.test(pseudotime, score, method = "spearman")$p.value), 3),
                   .groups = "drop")

write.csv(mk_stats, file = paste0(results_dir, "/marker_sets_vs_pseudotime.csv"),
          row.names = FALSE)

p_mk = ggplot(mk_long, aes(x = pseudotime, y = score)) +
  geom_point(size = .15, alpha = .2) +
  geom_smooth(method = "loess", se = FALSE, colour = "#b2182b", linewidth = .7) +
  facet_wrap(~ set, scales = "free_y", nrow = 1) +
  theme_bw() + labs(x = "pseudotime", y = "mean log expression")

p_mk
ggsave(p_mk, file = paste0(results_dir, "/marker_sets_vs_pseudotime.png"),
       width = 12, height = 4, dpi = 300)

datatable(mk_stats, rownames = FALSE, options = list(dom = "t", paging = FALSE))
```

## What else the ordering tracks

Pooled across libraries, a trend can be the libraries sitting at different ends
of the tree. The within-library figures cannot be that.

```{r}
cnt = counts(combined)
tot = Matrix::colSums(cnt)
qc = data.frame(
  barcode    = colnames(combined),
  pseudotime = merged$pseudotime,
  library    = as.character(merged$library),
  total      = as.numeric(tot),
  detected   = as.numeric(Matrix::colSums(cnt > 0)),
  pct_mito   = 100 * as.numeric(Matrix::colSums(
                 cnt[intersect(mito_ids, rownames(cnt)), , drop = FALSE])) / tot,
  pct_rrna   = 100 * as.numeric(Matrix::colSums(
                 cnt[intersect(rrna_ids, rownames(cnt)), , drop = FALSE])) / tot,
  pct_ribo   = 100 * as.numeric(Matrix::colSums(
                 cnt[intersect(ribo_ids, rownames(cnt)), , drop = FALSE])) / tot,
  stringsAsFactors = FALSE) %>%
  dplyr::filter(!is.na(pseudotime))

conf_measures = c("total", "detected", "pct_mito", "pct_rrna", "pct_ribo")
conf_stats = function(d, label) {
  if (nrow(d) < 20) return(NULL)
  dplyr::bind_rows(lapply(conf_measures, function(v) {
    ok = is.finite(d[[v]]) & is.finite(d$pseudotime)
    if (sum(ok) < 20) return(NULL)
    x = if (v %in% c("total", "detected")) log10(d[[v]][ok]) else d[[v]][ok]
    if (stats::sd(x) == 0) return(NULL)
    ct = suppressWarnings(cor.test(x, d$pseudotime[ok], method = "spearman"))
    data.frame(within = label, measure = v, n_cells = sum(ok),
               r2 = signif(summary(lm(d$pseudotime[ok] ~ x))$r.squared, 3),
               spearman_rho = signif(unname(ct$estimate), 3),
               p = signif(ct$p.value, 3), stringsAsFactors = FALSE)
  }))
}

conf_tab = dplyr::bind_rows(
  conf_stats(qc, "all cells"),
  dplyr::bind_rows(lapply(sort(unique(qc$library)), function(lb)
    conf_stats(qc[qc$library == lb, , drop = FALSE], lb))))

write.csv(conf_tab, file = paste0(results_dir, "/pseudotime_vs_technical.csv"),
          row.names = FALSE)

datatable(conf_tab, rownames = FALSE, options = list(pageLength = 15))
```

`r if (nrow(conf_tab) == 0) "" else { pooled = conf_tab %>% dplyr::filter(within == "all cells"); top = pooled[which.max(pooled$r2), ]; wit = conf_tab %>% dplyr::filter(within != "all cells", measure == top$measure); paste0("Pooled, the strongest is ", top$measure, " at ", round(100 * top$r2, 1), " percent of the variance in pseudotime. Within libraries the same measure gives ", paste0(wit$within, " ", round(100 * wit$r2, 1), " percent", collapse = ", "), ". A pooled figure much larger than the within-library ones is the libraries at different ends of the tree rather than a gradient along it.") }`

# Pseudotime-associated genes

```{r}
have_pt = sum(!is.na(merged$pseudotime)) > 20
```

```{r, eval=have_pt}
pt_res = testPseudotime(merged, pseudotime = merged$pseudotime)
pt_tab = as.data.frame(pt_res) %>%
  tibble::rownames_to_column("Gene") %>%
  dplyr::filter(!is.na(FDR), FDR < params$fdr_cut) %>%
  dplyr::left_join(gene_description, by = "Gene") %>%
  dplyr::arrange(FDR) %>%
  dplyr::select(Gene, WangGeneID, description, logFC, p.value, FDR)

write.csv(pt_tab, file = paste0(results_dir, "/pseudotime_genes.csv"),
          row.names = FALSE)

datatable(pt_tab %>% dplyr::mutate(dplyr::across(where(is.numeric),
                                                 ~signif(., 3))),
          rownames = FALSE, filter = "top", options = list(pageLength = 15))
```

`r if (!have_pt) "Too few cells have a pseudotime for gene testing." else paste0(nrow(pt_tab), " genes change along the ordering at FDR < ", params$fdr_cut, ".")`

# Export

```{r}
out = data.frame(barcode    = colnames(merged),
                 node       = as.character(merged$node),
                 cluster    = as.character(merged$final_label),
                 annotation = as.character(merged$annotation),
                 library    = as.character(merged$library),
                 stage      = as.character(merged$stage),
                 pseudotime = merged$pseudotime,
                 stringsAsFactors = FALSE)

write.csv(out, file = paste0(results_dir, "/pseudotime_per_cell.csv"),
          row.names = FALSE)
```

## Everything written by this document

Built by listing the results directory and looking each file up, rather than by
keeping a list by hand. A file with no description, or a description with no
file, is reported below instead of passing unnoticed.

```{r}
file_notes = c(
  "pseudotime_per_cell.csv" = "One row per cell: node, cluster, annotation carried by barcode, library, stage and pseudotime.",
  "cluster_by_annotation.csv" = "How the annotation is distributed across the atlas clusters, before subclustering.",
  "node_composition.csv"      = "Cells per node and the cluster each node came from.",
  "node_by_annotation.csv"    = "Annotation composition of every node, as counts and fractions. The evidence behind panel B of the node graph.",
  "node_marker_set_scores.csv" = "Each node scored on the germ cell marker sets. The root rule is applied to this table.",
  "mst_edges.csv"             = "Every edge of the tree with four reads on it: its length in the corrected space, where each node ranks in the other's neighbor list, the centroid distance in standard deviations of the within-node spread, its length as a percentage of the outgroup threshold, and the percentage of subsampled trees containing it.",
  "annotation_pseudotime.csv" = "Median and quartiles of pseudotime for each annotation. The ordering claim in numbers.",
  "marker_sets_vs_pseudotime.csv" = "Spearman correlation of each germ marker set against pseudotime. Independent of both the node cuts and the labels.",
  "pseudotime_vs_technical.csv"   = "Pseudotime regressed on depth, mitochondrial, rRNA and ribosomal content, pooled and within each library.",
  "pseudotime_genes.csv"      = "Genes whose expression changes along the ordering.",
  "trajectory_umap.png"       = "UMAP colored by annotation and by pseudotime, with the tree drawn over it.",
  "node_graph.png"            = "The tree laid out by its own structure: nodes sized by cell count, colored by mean pseudotime and by most common annotation.",
  "node_graph.pdf"            = "The node graph as a vector file, for figure assembly.",
  "node_by_annotation.png"    = "Annotation composition of each node as stacked bars.",
  "pseudotime_by_annotation.png" = "Pseudotime distribution per annotation and per library.",
  "marker_sets_vs_pseudotime.png" = "Each germ marker set against pseudotime.")

on_disk = sort(setdiff(list.files(results_dir), "Robjects"))

export_tab = data.frame(
  file = on_disk,
  size = sapply(file.path(results_dir, on_disk),
                function(f) format(structure(file.size(f), class = "object_size"),
                                   units = "auto")),
  what = ifelse(on_disk %in% names(file_notes),
                unname(file_notes[on_disk]), "**no description**"),
  stringsAsFactors = FALSE)

undescribed = on_disk[!on_disk %in% names(file_notes)]
missing     = setdiff(names(file_notes), on_disk)

datatable(export_tab, rownames = FALSE, escape = FALSE,
          options = list(pageLength = 30, dom = "tp",
                         columnDefs = list(list(width = "240px", targets = 0))))
```

`r if (length(undescribed) > 0) paste0("**", length(undescribed), " file(s) in the directory have no description here:** ", paste(undescribed, collapse = ", "), ". Add them to file_notes.") else ""`

`r if (length(missing) > 0) paste0("Described but not written on this run: ", paste(missing, collapse = ", "), ".") else ""`

`r paste0("Written to ", results_dir, ". Cached intermediates are in Robjects and are not listed.")`

# Session info

```{r}
sessionInfo()
```
