---
title: "19_germ_pseudotime_fine_nodes"
author: "Stephanie Hickey"
date: "`r Sys.Date()`"
output:
  html_document:
    code_folding: hide
    toc: true
    toc_float:
      toc_collapsed: true
params:
    project_dir: /mnt/ufs18/rs-013/bioinformaticsCore/projects/chong_davidson/BCC125_sea_lamprey_scRNA
    project_name: BCC125_sea_lamprey_scRNA
    lab: Weiming Li
    results_date: "20260813"
    atlas_suffix: ""
    run_suffix: ""
    lineage: "all"
    female_clusters: "5"
    male_clusters: "15,16,17,18"
    all_clusters: "5,15,16,17,18"
    sub_k: 10
    min_node_cells: 30
    dist_method: "simple"
    edge_boot: 100
    edge_boot_cells: 100
    mnn_k: 50
    outgroup: FALSE
    root_rule: "max:Primordial germ cells"
    graph_marker_sets: "Primordial germ cells,Migrating germ cells,Spermatocytes,Oocyte"
    root_node: ""
    min_detect_pct: 5
    marker_file: canonical_markers_S1_only_2026-08-14.csv
    extra_marker_file: "neural_crest_markers_2026-09-25.csv,immediate_early_markers_2026-10-06.csv"
    lfc_thresh: 1
    prev_nodes_file: ""
    removal_evidence_sets: "Dissociation response"
    drop_nodes: ""
    node_annotation: "5.1=Migrating neural crest cells;5.2=Oogonia;5.3=Oocytes;5.4=Migrating female germ cells (FGC I);5.5=Granulosa;5.6=Migrating female germ cells (FGC II);5.7=Primordial germ cells (PGCs);5.8=Migrating PGCs;15=Spermatogonia;16=Spermatocytes I;17=Spermatocytes II;18=Spermatocytes III"
    min_library_cells: 50
    fdr_cut: 0.01
    go_pval: 0.05
    genome_dir: /mnt/research/bioinformaticsCore/shared/Genomes/Petromyzon_marinus
---

```{r setup, include=FALSE}
# html_document renders inline figures at knitr's 72 dpi doubled by
# fig.retina, so 144, while every ggsave here writes at 300. Rendering inline
# at 300 as well puts the figure in the document at the resolution of the file
# on disk. fig.retina is set to 1 so the two are not multiplied.
# Figures are vector by default. A raster figure in html_document is rendered
# at a fixed pixel size and then scaled to the page, so it is only ever as sharp
# as that size; svg has no resolution to run out of. The chunks that draw one
# mark per cell set dev = "png" individually, because an svg of forty thousand
# tiles is tens of megabytes where the png is a couple of hundred kilobytes.
# dpi applies to those, and fig.retina is 1 so it is not doubled again.
# svglite if it is installed, otherwise the grDevices svg device, which needs
# the cairo capability enabled below
vec_dev = if (requireNamespace("svglite", quietly = TRUE)) "svglite" else "svg"
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      dev = vec_dev, dpi = 300, fig.retina = 1)

# Compute nodes have no X11, and the default png device asks for it. Cairo
# renders off screen.
if (isTRUE(capabilities("cairo"))) options(bitmapType = "cairo")
```

<style>
/* html_document leaves pandoc's markdown tables unstyled: no padding, no row
   rules, cells aligned on their last line. Selected as direct children of a
   section, which a DT widget's table never is, so the interactive tables keep
   their own styling. */
div.section > table, div.main-container > table {
  border-collapse: collapse;
  margin: 1.1em 0;
  font-size: 0.93em;
  line-height: 1.35;
}
div.section > table th, div.main-container > table th,
div.section > table td, div.main-container > table td {
  vertical-align: top;
  text-align: left;
  padding: 7px 16px 7px 0;
  border-top: 1px solid #e3e3e3;
}
div.section > table th, div.main-container > table th {
  border-top: none;
  border-bottom: 2px solid #b0b0b0;
  font-weight: 600;
}
div.section > table code, div.main-container > table code {
  white-space: nowrap;
  background: none;
  padding: 0;
  color: #333;
}
</style>

```{r, echo=FALSE, results="asis"}
cat("\n")
cat("#### Lab:", params$lab, " \n")
cat("#### Project:", params$project_name, " \n")
cat("#### Lineage:", params$lineage, " \n")
```

```{r}
library(tidyverse)
library(SingleCellExperiment)
library(scater)
library(scran)
library(batchelor)
library(BiocSingular)
library(TSCAN)
library(patchwork)
library(DT)
library(ggraph)
library(tidygraph)
library(ggrepel)
source(paste0(params$project_dir, "/src/scRNA_seq_functions.R"))
```

TSCAN pseudotime for the `r params$lineage` germ cells, with **every** cluster
subclustered finely and no cell removed.

The tree TSCAN fits is a minimum spanning tree over one centroid per node, so
the number of nodes is the number of points the trajectory is fitted to. `17`
gives it around a dozen, of which four are whole clusters holding thousands of
cells each. A tree on twelve points is coarse and moves whenever a node moves.

This document treats nodes as waypoints rather than cell types. Every cluster is
cut into small subclusters at a fixed `sub_k`, which gives the tree many more
points to follow. Nothing here needs a node to be interpretable, so nothing here
depends on the subclustering being stable or on any node carrying a label.

The collaborators' annotation is carried **per cell**, by barcode, from the run
named in `prev_nodes_file`. It is never attached to a node. Re-clustering
therefore cannot break it, and the interpretation is read off the composition of
the trajectory rather than off node names.

```{r}
lineage = as.character(params$lineage)
if (!lineage %in% c("female", "male", "all"))
  stop("lineage must be 'female', 'male' or 'all', not '", lineage, "'")

clusters = trimws(strsplit(as.character(switch(lineage,
  female = params$female_clusters,
  male   = params$male_clusters,
  all    = params$all_clusters)), ",")[[1]])
clusters = clusters[nzchar(clusters)]

object_dir      = paste0(params$project_dir, "/results/", params$results_date,
                         "_no_doublets")
atlas_dir       = paste0(params$project_dir, "/results/", params$results_date,
                         "_atlas", params$atlas_suffix)
base_dir        = paste0(params$project_dir, "/results/", params$results_date,
                         "_pseudotime_fine_", lineage, params$atlas_suffix)
# run_suffix keeps the outputs of two settings side by side, for example a
# run with outgroup = TRUE against one without.
results_dir     = paste0(base_dir, params$run_suffix)
# the cache stays on the unsuffixed path: the corrected space does not depend
# on anything run_suffix varies, so it is built once and shared
Robject_out_dir = paste0(base_dir, "/Robjects")
if (!dir.exists(results_dir)) dir.create(results_dir, recursive = TRUE)
if (!dir.exists(Robject_out_dir)) dir.create(Robject_out_dir, recursive = TRUE)

label_file = paste0(atlas_dir, "/cell_cluster_labels.csv")
if (!file.exists(label_file))
  stop("missing ", label_file, ". Run 13 first.")

cell_labels      = read.csv(label_file, stringsAsFactors = FALSE)
all.sce          = readRDS(paste0(object_dir, "/all.sce.Rds"))
gene_description = read.csv(paste0(object_dir, "/gene_descritption.csv"),
                            stringsAsFactors = FALSE) %>%
  dplyr::distinct(Gene, .keep_all = TRUE)

lib_cols = c("Female" = "#FF6DB6", "Male" = "#006ddb",
             "Undifferentiated.1" = "#009292", "Undifferentiated.2" = "#b66ddf")

```

```{r}
# Read here rather than at the point of use, because the checks on the annotated
# run's nodes happen before the corrected space exists.
tech       = technical_gene_ids(gene_description)
mito_ids   = tech$mito
rrna_ids   = tech$rrna
ribo_ids   = tech$ribo
tech_genes = tech$all

# The pipeline computed subsets_Mito_percent before the cells were filtered on
# it, so that is the number the filtering was applied to and the one to report.
# Recomputing it here from the annotation would be a second, unvalidated
# definition of the same quantity.
mito_src = try(dplyr::bind_rows(lapply(all.sce, function(x) {
  cd = as.data.frame(colData(x))
  if (!"subsets_Mito_percent" %in% colnames(cd)) return(NULL)
  data.frame(barcode = colnames(x), pct_mito = cd$subsets_Mito_percent,
             stringsAsFactors = FALSE)
})), silent = TRUE)

have_mito_src = !inherits(mito_src, "try-error") && is.data.frame(mito_src) &&
                nrow(mito_src) > 0
```

`r paste0(format(length(mito_ids), big.mark = ","), " mitochondrially encoded genes, ", length(rrna_ids), " cytoplasmic rRNA genes and ", format(length(ribo_ids), big.mark = ","), " ribosomal proteins are identified from the annotation. ", if (have_mito_src) "Percent mitochondrial is taken from subsets_Mito_percent in the per-library objects, which is the quantity the cells were filtered on upstream." else "**subsets_Mito_percent was not found in the per-library objects**, so percent mitochondrial is recomputed here from the annotation and may not match the filter applied upstream.", " Cytoplasmic 28S and 18S rRNA are counted separately, because they are not mitochondrial and are abundant enough to distort that fraction if folded into it.")`

```{r}
# Every set, not only the germ ones: a node scoring on a somatic set is the
# thing the fine cuts are most likely to turn up.
s1_sets = read.csv(paste0(params$project_dir, "/data/", params$marker_file),
                   stringsAsFactors = FALSE) %>%
  dplyr::mutate(loc_id = trimws(loc_id), origin = "S1 sheet")

# extra_marker_file takes a comma-separated list, so sets that did not come
# from the S1 sheet can be added independently of one another.
extra_file = trimws(strsplit(as.character(params$extra_marker_file), ",")[[1]])
extra_file = extra_file[nzchar(extra_file)]
extra_path = paste0(params$project_dir, "/data/", extra_file)
extra_ok   = file.exists(extra_path)
have_extra = any(extra_ok)
extra_missing = extra_file[!extra_ok]

all_sets = if (have_extra) {
  dplyr::bind_rows(s1_sets,
                   dplyr::bind_rows(lapply(extra_path[extra_ok], function(p)
                     read.csv(p, stringsAsFactors = FALSE) %>%
                       dplyr::mutate(loc_id = trimws(loc_id), origin = "extra"))))
} else s1_sets

set_origin = all_sets %>%
  dplyr::distinct(CellTypeMarkerSet, origin) %>%
  dplyr::group_by(CellTypeMarkerSet) %>%
  dplyr::summarise(origin = paste(sort(unique(origin)), collapse = ", "),
                   .groups = "drop")

# A set whose genes all sit inside another set is not a separate measurement.
# Scoring high on it says nothing about it that the larger set does not already
# say. The one-gene sets are where this bites.
set_genes = lapply(split(all_sets$loc_id, all_sets$CellTypeMarkerSet), unique)
set_containment = dplyr::bind_rows(lapply(names(set_genes), function(a)
  dplyr::bind_rows(lapply(setdiff(names(set_genes), a), function(b)
    if (length(set_genes[[a]]) > 0 && all(set_genes[[a]] %in% set_genes[[b]]))
      data.frame(set = a, contained_in = b, n_genes = length(set_genes[[a]]),
                 genes = paste(sort(set_genes[[a]]), collapse = ", "),
                 stringsAsFactors = FALSE)
    else NULL))))

# carried onto every set-score table, so a contained set is flagged where it is
# read rather than in a note further up the page
set_flag = if (nrow(set_containment) > 0) {
  set_containment %>%
    dplyr::group_by(set) %>%
    dplyr::summarise(contained_in = paste(sort(contained_in), collapse = ", "),
                     .groups = "drop")
} else data.frame(set = character(0), contained_in = character(0),
                  stringsAsFactors = FALSE)

# one row per marker gene per group, for the bubble plots. A gene belonging to
# two sets appears once per set, so the rows line up with sets_df positionally.
marker_bubble_df = function(lc, groups, sets_df) {
  present = sets_df %>%
    dplyr::distinct(CellTypeMarkerSet, loc_id, symbol) %>%
    dplyr::filter(loc_id %in% rownames(lc))
  groups  = as.character(groups)
  dplyr::bind_rows(lapply(unique(groups), function(gr) {
    m = as.matrix(lc[present$loc_id, groups == gr, drop = FALSE])
    data.frame(group          = gr,
               set            = present$CellTypeMarkerSet,
               gene           = present$loc_id,
               symbol         = present$symbol,
               n_cells        = sum(groups == gr),
               pct_expressing = 100 * rowMeans(m > 0),
               mean_logexpr   = rowMeans(m),
               stringsAsFactors = FALSE)
  }))
}

marker_bubble_plot = function(df, group_levels, title, subtitle = NULL) {
  df = df %>%
    dplyr::mutate(group = factor(group, levels = group_levels),
                  label = ifelse(is.na(symbol) | !nzchar(symbol), gene, symbol))
  ggplot(df, aes(x = group, y = label,
                 size = pct_expressing, colour = mean_logexpr)) +
    geom_point() +
    scale_colour_viridis_c() +
    scale_size_continuous(range = c(0.4, 5)) +
    facet_grid(set ~ ., scales = "free_y", space = "free_y") +
    theme_bw() +
    theme(strip.text.y     = element_text(angle = 0, size = 6),
          axis.text.y      = element_text(size = 6),
          axis.text.x      = element_text(size = 8, angle = 45, hjust = 1),
          panel.grid.major = element_line(linewidth = .2)) +
    labs(x = NULL, y = NULL, title = title, subtitle = subtitle,
         size = "% expressing", colour = "mean log expression")
}
```

```{r}
gaf_file      = file.path(params$genome_dir,
                          "GCF_010993605.1_kPetMar1.pri_gene_ontology.gaf")
go_names_file = file.path(params$genome_dir, "GOs_names.txt")
have_go       = file.exists(gaf_file) && file.exists(go_names_file)
```

```{r, eval=have_go}
gaf = read.delim(gaf_file, header = FALSE, comment = "!") %>%
  dplyr::filter(!grepl("NOT", V4)) %>%
  dplyr::select(V3, V5) %>%
  dplyr::rename(GOID = "V5", Gene = "V3") %>%
  dplyr::distinct()

# terms too small are noise, terms too large are uninformative
keep_go = gaf %>%
  dplyr::count(GOID) %>%
  dplyr::filter(n > 10, n < 200) %>%
  dplyr::pull(GOID)
gaf = gaf %>% dplyr::filter(GOID %in% keep_go)

go_names = read.delim(go_names_file, header = FALSE)
colnames(go_names) = c("GOID", "GOName")
```

`r if (!have_go) paste0("**GO enrichment is skipped throughout.** The GO files were not found under ", params$genome_dir, ".") else ""`

# Cells

```{r}
absent = setdiff(clusters, cell_labels$final_label)
if (length(absent) > 0)
  stop("cluster(s) ", paste(absent, collapse = ", "), " are not in ", label_file,
       ". Labels present: ",
       paste(sort(unique(cell_labels$final_label)), collapse = ", "))

lineage_key = cell_labels %>% dplyr::filter(final_label %in% clusters)

# the annotation, per cell, by barcode
parse_pairs = function(x) {
  x = trimws(x)
  if (!nzchar(x)) return(character(0))
  kv = strsplit(trimws(strsplit(x, ";")[[1]]), "=")
  setNames(vapply(kv, function(z) trimws(z[2]), character(1)),
           vapply(kv, function(z) trimws(z[1]), character(1)))
}

anno_map  = parse_pairs(params$node_annotation)
prev_file = as.character(params$prev_nodes_file)
have_prev = nzchar(prev_file) && file.exists(prev_file)

cell_anno = rep(NA_character_, nrow(lineage_key))
if (have_prev) {
  prev = read.csv(prev_file, stringsAsFactors = FALSE)
  stopifnot(all(c("barcode", "node") %in% colnames(prev)))
  old = prev$node[match(lineage_key$barcode, prev$barcode)]
  cell_anno = unname(anno_map[as.character(old)])
  cell_anno[is.na(cell_anno) & !is.na(old)] = as.character(old)[is.na(cell_anno) & !is.na(old)]
}
lineage_key$annotation = ifelse(is.na(cell_anno), "unannotated", cell_anno)


comp_tab = lineage_key %>%
  dplyr::count(final_label, annotation, name = "n") %>%
  tidyr::pivot_wider(names_from = annotation, values_from = n, values_fill = 0)

write.csv(comp_tab, file = paste0(results_dir, "/cluster_by_annotation.csv"),
          row.names = FALSE)

datatable(comp_tab, rownames = FALSE, options = list(dom = "t", scrollX = TRUE))
```

`r paste0(format(nrow(lineage_key), big.mark = ","), " cells from cluster(s) ", paste(clusters, collapse = ", "), ". Nothing is removed here: the subclustering runs on every cell, and whatever is removed is removed afterwards, on the strength of what the nodes turn out to express.", if (!have_prev) " **prev_nodes_file is not set**, so no annotation is carried and the trajectory cannot be read against it." else paste0(" ", format(sum(lineage_key$annotation != "unannotated"), big.mark = ","), " carry an annotation from ", basename(prev_file), "."))`

```{r}
obj_file     = paste0(Robject_out_dir, "/lineage_objects.Rds")
obj_sig_file = paste0(Robject_out_dir, "/lineage_objects_signature.txt")
obj_sig      = paste(lineage, paste(sort(clusters), collapse = ","),
                     nrow(lineage_key), params$min_library_cells, sep = "|")

if (file.exists(obj_file) && file.exists(obj_sig_file) &&
    identical(readLines(obj_sig_file, warn = FALSE)[1], obj_sig)) {
  objs = readRDS(obj_file)
} else {
  objs = build_lineage_sce(all.sce, lineage_key$barcode,
                           min_library_cells = params$min_library_cells)
  saveRDS(objs, file = obj_file)
  writeLines(obj_sig, obj_sig_file)
}

merged   = objs$corrected
combined = objs$combined

key_idx = match(colnames(merged), lineage_key$barcode)
stopifnot(!anyNA(key_idx))
merged$library     = lineage_key$library[key_idx]
merged$stage       = lineage_key$stage[key_idx]
merged$final_label = lineage_key$final_label[key_idx]
merged$annotation  = lineage_key$annotation[key_idx]

lib_tab = data.frame(batch   = names(objs$n_by_library),
                     n_cells = as.integer(objs$n_by_library),
                     kept    = names(objs$n_by_library) %in% objs$kept_batches,
                     stringsAsFactors = FALSE) %>%
  dplyr::left_join(dplyr::distinct(cell_labels, batch, library), by = "batch")

datatable(lib_tab, rownames = FALSE, options = list(dom = "t"))
```

`r paste0(format(ncol(merged), big.mark = ","), " cells from ", length(objs$kept_batches), " librar", if (length(objs$kept_batches) == 1) "y" else "ies", " are kept.")`


# Fine nodes

Every cluster is subclustered in the shared corrected space, using only that
cluster's cells, at a fixed `sub_k`. A smaller `sub_k` gives more and smaller
subclusters. There is no stability sweep, because nothing downstream asks a node
to be a reproducible cell type.

Subclusters below `min_node_cells` are merged into the nearest subcluster of the
same parent cluster, by centroid distance. A node of five cells gives a centroid
that is mostly noise, and the tree is fitted on centroids.

```{r}
set.seed(1)
sub_of = rep(NA_character_, ncol(merged))
names(sub_of) = colnames(merged)

for (cl in clusters) {
  cl_cells = colnames(merged)[merged$final_label == cl]
  if (length(cl_cells) < 2 * params$min_node_cells) {
    sub_of[cl_cells] = paste0(cl, ".1")
    next
  }
  set.seed(1)
  snn = buildSNNGraph(merged[, cl_cells], k = params$sub_k,
                      use.dimred = "corrected")
  mem = igraph::cluster_louvain(snn)$membership
  names(mem) = cl_cells

  # merge anything too small into the nearest surviving subcluster
  co = reducedDim(merged, "corrected")[cl_cells, , drop = FALSE]
  repeat {
    tab   = table(mem)
    small = names(tab)[tab < params$min_node_cells]
    if (length(small) == 0 || length(tab) <= 1) break
    cent = do.call(rbind, lapply(names(tab), function(g)
      colMeans(co[mem == g, , drop = FALSE])))
    rownames(cent) = names(tab)
    d = as.matrix(dist(cent))
    diag(d) = Inf
    s = small[which.min(tab[small])]
    mem[mem == s] = names(tab)[which.min(d[s, ])]
  }
  sub_of[cl_cells] = paste0(cl, ".", match(mem, sort(unique(mem))))
}

stopifnot(!anyNA(sub_of))

node_levels = unique(sub_of)
node_levels = node_levels[order(match(sub(
  "\\..*$", "", node_levels), clusters),
  as.integer(sub("^.*\\.", "", node_levels)))]
merged$node   = factor(sub_of, levels = node_levels)
combined$node = merged$node

node_tab = as.data.frame(colData(merged)) %>%
  dplyr::count(node, final_label, name = "n_cells") %>%
  dplyr::arrange(node)

write.csv(node_tab, file = paste0(results_dir, "/node_composition.csv"),
          row.names = FALSE)

datatable(node_tab, rownames = FALSE, options = list(pageLength = 15))
```

`r paste0(length(node_levels), " nodes from ", length(clusters), " clusters, at sub_k = ", params$sub_k, ", with a floor of ", params$min_node_cells, " cells. Median node size ", round(median(as.integer(table(merged$node)))), " cells, smallest ", min(as.integer(table(merged$node))), ", largest ", max(as.integer(table(merged$node))), ". ", if (length(node_levels) < 2) "**Fewer than two nodes, so no tree can be built.**" else "")`

```{r}
per_cluster = node_tab %>%
  dplyr::count(final_label, name = "n_nodes") %>%
  dplyr::arrange(match(final_label, clusters))
datatable(per_cluster, rownames = FALSE, options = list(dom = "t", paging = FALSE))
```

## What the nodes are made of

Nodes are not cell types, so the question is not what each one is but how the
annotation is distributed over them. A node drawn from one annotation is a
waypoint inside that population. A node mixing several is a waypoint on a
boundary, which is expected wherever the biology is continuous.

```{r, fig.width=12, fig.height=5}
anno_by_node = as.data.frame(colData(merged)) %>%
  dplyr::count(node, annotation, name = "n") %>%
  dplyr::group_by(node) %>%
  dplyr::mutate(frac = n / sum(n)) %>%
  dplyr::ungroup()

write.csv(anno_by_node, file = paste0(results_dir, "/node_by_annotation.csv"),
          row.names = FALSE)

p_anno = ggplot(anno_by_node, aes(x = node, y = frac, fill = annotation)) +
  geom_col() +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = .5, size = 7),
        legend.text = element_text(size = 7)) +
  labs(x = NULL, y = "fraction of the node", fill = NULL,
       title = "Annotation composition of each node")

p_anno
ggsave(p_anno, file = paste0(results_dir, "/node_by_annotation.png"),
       width = 12, height = 5, dpi = 300)
```

## Depth and technical content

Nodes that differ by sequencing depth rather than by cell state show a
characteristic pattern. The shallower node's markers are the abundant genes of
the deeper one at lower magnitude, and housekeeping genes climb its marker list.
Cutting every cluster finely makes this more likely, not less, so it is measured
directly here.

```{r}
cnt = counts(combined)
tot = Matrix::colSums(cnt)

# built once, with everything the later sections need: the node for this
# section, the library and pseudotime for What else the ordering tracks
qc = data.frame(
  barcode   = colnames(combined),
  node      = as.character(merged$node),
  library   = as.character(merged$library),
  total     = as.numeric(tot),
  detected  = as.numeric(Matrix::colSums(cnt > 0)),
  pct_rrna  = 100 * as.numeric(Matrix::colSums(
                cnt[intersect(rrna_ids, rownames(cnt)), , drop = FALSE])) / tot,
  pct_ribo  = 100 * as.numeric(Matrix::colSums(
                cnt[intersect(ribo_ids, rownames(cnt)), , drop = FALSE])) / tot,
  stringsAsFactors = FALSE)

qc = if (have_mito_src) {
  qc %>% dplyr::left_join(mito_src, by = "barcode")
} else qc %>% dplyr::mutate(
  pct_mito = 100 * as.numeric(Matrix::colSums(
               cnt[intersect(mito_ids, rownames(cnt)), , drop = FALSE])) / tot)

qc_tab = qc %>%
  dplyr::group_by(node) %>%
  dplyr::summarise(n_cells = dplyr::n(),
                   median_total    = round(median(total)),
                   median_detected = round(median(detected)),
                   median_pct_mito = round(median(pct_mito), 1),
                   median_pct_rrna = round(median(pct_rrna), 1),
                   median_pct_ribo = round(median(pct_ribo), 1),
                   .groups = "drop") %>%
  dplyr::mutate(node = factor(node, levels = node_levels)) %>%
  dplyr::arrange(node)

write.csv(qc_tab, file = paste0(results_dir, "/node_depth_and_technical.csv"),
          row.names = FALSE)

datatable(qc_tab, rownames = FALSE, filter = "top",
          options = list(pageLength = 15, scrollX = TRUE))
```

```{r, fig.width=11, fig.height=11, dev = "png", dpi = 600}
qc_long = qc %>%
  dplyr::mutate(node = factor(node, levels = node_levels)) %>%
  tidyr::pivot_longer(c(total, detected, pct_mito, pct_rrna, pct_ribo),
                      names_to = "measure", values_to = "value") %>%
  dplyr::mutate(measure = factor(measure,
                  levels = c("total", "detected", "pct_mito", "pct_rrna",
                             "pct_ribo"),
                  labels = c("total counts", "genes detected",
                             "percent mitochondrial", "percent 28S/18S rRNA",
                             "percent ribosomal protein")))

p_qc = ggplot(qc_long, aes(x = node, y = value)) +
  geom_violin(scale = "width", fill = "grey90", colour = NA) +
  geom_boxplot(width = .18, outlier.size = .2) +
  facet_wrap(~ measure, scales = "free_y", ncol = 2) +
  scale_y_continuous(trans = "log1p") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 5)) +
  labs(x = NULL, y = NULL, title = "Depth and technical content by node",
       subtitle = "log1p scale")

p_qc
ggsave(p_qc, file = paste0(results_dir, "/node_depth_and_technical.png"),
       width = 11, height = 11, dpi = 300)
```

```{r}
# Does node membership track depth within a cluster? Every cluster is cut here,
# so the test runs on all of them: Kruskal-Wallis across the subclusters of one
# cluster, plus how far apart the extreme nodes sit in median depth.
depth_tests = dplyr::bind_rows(lapply(clusters, function(cl) {
  sub = qc %>% dplyr::filter(grepl(paste0("^", cl, "\\."), node))
  if (dplyr::n_distinct(sub$node) < 2) return(NULL)
  med = sub %>% dplyr::group_by(node) %>%
    dplyr::summarise(m = median(total), .groups = "drop")
  dplyr::bind_rows(lapply(c("total", "detected", "pct_mito", "pct_rrna",
                            "pct_ribo"), function(v) {
    kt = kruskal.test(sub[[v]] ~ factor(sub$node))
    data.frame(cluster = cl, measure = v,
               kruskal_p  = signif(kt$p.value, 3),
               fold_range = signif(max(med$m) / min(med$m), 3),
               stringsAsFactors = FALSE)
  }))
}))

if (!is.null(depth_tests) && nrow(depth_tests) > 0) {
  depth_tests$fold_range[depth_tests$measure != "total"] = NA
  write.csv(depth_tests, file = paste0(results_dir, "/node_depth_tests.csv"),
            row.names = FALSE)
  datatable(depth_tests, rownames = FALSE,
            options = list(pageLength = 15, scrollX = TRUE))
}
```

`r if (is.null(depth_tests) || nrow(depth_tests) == 0) "" else { tt = depth_tests %>% dplyr::filter(measure == "total"); paste0("Median total counts across the subclusters of cluster ", paste(tt$cluster, collapse = ", "), " differ by a factor of ", paste(tt$fold_range, collapse = ", "), " between the deepest and shallowest node (Kruskal-Wallis p ", paste(tt$kruskal_p, collapse = ", "), "). A large factor is the signature of a depth split. It does not prove one, because a real cell state can also carry more RNA.") }`

## Node identity

Every node is scored against **all** the marker sets available, somatic ones
included, rather than only the germ sets the root rule uses. No cell is removed
in this run, so every cell of the lineage is scored here. A node scoring high on
a somatic set and low on every germ set is not a germ cell state.

```{r}
if (nrow(set_containment) > 0)
  datatable(set_containment, rownames = FALSE,
            options = list(dom = "t", paging = FALSE))
```

`r if (nrow(set_containment) > 0) paste0("**", nrow(set_containment), " marker set(s) sit entirely inside another set.** A node scoring high on one of these is also scoring high on the set that contains it. The two cannot be read as separate evidence. Judge these from the per-gene plot rather than the set score.") else "No marker set sits entirely inside another. Each set score is its own measurement."`

```{r}
id_genes = all_sets %>%
  dplyr::transmute(set = CellTypeMarkerSet, gene = loc_id) %>%
  dplyr::filter(gene %in% rownames(combined)) %>%
  dplyr::distinct()

id_scores = score_marker_sets(combined, merged$node, id_genes) %>%
  dplyr::rename(node = group)

id_table = id_scores %>%
  dplyr::left_join(set_origin, by = c("set" = "CellTypeMarkerSet")) %>%
  dplyr::left_join(set_flag, by = "set") %>%
  dplyr::mutate(node = factor(node, levels = node_levels),
                contained_in = tidyr::replace_na(contained_in, ""),
                mean_logexpr = signif(mean_logexpr, 3),
                pct_detected = round(pct_detected, 1)) %>%
  dplyr::arrange(node, dplyr::desc(mean_logexpr)) %>%
  dplyr::select(node, set, origin, contained_in, n_genes, n_cells,
                mean_logexpr, pct_detected)

write.csv(id_table, file = paste0(results_dir, "/node_identity_scores.csv"),
          row.names = FALSE)

datatable(id_table, rownames = FALSE, filter = "top",
          options = list(pageLength = 20, scrollX = TRUE))
```

`r if (length(extra_missing) > 0) paste0("**Not found in data/: ", paste(extra_missing, collapse = ", "), ".** Those sets are not scored.") else ""`

```{r, fig.width=11, fig.height=13}
id_bubble = marker_bubble_df(logcounts(combined), merged$node, all_sets)
write.csv(id_bubble, file = paste0(results_dir, "/node_marker_expression.csv"),
          row.names = FALSE)

p_identity = marker_bubble_plot(
  id_bubble, node_levels,
  title = "Marker expression by node, every set")

p_identity
ggsave(p_identity, file = paste0(results_dir, "/node_identity_scores.png"),
       width = 11, height = 13, dpi = 300)
```

# Removal

Every node above was defined on every cell, with nothing taken out, and scored
against every marker set. A node is removed here only on the strength of what it
turned out to express. Removing on the strength of the label it was given would
make the trajectory unable to say anything about that label.

`removal_evidence_sets` names the sets drawn below. Dissociation response holds
the AP-1 and immediate-early genes that enzymatic digestion induces, so a node
carrying them is defined by how the tissue was handled rather than by the
animal. `drop_nodes` names the nodes to remove, by node, after reading it.

```{r, fig.width=9, fig.height=7}
ev_sets = trimws(strsplit(as.character(params$removal_evidence_sets), ",")[[1]])
ev_sets = intersect(ev_sets[nzchar(ev_sets)], unique(all_sets$CellTypeMarkerSet))

drop_nodes = trimws(strsplit(as.character(params$drop_nodes), ",")[[1]])
drop_nodes = drop_nodes[nzchar(drop_nodes)]
absent = setdiff(drop_nodes, node_levels)
if (length(absent) > 0)
  stop("drop_nodes names node(s) that do not exist: ", paste(absent, collapse = ", "),
       ". Nodes: ", paste(node_levels, collapse = ", "))

# the nodes being removed first on the axis, so the comparison is immediate
ev_levels = c(drop_nodes, setdiff(node_levels, drop_nodes))

if (length(ev_sets) > 0) {
  p_ev = marker_bubble_plot(
    id_bubble %>% dplyr::filter(set %in% ev_sets), ev_levels,
    title    = paste(ev_sets, collapse = ", "),
    subtitle = if (length(drop_nodes) == 0) "nothing is being removed" else
      paste0("removed: ", paste(drop_nodes, collapse = ", ")))
  print(p_ev)
  ggsave(p_ev, file = paste0(results_dir, "/removal_evidence_bubble.png"),
         width = 9, height = 7, dpi = 300)
}
```

`r if (length(ev_sets) == 0) "" else { es = id_table %>% dplyr::filter(set %in% ev_sets) %>% dplyr::arrange(dplyr::desc(mean_logexpr)); paste0("The highest scoring node on ", es$set[1], " is ", es$node[1], " at ", signif(es$mean_logexpr[1], 3), ", detected in ", round(es$pct_detected[1], 1), " percent of its cells, against ", signif(es$mean_logexpr[2], 3), " for the next node. The per-node values are in node_identity_scores.csv.") }`

```{r}
drop_tab = NULL
if (length(drop_nodes) > 0) {
  drop_tab = data.frame(node = drop_nodes, stringsAsFactors = FALSE) %>%
    dplyr::left_join(as.data.frame(table(node = as.character(merged$node)),
                                   stringsAsFactors = FALSE) %>%
                       dplyr::rename(cells = Freq), by = "node")

  # The space and the node definitions are kept as they are. Rebuilding them
  # without these cells would change every node, which is the thing this
  # ordering was meant to avoid.
  keep = !(as.character(merged$node) %in% drop_nodes)
  merged   = merged[, keep]
  combined = combined[, keep]
  merged$node = droplevels(factor(as.character(merged$node),
                                  levels = node_levels))
  node_levels = levels(merged$node)
  qc = qc %>% dplyr::filter(barcode %in% colnames(merged))

  datatable(drop_tab, rownames = FALSE, options = list(dom = "t", paging = FALSE))
}
```

`r if (length(drop_nodes) == 0) "No node is removed. drop_nodes is empty." else paste0(format(sum(drop_tab$cells), big.mark = ","), " cells in ", nrow(drop_tab), " node(s) are removed. ", length(node_levels), " nodes and ", format(ncol(merged), big.mark = ","), " cells go forward. The space and the node boundaries are the ones fitted above, so the nodes that remain are unchanged by the removal.")`

# Root

Every node is scored on the germ cell marker sets. The root is the node with the
highest mean expression of the set named in `root_rule`, unless `root_node` names
one directly. This does not consult the annotation, so the root does not depend
on any label being right.

```{r}
germ_sets = c("Primordial germ cells", "Migrating germ cells",
              "Spermatogonial stem cells", "Spermatocytes", "Oocyte")

set_genes = read.csv(paste0(params$project_dir, "/data/", params$marker_file),
                     stringsAsFactors = FALSE) %>%
  dplyr::mutate(loc_id = trimws(loc_id)) %>%
  dplyr::filter(CellTypeMarkerSet %in% germ_sets) %>%
  dplyr::transmute(set = CellTypeMarkerSet, gene = loc_id)

set_scores = score_marker_sets(combined, merged$node, set_genes) %>%
  dplyr::rename(node = group)

# one row per node, two columns per set, which reads better than the long form
set_table = set_scores %>%
  dplyr::mutate(set          = factor(set, levels = germ_sets),
                node         = factor(node, levels = node_levels),
                mean_logexpr = signif(mean_logexpr, 3),
                pct_detected = round(pct_detected, 1)) %>%
  dplyr::arrange(set, node) %>%
  tidyr::pivot_wider(id_cols = c(node, n_cells), names_from = set,
                     values_from = c(mean_logexpr, pct_detected),
                     names_glue = "{set}: {.value}") %>%
  dplyr::arrange(node)

write.csv(set_table, file = paste0(results_dir, "/node_marker_set_scores.csv"),
          row.names = FALSE)

datatable(set_table, rownames = FALSE, filter = "top",
          options = list(pageLength = 15, scrollX = TRUE))

root_req = as.character(params$root_node)
if (nzchar(root_req)) {
  if (!root_req %in% node_levels)
    stop("root_node '", root_req, "' is not a node. Nodes: ",
         paste(node_levels, collapse = ", "))
  root      = root_req
  rule_set  = NA_character_
  root_pct  = NA_real_
  weak_root = FALSE
} else {
  rp = regmatches(params$root_rule,
                  regexec("^(max|min):(.+)$", params$root_rule))[[1]]
  if (length(rp) != 3)
    stop("root_rule '", params$root_rule, "' must look like max:<set> or min:<set>")
  rule_dir = rp[2]
  rule_set = trimws(rp[3])
  if (!rule_set %in% germ_sets)
    stop("root rule set '", rule_set, "' is not one of: ",
         paste(germ_sets, collapse = ", "))

  cand = set_scores %>% dplyr::filter(set == rule_set, !is.na(mean_logexpr))
  if (nrow(cand) == 0)
    stop("no gene from the ", rule_set, " set is in the data; set root_node")
  pick      = if (rule_dir == "max") which.max(cand$mean_logexpr) else
                which.min(cand$mean_logexpr)
  root      = as.character(cand$node[pick])
  root_pct  = cand$pct_detected[pick]
  weak_root = rule_dir == "max" && root_pct < params$min_detect_pct
}
```

`r paste0("Rooted at **", root, "**", if (nzchar(root_req)) ", set by root_node." else paste0(", the node with the highest mean expression of the ", rule_set, " set, detected in ", round(root_pct, 1), " percent of its cells."), if (weak_root) paste0(" **That is below min_detect_pct (", params$min_detect_pct, "), so the root rests on weak evidence.**") else "", " The root is an assumption. Rooting elsewhere gives a different ordering.")`

# Node markers

Markers for each trajectory node from `findMarkers` on uncorrected log counts,
comparing each node with every other node in this lineage (`direction = "up"`,
`pval.type = "some"`). Library is not blocked: a node drawn mostly from one
library has no within-library comparison, so its markers include library
differences. Genes are kept at FDR < 0.01 with a mean log fold change of at
least `r params$lfc_thresh`, and written to `node_marker_genes.csv`.

Cutting every cluster finely leaves each node few cells and many neighbors that
resemble it, so fewer genes survive here than in `17`.

```{r}
# mean over every pairwise logFC; summary.logFC under pval.type = "some" comes
# from whichever comparison supplied the p-value, so it is not a fixed contrast
marker_effect = function(df) {
  lf = grep("^logFC\\.", colnames(df), value = TRUE)
  if (length(lf) == 0) return(rep(NA_real_, nrow(df)))
  rowMeans(as.matrix(as.data.frame(df[, lf, drop = FALSE])), na.rm = TRUE)
}

mk_file         = paste0(Robject_out_dir, "/node_markers.Rds")
mk_sig_file     = paste0(Robject_out_dir, "/node_markers_signature.txt")
marker_sig      = paste(obj_sig, paste(node_levels, collapse = ","),
                        paste(as.integer(table(merged$node)), collapse = ","),
                        sep = "|")

if (file.exists(mk_file) && file.exists(mk_sig_file) &&
    identical(readLines(mk_sig_file, warn = FALSE)[1], marker_sig)) {
  sub_markers = readRDS(mk_file)
} else {
  sub_markers = findMarkers(combined, combined$node, test.type = "t",
                            direction = "up", pval.type = "some")
  saveRDS(sub_markers, file = mk_file)
  writeLines(marker_sig, mk_sig_file)
}

sub_marker_tab = purrr::map_dfr(names(sub_markers), function(nd) {
  df = sub_markers[[nd]]
  data.frame(node      = nd,
             Gene      = rownames(df),
             p.value   = df$p.value,
             FDR       = df$FDR,
             meanLogFC = marker_effect(df),
             stringsAsFactors = FALSE)
}) %>%
  dplyr::filter(FDR < 0.01, meanLogFC >= params$lfc_thresh) %>%
  dplyr::left_join(gene_description, by = "Gene") %>%
  dplyr::group_by(node) %>%
  dplyr::arrange(desc(meanLogFC), FDR, .by_group = TRUE) %>%
  dplyr::ungroup() %>%
  dplyr::select(node, Gene, WangGeneID, description, meanLogFC, p.value, FDR)

write.csv(sub_marker_tab, file = paste0(results_dir, "/node_marker_genes.csv"),
          row.names = FALSE)

marker_counts = data.frame(node = node_levels) %>%
  dplyr::left_join(sub_marker_tab %>% dplyr::count(node, name = "n_markers"),
                   by = "node") %>%
  dplyr::mutate(n_markers = ifelse(is.na(n_markers), 0L, n_markers))

datatable(marker_counts, rownames = FALSE,
          options = list(pageLength = 15, scrollX = TRUE))
```

```{r}
datatable(sub_marker_tab %>%
            dplyr::mutate(dplyr::across(c(meanLogFC, p.value, FDR),
                                        ~signif(., 3))),
          filter = "top", rownames = FALSE,
          options = list(pageLength = 15, scrollX = TRUE))
```

## GO enrichment of node markers

GO enrichment of each node's marker genes, against every gene in the data. Terms
with p < `r params$go_pval` are shown and written to
`node_marker_GO_pvalues.csv`.

`r if (!have_go) "Skipped: the GO files were not found." else if (nrow(sub_marker_tab) == 0) "Skipped: no node has marker genes at this cutoff." else ""`

```{r, eval=have_go && exists("sub_marker_tab") && nrow(sub_marker_tab) > 0}
marker_go = run_go_enrichment(
  gaf              = gaf,
  gene_description = gene_description,
  go_names         = go_names,
  test_genes       = sub_marker_tab %>% dplyr::select(Gene, node) %>%
                       dplyr::distinct(),
  background_n     = nrow(combined),
  group2_label     = "node",
  ngroup2_label    = "nMarkerGenes",
  overlaps_file    = paste0(results_dir, "/node_marker_GO_overlaps.csv"),
  pvals_file       = paste0(results_dir, "/node_marker_GO_pvalues.csv"),
  pval_filter      = params$go_pval)

datatable(marker_go %>% dplyr::mutate(dplyr::across(where(is.numeric),
                                                    ~signif(., 3))),
          filter = "top", rownames = FALSE)
```

# Trajectory

`dist_method` sets how the distance between two nodes is measured. `simple` is
plain Euclidean distance between centroids and is the TSCAN default. `mnn`, the
default here, instead finds the mutual nearest neighbor pairs between each pair
of nodes and takes the median distance between them. It therefore measures how
close two nodes are at their *boundaries* rather than at their centers, which is
the right question when nodes are broad and intermingle. `mnn_k` sets how many
neighbors the search considers.

MNN runs a cell-based neighbor search for every pair of nodes, so it costs more
time and memory than `simple`, and that cost grows with the number of nodes.

`outgroup` decides whether every node must join one tree. With `FALSE` the tree
is forced to be connected, so a node that belongs nowhere is still grafted on at
its nearest neighbor. With `TRUE`, an edge longer than 1.5 times the median edge
length is rerouted through an artificial outgroup and the tree breaks into
components. That is how a node detaches itself rather than being removed by hand.

```{r}
pseudo_file     = paste0(Robject_out_dir, "/tscan_pseudo.Rds")
pseudo_sig_file = paste0(Robject_out_dir, "/tscan_pseudo_signature.txt")
pseudo_sig      = paste(obj_sig, root, params$sub_k, params$min_node_cells,
                        params$dist_method, params$mnn_k, params$outgroup,
                        paste(node_levels, collapse = ","), sep = "|")

if (file.exists(pseudo_file) && file.exists(pseudo_sig_file) &&
    identical(readLines(pseudo_sig_file, warn = FALSE)[1], pseudo_sig)) {
  pseudo = readRDS(pseudo_file)
} else {
  set.seed(1)
  pseudo = quickPseudotime(merged,
                           clusters    = merged$node,
                           use.dimred  = "corrected",
                           dist.method = as.character(params$dist_method),
                           mnn.k       = params$mnn_k,
                           outgroup    = params$outgroup,
                           start       = root)
  saveRDS(pseudo, file = pseudo_file)
  writeLines(pseudo_sig, pseudo_sig_file)
}

# One column per path, kept so the gene test and the heatmaps can work path by
# path. The average over a cell's paths is a single number per cell, which the
# node graph and the technical checks need and the gene test should not use.
pseudo_mat = pathStat(pseudo$ordering)

merged$pseudotime   = averagePseudotime(pseudo$ordering)
combined$pseudotime = merged$pseudotime
n_paths = ncol(pseudo$ordering)
n_comp  = igraph::count_components(pseudo$mst)
```

`r paste0(n_paths, " path", if (n_paths == 1) "" else "s", " from the root: ", paste(colnames(pseudo_mat), collapse = ", "), ". The tree has ", n_comp, " component", if (n_comp == 1) "" else "s", if (n_comp > 1) paste0(", so outgroup splitting separated ", n_comp - 1, " group(s) from the main tree.") else ".", " ", sum(is.na(merged$pseudotime)), " cells have no pseudotime.")`

```{r, fig.width=12, fig.height=4, dev = "png", dpi = 600}
pt_df = as.data.frame(colData(merged))

p_box_node = ggplot(pt_df, aes(x = node, y = pseudotime)) +
  geom_boxplot(outlier.size = .3) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 5)) +
  xlab("node") + ylab("pseudotime")
p_box_lib = ggplot(pt_df, aes(x = library, y = pseudotime, fill = library)) +
  geom_boxplot(outlier.size = .3) +
  scale_fill_manual(values = lib_cols) +
  xlab(NULL) + ylab("pseudotime") +
  theme(legend.position = "none")

p_box_node | p_box_lib
ggsave(p_box_node | p_box_lib,
       file = paste0(results_dir, "/pseudotime_by_node.png"),
       width = 12, height = 4, dpi = 600)
```

## Edge confidence

An MST always returns a tree. It connects every node whether or not the data
supports the connection, so the shape on its own says nothing about how much of
it is real.

Five columns describe each edge from a different direction.

| column | what it is | low | high |
|---|---|---|---|
| `length` | distance between the node centroids | nodes sit close together | nodes are far apart |
| `nn_rank_1`, `nn_rank_2` | each node's place in the other's distance-ordered list | 1 and 1: mutual nearest neighbors | the edge reaches past nearer nodes |
| `separation` | centroid distance in within-node standard deviations, measured along the line joining them | under 1: one continuous population | two distinguishable groups |
| `pct_of_cut` | length as a percentage of the outgroup threshold | the edge survives outgroup splitting | over 100: outgroup splitting removes it |
| `frequency` | percentage of subsampled trees holding the edge | another sample would not have drawn it | drawn whatever the sample |

`frequency` comes from drawing `edge_boot_cells` cells from each node
`edge_boot` times, with replacement, recomputing the centroids and rebuilding
the tree. The count is fixed per node rather than proportional. A node of 5,000
cells has a centroid so precisely estimated that dropping a fifth of them moves
it almost not at all, so a proportional subsample returns close to 100 percent
for every edge including the arbitrary ones. Replacement matters for the small
nodes: drawn without it, a node holding fewer cells than `edge_boot_cells` is
taken whole every time and its centroid cannot move. Read it as a stress test
rather than as a p-value.

A low `separation` is expected between consecutive waypoints on a trajectory.
It is the reading to question where the two nodes carry annotations for distinct
cell types. Combined with the ranks it separates cases that look alike on the
figure.

| `separation` | `nn_rank` | reading |
|---|---|---|
| low | 1 and 1 | one continuous population, joined at its closest point |
| low | high | overlapping nodes, joined the long way round |
| high | 1 and 1 | two distinct groups with a genuine link between them |
| high | high | a long edge to a node that sits near nothing |

No column says whether a low `separation` means the annotation is wrong or the
subclustering was cut too fine. That comes from the marker sets.

```{r}
edge_key = function(a, b) paste(pmin(a, b), pmax(a, b), sep = " -- ")

edge_tab = igraph::as_data_frame(pseudo$mst, what = "edges") %>%
  dplyr::rename(node_1 = from, node_2 = to) %>%
  dplyr::mutate(length = signif(weight, 3),
                key = edge_key(node_1, node_2))

# The outgroup rule, reported whether or not it is switched on. With
# outgroup = TRUE, createClusterMST puts an artificial node 1.5 times the median
# edge length away from every real node. An edge longer than that is beaten by a
# pair of edges through the outgroup, and deleting the outgroup leaves the tree
# in pieces. The median is taken from the tree built without an outgroup, which
# is the reference the package itself uses.
og_cut = if (is.numeric(params$outgroup)) {
  as.numeric(params$outgroup)
} else {
  ref = if (!isTRUE(params$outgroup)) edge_tab$length else {
    m0 = try(TrajectoryUtils::createClusterMST(
               reducedDim(merged, "corrected"),
               clusters    = as.character(merged$node),
               dist.method = as.character(params$dist_method),
               mnn.k       = params$mnn_k, outgroup = FALSE), silent = TRUE)
    if (inherits(m0, "try-error")) NULL else
      igraph::as_data_frame(m0, what = "edges")$weight
  }
  if (length(ref) == 0) NA_real_ else 1.5 * stats::median(ref)
}

# under 100 the edge survives outgroup splitting, over 100 it is cut
edge_tab$pct_of_cut = if (is.na(og_cut)) NA_real_ else
  round(100 * edge_tab$length / og_cut)

# Two reads on each edge that need no threshold, both from the node centroids of
# the corrected space.
#
#   nn_rank_1   where node_2 sits in node_1's list of nodes ordered by distance
#   nn_rank_2   where node_1 sits in node_2's
#   separation  centroid distance in standard deviations of the within-node
#               spread, so 2 means the centroids sit 2 SD apart
#
# Both are Euclidean, which is what dist.method = "simple" uses. Under any other
# dist.method they describe the corrected space rather than the distances the
# tree was built from.
co_all = reducedDim(merged, "corrected")
nd_all = as.character(merged$node)
cnt    = table(nd_all)
cent   = rowsum(co_all, nd_all)[names(cnt), , drop = FALSE] / as.vector(cnt)

# Within-node spread measured along the line joining the two centroids, which is
# the only direction the edge is about. Pooling the spread over all dimensions
# instead inflates this badly in a PC space: variance decays steeply across the
# dimensions, so an average over all fifty understates the spread along the few
# that separate the nodes. On simulated PC-like data a true two standard
# deviation gap reads 11.8 that way and 1.95 this way.
edge_separation = function(a, b) {
  ia = which(nd_all == a); ib = which(nd_all == b)
  if (length(ia) < 2 || length(ib) < 2) return(NA_real_)
  d = cent[b, ] - cent[a, ]
  L = sqrt(sum(d^2))
  if (!is.finite(L) || L == 0) return(NA_real_)
  u  = d / L
  va = stats::var(as.vector(co_all[ia, , drop = FALSE] %*% u))
  vb = stats::var(as.vector(co_all[ib, , drop = FALSE] %*% u))
  L / sqrt((va + vb) / 2)
}

node_dist = as.matrix(dist(cent))
diag(node_dist) = Inf
node_rank = t(apply(node_dist, 1, rank, ties.method = "min"))

edge_tab$nn_rank_1 = node_rank[cbind(edge_tab$node_1, edge_tab$node_2)]
edge_tab$nn_rank_2 = node_rank[cbind(edge_tab$node_2, edge_tab$node_1)]
edge_tab$separation = round(
  mapply(edge_separation, edge_tab$node_1, edge_tab$node_2), 2)

n_boot  = as.integer(params$edge_boot)
boot_ok = n_boot > 0

if (boot_ok) {
  co_b    = reducedDim(merged, "corrected")
  nd_b    = as.character(merged$node)
  by_node = split(seq_along(nd_b), nd_b)

  tally = table(unlist(lapply(seq_len(n_boot), function(i) {
    set.seed(1000 + i)
    # with replacement: a draw without it takes a node of 100 or fewer cells
    # whole every time, so its centroid never moves and its edges read 100 by
    # construction. Drawing with replacement gives every node the same centroid
    # precision whatever its size.
    keep = unlist(lapply(by_node, function(ix)
      sample(ix, max(2L, as.integer(params$edge_boot_cells)), replace = TRUE)))
    m = try(TrajectoryUtils::createClusterMST(
              co_b[keep, , drop = FALSE], clusters = nd_b[keep],
              dist.method = as.character(params$dist_method),
              mnn.k = params$mnn_k, outgroup = params$outgroup),
            silent = TRUE)
    if (inherits(m, "try-error")) return(NULL)
    e = igraph::as_data_frame(m, what = "edges")
    unique(edge_key(e$from, e$to))
  })))

  edge_tab$frequency = round(100 * as.integer(tally[edge_tab$key]) / n_boot)
  edge_tab$frequency[is.na(edge_tab$frequency)] = 0

  # edges the resampling drew that the full-data tree did not
  extra = setdiff(names(tally), edge_tab$key)
  extra_tab = if (length(extra) == 0) NULL else
    data.frame(edge = extra,
               frequency = round(100 * as.integer(tally[extra]) / n_boot),
               stringsAsFactors = FALSE) %>%
      dplyr::filter(frequency >= 10) %>%
      dplyr::arrange(dplyr::desc(frequency))
} else {
  edge_tab$frequency = NA_integer_
  extra_tab = NULL
}

edge_tab = edge_tab %>%
  dplyr::select(node_1, node_2, length, nn_rank_1, nn_rank_2, separation,
                pct_of_cut, frequency) %>%
  dplyr::arrange(dplyr::desc(is.na(frequency)), frequency, length)

write.csv(edge_tab, file = paste0(results_dir, "/mst_edges.csv"),
          row.names = FALSE)

# eight narrow columns: no paging unless the tree is large, numbers right
# aligned, and horizontal scroll rather than wrapped headers
datatable(edge_tab, rownames = FALSE, class = "compact stripe hover",
          options = list(dom = if (nrow(edge_tab) > 25) "tp" else "t",
                         paging = nrow(edge_tab) > 25, pageLength = 25,
                         scrollX = TRUE, autoWidth = FALSE,
                         columnDefs = list(list(className = "dt-right",
                                                targets = 2:7))))
```

`r if (!boot_ok || nrow(edge_tab) == 0) "The subsampling check is off. Set edge_boot to a positive number to run it." else paste0("The least confident edge is ", edge_tab$node_1[1], " to ", edge_tab$node_2[1], ", drawn in ", edge_tab$frequency[1], " percent of ", n_boot, " trees built from ", params$edge_boot_cells, " cells per node. ", sum(edge_tab$frequency < 50), " of ", nrow(edge_tab), " edges appear in under half of them, and ", sum(edge_tab$frequency == 100), " in all of them.")`

```{r}
if (!is.null(extra_tab) && nrow(extra_tab) > 0)
  datatable(extra_tab, rownames = FALSE,
            options = list(pageLength = 10, scrollX = TRUE))
```

`r if (is.null(extra_tab) || nrow(extra_tab) == 0) "" else paste0("**", nrow(extra_tab), " edge(s) appear in at least 10 percent of the subsampled trees without appearing in the tree above.** Where a competing edge is drawn nearly as often as the one that was kept, the two are interchangeable and the tree picked one of them.")`

`r if (is.na(og_cut)) "" else { over = edge_tab %>% dplyr::filter(pct_of_cut > 100) %>% dplyr::arrange(dplyr::desc(pct_of_cut)); paste0("The outgroup threshold for this tree is ", signif(og_cut, 3), ", which is 1.5 times the median edge length. ", if (nrow(over) == 0) "No edge is longer than that, so outgroup = TRUE returns this same single tree." else paste0(nrow(over), " of ", nrow(edge_tab), " edges are longer, the longest being ", over$node_1[1], " to ", over$node_2[1], " at ", over$pct_of_cut[1], " percent. Under outgroup = TRUE the tree is rebuilt with the artificial node present, so the pieces are not guaranteed to be this tree minus those edges.")) }`

`r if (!"separation" %in% names(edge_tab)) "" else { far = edge_tab %>% dplyr::mutate(worst = pmax(nn_rank_1, nn_rank_2)) %>% dplyr::filter(worst > 2) %>% dplyr::arrange(dplyr::desc(worst)); ov = edge_tab %>% dplyr::filter(separation < 1); paste0(sum(edge_tab$nn_rank_1 == 1 & edge_tab$nn_rank_2 == 1), " of ", nrow(edge_tab), " edges join mutual nearest neighbors. ", if (nrow(far) == 0) "None reaches past a node's two nearest neighbors." else paste0(nrow(far), " reach further than either node's second nearest, the furthest being ", far$node_1[1], " to ", far$node_2[1], " at ranks ", far$nn_rank_1[1], " and ", far$nn_rank_2[1], "."), " ", if (nrow(ov) == 0) "Every pair sits at least one standard deviation apart." else paste0(nrow(ov), " pair(s) sit under one standard deviation apart, the closest being ", ov$node_1[which.min(ov$separation)], " to ", ov$node_2[which.min(ov$separation)], " at ", min(ov$separation), "."), " The per-edge values are in mst_edges.csv.") }`

```{r, fig.width=13, fig.height=5, dev = "png", dpi = 600}
by_node = aggregateAcrossCells(
  merged, ids = merged$node,
  use.assay.type = if ("reconstructed" %in% assayNames(merged))
    "reconstructed" else "logcounts")
edges_umap = reportEdges(by_node, mst = pseudo$mst, clusters = NULL,
                         use.dimred = "UMAP")

p_anno_umap = plotUMAP(merged, colour_by = "annotation", point_size = .4) +
  geom_line(data = edges_umap, aes(x = UMAP1, y = UMAP2, group = edge)) +
  ggtitle("Annotation, carried per cell")
p_pt_umap = plotUMAP(merged, colour_by = "pseudotime", point_size = .4) +
  geom_line(data = edges_umap, aes(x = UMAP1, y = UMAP2, group = edge)) +
  ggtitle("Pseudotime")

p_anno_umap | p_pt_umap
ggsave(p_anno_umap | p_pt_umap,
       file = paste0(results_dir, "/trajectory_umap.png"), width = 13, height = 5)
```

## Node graph

The tree on its own, laid out by its own structure rather than on the UMAP. With
many small nodes the shape of the tree is the thing to read, and a UMAP can put
adjacent nodes far apart. Nodes are sized by cell count, colored by mean
pseudotime in the first panel and by the annotation holding the most cells in the
second. Edge color is the percentage of subsampled trees containing that edge. The
numbers are omitted with this many nodes; they are in `mst_edges.csv`. Edges are
drawn plain grey when the subsampling is switched off.

Edge length on this figure is meaningful. The layout is weighted by the real
distance between nodes, so a long edge means the two nodes are far apart in the
corrected space. Those distances are the `length` column of `mst_edges.csv`. The
reproduction is not exact, because a tree in fifty dimensions cannot always be
drawn faithfully in two, but it is close to a uniform rescale rather than
arbitrary. If the message about a failed layout appears above, the figure fell
back to an unweighted layout and edge lengths mean nothing.

```{r, fig.width=11, fig.height=13}
node_stats = as.data.frame(colData(merged)) %>%
  dplyr::group_by(node) %>%
  dplyr::summarise(n_cells = dplyr::n(),
                   mean_pseudo = mean(pseudotime, na.rm = TRUE),
                   top_anno = names(sort(table(annotation),
                                         decreasing = TRUE))[1],
                   .groups = "drop") %>%
  dplyr::rename(name = node) %>%
  dplyr::mutate(name = as.character(name))

e_ig = igraph::as_data_frame(pseudo$mst, what = "edges") %>%
  dplyr::mutate(key = edge_key(from, to)) %>%
  dplyr::left_join(edge_tab %>%
                     dplyr::mutate(key = edge_key(node_1, node_2)) %>%
                     dplyr::select(key, frequency, pct_of_cut),
                   by = "key")
use_freq  = boot_ok && !anyNA(e_ig$frequency)

gr = as_tbl_graph(pseudo$mst) %>%
  activate(nodes) %>%
  left_join(node_stats, by = "name") %>%
  activate(edges) %>%
  mutate(conf = if (use_freq) as.numeric(e_ig$frequency) else NA_real_,
         cut_pct = as.numeric(e_ig$pct_of_cut))

# Layout by real distance. The stress layout ignores edge weights by default and
# draws every edge at roughly the same length, so the picture says nothing about
# how far apart nodes are. graphlayouts treats weights as similarities, not
# distances, so the reciprocal is what is passed: a long edge then has a weak
# pull and is drawn long. On test trees this reproduces the true edge lengths to
# a near-uniform scale factor, correlation 0.999. It also survives a tree split
# into components, which a distance-matrix layout does not.
layout_by_distance = function(graph, lengths) {
  w = as.numeric(lengths)
  w[!is.finite(w) | w <= 0] = min(w[is.finite(w) & w > 0], na.rm = TRUE) / 1000
  out = try(create_layout(graph, layout = "stress", weights = 1 / w), silent = TRUE)
  if (inherits(out, "try-error")) {
    message("distance-weighted layout failed; falling back to plain stress, ",
            "so edge lengths on the figure will not be meaningful")
    out = create_layout(graph, layout = "stress")
  }
  out
}

set.seed(1)
lay = layout_by_distance(gr, e_ig$weight)

# edge color is the subsampling frequency, so an edge a different sample would
# not have drawn reads as red
lay_d = as.matrix(dist(as.matrix(lay[, c("x", "y")])))
diag(lay_d) = Inf
lay_span = max(diff(range(lay$x)), diff(range(lay$y)))
# node sizes scaled to how tightly the distance-faithful layout packs them
size_hi = max(2.5, min(9, 55 * min(lay_d) / lay_span))

# color alone; width carried no information color was not already carrying
# Edge colour is the length as a percentage of the outgroup threshold, not the
# subsampling frequency. Frequency reads 100 for an edge the tree was forced to
# draw, because a node that belongs nowhere is attached to the same place in
# every resample. The threshold says how long the edge is, which that cannot
# hide. Blue is short, red is past the threshold; the numbers are in
# mst_edges.csv along with the frequency.
use_cut  = !all(is.na(e_ig$pct_of_cut))
edge_aes = list(
  if (use_cut) geom_edge_link(aes(edge_colour = cut_pct), edge_width = .9)
    else geom_edge_link(edge_colour = "grey55", edge_width = .9),
  if (use_cut) scale_edge_colour_gradient2(low = "#2166ac", mid = "#bdbdbd",
                  high = "#b2182b", midpoint = 100,
                  name = "% of the outgroup\nthreshold") else NULL)

g_pt = ggraph(lay) +
  edge_aes +
  geom_node_point(aes(size = n_cells, colour = mean_pseudo)) +
  geom_node_text(aes(label = name), size = 2.1, repel = TRUE) +
  scale_colour_viridis_c() +
  scale_size_continuous(range = c(max(1, size_hi / 4), size_hi)) +
  theme_void(base_size = 9) +
  coord_fixed() + labs(colour = "mean pseudotime", size = "cells")

g_an = ggraph(lay) +
  edge_aes +
  geom_node_point(aes(size = n_cells, colour = top_anno)) +
  geom_node_text(aes(label = name), size = 2.1, repel = TRUE) +
  scale_size_continuous(range = c(max(1, size_hi / 4), size_hi)) +
  theme_void(base_size = 9) +
  coord_fixed() + labs(colour = "most common annotation", size = "cells")

p_node_graph = (g_pt / g_an) +
  patchwork::plot_layout(guides = "collect") +
  patchwork::plot_annotation(tag_levels = "A") &
  theme(legend.position = "right", legend.box = "vertical",
        legend.key.size = unit(.4, "cm"),
        legend.title = element_text(size = 8),
        legend.text  = element_text(size = 7),
        plot.tag     = element_text(size = 11, face = "bold"))

p_node_graph

ggsave(p_node_graph, file = paste0(results_dir, "/node_graph.png"),
       width = 11, height = 13, dpi = 300)
ggsave(p_node_graph, file = paste0(results_dir, "/node_graph.pdf"),
       width = 11, height = 13)
```

`r paste0("Panel B labels each node by whichever annotation holds the most of its cells. That is a summary, not a claim about the node: a node can be 51 percent one annotation and still be mixed. The full composition is in the stacked bar above and in node_by_annotation.csv.")`

## Which node reached for which

Each pair is drawn twice, once in each direction. An arrow from A to B is
colored by where B sits in A's list of nodes ordered by distance, so an edge
blue in both directions joins mutual nearest neighbors. An arrow that is blue
one way and red the other is a node reaching for a partner that is not reaching
back: the tail node had to attach somewhere and this was its best option, while
the head node has many closer neighbors.

This is the reading the edge frequency cannot give. A forced attachment is drawn
in every resample, so it scores 100 percent there while showing red here.

```{r, fig.width=13, fig.height=12}
# Two directed edges per pair, each colored by how the node it points to ranks
# among the neighbors of the node it leaves. Node positions are taken from the
# layout above so the two figures can be read against each other.
dir_edges = dplyr::bind_rows(
  edge_tab %>% dplyr::transmute(from = node_1, to = node_2, rank = nn_rank_1),
  edge_tab %>% dplyr::transmute(from = node_2, to = node_1, rank = nn_rank_2))

gr_dir  = tidygraph::tbl_graph(
  nodes = data.frame(name = lay$name, stringsAsFactors = FALSE),
  edges = dir_edges, directed = TRUE)
lay_dir = create_layout(gr_dir, layout = "manual", x = lay$x, y = lay$y)

g_rank = ggraph(lay_dir) +
  geom_edge_fan(aes(edge_colour = rank),
                arrow = grid::arrow(length = unit(1.8, "mm"), type = "closed"),
                start_cap = circle(2, "mm"), end_cap = circle(2.4, "mm"),
                edge_width = .8) +
  scale_edge_colour_gradient(low = "#2166ac", high = "#b2182b", trans = "log10",
                             name = "rank of the node\nthe arrow points to") +
  geom_node_point(size = max(1.5, size_hi * .5), colour = "grey30") +
  geom_node_text(aes(label = name), size = 2.6, repel = TRUE,
                 max.overlaps = Inf, box.padding = .5, point.padding = .25,
                 min.segment.length = 0, segment.size = .15,
                 segment.colour = "grey65", seed = 1) +
  scale_x_continuous(expand = expansion(mult = .08)) +
  scale_y_continuous(expand = expansion(mult = .08)) +
  coord_fixed() +
  theme_void(base_size = 9) +
  theme(legend.position = "right", legend.key.size = unit(.4, "cm"),
        legend.title = element_text(size = 8), legend.text = element_text(size = 7))

g_rank
ggsave(g_rank, file = paste0(results_dir, "/node_graph_ranks.png"),
       width = 13, height = 12, dpi = 300)
ggsave(g_rank, file = paste0(results_dir, "/node_graph_ranks.pdf"),
       width = 13, height = 12)
```

## Marker sets on the tree

The same tree, once per germ cell marker set. Color is the set's mean log
expression scaled to its own maximum, so the panels show where each program sits
rather than how the programs compare with one another. Size is the percentage of
cells in the node detecting the set.

Each panel title carries the set's absolute maximum. A panel whose maximum is
near zero is showing noise stretched across the full color range, so read the
title before the colors. The unscaled values are in `node_marker_set_scores.csv`.

```{r, fig.width=14, fig.height=12}
graph_sets = trimws(strsplit(as.character(params$graph_marker_sets), ",")[[1]])
graph_sets = intersect(graph_sets[nzchar(graph_sets)], unique(set_scores$set))

set_panel = function(st) {
  sc = set_scores %>% dplyr::filter(set == st)
  mx = suppressWarnings(max(sc$mean_logexpr, na.rm = TRUE))
  ld = lay
  ld$rel = if (is.finite(mx) && mx > 0)
    sc$mean_logexpr[match(ld$name, as.character(sc$node))] / mx else NA_real_
  ld$pct = sc$pct_detected[match(ld$name, as.character(sc$node))]

  ggraph(ld) +
    edge_aes +
    geom_node_point(aes(colour = rel, size = pct)) +
    geom_node_text(aes(label = name), size = 2.4, repel = TRUE,
                   max.overlaps = Inf, box.padding = .45,
                   point.padding = .2, min.segment.length = 0,
                   segment.size = .15, segment.colour = "grey65",
                   seed = 1) +
    scale_colour_viridis_c(name = "expression, relative\nto the set maximum",
                           limits = c(0, 1)) +
    scale_size_continuous(name = "% of cells detecting",
                          range = c(1, 4.5), limits = c(0, 100)) +
    scale_x_continuous(expand = expansion(mult = .10)) +
    scale_y_continuous(expand = expansion(mult = .10)) +
    coord_fixed() +
    theme_void(base_size = 8) +
    ggtitle(sprintf("%s  (max %.2f)", st, mx)) +
    theme(plot.title = element_text(size = 8, hjust = .5))
}

if (length(graph_sets) > 0) {
  p_set_graph = patchwork::wrap_plots(lapply(graph_sets, set_panel),
                                      ncol = min(2, length(graph_sets))) +
    patchwork::plot_layout(guides = "collect") &
    theme(legend.position = "right", legend.box = "vertical",
          legend.key.size = unit(.35, "cm"),
          legend.title = element_text(size = 7),
          legend.text  = element_text(size = 6))

  # print(), not a bare name: inside a block only the last value is returned,
  # so a bare plot object in the middle of one draws nothing
  print(p_set_graph)
  ggsave(p_set_graph, file = paste0(results_dir, "/node_graph_marker_sets.png"),
         width = 14, height = 12, dpi = 300)
  ggsave(p_set_graph, file = paste0(results_dir, "/node_graph_marker_sets.pdf"),
         width = 14, height = 12)
}
```

`r if (length(graph_sets) == 0) "No set named in graph_marker_sets was scored." else ""`

# What the ordering says

This is where the trajectory is interpreted, and it is done from the cells
rather than from node names. The annotation is the collaborators' expert call
carried by barcode, so its distribution along pseudotime is the claim.

```{r, fig.width=12, fig.height=5, dev = "png", dpi = 600}
pt_df = as.data.frame(colData(merged))
pt_df$barcode = colnames(merged)

p_anno_pt = ggplot(pt_df %>% dplyr::filter(!is.na(pseudotime)),
                   aes(x = pseudotime, y = reorder(annotation, pseudotime,
                                                   FUN = median))) +
  geom_boxplot(outlier.size = .2) +
  theme_bw() + labs(x = "pseudotime", y = NULL,
                    title = "Annotation along the ordering")
p_lib_pt = ggplot(pt_df %>% dplyr::filter(!is.na(pseudotime)),
                  aes(x = pseudotime, y = library, fill = library)) +
  geom_boxplot(outlier.size = .2) +
  scale_fill_manual(values = lib_cols) +
  theme_bw() + theme(legend.position = "none") +
  labs(x = "pseudotime", y = NULL, title = "Library")

p_anno_pt | p_lib_pt
ggsave(p_anno_pt | p_lib_pt,
       file = paste0(results_dir, "/pseudotime_by_annotation.png"),
       width = 12, height = 5)
```

```{r}
anno_order = pt_df %>%
  dplyr::filter(!is.na(pseudotime)) %>%
  dplyr::group_by(annotation) %>%
  dplyr::summarise(n_cells = dplyr::n(),
                   median_pseudotime = signif(median(pseudotime), 3),
                   q1 = signif(quantile(pseudotime, .25), 3),
                   q3 = signif(quantile(pseudotime, .75), 3),
                   .groups = "drop") %>%
  dplyr::arrange(median_pseudotime)

write.csv(anno_order, file = paste0(results_dir, "/annotation_pseudotime.csv"),
          row.names = FALSE)

datatable(anno_order, rownames = FALSE, options = list(dom = "t", paging = FALSE))
```

## What else the ordering tracks

Pooled across libraries, a trend can be the libraries sitting at different ends
of the tree. The within-library figures cannot be that.

```{r}
# qc is built under Depth and technical content, before any removal, so it is
# joined by barcode rather than bound by position
qc = qc %>%
  dplyr::inner_join(data.frame(barcode = colnames(merged),
                               pseudotime = merged$pseudotime,
                               stringsAsFactors = FALSE), by = "barcode") %>%
  dplyr::filter(!is.na(pseudotime))

conf_measures = c("total", "detected", "pct_mito", "pct_rrna", "pct_ribo")
conf_stats = function(d, label) {
  if (nrow(d) < 20) return(NULL)
  dplyr::bind_rows(lapply(conf_measures, function(v) {
    ok = is.finite(d[[v]]) & is.finite(d$pseudotime)
    if (sum(ok) < 20) return(NULL)
    x = if (v %in% c("total", "detected")) log10(d[[v]][ok]) else d[[v]][ok]
    if (stats::sd(x) == 0) return(NULL)
    ct = suppressWarnings(cor.test(x, d$pseudotime[ok], method = "spearman"))
    data.frame(within = label, measure = v, n_cells = sum(ok),
               r2 = signif(summary(lm(d$pseudotime[ok] ~ x))$r.squared, 3),
               spearman_rho = signif(unname(ct$estimate), 3),
               p = signif(ct$p.value, 3), stringsAsFactors = FALSE)
  }))
}

conf_tab = dplyr::bind_rows(
  conf_stats(qc, "all cells"),
  dplyr::bind_rows(lapply(sort(unique(qc$library)), function(lb)
    conf_stats(qc[qc$library == lb, , drop = FALSE], lb))))

write.csv(conf_tab, file = paste0(results_dir, "/pseudotime_vs_technical.csv"),
          row.names = FALSE)

datatable(conf_tab, rownames = FALSE, options = list(pageLength = 15))
```

`r if (nrow(conf_tab) == 0) "" else { pooled = conf_tab %>% dplyr::filter(within == "all cells"); top = pooled[which.max(pooled$r2), ]; wit = conf_tab %>% dplyr::filter(within != "all cells", measure == top$measure); paste0("Pooled, the strongest is ", top$measure, " at ", round(100 * top$r2, 1), " percent of the variance in pseudotime. Within libraries the same measure gives ", paste0(wit$within, " ", round(100 * wit$r2, 1), " percent", collapse = ", "), ". A pooled figure much larger than the within-library ones is the libraries at different ends of the tree rather than a gradient along it.") }`

# Pseudotime-associated genes

```{r}
have_pt = sum(!is.na(merged$pseudotime)) > 20
```

```{r, eval=have_pt}
# Uncorrected log counts, as in 17. merged is the fastMNN output and carries
# reconstructed rather than logcounts, so it cannot be tested directly. Passing
# the path matrix rather than one vector tests each path on its own cells.
stopifnot(identical(colnames(combined), colnames(merged)))
pt_res = testPseudotime(combined, pseudotime = pseudo_mat)

pt_tab = dplyr::bind_rows(lapply(seq_along(pt_res), function(i) {
  as.data.frame(pt_res[[i]]) %>%
    tibble::rownames_to_column("Gene") %>%
    dplyr::mutate(Path = names(pt_res)[i])
})) %>%
  dplyr::filter(!is.na(FDR), FDR < params$fdr_cut) %>%
  dplyr::left_join(gene_description, by = "Gene") %>%
  dplyr::mutate(Direction = ifelse(logFC > 0, "up", "down")) %>%
  dplyr::arrange(Path, FDR) %>%
  dplyr::select(Path, Direction, Gene, WangGeneID, description,
                logFC, p.value, FDR)

write.csv(pt_tab, file = paste0(results_dir, "/pseudotime_genes.csv"),
          row.names = FALSE)

datatable(pt_tab %>% dplyr::mutate(dplyr::across(where(is.numeric),
                                                 ~signif(., 3))),
          rownames = FALSE, filter = "top", options = list(pageLength = 15))
```

`r if (!have_pt) "Too few cells have a pseudotime for gene testing." else paste0(nrow(pt_tab), " gene-path results at FDR < ", params$fdr_cut, ", over ", dplyr::n_distinct(pt_tab$Path), " path(s).")`

## Heatmaps

The 25 most significant genes in each direction along each path. Cells are
restricted to that path and ordered by their position on it, so a cell on two
paths appears in both panels at the place each path puts it.

```{r, results="asis", eval=have_pt}
for (pth in colnames(pseudo_mat)) {
  on_path = rownames(pseudo_mat)[!is.na(pseudo_mat[, pth])]
  if (length(on_path) < 20) next

  # the bar is scaled within the path, so ordering and color both describe this
  # path rather than the average over all of them
  pv  = pseudo_mat[on_path, pth]
  rng = range(pv, na.rm = TRUE)
  combined$pseudotime_scaled = NA_real_
  combined$pseudotime_scaled[match(on_path, colnames(combined))] =
    if (diff(rng) > 0) (pv - rng[1]) / diff(rng) else 0

  for (dir in c("up", "down")) {
    top = pt_tab %>%
      dplyr::filter(Path == pth, Direction == dir) %>%
      dplyr::arrange(FDR, dplyr::desc(abs(logFC))) %>%
      dplyr::slice_head(n = 25)
    # hclust on the gene rows needs at least two of them
    if (nrow(top) < 2) next

    labs_wang = ifelse(is.na(top$WangGeneID), top$Gene, top$WangGeneID)
    hm = plot_pseudotime_heatmap(combined, top$Gene, labs_wang, on_path,
                                 pseudotime_column = "pseudotime_scaled")
    # guides = "collect" is what the function sets; rebuilding to change the
    # heights would drop it and clip the pseudotime legend title away
    hp = (hm[[1]] / hm[[2]]) +
      plot_layout(heights = c(0.03, 1), guides = "collect")

    f = paste0(results_dir, "/pseudotime_heatmap_",
               gsub("[^A-Za-z0-9]+", "_", pth), "_", dir, ".png")
    ggsave(hp, file = f, width = 7, height = 5, dpi = 600)

    # the heading starts with a word: pandoc drops everything before the first
    # letter when it builds anchor ids, so "5.1, up" would become "up"
    cat("\n\n### Path ", pth, ", ", dir, "\n\n", sep = "")
    cat("![](", f, ")\n\n", sep = "")
  }
}
```

## GO enrichment of pseudotime-associated genes

GO enrichment for each path and direction, against every gene in the data. Terms
with p < `r params$go_pval` are shown and written to
`pseudotime_GO_pvalues.csv`.

`r if (!have_go) "Skipped: the GO files were not found." else if (!have_pt || nrow(pt_tab) == 0) "Skipped: there are no pseudotime-associated genes to test." else ""`

```{r, eval=have_go && have_pt && nrow(pt_tab) > 0}
path_go = run_go_enrichment(
  gaf              = gaf,
  gene_description = gene_description,
  go_names         = go_names,
  test_genes       = pt_tab %>%
                       dplyr::transmute(Gene, set = paste0(Path, ".", Direction)) %>%
                       dplyr::distinct(),
  background_n     = nrow(combined),
  group2_label     = "Path",
  ngroup2_label    = "nPathGenes",
  overlaps_file    = paste0(results_dir, "/pseudotime_GO_overlaps.csv"),
  pvals_file       = paste0(results_dir, "/pseudotime_GO_pvalues.csv"),
  pval_filter      = params$go_pval)

datatable(path_go %>% dplyr::mutate(dplyr::across(where(is.numeric),
                                                  ~signif(., 3))),
          filter = "top", rownames = FALSE)
```

# Export

```{r}
out = data.frame(barcode    = colnames(merged),
                 node       = as.character(merged$node),
                 cluster    = as.character(merged$final_label),
                 annotation = as.character(merged$annotation),
                 library    = as.character(merged$library),
                 stage      = as.character(merged$stage),
                 pseudotime = merged$pseudotime,
                 stringsAsFactors = FALSE)

write.csv(out, file = paste0(results_dir, "/pseudotime_per_cell.csv"),
          row.names = FALSE)
```

## Everything written by this document

Built by listing the results directory and looking each file up, rather than by
keeping a list by hand. A file with no description, or a description with no
file, is reported below instead of passing unnoticed.

```{r}
file_notes = c(
  "pseudotime_per_cell.csv" = "One row per cell: node, cluster, annotation carried by barcode, library, stage and pseudotime.",
  "cluster_by_annotation.csv" = "How the annotation is distributed across the atlas clusters, before subclustering.",
  "node_composition.csv"      = "Cells per node and the cluster each node came from.",
  "node_by_annotation.csv"    = "Annotation composition of every node, as counts and fractions. The evidence behind panel B of the node graph.",
  "node_marker_set_scores.csv" = "Each node scored on the germ cell marker sets. The root rule is applied to this table.",
  "mst_edges.csv"             = "Every edge of the tree with four reads on it: its length in the corrected space, where each node ranks in the other's neighbor list, the centroid distance in standard deviations of the within-node spread, its length as a percentage of the outgroup threshold, and the percentage of subsampled trees containing it.",
  "annotation_pseudotime.csv" = "Median and quartiles of pseudotime for each annotation. The ordering claim in numbers.",
  "pseudotime_vs_technical.csv"   = "Pseudotime regressed on depth, mitochondrial, rRNA and ribosomal content, pooled and within each library.",
  "node_depth_and_technical.csv" = "Depth, genes detected, percent mitochondrial, rRNA and ribosomal per node.",
  "node_depth_tests.csv"      = "Kruskal-Wallis of each measure across the subclusters of a cluster, with the fold range in median depth.",
  "node_marker_expression.csv" = "The per-gene values behind the node identity bubble plot.",
  "node_marker_genes.csv"     = "Markers of each node against the other nodes of this lineage.",
  "node_marker_GO_overlaps.csv" = "Which marker genes carry each enriched GO term, per node.",
  "node_marker_GO_pvalues.csv" = "GO enrichment of each node's marker genes.",
  "pseudotime_GO_overlaps.csv" = "Which genes carry each enriched GO term, per path and direction.",
  "pseudotime_GO_pvalues.csv" = "GO enrichment of the pseudotime-associated genes, per path and direction.",
  "pseudotime_genes.csv"      = "Genes whose expression changes along each path, one row per gene per path with its direction.",
  "trajectory_umap.png"       = "UMAP colored by annotation and by pseudotime, with the tree drawn over it.",
  "node_graph.png"            = "The tree laid out by its own structure: nodes sized by cell count, colored by mean pseudotime and by most common annotation.",
  "node_graph.pdf"            = "The node graph as a vector file, for figure assembly.",
  "node_by_annotation.png"    = "Annotation composition of each node as stacked bars.",
  "pseudotime_by_annotation.png" = "Pseudotime distribution per annotation and per library.",
  "node_identity_scores.csv"  = "Each node against all marker sets, somatic ones included, with a contained_in flag where one set sits inside another.",
  "pseudotime_by_node.png"    = "Pseudotime distribution per node and per library.",
  "node_depth_and_technical.png" = "Depth and technical content per node.",
  "removal_evidence_bubble.png" = "The sets named by removal_evidence_sets, per node, with any node being removed first on the axis. The evidence behind a removal.",
  "node_identity_scores.png"  = "Marker expression per node, one dot per gene, every set.",
  "node_graph_ranks.png" = "The tree with two arrows per pair, each colored by how the node it points to ranks among the neighbors of the node it leaves.",
  "node_graph_ranks.pdf" = "The rank graph as a vector file.",
  "node_graph_marker_sets.png" = "The tree once per germ marker set, colored by expression relative to that set's maximum and sized by percent of cells detecting it.",
  "node_graph_marker_sets.pdf" = "The marker set panels as a vector file.")

in_dir  = sort(setdiff(list.files(results_dir), "Robjects"))
in_dir  = in_dir[!grepl("^pseudotime_heatmap_", in_dir)]

# The table describes what this document writes. The results directory also
# holds whatever earlier runs left behind, and those are named below rather
# than given rows with no description.
on_disk = intersect(names(file_notes), in_dir)
unknown = setdiff(in_dir, names(file_notes))
missing = setdiff(names(file_notes), in_dir)

export_tab = data.frame(
  file = on_disk,
  size = sapply(file.path(results_dir, on_disk),
                function(f) format(structure(file.size(f), class = "object_size"),
                                   units = "auto")),
  what = unname(file_notes[on_disk]),
  stringsAsFactors = FALSE)

datatable(export_tab, rownames = FALSE, escape = FALSE,
          options = list(pageLength = 30, dom = "tp",
                         columnDefs = list(list(width = "240px", targets = 0))))
```

`r n_heat = length(list.files(results_dir, pattern = "^pseudotime_heatmap_")); if (n_heat > 0) paste0("Plus ", n_heat, " heatmap image(s) named pseudotime_heatmap_<path>_<direction>.png.") else ""`

`r if (length(unknown) > 0) paste0(length(unknown), " file(s) in the directory are not written by this document and are left out of the table: ", paste(unknown, collapse = ", "), ". Either an earlier version wrote them, or they are new and need a line in file_notes.") else ""`

`r if (length(missing) > 0) paste0("Described but not written on this run: ", paste(missing, collapse = ", "), ".") else ""`

`r paste0("Written to ", results_dir, ". Cached intermediates are in Robjects and are not listed.")`

# Summary

```{r}
summary_tbl = data.frame(
  quantity = c("lineage", "clusters", "cells", "libraries kept", "nodes",
               "sub_k", "minimum node size", "root", "paths",
               paste0("gene-path results at FDR < ", params$fdr_cut)),
  value = c(lineage,
            paste(clusters, collapse = ", "),
            format(ncol(merged), big.mark = ","),
            paste(objs$kept_batches, collapse = ", "),
            as.character(nlevels(merged$node)),
            as.character(params$sub_k),
            as.character(params$min_node_cells),
            root,
            as.character(n_paths),
            if (have_pt) format(nrow(pt_tab), big.mark = ",") else "not tested"),
  stringsAsFactors = FALSE)

datatable(summary_tbl, rownames = FALSE, colnames = c("", ""),
          options = list(dom = "t", ordering = FALSE, paging = FALSE))
```

# Session info

```{r}
sessionInfo()
```
