---
title: "Gene expression figures"
author: "Stephanie Hickey"
date: "`r Sys.Date()`"
output:
  html_document:
    code_folding: hide
    toc: true
    toc_float:
      toc_collapsed: true
params:
    project_dir: /mnt/ufs18/rs-013/bioinformaticsCore/projects/chong_davidson/BCC125_sea_lamprey_scRNA
    project_name: BCC125_sea_lamprey_scRNA
    lab: Weiming Li
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

```{r, echo=FALSE, results="asis"}
cat("\n")
cat("#### Lab:", params$lab, " \n")
cat("#### Project:", params$project_name, " \n")
```

```{r}
library(tidyverse)
library(SingleCellExperiment)
library(scater)
library(scran)
library(patchwork)
library(RColorBrewer)
library(readxl)
library(monocle3)
library(batchelor)
library(DT)
library(TSCAN)
library(rtracklayer)
library(BiocSingular)
library(ggforce)
library(ggraph)
library(tidygraph)
library(scatterpie)
library(ggrepel)
source(paste0(params$project_dir, "/src/scRNA_seq_functions.R"))
```

```{r}
results_dir = paste0(params$project_dir, "/results/20260615")
data_dir    = paste0(params$project_dir, "/data")
```

```{r, echo=FALSE}
# load pre-computed objects for rendering
Robject_dir          = paste0(params$project_dir, "/results/20260312/Robjects")
combined.sce         = readRDS(paste0(Robject_dir, "/combined.sce.Rds"))
all.sex.sce          = readRDS(paste0(Robject_dir, "/no.support.sex.sce.Rds"))
pseudo               = readRDS(paste0(Robject_dir, "/no.support.merged.sex.pseudo.Rds"))
scatter_markers      = read.csv(paste0(results_dir, "/scatter_cluster_marker_genes.csv"),row.names = 1)
germ_scatter_markers = read.csv(paste0(results_dir, "/no.support_sex_scatter_cluster_marker_genes.csv"), row.names = 1)
cell_meta            = read.delim(paste0(results_dir, "/cell_metadata.tsv"))
gene_descriptions    = read.csv(paste0(results_dir, "/gene_descritption.csv"))
pseudo_de            = read.csv(paste0(results_dir, "/no.support_sex_pseudotime_de_genes.csv"), row.names = 1)
```

### Build combined (uncorrected) sex cell SCE

```{r}
all.sex.dec  <- lapply(all.sex.sce, modelGeneVar)

universe     <- Reduce(intersect, lapply(all.sex.sce, rownames))
all.sex.sce2 <- lapply(all.sex.sce, "[", i = universe,)
all.sex.dec2 <- lapply(all.sex.dec, "[", i = universe,)

all.sex.sce2[["TM_C00004374"]]@colData$cluster = "noclust"

combined.sex.sce <- do.call(cbind, all.sex.sce2)
```

### Add meta data

```{r}
# set coldata
rownames(cell_meta) = cell_meta$barcode
cell_meta_ordered <- cell_meta[colnames(combined.sce), , drop = FALSE]
colData(combined.sce) <- DataFrame(cell_meta_ordered)

germ_cell_meta = 
  cell_meta %>%
  filter(!is.na(germ.cluster)) 

germ_cell_meta_ordered <- germ_cell_meta[colnames(combined.sex.sce), , drop = FALSE]
colData(combined.sex.sce) <- DataFrame(germ_cell_meta_ordered)


# set rowdata
rownames(gene_descriptions) = gene_descriptions$Gene
gene_descriptions_ordered = gene_descriptions[rownames(combined.sce), , drop = FALSE]
rowData(combined.sce) = DataFrame(gene_descriptions_ordered)
rowData(combined.sex.sce) = DataFrame(gene_descriptions_ordered)
```

### Set cluster annotations and pallettes

All cells
```{r}
anno = c(
  "1. Fibroblasts (migrating)",
  "2. Mesonephric cells",
  "3. Granulosa cells (migrating)",
  "4. Granulosa cells",
  "6. Endothelial cells",
  "8. Erythrocytes",
  "9. Pre-granulosa cells (migrating)",
  "10. Leukocytes",
  "11. Hemopoietic stem cells",
  "12. Podocytes",
  "13. Macrophages",
  "14. Pre-granulosa cells (mitotic)",
  "15. Immune cells (migrating, mitotic)",
  "7. Primordial germ cells (migrating)",
  "5. Oogonia/Oocytes",
  "17. Male germ cells (migrating)",
  "16. Spermatocytes (cytokinetic)",
  "18. Spermatocytes (apoptotic)",
  "19. Spermatocytes (GPCR-activated)"
)

level_order = c(1:4, 6, 8:15, 7, 5, 17, 16, 18, 19)
combined.sce$label <- factor(combined.sce$label, levels = level_order)

anno_order = c(
  "1. Fibroblasts (migrating)",
  "2. Mesonephric cells",
  "3. Granulosa cells (migrating)",
  "4. Granulosa cells",
  "5. Oogonia/Oocytes",
  "6. Endothelial cells",
  "7. Primordial germ cells (migrating)",
  "8. Erythrocytes",
  "9. Pre-granulosa cells (migrating)",
  "10. Leukocytes",
  "11. Hemopoietic stem cells",
  "12. Podocytes",
  "13. Macrophages",
  "14. Pre-granulosa cells (mitotic)",
  "15. Immune cells (migrating, mitotic)",
  "16. Spermatocytes (cytokinetic)",
  "17. Male germ cells (migrating)",
  "18. Spermatocytes (apoptotic)",
  "19. Spermatocytes (GPCR-activated)"
)

combined.sce$CellType <- anno_order[as.integer(as.character(combined.sce$label))]

batch_map <- c(
  "TF_C00004337" = "Female",
  "TM_C00004374" = "Male",
  "UND1_C00004336" = "UND1",
  "UND2_C00004362" = "UND2"
)

combined.sce$Sample <- batch_map[combined.sce$batch]

# Define a visible light blue and a deep dark blue from the Blues palette
start_blue <- brewer.pal(9, "Blues")[3] # #9ECAE1 (skips the two whitest shades)
end_blue   <- brewer.pal(9, "Blues")[9] # #08306B

# Generate 12 sequential shades starting from the darker base
color1_shades <- colorRampPalette(c(start_blue, end_blue))(13)

# Generate 6 distinct shades of Oranges
# Define custom boundaries for Orange to completely block the brown tones
start_org     <- "#FDBE85" # A highly visible, crisp light orange
end_org       <- "#A62B00" # A rich, intense burnt orange/terracotta (replaces the brown)

# Generate your 7 distinct, non-brown orange shades
color2_shades <- colorRampPalette(c(start_org, end_org))(6)


# Merge into your publication-ready master palette
publication_palette <- c(color1_shades, color2_shades)
publication_palette  <- setNames(publication_palette, anno)
```

Germ cells 

```{r}

# broad annotations (15 sex cell clusters)
germ_anno = c(
  "PGC",
  "Cyst.epithelial.cells",
  "FGC.migrating",
  "Oogonia.stem.cells",
  "Oogonia",
  "Oocytes",
  "MGC.SPC.group1",
  "MGC.SPC.group2",
  "MGC.SPC.group3",
  "SPC.cytokinetic"
)

combined.sex.sce$germ.broad.CellType <- factor(combined.sex.sce$germ.broad.CellType, levels = germ_anno)

# 1. Family One: 2 Shades (Gold/Amber)
# Custom anchors to guarantee a clean, high-contrast 2-step jump
color1_shades <- c("#FEC44F", "#D95F02")

# 2. Family Two: 4 Shades (Blues - skipping white/light tones)
start_blue    <- brewer.pal(9, "Blues")[4] # Clear light-medium blue
end_blue      <- brewer.pal(9, "Blues")[8] # Rich dark blue
color2_shades <- colorRampPalette(c(start_blue, end_blue))(4)

# 3. Family Three: 4 Shades (Greens - skipping white/light tones)
start_green   <- brewer.pal(9, "Greens")[4] # Vibrant light-medium green
end_green     <- brewer.pal(9, "Greens")[8] # Deep forest green
color3_shades <- colorRampPalette(c(start_green, end_green))(4)

# Combine into your final 10-color publication palette
three_group_palette <- c(color1_shades, color3_shades, color2_shades)
three_group_palette  <- setNames(three_group_palette, germ_anno)
```

### Fig.1: All cell tSNE

```{r, fig.width=8, fig.height=5}
level_order = c(1:4, 6, 8:15, 7, 5, 17, 16, 18, 19)
cell_meta$label = factor(cell_meta$label, levels = level_order)

cell_meta <- cell_meta %>%
  mutate(CellType = anno[as.integer(label)])  

plot_tsne(df           = cell_meta,
          color_col    = "CellType",
          label_col    = "label",
          palette      = publication_palette,
          color_levels = anno,
          point_size   = 0.5,
          text_color   = "black",
          bg_color     = "white",
          text_size    = 2)

ggsave(paste0(results_dir, "/Fig1_tsne.png"), width = 8, height = 5)
```

### Fig. 10: WNT Genes in germ cell clusters

```{r, fig.width=4, fig.height=4}
goi = c("WNT5B", "WNT5A-1", "WNT5B.1")

plot_violin_goi_named(sce              = combined.sex.sce,
                      markers          = germ_scatter_markers, 
                      goi              = goi, 
                      gene_description = gene_description, 
                      anno             = germ_anno,
                      palette          = three_group_palette,
                      cluster_col      = "germ.broad.CellType",
                      marker_col       = "cluster",
                      title            = NULL
                      )

ggsave(paste0(results_dir, "/Fig10_WNT_germ_cell_clusters.png"), width = 4, height = 4)
```

### Fig.12 Wnt5 pseudotime

```{r, fig.width=5, fig.height=5.5}
pseudo_mat  <- pathStat(pseudo$ordering)
goi = c("WNT5A-1", "WNT5B", "WNT5B.1", "WNT5B.2", "WNT5B.3")

plot_pseudotime_paths(sce = combined.sex.sce, 
                      pseudo_mat = pseudo_mat, 
                      goi = goi, 
                      gene_description = gene_descriptions, 
                      paths = c("Oocytes", "SPC.cytokinetic"),
                      tradeseq_res = pseudo_de,
                      title = NULL)

ggsave(paste0(results_dir, "/Fig12_WNT_pseudotime.png"), width = 5, height = 5.5)
```

### Fig.14 Chr 81 genes 

```{r, fig.width=7.5, fig.height=8}
id_oi = c("LOC116957683",
          "LOC116957684",
          "LOC116957694",
          "LOC116957681",
          "LOC116957685",
          "LOC116957682",
          "LOC116957686",
          "LOC116957687",
          "LOC116957688",
          "LOC116957695",
          "LOC116957689",
          "LOC116957690",
          "LOC116957691",
          "LOC116957693",
          "LOC116957692")

goi = 
  gene_descriptions %>%
  filter(Gene %in% id_oi) %>%
  pull(WangGeneID)

plot_violin_goi(sce              = combined.sce,
                markers          = scatter_markers,
                goi              = goi,
                gene_description = gene_descriptions,
                anno             = anno,
                palette          = publication_palette,
                label_col        = "label",
                level_order      = c(1:4, 6, 8:15, 7, 5, 17, 16, 18, 19),
                title            = NULL,
                ncol             = 2)

ggsave(paste0(results_dir, "/Fig14_Chr81_genes.png"), width = 7.5, height = 8)
```

### Fig. S6: CUDZ1 genes

```{r, fig.width=7.5, fig.height=8}
cuzd1_genes <- c(
  "CUZD1", "CUZD1.1", "CUZD1.10", "CUZD1.11", "CUZD1.12",
  "CUZD1.13", "CUZD1.14", "CUZD1.15", "CUZD1.16", "CUZD1.2",
  "CUZD1.3", "CUZD1.4", "CUZD1.5", "CUZD1.6", "CUZD1.7",
  "CUZD1.8", "CUZD1.9"
)

plot_violin_goi(sce              = combined.sce,
                markers          = scatter_markers,
                goi              = cuzd1_genes,
                gene_description = gene_descriptions,
                anno             = anno,
                palette          = publication_palette,
                label_col        = "label",
                level_order      = c(1:4, 6, 8:15, 7, 5, 17, 16, 18, 19),
                title            = NULL,
                ncol             = 2)

ggsave(paste0(results_dir, "/S6_CUDZ1_genes.png"), width = 7.5, height = 8)
```

### S7: DDX4

```{r, fig.width=5, fig.height=3}
plot_violin_goi_single(sce              = combined.sce,
                       markers          = scatter_markers,
                       goi              = "DDX4",
                       gene_description = gene_descriptions,
                       anno             = anno,
                       palette          = publication_palette,
                       title            = "DDX4",
                       level_order = c(1:4, 6, 8:15, 7, 5, 17, 16, 18, 19))

  
ggsave(paste0(results_dir, "/S7_DDX4.png"), width = 5, height = 3)
```

### S8: PIWIL1

```{r, fig.width=5, fig.height=3}
plot_violin_goi_single(sce              = combined.sce,
                       markers          = scatter_markers,
                       goi              = "PIWIL1",
                       gene_description = gene_descriptions,
                       anno             = anno,
                       palette          = publication_palette,
                       title            = "PIWIL1",
                       level_order = c(1:4, 6, 8:15, 7, 5, 17, 16, 18, 19))

  
ggsave(paste0(results_dir, "/S8_PIWIL1.png"), width = 5, height = 3)
```

### S9: SYCP3

```{r, fig.width=5, fig.height=3}
plot_violin_goi_single(sce              = combined.sce,
                       markers          = scatter_markers,
                       goi              = "SYCP3",
                       gene_description = gene_descriptions,
                       anno             = anno,
                       palette          = publication_palette,
                       title            = "SYCP3",
                       level_order = c(1:4, 6, 8:15, 7, 5, 17, 16, 18, 19))

  
ggsave(paste0(results_dir, "/S9_SYCP3.png"), width = 5, height = 3)
```

### S11

```{r, fig.width=7.5, fig.height=5}
fig_s11_genes <- c(
  "S100A1", "CXCR4", "RBPMS", "BOLL", "FOXL2", "SOX18"
)

plot_violin_goi(sce              = combined.sce,
                markers          = scatter_markers,
                goi              = fig_s11_genes,
                gene_description = gene_descriptions,
                anno             = anno,
                palette          = publication_palette,
                label_col        = "label",
                level_order      = c(1:4, 6, 8:15, 7, 5, 17, 16, 18, 19),
                title            = NULL,
                ncol             = 2)

ggsave(paste0(results_dir, "/S11_genes.png"), width = 7.5, height = 5)
```
### FOXL2 percentage

```{r, fig.width=7, fig.height=7}
plot_expr_proportion( combined.sce,
                      gene     = "LOC116946948",
                      gene_symbol    = "FOXL2",
                      celltype_col   = "CellType",
                      batch_col      = "Sample",
                      assay_name     = "counts",
                      threshold      = 0,
                      ncol           = 3,
                      celltype_order = anno_order)

ggsave(paste0(results_dir, "/FOXL2_percent.png"), width = 7, height = 7)
```


### S16: WNT5 genes all cells

```{r, fig.width=7.5, fig.height=5}
wnt5_genes <- c(
  "WNT5B", "WNT5B.2", "WNT5B.3", "WNT5B.1", "WNT5A-1"
)


plot_violin_goi(sce              = combined.sce,
                markers          = scatter_markers,
                goi              = wnt5_genes,
                gene_description = gene_descriptions,
                anno             = anno,
                palette          = publication_palette,
                label_col        = "label",
                level_order      = c(1:4, 6, 8:15, 7, 5, 17, 16, 18, 19),
                title            = NULL,
                ncol             = 2)

ggsave(paste0(results_dir, "/S16_WTN5_all_cells_genes.png"), width = 7.5, height = 5)
```

### S17: Immune/development genes

```{r, fig.width=7.5, fig.height=8}
immune_dev_genes <- c(
  "PFN2.1", "MMP19", "IQCG", "IGF1", "ALOX5", "IQUB",
  "STC2", "RPA1", "IGF1R", "MNS1", "RPA1.1", "PFN2", "IGFBP5"
)

plot_violin_goi(sce              = combined.sce,
                markers          = scatter_markers,
                goi              = immune_dev_genes,
                gene_description = gene_descriptions,
                anno             = anno,
                palette          = publication_palette,
                label_col        = "label",
                level_order      = c(1:4, 6, 8:15, 7, 5, 17, 16, 18, 19),
                title            = NULL,
                ncol             = 2)

ggsave(paste0(results_dir, "/S17_immune_genes.png"), width = 7.5, height = 8)
```

### S18: growth factors

```{r, fig.width=7.5, fig.height=6}
growth_factor_genes <- c(
  "IGF1R", "IGF1R.1", "GNRH2", "IGF1-1", "IGF1", "GNRHR",
  "GNRHR2", "GNRHR.1", "GNRH2.1", "IGF1R.2"
)

plot_violin_goi(sce              = combined.sce,
                markers          = scatter_markers,
                goi              = growth_factor_genes,
                gene_description = gene_descriptions,
                anno             = anno,
                palette          = publication_palette,
                label_col        = "label",
                level_order      = c(1:4, 6, 8:15, 7, 5, 17, 16, 18, 19),
                title            = NULL,
                ncol             = 2)

ggsave(paste0(results_dir, "/S18_growth_factor_genes.png"), width = 7.5, height = 6)
```

### S19: TGFB genes

```{r, fig.width=7.5, fig.height=8}
tgfb_genes <- c(
  "TGFBR2", "SMAD3", "SMAD6", "BMPR2", "BRINP1", "BMP2",
  "SNIP1", "BRINP3", "TGFB2", "BMPER", "KCP", "BMP2K"
)

plot_violin_goi(sce              = combined.sce,
                markers          = scatter_markers,
                goi              = tgfb_genes,
                gene_description = gene_descriptions,
                anno             = anno,
                palette          = publication_palette,
                label_col        = "label",
                level_order      = c(1:4, 6, 8:15, 7, 5, 17, 16, 18, 19),
                title            = NULL,
                ncol             = 2)

ggsave(paste0(results_dir, "/S19_tgfb_genes.png"), width = 7.5, height = 8)
```

### S20: Sex steriod genes

```{r, fig.width=7.5, fig.height=8}
sex_steroid_genes <- c(
  "RERG", "HSD3B7", "ESRRG", "RERG-1", "CYP17A1", "ESR2.1",
  "BCA+E7903R3", "MEDAG", "BCAR3", "ESR2", "NR5A1", "ESR2.2",
  "BCAR1-1", "ESRRG-2", "ESRRG-1"
)

plot_violin_goi(sce              = combined.sce,
                markers          = scatter_markers,
                goi              = sex_steroid_genes,
                gene_description = gene_descriptions,
                anno             = anno,
                palette          = publication_palette,
                label_col        = "label",
                level_order      = c(1:4, 6, 8:15, 7, 5, 17, 16, 18, 19),
                title            = NULL,
                ncol             = 2)

ggsave(paste0(results_dir, "/S20_sex_steroid_genes.png"), width = 7.5, height = 8)
```

### S21: percent cells expressing SOX9

```{r, fig.width=7, fig.height=7}

plot_expr_proportion( combined.sce,
                      gene = "LOC116951502",
                      gene_symbol  = "SOX9",
                      celltype_col = "CellType",
                      batch_col    = "Sample",
                      assay_name   = "counts",
                      threshold    = 0,
                      ncol         = 3,
                      celltype_order = anno_order)

ggsave(paste0(results_dir, "/S21_SOX9_percent.png"), width = 7, height = 7)

plot_expr_proportion( combined.sce,
                      gene     = "LOC103091769",
                      gene_symbol    = "sox9-a",
                      celltype_col   = "CellType",
                      batch_col      = "Sample",
                      assay_name     = "counts",
                      threshold      = 0,
                      ncol           = 3,
                      celltype_order = anno_order)

ggsave(paste0(results_dir, "/S21_sox9-a_percent.png"), width = 7, height = 7)
```


```{r, fig.width=5, fig.height=3}
plot_violin_goi_single(sce              = combined.sce,
                       markers          = scatter_markers,
                       goi              = "sox9-a",
                       gene_description = gene_descriptions,
                       anno             = anno,
                       palette          = publication_palette,
                       title            = "sox9-a",
                       level_order = c(1:4, 6, 8:15, 7, 5, 17, 16, 18, 19))

ggsave(paste0(results_dir, "/S21_sox9-a_violin.png"), width = 5, height = 3)

plot_violin_goi_single(sce              = combined.sce,
                       markers          = scatter_markers,
                       goi              = "SOX9",
                       gene_description = gene_descriptions,
                       anno             = anno,
                       palette          = publication_palette,
                       title            = "SOX9",
                       level_order = c(1:4, 6, 8:15, 7, 5, 17, 16, 18, 19))

ggsave(paste0(results_dir, "/S21_SOX9_violin.png"), width = 5, height = 3)

```

### S22: Germ cell fate genes

```{r, fig.width=7, fig.height=5}
germ_fate_genes <- c(
  "CDK9", "SRCAP", "PRDM1", "PRDM1.1"
)

plot_violin_goi(sce              = combined.sce,
                markers          = scatter_markers,
                goi              = germ_fate_genes,
                gene_description = gene_descriptions,
                anno             = anno,
                palette          = publication_palette,
                label_col        = "label",
                level_order      = c(1:4, 6, 8:15, 7, 5, 17, 16, 18, 19),
                title            = NULL,
                ncol             = 1)

ggsave(paste0(results_dir, "/S22_germ_cell_fate_genes.png"), width = 7, height = 5)
```
