---
title: "Trajectory and pseudotime analysis"
author: "Stephanie Hickey"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: true
    toc_float:
      toc_collapsed: true
params:
    project_dir: /mnt/ufs18/rs-013/bioinformaticsCore/projects/chong_davidson/BCC125_sea_lamprey_scRNA
    project_name: BCC125_sea_lamprey_scRNA
    lab: Weiming Li
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
```

```{r, echo=FALSE, results="asis"}
## parameterized text
cat("\n")
cat("#### Lab:", params$lab, " \n")
cat("#### Project:", params$project_name, " \n")
```

```{r}
# Set up
library(tidyverse)
#library(Matrix)
library(DT)
library(patchwork)
library(monocle3)
source(paste0(params$project_dir, "/src/scRNA_seq_functions.R"))
set.seed(1)
```

```{r}
# Set Paths
results_dir = paste0(params$project_dir, "/results")
# results_dir = "/mnt/ufs18/rs-013/bioinformaticsCore/projects/chong_davidson/BCC125_sea_lamprey_scRNA/results"
data_dir = paste0(params$project_dir, "/data")
# data_dir = "/mnt/ufs18/rs-013/bioinformaticsCore/projects/chong_davidson/BCC125_sea_lamprey_scRNA/data"
```

```{r}
# load seurat ob
cds_nb = readRDS(paste0(results_dir, "/cds_nb.Rds"))
```

## Plot clusters

UMAP plot labeled by cluster.

```{r}
plot_cells(cds_nb,
           show_trajectory_graph = FALSE,
           label_branch_points = FALSE,
           label_roots = FALSE,
           label_leaves = FALSE)

```

## Plot genes of interest

### TSNE plots

```{r, fig.width=18, fig.height=12}
goi = c("BOLL", "CCDC189", "CFAP69", "NANOS1", "PIWIL1", "SPAG5", "SPAG16", "SPATA22", "TEX11", "WNT11","LOC116937338", "LOC116937873", "LOC116937887", "LOC116938414", "LOC116940726", "LOC116941541", "LOC116944806", "LOC116948775")

plot_cells(cds_nb, genes=goi,
           show_trajectory_graph=FALSE,
           label_cell_groups=FALSE,
           label_leaves=FALSE,
           label_branch_points = FALSE)
```

### Violin plots

```{r, warning=FALSE, fig.height=40, fig.width=10}
cds_subset <- cds_nb[rowData(cds_nb)$gene_short_name %in% goi,]
plot_genes_violin(cds_subset, ncol=1, group_cells_by="cluster")
```


# Select female AND male germ cell development clusters

Cluster 18 -> 13 -> 15 (female germ cell development)
Cluster 18 -> 14 -> 12 -> 2 -> 29 (male germ cell development)

Here, I'll select all of the female and male germ cell development clusters and recluster.

```{r, eval=FALSE}
meta_df = as.data.frame(cds_nb@colData)
coi = 
  meta_df %>%
  filter(cluster %in% c(2, 12, 13, 14, 15, 18, 29)) %>%
  pull(barcode)
  
cds_coi = cds_nb[,coi]

cds_coi@colData$orig_cluster = cds_coi@colData$cluster

# recluster
cds_coi <- preprocess_cds(cds_coi, num_dim = 100)
cds_coi <- align_cds(cds_coi, alignment_group = "sample")
cds_coi <- reduce_dimension(cds_coi)
```

```{r}
cds_coi = readRDS(file = paste0(results_dir, "/cds_female_male_clusters.Rds"))
```

## Plot cells with original cluster labels

```{r}
plot_cells(cds_coi,
           color_cells_by = "orig_cluster",
           group_cells_by = "orig_cluster",
           show_trajectory_graph = FALSE,
           label_branch_points = FALSE,
           label_roots = FALSE,
           label_leaves = FALSE)
```

## Plot cells with new cluster labels

```{r, eval=FALSE}
cds_coi <- cluster_cells(cds_coi, resolution=5e-5)
```

```{r}
plot_cells(cds_coi,
           show_trajectory_graph = FALSE,
           label_branch_points = FALSE,
           label_roots = FALSE,
           label_leaves = FALSE)
```

Combine clusters 1 and 2

```{r}
colData(cds_coi)$cluster = clusters(cds_coi)
meta_df = as.data.frame(cds_coi@colData)

meta_df =
  meta_df %>%
  mutate(combine_clusters = case_when(cluster %in% c(1,2) ~ "1_2",
                                      !cluster %in% c(1,2) ~ cluster))

colData(cds_coi)$combine_clusters = meta_df$combine_clusters

plot_cells(cds_coi,
           color_cells_by = "combine_clusters",
           group_cells_by = "combine_clusters",
           show_trajectory_graph = FALSE,
           label_branch_points = FALSE,
           label_roots = FALSE,
           label_leaves = FALSE)

```

## Plot cells by sample

```{r}
plot_cells(cds_coi, 
           color_cells_by="sample", 
           label_cell_groups =FALSE,
           show_trajectory_graph = FALSE,
           label_branch_points = FALSE,
           label_roots = FALSE,
           label_leaves = FALSE) + facet_wrap(~sample, nrow = 2, ncol=2)
```

## Plot genes of interest

### TSNE plots

```{r, fig.width=12, fig.height=8}
goi = c("BOLL", "CCDC189", "CFAP69", "NANOS1", "PIWIL1", "SPAG5", "SPAG16", "SPATA22", "TEX11", "WNT11","LOC116937338", "LOC116937873", "LOC116937887", "LOC116938414", "LOC116940726", "LOC116941541", "LOC116944806", "LOC116948775")

plot_cells(cds_coi, genes=goi,
           show_trajectory_graph=FALSE,
           label_cell_groups=FALSE,
           label_leaves=FALSE,
           label_branch_points = FALSE)
```

### Violin plots

```{r, warning=FALSE, fig.height=20, fig.width=5}
cds_subset <- cds_coi[rowData(cds_coi)$gene_short_name %in% goi,]
plot_genes_violin(cds_subset, ncol=1, group_cells_by="cluster")
```


## Plot known cell-type markers

In each cell, the expression of genes within each group are summed. In the plots, higher expression is a brighter color. Expression is scaled to percent of maximum expression.

Use known markers unique to each cell type:

```{r}
# load known marker genes
library(readr)
chung_davidson_cell_type_markers <- read_csv(paste0(data_dir, "/chung_davidson_cell_type_markers.csv"), show_col_types = FALSE)

chung_davidson_cell_type_markers = 
  chung_davidson_cell_type_markers %>%
  select(-GeneName)

chung_davidson_unique = 
  chung_davidson_cell_type_markers %>%
  group_by(Gene) %>%
  tally() %>%
  filter(n == 1) %>%
  select(Gene) %>%
  left_join(chung_davidson_cell_type_markers, by = "Gene")

datatable(chung_davidson_unique, filter = "top", rownames = FALSE)
```

Plot early cell types

```{r, fig.width=12, fig.height=8}
rowData(cds_coi)$gene_short_name = row.names(cds_coi)
plot_cells(cds_coi, genes=chung_davidson_unique %>% 
             filter(Group %in% c("Germline cells during epigenetic reprogramming", 
                                 "Germline stem cells",
                                 "Embryonic stem cells",
                                 "Germline progenitor and amnionic cells",
                                 "Primordial Germ Cells (PGC",
                                 "Migrating germ cells",
                                 "Late PGC (meiotic entry genes")),
           show_trajectory_graph=FALSE,
           label_cell_groups=FALSE,
           label_leaves=FALSE,
           label_branch_points = FALSE)
```

Plot female cell types

```{r, fig.width=12, fig.height=4}
plot_cells(cds_coi, genes=chung_davidson_unique %>% 
             filter(Group %in% c("Female Germ cell", 
                                 "Mitotic oocyte",
                                 "Granulosa cell",
                                 "Oocyte")),
           show_trajectory_graph=FALSE,
           label_cell_groups=FALSE,
           label_leaves=FALSE,
           label_branch_points = FALSE)
```

Plot male cell types

```{r, fig.width=12, fig.height=8}
plot_cells(cds_coi, genes=chung_davidson_unique %>% 
             filter(Group %in% c("Male germ cells", 
                                 "Sertoli cell",
                                 "Mitotic Spermatogonia",
                                 "Meiotic Spermatogonia",
                                 "Spermatogonia",
                                 "Primary spermatocyte")),
           show_trajectory_graph=FALSE,
           label_cell_groups=FALSE,
           label_leaves=FALSE,
           label_branch_points = FALSE)
```


## Identify new cluster enriched genes

This table contains the top ~100 cluster enriched genes for each cluster. It is saved in `female_male_cluster_marker_genes.csv`. The markers can help us map a cluster to a cell type.

**Column descriptions**

* `gene_id` vector of gene names    
* `cell_group` character vector of the cell group to which the cell belongs   
* `marker_score` numeric vector of marker scores as the fraction expressing scaled by the specificity. The value ranges from 0 to 1.    
* `mean_expression` numeric vector of mean normalized expression of the gene in the cell group    
* `fraction_expressing` numeric vector of fraction of cells expressing the gene within the cell group   
* `specificity` numeric vector of a measure of how specific the gene's expression is to the cell group based on the Jensen-Shannon divergence. The value ranges from 0 to 1.   
* `pseudo_R2` numeric vector of pseudo R-squared values, a measure of how well the gene expression model fits the categorical data relative to the null model. The value ranges from 0 to 1.   
* `marker_test_p_value` numeric vector of likelihood ratio p-values   
* `marker_test_q_value` numeric vector of likelihood ratio q-values   

```{r, results='hide'}
cluster_marker_res <- top_markers(cds_coi, 
                                  genes_to_test_per_group = 100,
                                  group_cells_by="combine_clusters", cores=8)

write.csv(cluster_marker_res, file = paste0(results_dir, "/female_male_cluster_marker_genes.csv"), row.names = FALSE) 
```

```{r}
library(DT)
datatable(cluster_marker_res, filter = "top", rownames = FALSE)
```

## Construct single-cell trajectories

The black line shows the trajectory graph. It connects cell clusters or individual cells in the reduced space, ensuring smooth paths through the data structure. The graph models how cells progress through different states or fates based on gene expression similarity.

```{r, results='hide', eval=FALSE}
cds_coi <- learn_graph(cds_coi, use_partition = FALSE)
saveRDS(cds_coi, file = paste0(results_dir, "/cds_female_male_clusters.Rds"))
```

```{r}
plot_cells(cds_coi,
           color_cells_by = "combine_clusters",
           label_groups_by_cluster=TRUE,
           label_principal_points = FALSE,
           label_branch_points = FALSE,
           label_roots = FALSE,
           label_leaves = FALSE,
           show_trajectory_graph=TRUE)
```

I don't think this graph makes sense based on the cell type annotations provided in your email. I'll separate male and female clusters.

# Select female developmental trajectory clusters

8 -> 4 -> 6

```{r, eval=FALSE}
female = 
  meta_df %>%
  filter(cluster %in% c(4,6,8),
         sample != "TM_C00004374") %>%
  pull(barcode)
  
cds_female = cds_coi[,female]

# re-normalize
cds_female <- preprocess_cds(cds_female, num_dim = 100)
cds_female <- align_cds(cds_female, alignment_group = "sample")
cds_female <- reduce_dimension(cds_female)
```
```{r}
cds_female = readRDS(file = paste0(results_dir, "/cds_female_clusters.Rds"))
```

## Plot cells in TSNE space

```{r}
plot_cells(cds_female,
           show_trajectory_graph = FALSE,
           label_branch_points = FALSE,
           color_cells_by = "combine_clusters",
           group_cells_by = "combine_clusters",
           label_roots = FALSE,
           label_leaves = FALSE,
           cell_size = .7)

```

## Separate by sample

```{r}
plot_cells(cds_female, 
           color_cells_by="sample", 
           label_cell_groups =FALSE,
           show_trajectory_graph = FALSE,
           label_branch_points = FALSE,
           label_roots = FALSE,
           label_leaves = FALSE) + facet_wrap(~sample, nrow = 2, ncol=2)
```

```{r, eval=FALSE}
# set old clusters as new clusters
cds_female <- cluster_cells(cds_female, resolution=7e-4)
female_clusters = cds_female@colData$combine_clusters
names(female_clusters) = cds_female@colData$barcode
female_clusters = as.factor(female_clusters)

cds_female@clusters$UMAP$clusters = female_clusters
```

## Plot genes of interest

### TSNE plots

```{r, fig.width=18, fig.height=12}

goi = c("BOLL", "CCDC189", "CFAP69", "NANOS1", "PIWIL1", "SPAG5", "SPAG16", "SPATA22", "TEX11", "WNT11","LOC116937338", "LOC116937873", "LOC116937887", "LOC116938414", "LOC116940726", "LOC116941541", "LOC116944806", "LOC116948775")

plot_cells(cds_female, genes=goi,
           show_trajectory_graph=FALSE,
           label_cell_groups=FALSE,
           label_leaves=FALSE,
           label_branch_points = FALSE)
```

### Violin plots

```{r, warning=FALSE, fig.height=20, fig.width=3}
cds_subset <- cds_female[rowData(cds_female)$gene_short_name %in% goi,]
plot_genes_violin(cds_subset, ncol=1, group_cells_by="cluster")
```

## Plot known cell-type markers

Plot early cell types

```{r, fig.width=12, fig.height=8}
rowData(cds_female)$gene_short_name = row.names(cds_female)
plot_cells(cds_female, genes=chung_davidson_unique %>% 
             filter(Group %in% c("Germline cells during epigenetic reprogramming", 
                                 "Germline stem cells",
                                 "Embryonic stem cells",
                                 "Germline progenitor and amnionic cells",
                                 "Primordial Germ Cells (PGC",
                                 "Migrating germ cells",
                                 "Late PGC (meiotic entry genes")),
           show_trajectory_graph=FALSE,
           label_cell_groups=FALSE,
           label_leaves=FALSE,
           label_branch_points = FALSE,
           cell_size = .7)
```

Plot female cell types

```{r, fig.width=12, fig.height=4}
plot_cells(cds_female, genes=chung_davidson_unique %>% 
             filter(Group %in% c("Female Germ cell", 
                                 "Mitotic oocyte",
                                 "Granulosa cell",
                                 "Oocyte")),
           show_trajectory_graph=FALSE,
           label_cell_groups=FALSE,
           label_leaves=FALSE,
           label_branch_points = FALSE,
           cell_size = .7)
```

Plot male cell types

```{r, fig.width=12, fig.height=8}
plot_cells(cds_female, genes=chung_davidson_unique %>% 
             filter(Group %in% c("Male germ cells", 
                                 "Sertoli cell",
                                 "Mitotic Spermatogonia",
                                 "Meiotic Spermatogonia",
                                 "Spermatogonia",
                                 "Primary spermatocyte")),
           show_trajectory_graph=FALSE,
           label_cell_groups=FALSE,
           label_leaves=FALSE,
           label_branch_points = FALSE,
           cell_size = .7)
```

## Construct single-cell trajectories

The black line shows the trajectory graph. It connects cell clusters or individual cells in the reduced space, ensuring smooth paths through the data structure. The graph models how cells progress through different states or fates based on gene expression similarity.

```{r, results='hide', eval=FALSE}
cds_female <- learn_graph(cds_female, use_partition = FALSE, close_loop = FALSE)
saveRDS(cds_female, file = paste0(results_dir, "/cds_female_clusters.Rds"))
```

```{r}
plot_cells(cds_female,
           color_cells_by = "combine_clusters",
           group_cells_by = "combine_clusters",
           label_groups_by_cluster=TRUE,
           label_principal_points = FALSE,
           label_branch_points = FALSE,
           label_roots = FALSE,
           label_leaves = FALSE,
           show_trajectory_graph=TRUE,
           cell_size = .7,
            alpha = .3)
```

Principal points approximate key positions along the inferred trajectory and serve as summarized representations of the cells' progression through pseudotime. To define pseudotime, we need to choose 1 or more principal points that to represent the start of the trajectory. 

```{r}
plot_cells(cds_female,
           color_cells_by = "combine_clusters",
           label_groups_by_cluster=FALSE,
           label_principal_points = TRUE,
           show_trajectory_graph=TRUE,
           alpha = .1,
           cell_size = .7)
```

## Find genes that change along the trajectory 

Monocle3 introduces a new approach for finding genes that are differentially expressed across a single-cell trajectory that draws on a powerful technique in spatial correlation analysis, the Moran’s I test. Moran’s I is a measure of multi-directional and multi-dimensional spatial autocorrelation. The statistic tells you whether cells at nearby positions on a trajectory will have similar (or dissimilar) expression levels for the gene being tested. Although both Pearson correlation and Moran’s I ranges from -1 to 1, the interpretation of Moran’s I is slightly different: +1 means that nearby cells will have perfectly similar expression; 0 represents no correlation, and -1 means that neighboring cells will be anti-correlated.

```{r, results='hide', eval=FALSE}
female_traj_genes <- graph_test(cds_female, neighbor_graph="principal_graph", cores=10)

female_traj_genes$Gene = rownames(female_traj_genes)

saveRDS(female_traj_genes, file = paste0(results_dir, "/female_traj_genes.Rds"))
```

```{r}
female_traj_genes = readRDS(file = paste0(results_dir, "/female_traj_genes.Rds"))

datatable(female_traj_genes %>% 
            arrange(desc(morans_I)) %>% 
            filter(q_value < .01) %>%
            select(-status, -morans_test_statistic, -gene_short_name) %>%
            select(morans_I, p_value, q_value),
          filter = "top")
```

**Columns**

* `morans_I`: If you'd like to rank the genes by effect size, sort this table by the `morans_I` column, which ranges from -1 to +1. A value of 0 indicates no effect, while +1 indicates perfect positive autocorrelation and suggests that nearby cells have very similar values of a gene's expression.    
* `p_value`: p-value from the Moran's I test
* `q_value`: q-value from the Moran's I test (adjusted for multiple comparisons)

## Group the genes into modules

Grouping genes into modules helps us understand which genes have similar expression patterns across the trajectory.

```{r, fig.height=6, eval=FALSE}
deg_ids <- row.names(subset(female_traj_genes, q_value < 0.01))

gene_module_df <- find_gene_modules(cds_female[deg_ids,], resolution=c(10^seq(-20,-4)), random_seed=1)

write.csv(gene_module_df, file = paste0(results_dir, "/female_gene_module_df.csv"))
```

```{r, fig.height=6}
gene_module_df = read.csv(file = paste0(results_dir, "/female_gene_module_df.csv"), row.names = 1)

cell_group_df <- tibble::tibble(cell=row.names(colData(cds_female)), 
                                cell_group=colData(cds_female)$combine_clusters)

agg_mat <- aggregate_gene_expression(cds_female, gene_module_df, cell_group_df)

row.names(agg_mat) <- stringr::str_c("Module ", row.names(agg_mat))

library(pheatmap)

pheatmap::pheatmap(agg_mat,
                   scale="none", 
                   clustering_method="ward.D2")
```

## Plot aggregate module gene expression

```{r, fig.width=15, fig.height=10}
plot_cells(cds_female,
           genes=gene_module_df,
           label_cell_groups=FALSE,
           show_trajectory_graph=TRUE,
           label_leaves=FALSE,
           label_branch_points=FALSE,
           cell_size = 1)
```

## Table of module genes

```{r}
mod_tab = 
  gene_module_df %>%
  rename(id = "Gene") %>%
  select(Gene, module) %>%
  left_join(female_traj_genes) %>%
  select(-status, -morans_test_statistic)

datatable(mod_tab, filter = "top", rownames = FALSE)
```

## Module gene GO enrichment

```{r}

gaf = read.delim("/mnt/research/bioinformaticsCore/shared/Genomes/Petromyzon_marinus/GCF_010993605.1_kPetMar1.pri_gene_ontology.gaf", header=F, comment = "!")

gaf <- 
  gaf %>%
  filter(!grepl("NOT", V4)) %>%  
  dplyr::select(V3, V5) %>%      
  dplyr::rename(GOID = "V5",            
                Gene = "V3") %>%
  distinct() 

keep_go = 
  gaf %>%
  group_by(GOID) %>%
  tally() %>%
  filter(n > 10,
         n < 200) %>%
  pull(GOID)

gaf = 
  gaf %>%
  filter(GOID %in% keep_go)

go_names = read.delim("/mnt/research/bioinformaticsCore/shared/Genomes/Petromyzon_marinus/GOs_names.txt", header = F)

colnames(go_names) = c("GOID", "GOName")

module_genes = 
  gene_module_df %>%
  select(id, module) %>%
  mutate(module = paste0("Module", module))

go_ol =  phyperOverlapSets(gaf, 
                           module_genes, length(cds_female))

write.csv(go_ol$intersecting.genes, file = paste0(results_dir, "/GO_overlaps_female_modules.csv"), row.names = FALSE) 

scores = go_ol$pval.scores

scores = 
  scores %>%
  dplyr::rename(GOID = "Group1",
                Module = "Group2",
                nGOTermGenes = "nGroup1",
                nMonduleGenes = "nGroup2") %>%
  group_by(Module) %>%
    mutate(FDR = p.adjust(Pval, "BH")) %>%
  select(-Enrichment)

scores = 
  left_join(go_names, scores) %>%
  filter(!is.na(nGOTermGenes)) %>%
  filter(Pval < .01)

write.csv(scores, file = paste0(results_dir, "/GO_scores_female_modules.csv"), row.names = FALSE) 

datatable(scores, filter = "top", rownames = FALSE)
```

## Order the cells in pseudotime: Y_77

Setting Y_77 within the primordial germ cell cluster as zero.

```{r, results='hide', eval=FALSE}
start_point = "Y_77"
cds_female <- order_cells(cds_female, reduction_method = "UMAP", root_pr_nodes = start_point)
base::saveRDS(cds_female, file = paste0(results_dir, "/cds_female.Rds"))
```

```{r}
cds_female = readRDS(paste0(results_dir, "/cds_female.Rds"))
plot_cells(cds_female,
           color_cells_by = "pseudotime",
           label_cell_groups=FALSE,
           label_leaves=FALSE,
           label_branch_points=FALSE,
           graph_label_size=1.5,
           cell_size = .7)
```


## Plot aggregated module genes by pseudotime

```{r, fig.height=5, fig.width=5}
agg_mat2 <- aggregate_gene_expression(cds_female, 
                                      gene_module_df,
                                      scale_agg_values = FALSE)

row.names(agg_mat2) = paste0("Module", row.names(agg_mat2))

# make agg_mat2 into a cds
agg_cds <- new_cell_data_set(agg_mat2,
                             cell_metadata = colData(cds_female))

pst = pseudotime(cds_female)
clust = clusters(cds_female)

agg_cds@principal_graph_aux[["UMAP"]]$pseudotime = pst
agg_cds@clusters[["UMAP"]] <- list(cluster_result = clust)

rowData(agg_cds)$gene_short_name = row.names(agg_mat2)

p = plot_genes_in_pseudotime(agg_cds,
                             color_cells_by="cluster",
                             min_expr=0.5)

p
```

## Find modules correlated with pseudotime

```{r}
cor_results <- numeric(nrow(agg_mat2))
p_values <- numeric(nrow(agg_mat2))
  
  # Loop through each row, compute correlation and p-value
  for (i in 1:nrow(agg_mat2)) {
    test_result <- cor.test(agg_mat2[i,], pst, method = "spearman", alternative = "two.sided")
    cor_results[i] <- test_result$estimate       # Extract correlation coefficient
    p_values[i] <- test_result$p.value           # Extract p-value
  }
  
  cor_results_df <- data.frame(
    Module = rownames(agg_mat2),
    Correlation = cor_results,
    P_Value = p_values
  )
datatable(cor_results_df)
```

## Plot aggregated known cell type genes by pseudotime

```{r, fig.height=20, fig.width=7}
known_mat <- aggregate_gene_expression(cds_female, 
                                       chung_davidson_cell_type_markers,
                                       scale_agg_values = FALSE)

# make agg_mat2 into a cds
known_cds <- new_cell_data_set(known_mat,
                               cell_metadata = colData(cds_female))

pst = pseudotime(cds_female)
clust = clusters(cds_female)

known_cds@principal_graph_aux[["UMAP"]]$pseudotime = pst
known_cds@clusters[["UMAP"]] <- list(cluster_result = clust)

rowData(known_cds)$gene_short_name = row.names(known_cds)

p = plot_genes_in_pseudotime(known_cds,
                             color_cells_by="cluster",
                             min_expr=0.5)

p
```

## Find aggregated know cell type marker expression correlated with pseudotime

```{r}
cor_results <- numeric(nrow(known_mat))
p_values <- numeric(nrow(known_mat))
  
  # Loop through each row, compute correlation and p-value
  for (i in 1:nrow(known_mat)) {
    test_result <- cor.test(known_mat[i,], pst, method = "spearman", alternative = "two.sided")
    cor_results[i] <- test_result$estimate       # Extract correlation coefficient
    p_values[i] <- test_result$p.value           # Extract p-value
  }
  
  cor_results_df <- data.frame(
    CellType = rownames(known_mat),
    Correlation = cor_results,
    P_Value = p_values
  )
datatable(cor_results_df)
```
# Select male developmental trajectory clusters

Cluster 18 -> 14 -> 12 -> 2 -> 29 (male germ cell development)

```{r, eval=FALSE}
male = 
  meta_df %>%
  filter(orig_cluster %in% c(2, 12, 14, 18, 29),
         sample != "TF_C00004337") %>%
  pull(barcode)
  
cds_male = cds_coi[,male]

# re-normalize
cds_male <- preprocess_cds(cds_male, num_dim = 100)
cds_male <- align_cds(cds_male, alignment_group = "sample")
cds_male <- reduce_dimension(cds_male)
```
```{r}
cds_male = readRDS(file = paste0(results_dir, "/cds_male_clusters.Rds"))
```

## Plot cells by sample

```{r}
plot_cells(cds_male, 
           color_cells_by="sample", 
           label_cell_groups =FALSE,
           show_trajectory_graph = FALSE,
           label_branch_points = FALSE,
           label_roots = FALSE,
           label_leaves = FALSE) + facet_wrap(~sample, nrow = 2, ncol=2)
```

## Label with original clusters 

```{r}
plot_cells(cds_male,
           show_trajectory_graph = FALSE,
           label_branch_points = FALSE,
           color_cells_by = "orig_cluster",
           group_cells_by = "orig_cluster",
           label_roots = FALSE,
           label_leaves = FALSE)

```

## Label with sex development clusters 

```{r}
plot_cells(cds_male,
           show_trajectory_graph = FALSE,
           label_branch_points = FALSE,
           color_cells_by = "combine_clusters",
           group_cells_by = "combine_clusters",
           label_roots = FALSE,
           label_leaves = FALSE)

```

## Recluster male only

```{r, eval=FALSE}
cds_male <- cluster_cells(cds_male, resolution=7e-4)
```

```{r}
plot_cells(cds_male, 
           label_cell_groups =TRUE,
           show_trajectory_graph = FALSE,
           label_branch_points = FALSE,
           label_roots = FALSE,
           label_leaves = FALSE)

cds_male@colData$cluster = clusters(cds_male)
```

## Identify new cluster enriched genes

This table contains the top ~100 cluster enriched genes for each cluster. It is saved in `male_cluster_marker_genes.csv`. The markers can help us map a cluster to a cell type.

**Column descriptions**

* `gene_id` vector of gene names    
* `cell_group` character vector of the cell group to which the cell belongs   
* `marker_score` numeric vector of marker scores as the fraction expressing scaled by the specificity. The value ranges from 0 to 1.    
* `mean_expression` numeric vector of mean normalized expression of the gene in the cell group    
* `fraction_expressing` numeric vector of fraction of cells expressing the gene within the cell group   
* `specificity` numeric vector of a measure of how specific the gene's expression is to the cell group based on the Jensen-Shannon divergence. The value ranges from 0 to 1.   
* `pseudo_R2` numeric vector of pseudo R-squared values, a measure of how well the gene expression model fits the categorical data relative to the null model. The value ranges from 0 to 1.   
* `marker_test_p_value` numeric vector of likelihood ratio p-values   
* `marker_test_q_value` numeric vector of likelihood ratio q-values   

```{r, results='hide'}
cluster_marker_res <- top_markers(cds_male, 
                                  genes_to_test_per_group = 100,
                                  group_cells_by="cluster", cores=8)

write.csv(cluster_marker_res, file = paste0(results_dir, "/male_cluster_marker_genes.csv"), row.names = FALSE) 
```

```{r}
library(DT)
datatable(cluster_marker_res, filter = "top", rownames = FALSE)
```

## Plot genes of interest

### TSNE plots

```{r, fig.width=12, fig.height=8}
goi = c("BOLL", "CCDC189", "CFAP69", "NANOS1", "PIWIL1", "SPAG5", "SPAG16", "SPATA22", "TEX11", "WNT11","LOC116937338", "LOC116937873", "LOC116937887", "LOC116938414", "LOC116940726", "LOC116941541", "LOC116944806", "LOC116948775")

plot_cells(cds_male, genes=goi,
           show_trajectory_graph=FALSE,
           label_cell_groups=FALSE,
           label_leaves=FALSE,
           label_branch_points = FALSE)
```

### Violin plots

```{r, warning=FALSE, fig.height=20, fig.width=5}
cds_subset <- cds_male[rowData(cds_male)$gene_short_name %in% goi,]
plot_genes_violin(cds_subset, ncol=1, group_cells_by="cluster")
```

## Plot known cell-type markers

```{r, fig.width=12, fig.height=8}
rowData(cds_male)$gene_short_name = row.names(cds_male)
plot_cells(cds_male, genes=chung_davidson_unique %>% 
             filter(Group %in% c("Germline cells during epigenetic reprogramming", 
                                 "Germline stem cells",
                                 "Embryonic stem cells",
                                 "Germline progenitor and amnionic cells",
                                 "Primordial Germ Cells (PGC",
                                 "Migrating germ cells",
                                 "Late PGC (meiotic entry genes")),
           show_trajectory_graph=FALSE,
           label_cell_groups=FALSE,
           label_leaves=FALSE,
           label_branch_points = FALSE,
           cell_size = .7)
```

Plot female cell types

```{r, fig.width=12, fig.height=4}
plot_cells(cds_male, genes=chung_davidson_unique %>% 
             filter(Group %in% c("Female Germ cell", 
                                 "Mitotic oocyte",
                                 "Granulosa cell",
                                 "Oocyte")),
           show_trajectory_graph=FALSE,
           label_cell_groups=FALSE,
           label_leaves=FALSE,
           label_branch_points = FALSE,
           cell_size = .7)
```

Plot male cell types

```{r, fig.width=12, fig.height=8}
plot_cells(cds_male, genes=chung_davidson_unique %>% 
             filter(Group %in% c("Male germ cells", 
                                 "Sertoli cell",
                                 "Mitotic Spermatogonia",
                                 "Meiotic Spermatogonia",
                                 "Spermatogonia",
                                 "Primary spermatocyte")),
           show_trajectory_graph=FALSE,
           label_cell_groups=FALSE,
           label_leaves=FALSE,
           label_branch_points = FALSE,
           cell_size = .7)
```

## Construct single-cell trajectories

The black line shows the trajectory graph. It connects cell clusters or individual cells in the reduced space, ensuring smooth paths through the data structure. The graph models how cells progress through different states or fates based on gene expression similarity.

```{r, results='hide', eval=FALSE}
cds_male <- learn_graph(cds_male, use_partition = FALSE, close_loop = FALSE)
base::saveRDS(cds_male, file = paste0(results_dir, "/cds_male_clusters.Rds"))
```

## New male-specific clusters

```{r}
plot_cells(cds_male,
           label_groups_by_cluster=TRUE,
           label_principal_points = FALSE,
           label_branch_points = FALSE,
           label_roots = FALSE,
           label_leaves = FALSE,
           show_trajectory_graph=TRUE,
           cell_size = .7,
            alpha = .3)
```

## Original clusters

```{r}
plot_cells(cds_male,
           color_cells_by = "orig_cluster",
           group_cells_by = "orig_cluster",
           label_groups_by_cluster=TRUE,
           label_principal_points = FALSE,
           label_branch_points = FALSE,
           label_roots = FALSE,
           label_leaves = FALSE,
           show_trajectory_graph=TRUE,
           cell_size = .7,
            alpha = .3)
```

Principal points approximate key positions along the inferred trajectory and serve as summarized representations of the cells' progression through pseudotime. To define pseudotime, we need to choose 1 or more principal points that to represent the start of the trajectory. 

```{r}
plot_cells(cds_male,
           label_groups_by_cluster=FALSE,
           label_principal_points = TRUE,
           show_trajectory_graph=TRUE,
           alpha = .1,
           cell_size = .7)
```

## Find genes that change along the trajectory 

Monocle3 introduces a new approach for finding genes that are differentially expressed across a single-cell trajectory that draws on a powerful technique in spatial correlation analysis, the Moran’s I test. Moran’s I is a measure of multi-directional and multi-dimensional spatial autocorrelation. The statistic tells you whether cells at nearby positions on a trajectory will have similar (or dissimilar) expression levels for the gene being tested. Although both Pearson correlation and Moran’s I ranges from -1 to 1, the interpretation of Moran’s I is slightly different: +1 means that nearby cells will have perfectly similar expression; 0 represents no correlation, and -1 means that neighboring cells will be anti-correlated.

```{r, results='hide', eval=FALSE}
male_traj_genes <- graph_test(cds_male, neighbor_graph="principal_graph", cores=10)

male_traj_genes$Gene = rownames(male_traj_genes)

saveRDS(male_traj_genes, file = paste0(results_dir, "/male_traj_genes.Rds"))
```

```{r}
male_traj_genes = readRDS(file = paste0(results_dir, "/male_traj_genes.Rds"))

datatable(male_traj_genes %>% 
            arrange(desc(morans_I)) %>% 
            filter(q_value < .01) %>%
            select(-status, -morans_test_statistic, -gene_short_name) %>%
            select(morans_I, p_value, q_value),
          filter = "top")
```

**Columns**

* `morans_I`: If you'd like to rank the genes by effect size, sort this table by the `morans_I` column, which ranges from -1 to +1. A value of 0 indicates no effect, while +1 indicates perfect positive autocorrelation and suggests that nearby cells have very similar values of a gene's expression.    
* `p_value`: p-value from the Moran's I test
* `q_value`: q-value from the Moran's I test (adjusted for multiple comparisons)

## Group the genes into modules

Grouping genes into modules helps us understand which genes have similar expression patterns across the trajectory.

```{r, fig.height=6, eval=FALSE}
deg_ids <- row.names(subset(male_traj_genes, q_value < 0.01))

gene_module_df <- find_gene_modules(cds_male[deg_ids,], resolution=c(10^seq(-20,-4)), random_seed=1)

write.csv(gene_module_df, file = paste0(results_dir, "/male_gene_module_df.csv"))
```

```{r, fig.height=6}
gene_module_df = read.csv(file = paste0(results_dir, "/male_gene_module_df.csv"), row.names = 1)

cell_group_df <- tibble::tibble(cell=row.names(colData(cds_male)), 
                                cell_group=colData(cds_male)$cluster)

agg_mat <- aggregate_gene_expression(cds_male, gene_module_df, cell_group_df)

row.names(agg_mat) <- stringr::str_c("Module ", row.names(agg_mat))

library(pheatmap)

pheatmap::pheatmap(agg_mat,
                   scale="none", 
                   clustering_method="ward.D2")
```

## Plot aggregate module gene expression

```{r, fig.width=15, fig.height=10}
plot_cells(cds_male,
           genes=gene_module_df,
           label_cell_groups=FALSE,
           show_trajectory_graph=TRUE,
           label_leaves=FALSE,
           label_branch_points=FALSE,
           cell_size = .7)
```

## Table of module genes

```{r}
mod_tab = 
  gene_module_df %>%
  rename(id = "Gene") %>%
  select(Gene, module) %>%
  left_join(male_traj_genes) %>%
  select(-status, -morans_test_statistic)

datatable(mod_tab, filter = "top", rownames = FALSE)
```

## Module gene GO enrichment

```{r}

gaf = read.delim("/mnt/research/bioinformaticsCore/shared/Genomes/Petromyzon_marinus/GCF_010993605.1_kPetMar1.pri_gene_ontology.gaf", header=F, comment = "!")

gaf <- 
  gaf %>%
  filter(!grepl("NOT", V4)) %>%  
  dplyr::select(V3, V5) %>%      
  dplyr::rename(GOID = "V5",            
                Gene = "V3") %>%
  distinct() 

keep_go = 
  gaf %>%
  group_by(GOID) %>%
  tally() %>%
  filter(n > 10,
         n < 200) %>%
  pull(GOID)

gaf = 
  gaf %>%
  filter(GOID %in% keep_go)

go_names = read.delim("/mnt/research/bioinformaticsCore/shared/Genomes/Petromyzon_marinus/GOs_names.txt", header = F)

colnames(go_names) = c("GOID", "GOName")

module_genes = 
  gene_module_df %>%
  select(id, module) %>%
  mutate(module = paste0("Module", module))

go_ol =  phyperOverlapSets(gaf, 
                           module_genes, length(cds_female))

write.csv(go_ol$intersecting.genes, file = paste0(results_dir, "/GO_overlaps_male_modules.csv"), row.names = FALSE) 

scores = go_ol$pval.scores

scores = 
  scores %>%
  dplyr::rename(GOID = "Group1",
                Module = "Group2",
                nGOTermGenes = "nGroup1",
                nMonduleGenes = "nGroup2") %>%
  group_by(Module) %>%
    mutate(FDR = p.adjust(Pval, "BH")) %>%
  select(-Enrichment)

scores = 
  left_join(go_names, scores) %>%
  filter(!is.na(nGOTermGenes)) %>%
  filter(Pval < .01)

write.csv(scores, file = paste0(results_dir, "/GO_scores_male_modules.csv"), row.names = FALSE) 

datatable(scores, filter = "top", rownames = FALSE)
```


# Order the cells in pseudotime: Y_654

With "Y_646" set as 0. **Problem** Cluster 4 (mitotic spermatogonia) gets marked as the furthest along the trajectory here. 

```{r, results='hide'}
start_point = "Y_646"

cds_male <- order_cells(cds_male, reduction_method = "UMAP", root_pr_nodes = start_point)
base::saveRDS(cds_male, file = paste0(results_dir, "/male_cds_nb_", start_point, ".Rds"))
```

```{r}
plot_cells(cds_male,
           color_cells_by = "pseudotime",
           label_cell_groups=FALSE,
           label_leaves=FALSE,
           label_branch_points=FALSE,
           graph_label_size=1.5) 
```

## Plot aggregated module genes by pseudotime

```{r, fig.height=10, fig.width=7}
agg_mat2 <- aggregate_gene_expression(cds_male, 
                                      gene_module_df,
                                      scale_agg_values = FALSE)

row.names(agg_mat2) = paste0("Module", row.names(agg_mat2))

# make agg_mat2 into a cds
agg_cds <- new_cell_data_set(agg_mat2,
                             cell_metadata = colData(cds_male))

pst = pseudotime(cds_male)
clust = clusters(cds_male)

agg_cds@principal_graph_aux[["UMAP"]]$pseudotime = pst
agg_cds@clusters[["UMAP"]] <- list(cluster_result = clust)

rowData(agg_cds)$gene_short_name = row.names(agg_mat2)

p = plot_genes_in_pseudotime(agg_cds,
                             color_cells_by="cluster",
                             min_expr=0.5)

p

```

## Find modules correlated with pseudotime

```{r}
cor_results <- numeric(nrow(agg_mat2))
p_values <- numeric(nrow(agg_mat2))
  
  # Loop through each row, compute correlation and p-value
  for (i in 1:nrow(agg_mat2)) {
    test_result <- cor.test(agg_mat2[i,], pst, method = "spearman", alternative = "two.sided")
    cor_results[i] <- test_result$estimate       # Extract correlation coefficient
    p_values[i] <- test_result$p.value           # Extract p-value
  }
  
  cor_results_df <- data.frame(
    Row = rownames(agg_mat2),
    Correlation = cor_results,
    P_Value = p_values
  )
datatable(cor_results_df)
```

## Plot aggregated known cell type genes by pseudotime

```{r, fig.height=20, fig.width=7}
known_mat <- aggregate_gene_expression(cds_male, 
                                       chung_davidson_cell_type_markers,
                                       scale_agg_values = FALSE)

# make agg_mat2 into a cds
known_cds <- new_cell_data_set(known_mat,
                               cell_metadata = colData(cds_male))

pst = pseudotime(cds_male)
clust = clusters(cds_male)

known_cds@principal_graph_aux[["UMAP"]]$pseudotime = pst
known_cds@clusters[["UMAP"]] <- list(cluster_result = clust)

rowData(known_cds)$gene_short_name = row.names(known_cds)

p = plot_genes_in_pseudotime(known_cds,
                             color_cells_by="cluster",
                             min_expr=0.5)

p
```

## Find aggregated know cell type marker expression correlated with pseudotime

```{r}
cor_results <- numeric(nrow(known_mat))
p_values <- numeric(nrow(known_mat))
  
  # Loop through each row, compute correlation and p-value
  for (i in 1:nrow(known_mat)) {
    test_result <- cor.test(known_mat[i,], pst, method = "spearman", alternative = "two.sided")
    cor_results[i] <- test_result$estimate       # Extract correlation coefficient
    p_values[i] <- test_result$p.value           # Extract p-value
  }
  
  cor_results_df <- data.frame(
    CellType = rownames(known_mat),
    Correlation = cor_results,
    P_Value = p_values
  )
datatable(cor_results_df)
```

# Session info
```{r}
sessionInfo()
```

