---
title: "TSCAN trajectory analysis"
author: "Stephanie Hickey"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: true
    toc_float:
      toc_collapsed: true
params:
    project_dir: /mnt/ufs18/rs-013/bioinformaticsCore/projects/chong_davidson/BCC125_sea_lamprey_scRNA
    project_name: BCC125_sea_lamprey_scRNA
    lab: Weiming Li
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
```

```{r, echo=FALSE, results="asis"}
## parameterized text
cat("\n")
cat("#### Lab:", params$lab, " \n")
cat("#### Project:", params$project_name, " \n")
```

```{r}
library(tidyverse)
library(SingleCellExperiment)
library(scater)
library(scran)
library(patchwork)
library(pheatmap)
library(ggsankey)
library(RColorBrewer)
library(readxl)
library(monocle3)
library(batchelor)
library(BiocSingular)
library(DT)
library(TSCAN)
source(paste0(params$project_dir, "/src/scRNA_seq_functions.R"))
```

```{r}
# Set Paths
results_dir = paste0(params$project_dir, "/results")
# results_dir = "/mnt/ufs18/rs-013/bioinformaticsCore/projects/chong_davidson/BCC125_sea_lamprey_scRNA/results"
data_dir = paste0(params$project_dir, "/data")
# data_dir = "/mnt/ufs18/rs-013/bioinformaticsCore/projects/chong_davidson/BCC125_sea_lamprey_scRNA/data"
```

# Clustering of individual samples

```{r, eval=FALSE}
# Load QC'd data
sample_dirs = list.dirs(data_dir, full.names = TRUE, recursive = FALSE)
sample_names = basename(sample_dirs)

cds_list = list()

for(i in 1:length(sample_dirs)){
  
  cds_list[[i]] = readRDS(paste0(sample_dirs[i], 
                                 "/", 
                                 sample_names[i],
                                 "_cds_mito_filter.Rds"))
}

names(cds_list) = sample_names
all.sce = list()

# cds to sce
for (cds in names(cds_list)){
  
  raw_counts = counts(cds_list[[cds]])
  all.sce[[cds]] = 
    SingleCellExperiment(list(counts=raw_counts), 
                         colData=cds_list[[cds]]@colData)
  all.sce[[cds]]@colData$sample = cds
  
}

# https://bioconductor.org/books/3.14/OSCA.multisample/human-pbmcs-10x-genomics.html
# normalization, feature selection, dim reduction
all.sce <- lapply(all.sce, logNormCounts)
all.dec <- lapply(all.sce, modelGeneVar)
all.hvgs <- lapply(all.dec, getTopHVGs, prop=0.1)
all.sce <- mapply(FUN=runPCA, x=all.sce, subset_row=all.hvgs)

set.seed(1)
all.sce <- lapply(all.sce, runTSNE, dimred="PCA")
all.sce <- lapply(all.sce, runUMAP, dimred="PCA")

# individual clustering
for (n in names(all.sce)) {
    #print(paste0("start ", n, " buildSNNGraph"))
    set.seed(1)
    g <- buildSNNGraph(all.sce[[n]], k=30, use.dimred='PCA')
    print(paste0("start ", n, " clust"))
    clust <- igraph::cluster_walktrap(g)$membership
    colLabels(all.sce[[n]])  <- factor(clust)
}

saveRDS(all.sce, file=paste0(results_dir, "/all.sce.Rds"))
```
```{r, fig.height=10, fig.width=8}
all.sce = readRDS(paste0(results_dir, "/all.sce.Rds"))

all.umap <- list()
for (n in names(all.sce)) {
    all.umap[[n]] <- plotUMAP(all.sce[[n]], colour_by="label", point_size = .2) + ggtitle(n)
}

do.call(gridExtra::grid.arrange, c(all.umap, list(ncol=2)))
```

# Merge samples, removing sample batch effect

```{r, eval=FALSE}
# integration
# Intersecting the common genes.
universe <- Reduce(intersect, lapply(all.sce, rownames))
all.sce2 <- lapply(all.sce, "[", i=universe,)
all.dec2 <- lapply(all.dec, "[", i=universe,)

# Renormalizing to adjust for differences in depth.
normed.sce <- do.call(multiBatchNorm, all.sce2)

# Identifying a set of HVGs using stats from all batches.
combined.dec <- do.call(combineVar, all.dec2)
combined.hvg <- getTopHVGs(combined.dec, n=5000)

set.seed(1)
merged.sce <- do.call(fastMNN, c(normed.sce, 
    list(subset.row=combined.hvg, BSPARAM=RandomParam())))

# clustering in corrected space
g <- buildSNNGraph(merged.sce, use.dimred="corrected", k=30)
set.seed(1)
colLabels(merged.sce) <- factor(igraph::cluster_louvain(g)$membership)
# table(colLabels(merged.sce), merged.sce$batch)

# add monocle clusters
cds_nb = readRDS(paste0(results_dir, "/cds_nb.Rds"))

merged.sce@colData$monocole_cluster = cds_nb@colData$cluster

set.seed(1)
merged.sce <- runUMAP(merged.sce, dimred="corrected")
set.seed(1)
merged.sce <- runTSNE(merged.sce, dimred="corrected")

saveRDS(merged.sce, file = paste0(results_dir, "/merged.sce.Rds"))
```

The plot below shows the merged data in TSNE space separated by sample.

```{r, fig.width=4, fig.height=4}

merged.sce = readRDS(paste0(results_dir, "/merged.sce.Rds"))

# plot UMAP

p <- lapply(unique(merged.sce$batch), function(f) {
  sce.f = merged.sce[, which(merged.sce$batch == f)]
  plotTSNE(sce.f, point_size = .05) + ggtitle(f)
})

wrap_plots(p, ncol = 2)
```

This plot shows new clusters identified using the "scatter" package in the batch corrected expresion space.

```{r, fig.width=6, fig.height=6}
plotTSNE(merged.sce, colour_by="label", text_by="label", text_colour="red", point_size = .05)
```

This plot shows the proportion of cells from each sample within each new cluster.

```{r}
meta_df = as.data.frame(colData(merged.sce))

sample_tally = 
  meta_df %>%
  group_by(label, batch) %>%
  tally()

ggplot(sample_tally, aes(fill=batch, y=label, x=n)) + 
    geom_bar(position="fill", stat="identity") +
    xlab("Proportion of cells")

```

## Plot known cell-type markers

In each cell, the expression of genes within each group are summed. In the plots, higher expression is a brighter color. Expression is scaled to percent of maximum expression.

Use known markers unique to each cell type:

```{r}
# load known marker genes
library(readr)
chung_davidson_cell_type_markers <- read_csv(paste0(data_dir, "/chung_davidson_cell_type_markers.csv"))
chung_davidson_cell_type_markers = 
  chung_davidson_cell_type_markers %>%
  select(-GeneName)

chung_davidson_unique = 
  chung_davidson_cell_type_markers %>%
  group_by(Gene) %>%
  tally() %>%
  filter(n == 1) %>%
  select(Gene) %>%
  left_join(chung_davidson_cell_type_markers)

datatable(chung_davidson_unique, filter = "top", rownames = FALSE)
```
```{r}
# make cds for plotting with monocle
# Intersecting the common genes.
all.dec <- lapply(all.sce, modelGeneVar)
universe <- Reduce(intersect, lapply(all.sce, rownames))
all.sce2 <- lapply(all.sce, "[", i=universe,)
all.dec2 <- lapply(all.dec, "[", i=universe,)

all.sce2[["TM_C00004374"]]@colData$cluster="noclust" 

combined.sce <- do.call(cbind, all.sce2)

merged.cds <- new_cell_data_set(counts(combined.sce),
                                cell_metadata = merged.sce@colData)

reducedDims(merged.cds)$UMAP <- reducedDims(merged.sce)$UMAP
reducedDims(merged.cds)$tSNE <- reducedDims(merged.sce)$TSNE
```

Plot for early cell markers.

```{r, fig.width=12, fig.height=8}
rowData(merged.cds)$gene_short_name = row.names(merged.cds)
plot_cells(merged.cds, reduction_method = "tSNE",
           genes=chung_davidson_unique %>% 
             filter(Group %in% c("Germline cells during epigenetic reprogramming", 
                                 "Germline stem cells",
                                 "Embryonic stem cells",
                                 "Germline progenitor and amnionic cells",
                                 "Primordial Germ Cells (PGC",
                                 "Migrating germ cells",
                                 "Late PGC (meiotic entry genes")))
```

Female cell markers

```{r, fig.height=4, fig.width=12}
plot_cells(merged.cds, reduction_method = "tSNE",
           genes=chung_davidson_unique %>% 
             filter(Group %in% c("Female Germ cell", 
                                 "Mitotic oocyte",
                                 "Granulosa cell",
                                 "Oocyte")),
           show_trajectory_graph=FALSE,
           label_cell_groups=FALSE,
           label_leaves=FALSE,
           label_branch_points = FALSE)
```

Male cell markers.

```{r, fig.width=12, fig.height=8}
plot_cells(merged.cds, reduction_method = "tSNE",
           genes=chung_davidson_unique %>% 
             filter(Group %in% c("Male germ cells", 
                                 "Sertoli cell",
                                 "Mitotic Spermatogonia",
                                 "Meiotic Spermatogonia",
                                 "Spermatogonia",
                                 "Primary spermatocyte")),
           show_trajectory_graph=FALSE,
           label_cell_groups=FALSE,
           label_leaves=FALSE,
           label_branch_points = FALSE)
```

## Plot other genes of interest

Other markers of interest

```{r, fig.width=12, fig.height=12}
goi = c("BOLL", "CCDC189", "CFAP69", "NANOS1", "PIWIL1", "SPAG5", "SPAG16", "SPATA22", "TEX11", "WNT11","LOC116937338", "LOC116937873", "LOC116937887", "LOC116938414", "LOC116940726", "LOC116941541", "LOC116944806", "LOC116948775")
  
plot_cells(merged.cds, reduction_method = "tSNE", 
           genes=goi,
           show_trajectory_graph=FALSE,
           label_cell_groups=FALSE,
           label_leaves=FALSE,
           label_branch_points = FALSE)
```

Violin plots

```{r, fig.height=16, fig.width=10}
scatter.clusters = as.data.frame(colData(merged.sce)) 
scatter.clusters$barcode = rownames(scatter.clusters)
 
scatter.clusters =
  scatter.clusters %>%
  select(barcode, label) %>%
  deframe()

scatter.clusters = scatter.clusters[colnames(combined.sce)]
combined.sce$label = scatter.clusters

plotExpression(combined.sce, x = "label", features = goi)
```

## Compare scatter cluster enriched genes with known markers

### Cluster enriched genes

This table contains the top ~100 cluster enriched genes for each cluster. All of the cluster enriched genes are saved in `results/scatter_cluster_marker_genes.csv`. The markers can help us map a cluster to a cell type.

In this case, genes in each cluster were compared to genes in every other cluster in a pairwise manner, and the p.values and logFCs of those comparisons have been summarized for ease of interpretation. 

**Column descriptions**

* `cluster`: The cluster corresponding to the enriched genes.      
* `Gene`: The enriched gene   
* `p.value`: The combined p-value across all pairwise comparisons.     
* `FDR`: The BH-adjusted p-value for each gene.   
* `summary.logFC`: This contains the summary effect size, obtained by combining effect sizes from all pairwise comparison into a single value.    

```{r, eval=FALSE}
scatter_markers = findMarkers(combined.sce,
                              merged.sce$label,
                              test.type="t",
                              block=combined.sce$sample,
                              direction="up",
                              pval.type="some")

saveRDS(scatter_markers, file= paste0(results_dir, "/scatter_markers.Rds"))
```

```{r}
scatter_markers = readRDS(paste0(results_dir, "/scatter_markers.Rds"))

scatter_summary = list()

for (i in 1:length(scatter_markers)){
  
scatter_markers[[i]]$cluster = i
scatter_markers[[i]]$Gene = rownames(scatter_markers[[i]])
scatter_summary[[i]] = 
  as.data.frame(scatter_markers[[i]]) %>%
  select(cluster, Gene, p.value, FDR, summary.logFC)

}

scatter_markers_df =
  do.call(rbind, scatter_summary) %>%
  filter(FDR < .01)

write.csv(scatter_markers_df, file = paste0(results_dir, "/scatter_cluster_marker_genes.csv"))
  
scatter_markers_top100 = 
  do.call(rbind, scatter_summary) %>%
  filter(FDR < .01) %>%
  group_by(cluster) %>%
  slice_min(FDR, n = 100, with_ties=TRUE)
  
datatable(scatter_markers_top100, filter = "top", rownames = FALSE)  
```

### Comparison with known markers

We'll check for the significance of the overlap between the top 100 cluster enriched genes and the supplied known cell-type markers. In the plot below, the rows represent cell types with know marker genes, and the columns represent the cell clusters. The number in each box is the number of genes shared between the top 100 cluster enriched genes for the specified clusters and the specified known cell-type markers. Each box is colored by the significance of the overlap ($-log10(FDR)$). Significant overlaps may signify that the cells in the cluster belong to the specified cell type. 

* Overlap p-values are save in `results/scatter_cluster_marker_overlap_pvalues.csv`.    
* Overlaps between cluster enriched genes and known marker genes are save in `results/scatter_cluster_marker_gene_overlaps.csv`.    

```{r, fig.width=8}
marker_test_genes =
  scatter_markers_top100 %>%
  select(Gene, cluster) %>%
  mutate(cluster = paste0("cluster", cluster))

cluster_ol =  phyperOverlapSets(chung_davidson_cell_type_markers, 
                                marker_test_genes, length(merged.sce))

write.csv(cluster_ol$intersecting.genes, file = paste0(results_dir, "/scatter_cluster_marker_gene_overlaps.csv"), row.names = FALSE) 

scores = cluster_ol$pval.scores

scores = 
  scores %>%
  rename(Group1 = "CellTypes",
         Group2 = "Clusters") %>%
  group_by(CellTypes) %>%
    mutate(FDR = p.adjust(Pval, "BH"))

write.csv(scores, file = paste0(results_dir, "/scatter_cluster_marker_overlap_pvalues.csv"), row.names = FALSE) 

scores$FDR[scores$FDR > .01] = 1

FDR_wide <- 
  scores %>%
  mutate(logFDR = -log10(FDR)) %>%
  pivot_wider(id_cols = CellTypes, 
              names_from = Clusters, 
              values_from=logFDR)

CellTypes = FDR_wide$CellTypes
FDR_wide$CellTypes = NULL

FDR_wide = as.matrix(FDR_wide)
rownames(FDR_wide) = CellTypes

ord_row <- hclust( dist(FDR_wide), method = "ward.D" )$order
ord_col <- hclust( dist(t(FDR_wide)), method = "ward.D" )$order

scores$CellTypes = factor(scores$CellTypes, levels = rownames(FDR_wide)[ord_row])
scores$Clusters = factor(scores$Clusters, levels = colnames(FDR_wide)[ord_col])

plot1 =
    ggplot(scores,
           aes(Clusters, CellTypes)) + 
           geom_tile(aes(fill = -log10(FDR)), 
                     color = "light gray") +
        #Creating legend
        guides(fill=guide_colorbar(title = "-log10(FDR)",
                                   barheight = 10,
                                   barwidth = 2,
                                   default.unit = "mm")) +
        #Creating color range
        scale_fill_gradientn(colors=c("white",
                                      "#fd9668",
                                      "#f1605d",
                                      "#de4968",
                                      "#9e2f7f"),
                             guide="colorbar",
                             limits = c(0, ceiling(-log10(min(scores$FDR))))
                             ) +
        #Rotating labels
         theme(axis.text.x = element_text(angle = 30, 
                                          hjust = 1,vjust=1),
               ) +
        theme(text = element_text(size = 7),
              axis.text = element_text(size = 7)) +
        geom_text(aes(label = nSharedGenes), size = 1) +
        coord_fixed()

plot1

ggsave(plot1, file = paste0(results_dir, "/scatter_cluster_overlap_heatmap.png"))
```

## Gene ontology of top 100 scatter cluster enriched genes 

To help annotate cell type clusters, we can check for enriched gene ontology categories among cell-type enriched genes.

*Columns*

* `GOID`: Gene ontology ID            
* `GOName`: Gene ontology name    
* `nGOTermGenes`: Number of genes assigned to GO term        
* `Clusters`: Cluster   
* `nClusterEnrichedGenes`: Number of cluster enriched genes      
* `nSharedGenes`: Number of shared genes between the GO term and cluster enriched genes   
* `Pval`: p-value          
* `FDR`: p-value corrected for multiple comparisons         

Overlap p-values are save in `results/scatter_cluster_marker_gene_GO_overlap_pvalues.csv`. Overlaps between cluster enriched genes and known marker genes are save in `results/scatter_cluster_marker_gene_GO_overlaps.csv`.

```{r}
gaf = read.delim("/mnt/research/bioinformaticsCore/shared/Genomes/Petromyzon_marinus/GCF_010993605.1_kPetMar1.pri_gene_ontology.gaf", header=F, comment = "!")

gaf <- 
  gaf %>%
  filter(!grepl("NOT", V4)) %>%  
  dplyr::select(V3, V5) %>%      
  dplyr::rename(GOID = "V5",            
                Gene = "V3") %>%
  distinct() 

keep_go = 
  gaf %>%
  group_by(GOID) %>%
  tally() %>%
  filter(n > 10,
         n < 200) %>%
  pull(GOID)

gaf = 
  gaf %>%
  filter(GOID %in% keep_go)

go_names = read.delim("/mnt/research/bioinformaticsCore/shared/Genomes/Petromyzon_marinus/GOs_names.txt", header = F)
colnames(go_names) = c("GOID", "GOName")
  
go_ol =  phyperOverlapSets(gaf, 
                           marker_test_genes, length(merged.cds))

write.csv(go_ol$intersecting.genes, file = paste0(results_dir, "/scatter_cluster_marker_gene_GO_overlaps.csv"), row.names = FALSE) 

scores = go_ol$pval.scores

scores = 
  scores %>%
  dplyr::rename(GOID = "Group1",
                Clusters = "Group2",
                nGOTermGenes = "nGroup1",
                nClusterEnrichedGenes = "nGroup2") %>%
  group_by(Clusters) %>%
    mutate(FDR = p.adjust(Pval, "BH")) %>%
  select(-Enrichment)

scores = 
  left_join(go_names, scores, by="GOID") %>%
  filter(!is.na(nGOTermGenes)) %>%
  filter(Pval < .01)

write.csv(scores, file = paste0(results_dir, "/scatter_cluster_marker_gene_GO_overlap_pvalues.csv"), row.names = FALSE) 

datatable(scores, filter = "top", rownames = FALSE)
```

## Compare scatter clusters with monocle clusters

This plot show the original monocle clusters identified using the "monocle3" package in TSNE space.

```{r, fig.width=7, fig.height=6}
plotTSNE(merged.sce, colour_by="monocole_cluster", text_by="monocole_cluster", text_colour="red", point_size = .05)
```

### All clusters

This plot connects monocle clusters to scatter clusters by shared cells

```{r, fig.height=20, fig.width=10}
# Create a data frame of transitions
df <- data.frame(id = colnames(merged.sce),
                 scatter_cluster = paste0("scatter_", merged.sce@colData$label),
                 monocle_cluster =  paste0("monocle_",merged.sce@colData$monocole_cluster))

# Convert data to long format required by ggsankey
df_long <- df %>%
  make_long(monocle_cluster, scatter_cluster)
            

# Create the Sankey plot
ggplot(df_long, aes(x = x, 
                    next_x = next_x, 
                    node = node, 
                    next_node = next_node, 
                    fill = node,
                    label = node)) +
  geom_sankey(flow.alpha = 0.7, node.color = "black") +
  geom_sankey_label(size = 5) +
  theme_sankey(base_size = 16) +
  theme(legend.position = "none")
  
```

### Just sex development clusters

```{r, fig.height=10, fig.width=10}
sex_clust = c(18,14,12,2,29,13,15)
sex_clust = paste0("monocle_", sex_clust)

df_sex =
  df %>%
  filter(monocle_cluster %in% sex_clust)
  
df_long <- df_sex %>%
  make_long(monocle_cluster, scatter_cluster)            

# Create the Sankey plot
ggplot(df_long, aes(x = x, 
                    next_x = next_x, 
                    node = node, 
                    next_node = next_node, 
                    fill = node,
                    label = node)) +
  geom_sankey(flow.alpha = 0.7, node.color = "black") +
  geom_sankey_label(size = 5) +
  theme_sankey(base_size = 16) +
  theme(legend.position = "none")

```

### Monocle to scatter table

Below is a table showing how many cells from each monocle cluster went to each scatter cluster.

```{r}
df_count <- 
  df %>%
  dplyr::count(monocle_cluster, scatter_cluster) %>%
  group_by(monocle_cluster) %>%
  mutate(total_m1 = sum(n)) %>%
  ungroup() %>%
  group_by(scatter_cluster) %>%
  mutate(total_m2 = sum(n)) %>%
  ungroup() %>%
  # Compute the percentage of total in method 1 and method 2
  mutate(
    pct_of_monocle = round((n / total_m1) * 100, 2),
    pct_of_scatter = round((n / total_m2) * 100, 2)) %>%
  select(-total_m1, -total_m2)
  
datatable(df_count, filter = "top", rownames = FALSE)
```


### Monocle cluster enirched genes vs top 100 scatter cluster genes

Overlapping cluster enriched genes can be found in `results/scatter_monocle_cluster_marker_gene_overlaps.csv`

```{r, fig.height=8, fig.width=8, warnings=FALSE}

marker_test_genes =
  scatter_markers_top100 %>%
  select(Gene, cluster) %>%
  mutate(cluster = paste0("scatter", cluster))

monocle_genes = read.csv(paste0(results_dir, "/combined_samples_cluster_marker_genes_no_batch.csv"))

monocle_genes = 
  monocle_genes %>%
  select(gene_id, cell_group) %>%
  mutate(cell_group = paste0("monocle", cell_group))

cluster_ol =  phyperOverlapSets(monocle_genes, 
                                marker_test_genes, length(merged.sce))

write.csv(cluster_ol$intersecting.genes, file = paste0(results_dir, "/scatter_monocle_cluster_marker_gene_overlaps.csv"), row.names = FALSE) 

scores = cluster_ol$pval.scores

scores = 
  scores %>%
  dplyr::rename(monocle = "Group1",
                scatter = "Group2") %>%
  group_by(monocle) %>%
    mutate(FDR = p.adjust(Pval, "BH"))

scores$FDR[scores$FDR > .01] = 1

FDR_wide <- 
  scores %>%
  mutate(logFDR = -log10(FDR)) %>%
  pivot_wider(id_cols = monocle, 
              names_from = scatter, 
              values_from=logFDR)

monocle_clust = FDR_wide$monocle
FDR_wide$monocle = NULL

FDR_wide = as.matrix(FDR_wide)
rownames(FDR_wide) = monocle_clust

ord_row <- hclust( dist(FDR_wide), method = "ward.D" )$order
ord_col <- hclust( dist(t(FDR_wide)), method = "ward.D" )$order

scores$monocle = factor(scores$monocle, levels = rownames(FDR_wide)[ord_row])
scores$scatter = factor(scores$scatter, levels = colnames(FDR_wide)[ord_col])


plot1 =
    ggplot(scores,
           aes(scatter, monocle)) + 
           geom_tile(aes(fill = -log10(FDR)), 
                     color = "light gray") +
        #Creating legend
        guides(fill=guide_colorbar(title = "-log10(FDR)",
                                   barheight = 10,
                                   barwidth = 2,
                                   default.unit = "mm")) +
        #Creating color range
        scale_fill_gradientn(colors=c("white",
                                      "#fd9668",
                                      "#f1605d",
                                      "#de4968",
                                      "#9e2f7f"),
                             guide="colorbar",
                             limits = c(0, ceiling(-log10(min(scores$FDR))))
                             ) +
        #Rotating labels
         theme(axis.text.x = element_text(angle = 30, 
                                          hjust = 1,vjust=1),
               ) +
        theme(text = element_text(size = 7),
              axis.text = element_text(size = 7)) +
        geom_text(aes(label = nSharedGenes), size = 2) +
        coord_fixed()

plot1
```

### Potential map between scatter and monocle clusters 

| monocle | scatter | annotation                |
| ------- | ------- | ------------------------- |
| 3       | 1       | gonadal ridge fibroblasts |
| 5       | 2       | kidney cells              |
| 4       | 3       | Granulosa cells           |
| 1       | 4       | gonadal ridge epithelial  |
| 13      | 5       | oogonia                   |
| 15      | 5       | oocytes                   |
| 18      | 5       | preoogonia                |
| 26      | 5       | embryobic stem cells      |
| 8       | 6       | germ cell nest epithelial |
| 2       | 7       | primary spermatocyte      |
| 16      | 7       | kidney epithelial         |
| 9       | 8       | erythrocytes              |
| 19      | 8       | erythropoietic stem cells |
| 21      | 8       | erythrocytes              |
| 1       | 9       | gonadal ridge epithelial  |
| 10      | 10      | b-lymphocytes             |
| 6       | 11      | hemopoietic stem cells    |
| 17      | 12      | kidney podocytes          |
| 7       | 13      | gonadal ridge macrophages |
| 1       | 14      | gonadal ridge epithelial  |
| 11      | 15      | proliferating b-cells     |
| 12      | 16      | spermatogonia             |
| 2       | 17      | primary spermatocyte      |
| 14      | 17      | prespermatogonia          |
| 2       | 18      | primary spermatocyte      |
| 2       | 19      | primary spermatocyte      |
| 14      | 19      | PGCs                      |

## Trajectory analysis

The TSCAN algorithm is a straightforward way to track how cell states change over time. First, it groups similar cells  clusters. Then, it finds the center of each cluster by averaging the positions of its members. Finally, it connects these centers in the simplest way possible using a structure called a minimum spanning tree (MST). This MST creates a path that links all the clusters without forming loops, effectively showing how the data moves from one cluster to another.

### In TSNE space

```{r}
by.cluster <- aggregateAcrossCells(merged.sce, ids=colLabels(merged.sce), use.assay.type = "reconstructed")

centroids <- reducedDim(by.cluster, "corrected")

mst <- createClusterMST(centroids, clusters=NULL)

line.data <- reportEdges(by.cluster, mst=mst, clusters=NULL, use.dimred="TSNE")

plotTSNE(merged.sce, colour_by="label") + 
    geom_line(data=line.data, mapping=aes(x=TSNE1, y=TSNE2, group=edge))

```

```{r, eval=FALSE}
pseudo <- quickPseudotime(merged.sce, use.dimred="corrected", outgroup=FALSE)
saveRDS(pseudo, file = paste0(results_dir, "/all.scatter.pseudo.Rds"))
```

### By cluster

Here, each cluster is represented by a node.

```{r, fig.height=8, fig.width=8}
pseudo = readRDS(paste0(results_dir, "/all.scatter.pseudo.Rds"))
sample_list = list(F=c(3,8), M=c(19,18,17,16), U=c(1,9,10,13,14,15), UF=c(2,4,5,6,11,12), UMF=c(7))
sample_vec = unlist(sample_list)
sample_vec = sample_vec[order(sample_vec)]
sample_groups <- gsub("[0-9]", "", names(sample_vec))

smp = c("F", "M", "U", "UF", "UMF")
smp_i = match(sample_groups, smp)
colr= c("white", "#D9DDDC", "#D6CFC7", "#787276", "#3E424B")
sample_colors = colr[smp_i]

library(igraph)
V(pseudo$mst)$group = sample_groups


V(pseudo$mst)$color <- sample_colors

group_list = list(erythrocytes = 8,
                  germ.cell.nest = 6,
                  gonadal.ridge = c(1,4,9,14),
                  granulosa.cells = 3,
                  hemopoietic.stem.cells = 11,
                  immune = c(10, 15,13),
                  kidney = c(2,12),
                  oogonia.oocytes = 5,
                  early.sperm = c(7, 17, 18,19,16))

  plot(pseudo$mst, mark.groups = group_list, size = 20)
  legend("topright", 
         legend=c("U", "UF", "F", "UM", "M"), 
         col= colr,
         pt.bg =colr,
         pch=21, pt.cex=2, bty="n")
  legend("topleft", 
         legend=names(group_list), 
         col=rainbow(length(group_list)), 
         pch=21, 
         pt.bg=rainbow(length(group_list)), 
         pt.cex=2, bty="n")
```

# Sex cell development specific analysis

Filter for cells in clusters 7, 16, 17, 18, 19, 5 

```{r, eval=F}
# Load QC'd data
sample_dirs = list.dirs(data_dir, full.names = TRUE, recursive = FALSE)
sample_names = basename(sample_dirs)

cds_list = list()

for(i in 1:length(sample_dirs)){
  
  cds_list[[i]] = readRDS(paste0(sample_dirs[i], 
                                 "/", 
                                 sample_names[i],
                                 "_cds_mito_filter.Rds"))
}

names(cds_list) = sample_names

meta_df = as.data.frame(colData(merged.sce))
meta_df$cells = rownames(meta_df)

all.sex.sce = list()

# cds to sce
for (cds in names(cds_list)){
  
  raw_counts = counts(cds_list[[cds]])
  
  all.sce = 
    SingleCellExperiment(list(counts=raw_counts), 
                         colData=cds_list[[cds]]@colData)
  
  all.sce@colData$sample = cds
  
  sex.cells = 
   meta_df %>%
   filter(label %in% c(7, 16, 17, 18, 19, 5 ),
          batch == cds) %>%
   pull(cells)
  
  all.sex.sce[[cds]] = all.sce[,sex.cells]
}

# https://bioconductor.org/books/3.14/OSCA.multisample/human-pbmcs-10x-genomics.html
# normalization, feature selection, dim reduction
set.seed(1)
all.sex.sce <- lapply(all.sex.sce, logNormCounts)
all.sex.dec <- lapply(all.sex.sce, modelGeneVar)
all.sex.hvgs <- lapply(all.sex.dec, getTopHVGs, prop=0.1)
all.sex.sce <- mapply(FUN=runPCA, x=all.sex.sce, subset_row=all.sex.hvgs)

saveRDS(all.sex.sce, file = paste0(results_dir, "/all.sex.sce.Rds"))
```

```{r}
all.sex.sce = readRDS(paste0(results_dir, "/all.sex.sce.Rds"))
all.sex.dec <- lapply(all.sex.sce, modelGeneVar)

# integration
# Intersecting the common genes.
universe <- Reduce(intersect, lapply(all.sex.sce, rownames))
all.sex.sce2 <- lapply(all.sex.sce, "[", i=universe,)
all.sex.dec2 <- lapply(all.sex.dec, "[", i=universe,)

# Renormalizing to adjust for differences in depth.
normed.sce <- do.call(multiBatchNorm, all.sex.sce2)

# Identifying a set of HVGs using stats from all batches.
combined.dec <- do.call(combineVar, all.sex.dec2)
combined.hvg <- getTopHVGs(combined.dec, n=5000)
```

```{r, eval=FALSE}
set.seed(1)
merged.sex.sce <- do.call(fastMNN, c(normed.sce, 
    list(subset.row=combined.hvg, BSPARAM=RandomParam())))

# clustering in corrected space
set.seed(1)
g <- buildSNNGraph(merged.sex.sce, use.dimred="corrected")
set.seed(1)
colLabels(merged.sex.sce) <- factor(igraph::cluster_louvain(g)$membership)

# add all cells scatter clusters
all.scatter.clust = 
  meta_df %>%
  filter(label %in% c(7, 16, 17, 18, 19, 5)) %>%
  select(cells, label) %>%
  deframe()

all.scatter.clust <- all.scatter.clust[colnames(merged.sex.sce)]
merged.sex.sce$all.merged.clusters = all.scatter.clust

# add monocle clusters
cds_nb = readRDS(paste0(results_dir, "/cds_nb.Rds"))

monocle.clusters = 
  as.data.frame(colData(cds_nb)) %>%
  filter(barcode %in% colnames(merged.sex.sce)) %>%
  select(barcode, cluster) %>%
  deframe()

monocle.clusters = monocle.clusters[colnames(merged.sex.sce)]
merged.sex.sce$monocle.clusters = monocle.clusters

merged.sex.sce <- runUMAP(merged.sex.sce, dimred="corrected")
merged.sex.sce <- runTSNE(merged.sex.sce, dimred="corrected")

saveRDS(merged.sex.sce, file = paste0(results_dir, "/original.merged.sex.sce.Rds"))
```

## New sex cell development clusters

Cells in hemopoietic stem cell, kidney-, immune-, and erythrocyte-related clusters have been filtered out. New sex cell development clusters are shown below. 

```{r}
merged.sex.sce = readRDS(paste0(results_dir, "/original.merged.sex.sce.Rds"))
plotTSNE(merged.sex.sce, colour_by="label", text_by="label", text_colour="black", point_size = .05)
```

## Split by sample

```{r, fig.height=10, fig.width=10 }
# plot UMAP

p <- lapply(unique(merged.sex.sce$batch), function(f) {
  sce.f = merged.sex.sce[, which(merged.sex.sce$batch == f)]
  plotTSNE(sce.f, colour_by="label", text_by="label", text_colour="black", point_size = .05) + ggtitle(f)
})

wrap_plots(p, ncol = 2)
```

This plot shows the proportion of cells from each sample within each new cluster.

```{r}
meta_df = as.data.frame(colData(merged.sex.sce))

sample_tally = 
  meta_df %>%
  group_by(label, batch) %>%
  tally()

ggplot(sample_tally, aes(fill=batch, y=label, x=n)) + 
    geom_bar(position="fill", stat="identity") +
    xlab("Proportion of cells")

```

## Compare with Monocle clusters 

### TSNE labeled by Monocle clusters

```{r}
plotTSNE(merged.sex.sce, colour_by="monocle.clusters", text_by="monocle.clusters", text_colour="red", point_size = .05)
```


### Table showing how many cells from the origial monocle cluster are in each new cluster

```{r}
df <- data.frame(id = colnames(merged.sex.sce),
                 sex_scatter_cluster = paste0("sex_scatter_", merged.sex.sce@colData$label),
                 monocle_cluster =  paste0("monocle_",merged.sex.sce@colData$monocle.clusters))

df_count <- 
  df %>%
  dplyr::count(monocle_cluster, sex_scatter_cluster) %>%
  group_by(monocle_cluster) %>%
  mutate(total_m1 = sum(n)) %>%
  ungroup() %>%
  group_by(sex_scatter_cluster) %>%
  mutate(total_m2 = sum(n)) %>%
  ungroup() %>%
  # Compute the percentage of total in method 1 and method 2
  mutate(
    pct_of_monocle = round((n / total_m1) * 100, 2),
    pct_of_scatter = round((n / total_m2) * 100, 2)) %>%
  select(-total_m1, -total_m2)
  
datatable(df_count, filter = "top", rownames = FALSE)
```

```{r}
# make cds for plotting with monocle

all.sex.sce2[["TM_C00004374"]]@colData$cluster="noclust" 

combined.sex.sce <- do.call(cbind, all.sex.sce2)

merged.sex.cds <- new_cell_data_set(counts(combined.sex.sce),
                                cell_metadata = merged.sex.sce@colData)

reducedDims(merged.sex.cds)$tSNE <- reducedDims(merged.sex.sce)$TSNE

```

## Plot known markers

Plot for early cell markers.

```{r, fig.width=12, fig.height=8}
rowData(merged.sex.cds)$gene_short_name = row.names(merged.sex.cds)
plot_cells(merged.sex.cds, reduction_method = "tSNE", 
           genes=chung_davidson_unique %>% 
             filter(Group %in% c("Germline cells during epigenetic reprogramming", 
                                 "Germline stem cells",
                                 "Embryonic stem cells",
                                 "Germline progenitor and amnionic cells",
                                 "Primordial Germ Cells (PGC",
                                 "Migrating germ cells",
                                 "Late PGC (meiotic entry genes")))
```

Female cell markers

```{r, fig.height=4, fig.width=12}
plot_cells(merged.sex.cds, reduction_method = "tSNE", 
           genes=chung_davidson_unique %>% 
             filter(Group %in% c("Female Germ cell", 
                                 "Mitotic oocyte",
                                 "Granulosa cell",
                                 "Oocyte")),
           show_trajectory_graph=FALSE,
           label_cell_groups=FALSE,
           label_leaves=FALSE,
           label_branch_points = FALSE)
```

Male cell markers.

```{r, fig.width=12, fig.height=8}
plot_cells(merged.sex.cds, reduction_method = "tSNE", 
           genes=chung_davidson_unique %>%  
             filter(Group %in% c("Male germ cells", 
                                 "Sertoli cell",
                                 "Mitotic Spermatogonia",
                                 "Meiotic Spermatogonia",
                                 "Spermatogonia",
                                 "Primary spermatocyte")),
           show_trajectory_graph=FALSE,
           label_cell_groups=FALSE,
           label_leaves=FALSE,
           label_branch_points = FALSE)
```

## Plot other markers of interest

### TSNE plots

```{r, fig.width=12, fig.height=12}
goi = c("BOLL", "CCDC189", "CFAP69", "NANOS1", "PIWIL1", "SPAG5", "SPAG16", "SPATA22", "TEX11", "WNT11","LOC116937338", "LOC116937873", "LOC116937887", "LOC116938414", "LOC116940726", "LOC116941541", "LOC116944806", "LOC116948775")
  
plot_cells(merged.sex.cds, reduction_method = "tSNE", 
           genes=goi,
           show_trajectory_graph=FALSE,
           label_cell_groups=FALSE,
           label_leaves=FALSE,
           label_branch_points = FALSE)
```

### Violin plots

```{r, fig.height=16, fig.width=8}
sex.scatter.clusters = as.data.frame(colData(merged.sex.sce)) 
sex.scatter.clusters$barcode = rownames(sex.scatter.clusters)
 
sex.scatter.clusters =
  sex.scatter.clusters %>%
  select(barcode, label) %>%
  deframe()

sex.scatter.clusters = sex.scatter.clusters[colnames(combined.sex.sce)]
combined.sex.sce$label = sex.scatter.clusters

plotExpression(combined.sex.sce, x = "label", features = goi)
```

# Compare scatter sex cluster enriched genes with known markers

## Cluster enriched genes

This table contains the top ~100 cluster enriched genes for each cluster. All of the cluster enriched genes are saved in `original_sex_scatter_cluster_marker_genes.csv`. The markers can help us map a cluster to a cell type.

In this case, genes in each cluster were compared to genes in every other cluster in a pairwise manner, and the p.values and logFCs of those comparisons have been summarized for ease of interpretation. 

**Column descriptions**

* `cluster`: The cluster corresponding to the enriched genes.      
* `Gene`: The enriched gene   
* `p.value`: The combined p-value across all pairwise comparisons.     
* `FDR`: The BH-adjusted p-value for each gene.   
* `summary.logFC`: This contains the summary effect size, obtained by combining effect sizes from all pairwise comparison into a single value.    

```{r}
scatter_markers = findMarkers(combined.sex.sce,
                              merged.sex.sce$label,
                              test.type="t",
                              block=combined.sex.sce$sample,
                              direction="up",
                              pval.type="some")

saveRDS(scatter_markers, file= paste0(results_dir, "/original_sex_scatter_markers.Rds"))

scatter_summary = list()

for (i in 1:length(scatter_markers)){
  
scatter_markers[[i]]$cluster = i
scatter_markers[[i]]$Gene = rownames(scatter_markers[[i]])
scatter_summary[[i]] = 
  as.data.frame(scatter_markers[[i]]) %>%
  select(cluster, Gene, p.value, FDR, summary.logFC)

}

scatter_markers_df =
  do.call(rbind, scatter_summary) %>%
  filter(FDR < .01)

write.csv(scatter_markers_df, file = paste0(results_dir, "/original_sex_scatter_cluster_marker_genes.csv"))
  
scatter_markers_top100 = 
  do.call(rbind, scatter_summary) %>%
  filter(FDR < .01) %>%
  group_by(cluster) %>%
  slice_min(FDR, n = 100, with_ties=TRUE)
  
datatable(scatter_markers_top100, filter = "top", rownames = TRUE)  
```

## Comparison with known markers heatmap

We'll check for the significance of the overlap between the top 100 cluster enriched genes and the supplied known cell-type markers. In the plot below, the rows represent cell types with know marker genes, and the columns represent the cell clusters. The number in each box is the number of genes shared between the top 100 cluster enriched genes for the specified clusters and the specified known cell-type markers. Each box is colored by the significance of the overlap ($-log10(FDR)$). Significant overlaps may signify that the cells in the cluster belong to the specified cell type. 

```{r, fig.width=8}
marker_test_genes =
  scatter_markers_top100 %>%
  select(Gene, cluster) %>%
  mutate(cluster = paste0("cluster", cluster))

cluster_ol =  phyperOverlapSets(chung_davidson_cell_type_markers, 
                                marker_test_genes, length(merged.sce))

write.csv(cluster_ol$intersecting.genes, file = paste0(results_dir, "/original_sex_scatter_cluster_marker_gene_overlaps.csv"), row.names = FALSE) 

scores = cluster_ol$pval.scores

scores = 
  scores %>%
  rename(Group1 = "CellTypes",
         Group2 = "Clusters") %>%
  group_by(CellTypes) %>%
    mutate(FDR = p.adjust(Pval, "BH"))

scores$FDR[scores$FDR > .01] = 1

FDR_wide <- 
  scores %>%
  mutate(logFDR = -log10(FDR)) %>%
  pivot_wider(id_cols = CellTypes, 
              names_from = Clusters, 
              values_from=logFDR)

CellTypes = FDR_wide$CellTypes
FDR_wide$CellTypes = NULL

FDR_wide = as.matrix(FDR_wide)
rownames(FDR_wide) = CellTypes

ord_row <- hclust( dist(FDR_wide), method = "ward.D" )$order
ord_col <- hclust( dist(t(FDR_wide)), method = "ward.D" )$order

scores$CellTypes = factor(scores$CellTypes, levels = rownames(FDR_wide)[ord_row])
scores$Clusters = factor(scores$Clusters, levels = colnames(FDR_wide)[ord_col])

plot1 =
    ggplot(scores,
           aes(Clusters, CellTypes)) + 
           geom_tile(aes(fill = -log10(FDR)), 
                     color = "light gray") +
        #Creating legend
        guides(fill=guide_colorbar(title = "-log10(FDR)",
                                   barheight = 10,
                                   barwidth = 2,
                                   default.unit = "mm")) +
        #Creating color range
        scale_fill_gradientn(colors=c("white",
                                      "#fd9668",
                                      "#f1605d",
                                      "#de4968",
                                      "#9e2f7f"),
                             guide="colorbar",
                             limits = c(0, ceiling(-log10(min(scores$FDR))))
                             ) +
        #Rotating labels
         theme(axis.text.x = element_text(angle = 30, 
                                          hjust = 1,vjust=1),
               ) +
        theme(text = element_text(size = 7),
              axis.text = element_text(size = 7)) +
        geom_text(aes(label = nSharedGenes), size = 1) +
        coord_fixed()

plot1

ggsave(plot1, file = paste0(results_dir, "/original_sex_scatter_cluster_overlap_heatmap.png"))
```

# Gene ontology of top 100 scatter cluster enriched genes 

To help annotate cell type clusters, we can check for enriched gene ontology categories among cell-type enriched genes.

*Columns*

* `GOID`: Gene ontology ID            
* `GOName`: Gene ontology name    
* `nGOTermGenes`: Number of genes assigned to GO term        
* `Clusters`: Cluster   
* `nClusterEnrichedGenes`: Number of cluster enriched genes      
* `nSharedGenes`: Number of shared genes between the GO term and cluster enriched genes   
* `Pval`: p-value          
* `FDR`: p-value corrected for multiple comparisons            

```{r}
gaf = read.delim("/mnt/research/bioinformaticsCore/shared/Genomes/Petromyzon_marinus/GCF_010993605.1_kPetMar1.pri_gene_ontology.gaf", header=F, comment = "!")

gaf <- 
  gaf %>%
  filter(!grepl("NOT", V4)) %>%  
  dplyr::select(V3, V5) %>%      
  dplyr::rename(GOID = "V5",            
                Gene = "V3") %>%
  distinct() 

keep_go = 
  gaf %>%
  group_by(GOID) %>%
  tally() %>%
  filter(n > 10,
         n < 200) %>%
  pull(GOID)

gaf = 
  gaf %>%
  filter(GOID %in% keep_go)

go_names = read.delim("/mnt/research/bioinformaticsCore/shared/Genomes/Petromyzon_marinus/GOs_names.txt", header = F)
colnames(go_names) = c("GOID", "GOName")
  
go_ol =  phyperOverlapSets(gaf, 
                           marker_test_genes, length(merged.sce))

write.csv(go_ol$intersecting.genes, file = paste0(results_dir, "/original_sex_scatter_cluster_marker_gene_GO_overlaps.csv"), row.names = FALSE) 

scores = go_ol$pval.scores

scores = 
  scores %>%
  dplyr::rename(GOID = "Group1",
                Clusters = "Group2",
                nGOTermGenes = "nGroup1",
                nClusterEnrichedGenes = "nGroup2") %>%
  group_by(Clusters) %>%
    mutate(FDR = p.adjust(Pval, "BH")) %>%
  select(-Enrichment)

scores = 
  left_join(go_names, scores, by="GOID") %>%
  filter(!is.na(nGOTermGenes)) %>%
  filter(Pval < .01)

datatable(scores, filter = "top", rownames = FALSE)
```

# Sex cell developement trajectory analysis

## Trajectory shown in TSNE space

TSNE is best for discerning clusters, but not so great for visualizing trajectories. 

```{r}
set.seed(1)
by.cluster <- aggregateAcrossCells(merged.sex.sce, 
                                   ids=colLabels(merged.sex.sce), use.assay.type = "reconstructed")

centroids <- reducedDim(by.cluster, "corrected")

mst <- createClusterMST(centroids, clusters=NULL)

line.data <- reportEdges(by.cluster, mst=mst, clusters=NULL, use.dimred="TSNE")

plotTSNE(merged.sex.sce, colour_by="label", point_size = .5) + 
    geom_line(data=line.data, mapping=aes(x=TSNE1, y=TSNE2, group=edge))
```

## Trajectory shown in batch corrected gene expression space

It's hard to interpret this in TSNE space, so we'll plot in batch corrected expression space instead because the clusters and trajectory analysis were actually calculated in batch corrected expression space. 

```{r}
set.seed(1)
line.data_corrected <- reportEdges(by.cluster, mst=mst, clusters=NULL, use.dimred="corrected")

plotReducedDim(merged.sex.sce, colour_by="label", dimred = "corrected", point_size = .4) + 
    geom_line(data=line.data_corrected, mapping=aes(x=dim1, y=dim2, group=edge))
```

## Trajectory plotted with each cluster as a node

Simplify even further. 

```{r, fig.width=8}
pseudo <- quickPseudotime(merged.sex.sce, use.dimred="corrected", outgroup=FALSE, start=1)

# plot(pseudo$mst)

sample_list = list(M=c(2, 6:12), UF=c(1,3,4,5), UM=c(13))
sample_vec = unlist(sample_list)
sample_vec = sample_vec[order(sample_vec)]
sample_groups <- gsub("[0-9]", "", names(sample_vec))

smp = c("M", "UF", "UM")
smp_i = match(sample_groups, smp)
colr= c("white", "#D9DDDC", "#787276")
sample_colors = colr[smp_i]

library(igraph)
V(pseudo$mst)$group = sample_groups

V(pseudo$mst)$color <- sample_colors

group_list = list(preoogonia = 1,
                  oogonia.oocytes = 4,
                  primary.spermatocyte = c(2, 7:13),
                  spermatogonia = 6,
                  kidney.epithelial = 5,
                  kidney.cilliated.cells = 3)

plot(pseudo$mst, mark.groups = group_list, size = 20)
legend("topright", 
       legend = smp, 
       col= colr,
       pt.bg = colr,
       pt.cex=2, bty="n", pch=21)
legend("topleft", 
       legend=names(group_list), 
       col=rainbow(length(group_list)), 
       pch=21, 
       pt.bg=rainbow(length(group_list)), 
       pt.cex=2, bty="n")

saveRDS(pseudo, file = paste0(results_dir, "/original_sex_pseudo.Rds"))
```

## Plot pseudotime in tsne space

```{r}
common.pseudo <- averagePseudotime(pseudo$ordering) 
plotTSNE(merged.sex.sce, colour_by=I(common.pseudo), 
        text_by="label", text_colour="red",  point_size = .5) +
    geom_line(data=line.data, mapping=aes(x=TSNE1, y=TSNE2, group=edge))
```

## Plot pseudotime in batch corrected expression space

It's hard to interpret this in TSNE space, so we'll plot in batch corrected expression space instead because the clusters and trajectory analysis were actually calculated in batch corrected expression space. 

```{r}
line.data_corrected <- reportEdges(by.cluster, mst=mst, clusters=NULL, use.dimred="corrected")

plotReducedDim(merged.sex.sce, 
               colour_by=I(common.pseudo), 
               dimred = "corrected", point_size = .4) + 
  geom_line(data=line.data_corrected, mapping=aes(x=dim1, y=dim2, group=edge))
```

## Genes changing along each path

Top ~100 up and down-regulated genes by FDR changing with pseudotime along each path. These genes can be found in `/orig_sex_pseudotime_de_genes.csv`

```{r}
pseudo_mat = pathStat(pseudo$ordering)
test_pseudo <- TSCAN::testPseudotime(combined.sex.sce, 
                              pseudotime=pseudo_mat)

test_pseudo_filt = list()

for (i in 1:length(test_pseudo)){
  
  test_pseudo[[i]]$SYMBOL <- rownames(combined.sex.sce)
  test_pseudo[[i]]$Path <- names(test_pseudo)[i]
  
  test_pseudo_filt[[i]] =
    as.data.frame(test_pseudo[[i]]) %>%
    filter(FDR < .01) %>%
    arrange(desc(abs(logFC)))
  
}

test_pseudo_filt_df = do.call(rbind, test_pseudo_filt)
test_pseudo_filt_df = test_pseudo_filt_df[, c("Path", "SYMBOL", "logFC", "p.value", "FDR")]
write.csv(test_pseudo_filt_df, file = paste0(results_dir, "/orig_sex_pseudotime_de_genes.csv"))

test_pseudo100 = 
  test_pseudo_filt_df %>%
  mutate(Path = case_when(
    logFC > 0 ~ paste0(Path, ".Upregulated"),
    logFC < 0 ~  paste0(Path, ".Downregulated"))
    ) %>%
  group_by(Path) %>%
  slice_min(order_by = FDR, n=100)
  

datatable(test_pseudo100, filter = "top", rownames = FALSE)

```

## Genes differentially expressed immediately following the branch point in cluster 1

A more advanced analysis involves looking for differences in expression between paths of a branched trajectory. This is most interesting for cells close to the branch point between two or more paths where the differential expression analysis may highlight genes responsible for the branching event. The general strategy here is to fit one trend to the unique part of each path immediately following the branch point, followed by a comparison of the fits between paths.

These plots show pseudotime in cells in cluster 1 along paths 2 and 8 which separate male and female trajectories. 

### In TSNE space:

```{r, fig.height=4, fig.width=8}
combined.sex.sce$TSCAN.4 <- pathStat(pseudo$ordering)[,1]
combined.sex.sce$TSCAN.3 <- pathStat(pseudo$ordering)[,2]
combined.sex.sce$TSCAN.5 <- pathStat(pseudo$ordering)[,3]
combined.sex.sce$TSCAN.9 <- pathStat(pseudo$ordering)[,4]
combined.sex.sce$TSCAN.13 <- pathStat(pseudo$ordering)[,5]
combined.sex.sce$TSCAN.6 <- pathStat(pseudo$ordering)[,6]

merged.sex.sce$TSCAN.4 <- pathStat(pseudo$ordering)[,1]
merged.sex.sce$TSCAN.3 <- pathStat(pseudo$ordering)[,2]
merged.sex.sce$TSCAN.5 <- pathStat(pseudo$ordering)[,3]
merged.sex.sce$TSCAN.9 <- pathStat(pseudo$ordering)[,4]
merged.sex.sce$TSCAN.13 <- pathStat(pseudo$ordering)[,5]
merged.sex.sce$TSCAN.6 <- pathStat(pseudo$ordering)[,6]

# Subsetting to the desired cluster containing the branch point.
starter=1
keep <- colLabels(merged.sex.sce) == starter
combined.sex.sce1 <- combined.sex.sce[,keep]
merged.sex.sce1 <- merged.sex.sce[,keep]

# Showing only the lines to/from our cluster of interest.
line.data.sub <- line.data[grepl("^1--", line.data$edge) | grepl("--1$", line.data$edge),]
ggline <- geom_line(data=line.data.sub, mapping=aes(x=TSNE1, y=TSNE2, group=edge))

gridExtra::grid.arrange(
    plotTSNE(merged.sex.sce1, colour_by="TSCAN.4", point_size = .3) + ggline,
    plotTSNE(merged.sex.sce1, colour_by="TSCAN.9", point_size = .3) + ggline,
    ncol=2
)
```

### In batch corrected expression space:

It's hard to interpret this in TSNE space, so we'll plot in batch corrected expression space instead because the clusters and trajectory analysis were actually calculated in batch corrected expression space. 

```{r, fig.height=4, fig.width=8}
line.data.sub_corrected = line.data_corrected[grepl("^1--", line.data_corrected$edge) | grepl("--1$", line.data_corrected$edge),]
ggline_corrected <- geom_line(data=line.data.sub_corrected, mapping=aes(x=dim1, y=dim2, group=edge))

gridExtra::grid.arrange(
    plotTSNE(merged.sex.sce1, dimred = "corrected", colour_by="TSCAN.4", point_size = .5) + ggline_corrected,
    plotTSNE(merged.sex.sce1, dimred = "corrected", colour_by="TSCAN.9", point_size = .5) + ggline_corrected,
    ncol=2
)
```

```{r}
pseudo_test4 <- testPseudotime(combined.sex.sce1, df=1, pseudotime=combined.sex.sce1$TSCAN.4)
pseudo_test4$SYMBOL <- rowData(combined.sex.sce1)$SYMBOL
#pseudo_test2[order(pseudo_test2$p.value),]
```

```{r}
pseudo_test9 <- testPseudotime(combined.sex.sce1, df=1, pseudotime=combined.sex.sce1$TSCAN.9)
pseudo_test9$SYMBOL <- rowData(combined.sex.sce1)$SYMBOL
#pseudo_test4[order(pseudo_test4$p.value),]
```

## Genes that are significantly changing along path 4 and are not significant and/or changing in the opposite direction in path 9. 

Saved in `original_sex_onl4_DEgenes.csv`

```{r}
only4 <- pseudo_test4[which(pseudo_test4$FDR <= 0.05 & 
    (pseudo_test9$p.value >= 0.05 | sign(pseudo_test4$logFC)!=sign(pseudo_test9$logFC))),]

only4 = as.data.frame(only4) 
only4$Gene = rownames(only4) 
only4 = only4[c(4,1,2,3)]

write.csv(only4, file = paste0(results_dir, "/original_sex_onl4_DEgenes.csv"), row.names = FALSE) 

datatable(only4[order(only4$p.value),], filter="top", rownames = FALSE)
```

### Top 50 heatmap

```{r, fig.height=8, fig.width=6}
on.path4 <- !is.na(combined.sex.sce1$TSCAN.4)

top50 =
  only4 %>%
  slice_min(order_by = FDR, n=50) %>%
  pull(Gene)

plotHeatmap(combined.sex.sce1[,on.path4], 
            order_columns_by="TSCAN.4", 
            colour_columns_by="TSCAN.4", 
            features=top50,
            center=TRUE)
```

### Same genes on path 9

There should be no correlation between pseudotime and scaled expression or the opposite correlation.

```{r, fig.height=8, fig.width=6}
on.path9 <- !is.na(combined.sex.sce1$TSCAN.9)

plotHeatmap(combined.sex.sce1[,on.path9], 
            order_columns_by="TSCAN.9", 
            colour_columns_by="TSCAN.9", 
            features=top50,
            center=TRUE)
```

## Gene ontology enrichment of genes uniquely changing on path 4 vs 9

**Columns** 

* `GOID`: Gene ontology ID            
* `GOName`: Gene ontology name    
* `nGOTermGenes`: Number of genes assigned to GO term        
* `Path`: Path and whether the gene is up or down regulated      
* `nPathDEGenes`: Number of genes differentially expressed on the path         
* `nSharedGenes`: Number of shared genes between the GO term and path DE genes   
* `Pval`: p-value          
* `FDR`: p-value corrected for multiple comparisons        

These results are save in `results/origsex_only4_gene_GO_pvals.csv`. Overlapping genes are save in `results/origsex_only4_gene_GO_overlaps.csv`

```{r}

only4_genes = 
  only4 %>%
  mutate(Path = case_when(
    logFC > 0 ~ "Only4.Upregulated",
    logFC < 0 ~  "Only4.Downregulated")
    ) %>%
  select(Gene, Path)
  
  
go_ol =  phyperOverlapSets(gaf, 
                           only4_genes, length(merged.sce))

write.csv(go_ol$intersecting.genes, file = paste0(results_dir, "/origsex_only4_gene_GO_overlaps.csv"), row.names = FALSE) 

scores = go_ol$pval.scores

scores = 
  scores %>%
  dplyr::rename(GOID = "Group1",
                Path = "Group2",
                nGOTermGenes = "nGroup1",
                nPathDEGenes = "nGroup2") %>%
    mutate(FDR = p.adjust(Pval, "BH")) %>%
  select(-Enrichment)

scores = 
  left_join(go_names, scores, by="GOID") %>%
  filter(!is.na(nGOTermGenes)) %>%
  filter(Pval < .01)

datatable(scores, filter = "top", rownames = FALSE)
```

## Genes that are significantly changing along path 9 and are not significant and/or changing in the opposite direction in path 4. 

Saved in `original_sex_only9_DEgenes.csv`. 

```{r}
only9 <- pseudo_test9[which(pseudo_test9$FDR <= 0.05 & 
    (pseudo_test4$p.value >= 0.05 | sign(pseudo_test4$logFC)!=sign(pseudo_test9$logFC))),]

only9 = as.data.frame(only9) 
only9$Gene = rownames(only9) 
only9 = only9[c(4,1,2,3)]

write.csv(only9, file = paste0(results_dir, "/original_sex_only9_DEgenes.csv"), row.names = FALSE) 

datatable(only9[order(only9$p.value),], filter="top", rownames = FALSE)
```

### Top 50 heatmap

```{r, fig.height=8, fig.width=6}
on.path9 <- !is.na(combined.sex.sce1$TSCAN.9)

top50 =
  only9 %>%
  slice_min(order_by = FDR, n=50) %>%
  pull(Gene)

plotHeatmap(combined.sex.sce1[,on.path9], 
            order_columns_by="TSCAN.9", 
            colour_columns_by="TSCAN.9", 
            features=top50,
            center=TRUE)
```

### Same genes on path 4. 

These seem to be going the opposite direction.

```{r, fig.height=8, fig.width=6}
plotHeatmap(combined.sex.sce1[,on.path4], 
            order_columns_by="TSCAN.4", 
            colour_columns_by="TSCAN.4", 
            features=top50,
            center=TRUE)
```

## Gene ontology enrichment of genes uniquely changing on path 9 vs 4

**Columns** 

* `GOID`: Gene ontology ID            
* `GOName`: Gene ontology name    
* `nGOTermGenes`: Number of genes assigned to GO term        
* `Path`: Path and whether the gene is up or down regulated      
* `nPathDEGenes`: Number of genes differentially expressed on the path         
* `nSharedGenes`: Number of shared genes between the GO term and path DE genes   
* `Pval`: p-value          
* `FDR`: p-value corrected for multiple comparisons        

These results are save in `results/origsex_only9_gene_GO_pvals.csv`. Overlapping genes are save in `results/origsex_only9_gene_GO_overlaps.csv`

```{r}
only9_genes = 
  only9 %>%
  mutate(Path = case_when(
    logFC > 0 ~ "Only9.Upregulated",
    logFC < 0 ~  "Only9.Downregulated")
    ) %>%
  select(Gene, Path)
  
go_ol =  phyperOverlapSets(gaf, 
                           only9_genes, length(merged.cds))

write.csv(go_ol$intersecting.genes, file = paste0(results_dir, "/origsex_only9_gene_GO_overlaps.csv"), row.names = FALSE) 

scores = go_ol$pval.scores

scores = 
  scores %>%
  dplyr::rename(GOID = "Group1",
                Path = "Group2",
                nGOTermGenes = "nGroup1",
                nPathDEGenes = "nGroup2") %>%
    mutate(FDR = p.adjust(Pval, "BH")) %>%
  select(-Enrichment)

scores = 
  left_join(go_names, scores, by="GOID") %>%
  filter(!is.na(nGOTermGenes)) %>%
  filter(Pval < .01)

datatable(scores, filter = "top", rownames = FALSE)
```

# Remove cluster 2, 3 and 5 cells

I think cluster 2 might be sertoli cells. Cluster 5 is mostly made up of cells you originally annotated as kidney endothelial cells. Cluster 3 is mostly made up of cells you originally annotated as kidney cilliated cells.

```{r, eval=FALSE}
orig.combined.sex.sce = combined.sex.sce

meta_df = as.data.frame(colData(merged.sex.sce)) 
meta_df$cells = rownames(meta_df)

keep_cells = 
  meta_df %>%
  filter(!label %in% c(2,3,5)) %>%
  pull(cells)
  
# Load QC'd data
sample_dirs = list.dirs(data_dir, full.names = TRUE, recursive = FALSE)
sample_names = basename(sample_dirs)

cds_list = list()

for(i in 1:length(sample_dirs)){
  
  cds_list[[i]] = readRDS(paste0(sample_dirs[i], 
                                 "/", 
                                 sample_names[i],
                                 "_cds_mito_filter.Rds"))
}

names(cds_list) = sample_names

meta_df = as.data.frame(colData(merged.sce))
meta_df$cells = rownames(meta_df)

all.sex.sce = list()

# cds to sce
for (cds in names(cds_list)){
  
  raw_counts = counts(cds_list[[cds]])
  
  all.sce = 
    SingleCellExperiment(list(counts=raw_counts), 
                         colData=cds_list[[cds]]@colData)
  
  all.sce@colData$sample = cds
  
  sex.cells = 
   meta_df %>%
   filter(cells %in% keep_cells,
          batch == cds) %>%
   pull(cells)
  
  all.sex.sce[[cds]] = all.sce[,sex.cells]
}

# https://bioconductor.org/books/3.14/OSCA.multisample/human-pbmcs-10x-genomics.html
# normalization, feature selection, dim reduction
set.seed(1)
all.sex.sce <- lapply(all.sex.sce, logNormCounts)
all.sex.dec <- lapply(all.sex.sce, modelGeneVar)
all.sex.hvgs <- lapply(all.sex.dec, getTopHVGs, prop=0.1)
all.sex.sce <- mapply(FUN=runPCA, x=all.sex.sce, subset_row=all.sex.hvgs)

saveRDS(all.sex.sce, file = paste0(results_dir, "/fitered.all.sex.sce.Rds"))
```

```{r, eval=FALSE}
orig.merged.sex.sce = merged.sex.sce

all.sex.sce = readRDS(paste0(results_dir, "/fitered.all.sex.sce.Rds"))
all.sex.dec <- lapply(all.sex.sce, modelGeneVar)

# integration
# Intersecting the common genes.
universe <- Reduce(intersect, lapply(all.sex.sce, rownames))
all.sex.sce2 <- lapply(all.sex.sce, "[", i=universe,)
all.sex.dec2 <- lapply(all.sex.dec, "[", i=universe,)

# Renormalizing to adjust for differences in depth.
normed.sce <- do.call(multiBatchNorm, all.sex.sce2)

# Identifying a set of HVGs using stats from all batches.
combined.dec <- do.call(combineVar, all.sex.dec2)
combined.hvg <- getTopHVGs(combined.dec, n=5000)

set.seed(1)
merged.sex.sce <- do.call(fastMNN, c(normed.sce, 
    list(subset.row=combined.hvg, BSPARAM=RandomParam())))

# clustering in corrected space
set.seed(1)
g <- buildSNNGraph(merged.sex.sce, use.dimred="corrected", k = 8)
set.seed(1)
colLabels(merged.sex.sce) <- factor(igraph::cluster_louvain(g)$membership)

orig.sex.scatter.clust = colLabels(orig.merged.sex.sce)
names(orig.sex.scatter.clust) = colnames(orig.merged.sex.sce)
orig.sex.scatter.clust <- orig.sex.scatter.clust[colnames(merged.sex.sce)]
merged.sex.sce$orig.sex.scatter.clust = orig.sex.scatter.clust

# add monocle clusters
cds_nb = readRDS(paste0(results_dir, "/cds_nb.Rds"))

monocle.clusters = 
  as.data.frame(colData(cds_nb)) %>%
  filter(barcode %in% colnames(merged.sex.sce)) %>%
  select(barcode, cluster) %>%
  deframe()

monocle.clusters = monocle.clusters[colnames(merged.sex.sce)]
merged.sex.sce$monocle.clusters = monocle.clusters

merged.sex.sce <- runUMAP(merged.sex.sce, dimred="corrected")
merged.sex.sce <- runTSNE(merged.sex.sce, dimred="corrected")

saveRDS(merged.sex.sce, file = paste0(results_dir, "/filtered.merged.sex.sce.Rds"))
```

## Plot new clusters in TSNE space

```{r}
merged.sex.sce = readRDS(paste0(results_dir, "/filtered.merged.sex.sce.Rds"))
plotTSNE(merged.sex.sce, colour_by="label", text_by="label", text_colour="black", point_size = .05)
```

## Split by sample

```{r, fig.height=10, fig.width=10 }
# plot UMAP

p <- lapply(unique(merged.sex.sce$batch), function(f) {
  sce.f = merged.sex.sce[, which(merged.sex.sce$batch == f)]
  plotTSNE(sce.f, colour_by="label", text_by="label", text_colour="black", point_size = .05) + ggtitle(f)
})

wrap_plots(p, ncol = 2)
```

This plot shows the proportion of cells from each sample within each new cluster.

```{r}
meta_df = as.data.frame(colData(merged.sex.sce))

sample_tally = 
  meta_df %>%
  group_by(label, batch) %>%
  tally()

ggplot(sample_tally, aes(fill=batch, y=label, x=n)) + 
    geom_bar(position="fill", stat="identity") +
    xlab("Proportion of cells")

```

## Compare with monocle clusters

### TSNE labeled by Monocle clusters

```{r}
plotTSNE(merged.sex.sce, colour_by="monocle.clusters", text_by="monocle.clusters", text_colour="red", point_size = .05)
```

### Table showing how many cells from the origial monocle cluster are in each new cluster

```{r}
df <- data.frame(id = colnames(merged.sex.sce),
                 sex_scatter_cluster = paste0("sex_scatter_", merged.sex.sce@colData$label),
                 monocle_cluster =  paste0("monocle_",merged.sex.sce@colData$monocle.clusters))

df_count <- 
  df %>%
  dplyr::count(monocle_cluster, sex_scatter_cluster) %>%
  group_by(monocle_cluster) %>%
  mutate(total_m1 = sum(n)) %>%
  ungroup() %>%
  group_by(sex_scatter_cluster) %>%
  mutate(total_m2 = sum(n)) %>%
  ungroup() %>%
  # Compute the percentage of total in method 1 and method 2
  mutate(
    pct_of_monocle = round((n / total_m1) * 100, 2),
    pct_of_scatter = round((n / total_m2) * 100, 2)) %>%
  select(-total_m1, -total_m2)

datatable(df_count, filter = "top", rownames = FALSE)

```

```{r}
# make cds for plotting with monocle
all.sex.sce = readRDS(paste0(results_dir, "/fitered.all.sex.sce.Rds"))
all.sex.dec <- lapply(all.sex.sce, modelGeneVar)

# integration
# Intersecting the common genes.
universe <- Reduce(intersect, lapply(all.sex.sce, rownames))
all.sex.sce2 <- lapply(all.sex.sce, "[", i=universe,)
all.sex.dec2 <- lapply(all.sex.dec, "[", i=universe,)

all.sex.sce2[["TM_C00004374"]]@colData$cluster="noclust" 

combined.sex.sce <- do.call(cbind, all.sex.sce2)

saveRDS(combined.sex.sce, file = paste0(results_dir, "/fitered.combined.sex.sce.Rds"))

merged.sex.cds <- new_cell_data_set(counts(combined.sex.sce),
                                cell_metadata = merged.sex.sce@colData)

reducedDims(merged.sex.cds)$tSNE <- reducedDims(merged.sex.sce)$TSNE
rowData(merged.sex.cds)$gene_short_name = row.names(merged.sex.cds)

```

## Plot known markers

Plot for early cell markers.

```{r, fig.width=12, fig.height=8}
plot_cells(merged.sex.cds, reduction_method = "tSNE", 
           genes=chung_davidson_unique %>% 
             filter(Group %in% c("Germline cells during epigenetic reprogramming", 
                                 "Germline stem cells",
                                 "Embryonic stem cells",
                                 "Germline progenitor and amnionic cells",
                                 "Primordial Germ Cells (PGC",
                                 "Migrating germ cells",
                                 "Late PGC (meiotic entry genes")))
```

Female cell markers

```{r, fig.height=4, fig.width=12}
plot_cells(merged.sex.cds, reduction_method = "tSNE", 
           genes=chung_davidson_unique %>% 
             filter(Group %in% c("Female Germ cell", 
                                 "Mitotic oocyte",
                                 "Granulosa cell",
                                 "Oocyte")),
           show_trajectory_graph=FALSE,
           label_cell_groups=FALSE,
           label_leaves=FALSE,
           label_branch_points = FALSE)
```

Male cell markers.

```{r, fig.width=12, fig.height=8}
plot_cells(merged.sex.cds, reduction_method = "tSNE", 
           genes=chung_davidson_unique %>%  
             filter(Group %in% c("Male germ cells", 
                                 "Sertoli cell",
                                 "Mitotic Spermatogonia",
                                 "Meiotic Spermatogonia",
                                 "Spermatogonia",
                                 "Primary spermatocyte")),
           show_trajectory_graph=FALSE,
           label_cell_groups=FALSE,
           label_leaves=FALSE,
           label_branch_points = FALSE)
```

## Plot other markers of interest

### TSNE plots

```{r, fig.width=12, fig.height=12}
goi = c("BOLL", "CCDC189", "CFAP69", "NANOS1", "PIWIL1", "SPAG5", "SPAG16", "SPATA22", "TEX11", "WNT11","LOC116937338", "LOC116937873", "LOC116937887", "LOC116938414", "LOC116940726", "LOC116941541", "LOC116944806", "LOC116948775")
  
plot_cells(merged.sex.cds, reduction_method = "tSNE", 
           genes=goi,
           show_trajectory_graph=FALSE,
           label_cell_groups=FALSE,
           label_leaves=FALSE,
           label_branch_points = FALSE)
```

### Violin plots

```{r, fig.height=16, fig.width=8}
sex.scatter.clusters = as.data.frame(colData(merged.sex.sce)) 
sex.scatter.clusters$barcode = rownames(sex.scatter.clusters)
 
sex.scatter.clusters =
  sex.scatter.clusters %>%
  select(barcode, label) %>%
  deframe()

sex.scatter.clusters = sex.scatter.clusters[colnames(combined.sex.sce)]
combined.sex.sce$label = sex.scatter.clusters

plotExpression(combined.sex.sce, x = "label", features = goi)
```


## Cluster enriched genes

This table contains the top ~100 cluster enriched genes for each cluster. All of the cluster enriched genes are saved in `filtered_sex_scatter_cluster_marker_genes.csv`. The markers can help us map a cluster to a cell type.

In this case, genes in each cluster were compared to genes in every other cluster in a pairwise manner, and the p.values and logFCs of those comparisons have been summarized for ease of interpretation. 

**Column descriptions**

* `cluster`: The cluster corresponding to the enriched genes.      
* `Gene`: The enriched gene   
* `p.value`: The combined p-value across all pairwise comparisons.     
* `FDR`: The BH-adjusted p-value for each gene.   
* `summary.logFC`: This contains the summary effect size, obtained by combining effect sizes from all pairwise comparison into a single value.    

```{r, eval=FALSE}
scatter_markers = findMarkers(combined.sex.sce,
                              merged.sex.sce$label,
                              test.type="t",
                              block=combined.sex.sce$sample,
                              direction="up",
                              pval.type="some")

saveRDS(scatter_markers, file= paste0(results_dir, "/filtered_sex_scatter_markers.Rds"))
```

```{r}

scatter_markers = readRDS(paste0(results_dir, "/filtered_sex_scatter_markers.Rds"))

scatter_summary = list()

for (i in 1:length(scatter_markers)){
  
scatter_markers[[i]]$cluster = i
scatter_markers[[i]]$Gene = rownames(scatter_markers[[i]])
scatter_summary[[i]] = 
  as.data.frame(scatter_markers[[i]]) %>%
  select(cluster, Gene, p.value, FDR, summary.logFC)

}

scatter_markers_df =
  do.call(rbind, scatter_summary) %>%
  mutate(cluster = paste0("cluster_",cluster)) %>%
  filter(FDR < .01)

write.csv(scatter_markers_df, file = paste0(results_dir, "/filtered_sex_scatter_cluster_marker_genes.csv"))
  
scatter_markers_top100 = 
  do.call(rbind, scatter_summary) %>%
  mutate(cluster = paste0("cluster_",cluster)) %>%
  filter(FDR < .01) %>%
  group_by(cluster) %>%
  slice_min(FDR, n = 100, with_ties=TRUE)
  
datatable(scatter_markers_top100, filter = "top", rownames = FALSE)  
```

## Comparison with known markers

We'll check for the significance of the overlap between the top 100 cluster enriched genes and the supplied known cell-type markers. In the plot below, the rows represent cell types with know marker genes, and the columns represent the cell clusters. The number in each box is the number of genes shared between the top 100 cluster enriched genes for the specified clusters and the specified known cell-type markers. Each box is colored by the significance of the overlap ($-log10(FDR)$). Significant overlaps may signify that the cells in the cluster belong to the specified cell type. 

```{r, fig.width=8}
marker_test_genes =
  scatter_markers_top100 %>%
  select(Gene, cluster) 

cluster_ol =  phyperOverlapSets(chung_davidson_cell_type_markers, 
                                marker_test_genes, length(merged.sce))

write.csv(cluster_ol$intersecting.genes, file = paste0(results_dir, "/sex_scatter_cluster_marker_gene_overlaps.csv"), row.names = FALSE) 

scores = cluster_ol$pval.scores

scores = 
  scores %>%
  rename(Group1 = "CellTypes",
         Group2 = "Clusters") %>%
  group_by(CellTypes) %>%
    mutate(FDR = p.adjust(Pval, "BH"))

scores$FDR[scores$FDR > .01] = 1

FDR_wide <- 
  scores %>%
  mutate(logFDR = -log10(FDR)) %>%
  pivot_wider(id_cols = CellTypes, 
              names_from = Clusters, 
              values_from=logFDR)

CellTypes = FDR_wide$CellTypes
FDR_wide$CellTypes = NULL

FDR_wide = as.matrix(FDR_wide)
rownames(FDR_wide) = CellTypes

ord_row <- hclust( dist(FDR_wide), method = "ward.D" )$order
ord_col <- hclust( dist(t(FDR_wide)), method = "ward.D" )$order

scores$CellTypes = factor(scores$CellTypes, levels = rownames(FDR_wide)[ord_row])
scores$Clusters = factor(scores$Clusters, levels = colnames(FDR_wide)[ord_col])

plot1 =
    ggplot(scores,
           aes(Clusters, CellTypes)) + 
           geom_tile(aes(fill = -log10(FDR)), 
                     color = "light gray") +
        #Creating legend
        guides(fill=guide_colorbar(title = "-log10(FDR)",
                                   barheight = 10,
                                   barwidth = 2,
                                   default.unit = "mm")) +
        #Creating color range
        scale_fill_gradientn(colors=c("white",
                                      "#fd9668",
                                      "#f1605d",
                                      "#de4968",
                                      "#9e2f7f"),
                             guide="colorbar",
                             limits = c(0, ceiling(-log10(min(scores$FDR))))
                             ) +
        #Rotating labels
         theme(axis.text.x = element_text(angle = 30, 
                                          hjust = 1,vjust=1),
               ) +
        theme(text = element_text(size = 7),
              axis.text = element_text(size = 7)) +
        geom_text(aes(label = nSharedGenes), size = 1) +
        coord_fixed()

plot1

ggsave(plot1, file = paste0(results_dir, "/filtered_sex_scatter_cluster_overlap_heatmap.png"))
```

## Gene ontology of top 100 scatter cluster enriched genes 

To help annotate cell type clusters, we can check for enriched gene ontology categories among cell-type enriched genes.

*Columns*

* `GOID`: Gene ontology ID            
* `GOName`: Gene ontology name    
* `nGOTermGenes`: Number of genes assigned to GO term        
* `Clusters`: Cluster   
* `nClusterEnrichedGenes`: Number of cluster enriched genes      
* `nSharedGenes`: Number of shared genes between the GO term and cluster enriched genes   
* `Pval`: p-value          
* `FDR`: p-value corrected for multiple comparisons            

```{r}
gaf = read.delim("/mnt/research/bioinformaticsCore/shared/Genomes/Petromyzon_marinus/GCF_010993605.1_kPetMar1.pri_gene_ontology.gaf", header=F, comment = "!")

gaf <- 
  gaf %>%
  filter(!grepl("NOT", V4)) %>%  
  dplyr::select(V3, V5) %>%      
  dplyr::rename(GOID = "V5",            
                Gene = "V3") %>%
  distinct() 

keep_go = 
  gaf %>%
  group_by(GOID) %>%
  tally() %>%
  filter(n > 10,
         n < 200) %>%
  pull(GOID)

gaf = 
  gaf %>%
  filter(GOID %in% keep_go)

go_names = read.delim("/mnt/research/bioinformaticsCore/shared/Genomes/Petromyzon_marinus/GOs_names.txt", header = F)
colnames(go_names) = c("GOID", "GOName")
  
go_ol =  phyperOverlapSets(gaf, 
                           marker_test_genes, length(merged.sce))

write.csv(go_ol$intersecting.genes, file = paste0(results_dir, "/sex_scatter_cluster_marker_gene_GO_overlaps.csv"), row.names = FALSE) 

scores = go_ol$pval.scores

scores = 
  scores %>%
  dplyr::rename(GOID = "Group1",
                Clusters = "Group2",
                nGOTermGenes = "nGroup1",
                nClusterEnrichedGenes = "nGroup2") %>%
  group_by(Clusters) %>%
    mutate(FDR = p.adjust(Pval, "BH")) %>%
  select(-Enrichment)

scores = 
  left_join(go_names, scores, by="GOID") %>%
  filter(!is.na(nGOTermGenes)) %>%
  filter(Pval < .01)

datatable(scores, filter = "top", rownames = FALSE)
```

# Sex cell developement trajectory analysis

## Trajectory plotted in TSNE space

```{r}
set.seed(1)
by.cluster <- aggregateAcrossCells(merged.sex.sce, 
                                   ids=colLabels(merged.sex.sce), use.assay.type = "reconstructed")

centroids <- reducedDim(by.cluster, "corrected")

mst <- createClusterMST(centroids, clusters=NULL)

line.data <- reportEdges(by.cluster, mst=mst, clusters=NULL, use.dimred="TSNE")

plotTSNE(merged.sex.sce, colour_by="label", point_size = .5) + 
    geom_line(data=line.data, mapping=aes(x=TSNE1, y=TSNE2, group=edge))
```

## Trajectory plotted in batch corrected gene expression space

It's hard to interpret this in TSNE space, so we'll plot in batch corrected expression space instead because the clusters and trajectory analysis were actually calculated in batch corrected expression space. 

```{r}
line.data_corrected <- reportEdges(by.cluster, mst=mst, clusters=NULL, use.dimred="corrected")

plotReducedDim(merged.sex.sce, colour_by="label", dimred = "corrected", point_size = .4) + 
    geom_line(data=line.data_corrected, mapping=aes(x=dim1, y=dim2, group=edge))
```

```{r, eval=FALSE}
set.seed(1)
pseudo <- quickPseudotime(merged.sex.sce, use.dimred="corrected", outgroup=FALSE, start = 1)
#plot(pseudo$mst)
saveRDS(pseudo, file=paste0(results_dir, "/filtered_sex_pseudo.Rds"))
```

## Trajectory plotted with each cluster as a node

Simplify even further. 

```{r, fig.width=8, fig.height=6}
pseudo = readRDS(paste0(results_dir, "/filtered_sex_pseudo.Rds"))
sample_list = list(M=c(4,5,7,8,9,11,12,13), U=c(15), UF=c(3), UM=c(6,10,14),  UMF =1 , F=2)
sample_vec = unlist(sample_list)
sample_vec = sample_vec[order(sample_vec)]
sample_groups <- gsub("[0-9]", "", names(sample_vec))

smp = c("M", "U", "UF", "UM", "UMF", "F")
smp_i = match(sample_groups, smp)
colr= c("white", "#D9DDDC", "#787276","#BDB7Ab", "#999Da4", "#3E424B")
sample_colors = colr[smp_i]
  
library(igraph)
V(pseudo$mst)$group = sample_groups

V(pseudo$mst)$color <- sample_colors
V(pseudo$mst)$label.color <- "red"

group_list = list(preoogonia = 1,
                  oogonia = 3,
                  oocytes = 2,
                  spermatogonia = 5,
                  primary.spermatocyte = c(4, 6:14),
                  gonadal.ridge.fibroblasts = 15)

plot(pseudo$mst, mark.groups = group_list, size = 20)
legend("topright", 
       legend=smp, 
       col= colr,
       pt.bg = colr,
       pt.cex=2, bty="n", pch=21)
legend("topleft", 
       legend=names(group_list), 
       col=rainbow(length(group_list)), 
       pch=21, 
       pt.bg=rainbow(length(group_list)), 
       pt.cex=2, bty="n")

```

## Plot pseudotime

### In tSNE space

```{r}
common.pseudo <- averagePseudotime(pseudo$ordering) 
plotTSNE(merged.sex.sce, colour_by=I(common.pseudo), 
        text_by="label", text_colour="red",  point_size = .5) +
    geom_line(data=line.data, mapping=aes(x=TSNE1, y=TSNE2, group=edge))
```

### In batch corrected gene expression space

It's hard to interpret this in TSNE space, so we'll plot in batch corrected expression space instead because the clusters and trajectory analysis were actually calculated in batch corrected expression space. 

```{r}
plotTSNE(merged.sex.sce, dimred = "corrected", colour_by=I(common.pseudo), 
        text_by="label", text_colour="red",  point_size = .5) +
     geom_line(data=line.data_corrected, mapping=aes(x=dim1, y=dim2, group=edge))
```

## Genes changing with pseudotime along each path

Top ~100 up and down-regulated genes by FDR changing with pseudotime along each path. All of the significant genes are saved in `/filt_sex_pseudotime_de_genes.csv`

```{r}
pseudo_mat = pathStat(pseudo$ordering)
test_pseudo <- TSCAN::testPseudotime(combined.sex.sce, 
                              pseudotime=pseudo_mat)

test_pseudo_filt = list()

for (i in 1:length(test_pseudo)){
  
  test_pseudo[[i]]$SYMBOL <- rownames(combined.sex.sce)
  test_pseudo[[i]]$Path <- names(test_pseudo)[i]
  
  test_pseudo_filt[[i]] =
    as.data.frame(test_pseudo[[i]]) %>%
    filter(FDR < .01) %>%
    arrange(desc(abs(logFC)))
  
}

test_pseudo_filt_df = do.call(rbind, test_pseudo_filt)
test_pseudo_filt_df = test_pseudo_filt_df[, c("Path", "SYMBOL", "logFC", "p.value", "FDR")]
write.csv(test_pseudo_filt_df, file = paste0(results_dir, "/filt_sex_pseudotime_de_genes.csv"))

test_pseudo100 = 
  test_pseudo_filt_df %>%
  mutate(Path = case_when(
    logFC > 0 ~ paste0(Path, ".Upregulated"),
    logFC < 0 ~  paste0(Path, ".Downregulated"))
    ) %>%
  group_by(Path) %>%
  slice_min(order_by = FDR, n=100)
  

datatable(test_pseudo100, filter = "top", rownames = FALSE)
```

## Gene ontology enrichment, top ~100 genes

```{r}
test_pseudo100_genes =
  test_pseudo100 %>%
  select(SYMBOL, Path)

go_ol =  phyperOverlapSets(gaf, 
                           test_pseudo100_genes, length(merged.sce))

write.csv(go_ol$intersecting.genes, file = paste0(results_dir, "/filtered_sex_pseudotime_de_top100_GO_overlaps.csv"), row.names = FALSE) 

scores = go_ol$pval.scores

scores = 
  scores %>%
  dplyr::rename(GOID = "Group1",
                Path = "Group2",
                nGOTermGenes = "nGroup1",
                nPathDEGenes = "nGroup2") %>%
    mutate(FDR = p.adjust(Pval, "BH")) %>%
  select(-Enrichment)

scores = 
  left_join(go_names, scores, by="GOID") %>%
  filter(!is.na(nGOTermGenes)) %>%
  filter(Pval < .01)

datatable(scores, filter = "top", rownames = FALSE)
```

## Genes differentially expressed immediately following the branch point in cluster 1

A more advanced analysis involves looking for differences in expression between paths of a branched trajectory. This is most interesting for cells close to the branch point between two or more paths where the differential expression analysis may highlight genes responsible for the branching event. The general strategy here is to fit one trend to the unique part of each path immediately following the branch point, followed by a comparison of the fits between paths.

These plots show pseudotime in cells in cluster 1 along paths 2 and 8 which separate male and female trajectories. 

### In TSNE space

```{r, fig.height=4, fig.width=8}
combined.sex.sce$TSCAN.15 <- pathStat(pseudo$ordering)[,1]
combined.sex.sce$TSCAN.2 <- pathStat(pseudo$ordering)[,2]
combined.sex.sce$TSCAN.8 <- pathStat(pseudo$ordering)[,3]
combined.sex.sce$TSCAN.6 <- pathStat(pseudo$ordering)[,4]
combined.sex.sce$TSCAN.14 <- pathStat(pseudo$ordering)[,5]
combined.sex.sce$TSCAN.5 <- pathStat(pseudo$ordering)[,6]
combined.sex.sce$TSCAN.12 <- pathStat(pseudo$ordering)[,7]

merged.sex.sce$TSCAN.15 <- pathStat(pseudo$ordering)[,1]
merged.sex.sce$TSCAN.2 <- pathStat(pseudo$ordering)[,2]
merged.sex.sce$TSCAN.8 <- pathStat(pseudo$ordering)[,3]
merged.sex.sce$TSCAN.6 <- pathStat(pseudo$ordering)[,4]
merged.sex.sce$TSCAN.14 <- pathStat(pseudo$ordering)[,5]
merged.sex.sce$TSCAN.5 <- pathStat(pseudo$ordering)[,6]
merged.sex.sce$TSCAN.12 <- pathStat(pseudo$ordering)[,7]

# Subsetting to the desired cluster containing the branch point.
keep <- colLabels(merged.sex.sce) == 1
combined.sex.sce1 <- combined.sex.sce[,keep]
merged.sex.sce1 <- merged.sex.sce[,keep]

# Showing only the lines to/from our cluster of interest.
line.data.sub <- line.data[grepl("^1--", line.data$edge) | grepl("--1$", line.data$edge),]
line.data.sub <- line.data.sub[grepl("3--", line.data.sub$edge) | grepl("4--", line.data.sub$edge),]


ggline <- geom_line(data=line.data.sub, mapping=aes(x=TSNE1, y=TSNE2, group=edge)) 

gridExtra::grid.arrange(
    plotTSNE(merged.sex.sce1, colour_by="TSCAN.2", point_size = .3, text_by="label", text_colour="red") + ggline,
    plotTSNE(merged.sex.sce1, colour_by="TSCAN.8", point_size = .3, text_by="label", text_colour="red") + ggline,
    ncol=2
)
```

### In batch corrected space 

It's hard to interpret this in TSNE space, so we'll plot in batch corrected expression space instead because the clusters and trajectory analysis were actually calculated in batch corrected expression space. 

```{r, fig.height=4, fig.width=8}
line.data.sub_corrected = line.data_corrected[grepl("^1--", line.data_corrected$edge) | grepl("--1$", line.data_corrected$edge),]
line.data.sub_corrected <- line.data.sub_corrected[grepl("3--", line.data.sub_corrected$edge) | grepl("4--", line.data.sub_corrected$edge),]

ggline_corrected <- geom_line(data=line.data.sub_corrected, mapping=aes(x=dim1, y=dim2, group=edge))

gridExtra::grid.arrange(
    plotTSNE(merged.sex.sce1, dimred = "corrected", colour_by="TSCAN.2", point_size = .5, text_by="label", text_colour="red") + ggline_corrected,
    plotTSNE(merged.sex.sce1, dimred = "corrected", colour_by="TSCAN.8", point_size = .5, text_by="label", text_colour="red") + ggline_corrected,
    ncol=2
)
```

```{r}
pseudo_test2 <- testPseudotime(combined.sex.sce1, df=1, pseudotime=combined.sex.sce1$TSCAN.2)
pseudo_test2$SYMBOL <- rowData(combined.sex.sce1)$SYMBOL
#pseudo_test2[order(pseudo_test2$p.value),]
```

```{r}
pseudo_test8 <- testPseudotime(combined.sex.sce1, df=1, pseudotime=combined.sex.sce1$TSCAN.8)
pseudo_test8$SYMBOL <- rowData(combined.sex.sce1)$SYMBOL
#pseudo_test4[order(pseudo_test4$p.value),]
```

We want to find genes that are significant in path 2 and are not significant and/or changing in the opposite direction in path 8 and vise versa. We use the raw p-values to look for non-significant genes in order to increase the stringency of the definition of unique genes in our path.

## Genes uniquely changing on path 2 vs 8

These are saved in `results/filtered_sex_only2_DEgenes.csv`.

```{r}
only2 <- pseudo_test2[which(pseudo_test2$FDR <= 0.05 & 
    (pseudo_test8$p.value >= 0.05 | sign(pseudo_test2$logFC)!=sign(pseudo_test8$logFC))),]

only2 = as.data.frame(only2) 
only2$Gene = rownames(only2) 
only2 = only2[c(4,1,2,3)]

write.csv(only2, file = paste0(results_dir, "/filtered_sex_only2_DEgenes.csv"), row.names = FALSE) 

datatable(only2[order(only2$p.value),], filter="top", rownames = FALSE)
```

### Top 50 heatmap

```{r, fig.height=8, fig.width=6}
on.path2 <- !is.na(combined.sex.sce1$TSCAN.2)

top50 =
  only2 %>%
  slice_min(order_by = FDR, n=50) %>%
  pull(Gene)

plotHeatmap(combined.sex.sce1[,on.path2], 
            order_columns_by="TSCAN.2", 
            colour_columns_by="TSCAN.2", 
            features=top50,
            center=TRUE)
```

### Same genes on path 8. 

```{r, fig.height=8, fig.width=6}
on.path8 <- !is.na(combined.sex.sce1$TSCAN.8)

plotHeatmap(combined.sex.sce1[,on.path8], 
            order_columns_by="TSCAN.8", 
            colour_columns_by="TSCAN.8", 
            features=top50,
            center=TRUE)
```

## Gene ontology enrichment of genes uniquely changing on path 2 vs 8

**Columns**

* `GOID`: Gene ontology ID            
* `GOName`: Gene ontology name    
* `nGOTermGenes`: Number of genes assigned to GO term        
* `Path`: Path and whether the gene is up or down regulated      
* `nPathDEGenes`: Number of genes differentially expressed on the path         
* `nSharedGenes`: Number of shared genes between the GO term and path DE genes   
* `Pval`: p-value          
* `FDR`: p-value corrected for multiple comparisons            
          
These results are save in `results/filtered_sex_only2_gene_GO_pvals.csv`. Overlapping genes are save in `results/filtered_sex_only2_gene_GO_overlaps.csv`

```{r}
gaf = read.delim("/mnt/research/bioinformaticsCore/shared/Genomes/Petromyzon_marinus/GCF_010993605.1_kPetMar1.pri_gene_ontology.gaf", header=F, comment = "!")

gaf <- 
  gaf %>%
  filter(!grepl("NOT", V4)) %>%  
  dplyr::select(V3, V5) %>%      
  dplyr::rename(GOID = "V5",            
                Gene = "V3") %>%
  distinct() 

keep_go = 
  gaf %>%
  group_by(GOID) %>%
  tally() %>%
  filter(n > 10,
         n < 200) %>%
  pull(GOID)

gaf = 
  gaf %>%
  filter(GOID %in% keep_go)

go_names = read.delim("/mnt/research/bioinformaticsCore/shared/Genomes/Petromyzon_marinus/GOs_names.txt", header = F)
colnames(go_names) = c("GOID", "GOName")

only2_genes = 
  only2 %>%
  mutate(Path = case_when(
    logFC > 0 ~ "Only2.Upregulated",
    logFC < 0 ~  "Only2.Downregulated")
    ) %>%
  select(Gene, Path)
  
  
go_ol =  phyperOverlapSets(gaf, 
                           only2_genes, length(merged.sce))

write.csv(go_ol$intersecting.genes, file = paste0(results_dir, "/filtered_sex_only2_gene_GO_overlaps.csv"), row.names = FALSE) 

scores = go_ol$pval.scores

scores = 
  scores %>%
  dplyr::rename(GOID = "Group1",
                Path = "Group2",
                nGOTermGenes = "nGroup1",
                nPathDEGenes = "nGroup2") %>%
    mutate(FDR = p.adjust(Pval, "BH")) %>%
  select(-Enrichment)

scores = 
  left_join(go_names, scores, by="GOID") %>%
  filter(!is.na(nGOTermGenes)) %>%
  filter(Pval < .01)

write.csv(scores, file = paste0(results_dir, "/filtered_sex_only2_gene_GO_pvals.csv"), row.names = FALSE) 

datatable(scores, filter = "top", rownames = FALSE)
```

## Genes uniquely changing on path 8 vs 2

These are saved in `results/filtered_sex_only8_DEgenes.csv`.

```{r}
only8 <- pseudo_test8[which(pseudo_test8$FDR <= 0.05 & 
    (pseudo_test2$p.value >= 0.05 | sign(pseudo_test2$logFC)!=sign(pseudo_test8$logFC))),]

only8 = as.data.frame(only8) 
only8$Gene = rownames(only8) 
only8 = only8[c(4,1,2,3)]

write.csv(only8, file = paste0(results_dir, "/filtered_sex_only8_DEgenes.csv"), row.names = FALSE) 

datatable(only8[order(only8$p.value),], filter="top", rownames = FALSE)
```

### Top 50 heatmap

```{r, fig.height=8, fig.width=6}
on.path8 <- !is.na(combined.sex.sce1$TSCAN.8)

top50 =
  only8 %>%
  slice_min(order_by = FDR, n=50) %>%
  pull(Gene)

plotHeatmap(combined.sex.sce1[,on.path8], 
            order_columns_by="TSCAN.8", 
            colour_columns_by="TSCAN.8", 
            features=top50,
            center=TRUE)
```

### Same genes on path 2. 

There should not be a discernible correlation with pseudotime.  

```{r, fig.height=8, fig.width=6}
plotHeatmap(combined.sex.sce1[,on.path2], 
            order_columns_by="TSCAN.2", 
            colour_columns_by="TSCAN.2", 
            features=top50,
            center=TRUE)
```

## Gene ontology enrichment of genes uniquely changing on path 8 vs 2

*Columns*

* `GOID`: Gene ontology ID            
* `GOName`: Gene ontology name    
* `nGOTermGenes`: Number of genes assigned to GO term        
* `Path`: Path and whether the gene is up or down regulated      
* `nPathDEGenes`: Number of genes differentially expressed on the path         
* `nSharedGenes`: Number of shared genes between the GO term and path DE genes   
* `Pval`: p-value          
* `FDR`: p-value corrected for multiple comparisons            

These results are save in `results/filtered_sex_only8_gene_GO_pvals.csv`. Overlapping genes are save in `results/filtered_sex_only8_gene_GO_overlaps.csv`

```{r}
only8_genes = 
  only8 %>%
  mutate(Path = case_when(
    logFC > 0 ~ "Only8.Upregulated",
    logFC < 0 ~  "Only8.Downregulated")
    ) %>%
  select(Gene, Path)
  
  
go_ol =  phyperOverlapSets(gaf, 
                           only8_genes, length(merged.sce))

write.csv(go_ol$intersecting.genes, file = paste0(results_dir, "/filtered_sex_only8_gene_GO_overlaps.csv"), row.names = FALSE) 

scores = go_ol$pval.scores

scores = 
  scores %>%
  dplyr::rename(GOID = "Group1",
                Path = "Group2",
                nGOTermGenes = "nGroup1",
                nOnly8DEGenes = "nGroup2") %>%
    mutate(FDR = p.adjust(Pval, "BH")) %>%
  select(-Enrichment)

scores = 
  left_join(go_names, scores, by="GOID") %>%
  filter(!is.na(nGOTermGenes)) %>%
  filter(Pval < .01)

write.csv(scores, file = paste0(results_dir, "/filtered_sex_only8_gene_GO_pvals.csv"), row.names = FALSE) 

datatable(scores, filter = "top", rownames = FALSE)
```

# Session info
```{r}
sessionInfo()
```

